### 2023-08-22 report/esecfse24.tex
feat(report): init report

\documentclass[acmsmall]{acmart}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

\begin{document}

\title{We don't know yet...}

\author{Anon}

\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

\maketitle

\section{Introduction} % (fold)
\label{sec:Introduction}
Hello, this is the introduction of the report.
% section Introduction (end)

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliography}

\end{document}

### 2023-09-19 report/esecfse24.tex
feat(report): init methodology

@@
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen,review,anonymous]{acmart}
 
@@
 %%\acmSubmissionID{123-A56-BU3}
+\usepackage{multirow}
 
@@
 
-\title{We don't know yet...}
+\title{Bridging the Gap Between Visual and Analytical Machine Learning Testing}
 
@@
 
-\section{Introduction} % (fold)
-\label{sec:Introduction}
-Hello, this is the introduction of the report.
-% section Introduction (end)
+\section{Introduction}\label{sec:intro}
+% what problem are we trying to solve?
+% what is our motivation here?
 
+\section{Preliminaries}\label{sec:prelim}
+\subsection{Related Work}\label{sec:related}
+\subsection{Background}\label{sec:background}
+\subsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
+% we need to explain the json structure of jupyter notebooks, we
+% reference this in the methodology to explain the github query.
 \bibliographystyle{ACM-Reference-Format}
@@
 
+\section{Methodology}\label{sec:method}
+% TODO is there a scientific term in SE research to explain reading
+% code in detail?
+This section presents the methodology used to conduct the empirical
+study presented in Section~\ref{sec:intro}. Figure~\ref{fig:method}
+presents an overview of all relevant steps used in this study. The
+empirical analysis comprises of three distinct phases namely:
+\textit{the data collection phase}, \textit{identification of related
+visualisation and assertion pairs} and finally a \textit{complete
+analysis of the notebooks} to identify the final set of 29
+visualisation and assertion pairs.
+
+The yellow boxes in the data collection phase represent filters that
+were applied to the 54K notebooks obtained from Gihub, to obtain the
+final sample size of 1.7K notebooks used in the empirical study. The
+red boxes represent the exclusion criteria used during phase two to
+remove notebooks note relevant for this study. Finally, the green
+boxes in phase 3 represent the manual analysis conducted to identify
+the final 29 visualisation and assertion pairs.
+
+Boxes appearning next to one-another imply that they were applied
+using an "OR" operator. While boxes appearning below one-another imply
+an "AND" operator. Text in double quotes represent the string pattern
+used in the search query.
+
+More details regarding each phase of the methodology is presented below.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{method.pdf}
+  \caption{Methodology used to collect and analyse Jupyter notebooks
+    in this study.}
+  \label{fig:method}
+\end{figure}
+
+\subsection{Data Collection}\label{sec:data-collect}
+
+We mined public repositories from Github to collect Jupyter
+notebooks---files with the \texttt{.ipynb} extension---that contain
+the keyword ``assert'' in them.
+
+% TODO is it clear what I am trying to say here?
+% FIXME is "testing statement" the right terminology?
+
+We intentionally keep the search criteria on Github broad by only
+checking for the appropriate filetype and the presence of the
+``assert'' keyword. This is because, the ``assert'' keyword is used in
+python to define tests. Additionally, popular python libraries which
+provide a testing interface (such as nose, unittest and numpy),
+contain the same keyword in their API. Thus, our search criteria
+captures a large initial sample of $54,070$ notebooks with a variety
+of testing statements. Our approach also prevents the need to craft a
+custom search query based on an exhaustive list of keywords that
+appear in all python testing libraries.
+
+\begin{table}
+  \centering
+  \caption{Filters used during data collection phase.}
+  \begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
+    \toprule
+    \textbf{ID} &
+    \textbf{Filter} &
+    \textbf{Rationale}\\
+    \midrule
+    \emph{\textbf{F1}} &
+    Files with extension \texttt{.ipynb} &
+    This translates to the following Github search query:
+    \texttt{filename:"Jupyter Notebook"}.\\
+    \emph{\textbf{F2}} &
+    Notebooks with \texttt{"assert"} keyword &
+    \texttt{"assert"} is used in python to define tests. Additionally,
+    it also appears in other python libraries which provide their own
+    testing interface.\\
+    \emph{\textbf{F2}} &
+    Notebooks with popular data science or machine learning libraries &
+    The list of python libraries is derived by performing a web search
+    along with the author's prior knowledge and expertise in ML. We
+    combine the patterns using an \texttt{"OR"} operator. The final
+    regex pattern is the following:
+    \texttt{"tensorflow"|"dask"|"torch"|"sklearn"
+    |"scipy"|"numpy"|"pandas"}.\\
+    \emph{\textbf{F3}} &
+    Notebooks that contain visualisations &
+    Section~\ref{sec:nbformat} describes the internal json structure
+    of Jupyter notebooks. Visualisations are stored as binary strings
+    under the \texttt{"image/png"} field.\\
+    \emph{\textbf{F4}} &
+    Notebooks where the assertion appears below the visualisation &
+    The code cell containing the assertion must be defined below the
+    cell that produces the visualisation. Any number of markdown cells
+    may appear between the visualisation and assertion cells. See
+    Section~\ref{sec:cell-arrangement} for detailed explanation.\\
+    \emph{\textbf{F5}} &
+    Notebooks where the assertion cell is not longer than 10 lines &
+    The code cell containing the assertion should not be longer than
+    10 lines. The threshold for max lines was derived based on our
+    observations from a random sample of 50 notebooks. See
+    Section~\ref{sec:cell-arrangement} for detailed explanation.\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:filters}
+\end{table}
+
+Table~\ref{tab:filters} summarises the filters used during the data
+collection phase to obtain the final sample of 1.7K notebooks from the
+initial sample of 54K notebooks mined from Github. The table provides
+a short description of the filter along with our rationale for using
+them.
+
+% TODO don't mention quantity of random sample; then we need to
+% explain/defend this choice
+We used an iterative approach to determine the effectiveness of each
+filter that was used. We applied each filter in the same order as
+presented in Table~\ref{tab:filters}, starting with F1. We randomly
+sampled 50 notebooks from the population of notebooks obtained after
+applying the filter. We then conduct the qualitative assessment
+presented in Section~\ref{REFME} and observe the number of related
+visualisation and assertion pairs. In each new iteration, we
+progressively add the other filters and include the filter if the
+number of relevant visualisation and assertion pairs increases.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\textwidth]{nb-structure.pdf}
+  \caption{Arrangement of code cells containing assertion and
+    visualisation.}
+  \label{fig:cell-arrangement}
+\end{figure}
+
+For RQ1, we are interested in finding empirical evidence of
+visualisations and corresponding assertions used in the wild. To
+answer this question, we further reduce our notebook sample size by
+assuming a specific order in which the visualisation and assertion
+code cells appear in the notebooks. Figure~\ref{fig:cell-arrangement}
+presents a visual representation of various arrangements of code cells
+that may occur in notebooks. Note that the visual representation omits
+markdown cells since we only consider the order of code cells in our
+analysis. In the actual notebook however, one or more markdown cells
+may be present between the visualisation and assertion cells.
+
+As explained above, the selected arrangement of code cells was
+discovered by qualitatively analysing a random sample of notebooks
+with the code cell arrangements presented in
+Figure~\ref{fig:cell-arrangement}. We observed the signal-to-noise
+ratio for each cell arrangement. And picked the arrangement with the
+least signal-to-noise ratio. Table~\ref{tab:cell-arrangement}
+summarises our observations from the various cell arrangements.
+
+\begin{table}
+  \centering
+  \caption{Observations from random samples of notebooks with various
+  cell arrangement.}
+  \begin{tabular}{p{0.4\textwidth} p{0.4\textwidth}}
+    \toprule
+    \emph{\textbf{Cell arrangement}} &
+    \emph{\textbf{Observations}}\\
+    \midrule
+    Assertion above visualisation cell &
+    The assertions in this arrangement were typically not related to
+    the visualisation below. Often, a markdown cell present between
+    the assertion and visualisation code cells, would mark the
+    begining of a new section in the notebook. In other cases, the
+    assertions were typically pre-condition checks to ensure that the
+    visualisation code would not encounter any errors.\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:cell-arrangement}
+\end{table}
+
+We assume, that the visualisation and assertion pairs of interest, are
+more likely to occur in notebooks where the assertion follow
+
+\subsection{Identifying Related Visualisation and Assertion
+Pairs}\ref{sec:identify-related-pairs}
+% at some point, I think we will have to explain what a
+% pre/post-condition check is; then we can ellaborate on our
+% experiments with the strictness and proximity of the assert
+% statement
+% - we have to justify that most of the examples we found when the
+% assertion was in the same cell, were pre/post-conditions; whereas
+% looking at assertions in the next code cell was more fruitful
+
+\begin{table}
+  \centering
+  \caption{Exclusion criterion used to reduce sample size of notebooks
+    prior to conducting manual analysis to identify related
+    visualisation and assertion pairs.}
+  \begin{tabular}{l p{0.3\linewidth} p{0.3\linewidth}}
+    \toprule
+    \textbf{ID} &
+    \textbf{Exclusion Criteria(EC)} &
+    \textbf{Rationale}\\
+    \midrule
+    \textbf{EC1} &
+    NB does not contain popular data science or machine
+    learning libraries &
+    \texttt{"pandas|numpy|scipy|sklearn |torch|dask|tensorflow"}\\
+    \textbf{EC2} &
+    NB does not contain any visualisations &
+    TODO\\
+    \textbf{EC3} &
+    NB conforms to a strict structure as explained in~\ref{REFME} &
+    TODO\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:pre-exclusion-criteria}
+\end{table}
+\subsection{Deeper Analysis of Notebooks}\ref{sec:deep-dive}
+\section{Threats to Validity}\ref{sec:threats}
+% defend use of github as source for notebooks:
+% - why did kaggle not work (refer to notes)
+% - why did the existing replication packages not work?
+
+% motivation for keeping the search on github broad:
+% - not only does it pick up python assert statements, but also flags
+% - notebooks that use other testing libraries or modules that provide a
+% - testing interface. This is a good compromise between doing an
+% - exhaustive search of all testing method names in the python
+% - ecosystem and then doing a search for each of those patterns
+
 \end{document}

### 2023-09-19 presentation/siesta-2023.qmd
feat(presentation): init siesta 2023 presentation

---
title: "Bridging the Gap between Visual and Analytical ML Testing"
subtitle: "SIESTA 2023"
date: 2023-09-11
mainfont: "Iosevka Aile"
monofont: "Iosevka"
affiliations:
  - id: tud
    name: "Delft University of Technology"
author:
  - name: "[***Aru***]{.underline}moy Shome"
    affiliation:
      - ref: tud
  - name: "Luis Cruz"
    affiliation:
      - ref: tud
  - name: "Arie van Deursen"
    affiliation:
      - ref: tud
format:
  revealjs:
    theme: [default, custom.scss]
license: "CC BY"
copyright:
  holder: "Arumoy Shome"
  year: 2023
bibliography: siesta-2023.bib
---

## 🤫🤫🤫🤫

::: {.callout-tip title="Towards Understanding Machine Learning Testing in Practice [@shome2023towards]"}
:::

::: {.callout-important title="Discretion is advised!"}
~~Some~~Most parts of this presentation are unpublished!
:::

## ML Testing

::: {.callout-tip title="Machine Learning Testing: Survey, Landscapes, and Horizons[@zhang2020machine]"}
:::

::: {.fragment}
![ML lifecycle](ml-lifecycle.svg){#fig-ml-lifecycle}
:::
## From Implicit Expectations...

:::: {.columns}

::: {.column width="60%"}
![](histogram.jpg){width="50%"}

![](qqplot.png){width="50%"}
:::

::: {.column width="40%" .fragment}
+ Tend to be messy/quick/hacks
+ Exploratory, data understanding/sense making
+ Domain expertise and knowledge of practitioner *embedded* in visualisations
:::
::::
## ...To Explicit Tests
:::: {.columns}
::: {.column width="40%"}
+ Verifiable across entire ML lifecycle
+ Scales across org/team change
+ Scales to large-scale production systems and big data
+ Eg. *Shapiro-Wilk Test* of normality
:::

::: {.column width="60%" .fragment}
```python
import numpy as np
from scipy import stats

x = stats.norm.rvs(size=100)
shapiro_test = stats.shapiro(x)

assert shapiro_test.pvalue <= 0.05
```
:::
::::

## The Hunt for Data Properties {.smaller}

:::: {.columns}
::: {.column width="50%"}
**Literature**

+ Review of conferences and journals focusing on visualisations: HCI,
SE, VIZ, ML & DMM
+ Review of grey literature
:::

::: {.column width="50%"}
**Computational NBs**

+ Kaggle
+ Github
+ @pimentel2019large & @quaranta2021kgtorrent
:::
::::

## Here Come the Numbers {.smaller}

::: {.callout-important title="1.7K"}
Notebooks from Github with an `assert` statement
:::

::: {.callout-warning title="337"}
Notebooks with related `assert` & visualisation pairs
:::

::: {.callout-note title="90"}
Notebooks analysed in full depth
:::

::: {.callout-tip title="22"}
Examples of `assert` & visualisation pairs
:::

# Results

## Area under ROC curve {.smaller}

:::: {.columns}
::: {.column}
![](select-09.png)
:::

::: {.column}
```python
assert roc_auc_score(
    y_test, wv_model.predict_proba(X_test_wv)[:, 1]
) > 0.92,
"something's wrong with your features"
```
:::
::::
## Loss, convergence {.smaller}

:::: {.columns}
::: {.column}
![Type 1](select-12.png){#fig-left}
:::

::: {.column}
![Type 2](select-267.png){#fig-right}
:::
::::

:::: {.columns}
:::{.column}
```python
assert np.mean(history[:10]) >np.mean(history[-10:]),
"RNN didn't converge"
```
:::

::: {.column}
```python
# assert we actually descended at each step
for i in range(1, n_steps):
    assert (
        loss_array_true_grad[i] - loss_array_true_grad[i-1]
    ) <= 0
```
:::
::::

::: {.callout-warning title="Smell"}
The assertion for @fig-right is brittle compared to @fig-left.
:::
## Decision boundary & accuracy of classifier {.smaller}

:::: {.columns}
::: {.column}
![](select-04.png)
:::

::: {.column}
```python
assert accuracy_score(y_test, pred) >0.95
```
:::
::::

::: {.callout-note title="Open question"}
Some visualisations are more complex compared to their counter-part
assertion. Is it because we don't yet know how to write better
assertions for such visualisations?
:::
## Hyper-parameter grid search

:::: {.columns}
::: {.column}
![](select-290.png){width="60%"}
:::

::: {.column}
::: {.callout-note title="Qualitative analysis of hyper parameters"}
The author is visually checking all combinations of two hyper
parameters `C` and `gamma` to identify the combinations that cause the
model to {under,over}fit.
:::
:::
::::

# Our objective is to [*recommend*]{.underline} analytical tests automatically {.r-text-fit}

## Thank You

::: {layout-nrow="2"}
![](qr-cat.jpeg)

![](qr-lugano.jpeg)

![](qr-illustrate.jpeg)

![](qr-picaso.jpeg)
:::
## Refs {.smaller}

### 2023-09-19 catalogue/catalogue.qmd
feat(catalogue): init catalogue

---
title: Visualisation, assertion catalogue
author: Arumoy Shome
mainfont: "Iosevka Aile"
monofont: "Iosevka"
monofontoptions:
  - Scale=0.9
format: pdf
---

## (04) SVM decision boundary, accuracy
![](select-04.png)

```python
assert accuracy_score(y_test, pred) > 0.95
```

## (05) input image, accuracy
![](select-05.png)

```python
assert accuracy > 0.85, "Poor accuracy {:.3f}".format(accuracy)
```

## (09) roc, auc
![](select-09.png)

```python
assert roc_auc_score(y_test, wv_model.predict_proba(X_test_wv)[:, 1]) > 0.92,
"something's wrong with your features"
```

## (12) loss, convergence
![](select-12.png)

```python
assert np.mean(history[:10]) >np.mean(history[-10:]), "RNN didn't converge"
```

::: {.callout-note}
This is the most common visualisation,assertion pair.
:::

## (13) lineplot, correlation
![](select-13.png)

```python
assert answer_nine() >= -1. and answer_nine() <= 1.,
"Q9: A valid correlation should between -1 to 1!"
```

# (35) input image, `predict_proba`

![](select-35a.png)

```python
np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
```

![](select-35b.png)

```python
np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
```

::: {.callout-warning title="Smell"}
How did the author identify the probability threshold? Perhaps
comparision with ge/le is better?
:::

# (47) log-scale lineplot, coefficient & intercept

![](select-47.png)

```python
assert l.coef_ < 0, "Curve must be decreasing"
assert l.intercept_ > 1, "Curve must start higher"
assert l.score(np.log(X).reshape((-1,1)), np.log(values)) > 0.9, "Unusually large error."
```

::: {.callout-caution}
We need to revisit this one!
:::

# (61) various image rescaling techniques

::: {layout-ncol=3}
![Original input image](select-61a.png)

![Rescaled image](select-61b.png)

![Decomposed image](select-61c.png)
:::

# (63) visual and analytical assertion on mean recall

![](select-63.png)

```python
assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. "\
                                                  "You may need to change vectorizer model for that."
```

# (65) Scaling and Normalization

![](select-65a.png)

```python
assert scaled_data[scaled_data.argmax()] <= 1
assert scaled_data[scaled_data.argmin()] >= 0
```

::: {.callout-warning title="Smell"}
Perhaps the visualisation is redundant?
:::

![](select-65b.png)

```python
assert normalized_data[normalized_data.argmax()] < 10
assert normalized_data[normalized_data.argmin()] > -7
```

::: {.callout-warning title="Smell"}
How did the author identify the thresholds?
:::

::: {.callout-tip title="Visual testing"}
This is the most trivial example of visual testing - checking the
distribution of features.
:::

# (78) confusion matrix, false positive/negative

![](select-78.pdf){width="50%"}

```python
assert conf_mat.shape == (2,2)
assert conf_mat[0][1] < 7
assert conf_mat[1][0] < 3
```

::: {.callout-warning title="Smell"}
The values should be divided by $n$ (total number of observations) to
account for bias in the dataset.
:::

# (82) DNN mean reward history

![](select-82.png)

```python
assert np.mean(mean_rw_history[-10:]) > 10.
```

# (86) lolopy performance testing

![](select-86.png)

```python
lolopy_timing = interp1d(results['n'], results['lolopy_train'])
lolo_timing = interp1d(results['n'], results['scala_train'])
slowdown = lolopy_timing(100) / lolo_timing(100)
print('Training slowdown: {:.2f}'.format(slowdown))
assert slowdown < 2
```

::: {.callout-caution}
This is more devops/mlops imo, do we include in this our current
paper?
:::

# (88) histograms for outlier detection

![](select-88a.png)

::: {.callout-tip title="Visual testing"}
A very nice example of a visualisation to determine/understand the
property of the underlying dataset. Check the NB for more details.
:::

![](select-88b.png)

::: {.callout-warning title="Smell"}
The author identifies outliers in the dataset using histograms.
Typically a boxplot is better for outlier detection?
:::

# (110) data property from domain expertise

![](select-110.png)

```python
# In the plot it looks like there are no exceptions, but let's query
the dataset to be 100% sure
days_with_more_than_6_entries_per_day = df_daily_data.groupby('datum').filter(lambda x: len(x) > 6).index

assert len(days_with_more_than_6_entries_per_day) == 0
```

# (117) time-series spot checking

::: {layout-ncol=2}
![Spot checking $y$, $x$](select-117a.png){#fig-117a}

![Check error of prediction](select-117b.png){#fig-117b}
:::

In @fig-117a the author spot checks the values of $y$ at specific values
of $x$.

```python
# y_t[i] are the number of cases on the i-th day
assert y_t[0] == 85
assert y_t[10] == 1452
assert y_t[800] == 65032
```

::: {.callout-warning title="Smell"}
It is difficult to identify the exact value of $y$ from a
visualisation. When the data changes, the assert will fail right away.
:::

In @fig-117b the author evalues the performance of a predictor using a
custom measure/metric. Performance also visually checked using a
lineplot.

```python
# Result of the wfc function for the best choice of s
assert np.isclose(best_wfv, 7685.546728971963, 0.001)
```

# (143) input image, `predict_proba`

::: {layout-nrow=2}
![](select-143a.png)

![](select-143b.png)

![](select-143c.png)

![](select-143d.png)
:::

```python
assert 0 <= predict_image(face_model, TP_URL).first().prediction <= 1
assert 0 <= predict_image(face_model, TN_URL).first().prediction <= 1 
assert 0 <= predict_image(face_model, FP_URL).first().prediction <= 1
assert 0 <= predict_image(face_model, FN_URL).first().prediction <= 1
```

::: {.callout-warning title="Smell"}
The assertions are on the theoretical limits of probability.
:::

![](select-143g.png)

::: {.callout-tip title="Visual Testing"}
The author is visualising the decision boundary of a MLP along with the
accuracy of the model. 
:::

# (152) visual & statistical distribution check

![](select-152a.png)

```python
for feature in range(data_transformed.shape[1]):
    assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
```

![](select-152b.png)

::: {.callout-tip title="Visual Testing"}
We need to revisit this, but I think the author is trying to validate if the
output generated by the GAN is similar to the original data.
:::

# (153) bootstrap estimate of mean & variance

::: {layout-ncol=2}
![Mean](select-153a.png){#fig-153a}

![Variance](select-153b.png){#fig-153b}
:::

The assertions for Figure @fig-153a and @fig-153b are the following.

```python
assert 80 < boot_mean_mean < 81
assert 0.3 < boot_var_mean < 0.5
```

```python
assert 7 < boot_mean_std < 10
assert 0.23 < boot_var_std < 0.29
```

::: {.callout-warning title="Smell"}
While the mean of the sample can be identified from the histogram, I don't
think the same holds for the variance?
:::

# (165) pie chat

![](select-165.png)

```python
assert_equal(num_passengers_by_sex.index.tolist(), ["female", "male"])
```

# (172) image augmentation

::: {layout-ncol=2}
![Original image](select-172a.png)

![augmented image](select-172b.png)
:::

```python
assert not np.allclose(x_aug, x_original)
```

# (202) input image and predictions

![](select-202a.png)

::: {.callout-tip title="Visual Testing"}
This is a fairly common example of visual testing where the image input along
with the prediction of the model is visualised.

Although we can use various performance metrics (accuracy, precision, recall,
etc), this form of qualitative testing is still prominent when working with
images.
:::

# (204) logistic regression

![](select-204a.png)

::: {.callout-tip title="Visual Testing"}
I think the author is trying to visualise the results of the optimization here? Need second opinion.
:::

![](select-204b.png)

```python
ans4 = compute_loss(X_expanded, y, w)
np.testing.assert_almost_equal(ans4, 0.3042764)
```

::: {.callout-warning title="Smell"}
Better to use ge/le comparison rather than hardcoded threshold?
:::

# (210) great expectations

![](select-210.pdf)

```python
result = solar.expect_column_values_to_be_between(field_map["date"], 
                                                    solar[field_map["date"]].min(), 
                                                    pd.Timestamp.now(),
                                                    parse_strings_as_datetimes=True)
```

# (228) linear approximation using domain knowledge

::: {layout-ncol=2}
![](select-228a.png){#fig-228a}

![](select-228b.png){#fig-228b}
:::

For Figure @fig-228a.

```python
assert np.allclose(y_approx.mean(), 0)
assert np.allclose(y_approx.std(), 0.6721265191518054)
```

For Figure @fig-228b.

```python
assert np.allclose(y_long.mean(), 0.2742076173424)
```

::: {.callout-warning title="Smell"}
Its not possible to identify the mean of a distribution from the visualisation to such precision. I think a ge/le comparison is better.
:::

# (267) loss is decreasing

![](select-267.png)

```python
# assert we actually descended at each step
for i in range(1, n_steps): 
    assert (loss_array_true_grad[i] - loss_array_true_grad[i-1]) <= 0
```

::: {.callout-warning title="Smell"}
The author checks that the loss is decreasing at each step/epoch. How do we ensure that we are not stuck in a local minimum with this assertion style? Perhaps its better to keep it general and only check if the last $n$ observations is lower than the first $n$ observations?
:::

# (275) loss, accuracy

![Loss](select-275.png)

```python
acc = compute_test_accuracy(model)
assert acc>0.96, "Bidirectional RNNs are better than this!"
```

::: {.callout-caution title="Luis"}
The visualisation is of the loss, but the assertion is on the accuracy. They feel "kind of" related? What to do here?
:::

# (290) hyper-parameters

![](select-290.png)

::: {.callout-tip title="Visual testing"}
Qualitative assessment of hyper-paramer configuration that causes
over-fitting in model.
:::

# (332) recurrent neural networks

![True vs. predicted labels](select-332a.png){#fig-332a}

```python
r2_gru = r2_score(y_test, y_pred)
assert r2_gru > 0.6
```

::: {.callout-note}
The $R^2$ or coefficient of determination is derived from the predicted and actual labels.
:::

### 2023-09-19 report/esecfse24.tex
feat(report): init methodology figures

@@
     Assertion above visualisation cell &
-    The assertions in this arrangement were typically not related to
-    the visualisation below. Often, a markdown cell present between
-    the assertion and visualisation code cells, would mark the
-    begining of a new section in the notebook. In other cases, the
-    assertions were typically pre-condition checks to ensure that the
-    visualisation code would not encounter any errors.\\
+    The assertions in this arrangement were typically not related to the visualisation below.
+Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
+In other cases, the assertions were typically pre-condition checks to ensure that the visualisation code would not encounter any errors.
+Pre-condition checks typically include checking the shape of features that are used in the visualisation.
+Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.\\
+    Assertion in the visualisation cell &
+    There are two possible scenarios here:
+    \begin{enumerate}
+    \item[2.1]{This is the first one.}
+    \item[2.2]{This is the second one.}
+    \end{enumerate}\\
+
     \bottomrule

### 2023-09-21 report/esecfse24.tex
feat(report): 2023-09-21 10:32 updates

@@
 %%\acmSubmissionID{123-A56-BU3}
-\usepackage{multirow}
+
+\usepackage{listings}
+\usepackage{graphicx}
 
@@
 
+% NOTE our methodology of finding visualisation and assertion that are related is a (pretty good) contribution.
+
 \section{Preliminaries}\label{sec:prelim}
@@
 \subsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
-% we need to explain the json structure of jupyter notebooks, we
-% reference this in the methodology to explain the github query.
+% TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
 \bibliographystyle{ACM-Reference-Format}
@@
 \section{Methodology}\label{sec:method}
-% TODO is there a scientific term in SE research to explain reading
-% code in detail?
+% TODO is there a scientific term in SE research to explain reading code in detail?
+% NOTE: we looked for symmantic similarity between the visualisation and assertion code
 This section presents the methodology used to conduct the empirical
@@
     Notebooks that contain visualisations &
-    Section~\ref{sec:nbformat} describes the internal json structure
+    Section~\ref{sec:nbformat} describes the internal son structure
     of Jupyter notebooks. Visualisations are stored as binary strings
@@
 
-% TODO don't mention quantity of random sample; then we need to
-% explain/defend this choice
+% TODO don't mention quantity of random sample; then we need to explain/defend this choice
 We used an iterative approach to determine the effectiveness of each
@@
 
+% TODO maybe use colors from red-green to denote arrangements that produced better results? (instead of binary)
 \begin{figure}
@@
 
-For RQ1, we are interested in finding empirical evidence of
-visualisations and corresponding assertions used in the wild. To
-answer this question, we further reduce our notebook sample size by
-assuming a specific order in which the visualisation and assertion
-code cells appear in the notebooks. Figure~\ref{fig:cell-arrangement}
-presents a visual representation of various arrangements of code cells
-that may occur in notebooks. Note that the visual representation omits
-markdown cells since we only consider the order of code cells in our
-analysis. In the actual notebook however, one or more markdown cells
-may be present between the visualisation and assertion cells.
+For RQ1, we are interested in finding empirical evidence of visualisations and corresponding assertions used in the wild.
+To answer this question, we further reduce our notebook sample size by assuming a specific order in which the visualisation and assertion code cells appear in the notebooks.
+Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks.
+Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis.
+In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
 
-As explained above, the selected arrangement of code cells was
-discovered by qualitatively analysing a random sample of notebooks
-with the code cell arrangements presented in
-Figure~\ref{fig:cell-arrangement}. We observed the signal-to-noise
-ratio for each cell arrangement. And picked the arrangement with the
-least signal-to-noise ratio. Table~\ref{tab:cell-arrangement}
-summarises our observations from the various cell arrangements.
+As explained above, the selected arrangement of code cells was discovered by qualitatively analysing a random sample of notebooks with the code cell arrangements presented in Figure~\ref{fig:cell-arrangement}.
+We observed the signal-to-noise ratio for each cell arrangement.
+And picked the arrangement with the least signal-to-noise ratio.
+Table~\ref{tab:cell-arrangement} summarises our observations from the various cell arrangements.
 
+% TODO give more examples (from our data) of what a pre/post condition check is
 \begin{table}
@@
   cell arrangement.}
-  \begin{tabular}{p{0.4\textwidth} p{0.4\textwidth}}
+  \begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
     \toprule
+    \emph{\textbf{ID}}&
     \emph{\textbf{Cell arrangement}} &
@@
     \midrule
+    1 &
     Assertion above visualisation cell &
@@
 Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.\\
-    Assertion in the visualisation cell &
-    There are two possible scenarios here:
-    \begin{enumerate}
-    \item[2.1]{This is the first one.}
-    \item[2.2]{This is the second one.}
-    \end{enumerate}\\
-
+    2.1 &
+    Assertion in the visualisation cell: assertion above visualisation code &
+    Our observations were similar to that of arrangement 1. the assertions defined above the visualistion code were more likely to be pre-condition checks.\\
+    2.2 &
+    Assertion in the visualisation cell: assertion below visualisation code &
+    We found several examples where the assertion was related to the visualisation.
+Most often however, the assertions were post-condition checks to ensure the correctness of the visualisation itself.
+The signal-to-noise ratio in this arrangement was also higher compared to arrangement 3.2.\\
+    3.1 &
+    Assertion below visualisation cell: assertion cell more than 10 lines &
+    The assertions in this arrangement were frequently related to the visualisation above.
+In some cases, a markdown cell between the visualisation and assertion marked the begining of a new section in the notebook indicating that the visualisation and assertion were not related to each other.
+We also observed instances where the assertion cell defined a new class or several helper methods that were not related to the visualisation above.
+This motivated our decision to only flag notebooks where the assertion cell was no longer than 10 lines.\\
+    3.2 &
+    Assertion below visualisation cell: assertion cell has max 10 lines &
+    This arrangement had the highest signal-to-noise ratio.
+We observed that most notebook authors followed the convension of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
+The visualisations were often followed by another markdown cell to record the author's observations.
+And finally, the observations were translated to a analytical test using \texttt{``assert''} or other testing methods from external libraries.\\
     \bottomrule
@@
 
-We assume, that the visualisation and assertion pairs of interest, are
-more likely to occur in notebooks where the assertion follow
+\subsection{Identifying Related Visualisation and Assertion
+Pairs}\label{sec:identify-related-pairs}
 
-\subsection{Identifying Related Visualisation and Assertion
-Pairs}\ref{sec:identify-related-pairs}
-% at some point, I think we will have to explain what a
-% pre/post-condition check is; then we can ellaborate on our
-% experiments with the strictness and proximity of the assert
-% statement
-% - we have to justify that most of the examples we found when the
-% assertion was in the same cell, were pre/post-conditions; whereas
-% looking at assertions in the next code cell was more fruitful
+% TODO the numbers don't line up, because the annotations were not perfect. We 161 instances where I did not SELECT the notebook but also did not assign it an EC. How do we handle this?
+% NOTE move the 161 with the 337, then we can say we had 500+ notebooks and we cannot analyse them all in depth...
+The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
+We analysed all 1.7K notebooks manually to further identify 337 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
+Table~\ref{tab:exclusion-criteria} presents the exclusion criteria used in this phase of the analysis, along with the number of notebooks that were excluded based on each criteria.
 
@@
   \centering
-  \caption{Exclusion criterion used to reduce sample size of notebooks
-    prior to conducting manual analysis to identify related
-    visualisation and assertion pairs.}
-  \begin{tabular}{l p{0.3\linewidth} p{0.3\linewidth}}
+  \caption{Exclusion criteria used during phase 2 to identify related visualisation and assertion pairs.}
+  \begin{tabular}{l p{0.3\textwidth} p{0.4\textwidth} p{0.1\textwidth}}
     \toprule
     \textbf{ID} &
-    \textbf{Exclusion Criteria(EC)} &
-    \textbf{Rationale}\\
+    \textbf{Exclusion Criteria} &
+    \textbf{Rationale} &
+    \textbf{Num. excluded NBs}\\
     \midrule
     \textbf{EC1} &
-    NB does not contain popular data science or machine
-    learning libraries &
-    \texttt{"pandas|numpy|scipy|sklearn |torch|dask|tensorflow"}\\
+    Notebook not written in English &
+    % TODO find a better way to phrase this?
+    We excluded notebooks that were not authored in English since that is the primary language of all authors of this paper. &
+    159\\
     \textbf{EC2} &
-    NB does not contain any visualisations &
-    TODO\\
+    Notebook does not contain a trained ML model &
+    % TODO need to explain this better, we have notebooks with diverse python libraries; different ML models work differently so we need to understand the maths behind it.
+    We excluded notebooks that did not train an ML model.
+We found several notebooks that were on topics related to ML such as linear algebra, optimization and loss functions to name a few.
+However, identifying how the visualisation and assertion pairs can be applied to a production ML system, requires a deep understanding of several python libraries and the inner workings of various ML models.
+This was deemed beyond the scope of this research project.
+&
+    383\\
+    \textbf{EC2} &
+    Notebook contains incorrect code &
+    % TODO our filter excludes NBs without images, we have to defend why this EC is still required: jupyter adds the "image/png" field regardless of error in code
+    We excluded notebooks that contained incorrect code that did not produce a visualisation. We also excluded notebooks that used \texttt{assert} statements to stop execution of a code cell. &
+    31\\
     \textbf{EC3} &
-    NB conforms to a strict structure as explained in~\ref{REFME} &
-    TODO\\
+    % TODO is there technical terminology for what we did to identify if the visualisation & assertion are related to one another?
+    % TODO I think we need a brief paragraph on how we prompted ChatGPT, this stuff probably needs to go outside of a table!
+    The visualisation and assertion are not related &
+    We excluded notebooks where the visualisation and assertion pairs were unrelated.
+We manually analyse the source code of the visualisation and the assertion to determine if they are related to one another.
+In some instances, the tracing was not straightforward or the code was not well written.
+In such cases, we used one-shot prompt engineering with ChatGPT 4.0 model to derive the link between the visualisation and the assertion.
+&
+    440\\
+    \textbf{EC4} &
+    The assertion is a pre or post-condition check &
+    % TODO we have to give examples of pre/post condition checks. I think this content needs to be outside of a table!
+    We excluded &
+    200\\
     \bottomrule
   \end{tabular}
-  \label{tab:pre-exclusion-criteria}
+  \label{tab:exclusion-criteria}
 \end{table}
+
 \subsection{Deeper Analysis of Notebooks}\ref{sec:deep-dive}
-\section{Threats to Validity}\ref{sec:threats}
+% describe the steps used for the deeper analysis of NBs
+% - we read the NBs in top-down fashion
+% - reading the introductory md cell at the top often gives the context for the contents of the NB (eg. what type of ML model? which part of the ML lifecycle is this relevant for?)
+% - reading the md cell above visualisation to understand context/motivation of visualisation
+% - reading the source code that produces the visualisation
+% - reading md cell after visualisation to understand the author's interpretation/observations from visualisation
+% - reading the assertion
+% - we fould 29 unique patterns from the 90 NBs; we highlight the 5 most interesting patterns in results
+
+\section{Results}\label{sec:result}
+This study identified 26 patterns of testing using visualisation and assertion.
+This section highlights five interesting patterns in more detail.
+We provide the complete catalogue of visualisation and assertion pairs as supplemental material.
+
+\subsection{Decision boundary of SVM and accuracy}
+
+% % TODO try using the minipage environment
+% \begin{table}
+%   \centering
+%   \caption{TODO}
+%   \begin{tabular}[t]{p{0.5\textwidth} p{0.5\textwidth}}
+%   \toprule
+%   \textbf{Visualisation} &
+%   \textbf{Assertion}\\
+%   \midrule
+%   \includegraphics[width=\linewidth]{../catalogue/select-04.png} &
+%   \begin{verbatim}
+%   assert accuracy_score(y_test, pred) > 0.95
+%   \end{verbatim}\\
+%   \includegraphics[width=\linewidth]{../catalogue/select-09.png} &
+%   \begin{verbatim}
+%   assert roc_auc_score(
+%     y_test,
+%     wv_model.predict_proba(X_test_wv)[:, 1]
+%   ) > 0.92,
+%     "something's wrong with your featurs"
+%   \end{verbatim}\\
+%   \bottomrule
+%   \end{tabular}
+%   \label{TODO}
+% \end{table}
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+    assert accuracy_score(y_test, pred) > 0.95
+  \end{lstlisting}
+  \begin{minipage}{0.5\textwidth}
+    \includegraphics{../catalogue/select-04.png}
+  \end{minipage}
+\end{minipage}
+
+\section{Discussion}\label{sec:discuss}
+\section{Threats to Validity}\label{sec:threats}
 % defend use of github as source for notebooks:
@@
 % - ecosystem and then doing a search for each of those patterns
+\section{Conclusion}\label{sec:conclude}
 

### 2023-09-21 report/esecfse24.tex
feat(report): 2023-09-21 20:36 updates

@@
 \usepackage{listings}
+
+\lstdefinestyle{mystyle}{
+  basicstyle=\ttfamily\footnotesize,
+  breakatwhitespace=true,
+  breaklines=true,
+  keepspaces=true,
+  showspaces=false,
+  showstringspaces=false,
+  showtabs=false,
+  tabsize=2,
+  captionpos=b
+}
+\lstset{style=mystyle}
 \usepackage{graphicx}
+\usepackage{caption}
 
@@
 \begin{abstract}
-  A clear and well-documented \LaTeX\ document is presented as an
-  article formatted for publication by ACM in a conference proceedings
-  or journal publication. Based on the ``acmart'' document class, this
-  article presents and explains many of the common variations, as well
-  as many of the formatting elements an author may use in the
-  preparation of the documentation of their work.
+% what is the problem?
+% why is this an important problem to solve?
+% what did we do?
+% what did we find?
+
+Testing ML systems is a highly interactive process which demands a
+human-in-the-loop approach. In addition to writing tests for the code
+base, practitioners are required to analyse and interpret several
+visualisations using their domain expertise to validate if an ML system
+satisfies the required set of functional and non-functional properties.
+Visualisations are frequently used to qualitatively assess various
+parts of an ML pipeline. However, implicit knowledge gained from
+visualisations must be translated to explicit analytical tests that fail
+when there is change in any component of the ML pipeline. We conduct an
+empirical analysis of Jupyter notebooks to catalogue the state-of-the-
+art mappings between ML visualisations and assertions. We mine Github to
+collect Jupyter Notebooks that contain assertions written in Python. We
+develop a novel methodology to identify 1764 notebooks which contain an
+assertion below a visualisation. We manually analyse the 1.7K notebooks
+and identify 551 notebooks where at least one pair of visualisation and
+assertion are semantically related to one another. We further read and
+understand 90 notebooks in depth to identify 26 qualitative and their
+corresponding quantitative testing patterns. We showcase five such
+patterns in this paper and plan to make the entire catalogue publicly
+available in the near future.
+
 \end{abstract}
@@
 \subsection{Related Work}\label{sec:related}
+% summarise existing work in SE on computational NBs: bavishi2021vizsmith, pimentel2019large, quaranta2021kgtorrent
 \subsection{Background}\label{sec:background}
@@
 % TODO is there a scientific term in SE research to explain reading code in detail?
-% NOTE: we looked for symmantic similarity between the visualisation and assertion code
+% NOTE we looked for semantic similarity between the visualisation and assertion code
+
 This section presents the methodology used to conduct the empirical
@@
 were applied to the 54K notebooks obtained from Gihub, to obtain the
-final sample size of 1.7K notebooks used in the empirical study. The
-red boxes represent the exclusion criteria used during phase two to
-remove notebooks note relevant for this study. Finally, the green
-boxes in phase 3 represent the manual analysis conducted to identify
-the final 29 visualisation and assertion pairs.
+final sample size of 1.7K notebooks used in the empirical study. The red
+boxes represent the exclusion criteria used during phase two to remove
+notebooks note relevant for this study. Finally, the green boxes in
+phase 3 represent the manual analysis conducted to identify the final 29
+visualisation and assertion pairs.
 
@@
 
-\subsection{Deeper Analysis of Notebooks}\ref{sec:deep-dive}
+\subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}
 % describe the steps used for the deeper analysis of NBs
@@
 \section{Results}\label{sec:result}
-This study identified 26 patterns of testing using visualisation and assertion.
-This section highlights five interesting patterns in more detail.
-We provide the complete catalogue of visualisation and assertion pairs as supplemental material.
+This study identified 26 patterns of testing using visualisation and
+assertion. This section highlights five interesting patterns in more
+detail. We provide the complete catalogue of visualisation and
+assertion pairs as supplemental material.
 
-\subsection{Decision boundary of SVM and accuracy}
+\subsection{Decision boundary}\label{sec:svm}
 
-% % TODO try using the minipage environment
-% \begin{table}
-%   \centering
-%   \caption{TODO}
-%   \begin{tabular}[t]{p{0.5\textwidth} p{0.5\textwidth}}
-%   \toprule
-%   \textbf{Visualisation} &
-%   \textbf{Assertion}\\
-%   \midrule
-%   \includegraphics[width=\linewidth]{../catalogue/select-04.png} &
-%   \begin{verbatim}
-%   assert accuracy_score(y_test, pred) > 0.95
-%   \end{verbatim}\\
-%   \includegraphics[width=\linewidth]{../catalogue/select-09.png} &
-%   \begin{verbatim}
-%   assert roc_auc_score(
-%     y_test,
-%     wv_model.predict_proba(X_test_wv)[:, 1]
-%   ) > 0.92,
-%     "something's wrong with your featurs"
-%   \end{verbatim}\\
-%   \bottomrule
-%   \end{tabular}
-%   \label{TODO}
-% \end{table}
+% TODO try using captionof for the listing as well, so that the caption for the listing and figure align at the bottom
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python, caption=Assertion on the accuracy of the ML model.]
+assert accuracy_score(y_test, pred) > 0.95
+  \end{lstlisting}
+  \label{lst:svm}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-04.png}
+  \captionof{figure}{Visualisation of the decision boundary of a SVM.}
+  \label{fig:svm}
+\end{minipage}
+
+Figure~\ref{fig:svm} presents a visualisation of the decision boundaries of a Support Vector Machine (SVM).
+The author is visualising how the SVM is distinguishing between the two classes present in the dataset.
+Listing~\ref{lst:svm} shows the corresponding assertion defined below the visualisation.
+
+The assertion is  
+\subsection{ROC Curve}
 
@@
   \begin{lstlisting}[language=Python]
-    assert accuracy_score(y_test, pred) > 0.95
+assert roc_auc_score(y_test, wv_model.predict_proba(X_test_wv)[:, 1] ) > 0.92, "something's wrong with your features"
   \end{lstlisting}
+\end{minipage}
 \begin{minipage}{0.5\textwidth}
-    \includegraphics{../catalogue/select-04.png}
+  \includegraphics[width=\linewidth]{../catalogue/select-09.png}
 \end{minipage}
+
+\subsection{Training loss}
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+assert np.mean(history[:10]) >np.mean(history[-10:]), "RNN didn't converge"
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-12.png}
+\end{minipage}
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+# assert we actually descended at each step
+for i in range(1, n_steps):
+    assert loss_array_true_grad[i] - loss_array_true_grad[i-1]) <= 0
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-267.png}
+\end{minipage}
+
+\subsection{Coefficient of Determination}
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+r2_gru = r2_score(y_test, y_pred)
+assert r2_gru > 0.6
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
+\end{minipage}
+
+\subsection{Time-series Prediction}
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+# Result of the wfc function for the best choice of s
+assert np.isclose(best_wfv, 7685.546728971963, 0.001)
+  \end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-117b.png}
 \end{minipage}

### 2023-09-22 report/esecfse24.tex
feat(report): 2023-09-22 14:50 updates

@@
 % what did we find?
-
-Testing ML systems is a highly interactive process which demands a
-human-in-the-loop approach. In addition to writing tests for the code
-base, practitioners are required to analyse and interpret several
-visualisations using their domain expertise to validate if an ML system
-satisfies the required set of functional and non-functional properties.
-Visualisations are frequently used to qualitatively assess various
-parts of an ML pipeline. However, implicit knowledge gained from
-visualisations must be translated to explicit analytical tests that fail
-when there is change in any component of the ML pipeline. We conduct an
-empirical analysis of Jupyter notebooks to catalogue the state-of-the-
-art mappings between ML visualisations and assertions. We mine Github to
-collect Jupyter Notebooks that contain assertions written in Python. We
-develop a novel methodology to identify 1764 notebooks which contain an
-assertion below a visualisation. We manually analyse the 1.7K notebooks
-and identify 551 notebooks where at least one pair of visualisation and
-assertion are semantically related to one another. We further read and
-understand 90 notebooks in depth to identify 26 qualitative and their
-corresponding quantitative testing patterns. We showcase five such
-patterns in this paper and plan to make the entire catalogue publicly
-available in the near future.
+Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
+In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
+Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
+However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
+We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the- art mappings between ML visualisations and assertions.
+We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
+We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
+We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another.
+We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns.
+We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.
 
@@
 
+% TODO update the numbers
 \begin{figure}
@@
 
-Figure~\ref{fig:svm} presents a visualisation of the decision boundaries of a Support Vector Machine (SVM).
-The author is visualising how the SVM is distinguishing between the two classes present in the dataset.
-Listing~\ref{lst:svm} shows the corresponding assertion defined below the visualisation.
+% is the data from the training or testing set?
+% what is the info obtained from the visualisation?
+% what is the assertion doing? what is the info obtained here?
+% does the assertion reflect the info obtained from the visualisation?
 
-The assertion is  
-\subsection{ROC Curve}
+Figure~\ref{fig:svm} presents a visualisation obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
+The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}.
+The visualisation shows the two classes present in the training set, and the decision boundary of the model.
+
+Listing~\ref{lst:svm} shows the corresponding assertion that was defined below Figure~\ref{fig:svm} in the notebook.
+The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
+
+% NOTE it seems wrong that the visualisation is on the training set but the assertion is on the test set; but at the same time, it does seem "natural" to visualise the training set and calculate accuracy on the test set?
+
+\subsection{Input Image, \texttt{predict\_proba}}
+
+Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.
+
+Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
+The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
+Since the model is trained to detect tennis ball, and the input image is of a football, the specified probability estimate is very low.
 
@@
   \begin{lstlisting}[language=Python]
-assert roc_auc_score(y_test, wv_model.predict_proba(X_test_wv)[:, 1] ) > 0.92, "something's wrong with your features"
+np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check if the probability estimate of the model is close to the specified value.
+The input image is of a football.
+Since the model is trained to detect tennis balls, the specified value is very small.}
+  \label{lst:football}
 \end{minipage}
 \begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-09.png}
+  \includegraphics[width=\linewidth]{../catalogue/select-35a.png}
+  \captionof{figure}{Random Input image of a football from the test set.}
+  \label{fig:football}
+\end{minipage}
+
+In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.
+
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[language=Python]
+np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
+  \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check if the probability estimate of the model is close to the specified value.
+Compared to Figure~\ref{fig:football}, the specified value is much higher, since the input image is of tennis balls.}
+  \label{lst:tennisball}
+\end{minipage}
+\begin{minipage}{0.5\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-35b.png}
+  \captionof{figure}{Random input image of tennis balls from the test set.}
+  \label{fig:tennisball}
 \end{minipage}
@@
 \subsection{Training loss}
+
+% TODO this is also a condition0 example, the NB was flagged for another viz,assert pair that is not relevant to us. How do we handle this? Because this is a really nice example to have!
+Figure~\ref{fig:loss-relaxed} shows the loss of a Recurrent Neural Network (RNN) over the trainin epochs. The visualisation was obtained from a notebook that trains an RNN to generate human names.
 
@@
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that the mean of first 10 observations of training loss is higher than the mean of last 10 observations.
+In other words, the assertion checks that the loss decreased after training and the model converged to an optimal minima.}
+  \label{lst:loss-relaxed}
 \end{minipage}
@@
   \includegraphics[width=\linewidth]{../catalogue/select-12.png}
+  \captionof{figure}{Loss over training epochs of an ML model.}
+  \label{fig:loss-relaxed}
 \end{minipage}
@@
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that the loss at each iteration $n$ is less than or equal to the loss at $n-1$.}
+  \label{lst:loss-strict}
 \end{minipage}
@@
   \includegraphics[width=\linewidth]{../catalogue/select-267.png}
+  \captionof{figure}{Another example of the same visualisation presented in Figure~\ref{fig:loss-relaxed}. The visualisation shows the training loss of an ML model over time.}
+  \label{fig:loss-strict}
 \end{minipage}
@@
 \subsection{Coefficient of Determination}
+
+Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
+The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
+
+% I want to say that the spread of the points should be close to the diagonal to indicate a good fit which is not the case here.
+The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
+The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model learned something during training.
+However, the spread of the points is wide indicating that it can be improved with more fine tuning.
+
+Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
+The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn~\footnote{REFME}.
+The assertion checks that the $R^2$ is higher than a specified threshold.
 
@@
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}
+  \label{lst:r2}
 \end{minipage}
@@
   \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
+  \captionof{figure}{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}
+  \label{fig:r2}
 \end{minipage}
@@
 \section{Discussion}\label{sec:discuss}
+% NOTE comment on the over-whelming presence of magic thresholds in ML testing! where are authors getting these numbers from? link this to requirements engineering?
+
+% NOTE comment on the anti-patterns in testing that we observe in the 5 patterns presented
+
+% NOTE link the qualitative and quantitative testing to the ML lifecycle: how some tests will let practitioner know when something changes
+
+% NOTE comment on evolution? we have to look at the entire lifecycle of a visualisation: notebooks are run iteratively?
 \section{Threats to Validity}\label{sec:threats}

### 2023-09-23 report/esecfse24.tex
feat(report): 2023-09-23 21:21 notes on related work

@@
 \section{Preliminaries}\label{sec:prelim}
-\subsection{Related Work}\label{sec:related}
-% summarise existing work in SE on computational NBs: bavishi2021vizsmith, pimentel2019large, quaranta2021kgtorrent
-\subsection{Background}\label{sec:background}
-\subsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
+\subsection{ML Testing}\label{sec:ml-testing}
+% zhang2020machine
+
+% The emphasis is primarily on image datasets and DNNs, specifically
+% CNNs & RNNs. Almost all papers are vested in the notion of addressing
+% qualitative metrics such as robustness, safety & privacy. Of these 3
+% robustness appears most frequently because it is the easiest to define
+% in terms of quantitative metrics.
+
+% The two most common domains are safety-critical systems (such as
+% autonomous driving) & neural machine translation (NMT)
+% [ji2021automated, sun2020automatic, he2020structure].
+
+% The proposed testing techniques can be classified into 4
+% categories:
+
+% 1. Improvements over existing test adequacy metrics such as neuron
+%    coverage which was proposed by DeepExplore & DeepGauge
+%    [gerasimou2020importance].
+% 2. Formal verification methods which try to provide formal guarantee
+%    of robustness against adversarial examples [zhu2021deepmemory,
+%    baluta2021scalable].
+% 3. Data augmentation techniques which generate adversarial examples to
+%    be used during training to improve robustness of models. There are
+%    techniques based on fuzzing & seach based software testing
+%    (SBST) [braiek2019deepevolution, gao2020fuzz] and mutation testing
+%    [wang2021robot, zhang2020white].
+% 4. And finally some papers propose adversarial detection during
+%    runtime [xiao2021self, wang2020dissector, wang2019adversarial,
+%    berend2020cats].
+
+\subsection{Computational Notebooks and Software
+  Engineering}\label{sec:notebooks}
+
+% NOTE maybe this should be about jupyter notebook mining and work
+% done? Then we can say that these are the only two publicly available
+% datasets with notebooks; and the challenges we faced when trying to
+% use them!
+
+% TODO review Kastner's bib, he also has a section dedicated to prior
+% work done using notebooks!
+
+% TODO we need to defend why we did not use kgtorrent dataset for our analysis.
+We can say that we performed analysis on random samples of notebooks from kaggle and did not find that many interesting examples?
+
+% TODO some form of introduction to the notion of computational notebooks; literal programming?
+
+\citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
+The KGTorrent dataset is publicly available on Zenodo, in the form of a 81TB compressed file.
+
+\citeauthor{pimentel2019large} mine 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices when developing software using computational notebooks.
+The authors also make their dataset of notebooks publicly available on Zenodo~\footnote{https://zenodo.org/record/2592524}.
+Computational notebooks were originally designed to promote sharing reproducible research since they allow presenting information by interweaving natural text, code and other multimedia such as tables, images and videos.
+However, the authors discover that Jupyter notebooks are far from being reproducible due to bad coding practices such as lack of separation of concern, testing and versioning.
+
+% TODO yang2021subtle: data scientists spend most of their time preparing the data ('data wrangling').
+Authors propose WrangleDoc, a tool that automatically generates documentation of data wrangling code and what effect it has on the data.
+They integrate their tool with JupyterLab and evaluate using 100 notebooks from Kaggle and also with practitioners.
+
+% TODO head2019managing: notebooks do not follow best practices.
+authors typically work in an iterative fashion, often experimenting with different versions of code to analyse data and produce meaningful visualisations.
+This leads to duplicates and messy notebooks.
+The paper presents a code gathering tool that allows notebook authors to review various versions and only keep the relevant version.
+Paper performs qualitative usability study using 12 participants.
+
+% TODO kery2018story: study with 21 participants on coding practices when working in notebooks.
+Authors observe that practitioners often have multiple versions that lead to problems and reduced readability.
+Participants typically create narative through order of cells rather than with the use of markdown cells.
+Paper then conducts survey with 45 data scientists and provides design guidelines for future literate programming tools.
+
+% TODO chattopadhyay2020wrong: semi-structed interviews (n=20) and survey (n=156) to understand pain points when working with notebooks. Guidelines on design of computational notebooks.
+
+% TODO psallidas2019data: paper conducts a super massive empirical study: 6M python notebooks from github, 2M enterprise DS pipelines (anonymous company name), source code and metadata from 900+ releases 12 important DS libraries.
+Paper provides high-level guidelines on what technologies DS practitioners should focus and what system builders should focus on to serve practitioners better.
+
+% TODO kandel2012enterprise: semi-structured interview study (n=35) with data analysts from 25 organizations across diverse sectors.
+Relationship between data analysis and how oganizational features of enterprise affect it.
+Recurring painpoints, open challenges and guidelines for building better visual analysis research.
+
+% TODO liu2019understanding: 
+\subsection{Visualisations and Software Engineering}\label{sec:visualisations}
+
+% bavishi2021vizsmith
+% ??
+
+\subsection{Technical Background}\label{sec:background}
+\subsubsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
 % TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bibliography}
+
+\subsubsection{Pre and Post-Condition Checks}\ref{sec:pre-post-cond}
+% TODO explain and give examples of what we mean by pre/post-condition checks with examples
+
+% TODO contrast this to the type of tests we are actually looking for! Why are pre/post conditions not valuable to look at?
 
@@
 
+% TODO this is condition0! find an alternative!
+
 \begin{minipage}{0.5\textwidth}
@@
 
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibliography}
 \end{document}

### 2023-09-24 report/esecfse24.tex
feat(report): 2023-09-24 12:26 notes on related work
feat(report): 2023-09-24 18:23 updates

@@
 
+Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
+Although originally intended to promote reproducible software, computational notebooks are infamous for the exact opposite since they do not follow software engineering best practices.
+
+The process of understanding and cleaning the underlying training data---commonly referred to as "data wrangling"---tends to dominate the ML development workflow~\cite{CITME}.
+Data wrangling is high iterative.
+Practitioners often explore several hypotheses, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation.
+Studies have been conducted to understand how practitioners generate, evaluate and manage such alternatives.
+~\cite{liu2019understanding,kandel2012enterprise}.
+
+Developing software in computational notebooks introduces a new paradigm for writing code.
+In addition to code, notebooks also contain additional information in the form of natural language text and images.
+To better understand the process of working within computational notebooks and the challenges faced along the way, studies with human subjects have been conducted\cite{head2019managing}.
+
+ ML practitioners often experiment with several versions of code to implement the same functionality.
+Prior work has explored how practitioners manage the different versions of code and the challenges that they face along the way.
+
 \citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
@@
 
+% TODO yang2021subtle: data scientists spend most of their time preparing the data ('data wrangling').
+Authors propose WrangleDoc, a tool that automatically generates documentation of data wrangling code and what effect it has on the data.
+They integrate their tool with JupyterLab and evaluate using 100 notebooks from Kaggle and also with practitioners.
+
+% code quality; reproducibility; best practices...
 \citeauthor{pimentel2019large} mine 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices when developing software using computational notebooks.
@@
 
-% TODO yang2021subtle: data scientists spend most of their time preparing the data ('data wrangling').
-Authors propose WrangleDoc, a tool that automatically generates documentation of data wrangling code and what effect it has on the data.
-They integrate their tool with JupyterLab and evaluate using 100 notebooks from Kaggle and also with practitioners.
+% TODO psallidas2019data: paper conducts a super massive empirical study: 6M python notebooks from github, 2M enterprise DS pipelines (anonymous company name), source code and metadata from 900+ releases 12 important DS libraries.
+Paper provides high-level guidelines on what technologies DS practitioners should focus and what system builders should focus on to serve practitioners better.
 
@@
 
-% TODO psallidas2019data: paper conducts a super massive empirical study: 6M python notebooks from github, 2M enterprise DS pipelines (anonymous company name), source code and metadata from 900+ releases 12 important DS libraries.
-Paper provides high-level guidelines on what technologies DS practitioners should focus and what system builders should focus on to serve practitioners better.
-
-% TODO kandel2012enterprise: semi-structured interview study (n=35) with data analysts from 25 organizations across diverse sectors.
+% DONE kandel2012enterprise: semi-structured interview study (n=35) with data analysts from 25 organizations across diverse sectors.
 Relationship between data analysis and how oganizational features of enterprise affect it.
@@
 
-% TODO liu2019understanding: 
-\subsection{Visualisations and Software Engineering}\label{sec:visualisations}
+% DONE liu2019understanding: semi-structed interview study (n=12) with data workers with various expertise and domain knowledge.
+Paper wants to understand the process behind exploring "alternatives": different hypothesis, visual design, methods, tools, algorithms, data sources.
+The studies explores the motivation behind exploring such alternatives, how alternatives help with sense making process, high-level process of alternatives and how practitioners generate, explore and manage such alternatives.
 
-% bavishi2021vizsmith
-% ??
+computational notebooks have been widely adopted by the data science and machine learning community for developing ML models.
+Unlike traditional software, ML enabled systems evolve through change in not only the code base, but also the complexity of the model and the underlying training data.
+
+Naturally, research is active in understanding the process of developing using notebooks. Prior work has looked into best practices
+\subsection{Visualisations and Machine Learning}\label{sec:visualisations}
+
+% TODO bavishi2021vizsmith: paper presents VizSmith, an automatic
+% data visualisation framework that accepts data in the form of pandas
+% dataframe along with a list of columns the user wishes to visualise.
+% The tool then automatically produces visualisations for the
+% specified columns.
+
+% TODO amershi2015modeltracker: model building is a cognitively
+% disruptive step; practitioners are required to iterate between model
+% building and error analysis. Paper presents ModelTracker, a
+% visualisation tool that presents all the information in traditional
+% summary statistics along with the performance of the model. Tool
+% validated with users who used the tool for 6 months in a production
+% setting.
+
+% TODO gortler2022neo: Research with ML practitioners at Apple and
+% identify traditional confusion matrix does not support complex data
+% structures that are prevalent today. Paper proposes Neo, an
+% interactive and multi-output confusion matrix.
+
+% TODO yang2022pattern: Paper conducts study on interaction techniques for Multivariate matrix visualisations.
+
+% TODO wang2023gam: paper proposes GAM COACH, a tool that generates customizable counterfactual explanations for Generalized Additive Models (GAMs).
+The tools uses interactive visualisations to allow users to iteratively generate algorithmic recourse plans.
+The tool is qualitatively validated by users(n=41).
+
+% TODO ahn2023escape: paper proposes ESCAPE, a visual analytics tool that allows users to identify systematic errors in AI enabled systems.
+The tool promotes a human-in-the-loop approach to identify spurious associations and concept associated misclassifications.
+
+% TODO wang2023drava: paper proposes DRAVA, a visual analytics tool to relate human concepts with semantic dimensions extracted by ML models during Disentangled representation learning>.
+The tool also provides feedback to minimize the mismatches and helps practitioners obtain data insights fron concept-driven exploration.
+The tool is validated using application scenarios in an experimental setting.
+
+% TODO robertson2023angler: authors propose ANGLER, a visual analytics tools to help practitioners prioritise machine translation model improvements.
+Tool is validated with machine translation experts (n=7).
+The tool allows users to validate interesting hypotheses for prioritising model improvements.
+
+% TODO bernard2019visual: paper proposes visual-analytics tool to perform data pre-processing of multi-variate time series data.
+
+% TODO gschwandtner2018know: paper proposes visual analytics technique to perform data profiling of time series data
+
+% TODO yuan2021survey: survey of visual analytics techniques used in ML. Paper proposes taxonomy of visual analytics techniques: before model building, during model building and after model building.
+
+% TODO wexler2020if: paper proposes WHAT-IF tool, a model agnostic, interactive visualisation tool the understanding ML models.
+The tool allows users to visually analyse and verify their hypothesis regarding the training data and how the model works.
+The tool requires a trained model and a sample dataset.
+The tool is integrated with Tensorflow, Jupyter and Collaboratory notebooks.
+The tool allows conterfactual reasoning, investigate decision boundariesa and investigate how change in data affects the prediction.
+
+% TODO hohman2019visual: survey of visual analytics techniques specifically for deep learning models.
+Visual analytics has been rapidly used for model explanation, interpretation, debugging and improvement.
 
@@
 % TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
+
+\lstinputlisting[]{nbformat.json}
+\captionof{lstlisting}{Internal JSON structure of Jupyter Notebooks.}
+\label{lst:nbformat}
+
+The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
+Listing~\ref{lst:nbformat} shows the internal json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
+For simplicity, we only show the parts that are relevant for this study.
+
+Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
+The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
+\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
+The \texttt{cells} key consists of a list of all cells within the notebook.
+
+Listing~\ref{lst:nbformat} presents three examples of cells.
+Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
+Cells can be of three types: "markdown", "code" or "raw".
+The contents of each cell can be found within the \texttt{source} key, as a list of strings.
+
+The output generated by a code cell is stored within the \texttt{outputs} key.
+Static visualisations are stored as base64 encoded png images under the \texttt{"image/png"} key, which is nested within the \texttt{data} key inside \texttt{outputs}.
+Since a cell can generate multiple outputs, the \texttt{outputs} key is a list.
 
@@
 \begin{minipage}{0.5\textwidth}
-  \begin{lstlisting}[language=Python, caption=Assertion on the accuracy of the ML model.]
+  \begin{lstlisting}[language=Python]
 assert accuracy_score(y_test, pred) > 0.95
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion on the accuracy of the ML model.}
   \label{lst:svm}

### 2023-09-25 report/esecfse24.tex
feat(report): 2023-09-25 related work draft

--- 2023-09-22 report/esecfse24.tex
@@
 \section{Preliminaries}\label{sec:prelim}
-\subsection{Related Work}\label{sec:related}
-% summarise existing work in SE on computational NBs: bavishi2021vizsmith, pimentel2019large, quaranta2021kgtorrent
-\subsection{Background}\label{sec:background}
-\subsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
+\subsection{ML Testing}\label{sec:ml-testing}
+% zhang2020machine
+
+% The emphasis is primarily on image datasets and DNNs, specifically
+% CNNs & RNNs. Almost all papers are vested in the notion of addressing
+% qualitative metrics such as robustness, safety & privacy. Of these 3
+% robustness appears most frequently because it is the easiest to define
+% in terms of quantitative metrics.
+
+% The two most common domains are safety-critical systems (such as
+% autonomous driving) & neural machine translation (NMT)
+% [ji2021automated, sun2020automatic, he2020structure].
+
+% The proposed testing techniques can be classified into 4
+% categories:
+
+% 1. Improvements over existing test adequacy metrics such as neuron
+%    coverage which was proposed by DeepExplore & DeepGauge
+%    [gerasimou2020importance].
+% 2. Formal verification methods which try to provide formal guarantee
+%    of robustness against adversarial examples [zhu2021deepmemory,
+%    baluta2021scalable].
+% 3. Data augmentation techniques which generate adversarial examples to
+%    be used during training to improve robustness of models. There are
+%    techniques based on fuzzing & seach based software testing
+%    (SBST) [braiek2019deepevolution, gao2020fuzz] and mutation testing
+%    [wang2021robot, zhang2020white].
+% 4. And finally some papers propose adversarial detection during
+%    runtime [xiao2021self, wang2020dissector, wang2019adversarial,
+%    berend2020cats].
+
+\subsection{Computational Notebooks and Software
+  Engineering}\label{sec:notebooks}
+
+%% UNDERSTANDING PAIN POINTS:
+Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
+Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large, CITME}.
+Prior studies have been conducted using human participants to gain a deeper understanding of the challenges faced by ML practitioners when developing ML enabled software in notebooks.
+The studies find that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaninful visualisations.
+The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
+
+% sense-making; data wrangling
+The process of understanding and cleaning the underlying training data---commonly referred to as "data wrangling"---tends to dominate the ML development workflow~\cite{CITME}.
+Practitioners often explore several hypotheses, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation.
+Studies have been conducted to understand how practitioners generate, evaluate and manage such alternatives.
+~\cite{liu2019understanding,kandel2012enterprise}.
+
+% code quality; reproducibility; best practices...
+In an attempt to improve the reproducibility of computational notebooks, large-scale empirical studies have been conducted.
+\citeauthor{pimentel2019large} mine 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices when developing software using computational notebooks.
+The authors also make their dataset of notebooks publicly available on Zenodo~\footnote{https://zenodo.org/record/2592524}.
+\citeauthor{psallidas2019data} conducts a large-scale empirical analysis of six million python notebooks from Github, two million enterprise data science pipelines obtained from an anonymous company and source code and metadata from over 900 releases of 12 important data science libraries.
+The authors provide a bird-eye view into the evolving landscape of computational notebooks.
+Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
+To enable future research on computational notebooks, \citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
+The KGTorrent dataset is publicly available on Zenodo~\footnote{https://zenodo.org/record/4468523}.
+
+%% TOOLS for working with notebooks:
+Several tools have been proposed in the literature to aid ML practitioners when working with computational notebooks.
+\citeauthor{yang2021subtle} propose \textit{WrangleDoc}, a tool to automatically generate documentation of data wrangling code.
+Given a snippet of data wrangling code as input, WrangleDoc can explain what effect it has on the data.
+Since ML development is highly iterative, notebooks become messy over time and accumulate duplicate versions of code.
+\citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and keep the relevant version of code.
+\citeauthor{bavishi2021vizsmith} proposes \textit{VizSmith}, an automatic data visualisation framework.
+Given as input a pandas dataframe and a list of columns the user wishes to visualise, the tool can automatically produce visualisations of the specified columns.
+
+\subsection{Visualisations and Machine Learning}\label{sec:visualisations}
+
+% NOTE I think the literature in this section is irrelevant to our study...
+
+As ML slowly augment into software systems in safety-critical domains, emphasis has been put into explainability of ML models.
+ML models and the underlying data is complex and multi-dimensional.
+Naturally, visual analytics has been widely adopted by the ML community to aid in understanding the nature of the data and the internal workings of an ML model.
+Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
+\citeauthor{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\citeauthor{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
+The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
+
+In particular, several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
+\citeauthor{wexler2020if} propose the \textit{What-If} tool, which allows users to visually analyse and verify their hypotheses regarding the training data and how the model works.
+With a trained model and a sample dataset as input, the What-If tool allows users to produce counterfactual reasoning, investigate the decision boundaries of the model and how the model predictions are affected by change in data.
+\citeauthor{ahn2023escape} propose \textit{ESCAPE}, a visual analytics tool that allows users to identify systemmatic errors in ML enabled systems.
+Through a human-in-the-loop approach, the tool helps identify spurious associations and concept associated misclassifications in the model.
+To reduce the cognitive load of ML practitioners during model building phase, \citeauthor{amershi2015modeltracker} propose \textit{ModelTracker}.
+Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
+
+Due to the highly pervasive nature of ML, several specialised visual analytics tools have also been proposed.
+\citeauthor{wang2023gam} propose \textit{GAM Coach}, a visual analytics tool that generates counterfactual explanations for Generalised Additive Models (GAMs).
+\citeauthor{bernard2019visual} and \citeauthor{gschwandtner2018know} propose specialised visual analytics tools for time-series data.
+\citeauthor{gortler2022neo} and \citeauthor{yang2022pattern} study and suggest improvements to traditional confusion matrix and multivariate matrix visualisations respectively.
+\citeauthor{robertson2023angler} propose \textit{Angler}, a visual analytics tool to help practitioners prioritise machine translation model improvements.
+\citeauthor{wang2023drava} propose \textit{Drava}, a tool to help practitioners relate human concepts with semantic dimensions extracted by ML models during disentangled representation learning.
+
+\subsection{Technical Background}\label{sec:background}
+\subsubsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
 % TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bibliography}
+
+\lstinputlisting[]{nbformat.json}
+\captionof{lstlisting}{Internal JSON structure of Jupyter Notebooks.}
+\label{lst:nbformat}
+
+The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
+Listing~\ref{lst:nbformat} shows the internal json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
+For simplicity, we only show the parts that are relevant for this study.
+
+Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
+The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
+\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
+The \texttt{cells} key consists of a list of all cells within the notebook.
+
+Listing~\ref{lst:nbformat} presents three examples of cells.
+Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
+Cells can be of three types: "markdown", "code" or "raw".
+The contents of each cell can be found within the \texttt{source} key, as a list of strings.
+
+The output generated by a code cell is stored within the \texttt{outputs} key.
+Static visualisations are stored as base64 encoded png images under the \texttt{"image/png"} key, which is nested within the \texttt{data} key inside \texttt{outputs}.
+Since a cell can generate multiple outputs, the \texttt{outputs} key is a list.
+
+\subsubsection{Pre and Post-Condition Checks}\ref{sec:pre-post-cond}
+% TODO explain and give examples of what we mean by pre/post-condition checks with examples
+
+% TODO contrast this to the type of tests we are actually looking for! Why are pre/post conditions not valuable to look at?
 

### 2023-09-25 report/esecfse24.tex
feat(report): 2023-09-25 17:38 updates

@@
 However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
-We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the- art mappings between ML visualisations and assertions.
+We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
 We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
@@
 \section{Introduction}\label{sec:intro}
-% what problem are we trying to solve?
-% what is our motivation here?
+Artificial Intelligence (AI) is here to stay.
+Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
+With the advent of Large Language Models, the power of AI to directly influence the daily lives of humans is growing at a radical pace.
+As reports of "hallucinated" facts, racist predicts and autonomous car crashes steadily increase, ensuring that ML systems function as they were intended to, takes precedence.
 
-% NOTE our methodology of finding visualisation and assertion that are related is a (pretty good) contribution.
+Visualisations are used throughout the ML development lifecycle.
+In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and understanding its statistical properties.
+During the model development phase, visualisations are used to summarise various metrics and summary statistics of the model's performance.
+Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
+
+The research questions along with the contributions of this paper are as follows.
+\begin{description}
+\item[RQ1.] How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?
+
+We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
+We develop a novel technique to identify 1764 notebooks which an assertion below a visualisation.
+We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
+
+\item[RQ2.] What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?
+
+From the 553 notebooks, we further perform an in-depth analyse of 90 notebooks.
+We identify 26 unique visualisation and assertion pairs.
+We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.
+
+\end{description}
+
+This paper is organised as follows...
 
 \section{Preliminaries}\label{sec:prelim}
+
 \subsection{ML Testing}\label{sec:ml-testing}
-% zhang2020machine
 
-% The emphasis is primarily on image datasets and DNNs, specifically
-% CNNs & RNNs. Almost all papers are vested in the notion of addressing
-% qualitative metrics such as robustness, safety & privacy. Of these 3
-% robustness appears most frequently because it is the easiest to define
-% in terms of quantitative metrics.
+Testing ML enabled software systems poses more challenges compared to their traditional counter part.
+While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase.
+With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour and gender is of paramount importance.
+Existing scientific literature on ML testing broadly focuses on two aspects.
+First, the functional property of correctness of the model in making predictions for previously unseen data.
+And the non-functional property of fairness towards humans of different race, colour, gender, sexual orientation and disabilities.
 
-% The two most common domains are safety-critical systems (such as
-% autonomous driving) & neural machine translation (NMT)
-% [ji2021automated, sun2020automatic, he2020structure].
-
-% The proposed testing techniques can be classified into 4
-% categories:
-
-% 1. Improvements over existing test adequacy metrics such as neuron
-%    coverage which was proposed by DeepExplore & DeepGauge
-%    [gerasimou2020importance].
-% 2. Formal verification methods which try to provide formal guarantee
-%    of robustness against adversarial examples [zhu2021deepmemory,
-%    baluta2021scalable].
-% 3. Data augmentation techniques which generate adversarial examples to
-%    be used during training to improve robustness of models. There are
-%    techniques based on fuzzing & seach based software testing
-%    (SBST) [braiek2019deepevolution, gao2020fuzz] and mutation testing
-%    [wang2021robot, zhang2020white].
-% 4. And finally some papers propose adversarial detection during
-%    runtime [xiao2021self, wang2020dissector, wang2019adversarial,
-%    berend2020cats].
+% TODO verify that the papers cited are actually on test adequacy metrics
+To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
+Tools such as \textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepexplore, ma2018deepgauge, gerasimou2020importance}.
+Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
+Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
+Several works have been conducted on generating and detecting adversarial inputs for ML models.
+Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
+It is however not possible to include all variations and types of adversarial examples into the training data.
+Thus several methods have also been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
 
@@
 \subsection{Technical Background}\label{sec:background}
-\subsubsection{Internal Structure of Jupyter Notebooks}\ref{sec:nbformat}
+\subsubsection{Internal Structure of Jupyter Notebooks}\label{sec:nbformat}
 % TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
@@
 \section{Methodology}\label{sec:method}
-% TODO is there a scientific term in SE research to explain reading code in detail?
-% NOTE we looked for semantic similarity between the visualisation and assertion code
+This section presents the methodology used to conduct the empirical study introduced in Section~\ref{sec:intro}.
+Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
+The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.
 
-This section presents the methodology used to conduct the empirical
-study presented in Section~\ref{sec:intro}. Figure~\ref{fig:method}
-presents an overview of all relevant steps used in this study. The
-empirical analysis comprises of three distinct phases namely:
-\textit{the data collection phase}, \textit{identification of related
-visualisation and assertion pairs} and finally a \textit{complete
-analysis of the notebooks} to identify the final set of 29
-visualisation and assertion pairs.
+The yellow boxes in the data collection phase represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
+The red boxes represent the exclusion criteria used during phase two to remove notebooks note relevant for this study.
+Finally, the green boxes in phase 3 represent the manual analysis conducted to identify the final 26 visualisation and assertion pairs.
 
-The yellow boxes in the data collection phase represent filters that
-were applied to the 54K notebooks obtained from Gihub, to obtain the
-final sample size of 1.7K notebooks used in the empirical study. The red
-boxes represent the exclusion criteria used during phase two to remove
-notebooks note relevant for this study. Finally, the green boxes in
-phase 3 represent the manual analysis conducted to identify the final 29
-visualisation and assertion pairs.
-
-Boxes appearning next to one-another imply that they were applied
-using an "OR" operator. While boxes appearning below one-another imply
-an "AND" operator. Text in double quotes represent the string pattern
-used in the search query.
+Boxes appearning next to one-another imply that they were applied using an "OR" operator.
+While boxes appearning below one-another imply an "AND" operator.
+Text in double quotes represent the string pattern used in the search query.
 
@@
 
-% TODO update the numbers
 \begin{figure}
@@
 
-We mined public repositories from Github to collect Jupyter
-notebooks---files with the \texttt{.ipynb} extension---that contain
-the keyword ``assert'' in them.
+We mined public repositories from Github to collect Jupyter notebooks that contain the keyword ``assert'' in them.
+We intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
+This is because, the ``assert'' keyword is used in python to define tests.
+Additionally, popular python libraries which provide a testing interface (such as nose, unittest and numpy), contain the same keyword in their API.
+Our search criteria captures a large initial sample of $54,070$ notebooks with a variety of testing statements.
+Our approach also prevents the need to craft a custom search query based on an exhaustive list of keywords that appear in all python testing libraries.
 
-% TODO is it clear what I am trying to say here?
-% FIXME is "testing statement" the right terminology?
+We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook.
+This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}.
+Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
+The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
 
-We intentionally keep the search criteria on Github broad by only
-checking for the appropriate filetype and the presence of the
-``assert'' keyword. This is because, the ``assert'' keyword is used in
-python to define tests. Additionally, popular python libraries which
-provide a testing interface (such as nose, unittest and numpy),
-contain the same keyword in their API. Thus, our search criteria
-captures a large initial sample of $54,070$ notebooks with a variety
-of testing statements. Our approach also prevents the need to craft a
-custom search query based on an exhaustive list of keywords that
-appear in all python testing libraries.
+The above search query results in approximately 54K notebooks.
+To further reduce the sample size, we exclude notebooks that do not use popular ML libraries.
+We perform a string search to identify notebooks that import at least one of the ML libraries.
+The string patterns are derived from the python module name used in the import statements.
+The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
 
-\begin{table}
-  \centering
-  \caption{Filters used during data collection phase.}
-  \begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
-    \toprule
-    \textbf{ID} &
-    \textbf{Filter} &
-    \textbf{Rationale}\\
-    \midrule
-    \emph{\textbf{F1}} &
-    Files with extension \texttt{.ipynb} &
-    This translates to the following Github search query:
-    \texttt{filename:"Jupyter Notebook"}.\\
-    \emph{\textbf{F2}} &
-    Notebooks with \texttt{"assert"} keyword &
-    \texttt{"assert"} is used in python to define tests. Additionally,
-    it also appears in other python libraries which provide their own
-    testing interface.\\
-    \emph{\textbf{F2}} &
-    Notebooks with popular data science or machine learning libraries &
-    The list of python libraries is derived by performing a web search
-    along with the author's prior knowledge and expertise in ML. We
-    combine the patterns using an \texttt{"OR"} operator. The final
-    regex pattern is the following:
-    \texttt{"tensorflow"|"dask"|"torch"|"sklearn"
-    |"scipy"|"numpy"|"pandas"}.\\
-    \emph{\textbf{F3}} &
-    Notebooks that contain visualisations &
-    Section~\ref{sec:nbformat} describes the internal son structure
-    of Jupyter notebooks. Visualisations are stored as binary strings
-    under the \texttt{"image/png"} field.\\
-    \emph{\textbf{F4}} &
-    Notebooks where the assertion appears below the visualisation &
-    The code cell containing the assertion must be defined below the
-    cell that produces the visualisation. Any number of markdown cells
-    may appear between the visualisation and assertion cells. See
-    Section~\ref{sec:cell-arrangement} for detailed explanation.\\
-    \emph{\textbf{F5}} &
-    Notebooks where the assertion cell is not longer than 10 lines &
-    The code cell containing the assertion should not be longer than
-    10 lines. The threshold for max lines was derived based on our
-    observations from a random sample of 50 notebooks. See
-    Section~\ref{sec:cell-arrangement} for detailed explanation.\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:filters}
-\end{table}
+We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis.
+We map each cell as a row in the dataframe.
+For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
+We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.
+The internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}.
 
-Table~\ref{tab:filters} summarises the filters used during the data
-collection phase to obtain the final sample of 1.7K notebooks from the
-initial sample of 54K notebooks mined from Github. The table provides
-a short description of the filter along with our rationale for using
-them.
-
-% TODO don't mention quantity of random sample; then we need to explain/defend this choice
-We used an iterative approach to determine the effectiveness of each
-filter that was used. We applied each filter in the same order as
-presented in Table~\ref{tab:filters}, starting with F1. We randomly
-sampled 50 notebooks from the population of notebooks obtained after
-applying the filter. We then conduct the qualitative assessment
-presented in Section~\ref{REFME} and observe the number of related
-visualisation and assertion pairs. In each new iteration, we
-progressively add the other filters and include the filter if the
-number of relevant visualisation and assertion pairs increases.
-
-% TODO maybe use colors from red-green to denote arrangements that produced better results? (instead of binary)
 \begin{figure}
@@
 In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
+
+% TODO don't mention quantity of random sample; then we need to explain/defend this choice
+We used an iterative approach to determine the effectiveness of each
+filter that was used. We applied each filter in the same order as
+presented in Table~\ref{tab:filters}, starting with F1. We randomly
+sampled 50 notebooks from the population of notebooks obtained after
+applying the filter. We then conduct the qualitative assessment
+presented in Section~\ref{REFME} and observe the number of related
+visualisation and assertion pairs. In each new iteration, we
+progressively add the other filters and include the filter if the
+number of relevant visualisation and assertion pairs increases.
 

### 2023-09-25 report/esecfse24.tex
feat(report): 2023-09-25 23:22 updates

@@
 
-\subsection{Identifying Related Visualisation and Assertion
-Pairs}\label{sec:identify-related-pairs}
+\subsection{Identifying Related Visualisation and Assertion Pairs}\label{sec:identify-related-pairs}
 
@@
 \subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}
-% describe the steps used for the deeper analysis of NBs
-% - we read the NBs in top-down fashion
-% - reading the introductory md cell at the top often gives the context for the contents of the NB (eg. what type of ML model? which part of the ML lifecycle is this relevant for?)
-% - reading the md cell above visualisation to understand context/motivation of visualisation
-% - reading the source code that produces the visualisation
-% - reading md cell after visualisation to understand the author's interpretation/observations from visualisation
-% - reading the assertion
-% - we fould 29 unique patterns from the 90 NBs; we highlight the 5 most interesting patterns in results
+
+We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
+We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
+The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
+During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data.
+We manually read and comprehend each code cell in the notebooks.
+When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
+For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation.
+A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
 
 \section{Results}\label{sec:result}
-This study identified 26 patterns of testing using visualisation and
-assertion. This section highlights five interesting patterns in more
-detail. We provide the complete catalogue of visualisation and
-assertion pairs as supplemental material.
+This study identified 26 patterns of testing using visualisation and assertion.
+This section highlights five interesting patterns in more detail.
+We plan to release the entire catalogue of patterns publicly in the near future.
 
@@
 
-% TODO try using captionof for the listing as well, so that the caption for the listing and figure align at the bottom
 \begin{minipage}{0.5\textwidth}
@@
 \begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-04.png}
-  \captionof{figure}{Visualisation of the decision boundary of a SVM.}
+  \includegraphics[width=\linewidth]{../catalogue/select-04-linear.png}
+  \captionof{figure}{Visualisation of the decision boundary of a SVM with a linear kernel.}
+
+  \includegraphics[width=\linewidth]{../catalogue/select-04-rbf.png}
+  \captionof{figure}{Visualisation of the decision boundary of a SVM with a RBF kernel.}
   \label{fig:svm}
@@
 
-% is the data from the training or testing set?
-% what is the info obtained from the visualisation?
-% what is the assertion doing? what is the info obtained here?
-% does the assertion reflect the info obtained from the visualisation?
-
-Figure~\ref{fig:svm} presents a visualisation obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
+Figures~\ref{fig:svm} presents visualisations obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
 The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}.
-The visualisation shows the two classes present in the training set, and the decision boundary of the model.
-
-Listing~\ref{lst:svm} shows the corresponding assertion that was defined below Figure~\ref{fig:svm} in the notebook.
+The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel.
+Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations presented in Figure~\ref{fig:svm}.
 The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
@@
 \subsection{Input Image, \texttt{predict\_proba}}
-
-Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.
-
-Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
-The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
-Since the model is trained to detect tennis ball, and the input image is of a football, the specified probability estimate is very low.
 
@@
 
+Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.The visualisations are of two random input images, one from each class in the dataset.
+Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
+The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
+Since the model is trained to detect tennis ball, and the input image contains a footbal, the specified probability estimate is very low.
 In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.
+
+% NOTE when working with images, we tend to see that the input image is visualised (so human can manually validate), something to do with expensive labelling?
+% NOTE is this a good pattern? this test does not generalise across all possible input images?
 
@@
 The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
-
-% I want to say that the spread of the points should be close to the diagonal to indicate a good fit which is not the case here.
 The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
@@
 However, the spread of the points is wide indicating that it can be improved with more fine tuning.
-
 Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
-The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn~\footnote{REFME}.
+The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn.
 The assertion checks that the $R^2$ is higher than a specified threshold.
@@
 
-\subsection{Time-series Prediction}
-
-% TODO this is condition0! find an alternative!
+\subsection{Normalization}
 
@@
   \begin{lstlisting}[language=Python]
-# Result of the wfc function for the best choice of s
-assert np.isclose(best_wfv, 7685.546728971963, 0.001)
+assert normalized_data[normalized_data.argmax()] < 10
+assert normalized_data[normalized_data.argmin()] > -7
   \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that the min and max of a feature fall within specified thresholds derived from the visualisation presented in Figure~\ref{fig:normal}.}
+  \label{lst:normal}
 \end{minipage}
 \begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-117b.png}
+  \includegraphics[width=\linewidth]{../catalogue/select-65b.png}
+  \captionof{figure}{Histogram showing the distribution of a feature before and after normalisation.}
+  \label{fig:normal}
 \end{minipage}

### 2023-09-26 report/esecfse24.tex
feat(report): 2023-09-26 14:22 updates

@@
 
+\usepackage{xcolor}
 \usepackage{listings}
+\usepackage{caption}
+\usepackage{subcaption}
+
+\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
 
@@
   basicstyle=\ttfamily\footnotesize,
+  keywordstyle=\color{blue},
+  stringstyle=\color{red},
+  commentstyle=\color{green},
   breakatwhitespace=true,
@@
   showtabs=false,
-  tabsize=2,
-  captionpos=b
+  tabsize=4,
+  captionpos=b,
+  numbers=left,
+  numberstyle=\tiny\color{gray},
+  backgroundcolor=\color{backcolor},
 }
@@
 
-\begin{minipage}{0.5\textwidth}
+% NOTE something about how the visualisation helps with interpretability of how the model is classifying the data?
+
+% NOTE the author is using the visualisation to iteratively check/test how different kernels perform on the data; can also use it for hyper-parameter tuning?
+
+% NOTE: visualisation can provide qualitative assessment of under/overfitting
+
+% NOTE: better assertion: perhaps on the support vectors?
+
+\begin{figure}
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-linear.png}}
+  \hfill
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-rbf.png}}
+  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}
+  \label{fig:svm}
+\end{figure}
+
 \begin{lstlisting}[language=Python]
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion on the accuracy of the ML model.}
-  \label{lst:svm}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-04-linear.png}
-  \captionof{figure}{Visualisation of the decision boundary of a SVM with a linear kernel.}
-
-  \includegraphics[width=\linewidth]{../catalogue/select-04-rbf.png}
-  \captionof{figure}{Visualisation of the decision boundary of a SVM with a RBF kernel.}
-  \label{fig:svm}
-\end{minipage}
+\captionof{lstlisting}{Assertion on the accuracy of the ML model.}\label{lst:svm}
 
@@
 The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}.
-The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel.
-Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations presented in Figure~\ref{fig:svm}.
+The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel (Figure~\ref{fig:svm-linear}) and later with a RBF kernel (Figure~\ref{fig:svm-rbf}).
+Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
 The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
 
-% NOTE it seems wrong that the visualisation is on the training set but the assertion is on the test set; but at the same time, it does seem "natural" to visualise the training set and calculate accuracy on the test set?
+Visualising the decision boundary of a SVM provides a visual understanding of how the model is performing for a given dataset.
+Mis-classified examples can be seen on the wrong side of the decision boundary.
+The visualisation can also be used to check if the model is overfitting or underfitting the data.
+For instance, in Figure~\ref{fig:svm-linear} we observe that the model is underfitting since a linear kernel too simply for the underlying data.
 
@@
 
-\begin{minipage}{0.5\textwidth}
+\begin{figure}
+  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35a.png}}
+  \hfill
+  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35b.png}}
+  \caption{Visualisation of random input images for a pre-trained image classifier.}
+  \label{fig:balls}
+\end{figure}
+
+\begin{minipage}{0.45\textwidth}
   \begin{lstlisting}[language=Python]
@@
 \end{minipage}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-35a.png}
-  \captionof{figure}{Random Input image of a football from the test set.}
-  \label{fig:football}
-\end{minipage}
-
-Figures~\ref{fig:football} and~\ref{fig:tennisball} are obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.The visualisations are of two random input images, one from each class in the dataset.
-Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
-The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
-Since the model is trained to detect tennis ball, and the input image contains a footbal, the specified probability estimate is very low.
-In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.
-
-% NOTE when working with images, we tend to see that the input image is visualised (so human can manually validate), something to do with expensive labelling?
-% NOTE is this a good pattern? this test does not generalise across all possible input images?
-
-\begin{minipage}{0.5\textwidth}
+\hfill
+\begin{minipage}{0.45\textwidth}
   \begin{lstlisting}[language=Python]
@@
 \end{minipage}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-35b.png}
-  \captionof{figure}{Random input image of tennis balls from the test set.}
-  \label{fig:tennisball}
+
+Figure~\ref{fig:balls} presents visualisations obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.The visualisations are of two random input images, one from each class in the dataset.
+Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
+The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
+Since the model is trained to detect tennis ball, and the input image contains a footbal, the specified probability estimate is very low.
+In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.
+
+The above testing technique can be used to spot-check a few hand-picked examples of edge cases or adversarial examples which the model may find difficult to classify.
+The technique can also be generalized by constructing a large dataset containing all the edge cases that the develop may wish to check.
+And the assertion can be called on the probability estimate for image.
+
+% NOTE when working with images, we tend to see that the input image is visualised (so human can manually validate), something to do with expensive labelling?
+% NOTE is this a good pattern? this test does not generalise across all possible input images?
+
+\subsection{Normality}\label{sec:normal}
+
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}[language=Python]
+for feature in range(data_transformed.shape[1]):
+  assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
+  \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that each feature in a dataset is normal.
+The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}
+  \label{lst:kstest}
 \end{minipage}
-
-\subsection{Training loss}
-
-% TODO this is also a condition0 example, the NB was flagged for another viz,assert pair that is not relevant to us. How do we handle this? Because this is a really nice example to have!
-Figure~\ref{fig:loss-relaxed} shows the loss of a Recurrent Neural Network (RNN) over the trainin epochs. The visualisation was obtained from a notebook that trains an RNN to generate human names.
-
-\begin{minipage}{0.5\textwidth}
-  \begin{lstlisting}[language=Python]
-assert np.mean(history[:10]) >np.mean(history[-10:]), "RNN didn't converge"
-  \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the mean of first 10 observations of training loss is higher than the mean of last 10 observations.
-In other words, the assertion checks that the loss decreased after training and the model converged to an optimal minima.}
-  \label{lst:loss-relaxed}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-12.png}
-  \captionof{figure}{Loss over training epochs of an ML model.}
-  \label{fig:loss-relaxed}
-\end{minipage}
-
-\begin{minipage}{0.5\textwidth}
-  \begin{lstlisting}[language=Python]
-# assert we actually descended at each step
-for i in range(1, n_steps):
-    assert loss_array_true_grad[i] - loss_array_true_grad[i-1]) <= 0
-  \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the loss at each iteration $n$ is less than or equal to the loss at $n-1$.}
-  \label{lst:loss-strict}
-\end{minipage}
-\begin{minipage}{0.5\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-267.png}
-  \captionof{figure}{Another example of the same visualisation presented in Figure~\ref{fig:loss-relaxed}. The visualisation shows the training loss of an ML model over time.}
-  \label{fig:loss-strict}
+\hfill
+\begin{minipage}{0.45\textwidth}
+  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
+  \captionof{figure}{Visualisation of the distribution of a feature from the dataset.}
+  \label{fig:kstest}
 \end{minipage}
@@
 
-\begin{minipage}{0.5\textwidth}
+\begin{minipage}{0.45\textwidth}
   \begin{lstlisting}[language=Python]
@@
 \end{minipage}
-\begin{minipage}{0.5\textwidth}
+\hfill
+\begin{minipage}{0.45\textwidth}
   \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
@@
 \end{minipage}
-\begin{minipage}{0.5\textwidth}
+\hfill
+\begin{minipage}{0.45\textwidth}
   \includegraphics[width=\linewidth]{../catalogue/select-65b.png}
@@
 \section{Discussion}\label{sec:discuss}
+
 % NOTE comment on the over-whelming presence of magic thresholds in ML testing! where are authors getting these numbers from? link this to requirements engineering?
@@
 % NOTE comment on evolution? we have to look at the entire lifecycle of a visualisation: notebooks are run iteratively?
+
+\subsection{Evolution and Provenance of Visualisations}\label{sec:evolution}
+% NOTE ML development is iterative, visualisations are compared against one another, to qualitatively validate performance/different strategies of model performance?
+It will be an interesting direction to explore how visualisations evolve?
+
+\subsection{Something about magic thresholds}\label{sec:magic-threshold}
+% NOTE we observe that most analytical test translates to comparing a value against a threshold; this indicates the open challenges in ML testing; either we don't yet know how to translate analytical tests from visualisations or we cannot do this in the first place and we must use a hybrid approach?
+
+\subsection{Warning System}\label{sec:drift}
+% NOTE with the assertions, the implicit expectations are translated to explicit expectations. Which will fail if the expectations are no longer satisfied
+
 \section{Threats to Validity}\label{sec:threats}

### 2023-09-26 report/esecfse24.tex
feat(report): 2023-09-26 17:02 updates

@@
 
-Visualisations are used throughout the ML development lifecycle.
+Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
 In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and understanding its statistical properties.
@@
 Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
+
+Besides testing, information gained from ML visualisations have a profound influence in the design and implementation decisions taken for the next step of the ML pipeline.
+For instance, we may visualise the distribution of our training data and find that it is normally distributed.
+Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
+However, such decisions are violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
+Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer met.
+
+% However, in a production ML setting, such decisions quickly become implicit expectations since data constantly evolve and changes, as a reflection of the real world.
+These implicit expectations must be translated to more explicit, analytical tests which fail when our assumptions and expectations no longer hold.
 
@@
 
-This paper is organised as follows...
+% TODO This paper is organised as follows...
 
@@
 % TODO contrast this to the type of tests we are actually looking for! Why are pre/post conditions not valuable to look at?
+
+During the empirical analysis conducted for this study, we found several instances of where the assertions were related to the visualisation above.
+However, they were not relevant to this study since they did not test the property of the data or the model.
 
@@
 
+\subsection{Mining ML visualisations/notebooks}
+% NOTE we can perform further analysis by mining jupyter notebooks, in particular the visualisations?
+
 \section{Threats to Validity}\label{sec:threats}

### 2023-09-27 report/esecfse24.tex
feat(report): 2023-09-26 16:26 updates

@@
 \section{Results}\label{sec:result}
+
+% TODO move the stuff for RQ1 from methodology here!
+
+% NOTE the 5 pairs is to illustrate the 26 total pairs/approach we found for testing ML systems
+
+% NOTE visual+analytical test is a new "strategy" for testing AI
+
 This study identified 26 patterns of testing using visualisation and assertion.
@@
 
+% NOTE assertion created was a single metric (performance): we need tools/guidelines to help practitioners capture what they see in visualisations; community failing to help practitioners what they are visualising
+
+% NOTE assertion is not capturing the whole thing, only assert == we lose a lot of info, we need both
+
 \begin{figure}
@@
 \end{figure}
-
 \begin{lstlisting}[language=Python]
@@
 The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel (Figure~\ref{fig:svm-linear}) and later with a RBF kernel (Figure~\ref{fig:svm-rbf}).
+
+% NOTE maybe we need to ellaborate; the linear kernel was above the rbf; based on observations from the linear kernel, the author decides to try the rbf kernel; then after experimenting with hyper-parameters the author puts an assertion that captures the model's performance at that specific point in time.
+
+The visualisations presented in Figure~\ref{fig:svm} contain several useful information that can be used to test and debug ML models.
+Humans are tuned to recognise visual patterns~\cite{CITEME}.
+Visualisations in Figure~\ref{fig:svm}, allows ML practitioners to quickly identify mis-classified examples.
+Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
+For instance, in Figure~\ref{fig:svm-linear} we observe that the model is underfitting since a linear kernel too simply for the underlying data.
+The visualisation also allows ML practitioners to compare different ML models or variants of the same ML model to iteratively find a model that best fits the underlying training data.
+In this case for instance, the author of the notebook First trained the SVM with a linear kernel.
+And upon realising that the data is not linearly separable, chose the RBF kernel.
+
 Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
 The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
+The assertion is derived from the observations made from the visualisation in Figure~\ref{fig:svm-rbf}.
+The visulisation shows that the model is able to classify the data well which is why the notebook author sets the threshold for the accuracy in the assertion to a high value.
 
-Visualising the decision boundary of a SVM provides a visual understanding of how the model is performing for a given dataset.
-Mis-classified examples can be seen on the wrong side of the decision boundary.
-The visualisation can also be used to check if the model is overfitting or underfitting the data.
-For instance, in Figure~\ref{fig:svm-linear} we observe that the model is underfitting since a linear kernel too simply for the underlying data.
+The assertion however, is only able to capture the information pertaining to the performance of the model.
+While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
 
@@
   \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check if the probability estimate of the model is close to the specified value.
+  \captionof{lstlisting}{Assertion below Figure~\ref{fig:football}.
+It checks if the probability estimate of the model is close to the specified value.
 The input image is of a football.
@@
   \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check if the probability estimate of the model is close to the specified value.
-Compared to Figure~\ref{fig:football}, the specified value is much higher, since the input image is of tennis balls.}
+  \captionof{lstlisting}{Assertion below Figure~\ref{fig:tennisball}. Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}
   \label{lst:tennisball}
@@
 
-Figure~\ref{fig:balls} presents visualisations obtained from a notebook that uses a pre-trained image classifier from GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.The visualisations are of two random input images, one from each class in the dataset.
+Figure~\ref{fig:balls} presents visualisations obtained from a notebook that uses a pre-trained image classifier from the GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.
+The visualisations are of two random input images, one from each class in the dataset.
 Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
@@
 
-The above testing technique can be used to spot-check a few hand-picked examples of edge cases or adversarial examples which the model may find difficult to classify.
-The technique can also be generalized by constructing a large dataset containing all the edge cases that the develop may wish to check.
-And the assertion can be called on the probability estimate for image.
+The above visualisation and assertion pair can be used to spot-check a few edge-cases or adversarial examples which the model may find difficult to classify.
+It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
+The threshold may have been derived based on the domain expertise of the ML practitioner.
+Another alternative could be to define them during the Requirements Engineering phase, by discussing with all involved stakeholders\cite{CITEME}.
 
-% NOTE when working with images, we tend to see that the input image is visualised (so human can manually validate), something to do with expensive labelling?
-% NOTE is this a good pattern? this test does not generalise across all possible input images?
+In contrast to the visualisation and assertion pair in Section~\ref{sec:svm}, the assertion here is able to capture the author's motivation for visualising the input image along with the information obtained from it.
+Regardless, the notebook author still visualises the input image.
+This is a recurring phenomenon that we observe in other visualisation and assertion pairs in our catalogue.
+We discuss this phenomenon further in Section~\ref{sec:visual}.
 
@@
 
+% NOTE assertion and viz are perfect fit;
+
+% NOTE despite the analytical test, practitioners need to visualise this, we need visual confirmation; we cannot get rid of visualisations. we need both!
+
+The visualisation in Figure~\ref{fig:kstest} is obtained from a notebook that trains a Generative Adversarial Network (GAN) to generate data which simulates a neutrino particle detector.
+The notebook author transforms the training data by normalising it using the \texttt{QuantileTransform} from scikit-learn~\footnote{REFME}.
+After performing the transformation, the author performs a visual check by plotting the distribution of a feature from the dataset as shown in Figure~\ref{fig:kstest}.
+The visualisation shows a bell-shaped curve indicating that the data is normal and that the transformation worked.
+The author then defines the assertion presented in Listing~\ref{lst:kstest} which reflects their observation made from the visualisation.
+
+The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
+The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
+
+% NOTE I want to say that we don't even think about the fact that we always use a histogram to do this! its second nature, we are taught to do this, without thinking why!
+
+Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
+Several ML models assume normality in the underlying training data.
+The verification is often performed visually using a histogram and checking that the distribution shows a "bell-shaped" curve\cite{CITME}.
+However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
+The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
+And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
+
 \subsection{Coefficient of Determination}
 
-Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
-The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
-The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
-The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model learned something during training.
-However, the spread of the points is wide indicating that it can be improved with more fine tuning.
-Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
-The Coefficient of Determination ($R^2$) is calculated using the \texttt{r2\_score} method from scikit-learn.
-The assertion checks that the $R^2$ is higher than a specified threshold.
+% NOTE this vis,assert is model agnostic!
+
+% NOTE assertion captured what they visualised
+
+% NOTE visualisation helps to check for outliers: we are humans, we look at visual patterns, quick and easier to observe fro visualisations
 
@@
 
+Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
+The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
+The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
+The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model was able to learn certain patterns in the training data.
+The spread of $y_{pred}$ at any given $y_{true}$ is wide.
+For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
+This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.
+
+% TODO more context on r2_score, it fits a linear regression model and checks the blah blah blah
+Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
+The author uses \texttt{r2\_score} metric from the scikit-learn library to quantify the performance of the model.
+The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computed the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
+The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
+Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
+
+Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess is the model is under or overfitting the training data.
+If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
+In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
+The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
+The assertion in Listing~\ref{lst:r2} however, does not capture such information.
+It can only check if the performance of the model is better than a specific threshold.
+
 \subsection{Normalization}
 
+% TODO check the notebook again for implementation of normalization
+
+\begin{figure}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65a.png}}
+  \hfill
+  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65b.png}}
+  \caption{Visualisation of data before and after various data pre-processing techniques.}
+  \label{fig:data-pre-process}
+\end{figure}
+
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}[language=Python]
+assert scaled_data[scaled_data.argmax()] <= 1
+assert scaled_data[scaled_data.argmix()] >= 0
+  \end{lstlisting}
+  \captionof{lstlisting}{Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}
+  \label{lst:scale}
+\end{minipage}
+\hfill
 \begin{minipage}{0.45\textwidth}
@@
   \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the min and max of a feature fall within specified thresholds derived from the visualisation presented in Figure~\ref{fig:normal}.}
+  \captionof{lstlisting}{Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}
   \label{lst:normal}
 \end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-65b.png}
-  \captionof{figure}{Histogram showing the distribution of a feature before and after normalisation.}
-  \label{fig:normal}
-\end{minipage}
+
+Visualisations presented in Figure~\ref{fig:data-pre-process} were obtained from a notebook that presents a tutorial on scaling and normalization---two common data pre-processing techniques in ML.
+The visualisations show a comparison of the data before and after each pre-processing technique is applied to a feature from the data.
+The author scales the data using the \texttt{MinMaxScaler} from the scikit-learn library~\footnote{REFME}, between $[0, 1]$.
+The author performs a visual check, by comparing the distribution of the data before and after scaling, using a histogram (as shown in Figure~\ref{fig:scale}).
+Similarly, the author normalises the data using the \texttt{PowerTransformer} class from scikit-learn~footnote{REFME}, and performs a visual check as shown in Figure~\ref{fig:normal}.
+
+Listing~\ref{lst:scale}~and~\ref{lst:normal} show the assertions that were defined below Figure~\ref{fig:scale}~and~\ref{fig:normal} respectively.
+Both assertions check that the min and max of the feature is within the expected bounds after applying the respective pre-processing technique.
+The assertion in Listing~\ref{lst:scale} is felt to be unneccessary since the min and max thresholds are set to the theoretical limits.
+The assertions in Listing~\ref{normal} are more relevant, however it is not clear why the author include such a wide margin in the thresholds.
 
@@
 
-\subsection{Evolution and Provenance of Visualisations}\label{sec:evolution}
-% NOTE ML development is iterative, visualisations are compared against one another, to qualitatively validate performance/different strategies of model performance?
-It will be an interesting direction to explore how visualisations evolve?
+\subsection{From Implicit Expectations to Explicit Assertions}
+% NOTE regeneration of the model does not change this observation at this point in time
+
+% NOTE discuss the pros and cons of having a qualiative assessment: they are dense with useful information; humans need visual patterns!
+
+% NOTE then show that we also need qualitative assessment that can capture our expectations; this is important so that the model never violates that we understand from the visualisations
+
+% NOTE now is the time to talk about how assertions act as documentation of the domain expertise and knowledge extracted from the visualisation from the original creator; this also scales as the system complexity and composition of the team/organisation evolves!
 
@@
 
-\subsection{Warning System}\label{sec:drift}
-% NOTE with the assertions, the implicit expectations are translated to explicit expectations. Which will fail if the expectations are no longer satisfied
+% NOTE this is a better place to talk about the magic threshold phenomenon; link this to the magic number smell in software engineering, but in ML it is very prevalent and an alternative is not always clear
 
-\subsection{Mining ML visualisations/notebooks}
-% NOTE we can perform further analysis by mining jupyter notebooks, in particular the visualisations?
+% NOTE and then motivate how we need more tools/techniques to aid practitioners to extract/translate all the info from visualisations into qualitative assertions!
+
+\subsection{Humans need visual confirmation}\label{sec:visual}
+% NOTE even though analytical tests are sufficient; humans still prefer visual confirmation
+
+% NOTE when working with image dataset, we observe in our catalogue that practitioners prefers to also visually check the input image
+
+% NOTE more detailed discussion on how a lot of information can be condensed into a single visualition and how humans are good at observing visual patterns
 

### 2023-09-27 report/esecfse24.tex
feat(report): 2023-09-27 17:53 updates

@@
 \section{Introduction}\label{sec:intro}
-Artificial Intelligence (AI) is here to stay.
+
 Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
@@
 Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
-In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and understanding its statistical properties.
-During the model development phase, visualisations are used to summarise various metrics and summary statistics of the model's performance.
+In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and verify its statistical properties.
+During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
 Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
 
-Besides testing, information gained from ML visualisations have a profound influence in the design and implementation decisions taken for the next step of the ML pipeline.
+Building ML systems is highly iterative and experimental.
+Information gained from ML visualisations is used to make design and implementation decisions for the following steps of the ML pipeline.
 For instance, we may visualise the distribution of our training data and find that it is normally distributed.
 Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
-However, such decisions are violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
-Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer met.
+However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
+Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
 
-% However, in a production ML setting, such decisions quickly become implicit expectations since data constantly evolve and changes, as a reflection of the real world.
-These implicit expectations must be translated to more explicit, analytical tests which fail when our assumptions and expectations no longer hold.
+In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction.
+We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
+In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a qualitative form of testing (using analytical assertions).
+The research questions along with the contributions of this paper are as follows.
 
-The research questions along with the contributions of this paper are as follows.
 \begin{description}
@@
 We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
-We develop a novel technique to identify 1764 notebooks which an assertion below a visualisation.
+We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
 We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
@@
 
-From the 553 notebooks, we further perform an in-depth analyse of 90 notebooks.
+From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks.
 We identify 26 unique visualisation and assertion pairs.
@@
 
-% TODO This paper is organised as follows...
+The paper is organised as follows.
+Section~\ref{sec:prelim} summarises the related literature and technical information relevant to this study.
+The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
+The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity (Section~\ref{sec:threats}) and future work (Section~\ref{sec:conclude}).
 
 \section{Preliminaries}\label{sec:prelim}
+This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
+The section concludes with an overview of technical knowledge required for this paper.
 
@@
 Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase.
-With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour and gender is of paramount importance.
+While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase\cite{CITEME}.
+With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
 Existing scientific literature on ML testing broadly focuses on two aspects.
-First, the functional property of correctness of the model in making predictions for previously unseen data.
-And the non-functional property of fairness towards humans of different race, colour, gender, sexual orientation and disabilities.
+First, on functional properties such as correctness and robustness of the model towards unseen data.
+And second, on non-functional properties such as fairness, interpretability and privacy.
 
-% TODO verify that the papers cited are actually on test adequacy metrics
 To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
@@
 Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
-It is however not possible to include all variations and types of adversarial examples into the training data.
-Thus several methods have also been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
+It is however not possible to include all variations of adversarial examples into the training data.
+Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
+
+% TODO summarise literature on fairness testing
+Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.
 
@@
 
-%% UNDERSTANDING PAIN POINTS:
 Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large, CITME}.
-Prior studies have been conducted using human participants to gain a deeper understanding of the challenges faced by ML practitioners when developing ML enabled software in notebooks.
-The studies find that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaninful visualisations.
-The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
+Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large}.
 
-% sense-making; data wrangling
-The process of understanding and cleaning the underlying training data---commonly referred to as "data wrangling"---tends to dominate the ML development workflow~\cite{CITME}.
-Practitioners often explore several hypotheses, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation.
-Studies have been conducted to understand how practitioners generate, evaluate and manage such alternatives.
-~\cite{liu2019understanding,kandel2012enterprise}.
-
-% code quality; reproducibility; best practices...
-In an attempt to improve the reproducibility of computational notebooks, large-scale empirical studies have been conducted.
-\citeauthor{pimentel2019large} mine 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices when developing software using computational notebooks.
-The authors also make their dataset of notebooks publicly available on Zenodo~\footnote{https://zenodo.org/record/2592524}.
-\citeauthor{psallidas2019data} conducts a large-scale empirical analysis of six million python notebooks from Github, two million enterprise data science pipelines obtained from an anonymous company and source code and metadata from over 900 releases of 12 important data science libraries.
-The authors provide a bird-eye view into the evolving landscape of computational notebooks.
+\citeauthor{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, and source code and metadata from over 900 releases of 12 important data science libraries.
 Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
+\citeauthor{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices in computational notebooks.
+Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
 To enable future research on computational notebooks, \citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
 The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
-The KGTorrent dataset is publicly available on Zenodo~\footnote{https://zenodo.org/record/4468523}.
 
-%% TOOLS for working with notebooks:
+Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
+Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
+The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
+
+Since ML development is highly iterative, notebooks become messy over time and accumulate duplicate versions of code.
+\citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and keep the relevant version of code.
+
 Several tools have been proposed in the literature to aid ML practitioners when working with computational notebooks.
+
 \citeauthor{yang2021subtle} propose \textit{WrangleDoc}, a tool to automatically generate documentation of data wrangling code.
 Given a snippet of data wrangling code as input, WrangleDoc can explain what effect it has on the data.
-Since ML development is highly iterative, notebooks become messy over time and accumulate duplicate versions of code.
-\citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and keep the relevant version of code.
+
 \citeauthor{bavishi2021vizsmith} proposes \textit{VizSmith}, an automatic data visualisation framework.

### 2023-09-28 report/esecfse24.tex
feat(report): 2023-09-28 14:04 updates

@@
 % what did we find?
-Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
-In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
-Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
-However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
-We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
+Testing ML systems is a highly interactive process which demands
+  a human-in-the-loop approach. In addition to writing tests for the
+  code base, practitioners are required to analyse and interpret
+  several visualisations using their domain expertise to validate if
+  an ML system satisfies the required set of functional and
+  non-functional properties. Visualisations are frequently used to
+  qualitatively assess various parts of an ML pipeline. However,
+  implicit knowledge gained from visualisations must be translated to
+  explicit analytical tests that fail when there is change in any
+  component of the ML pipeline. We conduct an empirical analysis of
+  Jupyter notebooks to catalogue the state-of-the-art mappings between
+  ML visualisations and assertions.
 We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
@@
 
+% TODO what are the primary findings of this paper...
+
 The paper is organised as follows.
@@
 \section{Preliminaries}\label{sec:prelim}
+
+% TODO how is our work different from prior work?
+
 This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
@@
 
-Since ML development is highly iterative, notebooks become messy over time and accumulate duplicate versions of code.
-\citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and keep the relevant version of code.
-
-Several tools have been proposed in the literature to aid ML practitioners when working with computational notebooks.
-
-\citeauthor{yang2021subtle} propose \textit{WrangleDoc}, a tool to automatically generate documentation of data wrangling code.
-Given a snippet of data wrangling code as input, WrangleDoc can explain what effect it has on the data.
-
-\citeauthor{bavishi2021vizsmith} proposes \textit{VizSmith}, an automatic data visualisation framework.
-Given as input a pandas dataframe and a list of columns the user wishes to visualise, the tool can automatically produce visualisations of the specified columns.
+To manage and prune the multiple versions of code that accumulate over time when developing in notebooks, \citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
 
@@
 
-% NOTE I think the literature in this section is irrelevant to our study...
+As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
+ML models and the underlying data is complex and multi-dimensional.
+To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models.
 
-As ML slowly augment into software systems in safety-critical domains, emphasis has been put into explainability of ML models.
-ML models and the underlying data is complex and multi-dimensional.
-Naturally, visual analytics has been widely adopted by the ML community to aid in understanding the nature of the data and the internal workings of an ML model.
 Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
@@
 
-In particular, several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\citeauthor{wexler2020if} propose the \textit{What-If} tool, which allows users to visually analyse and verify their hypotheses regarding the training data and how the model works.
-With a trained model and a sample dataset as input, the What-If tool allows users to produce counterfactual reasoning, investigate the decision boundaries of the model and how the model predictions are affected by change in data.
-\citeauthor{ahn2023escape} propose \textit{ESCAPE}, a visual analytics tool that allows users to identify systemmatic errors in ML enabled systems.
-Through a human-in-the-loop approach, the tool helps identify spurious associations and concept associated misclassifications in the model.
+Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
+\citeauthor{wexler2020if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
 To reduce the cognitive load of ML practitioners during model building phase, \citeauthor{amershi2015modeltracker} propose \textit{ModelTracker}.
@@
 
-Due to the highly pervasive nature of ML, several specialised visual analytics tools have also been proposed.
-\citeauthor{wang2023gam} propose \textit{GAM Coach}, a visual analytics tool that generates counterfactual explanations for Generalised Additive Models (GAMs).
-\citeauthor{bernard2019visual} and \citeauthor{gschwandtner2018know} propose specialised visual analytics tools for time-series data.
-\citeauthor{gortler2022neo} and \citeauthor{yang2022pattern} study and suggest improvements to traditional confusion matrix and multivariate matrix visualisations respectively.
-\citeauthor{robertson2023angler} propose \textit{Angler}, a visual analytics tool to help practitioners prioritise machine translation model improvements.
-\citeauthor{wang2023drava} propose \textit{Drava}, a tool to help practitioners relate human concepts with semantic dimensions extracted by ML models during disentangled representation learning.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systemmatic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
 
-\subsection{Technical Background}\label{sec:background}
-\subsubsection{Internal Structure of Jupyter Notebooks}\label{sec:nbformat}
-% TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
+\subsection{Internal Structure of Jupyter Notebooks}\label{sec:nbformat}
 
-\lstinputlisting[]{nbformat.json}
-\captionof{lstlisting}{Internal JSON structure of Jupyter Notebooks.}
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}
+{
+  "metadata": {
+    "kernel_info": {...},
+    "language_info": {...},
+  },
+  "nbformat": 4,
+  "nbformat_minor": 0,
+  "cells": [ # dictionary of all cells in notebook ]
+}
+  \end{lstlisting}
+  \captionof{lstlisting}{Top-level JSON format of Jupyter notebooks. See Listing~\ref{lst:cell-md}, \ref{lst:cell-code} and \ref{lst:cell-output} for internal structure of various cells.}
   \label{lst:nbformat}
+\end{minipage}
+\hfill
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}
+{
+  "cell_type": "markdown",
+  "metadata": {},
+  "source": [
+    "# Hello world example\n",
+    "The following cell contains python code for greeting people!\n"
+  ]
+}
+  \end{lstlisting}
+  \captionof{lstlisting}{JSON structure of markdown cells.}
+  \label{lst:cell-md}
+\end{minipage}
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}
+{
+  "cell_type": "code",
+  "metadata": {},
+  "source": [
+    "def greet(name):\n",
+    "    print(f'Hello, {name}!')\n"
+  ]
+}
+  \end{lstlisting}
+  \captionof{lstlisting}{JSON structure of code cells.}
+  \label{lst:cell-code}
+\end{minipage}
+\hfill
+\begin{minipage}{0.45\textwidth}
+  \begin{lstlisting}
+{
+  "cell_type": "code",
+  "metadata": {},
+  "source": [...],
+  "outputs": [
+    {
+      "output_type": "display_data",
+      "data": {
+        "image/png": "<base64-encoded-png-data>"
+      }
+    }
+  ]
+}
+  \end{lstlisting}
+  \captionof{lstlisting}{JSON structure of code cells with an image output.}
+  \label{lst:cell-output}
+\end{minipage}
 
 The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
-Listing~\ref{lst:nbformat} shows the internal json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
-For simplicity, we only show the parts that are relevant for this study.
+Listing~\ref{lst:nbformat} shows the top-level json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
+For simplicity, we only show the parts that are relevant for this study and hide the irrelevant parts using ellipsis.
 
@@
 
-Listing~\ref{lst:nbformat} presents three examples of cells.
+Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and ~\ref{lst:cell-output} present three examples of cells.
 Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
-Cells can be of three types: "markdown", "code" or "raw".
+Cells can be of three types: ``markdown'', ``code'' or ``raw''.
 The contents of each cell can be found within the \texttt{source} key, as a list of strings.
 
-The output generated by a code cell is stored within the \texttt{outputs} key.
+Outputs generated by a code cell are stored as a list within the \texttt{outputs} key as seen in Listing~\ref{lst:cell-output}.
 Static visualisations are stored as base64 encoded png images under the \texttt{"image/png"} key, which is nested within the \texttt{data} key inside \texttt{outputs}.
-Since a cell can generate multiple outputs, the \texttt{outputs} key is a list.
 
-\subsubsection{Pre and Post-Condition Checks}\ref{sec:pre-post-cond}
-% TODO explain and give examples of what we mean by pre/post-condition checks with examples
+\subsection{Post-Condition Checks}\label{sec:post-cond}
 
-% TODO contrast this to the type of tests we are actually looking for! Why are pre/post conditions not valuable to look at?
+We found several instances of visualisation and assertion pairs which were realated to one another.
+However the assertions did not test a property of the ML system that was derived from the visualisation.
+Instead, the assertions were used to test the correctness of the code that produced the visualisation.
+Such assertions (and the corresponding visualisations) were excluded from this study.
+Table~\ref{tab:post-cond}provides examples of post-condition checks which were frequently observed in this study.
 
-During the empirical analysis conducted for this study, we found several instances of where the assertions were related to the visualisation above.
-However, they were not relevant to this study since they did not test the property of the data or the model.
+\begin{table}
+\centering
+\caption{Frequent examples of post-condition checks observed in this study.}
+\begin{tabular}{p{0.3\textwidth} p{0.4\textwidth} p{0.3\textwidth}}
+\toprule
+\emph{\textbf{Visualisation}}&
+\emph{\textbf{Post-condition check}}&
+\emph{\textbf{Rationale}}\\
+\midrule
+\includegraphics[width=\linewidth]{post-cond-01.png}&
+\begin{lstlisting}[language=Python]
+assert f1.gca().has_data()
+\end{lstlisting}&
+The assertion checks if the output of the visualisation code cell contains an image.\\
+\includegraphics[width=\linewidth]{post-cond-02.png}&
+\begin{lstlisting}[language=Python]
+assert_almost_equal(sigmoid(-50), 0, delta=1e-10)
+assert_almost_equal(sigmoid(0), 0.5, delta=1e-10)
+assert_almost_equal(sigmoid(50), 1, delta=1e-10)
+\end{lstlisting}&
+The assertions are spot-checking the sigmoid activation function.\\
+\includegraphics[width=\linewidth]{post-cond-03.png}&
+\begin{lstlisting}[language=Python]
+# The axis should contain 5 bars
+assert len(ax.containers[0]) == 5
+
+# One of the x tick labels should be "blue"
+tick_text = [tick.get_text() for tick in ax.get_xticklabels()]
+assert "blue" in tick_text
+\end{lstlisting}&
+The asserts are validating the correctness of the visualisation itself.\\
+\bottomrule
+\end{tabular}
+\label{tab:post-cond}
+\end{table}
 
@@
 
-\subsection{Coefficient of Determination}
+\subsection{Coefficient of Determination}\label{sec:r2}
 
@@
 
+\subsection{New form of testing}
+% NOTE some form of commentary on the fact that is form of qualitative and quantitative testing is happening but not that popular yet!
+
+% NOTE so we need more tools to aid practitioners to encourage this form of testing and also help them translate qualitative properties into quantitative tests!
+
 \subsection{From Implicit Expectations to Explicit Assertions}
@@
 
-% NOTE discuss the pros and cons of having a qualiative assessment: they are dense with useful information; humans need visual patterns!
+% NOTE discuss the pros and cons of having a qualitative assessment: they are dense with useful information; humans need visual patterns!
 
@@
 % NOTE now is the time to talk about how assertions act as documentation of the domain expertise and knowledge extracted from the visualisation from the original creator; this also scales as the system complexity and composition of the team/organisation evolves!
+
+% NOTE what are the benefits using visualisations to assess the
+% quality of ML systems? They are packed with a lot of
+% information. And humans are cognitively tuned to identify patterns
+% in visualisations.
+
+Visualisations condense valuable information that is easy to understand for humans.
 

### 2023-09-28 report/esecfse24.tex
feat(report): 2023-09-28 17:29 updates

@@
 
-We found several instances of visualisation and assertion pairs which were realated to one another.
-However the assertions did not test a property of the ML system that was derived from the visualisation.
-Instead, the assertions were used to test the correctness of the code that produced the visualisation.
-Such assertions (and the corresponding visualisations) were excluded from this study.
-Table~\ref{tab:post-cond}provides examples of post-condition checks which were frequently observed in this study.
-
 \begin{table}
@@
 
+We found several instances of visualisation and assertion pairs which were realated to one another.
+However the assertions did not test a property of the ML system that was derived from the visualisation.
+Instead, the assertions were used to test the correctness of the code that produced the visualisation.
+Such assertions (and the corresponding visualisations) were excluded from this study.
+Table~\ref{tab:post-cond}provides examples of post-condition checks which were frequently observed in this study.
+
 \section{Methodology}\label{sec:method}
-This section presents the methodology used to conduct the empirical study introduced in Section~\ref{sec:intro}.
 Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
@@
 Text in double quotes represent the string pattern used in the search query.
-
-More details regarding each phase of the methodology is presented below.
+Each phase of the methodology is explained in more detail below.
 
@@
 
-% TODO don't mention quantity of random sample; then we need to explain/defend this choice
-We used an iterative approach to determine the effectiveness of each
-filter that was used. We applied each filter in the same order as
-presented in Table~\ref{tab:filters}, starting with F1. We randomly
-sampled 50 notebooks from the population of notebooks obtained after
-applying the filter. We then conduct the qualitative assessment
-presented in Section~\ref{REFME} and observe the number of related
-visualisation and assertion pairs. In each new iteration, we
-progressively add the other filters and include the filter if the
-number of relevant visualisation and assertion pairs increases.
+We conduct the manual analysis presented in Section~\ref{sec:identify-related-pairs} for a small random sample of notebooks containing each cell arrangement.
+We observed the signal-to-noise ratio for each cell arrangement and picked the arrangement with the highest signal-to-noise ratio.
+Table~\ref{tab:cell-arrangement} presents the cell arrangement in ascending order of signal-to-noise ratio with a summary of our observations.
 
-As explained above, the selected arrangement of code cells was discovered by qualitatively analysing a random sample of notebooks with the code cell arrangements presented in Figure~\ref{fig:cell-arrangement}.
-We observed the signal-to-noise ratio for each cell arrangement.
-And picked the arrangement with the least signal-to-noise ratio.
-Table~\ref{tab:cell-arrangement} summarises our observations from the various cell arrangements.
+Assertions defined in a cell above the visualisation cell, were typically not related to each other.
+Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
+In other cases, the assertions were typically checks to ensure that the visualisation code would not encounter any errors.
+This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same cell.
+Common examples of such checks include checking the shape of features that are used in the visualisation.
+Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.
 
-% TODO give more examples (from our data) of what a pre/post condition check is
+The signal-to-noise ratio was observed to improve in the sample of notebooks where the assertions were defined below the visualisation code.
+The visualisation and assertions pairs observed in Arrangment 2.2 and 3.1 were often related to each other, but were post-condition checks.
+Section~\ref{sec:post-cond} provides more details on post-condition checks and our rationale for excluding them from this study.
+
+We observed the best results in notebooks where the assertion cell was not longer than 10 lines and was defined below the visualisation cell (Arrangement 3.2).
+We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
+The visualisations were often followed by another markdown cell to record the author's observations.
+And finally, the observations were translated to a analytical test using \texttt{``assert''} or other testing methods from external libraries.
+The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.
+
 \begin{table}
@@
     1 &
-    Assertion above visualisation cell &
-    The assertions in this arrangement were typically not related to the visualisation below.
-Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
-In other cases, the assertions were typically pre-condition checks to ensure that the visualisation code would not encounter any errors.
-Pre-condition checks typically include checking the shape of features that are used in the visualisation.
-Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.\\
+    Assertion cell above visualisation cell &
+    Least likely to be related to the visualisation below.\\
+    \midrule
+    \multicolumn{3}{l}{\textbf{Assertion and visualisation in the same code cell}}\\
+    \midrule
     2.1 &
-    Assertion in the visualisation cell: assertion above visualisation code &
-    Our observations were similar to that of arrangement 1. the assertions defined above the visualistion code were more likely to be pre-condition checks.\\
+    Assertion above visualisation code &
+    More likely to be a pre-condition check.\\
     2.2 &
-    Assertion in the visualisation cell: assertion below visualisation code &
-    We found several examples where the assertion was related to the visualisation.
-Most often however, the assertions were post-condition checks to ensure the correctness of the visualisation itself.
-The signal-to-noise ratio in this arrangement was also higher compared to arrangement 3.2.\\
+    Assertion below visualisation code &
+    More likely to be a post-condition check.\\
+    \midrule
+    \multicolumn{3}{l}{\textbf{Assertion cell below visualisation cell}}\\
+    \midrule
     3.1 &
-    Assertion below visualisation cell: assertion cell more than 10 lines &
-    The assertions in this arrangement were frequently related to the visualisation above.
-In some cases, a markdown cell between the visualisation and assertion marked the begining of a new section in the notebook indicating that the visualisation and assertion were not related to each other.
-We also observed instances where the assertion cell defined a new class or several helper methods that were not related to the visualisation above.
-This motivated our decision to only flag notebooks where the assertion cell was no longer than 10 lines.\\
+    Assertion cell more than 10 lines &
+    Better signal-to-noise ratio compared to Arrangement 2.1. However, More likely to be post-condition check or unrelated to the visualisation above.\\
     3.2 &
-    Assertion below visualisation cell: assertion cell has max 10 lines &
-    This arrangement had the highest signal-to-noise ratio.
-We observed that most notebook authors followed the convension of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
-The visualisations were often followed by another markdown cell to record the author's observations.
-And finally, the observations were translated to a analytical test using \texttt{``assert''} or other testing methods from external libraries.\\
+    Assertion cell has max 10 lines &
+    Most likely to be related to the visualisation above.\\
     \bottomrule
@@
 
-% TODO the numbers don't line up, because the annotations were not perfect. We 161 instances where I did not SELECT the notebook but also did not assign it an EC. How do we handle this?
-% NOTE move the 161 with the 337, then we can say we had 500+ notebooks and we cannot analyse them all in depth...
 The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
-We analysed all 1.7K notebooks manually to further identify 337 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
+We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
 Table~\ref{tab:exclusion-criteria} presents the exclusion criteria used in this phase of the analysis, along with the number of notebooks that were excluded based on each criteria.
+
+We manually analysed the source code of the visualisation and assertion.
+And excluded visualisation and assertion pairs that were not related to each other.
+We encountered several instances where the tracing was not straightforward due to poor coding practices.
+In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
+
+We excluded notebooks that did not train an ML model.
+We found several notebooks authored on topics such as linear algebra, optimisation and loss functions.
+Although these topics are related to ML, the visualisation and assertion pairs found in these notebooks could not be readily applied to a typically ML development workflow.
+Analysing such notebooks also required a working knowledge of all mathematical concepts related to ML.
+This was deemed beyond the scope of software engineering research.
+
+Finally, we excluded notebooks that were not authored in English or contained faulty code that did not produce a visualisation.
+Table~\ref{tab:exclusion-criteria} lists all exclusion criteria used along with the number of notebooks that were excluded from this study.
 
@@
   \centering
-  \caption{Exclusion criteria used during phase 2 to identify related visualisation and assertion pairs.}
-  \begin{tabular}{l p{0.3\textwidth} p{0.4\textwidth} p{0.1\textwidth}}
+  \caption{Exclusion criteria used to identify related visualisation and assertion pairs.}
+  \begin{tabular}{l l r}
     \toprule
@@
     \textbf{Exclusion Criteria} &
-    \textbf{Rationale} &
     \textbf{Num. excluded NBs}\\
@@
     \textbf{EC1} &
-    Notebook not written in English &
-    % TODO find a better way to phrase this?
-    We excluded notebooks that were not authored in English since that is the primary language of all authors of this paper. &
-    159\\
+    The visualisation and assertion are not related &
+    440\\
     \textbf{EC2} &
     Notebook does not contain a trained ML model &
-    % TODO need to explain this better, we have notebooks with diverse python libraries; different ML models work differently so we need to understand the maths behind it.
-    We excluded notebooks that did not train an ML model.
-We found several notebooks that were on topics related to ML such as linear algebra, optimization and loss functions to name a few.
-However, identifying how the visualisation and assertion pairs can be applied to a production ML system, requires a deep understanding of several python libraries and the inner workings of various ML models.
-This was deemed beyond the scope of this research project.
-&
     383\\
-    \textbf{EC2} &
+    \textbf{EC3} &
+    The assertion is a post-condition check &
+    200\\
+    \textbf{EC4} &
+    Notebook not written in English &
+    159\\
+    \textbf{EC5} &
     Notebook contains incorrect code &
-    % TODO our filter excludes NBs without images, we have to defend why this EC is still required: jupyter adds the "image/png" field regardless of error in code
-    We excluded notebooks that contained incorrect code that did not produce a visualisation. We also excluded notebooks that used \texttt{assert} statements to stop execution of a code cell. &
     31\\
-    \textbf{EC3} &
-    % TODO is there technical terminology for what we did to identify if the visualisation & assertion are related to one another?
-    % TODO I think we need a brief paragraph on how we prompted ChatGPT, this stuff probably needs to go outside of a table!
-    The visualisation and assertion are not related &
-    We excluded notebooks where the visualisation and assertion pairs were unrelated.
-We manually analyse the source code of the visualisation and the assertion to determine if they are related to one another.
-In some instances, the tracing was not straightforward or the code was not well written.
-In such cases, we used one-shot prompt engineering with ChatGPT 4.0 model to derive the link between the visualisation and the assertion.
-&
-    440\\
-    \textbf{EC4} &
-    The assertion is a pre or post-condition check &
-    % TODO we have to give examples of pre/post condition checks. I think this content needs to be outside of a table!
-    We excluded &
-    200\\
+    \midrule
+    \multicolumn{2}{c}{\emph{\textbf{Total number of NBs excluded:}}} &
+    1.213\\
     \bottomrule
@@
 
+For the 551 notebooks where
+
+We qualitatively assess the comments made for the
 We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
+
 We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
@@
 % - ecosystem and then doing a search for each of those patterns
+
+% TODO don't mention quantity of random sample; then we need to explain/defend this choice
+We used an iterative approach to determine the effectiveness of each
+filter that was used. We applied each filter in the same order as
+presented in Table~\ref{tab:filters}, starting with F1. We randomly
+sampled 50 notebooks from the population of notebooks obtained after
+applying the filter. We then conduct the qualitative assessment
+presented in Section~\ref{REFME} and observe the number of related
+visualisation and assertion pairs. In each new iteration, we
+progressively add the other filters and include the filter if the
+number of relevant visualisation and assertion pairs increases.
 \section{Conclusion}\label{sec:conclude}

### 2023-10-05 report/icst24.tex
ref(report): rename esecfse24 -> icst24

--- 2023-09-28 report/esecfse24.tex
@@
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
+\documentclass[conference]{IEEEtran}
+\IEEEoverridecommandlockouts
 
@@
 %%\acmSubmissionID{123-A56-BU3}
-
+\usepackage{cite}
+\usepackage{amsmath,amssymb,amsfonts}
+\usepackage{algorithmic}
+\usepackage{graphicx}
+\usepackage{textcomp}
 \usepackage{xcolor}
@@
 \usepackage{subcaption}
+\usepackage{array}
 
@@
 
+\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
+    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
 \begin{document}
 
-\title{Bridging the Gap Between Visual and Analytical Machine Learning Testing}
+\title{Bridging the Gap Between Visual and Analytical Machine Learning
+  Testing}
 
-\author{Anon}
-
-\begin{abstract}
-% what is the problem?
-% why is this an important problem to solve?
-% what did we do?
-% what did we find?
-Testing ML systems is a highly interactive process which demands
-  a human-in-the-loop approach. In addition to writing tests for the
-  code base, practitioners are required to analyse and interpret
-  several visualisations using their domain expertise to validate if
-  an ML system satisfies the required set of functional and
-  non-functional properties. Visualisations are frequently used to
-  qualitatively assess various parts of an ML pipeline. However,
-  implicit knowledge gained from visualisations must be translated to
-  explicit analytical tests that fail when there is change in any
-  component of the ML pipeline. We conduct an empirical analysis of
-  Jupyter notebooks to catalogue the state-of-the-art mappings between
-  ML visualisations and assertions.
-We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
-We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another.
-We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns.
-We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.
-
-\end{abstract}
+\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+\and
+\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
+}
 
@@
 
+\begin{abstract}
+Testing ML systems is a highly interactive process which demands a
+human-in-the-loop approach. In addition to writing tests for the code
+base, practitioners are required to analyse and interpret several
+visualisations using their domain expertise to validate if an ML
+system satisfies the required set of functional and non-functional
+properties. Visualisations are frequently used to qualitatively assess
+various parts of an ML pipeline. However, implicit knowledge gained
+from visualisations must be translated to explicit analytical tests
+that fail when there is change in any component of the ML pipeline. We
+conduct an empirical analysis of Jupyter notebooks to catalogue the
+state-of-the-art mappings between ML visualisations and assertions. We
+mine Github to collect Jupyter Notebooks that contain assertions
+written in Python. We develop a novel methodology to identify 1764
+notebooks which contain an assertion below a visualisation. We
+manually analyse the 1.7K notebooks and identify 551 notebooks where
+at least one pair of visualisation and assertion are semantically
+related to one another. We further read and understand 90 notebooks in
+depth to identify 26 qualitative and their corresponding quantitative
+testing patterns. We showcase five such patterns in this paper and
+plan to make the entire catalogue publicly available in the near
+future.
+\end{abstract}
+
+\begin{IEEEkeywords}
+  ML Testing, Visualisations, ???
+\end{IEEEkeywords}
+
 \section{Introduction}\label{sec:intro}
 
-Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
-With the advent of Large Language Models, the power of AI to directly influence the daily lives of humans is growing at a radical pace.
-As reports of "hallucinated" facts, racist predicts and autonomous car crashes steadily increase, ensuring that ML systems function as they were intended to, takes precedence.
+Recent years show a shift from chasing high correctness of ML models
+to ensuring that they are safe for application in safety-critical
+domains such as healthcare, finance and autonomous driving. With the
+advent of Large Language Models, the power of AI to directly influence
+the daily lives of humans is growing at a radical pace. As reports of
+``hallucinated'' facts, racist predicts and autonomous car crashes
+steadily increase, ensuring that ML systems function as they were
+intended to, takes precedence.
 
-Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
-In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and verify its statistical properties.
-During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
-Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
+Visualisations are used throughout the ML development lifecycle to
+test various properties of the ML system. In the early stages of the
+ML development lifecycle, visualisations are extensively used for make
+sense of the data and verify its statistical properties. During the
+model development phase, visualisations are used to summarise metrics
+and contrast different learning algorithms and iteratively fine-tune
+ML models. Once the model is deployed in production, visualisations
+are used to continually monitor their performance and trigger a new
+training cycle once their performance drops below a certain threshold.
 
-Building ML systems is highly iterative and experimental.
-Information gained from ML visualisations is used to make design and implementation decisions for the following steps of the ML pipeline.
-For instance, we may visualise the distribution of our training data and find that it is normally distributed.
-Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
-However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
-Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
+Building ML systems is highly iterative and experimental. Information
+gained from ML visualisations is used to make design and
+implementation decisions for the following steps of the ML
+pipeline. For instance, we may visualise the distribution of our
+training data and find that it is normally distributed. Based on this
+information, we may opt for a Linear Regression model which assumes
+normality in the underlying data. However, such expectations regarding
+the data may be violated once the ML system is deployed in production,
+where the data constantly changes as a reflect of the real
+world. Implicit expectations obtained from visualisations must
+therefore be translated to analytical tests that fail once our
+expectations regarding the ML system are no longer satisfied.
 
-In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction.
-We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
-In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a qualitative form of testing (using analytical assertions).
-The research questions along with the contributions of this paper are as follows.
+In contrast to prior work done in ML testing, we approach this
+pressing topic from a novel direction. We conduct an empirical
+analysis of computational notebooks obtained from Github, to
+understand the process of testing ML systems in practice. In
+particular, we focus on the combination of a qualitative form of
+testing (using visualisations) and a qualitative form of testing
+(using analytical assertions). The research questions along with the
+contributions of this paper are as follows.
 
 \begin{description}
-\item[RQ1.] How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?
+\item[RQ1.] How frequently do ML practitioners formulate analytical
+  tests from the visualisations they create to test ML systems?
 
-We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
-We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
+We mine 54K Jupyter notebooks from Github that contain an assertion
+    written in Python. We develop an automated technique to identify
+    1764 notebooks with an assertion below a visualisation. We
+    manually analyse the 1.7K notebooks and identify 551 notebooks
+    that contain at least one pair of visualisation and assertion that
+    are semantically related to each other.
 
-\item[RQ2.] What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?
+\item[RQ2.] What are the types of visualisation and assertion pairs
+  typically used by ML practitioners when testing ML systems?
 
-From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks.
-We identify 26 unique visualisation and assertion pairs.
-We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.
-
+From the 551 notebooks, we further perform an in-depth analysis of 90
+    notebooks. We identify 26 unique visualisation and assertion
+    pairs. We discuss five such pairs in this paper and plan to
+    release the full catalogue publicly in the near future.
 \end{description}
@@
 
-The paper is organised as follows.
-Section~\ref{sec:prelim} summarises the related literature and technical information relevant to this study.
-The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
-The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity (Section~\ref{sec:threats}) and future work (Section~\ref{sec:conclude}).
+The paper is organised as follows. Section~\ref{sec:prelim} summarises
+the related literature and technical information relevant to this
+study. The methodology for the empirical study is explained in
+Section~\ref{sec:method} and the results of the study are presented in
+Section~\ref{sec:result}. The paper concludes with a discussion on the
+implications of our results (Section~\ref{sec:discuss}), threats to
+validity (Section~\ref{sec:threats}) and future work
+(Section~\ref{sec:conclude}).
 
@@
 
-This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
-The section concludes with an overview of technical knowledge required for this paper.
+This section summarises the body of prior scientific work that has
+been done in ML Testing, computational notebooks and visual
+analytics. The section concludes with an overview of technical
+knowledge required for this paper.
 
@@
 
-Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase\cite{CITEME}.
-With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
-Existing scientific literature on ML testing broadly focuses on two aspects.
-First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy.
+Testing ML enabled software systems poses more challenges compared to
+their traditional counter part. While traditional software systems
+mature through change in their codebase, ML enabled systems experience
+change in the training data and the machine learning model in addition
+to the codebase\cite{CITEME}. With ML being adopted into
+safety-critical domains that can affect human lives, ensuring that ML
+systems are correct, robust to data perturbations and not biased by
+race, colour or gender is of paramount importance. Existing scientific
+literature on ML testing broadly focuses on two aspects. First, on
+functional properties such as correctness and robustness of the model
+towards unseen data. And second, on non-functional properties such as
+fairness, interpretability and privacy.
 
-To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
-Tools such as \textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepexplore, ma2018deepgauge, gerasimou2020importance}.
-Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
-Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
-Several works have been conducted on generating and detecting adversarial inputs for ML models.
-Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
-It is however not possible to include all variations of adversarial examples into the training data.
-Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
+To test the correctness of ML models, several improvements over
+existing test adequacy metrics have been proposed. Tools such as
+\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
+metrics such as neuron coverage adapted for ML enabled software
+systems~\cite{pei2017deepexplore, ma2018deepgauge,
+gerasimou2020importance}. Formal verification methods have also been
+proposed that try to provide formal guarantee of robustness against
+adversarial examples. Such methods however are only feasible for
+statistical ML models and become computationally expensive for more
+complex models such as deep neural networks~\cite{zhu2021deepmemory,
+baluta2021scalable}. Several works have been conducted on generating
+and detecting adversarial inputs for ML models. Data augmentation
+techniques based on fuzzing, search based software testing and
+mutation testing have been proposed to generate adversarial examples
+that can be used during model training to improve its
+robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
+zhang2020white}. It is however not possible to include all variations
+of adversarial examples into the training data. Thus, methods have
+been proposed to detect adversarial inputs during
+runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
+berend2020cats}.
 
 % TODO summarise literature on fairness testing
-Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.
+Bulk of the contributions on testing non-functional properties of ML
+systems, focuses on fairness.
 
@@
 
-Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large}.
+Computational notebooks have been ubiquitously adopted by the machine
+learning community for developing ML enabled systems. Although
+originally intended to promote reproducible software, computational
+notebooks are far from being reproducible due to lack of software
+engineering best practices such as separation of concern, testing and
+versioning~\cite{pimentel2019large}.
 
-\citeauthor{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, and source code and metadata from over 900 releases of 12 important data science libraries.
-Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
-\citeauthor{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices in computational notebooks.
-Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
-The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
+\cite{psallidas2019data} provide an overview of the evolving landscape
+of computational notebooks by analysing six million Python notebooks,
+two million enterprise data science pipelines, and source code and
+metadata from over 900 releases of 12 important data science
+libraries. Their findings can be used by system builders for improving
+data science tools and also by data science practitioners to
+understand the current trends and what technologies they should focus
+on. \cite{pimentel2019large} mined 1.4 million notebooks from Github
+to conduct an empirical study on the good and bad coding practices in
+computational notebooks. Based on their analysis, the authors propose
+guidelines on improving reproducibility of computational notebooks. To
+enable future research on computational notebooks,
+\cite{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com}
+and present \textit{KGTorrent}, a public dataset consisting of
+approximately 2.5 million Jupyter notebooks written in Python. The
+dataset also contains a relational database dump of metadata regarding
+publicly available notebooks on Kaggle.
 
-Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
-Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
-The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
+Studies with human subjects have been conducted to gain a deeper
+understanding of the challenges faced by ML practitioners when
+developing ML models inside notebooks. The results indicate that ML
+practitioners work in a highly iterative fashion, often experimenting
+with multiple strategies to analyse the data or produce meaningful
+visualisations~\cite{kandel2012enterprise, kery2018story,
+liu2019understanding, chattopadhyay2020wrong}. Studies have also been
+conducted to understand how practitioners generate, evaluate and
+manage alternative hypothesis, visual designs, methods, tools,
+algorithms and data sources to arrive at the final
+implementation~\cite{liu2019understanding,kandel2012enterprise}. The
+findings from these studies can be used as guidelines for improving
+existing notebook technologies or designing new ones.
 
-To manage and prune the multiple versions of code that accumulate over time when developing in notebooks, \citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
-Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
+To manage and prune the multiple versions of code that accumulate over
+time when developing in notebooks, \cite{head2019managing} propose
+a code gathering tool that allows practitioners to review and only
+keep the relevant version of code. Other tools such as
+\textit{WrangleDoc} and \textit{VizSmith} have also been proposed to
+aid ML practitioners when working within computational
+notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
 
@@
 
-As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
-ML models and the underlying data is complex and multi-dimensional.
-To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models.
+As ML augment software systems in safety-critical domains, emphasis
+has been put into explainability of ML models. ML models and the
+underlying data is complex and multi-dimensional. To combat the
+``curse of dimensionality'', visual analytics has been widely adopted
+by the ML community to understand the data and the internal workings
+of ML models.
 
-Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\citeauthor{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\citeauthor{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
-The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
+Prior studies have been conducted to understand how visual analytics
+techniques are currently being used in ML. \cite{yuan2021survey}
+conduct a systematic review of 259 papers and propose a taxonomy of
+visual analytics techniques for ML. \cite{hohman2019visual} conduct
+a survey of visual analytics techniques for Deep Learning Models. The
+findings of the study indicate that visual analytics has been widely
+adopted in ML for model explanation, interpretation, debugging and
+improvement.
 
-Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\citeauthor{wexler2020if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
-To reduce the cognitive load of ML practitioners during model building phase, \citeauthor{amershi2015modeltracker} propose \textit{ModelTracker}.
-Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
+Several tools have been proposed by the visual analytics research
+community to aid practitioners in understanding how their ML models
+operate. \cite{wexler2020if} propose \textit{What-If}, a visual
+analytics tool to explore alternative hypotheses, generate
+counterfactual reasoning, investigate decision boundaries of the model
+and how change in data affects the model predictions. To reduce the
+cognitive load of ML practitioners during model building phase,
+\cite{amershi2015modeltracker} propose \textit{ModelTracker}. Given
+a trained model and a sample dataset, ModelTracker presents all the
+information in traditional summary statistics along with the
+performance of the model.
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systemmatic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
+\textit{Drava} are a few other tools that have been proposed to handle
+specialised use-cases such as identifying systemmatic errors in ML
+models, generating counterfactual explanations for Generalised
+Additive Models, prioritising machine translation model improvements
+and relating human concepts with semantic dimensions extracted by ML
+models during disentangled representation learning\cite{ahn2023escape,
+wang2023gam, robertson2023angler, wang2023drava}.
 
@@
   \end{lstlisting}
-  \captionof{lstlisting}{Top-level JSON format of Jupyter notebooks. See Listing~\ref{lst:cell-md}, \ref{lst:cell-code} and \ref{lst:cell-output} for internal structure of various cells.}
-  \label{lst:nbformat}
+  \captionof{lstlisting}{Top-level JSON format of Jupyter notebooks.
+    See Listing~\ref{lst:cell-md},~\ref{lst:cell-code}
+    and~\ref{lst:cell-output} for internal structure of various
+    cells.}\label{lst:nbformat}
 \end{minipage}
@@
 
-The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
-Listing~\ref{lst:nbformat} shows the top-level json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
-For simplicity, we only show the parts that are relevant for this study and hide the irrelevant parts using ellipsis.
+The data comprised within Jupyter Notebooks is stored in the JSON
+format, and thus can be parsed
+programmatically. Listing~\ref{lst:nbformat} shows the top-level json
+structure of Jupyter Notebooks, adopted from the official nbformat
+website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}. For
+simplicity, we only show the parts that are relevant for this study
+and hide the irrelevant parts using ellipsis.
 
-Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
-The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
-\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
-The \texttt{cells} key consists of a list of all cells within the notebook.
+Notebooks contain four top level keys namely \texttt{metadata},
+\texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}. The
+\texttt{matadata} key consists of additional information regarding the
+notebook such as the kernel and programming
+language. \texttt{nbformat} and \texttt{nbformat\_minor} contain the
+version of nbformat used in the notebook. The \texttt{cells} key
+consists of a list of all cells within the notebook.
 
-Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and ~\ref{lst:cell-output} present three examples of cells.
-Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
-Cells can be of three types: ``markdown'', ``code'' or ``raw''.
-The contents of each cell can be found within the \texttt{source} key, as a list of strings.
+Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and
+~\ref{lst:cell-output} present three examples of cells. Each cell is
+represented as a dictionary with the \texttt{cell\_type},
+\texttt{metadata} and \texttt{source} keys. Cells can be of three
+types: ``markdown'', ``code'' or ``raw''. The contents of each cell
+can be found within the \texttt{source} key, as a list of strings.
 
-Outputs generated by a code cell are stored as a list within the \texttt{outputs} key as seen in Listing~\ref{lst:cell-output}.
-Static visualisations are stored as base64 encoded png images under the \texttt{"image/png"} key, which is nested within the \texttt{data} key inside \texttt{outputs}.
+Outputs generated by a code cell are stored as a list within the
+\texttt{outputs} key as seen in Listing~\ref{lst:cell-output}. Static
+visualisations are stored as base64 encoded png images under the
+\texttt{"image/png"} key, which is nested within the \texttt{data} key
+inside \texttt{outputs}.
 
@@
 
-\begin{table}
+\begin{table*}
 \centering
 \caption{Frequent examples of post-condition checks observed in this study.}
-\begin{tabular}{p{0.3\textwidth} p{0.4\textwidth} p{0.3\textwidth}}
-\toprule
+  \begin{tabular}{m{0.3\textwidth} m{0.4\textwidth} m{0.3\textwidth}}
+\hline
 \emph{\textbf{Visualisation}}&
@@
 \emph{\textbf{Rationale}}\\
-\midrule
+\hline
 \includegraphics[width=\linewidth]{post-cond-01.png}&
@@
 The asserts are validating the correctness of the visualisation itself.\\
-\bottomrule
+\hline
 \end{tabular}
 \label{tab:post-cond}
-\end{table}
+\end{table*}
 
-We found several instances of visualisation and assertion pairs which were realated to one another.
-However the assertions did not test a property of the ML system that was derived from the visualisation.
-Instead, the assertions were used to test the correctness of the code that produced the visualisation.
-Such assertions (and the corresponding visualisations) were excluded from this study.
-Table~\ref{tab:post-cond}provides examples of post-condition checks which were frequently observed in this study.
+We found several instances of visualisation and assertion pairs which
+were realated to one another. However the assertions did not test
+a property of the ML system that was derived from the
+visualisation. Instead, the assertions were used to test the
+correctness of the code that produced the visualisation. Such
+assertions (and the corresponding visualisations) were excluded from
+this study. Table~\ref{tab:post-cond}provides examples of
+post-condition checks which were frequently observed in this study.
 
 \section{Methodology}\label{sec:method}
-Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
-The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.
+Figure~\ref{fig:method} presents an overview of all relevant steps
+used in this study. The empirical analysis comprises of three distinct
+phases namely: \textit{the data collection phase},
+\textit{identification of related visualisation and assertion pairs}
+and finally a \textit{complete analysis of the notebooks} to identify
+the final set of 26 visualisation and assertion pairs.
 
-The yellow boxes in the data collection phase represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
-The red boxes represent the exclusion criteria used during phase two to remove notebooks note relevant for this study.
-Finally, the green boxes in phase 3 represent the manual analysis conducted to identify the final 26 visualisation and assertion pairs.
+The yellow boxes in the data collection phase represent filters that
+were applied to the initial corpus of 54K notebooks, to obtain the
+final sample size of 1.7K notebooks used in the empirical study. The
+red boxes represent the exclusion criteria used during phase two to
+remove notebooks note relevant for this study. Finally, the green
+boxes in phase 3 represent the manual analysis conducted to identify
+the final 26 visualisation and assertion pairs.
 
-Boxes appearning next to one-another imply that they were applied using an "OR" operator.
-While boxes appearning below one-another imply an "AND" operator.
-Text in double quotes represent the string pattern used in the search query.
-Each phase of the methodology is explained in more detail below.
+Boxes appearning next to one-another imply that they were applied
+using an "OR" operator. While boxes appearning below one-another imply
+an "AND" operator. Text in double quotes represent the string pattern
+used in the search query. Each phase of the methodology is explained
+in more detail below.
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:method}
-\end{figure}
+\end{figure*}
 
@@
 
-We mined public repositories from Github to collect Jupyter notebooks that contain the keyword ``assert'' in them.
-We intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-This is because, the ``assert'' keyword is used in python to define tests.
-Additionally, popular python libraries which provide a testing interface (such as nose, unittest and numpy), contain the same keyword in their API.
-Our search criteria captures a large initial sample of $54,070$ notebooks with a variety of testing statements.
-Our approach also prevents the need to craft a custom search query based on an exhaustive list of keywords that appear in all python testing libraries.
+We mined public repositories from Github to collect Jupyter notebooks
+that contain the keyword ``assert'' in them. We intentionally keep the
+search criteria on Github broad by only checking for the appropriate
+filetype and the presence of the ``assert'' keyword. This is because,
+the ``assert'' keyword is used in python to define
+tests. Additionally, popular python libraries which provide a testing
+interface (such as nose, unittest and numpy), contain the same keyword
+in their API. Our search criteria captures a large initial sample of
+$54,070$ notebooks with a variety of testing statements. Our approach
+also prevents the need to craft a custom search query based on an
+exhaustive list of keywords that appear in all python testing
+libraries.
 
-We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook.
-This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}.
-Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
-The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
+We use the Github advanced search syntax~\footnote{REFME} to isolate
+Jupyter Notebook. This translates to the following Github search
+query: \texttt{filename:"Jupyter Notebooks"}. Additionally, we include
+the ``assert'' keyword in the search query with an \texttt{AND}
+operator. The final search query is as follows:
+\texttt{filename:"Jupyter Notebooks" AND "assert"}.
 
-The above search query results in approximately 54K notebooks.
-To further reduce the sample size, we exclude notebooks that do not use popular ML libraries.
-We perform a string search to identify notebooks that import at least one of the ML libraries.
-The string patterns are derived from the python module name used in the import statements.
-The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
+The above search query results in approximately 54K notebooks. To
+further reduce the sample size, we exclude notebooks that do not use
+popular ML libraries. We perform a string search to identify notebooks
+that import at least one of the ML libraries. The string patterns are
+derived from the python module name used in the import statements. The
+regex search query is as follows:
+\texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
 
-We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis.
-We map each cell as a row in the dataframe.
-For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
-We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.
-The internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}.
+We programmatically parse the JSON structure of the notebooks and
+represent each notebook as a pandas dataframe for further analysis. We
+map each cell as a row in the dataframe. For each cell, we store the
+\texttt{cell\_type}, the \texttt{source} of the cell and the base64
+encoded PNG string, into the corresponding columns of the
+dataframe. We exclude notebooks that do not contain any visualisations
+by checking if the image column of the dataframe is empty. The
+internal structure of Jupyter notebooks is presented in
+Section~\ref{sec:nbformat}.
 
-\begin{figure}
+\begin{figure*}
   \centering
@@
   \label{fig:cell-arrangement}
-\end{figure}
+\end{figure*}
 
-For RQ1, we are interested in finding empirical evidence of visualisations and corresponding assertions used in the wild.
-To answer this question, we further reduce our notebook sample size by assuming a specific order in which the visualisation and assertion code cells appear in the notebooks.
-Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks.
-Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis.
-In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
+For RQ1, we are interested in finding empirical evidence of
+visualisations and corresponding assertions used in the wild. To
+answer this question, we further reduce our notebook sample size by
+assuming a specific order in which the visualisation and assertion
+code cells appear in the notebooks. Figure~\ref{fig:cell-arrangement}
+presents a visual representation of various arrangements of code cells
+that may occur in notebooks. Note that the visual representation omits
+markdown cells since we only consider the order of code cells in our
+analysis. In the actual notebook however, one or more markdown cells
+may be present between the visualisation and assertion cells.
 
-We conduct the manual analysis presented in Section~\ref{sec:identify-related-pairs} for a small random sample of notebooks containing each cell arrangement.
-We observed the signal-to-noise ratio for each cell arrangement and picked the arrangement with the highest signal-to-noise ratio.
-Table~\ref{tab:cell-arrangement} presents the cell arrangement in ascending order of signal-to-noise ratio with a summary of our observations.
+We conduct the manual analysis presented in
+Section~\ref{sec:identify-related-pairs} for a small random sample of
+notebooks containing each cell arrangement. We observed the
+signal-to-noise ratio for each cell arrangement and picked the
+arrangement with the highest signal-to-noise
+ratio. Table~\ref{tab:cell-arrangement} presents the cell arrangement
+in ascending order of signal-to-noise ratio with a summary of our
+observations.
 
-Assertions defined in a cell above the visualisation cell, were typically not related to each other.
-Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
-In other cases, the assertions were typically checks to ensure that the visualisation code would not encounter any errors.
-This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same cell.
-Common examples of such checks include checking the shape of features that are used in the visualisation.
-Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.
+Assertions defined in a cell above the visualisation cell, were
+typically not related to each other. Often, a markdown cell present
+between the assertion and visualisation code cells, would mark the
+begining of a new section in the notebook. In other cases, the
+assertions were typically checks to ensure that the visualisation code
+would not encounter any errors. This observation also holds for
+Arrangement 2.1, where the assertion is defined above the
+visualisation code, in the same cell. Common examples of such checks
+include checking the shape of features that are used in the
+visualisation. Or checking that the length of the ``x''and ``y``
+features are the same to ensure a continuous line in the plot.
 
-The signal-to-noise ratio was observed to improve in the sample of notebooks where the assertions were defined below the visualisation code.
-The visualisation and assertions pairs observed in Arrangment 2.2 and 3.1 were often related to each other, but were post-condition checks.
-Section~\ref{sec:post-cond} provides more details on post-condition checks and our rationale for excluding them from this study.
+The signal-to-noise ratio was observed to improve in the sample of
+notebooks where the assertions were defined below the visualisation
+code. The visualisation and assertions pairs observed in Arrangment
+2.2 and 3.1 were often related to each other, but were post-condition
+checks. Section~\ref{sec:post-cond} provides more details on
+post-condition checks and our rationale for excluding them from this
+study.
 
-We observed the best results in notebooks where the assertion cell was not longer than 10 lines and was defined below the visualisation cell (Arrangement 3.2).
-We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
-The visualisations were often followed by another markdown cell to record the author's observations.
-And finally, the observations were translated to a analytical test using \texttt{``assert''} or other testing methods from external libraries.
-The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.
+We observed the best results in notebooks where the assertion cell was
+not longer than 10 lines and was defined below the visualisation cell
+(Arrangement 3.2). We observed that most notebook authors followed the
+convention of writing their motivation for the visualisation in
+a markdown cell, before writing the code for the visualisation. The
+visualisations were often followed by another markdown cell to record
+the author's observations. And finally, the observations were
+translated to a analytical test using \texttt{``assert''} or other
+testing methods from external libraries. The restriction on the number
+of lines in the assertion cell was introduced to exclude instances
+where the assertion cell contained a new class definition or several
+helper functions that were not related to the visualisation above.
 
@@
   cell arrangement.}
-  \begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
-    \toprule
+  \begin{tabular}{l p{0.2\textwidth} p{0.2\textwidth}}
+    \hline
     \emph{\textbf{ID}}&
@@
     \emph{\textbf{Observations}}\\
-    \midrule
+    \hline
     1 &
@@
     Least likely to be related to the visualisation below.\\
-    \midrule
+    \hline
     \multicolumn{3}{l}{\textbf{Assertion and visualisation in the same code cell}}\\
-    \midrule
+    \hline
     2.1 &
@@
     More likely to be a post-condition check.\\
-    \midrule
+    \hline
     \multicolumn{3}{l}{\textbf{Assertion cell below visualisation cell}}\\
-    \midrule
+    \hline
     3.1 &
@@
     Most likely to be related to the visualisation above.\\
-    \bottomrule
+    \hline
   \end{tabular}
@@
 
-The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
-We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
-Table~\ref{tab:exclusion-criteria} presents the exclusion criteria used in this phase of the analysis, along with the number of notebooks that were excluded based on each criteria.
+The initial sample size of 54K notebooks from Github, was reduced to
+the final sample size of 1.7K after applying the filters presented in
+Section~\ref{sec:data-collect}. We analysed all 1.7K notebooks
+manually to further identify 551 notebooks that contained at least one
+pair of visualisation and assertions that were related to each
+other. Table~\ref{tab:exclusion-criteria} presents the exclusion
+criteria used in this phase of the analysis, along with the number of
+notebooks that were excluded based on each criteria.
 
-We manually analysed the source code of the visualisation and assertion.
-And excluded visualisation and assertion pairs that were not related to each other.
-We encountered several instances where the tracing was not straightforward due to poor coding practices.
-In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
+We manually analysed the source code of the visualisation and
+assertion. And excluded visualisation and assertion pairs that were
+not related to each other. We encountered several instances where the
+tracing was not straightforward due to poor coding practices. In such
+cases, we used one-shot prompt engineering~\cite{CITME} with the
+ChatGPT 4.0 model to determine if the visualisation and assertion were
+related to each other.
 
-We excluded notebooks that did not train an ML model.
-We found several notebooks authored on topics such as linear algebra, optimisation and loss functions.
-Although these topics are related to ML, the visualisation and assertion pairs found in these notebooks could not be readily applied to a typically ML development workflow.
-Analysing such notebooks also required a working knowledge of all mathematical concepts related to ML.
-This was deemed beyond the scope of software engineering research.
+We excluded notebooks that did not train an ML model. We found several
+notebooks authored on topics such as linear algebra, optimisation and
+loss functions. Although these topics are related to ML, the
+visualisation and assertion pairs found in these notebooks could not
+be readily applied to a typically ML development workflow. Analysing
+such notebooks also required a working knowledge of all mathematical
+concepts related to ML. This was deemed beyond the scope of software
+engineering research.
 
-Finally, we excluded notebooks that were not authored in English or contained faulty code that did not produce a visualisation.
-Table~\ref{tab:exclusion-criteria} lists all exclusion criteria used along with the number of notebooks that were excluded from this study.
+Finally, we excluded notebooks that were not authored in English or
+contained faulty code that did not produce
+a visualisation. Table~\ref{tab:exclusion-criteria} lists all
+exclusion criteria used along with the number of notebooks that were
+excluded from this study.
 
@@
   \caption{Exclusion criteria used to identify related visualisation and assertion pairs.}
-  \begin{tabular}{l l r}
-    \toprule
+  \begin{tabular}{l p{0.2\textwidth} r}
+    \hline
     \textbf{ID} &
@@
     \textbf{Num. excluded NBs}\\
-    \midrule
+    \hline
     \textbf{EC1} &
@@
     31\\
-    \midrule
+    \hline
     \multicolumn{2}{c}{\emph{\textbf{Total number of NBs excluded:}}} &
     1.213\\
-    \bottomrule
+    \hline
   \end{tabular}
@@
 
-For the 551 notebooks where
+We qualitatively assess the comments made for all notebooks during the
+analysis conducted in phase two of the study, as described in
+Section~\ref{sec:identify-related-pairs}.
 
-We qualitatively assess the comments made for the
-We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
-
-We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
-The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
-During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data.
-We manually read and comprehend each code cell in the notebooks.
-When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
-For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation.
-A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
+We perform an in-depth analysis of 90 short-listed notebooks to
+identify unique ML testing tasks. The in-depth analysis was conducted
+by reading and understanding the entire contents of the notebooks in
+a top-down order. During the analysis, we note several data points
+such as the primary objective of the notebook, the type of ML problem,
+the type of ML model and the type of data. We manually read and
+comprehend each code cell in the notebooks. When present, we also read
+the markdown cells above and below code cells to gain further context
+from the notebook author. For instance, we find that markdown cells
+above a visualisation cell typically motivates the need for the
+visualisation. A markdown cell directly below a visualisation often
+records the observations made by the notebook author from the
+visualisation.
 
@@
 
-This study identified 26 patterns of testing using visualisation and assertion.
-This section highlights five interesting patterns in more detail.
-We plan to release the entire catalogue of patterns publicly in the near future.
+This study identified 26 patterns of testing using visualisation and
+assertion. This section highlights five interesting patterns in more
+detail. We plan to release the entire catalogue of patterns publicly
+in the near future.
 
@@
 
-% NOTE comment on the over-whelming presence of magic thresholds in ML testing! where are authors getting these numbers from? link this to requirements engineering?
+% NOTE comment on the over-whelming presence of magic thresholds in ML
+% testing! where are authors getting these numbers from? link this to
+% requirements engineering?
 
@@
 
-\bibliographystyle{ACM-Reference-Format}
+\bibliographystyle{IEEEtran}
 \bibliography{bibliography}
+
 \end{document}
+\endinput

### 2023-10-12 report/icst24.tex
feat(report): 2023-10-12 init implications section

@@
 \usepackage{array}
+\usepackage{framed}
+\newcommand{\highlight}[1]{\begin{framed}%
+  \noindent\emph{#1}
+\end{framed}}
 
@@
 \begin{abstract}
-Testing ML systems is a highly interactive process which demands a
-human-in-the-loop approach. In addition to writing tests for the code
-base, practitioners are required to analyse and interpret several
-visualisations using their domain expertise to validate if an ML
-system satisfies the required set of functional and non-functional
-properties. Visualisations are frequently used to qualitatively assess
-various parts of an ML pipeline. However, implicit knowledge gained
-from visualisations must be translated to explicit analytical tests
-that fail when there is change in any component of the ML pipeline. We
-conduct an empirical analysis of Jupyter notebooks to catalogue the
-state-of-the-art mappings between ML visualisations and assertions. We
-mine Github to collect Jupyter Notebooks that contain assertions
-written in Python. We develop a novel methodology to identify 1764
-notebooks which contain an assertion below a visualisation. We
-manually analyse the 1.7K notebooks and identify 551 notebooks where
-at least one pair of visualisation and assertion are semantically
-related to one another. We further read and understand 90 notebooks in
-depth to identify 26 qualitative and their corresponding quantitative
-testing patterns. We showcase five such patterns in this paper and
-plan to make the entire catalogue publicly available in the near
-future.
+
+Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
+In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
+Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
+However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
+We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
+We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
+We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
+We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another.
+We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns.
+We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.
+
+% TODO what were the primary findings of this study?
+
 \end{abstract}
@@
 
-Recent years show a shift from chasing high correctness of ML models
-to ensuring that they are safe for application in safety-critical
-domains such as healthcare, finance and autonomous driving. With the
-advent of Large Language Models, the power of AI to directly influence
-the daily lives of humans is growing at a radical pace. As reports of
-``hallucinated'' facts, racist predicts and autonomous car crashes
-steadily increase, ensuring that ML systems function as they were
-intended to, takes precedence.
+Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
+With the advent of Large Language Models, the power of AI to directly influence the daily lives of humans is growing at a radical pace.
+As reports of ``hallucinated'' facts, racist predicts and autonomous car crashes steadily increase, ensuring that ML systems function as they were intended to, takes precedence.
 
-Visualisations are used throughout the ML development lifecycle to
-test various properties of the ML system. In the early stages of the
-ML development lifecycle, visualisations are extensively used for make
-sense of the data and verify its statistical properties. During the
-model development phase, visualisations are used to summarise metrics
-and contrast different learning algorithms and iteratively fine-tune
-ML models. Once the model is deployed in production, visualisations
-are used to continually monitor their performance and trigger a new
-training cycle once their performance drops below a certain threshold.
+Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
+In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
+During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
+Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
 
-Building ML systems is highly iterative and experimental. Information
-gained from ML visualisations is used to make design and
-implementation decisions for the following steps of the ML
-pipeline. For instance, we may visualise the distribution of our
-training data and find that it is normally distributed. Based on this
-information, we may opt for a Linear Regression model which assumes
-normality in the underlying data. However, such expectations regarding
-the data may be violated once the ML system is deployed in production,
-where the data constantly changes as a reflect of the real
-world. Implicit expectations obtained from visualisations must
-therefore be translated to analytical tests that fail once our
-expectations regarding the ML system are no longer satisfied.
+Building ML systems is highly iterative and experimental.
+Information gained from ML visualisations is used to make design and implementation decisions for the following steps of the ML pipeline.
+For instance, we may visualise the distribution of our training data and find that it is normally distributed.
+Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
+However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
+Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
 
-In contrast to prior work done in ML testing, we approach this
-pressing topic from a novel direction. We conduct an empirical
-analysis of computational notebooks obtained from Github, to
-understand the process of testing ML systems in practice. In
-particular, we focus on the combination of a qualitative form of
-testing (using visualisations) and a qualitative form of testing
-(using analytical assertions). The research questions along with the
-contributions of this paper are as follows.
+In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction.
+We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
+In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions).
+The research questions along with the contributions of this paper are as follows.
 
 \begin{description}
-\item[RQ1.] How frequently do ML practitioners formulate analytical
-  tests from the visualisations they create to test ML systems?
+\item[RQ1.] How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?
 
-We mine 54K Jupyter notebooks from Github that contain an assertion
-    written in Python. We develop an automated technique to identify
-    1764 notebooks with an assertion below a visualisation. We
-    manually analyse the 1.7K notebooks and identify 551 notebooks
-    that contain at least one pair of visualisation and assertion that
-    are semantically related to each other.
+We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
+We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
+We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
 
-\item[RQ2.] What are the types of visualisation and assertion pairs
-  typically used by ML practitioners when testing ML systems?
+\item[RQ2.] What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?
 
-From the 551 notebooks, we further perform an in-depth analysis of 90
-    notebooks. We identify 26 unique visualisation and assertion
-    pairs. We discuss five such pairs in this paper and plan to
-    release the full catalogue publicly in the near future.
+From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks.
+We identify 26 unique visualisation and assertion pairs.
+We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.
+
 \end{description}
@@
 
-The paper is organised as follows. Section~\ref{sec:prelim} summarises
-the related literature and technical information relevant to this
-study. The methodology for the empirical study is explained in
-Section~\ref{sec:method} and the results of the study are presented in
-Section~\ref{sec:result}. The paper concludes with a discussion on the
-implications of our results (Section~\ref{sec:discuss}), threats to
-validity (Section~\ref{sec:threats}) and future work
-(Section~\ref{sec:conclude}).
+The paper is organised as follows.
+Section~\ref{sec:prelim} summarises the related literature and technical information relevant to this study.
+The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
+The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity (Section~\ref{sec:threats}) and future work (Section~\ref{sec:conclude}).
 
@@
 
-This section summarises the body of prior scientific work that has
-been done in ML Testing, computational notebooks and visual
-analytics. The section concludes with an overview of technical
-knowledge required for this paper.
+This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
+The section concludes with an overview of technical knowledge required for this paper.
 
@@
 
-Testing ML enabled software systems poses more challenges compared to
-their traditional counter part. While traditional software systems
-mature through change in their codebase, ML enabled systems experience
-change in the training data and the machine learning model in addition
-to the codebase\cite{CITEME}. With ML being adopted into
-safety-critical domains that can affect human lives, ensuring that ML
-systems are correct, robust to data perturbations and not biased by
-race, colour or gender is of paramount importance. Existing scientific
-literature on ML testing broadly focuses on two aspects. First, on
-functional properties such as correctness and robustness of the model
-towards unseen data. And second, on non-functional properties such as
-fairness, interpretability and privacy.
+Testing ML enabled software systems poses more challenges compared to their traditional counter part.
+While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase\cite{CITEME}.
+With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
+Existing scientific literature on ML testing broadly focuses on two aspects.
+First, on functional properties such as correctness and robustness of the model towards unseen data.
+And second, on non-functional properties such as fairness, interpretability and privacy.
 
-To test the correctness of ML models, several improvements over
-existing test adequacy metrics have been proposed. Tools such as
-\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
-metrics such as neuron coverage adapted for ML enabled software
-systems~\cite{pei2017deepexplore, ma2018deepgauge,
-gerasimou2020importance}. Formal verification methods have also been
-proposed that try to provide formal guarantee of robustness against
-adversarial examples. Such methods however are only feasible for
-statistical ML models and become computationally expensive for more
-complex models such as deep neural networks~\cite{zhu2021deepmemory,
-baluta2021scalable}. Several works have been conducted on generating
-and detecting adversarial inputs for ML models. Data augmentation
-techniques based on fuzzing, search based software testing and
-mutation testing have been proposed to generate adversarial examples
-that can be used during model training to improve its
-robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
-zhang2020white}. It is however not possible to include all variations
-of adversarial examples into the training data. Thus, methods have
-been proposed to detect adversarial inputs during
-runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
-berend2020cats}.
+To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
+Tools such as \textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepexplore, ma2018deepgauge, gerasimou2020importance}.
+Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
+Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
+Several works have been conducted on generating and detecting adversarial inputs for ML models.
+Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
+It is however not possible to include all variations of adversarial examples into the training data.
+Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
 
 % TODO summarise literature on fairness testing
-Bulk of the contributions on testing non-functional properties of ML
-systems, focuses on fairness.
+%
+Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.
 
@@
 
-Computational notebooks have been ubiquitously adopted by the machine
-learning community for developing ML enabled systems. Although
-originally intended to promote reproducible software, computational
-notebooks are far from being reproducible due to lack of software
-engineering best practices such as separation of concern, testing and
-versioning~\cite{pimentel2019large}.
+Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
+Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large}.
 
-\cite{psallidas2019data} provide an overview of the evolving landscape
-of computational notebooks by analysing six million Python notebooks,
-two million enterprise data science pipelines, and source code and
-metadata from over 900 releases of 12 important data science
-libraries. Their findings can be used by system builders for improving
-data science tools and also by data science practitioners to
-understand the current trends and what technologies they should focus
-on. \cite{pimentel2019large} mined 1.4 million notebooks from Github
-to conduct an empirical study on the good and bad coding practices in
-computational notebooks. Based on their analysis, the authors propose
-guidelines on improving reproducibility of computational notebooks. To
-enable future research on computational notebooks,
-\cite{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com}
-and present \textit{KGTorrent}, a public dataset consisting of
-approximately 2.5 million Jupyter notebooks written in Python. The
-dataset also contains a relational database dump of metadata regarding
-publicly available notebooks on Kaggle.
+\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, and source code and metadata from over 900 releases of 12 important data science libraries.
+Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
+\cite{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices in computational notebooks.
+Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
+To enable future research on computational notebooks, \cite{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
 
-Studies with human subjects have been conducted to gain a deeper
-understanding of the challenges faced by ML practitioners when
-developing ML models inside notebooks. The results indicate that ML
-practitioners work in a highly iterative fashion, often experimenting
-with multiple strategies to analyse the data or produce meaningful
-visualisations~\cite{kandel2012enterprise, kery2018story,
-liu2019understanding, chattopadhyay2020wrong}. Studies have also been
-conducted to understand how practitioners generate, evaluate and
-manage alternative hypothesis, visual designs, methods, tools,
-algorithms and data sources to arrive at the final
-implementation~\cite{liu2019understanding,kandel2012enterprise}. The
-findings from these studies can be used as guidelines for improving
-existing notebook technologies or designing new ones.
+Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
+Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
+The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
 
-To manage and prune the multiple versions of code that accumulate over
-time when developing in notebooks, \cite{head2019managing} propose
-a code gathering tool that allows practitioners to review and only
-keep the relevant version of code. Other tools such as
-\textit{WrangleDoc} and \textit{VizSmith} have also been proposed to
-aid ML practitioners when working within computational
-notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
+To manage and prune the multiple versions of code that accumulate over time when developing in notebooks, \cite{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
 
@@
 
-As ML augment software systems in safety-critical domains, emphasis
-has been put into explainability of ML models. ML models and the
-underlying data is complex and multi-dimensional. To combat the
-``curse of dimensionality'', visual analytics has been widely adopted
-by the ML community to understand the data and the internal workings
-of ML models.
+As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
+ML models and the underlying data is complex and multi-dimensional.
+To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models.
 
-Prior studies have been conducted to understand how visual analytics
-techniques are currently being used in ML. \cite{yuan2021survey}
-conduct a systematic review of 259 papers and propose a taxonomy of
-visual analytics techniques for ML. \cite{hohman2019visual} conduct
-a survey of visual analytics techniques for Deep Learning Models. The
-findings of the study indicate that visual analytics has been widely
-adopted in ML for model explanation, interpretation, debugging and
-improvement.
+Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
+\cite{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\cite{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
+The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
 
-Several tools have been proposed by the visual analytics research
-community to aid practitioners in understanding how their ML models
-operate. \cite{wexler2020if} propose \textit{What-If}, a visual
-analytics tool to explore alternative hypotheses, generate
-counterfactual reasoning, investigate decision boundaries of the model
-and how change in data affects the model predictions. To reduce the
-cognitive load of ML practitioners during model building phase,
-\cite{amershi2015modeltracker} propose \textit{ModelTracker}. Given
-a trained model and a sample dataset, ModelTracker presents all the
-information in traditional summary statistics along with the
-performance of the model.
+Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
+\cite{wexler2020if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
+To reduce the cognitive load of ML practitioners during model building phase, \cite{amershi2015modeltracker} propose \textit{ModelTracker}.
+Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
-\textit{Drava} are a few other tools that have been proposed to handle
-specialised use-cases such as identifying systemmatic errors in ML
-models, generating counterfactual explanations for Generalised
-Additive Models, prioritising machine translation model improvements
-and relating human concepts with semantic dimensions extracted by ML
-models during disentangled representation learning\cite{ahn2023escape,
-wang2023gam, robertson2023angler, wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systemmatic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
 
@@
 
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}
+\begin{lstlisting}[caption={Top-level JSON format of Jupyter notebooks.
+See Listing~\ref{lst:cell-md},~\ref{lst:cell-code} and~\ref{lst:cell-output} for internal structure of various cells.}, label={lst:nbformat}]
 {
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Top-level JSON format of Jupyter notebooks.
-    See Listing~\ref{lst:cell-md},~\ref{lst:cell-code}
-    and~\ref{lst:cell-output} for internal structure of various
-    cells.}\label{lst:nbformat}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}
+\begin{lstlisting}[caption={JSON structure of markdown cells.}, label={lst:cell-md}]
 {
@@
 \end{lstlisting}
-  \captionof{lstlisting}{JSON structure of markdown cells.}
-  \label{lst:cell-md}
-\end{minipage}
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}
+
+\begin{lstlisting}[caption={JSON structure of code cells.}, label={lst:cell-code}]
 {
@@
 \end{lstlisting}
-  \captionof{lstlisting}{JSON structure of code cells.}
-  \label{lst:cell-code}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}
+
+\begin{lstlisting}[caption={JSON structure of code cells with an image output.}, label={lst:cell-output}]
 {
@@
 \end{lstlisting}
-  \captionof{lstlisting}{JSON structure of code cells with an image output.}
-  \label{lst:cell-output}
-\end{minipage}
 
-The data comprised within Jupyter Notebooks is stored in the JSON
-format, and thus can be parsed
-programmatically. Listing~\ref{lst:nbformat} shows the top-level json
-structure of Jupyter Notebooks, adopted from the official nbformat
-website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}. For
-simplicity, we only show the parts that are relevant for this study
-and hide the irrelevant parts using ellipsis.
+The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
+Listing~\ref{lst:nbformat} shows the top-level json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
+For simplicity, we only show the parts that are relevant for this study and hide the irrelevant parts using ellipsis.
 
-Notebooks contain four top level keys namely \texttt{metadata},
-\texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}. The
-\texttt{matadata} key consists of additional information regarding the
-notebook such as the kernel and programming
-language. \texttt{nbformat} and \texttt{nbformat\_minor} contain the
-version of nbformat used in the notebook. The \texttt{cells} key
-consists of a list of all cells within the notebook.
+Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
+The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
+\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
+The \texttt{cells} key consists of a list of all cells within the notebook.
 
-Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and
-~\ref{lst:cell-output} present three examples of cells. Each cell is
-represented as a dictionary with the \texttt{cell\_type},
-\texttt{metadata} and \texttt{source} keys. Cells can be of three
-types: ``markdown'', ``code'' or ``raw''. The contents of each cell
-can be found within the \texttt{source} key, as a list of strings.
-
+Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and ~\ref{lst:cell-output} present three examples of cells.
+Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
+Cells can be of three types: ``markdown'', ``code'' or ``raw''.
+The contents of each cell can be found within the \texttt{source} key, as a list of strings.
 Outputs generated by a code cell are stored as a list within the
@@
 
-We found several instances of visualisation and assertion pairs which
-were realated to one another. However the assertions did not test
-a property of the ML system that was derived from the
-visualisation. Instead, the assertions were used to test the
-correctness of the code that produced the visualisation. Such
-assertions (and the corresponding visualisations) were excluded from
-this study. Table~\ref{tab:post-cond}provides examples of
+We found several instances of visualisation and assertion pairs which were realated to one another.
+However the assertions did not test a property of the ML system that was derived from the visualisation.
+Instead, the assertions were used to test the correctness of the code that produced the visualisation.
+Such assertions (and the corresponding visualisations) were excluded from this study.
+Table~\ref{tab:post-cond}provides examples of
 post-condition checks which were frequently observed in this study.
@@
 \section{Methodology}\label{sec:method}
-Figure~\ref{fig:method} presents an overview of all relevant steps
-used in this study. The empirical analysis comprises of three distinct
-phases namely: \textit{the data collection phase},
-\textit{identification of related visualisation and assertion pairs}
-and finally a \textit{complete analysis of the notebooks} to identify
-the final set of 26 visualisation and assertion pairs.
 
-The yellow boxes in the data collection phase represent filters that
-were applied to the initial corpus of 54K notebooks, to obtain the
-final sample size of 1.7K notebooks used in the empirical study. The
-red boxes represent the exclusion criteria used during phase two to
-remove notebooks note relevant for this study. Finally, the green
-boxes in phase 3 represent the manual analysis conducted to identify
-the final 26 visualisation and assertion pairs.
+Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
+The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.
 
-Boxes appearning next to one-another imply that they were applied
-using an "OR" operator. While boxes appearning below one-another imply
-an "AND" operator. Text in double quotes represent the string pattern
-used in the search query. Each phase of the methodology is explained
-in more detail below.
+The yellow boxes in the data collection phase represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
+The red boxes represent the exclusion criteria used during phase two to remove notebooks note relevant for this study.
+Finally, the green boxes in phase 3 represent the manual analysis conducted to identify the final 26 visualisation and assertion pairs.
+
+Boxes appearning next to one-another imply that they were applied using an "OR" operator.
+While boxes appearning below one-another imply an "AND" operator.
+Text in double quotes represent the string pattern used in the search query.
+Each phase of the methodology is explained in more detail below.
 
@@
 
-We mined public repositories from Github to collect Jupyter notebooks
-that contain the keyword ``assert'' in them. We intentionally keep the
-search criteria on Github broad by only checking for the appropriate
-filetype and the presence of the ``assert'' keyword. This is because,
-the ``assert'' keyword is used in python to define
-tests. Additionally, popular python libraries which provide a testing
-interface (such as nose, unittest and numpy), contain the same keyword
-in their API. Our search criteria captures a large initial sample of
-$54,070$ notebooks with a variety of testing statements. Our approach
-also prevents the need to craft a custom search query based on an
-exhaustive list of keywords that appear in all python testing
-libraries.
+We mined public repositories from Github to collect $54,070$ Jupyter notebooks that contain the keyword ``assert'' in them.
+We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook.
+This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}.
+Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
+The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
 
-We use the Github advanced search syntax~\footnote{REFME} to isolate
-Jupyter Notebook. This translates to the following Github search
-query: \texttt{filename:"Jupyter Notebooks"}. Additionally, we include
-the ``assert'' keyword in the search query with an \texttt{AND}
-operator. The final search query is as follows:
-\texttt{filename:"Jupyter Notebooks" AND "assert"}.
+To further reduce the sample size, we exclude notebooks that do not use popular ML libraries.
+We perform a string search to identify notebooks that import at least one of the ML libraries.
+The string patterns are derived from the python module name used in the import statements.
+The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
 
-The above search query results in approximately 54K notebooks. To
-further reduce the sample size, we exclude notebooks that do not use
-popular ML libraries. We perform a string search to identify notebooks
-that import at least one of the ML libraries. The string patterns are
-derived from the python module name used in the import statements. The
-regex search query is as follows:
-\texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
-
-We programmatically parse the JSON structure of the notebooks and
-represent each notebook as a pandas dataframe for further analysis. We
-map each cell as a row in the dataframe. For each cell, we store the
-\texttt{cell\_type}, the \texttt{source} of the cell and the base64
-encoded PNG string, into the corresponding columns of the
-dataframe. We exclude notebooks that do not contain any visualisations
-by checking if the image column of the dataframe is empty. The
-internal structure of Jupyter notebooks is presented in
-Section~\ref{sec:nbformat}.
+We programmatically parse the JSON structure of the notebooks (more details regarding the internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}) and represent each notebook as a pandas dataframe for further analysis.
+We map each cell as a row in the dataframe.
+For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
+We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.
 
@@
 
-For RQ1, we are interested in finding empirical evidence of
-visualisations and corresponding assertions used in the wild. To
-answer this question, we further reduce our notebook sample size by
-assuming a specific order in which the visualisation and assertion
-code cells appear in the notebooks. Figure~\ref{fig:cell-arrangement}
-presents a visual representation of various arrangements of code cells
-that may occur in notebooks. Note that the visual representation omits
-markdown cells since we only consider the order of code cells in our
-analysis. In the actual notebook however, one or more markdown cells
-may be present between the visualisation and assertion cells.
+To identify visualisation and assertion pairs that are semantically related to one another, we only consider notebooks in which the visualisation and assertion code cells appear in a specific order.
+Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks.
+Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis.
+In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
 
-We conduct the manual analysis presented in
-Section~\ref{sec:identify-related-pairs} for a small random sample of
-notebooks containing each cell arrangement. We observed the
-signal-to-noise ratio for each cell arrangement and picked the
-arrangement with the highest signal-to-noise
-ratio. Table~\ref{tab:cell-arrangement} presents the cell arrangement
-in ascending order of signal-to-noise ratio with a summary of our
-observations.
+We conduct the manual analysis presented in Section~\ref{sec:identify-related-pairs} for a small random sample of notebooks containing each cell arrangement.
+We observed the signal-to-noise ratio for each cell arrangement and picked the arrangement with the highest signal-to-noise ratio.
+Table~\ref{tab:cell-arrangement} presents the cell arrangement in ascending order of signal-to-noise ratio with a summary of our observations.
 
-Assertions defined in a cell above the visualisation cell, were
-typically not related to each other. Often, a markdown cell present
-between the assertion and visualisation code cells, would mark the
-begining of a new section in the notebook. In other cases, the
-assertions were typically checks to ensure that the visualisation code
-would not encounter any errors. This observation also holds for
-Arrangement 2.1, where the assertion is defined above the
-visualisation code, in the same cell. Common examples of such checks
-include checking the shape of features that are used in the
-visualisation. Or checking that the length of the ``x''and ``y``
-features are the same to ensure a continuous line in the plot.
+Assertions defined in a cell above the visualisation cell, were typically not related to each other.
+Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
+In other cases, the assertions were typically checks to ensure that the visualisation code would not encounter any errors.
+This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same cell.
+Common examples of such checks include checking the shape of features that are used in the visualisation.
+Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.
 
-The signal-to-noise ratio was observed to improve in the sample of
-notebooks where the assertions were defined below the visualisation
-code. The visualisation and assertions pairs observed in Arrangment
-2.2 and 3.1 were often related to each other, but were post-condition
-checks. Section~\ref{sec:post-cond} provides more details on
-post-condition checks and our rationale for excluding them from this
-study.
+The signal-to-noise ratio was observed to improve in the sample of notebooks where the assertions were defined below the visualisation code.
+The visualisation and assertions pairs observed in Arrangment 2.2 and 3.1 were often related to each other, but were post-condition checks.
+Section~\ref{sec:post-cond} provides more details on post-condition checks and our rationale for excluding them from this study.
 
-We observed the best results in notebooks where the assertion cell was
-not longer than 10 lines and was defined below the visualisation cell
-(Arrangement 3.2). We observed that most notebook authors followed the
-convention of writing their motivation for the visualisation in
-a markdown cell, before writing the code for the visualisation. The
-visualisations were often followed by another markdown cell to record
-the author's observations. And finally, the observations were
-translated to a analytical test using \texttt{``assert''} or other
-testing methods from external libraries. The restriction on the number
-of lines in the assertion cell was introduced to exclude instances
-where the assertion cell contained a new class definition or several
-helper functions that were not related to the visualisation above.
+We observed the best results in notebooks where the assertion cell was not longer than 10 lines and was defined below the visualisation cell (Arrangement 3.2).
+We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
+The visualisations were often followed by another markdown cell to record the author's observations.
+And finally, the observations were translated to a analytical test using \texttt{assert} or other testing methods from external libraries.
+The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.
 
@@
 
-The initial sample size of 54K notebooks from Github, was reduced to
-the final sample size of 1.7K after applying the filters presented in
-Section~\ref{sec:data-collect}. We analysed all 1.7K notebooks
-manually to further identify 551 notebooks that contained at least one
-pair of visualisation and assertions that were related to each
-other. Table~\ref{tab:exclusion-criteria} presents the exclusion
-criteria used in this phase of the analysis, along with the number of
-notebooks that were excluded based on each criteria.
+The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
+We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
 
-We manually analysed the source code of the visualisation and
-assertion. And excluded visualisation and assertion pairs that were
-not related to each other. We encountered several instances where the
-tracing was not straightforward due to poor coding practices. In such
-cases, we used one-shot prompt engineering~\cite{CITME} with the
-ChatGPT 4.0 model to determine if the visualisation and assertion were
-related to each other.
+We manually analysed the source code of the visualisation and assertion code cells.
+And excluded visualisation and assertion pairs that were not related to each other.
+We encountered several instances where the tracing was not straightforward due to poor coding practices.
+In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
 
-We excluded notebooks that did not train an ML model. We found several
-notebooks authored on topics such as linear algebra, optimisation and
-loss functions. Although these topics are related to ML, the
-visualisation and assertion pairs found in these notebooks could not
-be readily applied to a typically ML development workflow. Analysing
-such notebooks also required a working knowledge of all mathematical
-concepts related to ML. This was deemed beyond the scope of software
-engineering research.
+We excluded notebooks that did not train an ML model.
+We found several notebooks authored on topics such as linear algebra, optimisation and loss functions.
+Although these topics are related to ML, the visualisation and assertion pairs found in these notebooks could not be readily applied to a typically ML development workflow.
+Analysing such notebooks also required a working knowledge of all mathematical concepts related to ML.
+This was deemed beyond the scope of this study.
 
-Finally, we excluded notebooks that were not authored in English or
-contained faulty code that did not produce
-a visualisation. Table~\ref{tab:exclusion-criteria} lists all
-exclusion criteria used along with the number of notebooks that were
-excluded from this study.
+Finally, we excluded notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation.
+Table~\ref{tab:exclusion-criteria} lists all exclusion criteria used along with the number of notebooks that were excluded from this study.
 
@@
 
-We qualitatively assess the comments made for all notebooks during the
-analysis conducted in phase two of the study, as described in
-Section~\ref{sec:identify-related-pairs}.
+We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
 
-We perform an in-depth analysis of 90 short-listed notebooks to
-identify unique ML testing tasks. The in-depth analysis was conducted
-by reading and understanding the entire contents of the notebooks in
-a top-down order. During the analysis, we note several data points
-such as the primary objective of the notebook, the type of ML problem,
-the type of ML model and the type of data. We manually read and
-comprehend each code cell in the notebooks. When present, we also read
-the markdown cells above and below code cells to gain further context
-from the notebook author. For instance, we find that markdown cells
-above a visualisation cell typically motivates the need for the
-visualisation. A markdown cell directly below a visualisation often
-records the observations made by the notebook author from the
-visualisation.
+We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
+The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
+During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data.
+We manually read and comprehend each code cell in the notebooks.
+When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
+For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation.
+A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
 
 \section{Results}\label{sec:result}
+\subsection{RQ1. How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?}\label{sec:result-rq1}
 
-% TODO move the stuff for RQ1 from methodology here!
+Of the 1764 Jupyter notebooks that were analysed in this study, we found 551 notebooks that contained at least one pair of visualisation and assertions which were semantically related to each other.
 
-% NOTE the 5 pairs is to illustrate the 26 total pairs/approach we found for testing ML systems
+Our observations indicate that AI practitioners formulate analytical tests based on the information gained from the visualisations that they create to test ML systems.
+However, this is an emerging testing practice since only 31\% of the notebooks analysed in this study exhibit this testing practice.
 
-% NOTE visual+analytical test is a new "strategy" for testing AI
+\subsection{RQ2. What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?}\label{sec:result-rq2}
 
-This study identified 26 patterns of testing using visualisation and
-assertion. This section highlights five interesting patterns in more
-detail. We plan to release the entire catalogue of patterns publicly
-in the near future.
+This study identified 26 patterns of testing using visualisations and assertions.
+This section highlights five interesting patterns in more detail.
+We plan to release the entire catalogue of patterns publicly.
 
-\subsection{Decision boundary}\label{sec:svm}
-
-% NOTE something about how the visualisation helps with interpretability of how the model is classifying the data?
-
-% NOTE the author is using the visualisation to iteratively check/test how different kernels perform on the data; can also use it for hyper-parameter tuning?
-
-% NOTE: visualisation can provide qualitative assessment of under/overfitting
-
-% NOTE: better assertion: perhaps on the support vectors?
-
-% NOTE assertion created was a single metric (performance): we need tools/guidelines to help practitioners capture what they see in visualisations; community failing to help practitioners what they are visualising
-
-% NOTE assertion is not capturing the whole thing, only assert == we lose a lot of info, we need both
+\subsubsection{Decision boundary}\label{sec:svm}
 
 \begin{figure}
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-linear.png}}
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
   \hfill
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-rbf.png}}
-  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}
-  \label{fig:svm}
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
+  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
 \end{figure}
-\begin{lstlisting}[language=Python]
+
+\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
 assert accuracy_score(y_test, pred) > 0.95
 \end{lstlisting}
-\captionof{lstlisting}{Assertion on the accuracy of the ML model.}\label{lst:svm}
 
@@
 The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel (Figure~\ref{fig:svm-linear}) and later with a RBF kernel (Figure~\ref{fig:svm-rbf}).
-
-% NOTE maybe we need to ellaborate; the linear kernel was above the rbf; based on observations from the linear kernel, the author decides to try the rbf kernel; then after experimenting with hyper-parameters the author puts an assertion that captures the model's performance at that specific point in time.
 
@@
 
-\subsection{Input Image, \texttt{predict\_proba}}
+\subsubsection{Input Image, \texttt{predict\_proba}}
 
 \begin{figure}
-  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35a.png}}
+  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35a.png}}
   \hfill
-  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35b.png}}
-  \caption{Visualisation of random input images for a pre-trained image classifier.}
-  \label{fig:balls}
+  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35b.png}}
+  \caption{Visualisation of random input images for a pre-trained image classifier.}\label{fig:balls}
 \end{figure}
 
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:football}.
+It checks if the probability estimate of the model is close to the specified value.
+The input image is of a football.
+Since the model is trained to detect tennis balls, the specified value is very small.}, label={lst:football}]
 np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion below Figure~\ref{fig:football}.
-It checks if the probability estimate of the model is close to the specified value.
-The input image is of a football.
-Since the model is trained to detect tennis balls, the specified value is very small.}
-  \label{lst:football}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+
+\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:tennisball}.
+Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}, label={lst:tennisball}]
 np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion below Figure~\ref{fig:tennisball}. Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}
-  \label{lst:tennisball}
-\end{minipage}
 
@@
 
-\subsection{Normality}\label{sec:normal}
+\subsubsection{Normality}\label{sec:normal}
 
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
+The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:ktest}]
 for feature in range(data_transformed.shape[1]):
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}
-  \label{lst:kstest}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
+
+\begin{figure}
   \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
-  \captionof{figure}{Visualisation of the distribution of a feature from the dataset.}
-  \label{fig:kstest}
-\end{minipage}
-
-% NOTE assertion and viz are perfect fit;
-
-% NOTE despite the analytical test, practitioners need to visualise this, we need visual confirmation; we cannot get rid of visualisations. we need both!
+  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
+\end{figure}
 
@@
 
-% NOTE I want to say that we don't even think about the fact that we always use a histogram to do this! its second nature, we are taught to do this, without thinking why!
-
 Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
@@
 
-\subsection{Coefficient of Determination}\label{sec:r2}
+\subsubsection{Coefficient of Determination}\label{sec:r2}
 
-% NOTE this vis,assert is model agnostic!
-
-% NOTE assertion captured what they visualised
-
-% NOTE visualisation helps to check for outliers: we are humans, we look at visual patterns, quick and easier to observe fro visualisations
-
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}, label={lst:r2}]
 r2_gru = r2_score(y_test, y_pred)
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}
-  \label{lst:r2}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
+
+\begin{figure}
   \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
-  \captionof{figure}{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}
-  \label{fig:r2}
-\end{minipage}
+  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
+\end{figure}
 
@@
 
-% TODO more context on r2_score, it fits a linear regression model and checks the blah blah blah
 Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
@@
 
-\subsection{Normalization}
-
-% TODO check the notebook again for implementation of normalization
+\subsubsection{Normalization}
 
 \begin{figure}
-  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65a.png}}
-  \hfill
-  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65b.png}}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=\linewidth]{../catalogue/select-65a.png}}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=\linewidth]{../catalogue/select-65b.png}}
   \caption{Visualisation of data before and after various data pre-processing techniques.}
@@
 
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
 assert scaled_data[scaled_data.argmax()] <= 1
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}
-  \label{lst:scale}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+
+\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
 assert normalized_data[normalized_data.argmax()] < 10
@@
 \end{lstlisting}
-  \captionof{lstlisting}{Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}
-  \label{lst:normal}
-\end{minipage}
 
@@
 
-\section{Discussion}\label{sec:discuss}
+\section{Implications}\label{sec:discuss}
 
-% NOTE comment on the over-whelming presence of magic thresholds in ML
-% testing! where are authors getting these numbers from? link this to
-% requirements engineering?
+This study mined 54K Jupyter Notebooks that contained an assertion.
+Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation.
+Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
 
-% NOTE comment on the anti-patterns in testing that we observe in the 5 patterns presented
+The results from RQ2 provide emprical evidence to support the use of visualisations to test various properties of ML systems.
+Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues.
+Visualisations used to test ML system properties however, demand manual verification effort.
+Information gained from visualisations and subsequent decisions made based on this information, must be manually verified in every iteration of the ML development cycle.
 
-% NOTE link the qualitative and quantitative testing to the ML lifecycle: how some tests will let practitioner know when something changes
+% NOTE link to accumulation of technical debt due to highly experimental nature, and constant change in the system
 
-% NOTE comment on evolution? we have to look at the entire lifecycle of a visualisation: notebooks are run iteratively?
+Assertions or analytical tests extracted from ML visualisations can reduce this manual verification effort.
+Assertions encapsulate the knowledge gained from the visualisation with regards to the property of the ML system which is currently under test.
+Formal assertions document the observations made by the AI practitioner regarding the model or the data, at a specific point in time.
+These observations should not be violated by the future model development cycles unless the requirements of the ML system are changed.
+Assertions scale across organizational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
 
-\subsection{New form of testing}
-% NOTE some form of commentary on the fact that is form of qualitative and quantitative testing is happening but not that popular yet!
-
-% NOTE so we need more tools to aid practitioners to encourage this form of testing and also help them translate qualitative properties into quantitative tests!
-
-\subsection{From Implicit Expectations to Explicit Assertions}
-% NOTE regeneration of the model does not change this observation at this point in time
-
-% NOTE discuss the pros and cons of having a qualitative assessment: they are dense with useful information; humans need visual patterns!
-
-% NOTE then show that we also need qualitative assessment that can capture our expectations; this is important so that the model never violates that we understand from the visualisations
-
-% NOTE now is the time to talk about how assertions act as documentation of the domain expertise and knowledge extracted from the visualisation from the original creator; this also scales as the system complexity and composition of the team/organisation evolves!
-
-% NOTE what are the benefits using visualisations to assess the
-% quality of ML systems? They are packed with a lot of
-% information. And humans are cognitively tuned to identify patterns
-% in visualisations.
-
-Visualisations condense valuable information that is easy to understand for humans.
-
-\subsection{Something about magic thresholds}\label{sec:magic-threshold}
-% NOTE we observe that most analytical test translates to comparing a value against a threshold; this indicates the open challenges in ML testing; either we don't yet know how to translate analytical tests from visualisations or we cannot do this in the first place and we must use a hybrid approach?
-
-% NOTE this is a better place to talk about the magic threshold phenomenon; link this to the magic number smell in software engineering, but in ML it is very prevalent and an alternative is not always clear
-
-% NOTE and then motivate how we need more tools/techniques to aid practitioners to extract/translate all the info from visualisations into qualitative assertions!
-
-\subsection{Humans need visual confirmation}\label{sec:visual}
-% NOTE even though analytical tests are sufficient; humans still prefer visual confirmation
-
-% NOTE when working with image dataset, we observe in our catalogue that practitioners prefers to also visually check the input image
-
-% NOTE more detailed discussion on how a lot of information can be condensed into a single visualition and how humans are good at observing visual patterns
+Results for RQ1 indicate that combining ML visualisations with analytical assertions is an emerging testing practice within the ML community.
+However in results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+Our observations indicate a need to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
+We find the need for automated tools that make this translation process easier for AI practitioners.
 
@@
 
-% motivation for keeping the search on github broad:
-% - not only does it pick up python assert statements, but also flags
-% - notebooks that use other testing libraries or modules that provide a
-% - testing interface. This is a good compromise between doing an
-% - exhaustive search of all testing method names in the python
-% - ecosystem and then doing a search for each of those patterns
+% TODO defend the use of a simple search query on github
+Our approach also prevents the need to craft a custom search query based on an exhaustive list of keywords that appear in all python testing libraries.
+We intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
+This is because, the ``assert'' keyword is used in python to define tests.
+Additionally, popular python libraries which provide a testing interface (such as nose, unittest and numpy), contain the same keyword in their API.
 
 % TODO don't mention quantity of random sample; then we need to explain/defend this choice
-We used an iterative approach to determine the effectiveness of each
-filter that was used. We applied each filter in the same order as
-presented in Table~\ref{tab:filters}, starting with F1. We randomly
-sampled 50 notebooks from the population of notebooks obtained after
-applying the filter. We then conduct the qualitative assessment
-presented in Section~\ref{REFME} and observe the number of related
-visualisation and assertion pairs. In each new iteration, we
-progressively add the other filters and include the filter if the
-number of relevant visualisation and assertion pairs increases.
+We used an iterative approach to determine the effectiveness of each filter that was used.
+We applied each filter in the same order as presented in Table~\ref{tab:filters}, starting with F1.
+We randomly sampled 50 notebooks from the population of notebooks obtained after applying the filter.
+We then conduct the qualitative assessment presented in Section~\ref{REFME} and observe the number of related visualisation and assertion pairs.
+In each new iteration, we progressively add the other filters and include the filter if the number of relevant visualisation and assertion pairs increases.
+
+% TODO defend use of one-shot prompt engineering for identifying related visualisation and assertion code cells; how did we mitigate this threat?
+
 \section{Conclusion}\label{sec:conclude}

### 2023-10-12 report/icst24.tex
feat(report): 2023-10-12 20:52 updates

@@
 
-\section{Threats to Validity}\label{sec:threats}
-% defend use of github as source for notebooks:
-% - why did kaggle not work (refer to notes)
-% - why did the existing replication packages not work?
+\section{Threats to Validity and Future Work}\label{sec:threats}
+The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
+To the best of our knowledge, \cite{pimentel2019large} and \cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
+We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github.
+We invite the scientific community to extend our catalogue of visual and analytical testing patterns, using the corpus of notebooks provided by the above authors.
 
-% TODO defend the use of a simple search query on github
-Our approach also prevents the need to craft a custom search query based on an exhaustive list of keywords that appear in all python testing libraries.
-We intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-This is because, the ``assert'' keyword is used in python to define tests.
-Additionally, popular python libraries which provide a testing interface (such as nose, unittest and numpy), contain the same keyword in their API.
+In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose, unittest and numpy).
+Our approach provides captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in all Python testing libraries.
 

### 2023-10-17 report/icst24.tex
feat(report): 2023-10-17 16:52 updates

@@
 
-Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
-With the advent of Large Language Models, the power of AI to directly influence the daily lives of humans is growing at a radical pace.
-As reports of ``hallucinated'' facts, racist predicts and autonomous car crashes steadily increase, ensuring that ML systems function as they were intended to, takes precedence.
-
 Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
@@
 
-In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction.
+In contrast to prior work, we approach ML testing from a novel direction.
 We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
@@
 The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
-The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity (Section~\ref{sec:threats}) and future work (Section~\ref{sec:conclude}).
+The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity and future work (Section~\ref{sec:threats}).
 
@@
 
+% TODO include shrestha2023detangler
 % TODO how is our work different from prior work?
 
-This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
+This section summarises prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
 The section concludes with an overview of technical knowledge required for this paper.
@@
 Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase\cite{CITEME}.
+While traditional software systems mature through change in their codebase, ML enabled systems additionally experience change in the training data and the machine learning model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
 With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
@@
 First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy.
+And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
 
@@
 
-% TODO summarise literature on fairness testing
-%
-Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.
+In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems in practice.
+Specifically, we focus on how ML practitioners use visualisations to test specific properties of ML systems.
+We further investigate how frequently ML practitioners formulate analytical tests based on the information gained from the visualisations.
 
@@
 Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large}.
+Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.
 
-\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, and source code and metadata from over 900 releases of 12 important data science libraries.
-Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
-\cite{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices in computational notebooks.
+\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, source code and metadata from over 900 releases of 12 important data science libraries.
+Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends in technologies they should focus on.
+\cite{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
 Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
@@
 
+% TODO still needs improvement
+This study uses Jupyter Notebooks mined from Github.
+Jupyter notebooks have been widely adopted by the Data Science and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.
+Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
+We leverage the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
+
 \subsection{Visualisations and Machine Learning}\label{sec:visualisations}
@@
 ML models and the underlying data is complex and multi-dimensional.
-To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models.
+To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.
 
@@
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systemmatic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+
+In contrast to prior work which propose dedicated visual analytics tools, this study focuses on visualisations within computational notebooks.
+We focus on visualisations produced by AI practitioners using Python libraries (such as Matplotlib or Seaborn).
+Additional context provided by surrounding code and markdown cells are used to understand the motivation for creating the visualisation and the information gained from it.
 
@@
 The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
-Listing~\ref{lst:nbformat} shows the top-level json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
-For simplicity, we only show the parts that are relevant for this study and hide the irrelevant parts using ellipsis.
+Listing~\ref{lst:nbformat} shows the top-level JSON structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
+For simplicity, only parts that are relevant for this study are shown and irrelevant parts are hidden using ellipsis.
 
 Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
-The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
-\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
+The \texttt{metadata}, \texttt{nbformat} and \texttt{nbformat\_minor} keys contain additional information regarding the notebook such as information regarding the kernel and the programming language used in the notebook.
 The \texttt{cells} key consists of a list of all cells within the notebook.
 
-Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and ~\ref{lst:cell-output} present three examples of cells.
+Listings~\ref{lst:cell-md},~\ref{lst:cell-code} and~\ref{lst:cell-output} present three examples of cells.
 Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
@@
 The contents of each cell can be found within the \texttt{source} key, as a list of strings.
-Outputs generated by a code cell are stored as a list within the
-\texttt{outputs} key as seen in Listing~\ref{lst:cell-output}. Static
-visualisations are stored as base64 encoded png images under the
-\texttt{"image/png"} key, which is nested within the \texttt{data} key
-inside \texttt{outputs}.
+Outputs generated by a code cell are stored as a list within the \texttt{outputs} key as seen in Listing~\ref{lst:cell-output}.
+Static visualisations are stored as base64 encoded png images under the \texttt{outputs.data.image/png} key.
 
@@
 \caption{Frequent examples of post-condition checks observed in this study.}
-  \begin{tabular}{m{0.3\textwidth} m{0.4\textwidth} m{0.3\textwidth}}
+  \begin{tabular}{m{0.2\textwidth} m{0.4\textwidth} m{0.3\textwidth}}
 \hline
@@
 
-We found several instances of visualisation and assertion pairs which were realated to one another.
+We found several instances of visualisation and assertion pairs which were related to one another.
 However the assertions did not test a property of the ML system that was derived from the visualisation.
@@
 Such assertions (and the corresponding visualisations) were excluded from this study.
-Table~\ref{tab:post-cond}provides examples of
-post-condition checks which were frequently observed in this study.
+Table~\ref{tab:post-cond} provides examples of such post-condition checks that were frequently observed in this study.
 
@@
 Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
-The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.
+The empirical analysis comprises of three distinct phases namely: \textit{data collection}, \textit{identifying related visualisation and assertion pairs} and finally a \textit{identifying qualitative and quantitative testing patterns} to identify the final set of 26 unique testing patterns using visualisations and assertions.
 
@@
 
-Boxes appearning next to one-another imply that they were applied using an "OR" operator.
-While boxes appearning below one-another imply an "AND" operator.
+Boxes appearing next to one-another imply that they were applied using an ``OR'' operator.
+While boxes appearing below one-another imply an ``AND'' operator.
 Text in double quotes represent the string pattern used in the search query.
@@
 We mined public repositories from Github to collect $54,070$ Jupyter notebooks that contain the keyword ``assert'' in them.
-We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook.
-This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}.
+We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook.
 Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
@@
 We programmatically parse the JSON structure of the notebooks (more details regarding the internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}) and represent each notebook as a pandas dataframe for further analysis.
-We map each cell as a row in the dataframe.
+Each cell is mapped as a row in the dataframe.
 For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
-We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.
+Notebooks without any visualisations are excluded by checking if the image column of  the dataframe is empty.
 
@@
 Assertions defined in a cell above the visualisation cell, were typically not related to each other.
-Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
+A markdown cell present between the assertion and visualisation code cells, would often mark the beginning of a new section in the notebook.
 In other cases, the assertions were typically checks to ensure that the visualisation code would not encounter any errors.
-This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same cell.
+This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same code cell.
 Common examples of such checks include checking the shape of features that are used in the visualisation.
-Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.
+Or checking that the length of the features used as the ``x''and ``y`` axes in the figure are the same to ensure a continuous line in the plot.
 
-The signal-to-noise ratio was observed to improve in the sample of notebooks where the assertions were defined below the visualisation code.
+The signal-to-noise ratio improved in the sample of notebooks where the assertions were defined below the visualisation code.
 The visualisation and assertions pairs observed in Arrangment 2.2 and 3.1 were often related to each other, but were post-condition checks.
@@
 
-We observed the best results in notebooks where the assertion cell was not longer than 10 lines and was defined below the visualisation cell (Arrangement 3.2).
+We observed the best results in notebooks with Arrangement 3.2---where the assertion cell was not longer than 10 lines and was defined below the visualisation cell.
 We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
 The visualisations were often followed by another markdown cell to record the author's observations.
-And finally, the observations were translated to a analytical test using \texttt{assert} or other testing methods from external libraries.
+And finally, the observations were translated to a analytical test using an \texttt{assert} statement or other testing methods from external libraries.
 The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.
@@
 The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
-We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
+We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were semantically related to each other.
 
@@
 And excluded visualisation and assertion pairs that were not related to each other.
-We encountered several instances where the tracing was not straightforward due to poor coding practices.
+We encountered several instances where the tracing was not straightforward due to poor coding practices---such as use of obscure variable names or deeply nested function calls.
 In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
 
-We excluded notebooks that did not train an ML model.
+% TODO perhaps this requires more clarification? What do we mean that they could not be "readily" applied?
+Notebooks that did not train an ML model were excluded from this study.
 We found several notebooks authored on topics such as linear algebra, optimisation and loss functions.
@@
 
-Finally, we excluded notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation.
-Table~\ref{tab:exclusion-criteria} lists all exclusion criteria used along with the number of notebooks that were excluded from this study.
+Finally, notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation were excluded from this study.
+Table~\ref{tab:exclusion-criteria} lists all exclusion criteria along with the number of notebooks that were excluded from this study.
 
@@
 
-We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
+% TODO I think its not clear that the qualitative assessment of notes was done to identify the 90 notebooks?
+We qualitatively assessed our notes and observations made during Phase 2 (as described in Section~\ref{sec:identify-related-pairs}).
+And perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
+The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
+During the analysis, several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data were captured.
 
-We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
-The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
-During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data.
-We manually read and comprehend each code cell in the notebooks.
+% TODO repeat of what we say for Figure 2.
+We read each code cell in the notebooks.
 When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
@@
 A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
+
+% TODO transition between methodology and results is not strong.
+Specifically, Phase 3 of the analysis is not substantial.
+The distinction between the vis-assert pairs and the ML testing tasks is not clear.
+We need to describe that the vis-assert pairs are used to accomplish the ML testing tasks.
 
@@
 Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation.
-Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
+Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.
+A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
 
@@
 \section{Threats to Validity and Future Work}\label{sec:threats}
+% TODO discuss that we only consider viz-assert pairs that could be readily applied to ML development lifecycle.
+However, we also found several visualisation-assert pairs that target the more fundamental concepts of ML.
+Further work is required to explore such pairs and check if they are useful...
+%
 The empirical analysis presented in this study uses Jupyter notebooks mined from Github.

### 2023-10-17 bin/filter.py

@@ -1,14 +1,10 @@
 """Check for assert statements in juptyer notebooks.
 
-Without any flags, the script will check for assert statements only in code cells that produce
-a visualisation. The search is not greedy by default. It will print the notebook to stdin if at
-least 1 assert statement exists in the notebook.
+Without any flags, the script will check for assert statements only in code cells that produce a visualisation. The search is not greedy by default. It will print the notebook to stdin if at least 1 assert statement exists in the notebook.
 
-With the `--proximity` flag, the script will also check in code cells directly below the
-visualisation cells.
+With the `--proximity` flag, the script will also check in code cells directly below the visualisation cells.
 
-With the `--strict` flag, the script will only flag notebooks if the assert statements is in the
-code cell below the visualisation cell. This flag takes precedence over `--proximity`.
+With the `--strict` flag, the script will only flag notebooks if the assert statements is in the code cell below the visualisation cell. This flag takes precedence over `--proximity`.
 """
 
 import ast

### 2023-10-24 report/icst24.tex
feat(report): 2023-10-24 updates

@@
 email address or ORCID}
+\and
+\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
+\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+\textit{name of organization (of Aff.)}\\
+City, Country \\
+email address or ORCID}
 }
@@
 We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another.
-We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns.
-We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.
-
-% TODO what were the primary findings of this study?
+We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
+We further investigate of the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns.
+We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part.
+Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems.
+And help ML practitioners formulate better analytical assertions.
 
@@
 
+% TODO make sure that we use the terminology of "Testing Tasks" everywhere for RQ2
 Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
@@
 
-In contrast to prior work, we approach ML testing from a novel direction.
+In contrast to prior work, we approach ML testing from a new perspective.
 We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
@@
 \begin{description}
-\item[RQ1.] How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?
+\item[RQ1.] How frequently are analytical tests formulated from visualisations created to test ML systems?
 
@@
 We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
+We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion (VA) pairs that are semantically related to each other.
+We plan to release this catalogue of VA pairs publicly.
 
-\item[RQ2.] What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?
+\item[RQ2.] What patterns are frequently observed in VA pairs used to test ML systems?
 
-From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks.
-We identify 26 unique visualisation and assertion pairs.
-We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.
+We analyse the 269 VA pairs from RQ1, and observe three testing patterns that frequently occur in the VA pairs.
+
+\item[RQ3.] Do the assertions capture all the information obtained from the corresponding visualisation?
+
+We further perform an in-depth analyse of 34 VA pairs (and their corresponding notebooks). We observe that assertions often cannot capture all the information obtained from the visualisation.
 
@@
 
-% TODO what are the primary findings of this paper...
-
-The paper is organised as follows.
-Section~\ref{sec:prelim} summarises the related literature and technical information relevant to this study.
-The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
-The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity and future work (Section~\ref{sec:threats}).
+Our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique.
+However, the assertions found in this study often do not reflect the information obtained from their visual counter-part.
+This study finds the need for automated tools that can reduce the manual effort necessary to validate visualisations in ML systems.
+And help ML practitioners to formulate better analytical assertions from visualisations.
 
 \section{Preliminaries}\label{sec:prelim}
-
-% TODO include shrestha2023detangler
-% TODO how is our work different from prior work?
 
@@
 
-\subsection{Computational Notebooks and Software
-  Engineering}\label{sec:notebooks}
+\subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
 
@@
 
-Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
-The empirical analysis comprises of three distinct phases namely: \textit{data collection}, \textit{identifying related visualisation and assertion pairs} and finally a \textit{identifying qualitative and quantitative testing patterns} to identify the final set of 26 unique testing patterns using visualisations and assertions.
-
-The yellow boxes in the data collection phase represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
-The red boxes represent the exclusion criteria used during phase two to remove notebooks note relevant for this study.
-Finally, the green boxes in phase 3 represent the manual analysis conducted to identify the final 26 visualisation and assertion pairs.
-
-Boxes appearing next to one-another imply that they were applied using an ``OR'' operator.
-While boxes appearing below one-another imply an ``AND'' operator.
+Figure~\ref{fig:method} summarises the methodology used to collect related VA pairs from Jupyter notebooks.
+The yellow boxes represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
+The red boxes represent the exclusion criteria used to exclude VA pairs not relevant for this study.
 Text in double quotes represent the string pattern used in the search query.
@@
 
-To further reduce the sample size, we exclude notebooks that do not use popular ML libraries.
-We perform a string search to identify notebooks that import at least one of the ML libraries.
+To further reduce the sample size, we exclude notebooks that do not use popular ML or Data Science (DS) libraries.
+We perform a string search to identify notebooks that import at least one of the libraries.
 The string patterns are derived from the python module name used in the import statements.
@@
 
-\begin{figure*}
+\begin{figure}
   \centering
-  \includegraphics[width=\textwidth]{nb-structure.pdf}
+  \includegraphics[width=\linewidth]{nb-structure.pdf}
   \caption{Arrangement of code cells containing assertion and
-    visualisation.}
-  \label{fig:cell-arrangement}
-\end{figure*}
+    visualisation.}\label{fig:cell-arrangement}
+\end{figure}
 
@@
 We conduct the manual analysis presented in Section~\ref{sec:identify-related-pairs} for a small random sample of notebooks containing each cell arrangement.
-We observed the signal-to-noise ratio for each cell arrangement and picked the arrangement with the highest signal-to-noise ratio.
+We observed the signal-to-noise ratio---notebooks with related VA pairs versus not---for each cell arrangement and picked the arrangement with the best results.
 Table~\ref{tab:cell-arrangement} presents the cell arrangement in ascending order of signal-to-noise ratio with a summary of our observations.
@@
 The signal-to-noise ratio improved in the sample of notebooks where the assertions were defined below the visualisation code.
-The visualisation and assertions pairs observed in Arrangment 2.2 and 3.1 were often related to each other, but were post-condition checks.
+The visualisation and assertions pairs observed in Arrangement 2.2 and 3.1 were often related to each other, but were post-condition checks.
 Section~\ref{sec:post-cond} provides more details on post-condition checks and our rationale for excluding them from this study.
@@
 
-\subsection{Identifying Related Visualisation and Assertion Pairs}\label{sec:identify-related-pairs}
+\subsection{Identifying Related Visualisation-Assertion Pairs}\label{sec:identify-related-pairs}
 
 The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
-We analysed all 1.7K notebooks manually to further identify 551 notebooks that contained at least one pair of visualisation and assertions that were semantically related to each other.
+We analysed all 1.7K notebooks manually to further identify 269 VA pairs that were semantically related to each other.
 
 We manually analysed the source code of the visualisation and assertion code cells.
-And excluded visualisation and assertion pairs that were not related to each other.
-We encountered several instances where the tracing was not straightforward due to poor coding practices---such as use of obscure variable names or deeply nested function calls.
-In such cases, we used one-shot prompt engineering~\cite{CITME} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
+And excluded VA pairs that were not related to each other.
+We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls.
+In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
 
-% TODO perhaps this requires more clarification? What do we mean that they could not be "readily" applied?
-Notebooks that did not train an ML model were excluded from this study.
-We found several notebooks authored on topics such as linear algebra, optimisation and loss functions.
-Although these topics are related to ML, the visualisation and assertion pairs found in these notebooks could not be readily applied to a typically ML development workflow.
-Analysing such notebooks also required a working knowledge of all mathematical concepts related to ML.
-This was deemed beyond the scope of this study.
+Notebooks not related to ML or Data Science (DS) were excluded from this study.
+We found several notebooks from other domains such as cryptography, scientific simulations, genetic algorithms and fluid dynamics.
+Such notebooks were captured in Phase 1 because they imported one of the popular python libraries (typically numpy or pandas).
+However, the visualisation and assertion pairs discovered within these notebooks were not related to ML or DS.
 
-Finally, notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation were excluded from this study.
+As explained in Section~\ref{sec:post-cond}, this study excluded assertions which did not test a property of an ML system but rather the correctness of the visualisation.
+Such assertions or post-condition checks were excluded from this study.
+
+We also excluded duplicate notebooks from the study. For the purposes of this study, a notebook is labelled as a duplicate of another, when all the following conditions hold.
+
+\begin{enumerate}
+  \item Both notebooks are authored by the same author and found in the same Github repository
+  \item The visualisation-assertion pairs found in the notebooks has previously been recorded in this study
+  \item And the contents of the notebook is identical to a prior notebook analysed in this study
+\end{enumerate}
+
+Finally, notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation were excluded.
 Table~\ref{tab:exclusion-criteria} lists all exclusion criteria along with the number of notebooks that were excluded from this study.
@@
   \centering
-  \caption{Exclusion criteria used to identify related visualisation and assertion pairs.}
+  \caption{Inclusion-Exclusion criteria used to identify related visualisation-assertion pairs.}\label{tab:exclusion-criteria}
   \begin{tabular}{l p{0.2\textwidth} r}
@@
     The visualisation and assertion are not related &
-    440\\
+    560\\
     \textbf{EC2} &
-    Notebook does not contain a trained ML model &
-    383\\
+    Notebook is not ML or DS related &
+    419\\
     \textbf{EC3} &
     The assertion is a post-condition check &
-    200\\
+    201\\
     \textbf{EC4} &
-    Notebook not written in English &
-    159\\
+    Notebook not entirely written in English &
+    164\\
     \textbf{EC5} &
+    Notebook is a duplicate &
+    129\\
+    \textbf{EC6} &
     Notebook contains incorrect code &
-    31\\
+    34\\
     \hline
     \multicolumn{2}{c}{\emph{\textbf{Total number of NBs excluded:}}} &
-    1.213\\
+    1507\\
     \hline
   \end{tabular}
-  \label{tab:exclusion-criteria}
 \end{table}
 
-\subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}
+\section{Results}\label{sec:result}
+\subsection{RQ1: How frequently are analytical tests formulated from visualisations created to test ML systems?}\label{sec:result-rq1}
 
-% TODO I think its not clear that the qualitative assessment of notes was done to identify the 90 notebooks?
-We qualitatively assessed our notes and observations made during Phase 2 (as described in Section~\ref{sec:identify-related-pairs}).
-And perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
-The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
-During the analysis, several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data were captured.
-
-% TODO repeat of what we say for Figure 2.
-We read each code cell in the notebooks.
-When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
-For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation.
-A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
-
-% TODO transition between methodology and results is not strong.
-Specifically, Phase 3 of the analysis is not substantial.
-The distinction between the vis-assert pairs and the ML testing tasks is not clear.
-We need to describe that the vis-assert pairs are used to accomplish the ML testing tasks.
-
-\section{Results}\label{sec:result}
-\subsection{RQ1. How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?}\label{sec:result-rq1}
-
-Of the 1764 Jupyter notebooks that were analysed in this study, we found 551 notebooks that contained at least one pair of visualisation and assertions which were semantically related to each other.
+From the 1764 Jupyter notebooks that were analysed in this study, we found 269 VA pairs that were semantically related to each other.
 
 Our observations indicate that AI practitioners formulate analytical tests based on the information gained from the visualisations that they create to test ML systems.
-However, this is an emerging testing practice since only 31\% of the notebooks analysed in this study exhibit this testing practice.
+However, this is an emerging testing practice since only 15\% of the notebooks analysed in this study exhibit this testing practice.
 
-\subsection{RQ2. What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?}\label{sec:result-rq2}
+\subsection{RQ2: What patterns are frequently observed in VA pairs used to test ML systems?}\label{sec:result-rq2}
 
-This study identified 26 patterns of testing using visualisations and assertions.
-This section highlights five interesting patterns in more detail.
-We plan to release the entire catalogue of patterns publicly.
+We further analysed the 269 VA pairs obtained from RQ1, and identified three testing patterns which are presented in Table~\ref{tab:testing-patterns} and discussed further below.
 
-\subsubsection{Decision boundary}\label{sec:svm}
+\begin{table}
+  \centering
+  \caption{Testing patterns observed in the visualisation-assertion pairs.}
+  \begin{tabular}{l p{0.3\textwidth} c c}
+    \hline
+    \textbf{ID} &
+    \textbf{Description} &
+    \multicolumn{2}{c}{\textbf{Count}}\\
+    & & Yes & No\\
+    \hline
+    \textbf{P1} &
+    Does the assertion use a comparison operator (such as \emph{eq, lt, gt})? &
+    202 & 67\\
+    \textbf{P2} &
+    Does the assertion use magic thresholds? &
+    101 & 168\\
+    \textbf{P3} &
+    Does the assertion use an external testing library (such as \texttt{numpy.testing}? &
+    42 & 227\\
+    \hline
+  \end{tabular}
+  \label{tab:testing-patterns}
+\end{table}
+
+\begin{table}
+  \centering
+  \caption{Most frequently occurring assertions that use a comparison operator.}
+  \begin{tabular}{m{0.9\linewidth}}
+    \hline
+    \begin{lstlisting}[language=Python]
+assert accuracy_score(y_test, pred) > 0.95
+    \end{lstlisting}\\
+    \begin{lstlisting}[language=Python]
+assert HISTORY['test_accuracy'][-1] > 0.92
+    \end{lstlisting}\\
+    \begin{lstlisting}[language=Python]
+assert np.mean(mean_rw_history[-10:]) > 10.
+print("That's good enough for tutorial.")
+    \end{lstlisting}\\
+    \begin{lstlisting}[language=Python]
+assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."
+    \end{lstlisting}\\
+    \begin{lstlisting}
+assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
+    \end{lstlisting}\\
+    \hline
+  \end{tabular}
+  \label{tab:compare-op-asserts}
+\end{table}
+
+202 assertions from the 269 VA pairs (75\%) contain a comparison operator such as \emph{equal to}, \emph{lesser than} or \emph{greater than}.
+Table~\ref{tab:compare-op-asserts} presents the most frequently observed assertions that conform to this pattern.
+An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
+
+46\% of the assertions with a comparison operator, also contain a ``magic threshold'' defined by the author.
+It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
+The threshold could have been derived based on the domain expertise of the ML practitioner.
+Another alternative could be to define them during the Requirements Engineering phase, though discussions with all involved stakeholders\cite{CITEME}.
 
 \begin{figure}
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
-  \hfill
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
-  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
+  \includegraphics[width=\linewidth]{../catalogue/select-16.png}
+  \caption{Visualisation of the training loss of an ML model.}\label{fig:loss}
 \end{figure}
 
-\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
-assert accuracy_score(y_test, pred) > 0.95
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
+In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]
+assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."
 \end{lstlisting}
 
-Figures~\ref{fig:svm} presents visualisations obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
-The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}.
-The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel (Figure~\ref{fig:svm-linear}) and later with a RBF kernel (Figure~\ref{fig:svm-rbf}).
-
-The visualisations presented in Figure~\ref{fig:svm} contain several useful information that can be used to test and debug ML models.
-Humans are tuned to recognise visual patterns~\cite{CITEME}.
-Visualisations in Figure~\ref{fig:svm}, allows ML practitioners to quickly identify mis-classified examples.
-Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
-For instance, in Figure~\ref{fig:svm-linear} we observe that the model is underfitting since a linear kernel too simply for the underlying data.
-The visualisation also allows ML practitioners to compare different ML models or variants of the same ML model to iteratively find a model that best fits the underlying training data.
-In this case for instance, the author of the notebook First trained the SVM with a linear kernel.
-And upon realising that the data is not linearly separable, chose the RBF kernel.
-
-Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
-The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
-The assertion is derived from the observations made from the visualisation in Figure~\ref{fig:svm-rbf}.
-The visulisation shows that the model is able to classify the data well which is why the notebook author sets the threshold for the accuracy in the assertion to a high value.
-
-The assertion however, is only able to capture the information pertaining to the performance of the model.
-While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
-
-\subsubsection{Input Image, \texttt{predict\_proba}}
-
-\begin{figure}
-  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35a.png}}
-  \hfill
-  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35b.png}}
-  \caption{Visualisation of random input images for a pre-trained image classifier.}\label{fig:balls}
-\end{figure}
-
-\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:football}.
-It checks if the probability estimate of the model is close to the specified value.
-The input image is of a football.
-Since the model is trained to detect tennis balls, the specified value is very small.}, label={lst:football}]
-np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
-\end{lstlisting}
-
-\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:tennisball}.
-Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}, label={lst:tennisball}]
-np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
-\end{lstlisting}
-
-Figure~\ref{fig:balls} presents visualisations obtained from a notebook that uses a pre-trained image classifier from the GluonCV library~\footnote{https://cv.gluon.ai}, to identify images that contain tennis balls.
-The visualisations are of two random input images, one from each class in the dataset.
-Listing~\ref{lst:football} presents the assertion related to Figure~\ref{fig:football}.
-The assertion checks if the probability estimate of the model for the given input image, is close to the specified value with a precision of up to three decimal places.
-Since the model is trained to detect tennis ball, and the input image contains a footbal, the specified probability estimate is very low.
-In contrast, Listing~\ref{lst:tennisball} specifies a much higher probability estimate since the input image contains tennis balls.
-
-The above visualisation and assertion pair can be used to spot-check a few edge-cases or adversarial examples which the model may find difficult to classify.
-It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
-The threshold may have been derived based on the domain expertise of the ML practitioner.
-Another alternative could be to define them during the Requirements Engineering phase, by discussing with all involved stakeholders\cite{CITEME}.
-
-In contrast to the visualisation and assertion pair in Section~\ref{sec:svm}, the assertion here is able to capture the author's motivation for visualising the input image along with the information obtained from it.
-Regardless, the notebook author still visualises the input image.
-This is a recurring phenomenon that we observe in other visualisation and assertion pairs in our catalogue.
-We discuss this phenomenon further in Section~\ref{sec:visual}.
-
-\subsubsection{Normality}\label{sec:normal}
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:ktest}]
-for feature in range(data_transformed.shape[1]):
-assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
-\end{lstlisting}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
-  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
-\end{figure}
-
-The visualisation in Figure~\ref{fig:kstest} is obtained from a notebook that trains a Generative Adversarial Network (GAN) to generate data which simulates a neutrino particle detector.
-The notebook author transforms the training data by normalising it using the \texttt{QuantileTransform} from scikit-learn~\footnote{REFME}.
-After performing the transformation, the author performs a visual check by plotting the distribution of a feature from the dataset as shown in Figure~\ref{fig:kstest}.
-The visualisation shows a bell-shaped curve indicating that the data is normal and that the transformation worked.
-The author then defines the assertion presented in Listing~\ref{lst:kstest} which reflects their observation made from the visualisation.
-
-The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
-The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
-
-Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
-Several ML models assume normality in the underlying training data.
-The verification is often performed visually using a histogram and checking that the distribution shows a "bell-shaped" curve\cite{CITME}.
-However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
-The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
-And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
-
-\subsubsection{Coefficient of Determination}\label{sec:r2}
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}, label={lst:r2}]
-r2_gru = r2_score(y_test, y_pred)
-assert r2_gru > 0.6
-\end{lstlisting}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
-  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
-\end{figure}
-
-Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
-The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
-The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
-The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model was able to learn certain patterns in the training data.
-The spread of $y_{pred}$ at any given $y_{true}$ is wide.
-For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
-This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.
-
-Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
-The author uses \texttt{r2\_score} metric from the scikit-learn library to quantify the performance of the model.
-The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computed the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
-The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
-Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
-
-Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess is the model is under or overfitting the training data.
-If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
-In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
-The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
-The assertion in Listing~\ref{lst:r2} however, does not capture such information.
-It can only check if the performance of the model is better than a specific threshold.
-
-\subsubsection{Normalization}
+In contrast to the assertions presented in Table~\ref{tab:compare-op-asserts}, Listing~\ref{lst:loss} presents an assertion that ensures that the model performs well, without using a magic threshold.
+The assertion checks if the mean of the first 10 observations of the loss is higher compared to the mean of the last 10 observations.
+In other words, the assertion ensures that the loss of the model decreased after training, and that the model was able to ``learn'' from the training set.
+The assertion is derived from the visualisation presented in Figure~\ref{fig:loss}, which shows the loss of the model over the training epochs.
 
@@
 
-Visualisations presented in Figure~\ref{fig:data-pre-process} were obtained from a notebook that presents a tutorial on scaling and normalization---two common data pre-processing techniques in ML.
-The visualisations show a comparison of the data before and after each pre-processing technique is applied to a feature from the data.
-The author scales the data using the \texttt{MinMaxScaler} from the scikit-learn library~\footnote{REFME}, between $[0, 1]$.
-The author performs a visual check, by comparing the distribution of the data before and after scaling, using a histogram (as shown in Figure~\ref{fig:scale}).
-Similarly, the author normalises the data using the \texttt{PowerTransformer} class from scikit-learn~footnote{REFME}, and performs a visual check as shown in Figure~\ref{fig:normal}.
+Visualisations in Figure~\ref{fig:data-pre-process} were obtained from a notebook that presents a tutorial on scaling and normalization---two common data pre-processing techniques in ML.
+The visualisations show a comparison of the data before and after each pre-processing technique is applied to a feature from the dataset.
+The author performs a visual check, by comparing the distribution of the data before and after applying the pre-processing technique, using a histogram.
 
-Listing~\ref{lst:scale}~and~\ref{lst:normal} show the assertions that were defined below Figure~\ref{fig:scale}~and~\ref{fig:normal} respectively.
+Listing~\ref{lst:scale} and~\ref{lst:normal} show the assertions that were defined below Figure~\ref{fig:scale} and~\ref{fig:normal} respectively.
 Both assertions check that the min and max of the feature is within the expected bounds after applying the respective pre-processing technique.
-The assertion in Listing~\ref{lst:scale} is felt to be unneccessary since the min and max thresholds are set to the theoretical limits.
-The assertions in Listing~\ref{normal} are more relevant, however it is not clear why the author include such a wide margin in the thresholds.
+In contrast to assertions presented in Table~\ref{tab:compare-op-asserts}, it is clear that the thresholds are derived from the corresponding visualisations.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]
+assert not np.allclose(x_aug, x_original)
+\end{lstlisting}
+
+22\% of the assertions did not use a comparison operator, nor a magic threshold.
+Listing~\ref{lst:no-comp-magic} highlights such an assertion which is obtained from a notebook that performs image augmentation using Weighted K-Nearest Neighbours.
+The assertion compares the matrix representation of two images, and checks that the augmented image is different from the original.
+
+Most notebooks examined in this study do no use any external testing libraries.
+227 or 84\% of the VA pairs used the Python \texttt{assert} statement to define the test directly.
+This aligns with our observations that majority of the assertions compare the value of a performance metric against a user defined threshold, and do not require the use of a specialised testing library.
+
+This study found 42 VA pairs where an external testing library was used.
+Listing~\ref{lst:np-almost-equal} shows an assertion that uses the \texttt{assert\_almost\_equal} method from the \texttt{numpy.testing} module.
+The assertion checks if the loss of the model is similar to the specified value, up to a desired precision.
+Due to the stochastic nature of ML, results obtained from two separate development cycles will slightly differ.
+The assertion presented in Listing~\ref{lst:np-almost-equal} is found to be robust towards such perturbations and will only fail if there is a significant change in the performance of the model.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]
+ans4 = compute_loss(X_expanded, y, w)
+np.testing.assert_almost_equal(ans4, 0.3042764) 
+\end{lstlisting}
+
+\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq2-p3}
+
+We performed an in-depth analysis of 34 unique VA pairs.
+This was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
+During the analysis, several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data were captured.
+We find several instances of VA pairs where the assertion is unable to capture all the information obtained from the visualisation.
+
+\begin{figure}
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
+  \hfill
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
+  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
+\end{figure}
+
+\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
+assert accuracy_score(y_test, pred) > 0.95
+\end{lstlisting}
+
+Figure~\ref{fig:svm} and Listing~\ref{lst:svm} present one such pair.
+Visualisations~\ref{fig:svm-linear} and~\ref{fig:svm-rbf} are obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
+The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel.
+
+The visualisations contain several useful information that can be used to test and debug ML models.
+For instance, using Visualisation~\ref{fig:svm}, ML practitioners can get a visual understanding of how well the model is able to classify the two classes in the dataset.
+Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
+The visualisation also allows ML practitioners to compare different ML models or variants of the same ML model to iteratively find a model that best fits the underlying training data.
+In this case for instance, the author of the notebook first trained the SVM with a linear kernel.
+And upon realising that the data is not linearly separable, chose the RBF kernel.
+
+Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
+The assertion checks if the model accuracy is above a threshold specified by the author.
+The assertion however, is only able to capture the information pertaining to the performance of the model.
+While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]
+r2_gru = r2_score(y_test, y_pred)
+assert r2_gru > 0.6
+\end{lstlisting}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
+  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
+\end{figure}
+
+Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
+The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
+A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data.
+The spread of $y_{pred}$ at any given $y_{true}$ is wide.
+For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
+This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.
+
+Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
+The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computes the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
+The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
+Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
+
+Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess if the model is under or overfitting the training data.
+If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
+In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
+The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
+The assertion in Listing~\ref{lst:r2} however, does not capture such information.
+It can only check if the performance of the model is better than a specific threshold.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
+The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
+for feature in range(data_transformed.shape[1]):
+assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
+\end{lstlisting}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
+  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
+\end{figure}
+
+In contrast to Listing~\ref{lst:svm} and Listing~\ref{lst:r2}, Figure~\ref{fig:kstest} and Listing~\ref{lst:kstest} present a VA pair where the assertion captures all the information obtained from the visualisation.
+
+The notebook author normalises the training data and performs a visual check using the visualisation presented in Figure~\ref{fig:kstest}.
+The visualisation shows a bell-shaped curve thus indicating that the data is normal and that the transformation was successful.
+The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
+The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
+
+Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
+Several ML models assume normality in the underlying training data.
+However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
+The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
+And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
 
 \section{Implications}\label{sec:discuss}
-
-This study mined 54K Jupyter Notebooks that contained an assertion.
-Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation.
-Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.
-A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
 
@@
 \section{Threats to Validity and Future Work}\label{sec:threats}
-% TODO discuss that we only consider viz-assert pairs that could be readily applied to ML development lifecycle.
-However, we also found several visualisation-assert pairs that target the more fundamental concepts of ML.
-Further work is required to explore such pairs and check if they are useful...
-%
+
 The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
@@
 
+This study mined 54K Jupyter Notebooks that contained an assertion.
+Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation.
+Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.
+A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
+
 \bibliographystyle{IEEEtran}

### 2023-10-25 report/icst24.tex
small adjustment results

@@
 
-202 assertions from the 269 VA pairs (75\%) contain a comparison operator such as \emph{equal to}, \emph{lesser than} or \emph{greater than}.
+In a total of 269 VA pairs, 202 assertions (75\%) contain a comparison operator such as \emph{equal to}, \emph{lesser than} or \emph{greater than}.
 Table~\ref{tab:compare-op-asserts} presents the most frequently observed assertions that conform to this pattern.
@@
 
-46\% of the assertions with a comparison operator, also contain a ``magic threshold'' defined by the author.
+We also observe that 46\% of the assertions with a comparison operator contain a ``magic threshold'' defined by the author.
 It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.

### 2023-10-25 report/icst24.tex
minor adjustment results

@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+%%%TRASH luis
+\usepackage{pifont}
+\newcommand{\luis}[1]{\textcolor{teal}{\ding{46}~\textbf{Luis:~}#1}}
+%%%%%
 
@@
 Table~\ref{tab:compare-op-asserts} presents the most frequently observed assertions that conform to this pattern.
+
 An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
-
-We also observe that 46\% of the assertions with a comparison operator contain a ``magic threshold'' defined by the author.
+This covers 46\% of the assertions with a comparison operator.
 It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
-The threshold could have been derived based on the domain expertise of the ML practitioner.
-Another alternative could be to define them during the Requirements Engineering phase, though discussions with all involved stakeholders\cite{CITEME}.
+The threshold could have been derived based on domain expertise.
+Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{CITEME}.\luis{Agathe Balayn might have something on these lines}
 

### 2023-10-25 report/icst24.tex
squash

@@
 \usepackage{framed}
+\usepackage{tabularx}
+\usepackage{booktabs}
 \newcommand{\highlight}[1]{\begin{framed}%
@@
 \definecolor{backcolor}{rgb}{0.95,0.95,0.92}
+\colorlet{punct}{red!60!black}
+\definecolor{delim}{RGB}{20,105,176}
+\colorlet{numb}{magenta!60!black}
 
 \lstdefinestyle{mystyle}{
+  language=Python,
   basicstyle=\ttfamily\footnotesize,
-  keywordstyle=\color{blue},
-  stringstyle=\color{red},
-  commentstyle=\color{green},
+  keywordstyle=\color{delim},
+  stringstyle=\color{punct},
+  commentstyle=\color{numb},
   breakatwhitespace=true,
@@
   showtabs=false,
-  tabsize=4,
+  tabsize=2,
+  stepnumber=1,
+  numbersep=3pt,
   captionpos=b,
@@
   numberstyle=\tiny\color{gray},
-  backgroundcolor=\color{backcolor},
+  backgroundcolor=\color{backcolor}
 }
 \lstset{style=mystyle}
+
+\lstset{emph={%  
+    assert%
+    },emphstyle={\color{delim}}%
+}%
+
 \usepackage{graphicx}
@@
 \begin{description}
-\item[RQ1.] How frequently are analytical tests formulated from visualisations created to test ML systems?
+  \item[RQ1.] \textbf{How frequently are analytical tests formulated from visualisations created to test ML systems?}
 
@@
 
-\item[RQ2.] What patterns are frequently observed in VA pairs used to test ML systems?
+\item[RQ2.] \textbf{What patterns are frequently observed in VA pairs used to test ML systems?}
 
@@
 
-\item[RQ3.] Do the assertions capture all the information obtained from the corresponding visualisation?
+\item[RQ3.] \textbf{Do the assertions capture all the information obtained from the corresponding visualisation?}
 
@@
 And help ML practitioners to formulate better analytical assertions from visualisations.
+
+The replication package for this study is made public on Figshare\footnote{The replication package can be found at this URL: https://figshare.com/s/9347228011999cba29f5}.
 
@@
 \caption{Frequent examples of post-condition checks observed in this study.}
-  \begin{tabular}{m{0.2\textwidth} m{0.4\textwidth} m{0.3\textwidth}}
-\hline
+  \begin{tabular}{@{}m{0.2\textwidth} m{0.42\textwidth} m{0.28\textwidth}@{}}
+\toprule
 \emph{\textbf{Visualisation}}&
@@
 \emph{\textbf{Rationale}}\\
-\hline
+\midrule
 \includegraphics[width=\linewidth]{post-cond-01.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
 assert f1.gca().has_data()
@@
 \includegraphics[width=\linewidth]{post-cond-02.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
 assert_almost_equal(sigmoid(-50), 0, delta=1e-10)
@@
 \includegraphics[width=\linewidth]{post-cond-03.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
 # The axis should contain 5 bars
@@
 The asserts are validating the correctness of the visualisation itself.\\
-\hline
+\bottomrule
 \end{tabular}
@@
 Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
-The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
+The final search query is as follows: \texttt{"image/png" AND "assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
 
@@
 The string patterns are derived from the python module name used in the import statements.
-The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
+The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy| pandas}.
 
@@
   cell arrangement.}
-  \begin{tabular}{l p{0.2\textwidth} p{0.2\textwidth}}
-    \hline
+  \begin{tabular}{@{}l p{0.13\textwidth} p{0.27\textwidth}@{}}
+    \toprule
     \emph{\textbf{ID}}&
@@
     \emph{\textbf{Observations}}\\
-    \hline
+    \midrule
     1 &
@@
     Least likely to be related to the visualisation below.\\
-    \hline
+    \midrule
     \multicolumn{3}{l}{\textbf{Assertion and visualisation in the same code cell}}\\
-    \hline
+    \midrule
     2.1 &
@@
     More likely to be a post-condition check.\\
-    \hline
+    \midrule
     \multicolumn{3}{l}{\textbf{Assertion cell below visualisation cell}}\\
-    \hline
+    \midrule
     3.1 &
@@
     Most likely to be related to the visualisation above.\\
-    \hline
+    \bottomrule
   \end{tabular}
@@
   \caption{Inclusion-Exclusion criteria used to identify related visualisation-assertion pairs.}\label{tab:exclusion-criteria}
-  \begin{tabular}{l p{0.2\textwidth} r}
-    \hline
+  \begin{tabularx}{0.47\textwidth}{@{}l X r@{}}
+    \toprule
     \textbf{ID} &
     \textbf{Exclusion Criteria} &
-    \textbf{Num. excluded NBs}\\
-    \hline
+    \textbf{\# excluded NBs}\\
+    \midrule
     \textbf{EC1} &
@@
     34\\
-    \hline
+    \midrule
     \multicolumn{2}{c}{\emph{\textbf{Total number of NBs excluded:}}} &
     1507\\
-    \hline
-  \end{tabular}
+    \bottomrule
+  \end{tabularx}
 \end{table}
@@
   \caption{Testing patterns observed in the visualisation-assertion pairs.}
-  \begin{tabular}{l p{0.3\textwidth} c c}
-    \hline
+  \begin{tabularx}{0.45\textwidth}{@{}l X r r@{}}
+    \toprule
     \textbf{ID} &
@@
     & & Yes & No\\
-    \hline
+    \midrule
     \textbf{P1} &
@@
     \textbf{P3} &
-    Does the assertion use an external testing library (such as \texttt{numpy.testing}? &
+    Does the assertion use an external testing library (such as \texttt{numpy.testing})? &
     42 & 227\\
-    \hline
-  \end{tabular}
+    \bottomrule
+  \end{tabularx}
   \label{tab:testing-patterns}
@@
   \caption{Most frequently occurring assertions that use a comparison operator.}
-  \begin{tabular}{m{0.9\linewidth}}
-    \hline
-    \begin{lstlisting}[language=Python]
-assert accuracy_score(y_test, pred) > 0.95
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
-assert HISTORY['test_accuracy'][-1] > 0.92
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
+  \begin{tabular}{@{}m{0.9\linewidth}@{}}
+    \toprule
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
 assert np.mean(mean_rw_history[-10:]) > 10.
-print("That's good enough for tutorial.")
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
-assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."
-    \end{lstlisting}\\
-    \begin{lstlisting}
-assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
-    \end{lstlisting}\\
-    \hline
+print("That's good enough for tutorial.")\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
+    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."\end{lstlisting}\\
+    \bottomrule
   \end{tabular}
@@
 
-An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
+An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author\luis{we need to absolute values here (out of 269)}.
 This covers 46\% of the assertions with a comparison operator.
@@
 
-Most notebooks examined in this study do no use any external testing libraries.
-227 or 84\% of the VA pairs used the Python \texttt{assert} statement to define the test directly.
+Most notebooks examined in this study do not use any external testing libraries.
+227 or 84\% of the VA pairs use the low-level Python \texttt{assert} statement to define the test directly.
 This aligns with our observations that majority of the assertions compare the value of a performance metric against a user defined threshold, and do not require the use of a specialised testing library.
@@
 
-The results from RQ2 provide emprical evidence to support the use of visualisations to test various properties of ML systems.
+The results from RQ2 provide empirical evidence to support the use of visualisations to test various properties of ML systems.
 Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues.
@@
 These observations should not be violated by the future model development cycles unless the requirements of the ML system are changed.
-Assertions scale across organizational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
+Assertions scale across organisational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
 
-Results for RQ1 indicate that combining ML visualisations with analytical assertions is an emerging testing practice within the ML community.
-However in results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-Our observations indicate a need to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
-We find the need for automated tools that make this translation process easier for AI practitioners.
+The results for RQ1 indicate that the combination of machine learning visualisations with analytical assertions is an emerging testing practice within the machine learning community.
+We found only 54,000 notebooks that contain the keyword ``assert'' and an image, while there are more than a million Jupyter notebooks available on GitHub.
+This suggests that the practice of assertion is still relatively uncommon.
+We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
+We argue that we need more guidelines on ML coding practices and that education and research of software engineering practices ought to be steered towards the specific challenges entailed from contributing to ML software.
+We delve into this issue further in Section~\ref{sec:fw} \luis{I dont see this in fw and I wouldr remove this sentence because it breaks the flow of the narrative}.
+
+\luis{Amother thought: there are still many visualisations without asserts.
+We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors.
+this is already in the motivation but we can frame it as a visible open problem here as well.}
+
+Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
+Very rarely assertions use dedicated API test methods.
+We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
+
+\luis{some thought that is blended here but maybe we want to make it clear.
+ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics).
+This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}
+
+The wide adoption of ``magic'' thresholds (\luis{add numbers from results here}) is also concerning as it can hinder to the maintainability of ML projects.
+While in some cases there these thresholds come a results of domain expertise and experience acquired while working in a give project, we argue that such thresholds need to be properly documented.
+Our results show that there is a need to help practitioners keeping these decisions documented.
+This could be automated using a tool to flag undocumented decisions or that would suggest boilerplate documentation in Markdown or code annotations.
+
+Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met.
+After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer.
+For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf.
+Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6).
+Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
+In other words, there is a need for automated tools that make this translation process easier for AI practitioners.
+In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.
 
@@
 We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github.
-We invite the scientific community to extend our catalogue of visual and analytical testing patterns, using the corpus of notebooks provided by the above authors.
+However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}).
+We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.
 
 In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose, unittest and numpy).
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose\footnote{}, unittest and numpy).
 Our approach provides captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in all Python testing libraries.
@@
 
-This study mined 54K Jupyter Notebooks that contained an assertion.
-Further analysis identified 1.7K notebooks which contained an assertion directly below a visualisation.
-Of the 1.7K notebooks, 551 notebooks contained a visualisation and assertion pair which were semantically related to each other.
-A deeper analysis of 90 Jupyter notebooks revealved 26 unique ML testing tasks which utilised a combination of visualisations and assertions.
+In this study, we delved into the fusion of visualizations and analytical assertions in the testing of machine learning (ML) systems.
+Our investigation unveiled the emerging practice of crafting analytical assertions based on visual insights, a technique that bridges the qualitative and quantitative aspects of ML testing.
+
+Testing ML systems in Jupyter Notebooks remains a burgeoning practice, with only a minority of notebooks adopting this approach.
+In total, we identified 269 pairs of visualizations and assertions that exhibited semantic correlation within 1.7k notebooks.
+Within these pairs, our analysis unveiled several key insights: nearly 40\% of the pairs included magic thresholds, 75\% of the assertions employed comparison operators, and 16\% of the assertions made use of external testing libraries.
+Furthermore, we observed that 22\% of the assertions did not utilize a comparison operator or a magic threshold.
+
+Additionally, our study brought to light recurring testing patterns, underscoring the need yo enhance the alignment between visualizations and their corresponding assertions.
+This research serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualizations and elevate the quality of analytical assertions.
+
+\section{Future Work}\label{sec:fw}
+
+This study underscores the imperative to enhance the current state of machine learning (ML) testing. Our observations have pointed to several potential future directions that could advance this field:
+
+1.
+Tailored Assertion Tools: There is a clear need for assertion tools specifically designed to cater to the requirements of ML practitioners.
+These tools should provide assistance in identifying the most suitable visualizations for testing specific properties of ML systems.
+Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.
+
+2.
+Automation of Assertions: The manual translation of insights derived from visualizations into analytical assertions presents a substantial challenge.
+Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
+By automating the generation of assertions from visual information, these tools can promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort.
+
+3.
+Dynamic Testing in Real-World Scenarios: Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration.
+This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
+Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.
 

### 2023-10-25 report/icst24.tex
add assert as a keyword for listings

--- 2023-10-25 report/icst24.tex
@@
   backgroundcolor=\color{backcolor},
+  morekeywords= [2]{assert},
 }
@@
 \includegraphics[width=\linewidth]{post-cond-01.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}{language=Python}
 assert f1.gca().has_data()
@@
 \includegraphics[width=\linewidth]{post-cond-02.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}{language=Python}
 assert_almost_equal(sigmoid(-50), 0, delta=1e-10)
@@
 \includegraphics[width=\linewidth]{post-cond-03.png}&
-\begin{lstlisting}[language=Python]
+\begin{lstlisting}{language=Python}
 # The axis should contain 5 bars
@@
     \hline
-    \begin{lstlisting}[language=Python]
+    \begin{lstlisting}{language=Python}
 assert accuracy_score(y_test, pred) > 0.95
     \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
+    \begin{lstlisting}{language=Python}
 assert HISTORY['test_accuracy'][-1] > 0.92
     \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
+    \begin{lstlisting}{language=Python}
 assert np.mean(mean_rw_history[-10:]) > 10.
@@
     \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
+    \begin{lstlisting}{language=Python}
 assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."
     \end{lstlisting}\\
-    \begin{lstlisting}
+    \begin{lstlisting}{language=Python}
 assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
-In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]
+\begin{lstlisting}[caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
+In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]{language=Python}
 assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
+\begin{lstlisting}[caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]{language=Python}
 assert scaled_data[scaled_data.argmax()] <= 1
@@
 
-\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
+\begin{lstlisting}[caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]{language=Python}
 assert normalized_data[normalized_data.argmax()] < 10
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]
+\begin{lstlisting}[caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]{language=Python}
 assert not np.allclose(x_aug, x_original)
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]
+\begin{lstlisting}[caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]{language=Python}
 ans4 = compute_loss(X_expanded, y, w)
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
+\begin{lstlisting}[caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]{language=Python}
 assert accuracy_score(y_test, pred) > 0.95
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]
+\begin{lstlisting}[caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]{language=Python}
 r2_gru = r2_score(y_test, y_pred)
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
+\begin{lstlisting}[caption={Assertion to check that each feature in a dataset is normal.
+The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]{language=Python}
 for feature in range(data_transformed.shape[1]):

### 2023-10-25 report/icst24.tex
improve Implications, conclusion and future work

@@
 
-Results for RQ1 indicate that combining ML visualisations with analytical assertions is an emerging testing practice within the ML community.
-However in results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+The results for RQ1 indicate that the combination of machine learning visualizations with analytical assertions is an emerging testing practice within the machine learning community.
+We found only 54,000 notebooks that contain the keyword "assert" and an image, while there are more than a million Jupyter notebooks available on GitHub.
+This suggests that the practice of assertion is still relatively uncommon.
+We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
+We delve into this issue further in Section~\ref{sec:fw}.
+
+Additionally the results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
 Our observations indicate a need to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.

### 2023-10-25 report/icst24.tex
revert listing config

@@
   numberstyle=\tiny\color{gray},
-  backgroundcolor=\color{backcolor},
-  morekeywords= [2]{assert},
+  backgroundcolor=\color{backcolor}
 }
@@
 \includegraphics[width=\linewidth]{post-cond-01.png}&
-\begin{lstlisting}{language=Python}
+\begin{lstlisting}[language=Python]
 assert f1.gca().has_data()
@@
 \includegraphics[width=\linewidth]{post-cond-02.png}&
-\begin{lstlisting}{language=Python}
+\begin{lstlisting}[language=Python]
 assert_almost_equal(sigmoid(-50), 0, delta=1e-10)
@@
 \includegraphics[width=\linewidth]{post-cond-03.png}&
-\begin{lstlisting}{language=Python}
+\begin{lstlisting}[language=Python]
 # The axis should contain 5 bars
@@
     \hline
-    \begin{lstlisting}{language=Python}
+    \begin{lstlisting}[language=Python]
 assert accuracy_score(y_test, pred) > 0.95
     \end{lstlisting}\\
-    \begin{lstlisting}{language=Python}
+    \begin{lstlisting}[language=Python]
 assert HISTORY['test_accuracy'][-1] > 0.92
     \end{lstlisting}\\
-    \begin{lstlisting}{language=Python}
+    \begin{lstlisting}[language=Python]
 assert np.mean(mean_rw_history[-10:]) > 10.
@@
     \end{lstlisting}\\
-    \begin{lstlisting}{language=Python}
+    \begin{lstlisting}[language=Python]
 assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."
     \end{lstlisting}\\
-    \begin{lstlisting}{language=Python}
+    \begin{lstlisting}[language=python]
 assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
@@
 
-\begin{lstlisting}[caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
-In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
+In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]
 assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."
@@
 
-\begin{lstlisting}[caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
 assert scaled_data[scaled_data.argmax()] <= 1
@@
 
-\begin{lstlisting}[caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
 assert normalized_data[normalized_data.argmax()] < 10
@@
 
-\begin{lstlisting}[caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]
 assert not np.allclose(x_aug, x_original)
@@
 
-\begin{lstlisting}[caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]
 ans4 = compute_loss(X_expanded, y, w)
@@
 
-\begin{lstlisting}[caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
 assert accuracy_score(y_test, pred) > 0.95
@@
 
-\begin{lstlisting}[caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]
 r2_gru = r2_score(y_test, y_pred)
@@
 
-\begin{lstlisting}[caption={Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]{language=Python}
+\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
+The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
 for feature in range(data_transformed.shape[1]):

### 2023-10-25 report/icst24.tex
squash

@@
 \lstset{style=mystyle}
+
+\lstset{emph={%  
+    assert%
+    },emphstyle={\color{delim}}%
+}%
+
 \usepackage{graphicx}
@@
 We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls.
-In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
+In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation.
+Listing~\ref{lst:prompt} shows the prompt used for such cases.
+In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair.
+Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.
+
+\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to understand visualisation code.}, label={lst:prompt}]
+Explain the following Python code:
+
+<code for visualisation>
+
+\end{lstlisting}
+
+\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to confirm if a VA pair are related to each other.}, label={lst:prompt-follow-up}]
+Is the following Python assertion related to the visualisation code?
+
+<code for assertion>
+
+\end{lstlisting}
 
@@
   \caption{Most frequently occurring assertions that use a comparison operator.}
-  \begin{tabular}{m{0.9\linewidth}}
-    \hline
-    \begin{lstlisting}[language=Python]
-assert accuracy_score(y_test, pred) > 0.95
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
-assert HISTORY['test_accuracy'][-1] > 0.92
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
+  \begin{tabular}{@{}m{0.9\linewidth}@{}}
+    \toprule
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
 assert np.mean(mean_rw_history[-10:]) > 10.
-print("That's good enough for tutorial.")
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=Python]
-assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."
-    \end{lstlisting}\\
-    \begin{lstlisting}[language=python]
-assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
-    \end{lstlisting}\\
-    \hline
+print("That's good enough for tutorial.")\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
+    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."\end{lstlisting}\\
+    \bottomrule
   \end{tabular}
@@
 These observations should not be violated by the future model development cycles unless the requirements of the ML system are changed.
-Assertions scale across organizational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
+Assertions scale across organisational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
 
-The results for RQ1 indicate that the combination of machine learning visualizations with analytical assertions is an emerging testing practice within the machine learning community.
-We found only 54,000 notebooks that contain the keyword "assert" and an image, while there are more than a million Jupyter notebooks available on GitHub.
+The results for RQ1 indicate that the combination of machine learning visualisations with analytical assertions is an emerging testing practice within the machine learning community.
+We found only 54,000 notebooks that contain the keyword ``assert'' and an image, while there are more than a million Jupyter notebooks available on GitHub.
 This suggests that the practice of assertion is still relatively uncommon.
 We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
-We delve into this issue further in Section~\ref{sec:fw}.
+We argue that we need more guidelines on ML coding practices and that education and research of software engineering practices ought to be steered towards the specific challenges entailed from contributing to ML software.
+We delve into this issue further in Section~\ref{sec:fw} \luis{I dont see this in fw and I wouldr remove this sentence because it breaks the flow of the narrative}.
 
-Additionally the results from RQ2, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-Our observations indicate a need to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
-We find the need for automated tools that make this translation process easier for AI practitioners.
+\luis{Amother thought: there are still many visualisations without asserts.
+We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors.
+this is already in the motivation but we can frame it as a visible open problem here as well.}
+
+Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
+Very rarely assertions use dedicated API test methods.
+We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
+
+\luis{some thought that is blended here but maybe we want to make it clear.
+ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics).
+This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}
+
+The wide adoption of ``magic'' thresholds (\luis{add numbers from results here}) is also concerning as it can hinder to the maintainability of ML projects.
+While in some cases there these thresholds come a results of domain expertise and experience acquired while working in a give project, we argue that such thresholds need to be properly documented.
+Our results show that there is a need to help practitioners keeping these decisions documented.
+This could be automated using a tool to flag undocumented decisions or that would suggest boilerplate documentation in Markdown or code annotations.
+
+Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met.
+After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer.
+For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf.
+Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6).
+Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
+In other words, there is a need for automated tools that make this translation process easier for AI practitioners.
+In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.
 
@@
 The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
-To the best of our knowledge, \cite{pimentel2019large} and \cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
+To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
 We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github.
-We invite the scientific community to extend our catalogue of visual and analytical testing patterns, using the corpus of notebooks provided by the above authors.
+However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}).
+We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.
 
 In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose, unittest and numpy).
-Our approach provides captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in all Python testing libraries.
-
-% TODO don't mention quantity of random sample; then we need to explain/defend this choice
-We used an iterative approach to determine the effectiveness of each filter that was used.
-We applied each filter in the same order as presented in Table~\ref{tab:filters}, starting with F1.
-We randomly sampled 50 notebooks from the population of notebooks obtained after applying the filter.
-We then conduct the qualitative assessment presented in Section~\ref{REFME} and observe the number of related visualisation and assertion pairs.
-In each new iteration, we progressively add the other filters and include the filter if the number of relevant visualisation and assertion pairs increases.
-
-% TODO defend use of one-shot prompt engineering for identifying related visualisation and assertion code cells; how did we mitigate this threat?
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
+Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in several Python testing libraries.
 

### 2023-10-25 report/icst24.tex
improve table and listing presentation

--- 2023-10-25 report/icst24.tex
@@
 \caption{Frequent examples of post-condition checks observed in this study.}
-  \begin{tabular}{m{0.2\textwidth} m{0.4\textwidth} m{0.3\textwidth}}
-\hline
+  \begin{tabular}{@{}m{0.2\textwidth} m{0.4\textwidth} m{0.3\textwidth}@{}}
+\toprule
 \emph{\textbf{Visualisation}}&
@@
 Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
-The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
+The final search query is as follows: \texttt{"image/png" AND "assert" language:"Jupyter Notebook"}.
 

### 2023-10-25 report/icst24.tex
highlight asserts
add collection date
feat(report): 2023-10-25 13:54 updates

--- 2023-10-25 report/icst24.tex
@@
 We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
-We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
+We mine Github to collect $54,070$ Jupyter Notebooks that contain assertions written in Python.
 We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
 We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
-We further investigate of the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns.
+We further investigate the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns.
 We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part.
@@
 
-% TODO make sure that we use the terminology of "Testing Tasks" everywhere for RQ2
 Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
@@
 
-We manually analysed the source code of the visualisation and assertion code cells.
+We malysed the source code of the visualisation and assertion code cells.
 And excluded VA pairs that were not related to each other.
@@
 \subsection{RQ1: How frequently are analytical tests formulated from visualisations created to test ML systems?}\label{sec:result-rq1}
+
+% TODO add the 54K NB found with assert here as well
 
@@
 
-\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq2-p3}
+\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq3}
 
@@
 
-The results for RQ1 indicate that the combination of machine learning visualisations with analytical assertions is an emerging testing practice within the machine learning community.
-We found only 54,000 notebooks that contain the keyword ``assert'' and an image, while there are more than a million Jupyter notebooks available on GitHub.
+The results for RQ1 indicate that the combination of ML visualisations with analytical assertions is an emerging testing practice within the machine learning community.
+We found only $54,070$ notebooks that contain the keyword ``assert'', while there are more than a million Jupyter notebooks available on GitHub.
 This suggests that the practice of assertion is still relatively uncommon.
 We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
-We argue that we need more guidelines on ML coding practices and that education and research of software engineering practices ought to be steered towards the specific challenges entailed from contributing to ML software.
-We delve into this issue further in Section~\ref{sec:fw} \luis{I dont see this in fw and I wouldr remove this sentence because it breaks the flow of the narrative}.
+We argue that guidelines on ML coding practices require improvement.
+And the need to steer education and research of Software Engineering practices towards the specific challenges faced  when contributing to ML software systems.
 
-\luis{Amother thought: there are still many visualisations without asserts.
-We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors.
-this is already in the motivation but we can frame it as a visible open problem here as well.}
+In the 1764 notebooks analysed in this study, several visualisations lack assertions.
+However, unformalised expectations still exist within each visualisation which cannot be readily validated by other participants.
+ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts.
+This highlights the need for refining the education for these professionals and crafting tools to compensate for potential gaps in their knowledge.
 
 Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
-Very rarely assertions use dedicated API test methods.
-We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
+We hypothesise that this problem stems from two fundamental sources.
+The assertions in the VA pairs identified in this study lack dedicated testing libraries.
+This indicates that the current culture of testing ML code is still in its infancy and ML practitioners are not yet exploring software testing libraries.
+Second, the existing testing libraries are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.
 
-\luis{some thought that is blended here but maybe we want to make it clear.
-ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics).
-This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}
+% TODO \luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
 
-The wide adoption of ``magic'' thresholds (\luis{add numbers from results here}) is also concerning as it can hinder to the maintainability of ML projects.
-While in some cases there these thresholds come a results of domain expertise and experience acquired while working in a give project, we argue that such thresholds need to be properly documented.
-Our results show that there is a need to help practitioners keeping these decisions documented.
-This could be automated using a tool to flag undocumented decisions or that would suggest boilerplate documentation in Markdown or code annotations.
+Results from RQ2 show the prevalent use of ``magic'' thresholds in assertions, which can impede the maintainability of ML projects.
+Even if these thresholds arise from domain expertise or experience obtained over time, it is crucial they are thoroughly documented.
+Our findings indicate a clear need to assist practitioners in documenting these decisions.
+Potential solutions include automated tools that highlight undocumented choices or suggest standardized documentation in Markdown or code annotations.
 
-Additionally, in the results from RQ2\luis{this is RQ3}, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-We argue that this can lead to a mismatch of expectations in future versions of the ML software: for example, another collaborator of the project might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also being met.
-After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer.
-For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf.
-Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6).
-Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
-In other words, there is a need for automated tools that make this translation process easier for AI practitioners.
-In sum, our results showcase a gap in the current state of the art in software testing: it is failing to address the idiosyncrasies of ML code.
+In the results from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+We argue that this can lead to a mismatch of expectations in future versions of the ML software.
+For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
+The examples presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
+Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of a ML developer.
+
+For example, a linear correlation between the true and predicted labels is evident in Figure~{fig:r2}.
+Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric.
+Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
+There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
 
@@
 
+This study finds a clear need to improve the current ML testing process.
+There is a clear need for tools specifically designed to aid ML practitioners when testing ML systems.
+These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems.
+Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.
+
+The manual translation of insights derived from visualisations into analytical assertions presents a substantial challenge.
+Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
+By automating the generation of assertions from visual information, these tools will promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort and bringing the code developed within the notebooks, closer to a production-ready codebase.
+
+Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration.
+This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
+Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.
+
 \section{Conclusion}\label{sec:conclude}
 
-In this study, we delved into the fusion of visualizations and analytical assertions in the testing of machine learning (ML) systems.
-Our investigation unveiled the emerging practice of crafting analytical assertions based on visual insights, a technique that bridges the qualitative and quantitative aspects of ML testing.
+This study study revealed the emerging practice of crafting analytical assertions based on visual insights to bridge the gap between qualitative and quantitative aspects of ML testing.
 
-Testing ML systems in Jupyter Notebooks remains a burgeoning practice, with only a minority of notebooks adopting this approach.
-In total, we identified 269 pairs of visualizations and assertions that exhibited semantic correlation within 1.7k notebooks.
-Within these pairs, our analysis unveiled several key insights: nearly 40\% of the pairs included magic thresholds, 75\% of the assertions employed comparison operators, and 16\% of the assertions made use of external testing libraries.
-Furthermore, we observed that 22\% of the assertions did not utilize a comparison operator or a magic threshold.
+We identified 269 visualisation-assertion pairs from 1.7K Jupyter notebooks, that were semantically related to one another.
+And identified three recurring testing patterns within these pairs.
+Nearly 40\% of the assertions used ``magic'' thresholds, 75\% of the assertions used comparison operators, and 16\% of the assertions used external testing libraries.
 
-Additionally, our study brought to light recurring testing patterns, underscoring the need yo enhance the alignment between visualizations and their corresponding assertions.
-This research serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualizations and elevate the quality of analytical assertions.
-
-\section{Future Work}\label{sec:fw}
-
-This study underscores the imperative to enhance the current state of machine learning (ML) testing. Our observations have pointed to several potential future directions that could advance this field:
-
-1.
-Tailored Assertion Tools: There is a clear need for assertion tools specifically designed to cater to the requirements of ML practitioners.
-These tools should provide assistance in identifying the most suitable visualizations for testing specific properties of ML systems.
-Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.
-
-2.
-Automation of Assertions: The manual translation of insights derived from visualizations into analytical assertions presents a substantial challenge.
-Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
-By automating the generation of assertions from visual information, these tools can promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort.
-
-3.
-Dynamic Testing in Real-World Scenarios: Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration.
-This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
-Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.
+This study serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualisations and elevate the quality of analytical assertions.
 

### 2023-10-25 report/icst24.tex
add items to implications

--- 2023-10-25 report/icst24.tex
@@
 \begin{description}
-  \item[RQ1.] \textbf{How frequently are analytical tests formulated from visualisations created to test ML systems?}
+\item[RQ1.] How frequently are analytical tests formulated from visualisations created to test ML systems?
 
@@
 
-\item[RQ2.] \textbf{What patterns are frequently observed in VA pairs used to test ML systems?}
+\item[RQ2.] What patterns are frequently observed in VA pairs used to test ML systems?
 
@@
 
-\item[RQ3.] \textbf{Do the assertions capture all the information obtained from the corresponding visualisation?}
+\item[RQ3.] Do the assertions capture all the information obtained from the corresponding visualisation?
 
@@
 And help ML practitioners to formulate better analytical assertions from visualisations.
-
-The replication package for this study is made public on Figshare\footnote{The replication package can be found at this URL: https://figshare.com/s/9347228011999cba29f5}.
 
@@
 We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls.
-In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation.
-Listing~\ref{lst:prompt} shows the prompt used for such cases.
-In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair.
-Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.
-
-\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to understand visualisation code.}, label={lst:prompt}]
-Explain the following Python code:
-
-<code for visualisation>
-
-\end{lstlisting}
-
-\begin{lstlisting}[caption={Prompt used for ChatGPT 4.0 model to confirm if a VA pair are related to each other.}, label={lst:prompt-follow-up}]
-Is the following Python assertion related to the visualisation code?
-
-<code for assertion>
-
-\end{lstlisting}
+In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to determine if the visualisation and assertion were related to each other.
 
@@
 
-\luis{Amother thought: there are still many visualisations without asserts.
-We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors.
-this is already in the motivation but we can frame it as a visible open problem here as well.}
-
 Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
@@
 We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
-
-\luis{some thought that is blended here but maybe we want to make it clear.
-ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics).
-This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}
 
@@
 After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer.
-For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf.
-Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6).
-Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
+We ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
 In other words, there is a need for automated tools that make this translation process easier for AI practitioners.
@@
 The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
-To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
+To the best of our knowledge, \cite{pimentel2019large} and \cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
 We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github.
-However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}).
-We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.
+We invite the scientific community to extend our catalogue of visual and analytical testing patterns, using the corpus of notebooks provided by the above authors.
 
 In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
-Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in several Python testing libraries.
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose, unittest and numpy).
+Our approach provides captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in all Python testing libraries.
+
+% TODO don't mention quantity of random sample; then we need to explain/defend this choice
+We used an iterative approach to determine the effectiveness of each filter that was used.
+We applied each filter in the same order as presented in Table~\ref{tab:filters}, starting with F1.
+We randomly sampled 50 notebooks from the population of notebooks obtained after applying the filter.
+We then conduct the qualitative assessment presented in Section~\ref{REFME} and observe the number of related visualisation and assertion pairs.
+In each new iteration, we progressively add the other filters and include the filter if the number of relevant visualisation and assertion pairs increases.
+
+% TODO defend use of one-shot prompt engineering for identifying related visualisation and assertion code cells; how did we mitigate this threat?
 

### 2023-10-25 report/icst24.tex
typo indent
adjust implications again

@@
 
+\luis{Amother thought: there are still many visualisations without asserts.
+We believe in each visualisation there is an expectation that has not been formalised and that it will not be easily validated by other contributors.
+this is already in the motivation but we can frame it as a visible open problem here as well.}
+
 Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
@@
 We argue that the root of this problem is two-fold: 1) the culture of testing ML code is still in its early stages and ML practitioners are not yet exploring software testing libraries, and 2) the existing testing APIs are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.\luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
+
+\luis{some thought that is blended here but maybe we want to make it clear.
+ML testing is a task that requires practitioners to have a strong background in machine learning, software testing, and mathematics (or statistics).
+This reinforces the need to adjust the education of these professionals and to design tools that aim mitigate the lack of background in some of these directions.}
 
@@
 After delving into some examples, we also observe that deriving assertions from validations is far from trivial: basic visualisation patterns require complex mathematical techniques that might be out of the expertise of the ML developer.
-We ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
+For example, the linear correlation observed with the scatter plot of Figure 6\luis{add ref} is clear to any non-expert but it had to be formalised (cf.
+Listing 11\luis{add ref}) using the metric coefficient of determination bounded to a sound threshold (0.6).
+Hence, we ought to bridge the gap between qualitative assessment of ML systems using visualisations and deriving analytical assertions from the visualisations.
 In other words, there is a need for automated tools that make this translation process easier for AI practitioners.

### 2023-10-25 report/icst24.tex
feat(report): 2023-10-25 Wed 14:22 updates

--- 2023-10-25 report/icst24.tex
@@
 \begin{IEEEkeywords}
-  ML Testing, Visualisations, ???
+  ML Testing, Visualisations, Assertions, Computational Notebooks
 \end{IEEEkeywords}
@@
 During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
-Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
+Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
 
@@
 Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
-However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
+However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
 Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.

### 2023-10-25 report/icst24.tex
feat(report): 2023-10-25 17:10 updates

@@
 We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
-We further investigate the 269 visualisation-assertion pairs and identify three frequently occuring testing patterns.
+We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns.
 We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part.
@@
 
-% TODO still needs improvement
-This study uses Jupyter Notebooks mined from Github.
-Jupyter notebooks have been widely adopted by the Data Science and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
 Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.
 Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
-We leverage the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
+This study leverages the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
 
@@
 
-We programmatically parse the JSON structure of the notebooks (more details regarding the internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}) and represent each notebook as a pandas dataframe for further analysis.
+We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis.
 Each cell is mapped as a row in the dataframe.
@@
 
-We malysed the source code of the visualisation and assertion code cells.
+We analysed the source code of the visualisation and assertion code cells.
 And excluded VA pairs that were not related to each other.
@@
 Listing~\ref{lst:prompt} shows the prompt used for such cases.
-In a few cases, where the explaination of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair.
+In a few cases, where the explanation of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair.
 Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.
@@
 
-% TODO add the 54K NB found with assert here as well
-
+Although Github contains more than a million Jupyter notebooks, only 54K notebooks contained the keyword ``assert'' in them.
 From the 1764 Jupyter notebooks that were analysed in this study, we found 269 VA pairs that were semantically related to each other.
@@
     \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
-assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."\end{lstlisting}\\
+assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
+    \end{lstlisting}\\
     \bottomrule
@@
 The threshold could have been derived based on domain expertise.
-Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{CITEME}.\luis{Agathe Balayn might have something on these lines}
+Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirementsi,rahimi2019toward}.
 
@@
   \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
-  \caption{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
+  \caption{Scatter plot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
 \end{figure}
@@
 Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
-The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
+The visualisation shows a scatter plot of the real and predicted values of $NO_2$ from the test set.
 A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data.
@@
 
-The results from RQ2 provide empirical evidence to support the use of visualisations to test various properties of ML systems.
+The results from RQ1 provide empirical evidence to support the use of visualisations to test various properties of ML systems.
 Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues.
@@
 In the 1764 notebooks analysed in this study, several visualisations lack assertions.
-However, unformalised expectations still exist within each visualisation which cannot be readily validated by other participants.
+However, implicit expectations still exist within each visualisation which cannot be readily validated by other participants.
 ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts.
@@
 Second, the existing testing libraries are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.
-
-% TODO \luis{Side thought: we could also frame this is as: our results call for more tools that help testing ML code and that educational materials are needed to guide practitioners to testing best practices (I leave it to your judgement)}
 
@@
 
-For example, a linear correlation between the true and predicted labels is evident in Figure~{fig:r2}.
+For example, a linear correlation between the true and predicted labels is evident in Figure~\ref{fig:r2}.
 Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric.
@@
 However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}).
-We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targetted towards the ML and DS community.
+We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targeted towards the ML and DS community.
 
@@
 Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
-Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an enhaustive list of keywords that appear in several Python testing libraries.
+Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in several Python testing libraries.
 
@@
 
-This study study revealed the emerging practice of crafting analytical assertions based on visual insights to bridge the gap between qualitative and quantitative aspects of ML testing.
+This study empirically supports the efficacy of visualisations for testing various properties of machine learning systems.
+Visualisations enhance the ability of ML practitioners to interpret complex information through visual cues.
+However, decisions based on such visualisations demand manual verification in each development iteration.
+Analytical assertions derived from ML visualisations can reduce this manual effort by encapsulating key insights, and ensuring continuity in interpretation across different development stages and project collaborators.
 
-We identified 269 visualisation-assertion pairs from 1.7K Jupyter notebooks, that were semantically related to one another.
-And identified three recurring testing patterns within these pairs.
-Nearly 40\% of the assertions used ``magic'' thresholds, 75\% of the assertions used comparison operators, and 16\% of the assertions used external testing libraries.
-
-This study serves as a first step in comprehending the landscape of ML testing in Notebooks, emphasizing the potential for the development of more effective tools and practices to streamline the validation of visualisations and elevate the quality of analytical assertions.
+Our results indicate that crafting analytical assertions from ML visualisations, is an emerging testing technique.
+Majority of the assertions however, are found to be primitive and unable to convey all information discerned from their visual counter-parts.
+This study finds a gap in the current software testing methods to address the unique challenges of ML.
+And emphasises the urgent need for automated tools to bridge visual assessments and analytical assertions.
 

### 2023-10-25 report/icst24.tex
feat(report): 2023-10-25 21:53 updates

@@
 We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
-We mine Github to collect $54,070$ Jupyter Notebooks that contain assertions written in Python.
+We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python.
 We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
@@
 We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns.
-We perform an in-depth analysis of 34 visualisation-assertion pairs by reading and understanding the corresponding notebooks and find that the assertions often fail to capture all the information present in their visual counter-part.
-Our results indicate the need for automated tools that can reduce the manual effort required to verify visualisations used to test ML systems.
-And help ML practitioners formulate better analytical assertions.
+We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part.
+Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML.
+And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.
 
@@
 In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
-During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
+During the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ML models.
 Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
@@
 However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
+
 Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
@@
 
-In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems in practice.
-Specifically, we focus on how ML practitioners use visualisations to test specific properties of ML systems.
-We further investigate how frequently ML practitioners formulate analytical tests based on the information gained from the visualisations.
+In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems.
+Specifically, we focus on how visualisations are used to test specific properties of ML systems.
+We further investigate how frequently analytical tests are formulated based on the information gained from the visualisations.
 
@@
 
-\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, source code and metadata from over 900 releases of 12 important data science libraries.
-Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends in technologies they should focus on.
-\cite{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
+\emph{Psallidas et al} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
+Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
+\emph{Pimentel et al} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
 Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \cite{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+To enable future research on computational notebooks, \emph{Quaranta et al} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
 The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
@@
 
-To manage and prune the multiple versions of code that accumulate over time when developing in notebooks, \cite{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
 Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
@@
 Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\cite{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\cite{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
+\emph{Yuan et al} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\emph{Hohman et al} conduct a survey of visual analytics techniques for Deep Learning Models.
 The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
@@
 Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\cite{wexler2020if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
-To reduce the cognitive load of ML practitioners during model building phase, \cite{amershi2015modeltracker} propose \textit{ModelTracker}.
+\emph{Wexler et al} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
+To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al} propose \textit{ModelTracker}.
 Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
@@
 Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
-The \texttt{metadata}, \texttt{nbformat} and \texttt{nbformat\_minor} keys contain additional information regarding the notebook such as information regarding the kernel and the programming language used in the notebook.
+The \texttt{metadata}, \texttt{nbformat} and \texttt{nbformat\_minor} keys contain additional notebook metadata such as information regarding the kernel and the programming language used in the notebook.
 The \texttt{cells} key consists of a list of all cells within the notebook.
@@
 
-Figure~\ref{fig:method} summarises the methodology used to collect related VA pairs from Jupyter notebooks.
+Figure~\ref{fig:method} summarises the methodology used to collect related visualisation-assertion (VA) pairs from Jupyter notebooks.
 The yellow boxes represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
@@
 
-To further reduce the sample size, we exclude notebooks that do not use popular ML or Data Science (DS) libraries.
-We perform a string search to identify notebooks that import at least one of the libraries.
+To further reduce the sample size, notebooks that do not use popular ML or Data Science (DS) libraries are excluded.
+A string search is used to identify notebooks that import at least one of the libraries.
 The string patterns are derived from the python module name used in the import statements.
@@
 A markdown cell present between the assertion and visualisation code cells, would often mark the beginning of a new section in the notebook.
-In other cases, the assertions were typically checks to ensure that the visualisation code would not encounter any errors.
-This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, in the same code cell.
+In other cases, the assertions were checks to ensure that the visualisation code would not encounter any errors.
+This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, but within the same code cell.
 Common examples of such checks include checking the shape of features that are used in the visualisation.
@@
 We observed the best results in notebooks with Arrangement 3.2---where the assertion cell was not longer than 10 lines and was defined below the visualisation cell.
-We observed that most notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
+We observed that some notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
 The visualisations were often followed by another markdown cell to record the author's observations.
@@
 We encountered several instances where the tracing was not straightforward due to lack of sufficient context from surrounding markdown cells or poor coding practices---such as use of obscure variable names or deeply nested function calls.
-In such cases, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation.
+Here, we used one-shot prompt engineering~\cite{liu2023pretrain} with the ChatGPT 4.0 model to understand the code used to create the visualisation.
 Listing~\ref{lst:prompt} shows the prompt used for such cases.
-In a few cases, where the explanation of the visualisation code was not sufficient, we used the ChatGPT model to confirm our annotation of the VA pair.
+For VA pairs, where the explanation of the visualisation code was not sufficient, we used the ChatGPT model to confirm if they were related to each other.
 Listing~\ref{lst:prompt-follow-up} shows the prompt used in such cases.
+A sample of positive annotations from the ChatGPT model were additionally verified by the authors of this paper.
+No errors were found and the annotations were included in the study.
 
@@
 
-Notebooks not related to ML or Data Science (DS) were excluded from this study.
+Table~\ref{tab:exclusion-criteria} lists all exclusion criteria along with the number of notebooks that were excluded from this study.
+Notebooks not related to ML or DS were excluded from this study.
 We found several notebooks from other domains such as cryptography, scientific simulations, genetic algorithms and fluid dynamics.
-Such notebooks were captured in Phase 1 because they imported one of the popular python libraries (typically numpy or pandas).
+Such notebooks were captured in Phase 1 because they imported one of the popular python libraries.
 However, the visualisation and assertion pairs discovered within these notebooks were not related to ML or DS.
@@
 Finally, notebooks that were not authored entirely in English or contained faulty code that did not produce a visualisation were excluded.
-Table~\ref{tab:exclusion-criteria} lists all exclusion criteria along with the number of notebooks that were excluded from this study.
 
@@
 Our observations indicate that AI practitioners formulate analytical tests based on the information gained from the visualisations that they create to test ML systems.
-However, this is an emerging testing practice since only 15\% of the notebooks analysed in this study exhibit this testing practice.
+However, this is an emerging testing practice since only 15\% of the notebooks analysed in this study exhibit this form of testing.
 
@@
 
-An overwhelming majority of assertions compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author\luis{we need to absolute values here (out of 269)}.
+101 of 269 assertions (37\%) compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
 This covers 46\% of the assertions with a comparison operator.
@@
 The threshold could have been derived based on domain expertise.
-Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirementsi,rahimi2019toward}.
+Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirements,rahimi2019toward}.
 
@@
 
-\subsection{RQ3: Does the assertion capture all the information present in the visualisation?}\label{sec:result-rq3}
-
+\subsection{Do the assertions capture all the information obtained from the corresponding visualisation?}\label{sec:result-rq3}
 We performed an in-depth analysis of 34 unique VA pairs.
@@
 Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
-The visualisation also allows ML practitioners to compare different ML models or variants of the same ML model to iteratively find a model that best fits the underlying training data.
+The visualisation also allows comparison of different ML models or variants, to iteratively find a model and hyper-parameters that best fits the underlying training data.
 In this case for instance, the author of the notebook first trained the SVM with a linear kernel.
@@
 For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
-This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.
+This indicates that the model was not able to fit the training data that well, and can be fine-tuned.
 
@@
 
-\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
+\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
 for feature in range(data_transformed.shape[1]):
@@
 We found only $54,070$ notebooks that contain the keyword ``assert'', while there are more than a million Jupyter notebooks available on GitHub.
-This suggests that the practice of assertion is still relatively uncommon.
+This suggests that writing assertions in computational notebooks is still relatively uncommon.
 We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
 We argue that guidelines on ML coding practices require improvement.
-And the need to steer education and research of Software Engineering practices towards the specific challenges faced  when contributing to ML software systems.
+And advocate the need to steer education and research of Software Engineering practices towards the specific challenges faced when contributing to ML software systems.
 
 In the 1764 notebooks analysed in this study, several visualisations lack assertions.
-However, implicit expectations still exist within each visualisation which cannot be readily validated by other participants.
+However, implicit expectations still exist within each visualisation which cannot be readily validated by other project collaborators.
 ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts.
-This highlights the need for refining the education for these professionals and crafting tools to compensate for potential gaps in their knowledge.
+This further highlights the need for refining the education for these professionals and crafting tools to compensate for gaps in their knowledge.
 
-Despite the efforts from some practitioners to automatically validate patterns from plots, results from RQ2 show that the classic software engineering testing practices are failing to adapt to the context of testing ML code.
-We hypothesise that this problem stems from two fundamental sources.
+Results from RQ2 reveal two challenges in the current landscape of ML testing.
 The assertions in the VA pairs identified in this study lack dedicated testing libraries.
@@
 
-Results from RQ2 show the prevalent use of ``magic'' thresholds in assertions, which can impede the maintainability of ML projects.
-Even if these thresholds arise from domain expertise or experience obtained over time, it is crucial they are thoroughly documented.
-Our findings indicate a clear need to assist practitioners in documenting these decisions.
-Potential solutions include automated tools that highlight undocumented choices or suggest standardized documentation in Markdown or code annotations.
+The widespread use of ``magic'' thresholds---often rooted in domain expertise or experience obtained over time---poses risks to maintainability of ML projects.
+We advocate that such thresholds are documented for future reference.
+And find a clear need for automated tools that highlight undocumented choices or suggest standardized documentation in Markdown or code annotations.
 
 In the results from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-We argue that this can lead to a mismatch of expectations in future versions of the ML software.
+This can lead to a mismatch of expectations in future versions of the ML software.
 For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
 The examples presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
-Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of a ML developer.
+Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
 
@@
 The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
-To the best of our knowledge,~\cite{pimentel2019large} and~\cite{quaranta2021kgtorrent} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
-We faced several technical challenges while trying to access the notebooks provided by these authors and thus decided to mine our own data from Github.
-However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS (Section~\ref{sec:identify-related-pairs} and Table~\ref{tab:exclusion-criteria}).
-We wish to extend our analysis to notebooks collected from Kaggle since Kaggle is predominantly targeted towards the ML and DS community.
+To the best of our knowledge,~\emph{Pimentel et al} and~\emph{Quaranta et al} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
+We faced several technical challenges while trying to access the notebooks provided by these replication packages and thus decided to mine our own data from Github.
+However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS.
+We wish to extend our analysis to notebooks collected from Kaggle since it is predominantly targeted towards the ML and DS community.
 
-In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python libraries that provide a testing interface (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
-Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in several Python testing libraries.
+In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate file type and the presence of the ``assert'' keyword.
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python testing libraries (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
+Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in python testing libraries.
 
-This study finds a clear need to improve the current ML testing process.
-There is a clear need for tools specifically designed to aid ML practitioners when testing ML systems.
+There is a clear need for tools designed to aid ML practitioners when testing ML systems.
 These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems.
@@
 Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
-By automating the generation of assertions from visual information, these tools will promote broader adoption of assertions in ML notebooks, reducing the burden of manual effort and bringing the code developed within the notebooks, closer to a production-ready codebase.
+Generating assertions from visual information automatically, will promote adoption of assertions in ML notebooks.
+And bring the code developed within notebooks, closer to a production-ready codebase.
 
-Extending ML testing to dynamic real-world scenarios is a critical avenue for future exploration.
+% TODO remove this is no space left
+Extending ML testing to dynamic real-world scenarios should be explored further.
 This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
-Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycles.
+Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycle.
 
@@
 This study finds a gap in the current software testing methods to address the unique challenges of ML.
-And emphasises the urgent need for automated tools to bridge visual assessments and analytical assertions.
+And emphasises the need for automated tools to bridge visual assessments and analytical assertions.
 

### 2024-01-31 report/bsc-research-project-proposal.qmd
refactor: migrate shome2023llm files
It does not make sense to keep shome2023llm as a separate project
since its a sibling of shome2023notebook.

This commit moves all relevant files of shome2023llm into this repo.

---
title: "Bsc. Research Project Proposal"
author: "Arumoy Shome"
format:
  pdf:
    mainfont: "Iosevka Aile"
    monofont: "Iosevka"
bibliography: bibliography.bib
csl: acm-sig.csl
---

# Introduction

Visualisations are employed at various stages of a Machine Learning
(ML) pipeline. They are used to understand and verify data properties
during the early stages, summarise metrics and fine-tune models during
development, and monitor performance post-deployment. The iterative
and experimental nature of building ML systems heavily relies on
insights from visualisations to guide design and implementation
decisions
[@yuan2020survey;@hohman2019visual;@amershi2015modeltracker;@wexler2019what-if;@shome2022data;@haakman2021ai].

However, real-world data that ML systems encounter post-deployment
seldom remain static. They often change as a reflection of the world,
potentially violating initial assumptions made during development
[@amershi2019software;@lwakatare2021on]. Every subsequent iteration of
the ML development cycle used to retrain and update the ML model,
therefore demands manual validation of the visualisations that were
used to test ML system properties.

Assertions or analytical tests derived from ML visualisations can
significantly reduce manual verification efforts. Such formal
assertions record the AI practitioner's observations about the model
or data at a specific moment. They also serve as a reference point for
future AI practitioners to understand the interpretations made from
earlier visualisations.

In a prior study, we mined $54,070$ Jupyter notebooks from Github and
created a high-quality dataset of $269$ semantically related
visualisation-assertion (VA) pairs. The input feature space comprises
of a rich source of information comprising of visualisations, Python
source code, and associated markdown text.

This research project will focus on developing automated tools using
state-of-the-art Large Language Models (LLMs) that aid ML
practitioners when working with visualisations. The research questions
for this project are categorised into 3 distinct tracks and presented
below.

# Track 1: Exploring Multi-Model Input Feature Space for Translating ML Visualisation to Python Assertions

This track will focus on exploring various multi-model input feature
space for the LLMs. Students will work with various open-source LLMs
from the HuggingFace library and develop an optimal training set that
produces the best results. There are 3 research questions here:

+ **RQ1**: Can image-to-text models translate visual properties
  observed in ML visualisation to Python assertions?

  The objective here is to experiment with various image-to-text
  models with a dataset comprising of PNG images containing
  visualisations obtained from Jupyter Notebooks.

+ **RQ2**: Does a multi-modal dataset improve the results of the AI
  models?

  Here we will augment the dataset with additional features extracted
  from the Python code that produces the visualisation and the
  surrounding Markdown text written by the notebook authors.

+ **RQ3**: Do LLMs4Code models perform better compared to
  general-purpose LLMs?

  There are several LLMs that have been trained on a large corpus of
  code written in programming languages. The objective of RQ3 is to
  compare such models with general purpose LLMs and determine if they
  are better suited for the task.

# Track 2: Detecting "Smelly" Python Visualisation Code

This track will focus on creating an automated tool that can detect
code smells in the Python code that generate ML visualisations.

+ **RQ4**: What are the most frequently occuring code smells in ML
  visualisation code written in Python?

  The focus here will be to propose code smells observed in Python
  code that generate visualisations. This track is more open-ended as
  students can propose code smells from different aspects such as
  design of the visualisation and performance of the underlying code.

# Track 3: Automatically Detecting "Related" VA Pairs in Jupyter Notebooks

Creating a dataset of semantically related VA pairs requires a lot of
manual effort. This track will focus on exploring strategies to
automate this process.

+ **RQ5**: Can code similarity metrics automatically detect related
  VA pairs?

  In this RQ the focus will be on experimenting with various code
  similarity metrics [@eghbali2022crystalbleu] which can automatically
  detect if a given VA pair are related to each other.

# References

::: {#refs}
:::

### 2024-01-31 report/icse24-nlbse.tex
refactor: migrate shome2023llm files
It does not make sense to keep shome2023llm as a separate project
since its a sibling of shome2023notebook.

This commit moves all relevant files of shome2023llm into this repo.

%% This is file `sample-sigconf.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigconf.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigconf]{acmart}

\usepackage[most]{tcolorbox}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
    language=Python,
    basicstyle=\ttfamily,
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    commentstyle=\color{green},
    morecomment=[l][\color{magenta}]{\#}
}

%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2024}
\acmYear{2024}
\acmDOI{XXXXXXX.XXXXXXX}

%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[ICSE 2024]{46th International Conference on Software Engineering}{April 2024}{Lisbon, Portugal}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
% \acmPrice{15.00}
% \acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Towards Automatic Translation of Machine Learning Visual
Insights to Analytical Assertions}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.

\author{Arumoy Shome}
\email{a.shome@tudelft.nl}
\affiliation{%
  \institution{Delft University of Technology}
  \country{Netherlands}
}

\author{Lu{\`\i}s Cruz}
\email{l.cruz@tudelft.nl}
\affiliation{%
  \institution{Delft University of Technology}
  \country{Netherlands}}

\author{Arie van Deursen}
\email{arie.vandeursen@tudelft.nl}
\affiliation{%
  \institution{Delft University of Technology}
  \country{Netherlands}
}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

We present our vision for developing an automated tool capable of
translating visual properties observed in Machine Learning (ML)
visualisations into Python assertions. The tool aims to streamline the
process of manually verifying these visualisations in the ML
development cycle, which is critical as real-world data and
assumptions often change post-deployment. In a prior study, we mined
$54,070$ Jupyter notebooks from Github and created a catalogue of
$269$ semantically related visualisation-assertion (VA) pairs.
Building on this catalogue, we propose to build a taxonomy that
organises the VA pairs based on ML verification tasks. The input
feature space comprises of a rich source of information mined from the
Jupyter notebooks---visualisations, Python source code, and associated
markdown text. The effectiveness of various AI models, including
traditional NLP4Code models and modern Large Language Models, will be
compared using established machine translation metrics and evaluated
through a qualitative study with human participants. The paper also
plans to address the challenge of extending the existing VA pair
dataset with additional pairs from Kaggle and to compare the tool's
effectiveness with commercial generative AI models like ChatGPT. This
research not only contributes to the field of ML system validation but
also explores novel ways to leverage AI for automating and enhancing
software engineering practices in ML.

\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{SE4AI, NLP4Code, ML Testing, Visualisations, Assertions,
Computational Notebooks, Automated Tool}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

Visualisations are employed at various stages of a Machine Learning
(ML) pipeline. They are used to understand and verify data properties
during the early stages, summarise metrics and fine-tune models during
development, and monitor performance post-deployment. The iterative
and experimental nature of building ML systems heavily relies on
insights from visualisations to guide design and implementation
decisions~\cite{yuan2020survey,hohman2019visual,amershi2015modeltracker,wexler2019what-if,shome2022data,haakman2021ai}.

However, real-world data that ML systems encounter post- deployment
seldom remain static. They often change as a reflection of the world,
potentially violating initial assumptions made during
development~\cite{amershi2019software,lwakatare2021on}. Every
subsequent iteration of the ML development cycle used to retrain and
update the ML model, therefore demands manual validation of the
visualisations that were used to test ML system properties.

Assertions or analytical tests derived from ML visualisations can
significantly reduce manual verification efforts. Such formal
assertions record the AI practitioner's observations about the model
or data at a specific moment. They also serve as a reference point for
future AI practitioners to understand the interpretations made from
earlier visualisations.

In a prior study currently under review, we investigate how
frequently analytical tests are formulated from visualisations in ML
systems and analysed the effectiveness of these tests. We mined
$54,070$ Jupyter notebooks from GitHub that contain assertions written
in Python and developed a novel methodology to identify $1,764$
notebooks which contain an assertion below a visualisation. We
manually analysed the $1,764$ notebooks and catalogued $269$
semantically related visualisation-assertion (VA) pairs (to be
released publicly with the camera-ready version of the paper). Further
in-depth analysis of the VA pairs revealed that assertions often fail
to capture all the information obtained from the corresponding
visualisations. Our results indicate that formulating analytical
assertions from visualisations is an emerging testing technique in ML.
However, it also highlights the limitations of current practices,
demonstrating a need for automated tools that can assist ML
practitioners in validating visualisations and formulating more
comprehensive analytical assertions.

In this paper, we propose our research vision to develop an automated
tool to generate analytical assertions from ML visualisations. The
research questions we plan to address along with the contributions we
envision to make are presented below.

\begin{description}
  \item[RQ1.] \textbf{How are VA pairs used to perform ML verification
    tasks?}

    We begin by creating a taxonomy that clusters the existing $269$
    VA pairs by specific ML verification tasks. This taxonomy not only
    enables us to fine-tune the data and model for the automated tool,
    but also serves as valuable reference for future ML practitioners.

  \item[RQ2.] \textbf{What kind of input features enable AI models to
    generate assertions from visualisations?}

    The dataset consists of visualisations, source code and natural
    text. Based on the manual analysis we previously conducted to
    identify semantically related VA pairs, we hypothesis that the
    source code for the visualisation is sufficient for ML models to
    generate meaningful assertions. We plan to compare the results to
    multi-modal input feature space where a combination of image, code
    and text are used.

  \item[RQ3.] \textbf{What kind of AI models generate the best
    assertions from visualisations?}

    A empirical study will be conducted to compare traditional
    NLP4Code models with modern Large Language Models. The
    effectiveness of these models will be evaluated both using
    a holdout set and additionally using a qualitative study with
    human participants.

  \item[RQ3.] \textbf{How does our tool compare to commercial
    Generative AI models?}

    This paper presents results from a preliminary study which we
    conducted to evaluate the effectiveness of the ChatGPT 4.0 model
    to automatically generate assertions from ML visualisations. We
    plan to extend the preliminary study to the full catalogue on VA
    pairs and compare the results to our proposed tool.

\end{description}

\section{Our Vision}

Figure~\ref{fig:vision} provides a visual summary of the research
vision proposed in this paper. During the data collection phase, an
extended catalogue of VA pairs will be created using VA pairs mined
from Github and Kaggle. A taxonomy of VA pairs used for ML
verification tasks will be created and used to fine-tune the dataset
and model. Quantitative studies will determine the best input feature
space and AI models that are able to generate meaningful assertions
from visualisations. Quantitative and qualitative evaluations will be
employed to position the proposed tool with-respect-to ChatGPT.

\begin{figure*} \includegraphics[width=\textwidth]{vision.pdf}
\caption{Vision for the automated tool proposed in this paper to
generate analytical assertions from ML visualisations.}
\label{fig:vision} \end{figure*}

\subsection{RQ1: How are VA pairs used to perform ML verification tasks?}

The existing catalogue of $269$ VA pairs was created by mining public
Jupyter notebooks on Github, that contained both visualisations and
assertions. We searched the entirely of Github and obtained $54,070$
Jupyter Notebooks written in Python, containing the keyword ``assert''
in them. The sample was further refined to $1,764$ notebooks that
imported popular Machine Learning or Data Science libraries and had
specific cell arrangements that indicated a potential relationship
between visualisations and assertions. We manually reviewed these
notebooks to identify $269$ VA pairs that were semantically related to
each other. This analysis involved manually examining the source code
of the visualisation and assertion cells and excluding unrelated
pairs.

The $269$ VA pairs however lack structure. Therefore, We plan to
organise them using a taxonomy that categories VA pairs according to
specific ML verification tasks. This taxonomy is critical for the
creation of an automated tool, as it allows for more targeted
fine-tuning of the training data and model. We hope that by aligning
the model with distinct verification tasks identified in the taxonomy,
the tool can generate more accurate and relevant assertions.

The taxonomy also organises the domain knowledge of ML visualisations
and assertions. It can be used by future ML practitioners as reference
to identify the types of visualisation that can be used for a given
verification task, or formulate comprehensive assertions given a visualisation.

\subsection{RQ2: What kind of input features enables AI models to
generate assertions from visualisations?}

Using the taxonomy created in the prior phase, we plan to create
a high quality dataset of VA pairs which can be used to train and test
AI models that automatically generate the assertions. The input
feature space comprises of three rich sources of information---the
visualisations stored as PNG images, Python source code used to create
the visualisation and when available, natural text written by the
notebook authors above and below the visualisation.

In our prior study, we analysed $1,764$ notebooks manually to identify
$269$ VA pairs. We reflected on our manual analysis procedure and
realised that the code for the visualisation and the assertion provide
sufficient context to determine whether they are related to each other
or not. Our hypothesis is that the Python source code along with the
Markdown text should provide sufficient information for the ML model
to generate meaningful assertions. However, it will also be
interesting to compare the results with multi-modal AI models which
also use the visualisations as input.

\subsection{RQ3: What kind of AI models generate the best assertions
from visualisations?}

We plan to conduct an extensive empirical study to determine what type
of AI models are able to produce the best assertions. Here, we plan to
compare and contrast traditional NLP4Code models with open-source
Large Language Models (LLMs).

From an NLP perspective, creating assertions given the source code for
a visualisation is a translation task. Thus, the effectiveness of the
above models can be quantified using existing translation metrics such
as \textit{BLEU}, \textit{Meteor} and
\textit{Rouge-L}~\cite{zhu2023revisiting}.

We further plan to evaluate the practical relevance and impact of our
proposed tool by conducting a qualitative study with human
participants. We are currently exploring two approaches here. First,
with human participants that compare the ground truth assertions with
the ones generated by our tool, and rate them on a Likert Scale on
Naturalness and Correctness~\cite{izadi2022on}. An alternative
approach would be to collect feedback from the notebook authors by
opening Pull Requests on Github that propose the generated assertion.

\subsection{RQ4: How does our solution compare to commercial
generative AI models?}

The ``Elephant in the room'' must be addressed. OpenAI's suite of
commercial generative AI models has produced a tectonic shift in the
Software Engineering landscape. We plan to evaluate the effectiveness
of our proposed tool against commercial AI models such as ChatGPT and
DALL-E.

In a preliminary study, we assessed the capability of ChatGPT 4.0 to
generate assertions from visualisations, utilising six VA pairs that
were highlighted in our prior study. The study employed one-shot
prompting with three different inputs---providing ChatGPT with only
the visualisation image, only the code generating the visualisation,
and both the image and code. The responses were evaluated based on
three criteria---whether the generated assertions reflected properties
or trends evident in the visualisations, the similarity of the
response to the ground truth assertions, and whether ChatGPT proposed
alternative assertions that validated the same visual properties as
the ground truth.

All 18 responses from ChatGPT contained assertions derived from visual
properties, notably even when prompted solely with the code. Out of
these, 8 responses closely matched the ground truth assertions, and in
the remaining 10, ChatGPT suggested valid alternative assertions
aligned with the same visual properties as the ground truth. We plan
to extend this study to the entire catalogue of VA pairs which will
allow us to compare and contrast our proposed tool to the abilities of
ChatGPT.

\section{Challenges}

\subsection{Dataset Limitations and Bias}

The quality and diversity of the dataset used to train the tool will
heavily influence its effectiveness. There is a risk of bias if the
datasets are not representative of real-world scenarios or if they
lack diversity in terms of the types of visualisations and assertions.

In RQ1, the existing $269$ VA pairs were obtained from Github which
might not be representative of all VA pairs observed in ML. To
mitigate this threat, we plan to extend the catalogue with VA
pairs mined from Kaggle. This also allows us to simplify our data
mining process by eliminating the need to filter notebooks for
relevant libraries since Kaggle is primarily targeted towards
the Data Science and Machine Learning communities.

In RQ2, we expect to face challenges with respect to the quantity and
quality of the training data. Since translating visual properties to
analytical assertions is an emerging testing practise, we may not have
sufficient training data for our AI models. The extension using VA
pairs from Kaggle should alleviate this threat to a certain extent.
Another alternative would be to fine-tune the NLP4Code models using
our catalogue. With LLMs, we can experiment with different prompting
techniques such as few-shot prompting~\cite{liu2023pretrain}.

% TODO come back to this one
With respect to quality, we want to reinforce the manual analysis
required to determine whether a given visualisation and assertion are
related, with multiple annotators. The quality of the annotations can
be determined by calculating the inter-rater agreement using the
\textit{Kappa} metric.

In RQ4, we do not know the contents of the training corpus used to
train ChatGPT. However given its colossal size, it is safe to assume
that our catalogue of VA pairs exists in its training data (or will be
included in future versions). This supports our decision to use
open-source LLMs where we can validate that no such data leakage
occurs. We can also extend the proposed evaluation for our tool (RQ3)
and ChatGPT (RQ4) using VA pairs that have not been publicly released.
One potential solution here would be to collect VA pairs created by
students in a Software Engineering for ML course.

\subsection{Complexity of Visual Analytics}

Translating visual properties accurately into analytical assertions is
a complex task. The tool must be capable of understanding and
interpreting a wide range of visual patterns and nuances, which is
a significant challenge. In our prior paper, we identified several
instances where the analytical assertions were not able to capture all
the information present in their visual counter-part. While our
proposed tool may not be able to capture all the nuances of
a visualisation every time, identifying this boundary of automation
and human intervention remains an important and exciting research
direction.

\subsection{Integration into Existing Workflows}

For the tool to be widely adopted, it needs to be seamlessly
integrated into existing ML development workflows. This includes
compatibility with various programming environments and ML frameworks.
While the outcome of our research endeavours may be a simple
prototype, we hope to lay the foundations that enable future
researchers to integrate our tool into existing computational notebook
environments such as Jupyter Lab or Visual Studio Code. Such an
integration may enable the tool to provide better assertions using
additional context either obtained directly from the user through
a chat interface~\footnote{Such as the one provided by Github Co-Pilot
extension for Visual Studio Code:
https://code.visualstudio.com/blogs/2023/03/30/vscode-copilot} or
derived from the entire software codebase~\footnote{One example of
such an editor extension is Cody which is available for Visual Studio
Code and JetBrains: https://github.com/sourcegraph/cody}.

\section{Expected Outcomes}

The proposed tool automates the translation of visual properties from ML
visualisations into Python assertions. This automation reduces the
manual effort required in the model validation process, making it
faster and more efficient. By automating a critical part of the ML
development process, the tool can streamline the entire cycle from
data preprocessing to model deployment, enabling quicker iterations
and enhancements. Automated assertions make it easier to handle large
datasets and complex models, as it removes the bottleneck of manual
analysis, thereby enhancing scalability.

Manual processes are prone to human errors, such as overlooking subtle
visual cues in data. In contrast, an automated tool can consistently
capture and translate these nuances, reducing the likelihood of
errors. This form of automation leads to a more standardized approach
to validating ML models, ensuring that the same criteria are applied
consistently across different models and datasets. Furthermore,
automated assertions provide a reliable record of the state of the
data and model at various points in time, aiding in historical
analysis and model auditing.

An automated tool also enables ML practitioners to adapt assertions as
real-world data evolves. By providing immediate feedback on visual
data, the tool will enable faster updates and refinements to ML models
in response to changing data trends. The tool could potentially
identify anomalies or shifts in data patterns sooner than manual
processes. This not only allows for proactive adjustments to models,
but also significantly reduces effort required to maintain the
relevance and accuracy of ML models over time.

\section{Expected Outcomes Beyond ML Testing}

As ML augments software systems in safety-critical domains, there has
been a significant emphasis on the explainability of ML models. The
complexity and multi-dimensionality of such models and their
underlying data has lead to the adoption of Visual Analytics by the ML
community. However, interpretation of a visualisation may vary between
developers with varying domain expertise and background. Converting
such visual interpretations into assertions can help bring consistency
into the interpretability of ML visualisations across all project
members.

The concept of translating visual properties to formal assertions can
be extended to other domains (such as scientific computing and
simulation), where visual analytics also plays a prominent role in the
decision making process. Future research could explore ways to
customise these tools for specific industries or types of data,
increasing their applicability and effectiveness.

The effectiveness of the tool in real-world scenarios could also open
research into how human practitioners and AI tools can best
collaborate, particularly in the context of complex problem-solving
tasks like ML model validation.

\section{Conclusion}

Developing an automated tool to convert visual properties from ML
visualisations into analytical assertions marks a timely advancement
in the domains of Machine Learning and Software Engineering. This tool
has the potential to improve the efficiency, accuracy, and consistency
of ML model testing and validation. Nevertheless, it also presents
considerable challenges, especially in areas such as visual analysis,
data set quality, and integration with current workflows. The research
initiates various new opportunities, such as enhancing the
interpretability of AI, enabling automated testing, and exploring the
dynamics of human-AI collaboration.

\bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

### 2024-01-31 report/icst24.tex
feat(readme): 2023-10-26 15:09 updates
refactor: migrate shome2023llm files
It does not make sense to keep shome2023llm as a separate project
since its a sibling of shome2023notebook.

This commit moves all relevant files of shome2023llm into this repo.

@@
 
-  Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
-In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
-Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
-However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
-We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
-We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python.
-We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
-We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns.
-We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part.
-Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML.
-And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.
+  Testing ML systems is a highly interactive process which demands a
+  human-in-the-loop approach. In addition to writing tests for the
+  code base, practitioners are required to analyse and interpret
+  several visualisations using their domain expertise to validate if
+  an ML system satisfies the required set of functional and
+  non-functional properties. Visualisations are frequently used to
+  qualitatively assess various parts of an ML pipeline. However,
+  implicit knowledge gained from visualisations must be translated to
+  explicit analytical tests that fail when there is change in any
+  component of the ML pipeline. We conduct an empirical analysis of
+  Jupyter notebooks to catalogue the state-of-the-art mappings between
+  ML visualisations and assertions. We mine Github to collect 54K
+  Jupyter Notebooks that contain assertions written in Python. We
+  develop a novel methodology to identify 1764 notebooks which contain
+  an assertion below a visualisation. We manually analyse the 1.7K
+  notebooks and identify 269 visualisation-assertion pairs that are
+  semantically related to one another. We further investigate the 269
+  visualisation-assertion pairs and identify three frequently
+  occurring testing patterns. We perform an in-depth analysis of 34
+  visualisation-assertion pairs and find that the assertions often
+  fail to capture all the information present in their visual
+  counter-part. Empirical evidence obtained in this study indicates
+  that current software testing methods fail to address the unique
+  challenges of ML. And emphasises the need for automated tools that
+  bridge the gap between visual assessments and analytical assertions.
 
@@
 \begin{IEEEkeywords}
-  ML Testing, Visualisations, Assertions, Computational Notebooks
+  ML Testing, SE4AI, Visualisations, Assertions, Computational Notebooks
 \end{IEEEkeywords}
@@
 
-Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
-In the early stages of the ML development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
-During the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ML models.
-Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
+Visualisations are used throughout the ML development lifecycle to
+test various properties of the ML system. In the early stages of the
+ML development lifecycle, visualisations are extensively used to make
+sense of the data and verify its statistical properties. During the
+model development phase, visualisations are used to summarise metrics,
+contrast different learning algorithms and iteratively fine-tune ML
+models. Once the model is deployed in production, visualisations are
+used to continually monitor their performance and trigger a new
+training cycle once their performance drops below a certain
+threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
 
-Building ML systems is highly iterative and experimental.
-Information gained from ML visualisations is used to make design and implementation decisions for the following steps of the ML pipeline.
-For instance, we may visualise the distribution of our training data and find that it is normally distributed.
-Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
-However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
+Building ML systems is highly iterative and experimental. Information
+gained from ML visualisations is used to make design and
+implementation decisions for the following steps of the ML pipeline.
+For instance, we may visualise the distribution of our training data
+and find that it is normally distributed. Based on this information,
+we may opt for a Linear Regression model which assumes normality in
+the underlying data. However, such expectations regarding the data may
+be violated once the ML system is deployed in production, where the
+data constantly changes as a reflect of the real
+world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
 
-Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
+Implicit expectations obtained from visualisations must therefore be
+translated to analytical tests that fail once our expectations
+regarding the ML system are no longer satisfied.
 
-In contrast to prior work, we approach ML testing from a new perspective.
-We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
-In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions).
-The research questions along with the contributions of this paper are as follows.
+In contrast to prior work, we approach ML testing from a new
+perspective. We conduct an empirical analysis of computational
+notebooks obtained from Github, to understand the process of testing
+ML systems in practice. In particular, we focus on the combination of
+a qualitative form of testing (using visualisations) and a
+quantitative form of testing (using analytical assertions). The
+research questions along with the contributions of this paper are as
+follows.
 
 \begin{description}
-  \item[RQ1.] \textbf{How frequently are analytical tests formulated from visualisations created to test ML systems?}
+  \item[RQ1.] \textbf{How frequently are analytical tests formulated
+    from visualisations created to test ML systems?}
 
-We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
-We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion (VA) pairs that are semantically related to each other.
-We plan to release this catalogue of VA pairs publicly.
+    We mine 54K Jupyter notebooks from Github that contain an assertion
+    written in Python. We develop an automated technique to identify 1764
+    notebooks with an assertion below a visualisation. We manually analyse
+    the 1.7K notebooks and identify 269 visualisation-assertion (VA) pairs
+    that are semantically related to each other. We plan to release this
+    catalogue of VA pairs publicly.
 
-\item[RQ2.] \textbf{What patterns are frequently observed in VA pairs used to test ML systems?}
+\item[RQ2.] \textbf{What patterns are frequently observed in VA pairs
+  used to test ML systems?}
 
-We analyse the 269 VA pairs from RQ1, and observe three testing patterns that frequently occur in the VA pairs.
+  We analyse the 269 VA pairs from RQ1, and observe three testing
+  patterns that frequently occur in the VA pairs.
 
-\item[RQ3.] \textbf{Do the assertions capture all the information obtained from the corresponding visualisation?}
+\item[RQ3.] \textbf{Do the assertions capture all the information
+  obtained from the corresponding visualisation?}
 
-We further perform an in-depth analyse of 34 VA pairs (and their corresponding notebooks). We observe that assertions often cannot capture all the information obtained from the visualisation.
+  We further perform an in-depth analyse of 34 VA pairs (and their
+  corresponding notebooks). We observe that assertions often cannot
+  capture all the information obtained from the visualisation.
 
@@
 
-Our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique.
-However, the assertions found in this study often do not reflect the information obtained from their visual counter-part.
-This study finds the need for automated tools that can reduce the manual effort necessary to validate visualisations in ML systems.
-And help ML practitioners to formulate better analytical assertions from visualisations.
+Our observations indicate that formulating analytical assertions from
+visualisations is an emerging testing technique. However, the
+assertions found in this study often do not reflect the information
+obtained from their visual counter-part. This study finds the need for
+automated tools that can reduce the manual effort necessary to
+validate visualisations in ML systems. And help ML practitioners to
+formulate better analytical assertions from visualisations.
 
-The replication package for this study is made public on Figshare\footnote{The replication package can be found at this URL: https://figshare.com/s/9347228011999cba29f5}.
+The replication package for this study is made public on
+Figshare\footnote{The replication package can be found at this URL:
+https://figshare.com/s/9347228011999cba29f5}.
 
@@
 
-This section summarises prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
-The section concludes with an overview of technical knowledge required for this paper.
+This section summarises prior scientific work that has been done in ML
+Testing, computational notebooks and visual analytics. The section
+concludes with an overview of technical knowledge required for this
+paper.
 
@@
 
-Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems additionally experience change in the training data and the machine learning model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
-With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
-Existing scientific literature on ML testing broadly focuses on two aspects.
-First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
+Testing ML enabled software systems poses more challenges compared to
+their traditional counter part. While traditional software systems
+mature through change in their codebase, ML enabled systems
+additionally experience change in the training data and the machine
+learning
+model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
+With ML being adopted into safety-critical domains that can affect
+human lives, ensuring that ML systems are correct, robust to data
+perturbations and not biased by race, colour or gender is of paramount
+importance. Existing scientific literature on ML testing broadly
+focuses on two aspects. First, on functional properties such as
+correctness and robustness of the model towards unseen data. And
+second, on non-functional properties such as fairness,
+interpretability and
+privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
 
-To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
-Tools such as \textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepexplore, ma2018deepgauge, gerasimou2020importance}.
-Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
-Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
-Several works have been conducted on generating and detecting adversarial inputs for ML models.
-Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
-It is however not possible to include all variations of adversarial examples into the training data.
-Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
+To test the correctness of ML models, several improvements over
+existing test adequacy metrics have been proposed. Tools such as
+\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
+metrics such as neuron coverage adapted for ML enabled software
+systems~\cite{pei2017deepexplore, ma2018deepgauge,
+  gerasimou2020importance}. Formal verification methods have also been
+proposed that try to provide formal guarantee of robustness against
+adversarial examples. Such methods however are only feasible for
+statistical ML models and become computationally expensive for more
+complex models such as deep neural networks~\cite{zhu2021deepmemory,
+  baluta2021scalable}. Several works have been conducted on generating
+and detecting adversarial inputs for ML models. Data augmentation
+techniques based on fuzzing, search based software testing and
+mutation testing have been proposed to generate adversarial examples
+that can be used during model training to improve its
+robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
+  zhang2020white}. It is however not possible to include all
+variations of adversarial examples into the training data. Thus,
+methods have been proposed to detect adversarial inputs during
+runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
+  berend2020cats}.
 
-In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems.
-Specifically, we focus on how visualisations are used to test specific properties of ML systems.
-We further investigate how frequently analytical tests are formulated based on the information gained from the visualisations.
+In contrast to existing scientific contributions, we conduct a
+large-scale empirical analysis of Jupyter Notebooks to understand the
+process of testing ML systems. Specifically, we focus on how
+visualisations are used to test specific properties of ML systems. We
+further investigate how frequently analytical tests are formulated
+based on the information gained from the visualisations.
 
@@
 
-Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.
+Computational notebooks have been ubiquitously adopted by the machine
+learning community for developing ML enabled systems. Although
+originally intended to promote reproducible software, computational
+notebooks are far from being reproducible due to lack of software
+engineering best practices such as separation of concern, testing and
+versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.
 
-\emph{Psallidas et al} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
-Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
-\emph{Pimentel et al} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
-Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \emph{Quaranta et al} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
-The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
+\emph{Psallidas et al} provide an overview of the evolving landscape
+of computational notebooks by analysing six million Python notebooks,
+two million enterprise Data Science (DS) pipelines, source code and
+metadata from over 900 releases of 12 important DS libraries. Their
+findings can be used by system builders for improving DS tools and
+also by DS practitioners to understand the current trends in
+technologies they should focus on. \emph{Pimentel et al} mined 1.4
+million notebooks from Github to conduct an empirical study on the
+coding practices in computational notebooks. Based on their analysis,
+the authors propose guidelines on improving reproducibility of
+computational notebooks. To enable future research on computational
+notebooks, \emph{Quaranta et al} mine Kaggle and present
+\textit{KGTorrent}, a public dataset consisting of approximately 2.5
+million Jupyter notebooks written in Python. The dataset also contains
+a relational database dump of metadata regarding publicly available
+notebooks on Kaggle.
 
-Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
-Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
-The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
+Studies with human subjects have been conducted to gain a deeper
+understanding of the challenges faced by ML practitioners when
+developing ML models inside notebooks. The results indicate that ML
+practitioners work in a highly iterative fashion, often experimenting
+with multiple strategies to analyse the data or produce meaningful
+visualisations~\cite{kandel2012enterprise, kery2018story,
+  liu2019understanding, chattopadhyay2020wrong}. Studies have also
+been conducted to understand how practitioners generate, evaluate and
+manage alternative hypothesis, visual designs, methods, tools,
+algorithms and data sources to arrive at the final
+implementation~\cite{liu2019understanding,kandel2012enterprise}. The
+findings from these studies can be used as guidelines for improving
+existing notebook technologies or designing new ones.
 
-To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
-Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
+To manage and prune multiple versions of code that accumulate over
+time when developing in notebooks, \emph{Head et al} propose a code
+gathering tool that allows practitioners to review and only keep the
+relevant version of code. Other tools such as \textit{WrangleDoc} and
+\textit{VizSmith} have also been proposed to aid ML practitioners when
+working within computational notebooks~\cite{yang2021subtle,
+  bavishi2021vizsmith}.
 
-Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
-Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.
-Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
-This study leverages the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
+Jupyter notebooks have been widely adopted by the DS and ML
+communities to develop ML
+pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Computational notebooks provide a rich source of data in the form of
+natural text, code and visualisations. Computational notebooks also
+allow ML practitioners to present the knowledge gained from the
+analysis in a narrative that can be shared with
+others~\cite{rule2018exploration}. This study leverages the
+computational narrative present within notebooks to identify
+visualisations and assertions that are semantically linked to each
+other.
 
@@
 
-As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
-ML models and the underlying data is complex and multi-dimensional.
-To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.
+As ML augment software systems in safety-critical domains, emphasis
+has been put into explainability of ML models. ML models and the
+underlying data is complex and multi-dimensional. To combat the
+``curse of dimensionality'', visual analytics has been widely adopted
+by the ML community to understand the data and the internal workings
+of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.
 
-Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\emph{Yuan et al} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\emph{Hohman et al} conduct a survey of visual analytics techniques for Deep Learning Models.
-The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
+Prior studies have been conducted to understand how visual analytics
+techniques are currently being used in ML. \emph{Yuan et al} conduct a
+systematic review of 259 papers and propose a taxonomy of visual
+analytics techniques for ML. \emph{Hohman et al} conduct a survey of
+visual analytics techniques for Deep Learning Models. The findings of
+the study indicate that visual analytics has been widely adopted in ML
+for model explanation, interpretation, debugging and improvement.
 
-Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\emph{Wexler et al} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
-To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al} propose \textit{ModelTracker}.
-Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
+Several tools have been proposed by the visual analytics research
+community to aid practitioners in understanding how their ML models
+operate. \emph{Wexler et al} propose \textit{What-If}, a visual
+analytics tool to explore alternative hypotheses, generate
+counterfactual reasoning, investigate decision boundaries of the model
+and how change in data affects the model predictions. To reduce the
+cognitive load of ML practitioners during model building phase,
+\emph{Amershi et al} propose \textit{ModelTracker}. Given a trained
+model and a sample dataset, ModelTracker presents all the information
+in traditional summary statistics along with the performance of the
+model.
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
+\textit{Drava} are a few other tools that have been proposed to handle
+specialised use-cases such as identifying systematic errors in ML
+models, generating counterfactual explanations for Generalised
+Additive Models, prioritising machine translation model improvements
+and relating human concepts with semantic dimensions extracted by ML
+models during disentangled representation learning\cite{ahn2023escape,
+  wang2023gam, robertson2023angler, wang2023drava}.
 
-In contrast to prior work which propose dedicated visual analytics tools, this study focuses on visualisations within computational notebooks.
-We focus on visualisations produced by AI practitioners using Python libraries (such as Matplotlib or Seaborn).
-Additional context provided by surrounding code and markdown cells are used to understand the motivation for creating the visualisation and the information gained from it.
+In contrast to prior work which propose dedicated visual analytics
+tools, this study focuses on visualisations within computational
+notebooks. We focus on visualisations produced by AI practitioners
+using Python libraries (such as Matplotlib or Seaborn). Additional
+context provided by surrounding code and markdown cells are used to
+understand the motivation for creating the visualisation and the
+information gained from it.
 
@@
 
-Figure~\ref{fig:method} summarises the methodology used to collect related visualisation-assertion (VA) pairs from Jupyter notebooks.
-The yellow boxes represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
-The red boxes represent the exclusion criteria used to exclude VA pairs not relevant for this study.
-Text in double quotes represent the string pattern used in the search query.
+Figure~\ref{fig:method} summarises the methodology used to collect
+related visualisation-assertion (VA) pairs from Jupyter notebooks. The
+yellow boxes represent filters that were applied to the initial corpus
+of 54K notebooks, to obtain the final sample size of 1.7K notebooks
+used in the empirical study. The red boxes represent the exclusion
+criteria used to exclude VA pairs not relevant for this study. Text in
+double quotes represent the string pattern used in the search query.
 Each phase of the methodology is explained in more detail below.
@@
 
-We mined public repositories from Github to collect $54,070$ Jupyter notebooks that contain the keyword ``assert'' in them.
-We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook.
-Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
-The final search query is as follows: \texttt{"image/png" AND "assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
+We mined public repositories from Github to collect $54,070$ Jupyter
+notebooks that contain the keyword ``assert'' in them. We use the
+Github advanced search
+syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}
+to isolate Jupyter Notebook. Additionally, we include the ``assert''
+keyword in the search query with an \texttt{AND} operator. The final
+search query is as follows: \texttt{"image/png" AND "assert"
+  language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
 
-To further reduce the sample size, notebooks that do not use popular ML or Data Science (DS) libraries are excluded.
-A string search is used to identify notebooks that import at least one of the libraries.
-The string patterns are derived from the python module name used in the import statements.
-The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy| pandas}.
+To further reduce the sample size, notebooks that do not use popular
+ML or Data Science (DS) libraries are excluded. A string search is
+used to identify notebooks that import at least one of the libraries.
+The string patterns are derived from the python module name used in
+the import statements. The regex search query is as follows:
+\texttt{tensorflow|dask|torch|sklearn|scipy|numpy| pandas}.
 
-We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis.
-Each cell is mapped as a row in the dataframe.
-For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
-Notebooks without any visualisations are excluded by checking if the image column of the dataframe is empty.
+We programmatically parse the JSON structure of the notebooks and
+represent each notebook as a pandas dataframe for further analysis.
+Each cell is mapped as a row in the dataframe. For each cell, we store
+the \texttt{cell\_type}, the \texttt{source} of the cell and the
+base64 encoded PNG string, into the corresponding columns of the
+dataframe. Notebooks without any visualisations are excluded by
+checking if the image column of the dataframe is empty.
 
@@
 
-To identify visualisation and assertion pairs that are semantically related to one another, we only consider notebooks in which the visualisation and assertion code cells appear in a specific order.
-Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks.
-Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis.
-In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
+To identify visualisation and assertion pairs that are semantically
+related to one another, we only consider notebooks in which the
+visualisation and assertion code cells appear in a specific order.
+Figure~\ref{fig:cell-arrangement} presents a visual representation of
+various arrangements of code cells that may occur in notebooks. Note
+that the visual representation omits markdown cells since we only
+consider the order of code cells in our analysis. In the actual
+notebook however, one or more markdown cells may be present between
+the visualisation and assertion cells.
 
-We conduct the manual analysis presented in Section~\ref{sec:identify-related-pairs} for a small random sample of notebooks containing each cell arrangement.
-We observed the signal-to-noise ratio---notebooks with related VA pairs versus not---for each cell arrangement and picked the arrangement with the best results.
-Table~\ref{tab:cell-arrangement} presents the cell arrangement in ascending order of signal-to-noise ratio with a summary of our observations.
+We conduct the manual analysis presented in
+Section~\ref{sec:identify-related-pairs} for a small random sample of
+notebooks containing each cell arrangement. We observed the
+signal-to-noise ratio---notebooks with related VA pairs versus
+not---for each cell arrangement and picked the arrangement with the
+best results. Table~\ref{tab:cell-arrangement} presents the cell
+arrangement in ascending order of signal-to-noise ratio with a summary
+of our observations.
 
-Assertions defined in a cell above the visualisation cell, were typically not related to each other.
-A markdown cell present between the assertion and visualisation code cells, would often mark the beginning of a new section in the notebook.
-In other cases, the assertions were checks to ensure that the visualisation code would not encounter any errors.
-This observation also holds for Arrangement 2.1, where the assertion is defined above the visualisation code, but within the same code cell.
-Common examples of such checks include checking the shape of features that are used in the visualisation.
-Or checking that the length of the features used as the ``x''and ``y`` axes in the figure are the same to ensure a continuous line in the plot.
+Assertions defined in a cell above the visualisation cell, were
+typically not related to each other. A markdown cell present between
+the assertion and visualisation code cells, would often mark the
+beginning of a new section in the notebook. In other cases, the
+assertions were checks to ensure that the visualisation code would not
+encounter any errors. This observation also holds for Arrangement 2.1,
+where the assertion is defined above the visualisation code, but
+within the same code cell. Common examples of such checks include
+checking the shape of features that are used in the visualisation. Or
+checking that the length of the features used as the ``x''and ``y``
+axes in the figure are the same to ensure a continuous line in the
+plot.
 
-The signal-to-noise ratio improved in the sample of notebooks where the assertions were defined below the visualisation code.
-The visualisation and assertions pairs observed in Arrangement 2.2 and 3.1 were often related to each other, but were post-condition checks.
-Section~\ref{sec:post-cond} provides more details on post-condition checks and our rationale for excluding them from this study.
+The signal-to-noise ratio improved in the sample of notebooks where
+the assertions were defined below the visualisation code. The
+visualisation and assertions pairs observed in Arrangement 2.2 and 3.1
+were often related to each other, but were post-condition checks.
+Section~\ref{sec:post-cond} provides more details on post-condition
+checks and our rationale for excluding them from this study.
 
-We observed the best results in notebooks with Arrangement 3.2---where the assertion cell was not longer than 10 lines and was defined below the visualisation cell.
-We observed that some notebook authors followed the convention of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
-The visualisations were often followed by another markdown cell to record the author's observations.
-And finally, the observations were translated to a analytical test using an \texttt{assert} statement or other testing methods from external libraries.
-The restriction on the number of lines in the assertion cell was introduced to exclude instances where the assertion cell contained a new class definition or several helper functions that were not related to the visualisation above.
+We observed the best results in notebooks with Arrangement 3.2---where
+the assertion cell was not longer than 10 lines and was defined below
+the visualisation cell. We observed that some notebook authors
+followed the convention of writing their motivation for the
+visualisation in a markdown cell, before writing the code for the
+visualisation. The visualisations were often followed by another
+markdown cell to record the author's observations. And finally, the
+observations were translated to a analytical test using an
+\texttt{assert} statement or other testing methods from external
+libraries. The restriction on the number of lines in the assertion
+cell was introduced to exclude instances where the assertion cell
+contained a new class definition or several helper functions that were
+not related to the visualisation above.
 
@@
 
-The results from RQ1 provide empirical evidence to support the use of visualisations to test various properties of ML systems.
-Visualisations are an efficient and effective medium to present complex information, that is easier to interpret by humans with the aid of visual cues.
-Visualisations used to test ML system properties however, demand manual verification effort.
-Information gained from visualisations and subsequent decisions made based on this information, must be manually verified in every iteration of the ML development cycle.
+The results from RQ1 provide empirical evidence to support the use of
+visualisations to test various properties of ML systems.
+Visualisations are an efficient and effective medium to present
+complex information, that is easier to interpret by humans with the
+aid of visual cues. Visualisations used to test ML system properties
+however, demand manual verification effort. Information gained from
+visualisations and subsequent decisions made based on this
+information, must be manually verified in every iteration of the ML
+development cycle.
 
-% NOTE link to accumulation of technical debt due to highly experimental nature, and constant change in the system
+% NOTE link to accumulation of technical debt due to highly
+% experimental nature, and constant change in the system
 
-Assertions or analytical tests extracted from ML visualisations can reduce this manual verification effort.
-Assertions encapsulate the knowledge gained from the visualisation with regards to the property of the ML system which is currently under test.
-Formal assertions document the observations made by the AI practitioner regarding the model or the data, at a specific point in time.
-These observations should not be violated by the future model development cycles unless the requirements of the ML system are changed.
-Assertions scale across organisational or personnel changes as future AI practitioners can refer to the assertions to understand how the visualisations were previously interpreted.
+Assertions or analytical tests extracted from ML visualisations can
+reduce this manual verification effort. Assertions encapsulate the
+knowledge gained from the visualisation with regards to the property
+of the ML system which is currently under test. Formal assertions
+document the observations made by the AI practitioner regarding the
+model or the data, at a specific point in time. These observations
+should not be violated by the future model development cycles unless
+the requirements of the ML system are changed. Assertions scale across
+organisational or personnel changes as future AI practitioners can
+refer to the assertions to understand how the visualisations were
+previously interpreted.
 
-The results for RQ1 indicate that the combination of ML visualisations with analytical assertions is an emerging testing practice within the machine learning community.
-We found only $54,070$ notebooks that contain the keyword ``assert'', while there are more than a million Jupyter notebooks available on GitHub.
-This suggests that writing assertions in computational notebooks is still relatively uncommon.
-We believe that this is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of model properties.
-We argue that guidelines on ML coding practices require improvement.
-And advocate the need to steer education and research of Software Engineering practices towards the specific challenges faced when contributing to ML software systems.
+The results for RQ1 indicate that the combination of ML visualisations
+with analytical assertions is an emerging testing practice within the
+machine learning community. We found only $54,070$ notebooks that
+contain the keyword ``assert'', while there are more than a million
+Jupyter notebooks available on GitHub. This suggests that writing
+assertions in computational notebooks is still relatively uncommon. We
+believe that this is due to a lack of knowledge among practitioners
+regarding testing approaches and the absence of adequate tooling to
+support the assertion of model properties. We argue that guidelines on
+ML coding practices require improvement. And advocate the need to
+steer education and research of Software Engineering practices towards
+the specific challenges faced when contributing to ML software
+systems.
 
-In the 1764 notebooks analysed in this study, several visualisations lack assertions.
-However, implicit expectations still exist within each visualisation which cannot be readily validated by other project collaborators.
-ML testing demands proficiency in Machine Learning, Software Testing, and the underlying Mathematical concepts.
-This further highlights the need for refining the education for these professionals and crafting tools to compensate for gaps in their knowledge.
+In the 1764 notebooks analysed in this study, several visualisations
+lack assertions. However, implicit expectations still exist within
+each visualisation which cannot be readily validated by other project
+collaborators. ML testing demands proficiency in Machine Learning,
+Software Testing, and the underlying Mathematical concepts. This
+further highlights the need for refining the education for these
+professionals and crafting tools to compensate for gaps in their
+knowledge.
 
-Results from RQ2 reveal two challenges in the current landscape of ML testing.
-The assertions in the VA pairs identified in this study lack dedicated testing libraries.
-This indicates that the current culture of testing ML code is still in its infancy and ML practitioners are not yet exploring software testing libraries.
-Second, the existing testing libraries are not tuned to the specifics of ML code and practitioners have to fall back to basic built-in approaches.
+Results from RQ2 reveal two challenges in the current landscape of ML
+testing. The assertions in the VA pairs identified in this study lack
+dedicated testing libraries. This indicates that the current culture
+of testing ML code is still in its infancy and ML practitioners are
+not yet exploring software testing libraries. Second, the existing
+testing libraries are not tuned to the specifics of ML code and
+practitioners have to fall back to basic built-in approaches.
 
-The widespread use of ``magic'' thresholds---often rooted in domain expertise or experience obtained over time---poses risks to maintainability of ML projects.
-We advocate that such thresholds are documented for future reference.
-And find a clear need for automated tools that highlight undocumented choices or suggest standardized documentation in Markdown or code annotations.
+The widespread use of ``magic'' thresholds---often rooted in domain
+expertise or experience obtained over time---poses risks to
+maintainability of ML projects. We advocate that such thresholds are
+documented for future reference. And find a clear need for automated
+tools that highlight undocumented choices or suggest standardized
+documentation in Markdown or code annotations.
 
-In the results from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-This can lead to a mismatch of expectations in future versions of the ML software.
-For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
-The examples presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
-Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
+In the results from RQ3, we see several examples of assertions that
+fail to capture all the information gained from its visual
+counter-part. This can lead to a mismatch of expectations in future
+versions of the ML software. For instance, another project
+collaborator might run the code with a new version of the dataset or a
+different learning technique and assume that if the assertions run
+successfully, the expected patterns from the visualisation are also
+satisfied. The examples presented in Section~\ref{sec:result-rq3} show
+that deriving assertions from visualisations is not trivial. Basic
+visualisation patterns require complex mathematical techniques which
+may be beyond the expertise of an ML developer.
 
-For example, a linear correlation between the true and predicted labels is evident in Figure~\ref{fig:r2}.
-Yet, its formal assertion in Listing~\ref{lst:r2} necessitates defining an appropriate threshold for the coefficient of determination metric.
-Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
-There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
+For example, a linear correlation between the true and predicted
+labels is evident in Figure~\ref{fig:r2}. Yet, its formal assertion in
+Listing~\ref{lst:r2} necessitates defining an appropriate threshold
+for the coefficient of determination metric. Our results show that
+current software testing practices fail to capture the idiosyncrasies
+of ML code. There is a pressing need for automated tools, that bridge
+the disparity between the qualitative evaluation of ML systems using
+visualisations and the extraction of analytical assertions from these
+visuals.
 

### 2024-01-31 report/vis-assert-pairs.tex
refactor: migrate shome2023llm files
It does not make sense to keep shome2023llm as a separate project
since its a sibling of shome2023notebook.

This commit moves all relevant files of shome2023llm into this repo.

--- 2023-09-27 report/esecfse24.tex
@@
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
-
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    Bib\TeX}}}
-
-%% Rights management information.  This information is sent to you
-%% when you complete the rights form.  These commands have SAMPLE
-%% values in them; it is your responsibility as an author to replace
-%% the commands and values with those provided to you when you
-%% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-
-%%
-%% Submission ID.
-%% Use this when submitting an article to a sponsored event. You'll
-%% receive a unique submission ID from the organizers
-%% of the event, and this ID should be used as the parameter to this command.
-%%\acmSubmissionID{123-A56-BU3}
-
-\usepackage{xcolor}
-\usepackage{listings}
-\usepackage{caption}
-\usepackage{subcaption}
-
-\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
-
-\lstdefinestyle{mystyle}{
-  basicstyle=\ttfamily\footnotesize,
-  keywordstyle=\color{blue},
-  stringstyle=\color{red},
-  commentstyle=\color{green},
-  breakatwhitespace=true,
-  breaklines=true,
-  keepspaces=true,
-  showspaces=false,
-  showstringspaces=false,
-  showtabs=false,
-  tabsize=4,
-  captionpos=b,
-  numbers=left,
-  numberstyle=\tiny\color{gray},
-  backgroundcolor=\color{backcolor},
-}
-\lstset{style=mystyle}
-\usepackage{graphicx}
-\usepackage{caption}
-
-\begin{document}
-
-\title{Bridging the Gap Between Visual and Analytical Machine Learning Testing}
-
-\author{Anon}
-
-\begin{abstract}
-% what is the problem?
-% why is this an important problem to solve?
-% what did we do?
-% what did we find?
-Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
-In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
-Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
-However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
-We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
-We mine Github to collect Jupyter Notebooks that contain assertions written in Python.
-We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks where at least one pair of visualisation and assertion are semantically related to one another.
-We further read and understand 90 notebooks in depth to identify 26 qualitative and their corresponding quantitative testing patterns.
-We showcase five such patterns in this paper and plan to make the entire catalogue publicly available in the near future.
-
-\end{abstract}
-
-\maketitle
-
-\section{Introduction}\label{sec:intro}
-
-Recent years show a shift from chasing high correctness of ML models to ensuring that they are safe for application in safety-critical domains such as healthcare, finance and autonomous driving.
-With the advent of Large Language Models, the power of AI to directly influence the daily lives of humans is growing at a radical pace.
-As reports of "hallucinated" facts, racist predicts and autonomous car crashes steadily increase, ensuring that ML systems function as they were intended to, takes precedence.
-
-Visualisations are used throughout the ML development lifecycle to test various properties of the ML system.
-In the early stages of the ML development lifecycle, visualisations are extensively used for make sense of the data and verify its statistical properties.
-During the model development phase, visualisations are used to summarise metrics and contrast different learning algorithms and iteratively fine-tune ML models.
-Once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold.
-
-Building ML systems is highly iterative and experimental.
-Information gained from ML visualisations is used to make design and implementation decisions for the following steps of the ML pipeline.
-For instance, we may visualise the distribution of our training data and find that it is normally distributed.
-Based on this information, we may opt for a Linear Regression model which assumes normality in the underlying data.
-However, such expectations regarding the data may be violated once the ML system is deployed in production, where the data constantly changes as a reflect of the real world.
-Implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ML system are no longer satisfied.
-
-In contrast to prior work done in ML testing, we approach this pressing topic from a novel direction.
-We conduct an empirical analysis of computational notebooks obtained from Github, to understand the process of testing ML systems in practice.
-In particular, we focus on the combination of a qualitative form of testing (using visualisations) and a qualitative form of testing (using analytical assertions).
-The research questions along with the contributions of this paper are as follows.
-
-\begin{description}
-\item[RQ1.] How frequently do ML practitioners formulate analytical tests from the visualisations they create to test ML systems?
-
-We mine 54K Jupyter notebooks from Github that contain an assertion written in Python.
-We develop an automated technique to identify 1764 notebooks with an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 551 notebooks that contain at least one pair of visualisation and assertion that are semantically related to each other.
-
-\item[RQ2.] What are the types of visualisation and assertion pairs typically used by ML practitioners when testing ML systems?
-
-From the 551 notebooks, we further perform an in-depth analysis of 90 notebooks.
-We identify 26 unique visualisation and assertion pairs.
-We discuss five such pairs in this paper and plan to release the full catalogue publicly in the near future.
-
-\end{description}
-
-The paper is organised as follows.
-Section~\ref{sec:prelim} summarises the related literature and technical information relevant to this study.
-The methodology for the empirical study is explained in Section~\ref{sec:method} and the results of the study are presented in Section~\ref{sec:result}.
-The paper concludes with a discussion on the implications of our results (Section~\ref{sec:discuss}), threats to validity (Section~\ref{sec:threats}) and future work (Section~\ref{sec:conclude}).
-
-\section{Preliminaries}\label{sec:prelim}
-This section summarises the body of prior scientific work that has been done in ML Testing, computational notebooks and visual analytics.
-The section concludes with an overview of technical knowledge required for this paper.
-
-\subsection{ML Testing}\label{sec:ml-testing}
-
-Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems experience change in the training data and the machine learning model in addition to the codebase\cite{CITEME}.
-With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
-Existing scientific literature on ML testing broadly focuses on two aspects.
-First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy.
-
-To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
-Tools such as \textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepexplore, ma2018deepgauge, gerasimou2020importance}.
-Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
-Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
-Several works have been conducted on generating and detecting adversarial inputs for ML models.
-Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white}.
-It is however not possible to include all variations of adversarial examples into the training data.
-Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial, berend2020cats}.
-
-% TODO summarise literature on fairness testing
-Bulk of the contributions on testing non-functional properties of ML systems, focuses on fairness.
-
-\subsection{Computational Notebooks and Software
-  Engineering}\label{sec:notebooks}
-
-Computational notebooks have been ubiquitously adopted by the machine learning community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing and versioning~\cite{pimentel2019large}.
-
-\citeauthor{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise data science pipelines, and source code and metadata from over 900 releases of 12 important data science libraries.
-Their findings can be used by system builders for improving data science tools and also by data science practitioners to understand the current trends and what technologies they should focus on.
-\citeauthor{pimentel2019large} mined 1.4 million notebooks from Github to conduct an empirical study on the good and bad coding practices in computational notebooks.
-Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \citeauthor{quaranta2021kgtorrent} mine Kaggle~\footnote{https://kaggle.com} and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
-The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
-
-Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020wrong}.
-Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
-The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
-
-Since ML development is highly iterative, notebooks become messy over time and accumulate duplicate versions of code.
-\citeauthor{head2019managing} propose a code gathering tool that allows practitioners to review and keep the relevant version of code.
-
-Several tools have been proposed in the literature to aid ML practitioners when working with computational notebooks.
-
-\citeauthor{yang2021subtle} propose \textit{WrangleDoc}, a tool to automatically generate documentation of data wrangling code.
-Given a snippet of data wrangling code as input, WrangleDoc can explain what effect it has on the data.
-
-\citeauthor{bavishi2021vizsmith} proposes \textit{VizSmith}, an automatic data visualisation framework.
-Given as input a pandas dataframe and a list of columns the user wishes to visualise, the tool can automatically produce visualisations of the specified columns.
-
-\subsection{Visualisations and Machine Learning}\label{sec:visualisations}
-
-% NOTE I think the literature in this section is irrelevant to our study...
-
-As ML slowly augment into software systems in safety-critical domains, emphasis has been put into explainability of ML models.
-ML models and the underlying data is complex and multi-dimensional.
-Naturally, visual analytics has been widely adopted by the ML community to aid in understanding the nature of the data and the internal workings of an ML model.
-Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\citeauthor{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\citeauthor{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
-The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
-
-In particular, several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\citeauthor{wexler2020if} propose the \textit{What-If} tool, which allows users to visually analyse and verify their hypotheses regarding the training data and how the model works.
-With a trained model and a sample dataset as input, the What-If tool allows users to produce counterfactual reasoning, investigate the decision boundaries of the model and how the model predictions are affected by change in data.
-\citeauthor{ahn2023escape} propose \textit{ESCAPE}, a visual analytics tool that allows users to identify systemmatic errors in ML enabled systems.
-Through a human-in-the-loop approach, the tool helps identify spurious associations and concept associated misclassifications in the model.
-To reduce the cognitive load of ML practitioners during model building phase, \citeauthor{amershi2015modeltracker} propose \textit{ModelTracker}.
-Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
-
-Due to the highly pervasive nature of ML, several specialised visual analytics tools have also been proposed.
-\citeauthor{wang2023gam} propose \textit{GAM Coach}, a visual analytics tool that generates counterfactual explanations for Generalised Additive Models (GAMs).
-\citeauthor{bernard2019visual} and \citeauthor{gschwandtner2018know} propose specialised visual analytics tools for time-series data.
-\citeauthor{gortler2022neo} and \citeauthor{yang2022pattern} study and suggest improvements to traditional confusion matrix and multivariate matrix visualisations respectively.
-\citeauthor{robertson2023angler} propose \textit{Angler}, a visual analytics tool to help practitioners prioritise machine translation model improvements.
-\citeauthor{wang2023drava} propose \textit{Drava}, a tool to help practitioners relate human concepts with semantic dimensions extracted by ML models during disentangled representation learning.
-
-\subsection{Technical Background}\label{sec:background}
-\subsubsection{Internal Structure of Jupyter Notebooks}\label{sec:nbformat}
-% TODO we need to explain the json structure of jupyter notebooks, we reference this in the methodology to explain the github query.
-
-\lstinputlisting[]{nbformat.json}
-\captionof{lstlisting}{Internal JSON structure of Jupyter Notebooks.}
-\label{lst:nbformat}
-
-The data comprised within Jupyter Notebooks is stored in the JSON format, and thus can be parsed programmatically.
-Listing~\ref{lst:nbformat} shows the internal json structure of Jupyter Notebooks, adopted from the official nbformat website.~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}.
-For simplicity, we only show the parts that are relevant for this study.
-
-Notebooks contain four top level keys namely \texttt{metadata}, \texttt{nbformat}, \texttt{nbformat\_minor} and \texttt{cells}.
-The \texttt{matadata} key consists of additional information regarding the notebook such as the kernel and programming language.
-\texttt{nbformat} and \texttt{nbformat\_minor} contain the version of nbformat used in the notebook.
-The \texttt{cells} key consists of a list of all cells within the notebook.
-
-Listing~\ref{lst:nbformat} presents three examples of cells.
-Each cell is represented as a dictionary with the \texttt{cell\_type}, \texttt{metadata} and \texttt{source} keys.
-Cells can be of three types: "markdown", "code" or "raw".
-The contents of each cell can be found within the \texttt{source} key, as a list of strings.
-
-The output generated by a code cell is stored within the \texttt{outputs} key.
-Static visualisations are stored as base64 encoded png images under the \texttt{"image/png"} key, which is nested within the \texttt{data} key inside \texttt{outputs}.
-Since a cell can generate multiple outputs, the \texttt{outputs} key is a list.
-
-\subsubsection{Pre and Post-Condition Checks}\ref{sec:pre-post-cond}
-% TODO explain and give examples of what we mean by pre/post-condition checks with examples
-
-% TODO contrast this to the type of tests we are actually looking for! Why are pre/post conditions not valuable to look at?
-
-During the empirical analysis conducted for this study, we found several instances of where the assertions were related to the visualisation above.
-However, they were not relevant to this study since they did not test the property of the data or the model.
-
-\section{Methodology}\label{sec:method}
-This section presents the methodology used to conduct the empirical study introduced in Section~\ref{sec:intro}.
-Figure~\ref{fig:method} presents an overview of all relevant steps used in this study.
-The empirical analysis comprises of three distinct phases namely: \textit{the data collection phase}, \textit{identification of related visualisation and assertion pairs} and finally a \textit{complete analysis of the notebooks} to identify the final set of 26 visualisation and assertion pairs.
-
-The yellow boxes in the data collection phase represent filters that were applied to the initial corpus of 54K notebooks, to obtain the final sample size of 1.7K notebooks used in the empirical study.
-The red boxes represent the exclusion criteria used during phase two to remove notebooks note relevant for this study.
-Finally, the green boxes in phase 3 represent the manual analysis conducted to identify the final 26 visualisation and assertion pairs.
-
-Boxes appearning next to one-another imply that they were applied using an "OR" operator.
-While boxes appearning below one-another imply an "AND" operator.
-Text in double quotes represent the string pattern used in the search query.
-
-More details regarding each phase of the methodology is presented below.
+\subsubsection{Input Image, \texttt{predict\_proba}}
 
 \begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{method.pdf}
-  \caption{Methodology used to collect and analyse Jupyter notebooks
-    in this study.}
-  \label{fig:method}
+  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35a.png}}
+  \hfill
+  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-35b.png}}
+  \caption{Visualisation of random input images for a pre-trained image classifier.}\label{fig:balls}
 \end{figure}
 
-\subsection{Data Collection}\label{sec:data-collect}
-
-We mined public repositories from Github to collect Jupyter notebooks that contain the keyword ``assert'' in them.
-We intentionally keep the search criteria on Github broad by only checking for the appropriate filetype and the presence of the ``assert'' keyword.
-This is because, the ``assert'' keyword is used in python to define tests.
-Additionally, popular python libraries which provide a testing interface (such as nose, unittest and numpy), contain the same keyword in their API.
-Our search criteria captures a large initial sample of $54,070$ notebooks with a variety of testing statements.
-Our approach also prevents the need to craft a custom search query based on an exhaustive list of keywords that appear in all python testing libraries.
-
-We use the Github advanced search syntax~\footnote{REFME} to isolate Jupyter Notebook.
-This translates to the following Github search query: \texttt{filename:"Jupyter Notebooks"}.
-Additionally, we include the ``assert'' keyword in the search query with an \texttt{AND} operator.
-The final search query is as follows: \texttt{filename:"Jupyter Notebooks" AND "assert"}.
-
-The above search query results in approximately 54K notebooks.
-To further reduce the sample size, we exclude notebooks that do not use popular ML libraries.
-We perform a string search to identify notebooks that import at least one of the ML libraries.
-The string patterns are derived from the python module name used in the import statements.
-The regex search query is as follows: \texttt{tensorflow|dask|torch|sklearn|scipy|numpy|pandas}.
-
-We programmatically parse the JSON structure of the notebooks and represent each notebook as a pandas dataframe for further analysis.
-We map each cell as a row in the dataframe.
-For each cell, we store the \texttt{cell\_type}, the \texttt{source} of the cell and the base64 encoded PNG string, into the corresponding columns of the dataframe.
-We exclude notebooks that do not contain any visualisations by checking if the image column of the dataframe is empty.
-The internal structure of Jupyter notebooks is presented in Section~\ref{sec:nbformat}.
-
-\begin{figure}
-  \centering
-  \includegraphics[width=\textwidth]{nb-structure.pdf}
-  \caption{Arrangement of code cells containing assertion and
-    visualisation.}
-  \label{fig:cell-arrangement}
-\end{figure}
-
-For RQ1, we are interested in finding empirical evidence of visualisations and corresponding assertions used in the wild.
-To answer this question, we further reduce our notebook sample size by assuming a specific order in which the visualisation and assertion code cells appear in the notebooks.
-Figure~\ref{fig:cell-arrangement} presents a visual representation of various arrangements of code cells that may occur in notebooks.
-Note that the visual representation omits markdown cells since we only consider the order of code cells in our analysis.
-In the actual notebook however, one or more markdown cells may be present between the visualisation and assertion cells.
-
-% TODO don't mention quantity of random sample; then we need to explain/defend this choice
-We used an iterative approach to determine the effectiveness of each
-filter that was used. We applied each filter in the same order as
-presented in Table~\ref{tab:filters}, starting with F1. We randomly
-sampled 50 notebooks from the population of notebooks obtained after
-applying the filter. We then conduct the qualitative assessment
-presented in Section~\ref{REFME} and observe the number of related
-visualisation and assertion pairs. In each new iteration, we
-progressively add the other filters and include the filter if the
-number of relevant visualisation and assertion pairs increases.
-
-As explained above, the selected arrangement of code cells was discovered by qualitatively analysing a random sample of notebooks with the code cell arrangements presented in Figure~\ref{fig:cell-arrangement}.
-We observed the signal-to-noise ratio for each cell arrangement.
-And picked the arrangement with the least signal-to-noise ratio.
-Table~\ref{tab:cell-arrangement} summarises our observations from the various cell arrangements.
-
-% TODO give more examples (from our data) of what a pre/post condition check is
-\begin{table}
-  \centering
-  \caption{Observations from random samples of notebooks with various
-  cell arrangement.}
-  \begin{tabular}{l p{0.4\textwidth} p{0.4\textwidth}}
-    \toprule
-    \emph{\textbf{ID}}&
-    \emph{\textbf{Cell arrangement}} &
-    \emph{\textbf{Observations}}\\
-    \midrule
-    1 &
-    Assertion above visualisation cell &
-    The assertions in this arrangement were typically not related to the visualisation below.
-Often, a markdown cell present between the assertion and visualisation code cells, would mark the begining of a new section in the notebook.
-In other cases, the assertions were typically pre-condition checks to ensure that the visualisation code would not encounter any errors.
-Pre-condition checks typically include checking the shape of features that are used in the visualisation.
-Or checking that the length of the ``x''and ``y`` features are the same to ensure a continuous line in the plot.\\
-    2.1 &
-    Assertion in the visualisation cell: assertion above visualisation code &
-    Our observations were similar to that of arrangement 1. the assertions defined above the visualistion code were more likely to be pre-condition checks.\\
-    2.2 &
-    Assertion in the visualisation cell: assertion below visualisation code &
-    We found several examples where the assertion was related to the visualisation.
-Most often however, the assertions were post-condition checks to ensure the correctness of the visualisation itself.
-The signal-to-noise ratio in this arrangement was also higher compared to arrangement 3.2.\\
-    3.1 &
-    Assertion below visualisation cell: assertion cell more than 10 lines &
-    The assertions in this arrangement were frequently related to the visualisation above.
-In some cases, a markdown cell between the visualisation and assertion marked the begining of a new section in the notebook indicating that the visualisation and assertion were not related to each other.
-We also observed instances where the assertion cell defined a new class or several helper methods that were not related to the visualisation above.
-This motivated our decision to only flag notebooks where the assertion cell was no longer than 10 lines.\\
-    3.2 &
-    Assertion below visualisation cell: assertion cell has max 10 lines &
-    This arrangement had the highest signal-to-noise ratio.
-We observed that most notebook authors followed the convension of writing their motivation for the visualisation in a markdown cell, before writing the code for the visualisation.
-The visualisations were often followed by another markdown cell to record the author's observations.
-And finally, the observations were translated to a analytical test using \texttt{``assert''} or other testing methods from external libraries.\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:cell-arrangement}
-\end{table}
-
-\subsection{Identifying Related Visualisation and Assertion Pairs}\label{sec:identify-related-pairs}
-
-% TODO the numbers don't line up, because the annotations were not perfect. We 161 instances where I did not SELECT the notebook but also did not assign it an EC. How do we handle this?
-% NOTE move the 161 with the 337, then we can say we had 500+ notebooks and we cannot analyse them all in depth...
-The initial sample size of 54K notebooks from Github, was reduced to the final sample size of 1.7K after applying the filters presented in Section~\ref{sec:data-collect}.
-We analysed all 1.7K notebooks manually to further identify 337 notebooks that contained at least one pair of visualisation and assertions that were related to each other.
-Table~\ref{tab:exclusion-criteria} presents the exclusion criteria used in this phase of the analysis, along with the number of notebooks that were excluded based on each criteria.
-
-\begin{table}
-  \centering
-  \caption{Exclusion criteria used during phase 2 to identify related visualisation and assertion pairs.}
-  \begin{tabular}{l p{0.3\textwidth} p{0.4\textwidth} p{0.1\textwidth}}
-    \toprule
-    \textbf{ID} &
-    \textbf{Exclusion Criteria} &
-    \textbf{Rationale} &
-    \textbf{Num. excluded NBs}\\
-    \midrule
-    \textbf{EC1} &
-    Notebook not written in English &
-    % TODO find a better way to phrase this?
-    We excluded notebooks that were not authored in English since that is the primary language of all authors of this paper. &
-    159\\
-    \textbf{EC2} &
-    Notebook does not contain a trained ML model &
-    % TODO need to explain this better, we have notebooks with diverse python libraries; different ML models work differently so we need to understand the maths behind it.
-    We excluded notebooks that did not train an ML model.
-We found several notebooks that were on topics related to ML such as linear algebra, optimization and loss functions to name a few.
-However, identifying how the visualisation and assertion pairs can be applied to a production ML system, requires a deep understanding of several python libraries and the inner workings of various ML models.
-This was deemed beyond the scope of this research project.
-&
-    383\\
-    \textbf{EC2} &
-    Notebook contains incorrect code &
-    % TODO our filter excludes NBs without images, we have to defend why this EC is still required: jupyter adds the "image/png" field regardless of error in code
-    We excluded notebooks that contained incorrect code that did not produce a visualisation. We also excluded notebooks that used \texttt{assert} statements to stop execution of a code cell. &
-    31\\
-    \textbf{EC3} &
-    % TODO is there technical terminology for what we did to identify if the visualisation & assertion are related to one another?
-    % TODO I think we need a brief paragraph on how we prompted ChatGPT, this stuff probably needs to go outside of a table!
-    The visualisation and assertion are not related &
-    We excluded notebooks where the visualisation and assertion pairs were unrelated.
-We manually analyse the source code of the visualisation and the assertion to determine if they are related to one another.
-In some instances, the tracing was not straightforward or the code was not well written.
-In such cases, we used one-shot prompt engineering with ChatGPT 4.0 model to derive the link between the visualisation and the assertion.
-&
-    440\\
-    \textbf{EC4} &
-    The assertion is a pre or post-condition check &
-    % TODO we have to give examples of pre/post condition checks. I think this content needs to be outside of a table!
-    We excluded &
-    200\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:exclusion-criteria}
-\end{table}
-
-\subsection{Deeper Analysis of Notebooks}\label{sec:deep-dive}
-
-We qualitatively assess the comments made for all notebooks during the analysis conducted in phase two of the study, as described in Section~\ref{sec:identify-related-pairs}.
-We perform an in-depth analysis of 90 short-listed notebooks to identify unique ML testing tasks.
-The in-depth analysis was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
-During the analysis, we note several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data.
-We manually read and comprehend each code cell in the notebooks.
-When present, we also read the markdown cells above and below code cells to gain further context from the notebook author.
-For instance, we find that markdown cells above a visualisation cell typically motivates the need for the visualisation.
-A markdown cell directly below a visualisation often records the observations made by the notebook author from the visualisation.
-
-\section{Results}\label{sec:result}
-
-% TODO move the stuff for RQ1 from methodology here!
-
-% NOTE the 5 pairs is to illustrate the 26 total pairs/approach we found for testing ML systems
-
-% NOTE visual+analytical test is a new "strategy" for testing AI
-
-This study identified 26 patterns of testing using visualisation and assertion.
-This section highlights five interesting patterns in more detail.
-We plan to release the entire catalogue of patterns publicly in the near future.
-
-\subsection{Decision boundary}\label{sec:svm}
-
-% NOTE something about how the visualisation helps with interpretability of how the model is classifying the data?
-
-% NOTE the author is using the visualisation to iteratively check/test how different kernels perform on the data; can also use it for hyper-parameter tuning?
-
-% NOTE: visualisation can provide qualitative assessment of under/overfitting
-
-% NOTE: better assertion: perhaps on the support vectors?
-
-% NOTE assertion created was a single metric (performance): we need tools/guidelines to help practitioners capture what they see in visualisations; community failing to help practitioners what they are visualising
-
-% NOTE assertion is not capturing the whole thing, only assert == we lose a lot of info, we need both
-
-\begin{figure}
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-linear.png}}
-  \hfill
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-04-rbf.png}}
-  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}
-  \label{fig:svm}
-\end{figure}
-\begin{lstlisting}[language=Python]
-assert accuracy_score(y_test, pred) > 0.95
-\end{lstlisting}
-\captionof{lstlisting}{Assertion on the accuracy of the ML model.}\label{lst:svm}
-
-Figures~\ref{fig:svm} presents visualisations obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
-The model is trained on the \texttt{make\_circles}~\footnote{REFME} dataset obtained from scikit-learn~\footnote{REFME}.
-The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel (Figure~\ref{fig:svm-linear}) and later with a RBF kernel (Figure~\ref{fig:svm-rbf}).
-
-% NOTE maybe we need to ellaborate; the linear kernel was above the rbf; based on observations from the linear kernel, the author decides to try the rbf kernel; then after experimenting with hyper-parameters the author puts an assertion that captures the model's performance at that specific point in time.
-
-The visualisations presented in Figure~\ref{fig:svm} contain several useful information that can be used to test and debug ML models.
-Humans are tuned to recognise visual patterns~\cite{CITEME}.
-Visualisations in Figure~\ref{fig:svm}, allows ML practitioners to quickly identify mis-classified examples.
-Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
-For instance, in Figure~\ref{fig:svm-linear} we observe that the model is underfitting since a linear kernel too simply for the underlying data.
-The visualisation also allows ML practitioners to compare different ML models or variants of the same ML model to iteratively find a model that best fits the underlying training data.
-In this case for instance, the author of the notebook First trained the SVM with a linear kernel.
-And upon realising that the data is not linearly separable, chose the RBF kernel.
-
-Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
-The assertion checks if the model accuracy is above a threshold specified by the author of the notebook.
-The assertion is derived from the observations made from the visualisation in Figure~\ref{fig:svm-rbf}.
-The visulisation shows that the model is able to classify the data well which is why the notebook author sets the threshold for the accuracy in the assertion to a high value.
-
-The assertion however, is only able to capture the information pertaining to the performance of the model.
-While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
-
-\subsection{Input Image, \texttt{predict\_proba}}
-
-\begin{figure}
-  \subcaptionbox{Random input image of a football from the test set.\label{fig:football}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35a.png}}
-  \hfill
-  \subcaptionbox{Random input image of tennis balls from the test set.\label{fig:tennisball}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-35b.png}}
-  \caption{Visualisation of random input images for a pre-trained image classifier.}
-  \label{fig:balls}
-\end{figure}
-
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:football}.
+It checks if the probability estimate of the model is close to the specified value.
+The input image is of a football.
+Since the model is trained to detect tennis balls, the specified value is very small.}, label={lst:football}]
 np.testing.assert_almost_equal(pred_proba, 2.0355723e-05, decimal=3)
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion below Figure~\ref{fig:football}.
-It checks if the probability estimate of the model is close to the specified value.
-The input image is of a football.
-Since the model is trained to detect tennis balls, the specified value is very small.}
-  \label{lst:football}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
+
+\begin{lstlisting}[language=Python, caption={Assertion below Figure~\ref{fig:tennisball}.
+Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}, label={lst:tennisball}]
 np.testing.assert_almost_equal(pred_proba, 0.9988895, decimal=3)
 \end{lstlisting}
-  \captionof{lstlisting}{Assertion below Figure~\ref{fig:tennisball}. Compared to Listing~\ref{lst:football}, the specified value is much higher, since the input image contains tennis balls.}
-  \label{lst:tennisball}
-\end{minipage}
 
@@
 
-\subsection{Normality}\label{sec:normal}
+\subsubsection{Normality}\label{sec:normal}
 
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
-for feature in range(data_transformed.shape[1]):
-  assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
-  \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that each feature in a dataset is normal.
-The distribution of each feature is compared to that of a normal distribution using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}
-  \label{lst:kstest}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
-  \captionof{figure}{Visualisation of the distribution of a feature from the dataset.}
-  \label{fig:kstest}
-\end{minipage}
+\subsubsection{Coefficient of Determination}\label{sec:r2}
 
-% NOTE assertion and viz are perfect fit;
-
-% NOTE despite the analytical test, practitioners need to visualise this, we need visual confirmation; we cannot get rid of visualisations. we need both!
-
-The visualisation in Figure~\ref{fig:kstest} is obtained from a notebook that trains a Generative Adversarial Network (GAN) to generate data which simulates a neutrino particle detector.
-The notebook author transforms the training data by normalising it using the \texttt{QuantileTransform} from scikit-learn~\footnote{REFME}.
-After performing the transformation, the author performs a visual check by plotting the distribution of a feature from the dataset as shown in Figure~\ref{fig:kstest}.
-The visualisation shows a bell-shaped curve indicating that the data is normal and that the transformation worked.
-The author then defines the assertion presented in Listing~\ref{lst:kstest} which reflects their observation made from the visualisation.
-
-The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
-The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
-
-% NOTE I want to say that we don't even think about the fact that we always use a histogram to do this! its second nature, we are taught to do this, without thinking why!
-
-Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
-Several ML models assume normality in the underlying training data.
-The verification is often performed visually using a histogram and checking that the distribution shows a "bell-shaped" curve\cite{CITME}.
-However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
-The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
-And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
-
-\subsection{Coefficient of Determination}
-
-% NOTE this vis,assert is model agnostic!
-
-% NOTE assertion captured what they visualised
-
-% NOTE visualisation helps to check for outliers: we are humans, we look at visual patterns, quick and easier to observe fro visualisations
-
-\begin{minipage}{0.45\textwidth}
-  \begin{lstlisting}[language=Python]
-r2_gru = r2_score(y_test, y_pred)
-assert r2_gru > 0.6
-  \end{lstlisting}
-  \captionof{lstlisting}{Assertion to check that the Coefficient of Determination ($R^2$) is higer than the specified threshold.}
-  \label{lst:r2}
-\end{minipage}
-\hfill
-\begin{minipage}{0.45\textwidth}
-  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
-  \captionof{figure}{Scatterplot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}
-  \label{fig:r2}
-\end{minipage}
-
-Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
-The model is trained using time-series data consisting of historical readings of $NO_2$ and meteorological data such as temparate, humidity, wind speed, and such.
-The visualisation shows a scatterplot of the real and predicted values of $NO_2$ from the test set.
-The scatterplot shows a linear relationship between the ground truth and predictions indicating that the model was able to learn certain patterns in the training data.
-The spread of $y_{pred}$ at any given $y_{true}$ is wide.
-For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
-This Indicates that the model was not able to fit the training data that well, and can be improved with fine-tuning.
-
-% TODO more context on r2_score, it fits a linear regression model and checks the blah blah blah
-Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
-The author uses \texttt{r2\_score} metric from the scikit-learn library to quantify the performance of the model.
-The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computed the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
-The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
-Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
-
-Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess is the model is under or overfitting the training data.
-If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
-In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
-The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
-The assertion in Listing~\ref{lst:r2} however, does not capture such information.
-It can only check if the performance of the model is better than a specific threshold.
-
-\subsection{Normalization}
-
-% TODO check the notebook again for implementation of normalization
+\subsubsection{Normalization}
 
 \begin{figure}
-  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65a.png}}
-  \hfill
-  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=0.45\linewidth]{../catalogue/select-65b.png}}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=\linewidth]{../catalogue/select-65a.png}}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=\linewidth]{../catalogue/select-65b.png}}
   \caption{Visualisation of data before and after various data pre-processing techniques.}
@@
 The assertions in Listing~\ref{normal} are more relevant, however it is not clear why the author include such a wide margin in the thresholds.
-
-\section{Discussion}\label{sec:discuss}
-
-% NOTE comment on the over-whelming presence of magic thresholds in ML testing! where are authors getting these numbers from? link this to requirements engineering?
-
-% NOTE comment on the anti-patterns in testing that we observe in the 5 patterns presented
-
-% NOTE link the qualitative and quantitative testing to the ML lifecycle: how some tests will let practitioner know when something changes
-
-% NOTE comment on evolution? we have to look at the entire lifecycle of a visualisation: notebooks are run iteratively?
-
-\subsection{From Implicit Expectations to Explicit Assertions}
-% NOTE regeneration of the model does not change this observation at this point in time
-
-% NOTE discuss the pros and cons of having a qualiative assessment: they are dense with useful information; humans need visual patterns!
-
-% NOTE then show that we also need qualitative assessment that can capture our expectations; this is important so that the model never violates that we understand from the visualisations
-
-% NOTE now is the time to talk about how assertions act as documentation of the domain expertise and knowledge extracted from the visualisation from the original creator; this also scales as the system complexity and composition of the team/organisation evolves!
-
-\subsection{Something about magic thresholds}\label{sec:magic-threshold}
-% NOTE we observe that most analytical test translates to comparing a value against a threshold; this indicates the open challenges in ML testing; either we don't yet know how to translate analytical tests from visualisations or we cannot do this in the first place and we must use a hybrid approach?
-
-% NOTE this is a better place to talk about the magic threshold phenomenon; link this to the magic number smell in software engineering, but in ML it is very prevalent and an alternative is not always clear
-
-% NOTE and then motivate how we need more tools/techniques to aid practitioners to extract/translate all the info from visualisations into qualitative assertions!
-
-\subsection{Humans need visual confirmation}\label{sec:visual}
-% NOTE even though analytical tests are sufficient; humans still prefer visual confirmation
-
-% NOTE when working with image dataset, we observe in our catalogue that practitioners prefers to also visually check the input image
-
-% NOTE more detailed discussion on how a lot of information can be condensed into a single visualition and how humans are good at observing visual patterns
-
-\section{Threats to Validity}\label{sec:threats}
-% defend use of github as source for notebooks:
-% - why did kaggle not work (refer to notes)
-% - why did the existing replication packages not work?
-
-% motivation for keeping the search on github broad:
-% - not only does it pick up python assert statements, but also flags
-% - notebooks that use other testing libraries or modules that provide a
-% - testing interface. This is a good compromise between doing an
-% - exhaustive search of all testing method names in the python
-% - ecosystem and then doing a search for each of those patterns
-\section{Conclusion}\label{sec:conclude}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{bibliography}
-\end{document}

### 2024-02-13 bin/filter.py
refactor(bin): filter.py revision
1. I removed bin/ipynb2csv.{bash,py}, bin/filter.py handles reading in
   the notebook directly now.
2. bin/filter.py has some minor updates. Since `pandas.read_json` does
   not support reading a literal json string (since version 2.1.0),
   I now wrap it in a StringIO object. However, this caused the
   `pandas.Series.str.contains(*)` methods to fail silently. The
   contents of a cell are stored as a list of strings, so I join them
   into one single string using a lambda function prior to calling the
   `str.contains` method.

@@ -1,21 +1,32 @@
 """Check for assert statements in juptyer notebooks.
 
-Without any flags, the script will check for assert statements only in code cells that produce a visualisation. The search is not greedy by default. It will print the notebook to stdin if at least 1 assert statement exists in the notebook.
+Without any flags, the script will check for assert statements only in code cells that produce
+a visualisation. The search is not greedy by default. It will print the notebook to stdin if at
+least 1 assert statement exists in the notebook.
 
-With the `--proximity` flag, the script will also check in code cells directly below the visualisation cells.
+With the `--proximity` flag, the script will also check in code cells directly below the
+visualisation cells.
 
-With the `--strict` flag, the script will only flag notebooks if the assert statements is in the code cell below the visualisation cell. This flag takes precedence over `--proximity`.
+With the `--strict` flag, the script will only flag notebooks if the assert statements is in the
+code cell below the visualisation cell. This flag takes precedence over `--proximity`.
 """
 
+import json
 import ast
 import argparse
 import pandas as pd
+import numpy as np
+from io import StringIO
 
 pd.set_option("display.max_columns", 10)
 
 
 def check(df: pd.DataFrame) -> bool:
-    return not df.empty and df["source"].str.contains("assert").any()
+    decision = (
+            not df.empty and
+            df["source"].apply(lambda x: "".join(x)).str.contains("assert").any()
+            )
+    return decision
 
 
 def get_proximity_cells(
@@ -52,16 +63,67 @@ def filter(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
         ]
 
     if args.max_num_lines:
-        filtered["source_num_lines"] = filtered.source.apply(ast.literal_eval).apply(
+        filtered["source_num_lines"] = filtered.source.apply(
             lambda x: len(x)
-        )
+            )
         filtered = filtered.loc[(filtered.source_num_lines <= args.max_num_lines)]
 
     return filtered
 
 
 def print_match() -> None:
-    print(f'{args.csv.replace("csv", "ipynb")}')
+    print(f'{args.notebook}')
+
+
+def ipynb_to_dataframe(file: str) -> pd.DataFrame:
+    frames = []
+
+    with open(file) as f:
+        cells = json.load(f)['cells']
+        df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
+
+        # NOTE fill missing outputs with empty list
+        # NOTE we have to iterate through them, cannot do a bulk assignment
[…]

### 2024-02-13 bin/ipynb2csv.py

@@ -1,63 +1,0 @@
-import argparse
-import json
-import numpy as np
-import pandas as pd
-pd.set_option("display.max_columns", 10)
-
-if __name__ == "__main__":
-    frames = []
-
-    parser = argparse.ArgumentParser()
-    parser.add_argument(
-        "notebook",
-        help="Path to Jupyter Notebook",
-    )
-    args = parser.parse_args()
-
-    with open(args.notebook) as f:
-        cells = json.load(f)['cells']
-        df = pd.read_json(json.dumps(cells), orient="records")
-
-        # fill missing outputs with empty list
-        # we have to iterate through them, cannot do a bulk assignment
-        for idx, row in df.loc[df["outputs"].isna()].iterrows():
-            df.at[idx, "outputs"] = []
-
-        for row in df.itertuples():
-            frame = {
-                "index": [],
-                "cell_type": [],
-                "source": [],
-                "image": [],
-            }
-
-            if len(row.outputs) == 0:
-                frame["index"].append(row.Index)
-                frame["cell_type"].append(row.cell_type)
-                frame["source"].append(row.source)
-                frame["image"].append(np.nan)
-                frames.append(pd.DataFrame(
-                    data=frame,
-                    index=frame["index"],
-                    columns=["cell_type", "source", "image"]
-                ))
-                continue
-
-            for output in row.outputs:
-                try:
-                    frame["image"].append(output["data"]["image/png"])
-                except KeyError:
-                    frame["image"].append(np.nan)
-                finally:
-                    frame["index"].append(row.Index)
-                    frame["cell_type"].append(row.cell_type)
-                    frame["source"].append(row.source)
-
-                frames.append(pd.DataFrame(
-                    data=frame,
-                    index=frame["index"],
-                    columns=["cell_type", "source", "image"]
-                ))
-
-        pd.concat(frames).to_csv(args.notebook.replace("ipynb", "csv"))
-        print(f'INFO:ipynb2csv:wrote {args.notebook.replace("ipynb", "csv")}')

### 2024-02-13 bin/kgtorrent.py
Revert "feat(bin): init kgtorrent download script"
This reverts commit 710db18c2771bd616cd58ef982a96a57da1225bd.

@@ -1,25 +1,0 @@
-import requests
-import os
-
-if os.getenv("ACCESS_TOKEN"):
-
-    ACCESS_TOKEN = os.getenv("ACCESS_TOKEN")
-else:
-    print("kgtorrent:ERROR:ACCESS_TOKEN not found.")
-    quit()
-
-RECORD_ID = "4468523"
-RECORD_FILE = "files/KT_dataset.tar.bz2/content"
-
-r = requests.get(
-    f"https://zenodo.org/api/records/{RECORD_ID}/{RECORD_FILE}",
-    params={'access_token': ACCESS_TOKEN}
-)
-download_urls = [f['links']['self'] for f in r.json()['files']]
-filenames = [f['key'] for f in r.json()['files']]
-
-for filename, url in zip(filenames, download_urls):
-    print("Downloading:", filename)
-    r = requests.get(url, params={'access_token': ACCESS_TOKEN})
-    with open(filename, 'wb') as f:
-        f.write(r.content)

### 2024-02-13 report/icse24-nlbse.tex
feat(report): add icse24-nlbse copyright info

@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
 \copyrightyear{2024}
 \acmYear{2024}
-\acmDOI{XXXXXXX.XXXXXXX}
+\setcopyright{rightsretained}
 
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[ICSE 2024]{46th International Conference on Software Engineering}{April 2024}{Lisbon, Portugal}
+\acmConference[NLBSE '24]{2024 ACM/IEEE International Workshop on
+NL-based Software Engineering}{April 20, 2024}{Lisbon, Portugal}
+\acmBooktitle{2024 ACM/IEEE International Workshop on NL-based
+Software Engineering (NLBSE '24), April 20, 2024, Lisbon, Portugal}
+\acmDOI{10.1145/3643787.3648032}
+\acmISBN{979-8-4007-0576-2/24/04}
 %%

### 2024-02-14 bin/validate.py
feat(bin): init nb validation script
bin/validate.py uses the nbformat package to ensure that the notebook
does not contain invalid JSON and confirms to the latest Jupyter
Notebook specs.

The accompanying bin/validate.bash script runs bin/validate.py in
parallel over all notebooks within a directory.

The expected usage is to filter the raw list of notebooks through
validate.py before passing on the results to bin/filter.py which
identifies notebooks with related VA-pairs.

@@ -1,0 +1,16 @@
+"""Check if the given Jupyter notebooks contains valid JSON."""
+
+import argparse
+import nbformat
+
+if __name__ == "__main__":
+    parser = argparse.ArgumentParser()
+    parser.add_argument("notebook", help="Path to Jupyter Notebook.")
+    args = parser.parse_args()
+
+    try:
+        nb_dict = nbformat.read(args.notebook, nbformat.NO_CONVERT)
+    except Exception as err:
+        print(f"ERROR:{args.notebook}:{err}")
+    else:
+        print(f"INFO:{args.notebook}:is valid")

### 2024-02-29 bin/filter.py
refactor,fix(bin): filter.py
Minor refactor to improve readability.

Also fix a bug in the ipynb_to_dataframe method which threw an
exception when notebooks were empty.

@@ -12,7 +12,6 @@ code cell below the visualisation cell. This flag takes precedence over `--proxi
 """
 
 import json
-import ast
 import argparse
 import pandas as pd
 import numpy as np
@@ -21,16 +20,19 @@ from io import StringIO
 pd.set_option("display.max_columns", 10)
 
 
+def _list_to_str(x):
+    return "".join(x)
+
+
 def check(df: pd.DataFrame) -> bool:
     decision = (
-            not df.empty and
-            df["source"].apply(lambda x: "".join(x)).str.contains("assert").any()
-            )
+        not df.empty and df["source"].apply(_list_to_str).str.contains("assert").any()
+    )
     return decision
 
 
-def get_proximity_cells(
-    all_cells: pd.DataFrame,
+def get_cells(
+    code_cells: pd.DataFrame,
     viz_cells: pd.DataFrame,
     strict: bool = False,
 ) -> pd.DataFrame:
@@ -40,10 +42,10 @@ def get_proximity_cells(
             # add cell with visualisation
             cells.append(cell)
 
-        chunk = all_cells.loc[all_cells.index > idx]
+        chunk = code_cells.loc[code_cells.index > idx]
 
         try:
-            below = chunk.loc[(chunk.cell_type == "code")].iloc[0]
+            below = chunk.iloc[0]
         except IndexError:
             # chunk is empty or there were no code cells below the cell with visualistion
             # don't do anything
@@ -54,7 +56,7 @@ def get_proximity_cells(
     return pd.DataFrame(cells)
 
 
-def filter(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
+def filter(df: pd.DataFrame) -> pd.DataFrame:
     filtered = df.copy()
     if args.no_shape:
         filtered = filtered.loc[
@@ -63,24 +65,26 @@ def filter(df: pd.DataFrame, args: argparse.Namespace) -> pd.DataFrame:
         ]
 
     if args.max_num_lines:
-        filtered["source_num_lines"] = filtered.source.apply(
-            lambda x: len(x)
-            )
+        filtered["source_num_lines"] = filtered.source.apply(lambda x: len(x))
         filtered = filtered.loc[(filtered.source_num_lines <= args.max_num_lines)]
 
     return filtered
 
 
 def print_match() -> None:
-    print(f'{args.notebook}')
+    print(f"{args.notebook}")
 
 
[…]

### 2024-03-01 report/bsc-research-project-proposal.qmd
chore(docs): commit notes & documents

@@
 ---
-title: "Bsc. Research Project Proposal"
+title: "Automating Insight Extracting from ML Visual Data with Large Language Models"
 author: "Arumoy Shome"
-format:
-  pdf:
-    mainfont: "Iosevka Aile"
-    monofont: "Iosevka"
+format: pdf
 bibliography: bibliography.bib

### 2024-04-10 report/esem24.tex
feat,refactor(bin,docs,report)
+ Update to data collection scripts
+ init case study observations
+ init ESEM report

--- 2024-01-31 report/icse24-nlbse.tex
@@
 %% complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2024}
-\acmYear{2024}
-\acmDOI{XXXXXXX.XXXXXXX}
+% \copyrightyear{2024}
+% \acmYear{2024}
+% \setcopyright{rightsretained}
 
 %% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[ICSE 2024]{46th International Conference on Software Engineering}{April 2024}{Lisbon, Portugal}
+% \acmConference[NLBSE '24]{2024 ACM/IEEE International Workshop on
+% NL-based Software Engineering}{April 20, 2024}{Lisbon, Portugal}
+% \acmBooktitle{2024 ACM/IEEE International Workshop on NL-based
+% Software Engineering (NLBSE '24), April 20, 2024, Lisbon, Portugal}
+% \acmDOI{10.1145/3643787.3648032}
+% \acmISBN{979-8-4007-0576-2/24/04}
 %%
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Towards Automatic Translation of Machine Learning Visual
-Insights to Analytical Assertions}
+\title{TODO}
 
@@
 
-We present our vision for developing an automated tool capable of
-translating visual properties observed in Machine Learning (ML)
-visualisations into Python assertions. The tool aims to streamline the
-process of manually verifying these visualisations in the ML
-development cycle, which is critical as real-world data and
-assumptions often change post-deployment. In a prior study, we mined
-$54,070$ Jupyter notebooks from Github and created a catalogue of
-$269$ semantically related visualisation-assertion (VA) pairs.
-Building on this catalogue, we propose to build a taxonomy that
-organises the VA pairs based on ML verification tasks. The input
-feature space comprises of a rich source of information mined from the
-Jupyter notebooks---visualisations, Python source code, and associated
-markdown text. The effectiveness of various AI models, including
-traditional NLP4Code models and modern Large Language Models, will be
-compared using established machine translation metrics and evaluated
-through a qualitative study with human participants. The paper also
-plans to address the challenge of extending the existing VA pair
-dataset with additional pairs from Kaggle and to compare the tool's
-effectiveness with commercial generative AI models like ChatGPT. This
-research not only contributes to the field of ML system validation but
-also explores novel ways to leverage AI for automating and enhancing
-software engineering practices in ML.
-
 \end{abstract}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{SE4AI, NLP4Code, ML Testing, Visualisations, Assertions,
-Computational Notebooks, Automated Tool}
+% \keywords{SE4AI, NLP4Code, ML Testing, Visualisations, Assertions,
+% Computational Notebooks, Automated Tool}
 
@@
 
-Visualisations are employed at various stages of a Machine Learning
-(ML) pipeline. They are used to understand and verify data properties
-during the early stages, summarise metrics and fine-tune models during
-development, and monitor performance post-deployment. The iterative
-and experimental nature of building ML systems heavily relies on
-insights from visualisations to guide design and implementation
-decisions~\cite{yuan2020survey,hohman2019visual,amershi2015modeltracker,wexler2019what-if,shome2022data,haakman2021ai}.
-
-However, real-world data that ML systems encounter post- deployment
-seldom remain static. They often change as a reflection of the world,
-potentially violating initial assumptions made during
-development~\cite{amershi2019software,lwakatare2021on}. Every
-subsequent iteration of the ML development cycle used to retrain and
-update the ML model, therefore demands manual validation of the
-visualisations that were used to test ML system properties.
-
-Assertions or analytical tests derived from ML visualisations can
-significantly reduce manual verification efforts. Such formal
-assertions record the AI practitioner's observations about the model
-or data at a specific moment. They also serve as a reference point for
-future AI practitioners to understand the interpretations made from
-earlier visualisations.
-
-In a prior study currently under review, we investigate how
-frequently analytical tests are formulated from visualisations in ML
-systems and analysed the effectiveness of these tests. We mined
-$54,070$ Jupyter notebooks from GitHub that contain assertions written
-in Python and developed a novel methodology to identify $1,764$
-notebooks which contain an assertion below a visualisation. We
-manually analysed the $1,764$ notebooks and catalogued $269$
-semantically related visualisation-assertion (VA) pairs (to be
-released publicly with the camera-ready version of the paper). Further
-in-depth analysis of the VA pairs revealed that assertions often fail
-to capture all the information obtained from the corresponding
-visualisations. Our results indicate that formulating analytical
-assertions from visualisations is an emerging testing technique in ML.
-However, it also highlights the limitations of current practices,
-demonstrating a need for automated tools that can assist ML
-practitioners in validating visualisations and formulating more
-comprehensive analytical assertions.
-
-In this paper, we propose our research vision to develop an automated
-tool to generate analytical assertions from ML visualisations. The
-research questions we plan to address along with the contributions we
-envision to make are presented below.
-
-\begin{description}
-  \item[RQ1.] \textbf{How are VA pairs used to perform ML verification
-    tasks?}
-
-    We begin by creating a taxonomy that clusters the existing $269$
-    VA pairs by specific ML verification tasks. This taxonomy not only
-    enables us to fine-tune the data and model for the automated tool,
-    but also serves as valuable reference for future ML practitioners.
-
-  \item[RQ2.] \textbf{What kind of input features enable AI models to
-    generate assertions from visualisations?}
-
-    The dataset consists of visualisations, source code and natural
-    text. Based on the manual analysis we previously conducted to
-    identify semantically related VA pairs, we hypothesis that the
-    source code for the visualisation is sufficient for ML models to
-    generate meaningful assertions. We plan to compare the results to
-    multi-modal input feature space where a combination of image, code
-    and text are used.
-
-  \item[RQ3.] \textbf{What kind of AI models generate the best
-    assertions from visualisations?}
-
-    A empirical study will be conducted to compare traditional
-    NLP4Code models with modern Large Language Models. The
-    effectiveness of these models will be evaluated both using
-    a holdout set and additionally using a qualitative study with
-    human participants.
-
-  \item[RQ3.] \textbf{How does our tool compare to commercial
-    Generative AI models?}
-
-    This paper presents results from a preliminary study which we
-    conducted to evaluate the effectiveness of the ChatGPT 4.0 model
-    to automatically generate assertions from ML visualisations. We
-    plan to extend the preliminary study to the full catalogue on VA
-    pairs and compare the results to our proposed tool.
-
-\end{description}
-
-\section{Our Vision}
-
-Figure~\ref{fig:vision} provides a visual summary of the research
-vision proposed in this paper. During the data collection phase, an
-extended catalogue of VA pairs will be created using VA pairs mined
-from Github and Kaggle. A taxonomy of VA pairs used for ML
-verification tasks will be created and used to fine-tune the dataset
-and model. Quantitative studies will determine the best input feature
-space and AI models that are able to generate meaningful assertions
-from visualisations. Quantitative and qualitative evaluations will be
-employed to position the proposed tool with-respect-to ChatGPT.
-
-\begin{figure*} \includegraphics[width=\textwidth]{vision.pdf}
-\caption{Vision for the automated tool proposed in this paper to
-generate analytical assertions from ML visualisations.}
-\label{fig:vision} \end{figure*}
-
-\subsection{RQ1: How are VA pairs used to perform ML verification tasks?}
-
-The existing catalogue of $269$ VA pairs was created by mining public
-Jupyter notebooks on Github, that contained both visualisations and
-assertions. We searched the entirely of Github and obtained $54,070$
-Jupyter Notebooks written in Python, containing the keyword ``assert''
-in them. The sample was further refined to $1,764$ notebooks that
-imported popular Machine Learning or Data Science libraries and had
-specific cell arrangements that indicated a potential relationship
-between visualisations and assertions. We manually reviewed these
-notebooks to identify $269$ VA pairs that were semantically related to
-each other. This analysis involved manually examining the source code
-of the visualisation and assertion cells and excluding unrelated
-pairs.
-
-The $269$ VA pairs however lack structure. Therefore, We plan to
-organise them using a taxonomy that categories VA pairs according to
-specific ML verification tasks. This taxonomy is critical for the
-creation of an automated tool, as it allows for more targeted
-fine-tuning of the training data and model. We hope that by aligning
-the model with distinct verification tasks identified in the taxonomy,
-the tool can generate more accurate and relevant assertions.
-
-The taxonomy also organises the domain knowledge of ML visualisations
-and assertions. It can be used by future ML practitioners as reference
-to identify the types of visualisation that can be used for a given
-verification task, or formulate comprehensive assertions given a visualisation.
-
-\subsection{RQ2: What kind of input features enables AI models to
-generate assertions from visualisations?}
-
-Using the taxonomy created in the prior phase, we plan to create
-a high quality dataset of VA pairs which can be used to train and test
-AI models that automatically generate the assertions. The input
-feature space comprises of three rich sources of information---the
-visualisations stored as PNG images, Python source code used to create
-the visualisation and when available, natural text written by the
-notebook authors above and below the visualisation.
-
-In our prior study, we analysed $1,764$ notebooks manually to identify
-$269$ VA pairs. We reflected on our manual analysis procedure and
-realised that the code for the visualisation and the assertion provide
-sufficient context to determine whether they are related to each other
-or not. Our hypothesis is that the Python source code along with the
-Markdown text should provide sufficient information for the ML model
-to generate meaningful assertions. However, it will also be
-interesting to compare the results with multi-modal AI models which
-also use the visualisations as input.
-
-\subsection{RQ3: What kind of AI models generate the best assertions
-from visualisations?}
-
-We plan to conduct an extensive empirical study to determine what type
-of AI models are able to produce the best assertions. Here, we plan to
-compare and contrast traditional NLP4Code models with open-source
-Large Language Models (LLMs).
-
-From an NLP perspective, creating assertions given the source code for
-a visualisation is a translation task. Thus, the effectiveness of the
-above models can be quantified using existing translation metrics such
-as \textit{BLEU}, \textit{Meteor} and
-\textit{Rouge-L}~\cite{zhu2023revisiting}.
-
-We further plan to evaluate the practical relevance and impact of our
-proposed tool by conducting a qualitative study with human
-participants. We are currently exploring two approaches here. First,
-with human participants that compare the ground truth assertions with
-the ones generated by our tool, and rate them on a Likert Scale on
-Naturalness and Correctness~\cite{izadi2022on}. An alternative
-approach would be to collect feedback from the notebook authors by
-opening Pull Requests on Github that propose the generated assertion.
-
-\subsection{RQ4: How does our solution compare to commercial
-generative AI models?}
-
-The ``Elephant in the room'' must be addressed. OpenAI's suite of
-commercial generative AI models has produced a tectonic shift in the
-Software Engineering landscape. We plan to evaluate the effectiveness
-of our proposed tool against commercial AI models such as ChatGPT and
-DALL-E.
-
-In a preliminary study, we assessed the capability of ChatGPT 4.0 to
-generate assertions from visualisations, utilising six VA pairs that
-were highlighted in our prior study. The study employed one-shot
-prompting with three different inputs---providing ChatGPT with only
-the visualisation image, only the code generating the visualisation,
-and both the image and code. The responses were evaluated based on
-three criteria---whether the generated assertions reflected properties
-or trends evident in the visualisations, the similarity of the
-response to the ground truth assertions, and whether ChatGPT proposed
-alternative assertions that validated the same visual properties as
-the ground truth.
-
-All 18 responses from ChatGPT contained assertions derived from visual
-properties, notably even when prompted solely with the code. Out of
-these, 8 responses closely matched the ground truth assertions, and in
-the remaining 10, ChatGPT suggested valid alternative assertions
-aligned with the same visual properties as the ground truth. We plan
-to extend this study to the entire catalogue of VA pairs which will
-allow us to compare and contrast our proposed tool to the abilities of
-ChatGPT.
-
-\section{Challenges}
-
-\subsection{Dataset Limitations and Bias}
-
-The quality and diversity of the dataset used to train the tool will
-heavily influence its effectiveness. There is a risk of bias if the
-datasets are not representative of real-world scenarios or if they
-lack diversity in terms of the types of visualisations and assertions.
-
-In RQ1, the existing $269$ VA pairs were obtained from Github which
-might not be representative of all VA pairs observed in ML. To
-mitigate this threat, we plan to extend the catalogue with VA
-pairs mined from Kaggle. This also allows us to simplify our data
-mining process by eliminating the need to filter notebooks for
-relevant libraries since Kaggle is primarily targeted towards
-the Data Science and Machine Learning communities.
-
-In RQ2, we expect to face challenges with respect to the quantity and
-quality of the training data. Since translating visual properties to
-analytical assertions is an emerging testing practise, we may not have
-sufficient training data for our AI models. The extension using VA
-pairs from Kaggle should alleviate this threat to a certain extent.
-Another alternative would be to fine-tune the NLP4Code models using
-our catalogue. With LLMs, we can experiment with different prompting
-techniques such as few-shot prompting~\cite{liu2023pretrain}.
-
-% TODO come back to this one
-With respect to quality, we want to reinforce the manual analysis
-required to determine whether a given visualisation and assertion are
-related, with multiple annotators. The quality of the annotations can
-be determined by calculating the inter-rater agreement using the
-\textit{Kappa} metric.
-
-In RQ4, we do not know the contents of the training corpus used to
-train ChatGPT. However given its colossal size, it is safe to assume
-that our catalogue of VA pairs exists in its training data (or will be
-included in future versions). This supports our decision to use
-open-source LLMs where we can validate that no such data leakage
-occurs. We can also extend the proposed evaluation for our tool (RQ3)
-and ChatGPT (RQ4) using VA pairs that have not been publicly released.
-One potential solution here would be to collect VA pairs created by
-students in a Software Engineering for ML course.
-
-\subsection{Complexity of Visual Analytics}
-
-Translating visual properties accurately into analytical assertions is
-a complex task. The tool must be capable of understanding and
-interpreting a wide range of visual patterns and nuances, which is
-a significant challenge. In our prior paper, we identified several
-instances where the analytical assertions were not able to capture all
-the information present in their visual counter-part. While our
-proposed tool may not be able to capture all the nuances of
-a visualisation every time, identifying this boundary of automation
-and human intervention remains an important and exciting research
-direction.
-
-\subsection{Integration into Existing Workflows}
-
-For the tool to be widely adopted, it needs to be seamlessly
-integrated into existing ML development workflows. This includes
-compatibility with various programming environments and ML frameworks.
-While the outcome of our research endeavours may be a simple
-prototype, we hope to lay the foundations that enable future
-researchers to integrate our tool into existing computational notebook
-environments such as Jupyter Lab or Visual Studio Code. Such an
-integration may enable the tool to provide better assertions using
-additional context either obtained directly from the user through
-a chat interface~\footnote{Such as the one provided by Github Co-Pilot
-extension for Visual Studio Code:
-https://code.visualstudio.com/blogs/2023/03/30/vscode-copilot} or
-derived from the entire software codebase~\footnote{One example of
-such an editor extension is Cody which is available for Visual Studio
-Code and JetBrains: https://github.com/sourcegraph/cody}.
-
-\section{Expected Outcomes}
-
-The proposed tool automates the translation of visual properties from ML
-visualisations into Python assertions. This automation reduces the
-manual effort required in the model validation process, making it
-faster and more efficient. By automating a critical part of the ML
-development process, the tool can streamline the entire cycle from
-data preprocessing to model deployment, enabling quicker iterations
-and enhancements. Automated assertions make it easier to handle large
-datasets and complex models, as it removes the bottleneck of manual
-analysis, thereby enhancing scalability.
-
-Manual processes are prone to human errors, such as overlooking subtle
-visual cues in data. In contrast, an automated tool can consistently
-capture and translate these nuances, reducing the likelihood of
-errors. This form of automation leads to a more standardized approach
-to validating ML models, ensuring that the same criteria are applied
-consistently across different models and datasets. Furthermore,
-automated assertions provide a reliable record of the state of the
-data and model at various points in time, aiding in historical
-analysis and model auditing.
-
-An automated tool also enables ML practitioners to adapt assertions as
-real-world data evolves. By providing immediate feedback on visual
-data, the tool will enable faster updates and refinements to ML models
-in response to changing data trends. The tool could potentially
-identify anomalies or shifts in data patterns sooner than manual
-processes. This not only allows for proactive adjustments to models,
-but also significantly reduces effort required to maintain the
-relevance and accuracy of ML models over time.
-
-\section{Expected Outcomes Beyond ML Testing}
-
-As ML augments software systems in safety-critical domains, there has
-been a significant emphasis on the explainability of ML models. The
-complexity and multi-dimensionality of such models and their
-underlying data has lead to the adoption of Visual Analytics by the ML
-community. However, interpretation of a visualisation may vary between
-developers with varying domain expertise and background. Converting
-such visual interpretations into assertions can help bring consistency
-into the interpretability of ML visualisations across all project
-members.
-
-The concept of translating visual properties to formal assertions can
-be extended to other domains (such as scientific computing and
-simulation), where visual analytics also plays a prominent role in the
-decision making process. Future research could explore ways to
-customise these tools for specific industries or types of data,
-increasing their applicability and effectiveness.
-
-The effectiveness of the tool in real-world scenarios could also open
-research into how human practitioners and AI tools can best
-collaborate, particularly in the context of complex problem-solving
-tasks like ML model validation.
-
-\section{Conclusion}
-
-Developing an automated tool to convert visual properties from ML
-visualisations into analytical assertions marks a timely advancement
-in the domains of Machine Learning and Software Engineering. This tool
-has the potential to improve the efficiency, accuracy, and consistency
-of ML model testing and validation. Nevertheless, it also presents
-considerable challenges, especially in areas such as visual analysis,
-data set quality, and integration with current workflows. The research
-initiates various new opportunities, such as enhancing the
-interpretability of AI, enabling automated testing, and exploring the
-dynamics of human-AI collaboration.
-
 \bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}

### 2024-04-10 bin/create-master-assert-content.py

@@ -1,0 +1,54 @@
+import pandas as pd
+import ast
+
+
+def get_ast(source: str) -> ast.Module:
+    try:
+        tree = ast.parse(source)
+    except:
+        tree = None
+    finally:
+        return tree
+
+
+def get_assert_nodes(tree: ast.Module) -> list[ast.Assert]:
+    collector = AssertCollector()
+    collector.visit(tree)
+    return collector.nodes
+
+
+class AssertCollector(ast.NodeVisitor):
+    def __init__(self):
+        self.nodes = []
+
+    def visit_Assert(self, node: ast.Assert) -> None:
+        self.nodes.append(node)
+
+
+if __name__ == "__main__":
+    github = pd.read_csv(
+        "data/shome2023notebook/github-assert-content.csv",
+        header=None,
+        names=["cell_type", "source", "notebook"],
+    )
+
+    kaggle = pd.read_csv(
+        "data/shome2023notebook/quaranta2021kgtorrent-assert-content.csv",
+        header=None,
+        names=["cell_type", "source", "notebook"],
+    )
+
+    asserts = pd.concat([github, kaggle])
+
+    asserts.loc[:, "ast"] = asserts["source"].apply(get_ast)
+    asserts = asserts.loc[asserts["ast"].notna()]
+    asserts = asserts.loc[
+        asserts["ast"].map(lambda x: True if list(ast.walk(x)) else False)
+    ]
+    asserts.loc[:, "assert"] = asserts.loc[:, "ast"].apply(get_assert_nodes)
+    asserts = asserts.explode("assert")
+    asserts = asserts.loc[asserts["assert"].notna()]
+    asserts.loc[:, "assert"] = asserts.loc[:, "assert"].apply(lambda x: ast.unparse(x))
+    asserts = asserts.drop_duplicates(subset=["assert"])
+
+    asserts.to_csv("data/shome2023notebook/assert-content.csv", columns=["notebook", "assert", "source"])

### 2024-04-10 bin/data-collection-asserts.py

@@ -14,8 +14,6 @@ def ipynb_to_dataframe() -> pd.DataFrame:
     with open(args.notebook) as f:
         cells = json.load(f)["cells"]
         df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
-        # NOTE: remove outputs column
-        df = df.loc[:, ["cell_type", "source"]]
 
     # NOTE: early exit if empty notebook or no code cells
     (df.empty or df.loc[df["cell_type"] == "code"].empty) and exit()
@@ -31,7 +29,7 @@ def ipynb_to_dataframe() -> pd.DataFrame:
     # NOTE: `source` can be str or list; following lambda function works on both cases
     df.loc[:, "source"] = df["source"].apply(lambda x: "".join(x))
 
-    return df.loc[:, ["cell_type", "source", "outputs"]]
+    return df.loc[:, ["cell_type", "source"]]
 
 
 if __name__ == "__main__":
@@ -48,6 +46,16 @@ if __name__ == "__main__":
     all_cells = ipynb_to_dataframe()
     code_cells = all_cells.loc[all_cells["cell_type"] == "code"]
     md_cells = all_cells.loc[all_cells["cell_type"] == "markdown"]
+
+    import_cells = code_cells.loc[code_cells["source"].str.contains("import")]
+    (
+        import_cells.empty
+        or import_cells.loc[
+            import_cells["source"].str.contains(
+                r"sklearn|torch|tensorflow|keras", regex=True
+            )
+        ].empty
+    ) and exit()
 
     # NOTE: this may return false positives (such as the keyword `assert` appearing in a comment)
     assert_cells = code_cells.loc[code_cells["source"].str.contains("assert")]
@@ -84,13 +92,12 @@ if __name__ == "__main__":
     # NOTE: make sure that the term `ipynb` only occurs at the end of filename
     # I did this using the following command and checking that it *does not* print anything:
     # $ find data/ -name '*.ipynb' -type f -not -path '*ipynb_checkpoints*' |grep -v 'ipynb$'
-    basename, _ = os.path.splitext(args.notebook)
-    dirname = os.path.dirname(args.notebook).replace(
-        "data/", "data/shome2023notebook/", 1
-    )
+    dirname, filename = os.path.split(args.notebook)
+    dirname = dirname.replace("data/", "data/shome2023notebook/", 1)
     if not os.path.exists(dirname):
[…]

### 2024-04-10 bin/data-collection-outputs.py

@@ -86,6 +86,17 @@ if __name__ == "__main__":
 
     all_cells = ipynb_to_dataframe()
     code_cells = all_cells.loc[all_cells["cell_type"] == "code"]
+
+    import_cells = code_cells.loc[code_cells["source"].str.contains("import")]
+    (
+        import_cells.empty
+        or import_cells.loc[
+            import_cells["source"].str.contains(
+                r"sklearn|torch|tensorflow|keras", regex=True
+            )
+        ].empty
+    ) and exit()
+
     outputs = get_outputs()
     # NOTE: early exit if neither outputs.data.{"image/png","text/plain"} were present
     outputs.empty and exit()
@@ -96,6 +107,6 @@ if __name__ == "__main__":
         os.makedirs(dirname)
 
     # NOTE: order of headers ["index", "notebook", "source", "image/png", "text/plain"]
-    name = dirname + os.path.splitext(filename)[0] + "-outputs.csv"
+    name = os.path.join(dirname, os.path.splitext(filename)[0] + "-outputs.csv")
     outputs.to_csv(name, header=False)
     print(f"OUTPUT:{name}")

### 2024-05-02 report/esem24.tex
feat(report): init esem24 paper

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{TODO}
+\title{Exploring the Role of Feedback in Developing Machine Learning
+Code}
 
@@
   \country{Netherlands}}
+
+\author{Diomidis Spinellis}
+\email{d.spinellis@tudelft.nl}
+\affiliation{%
+  \institution{Delft University of Technology}
+  \country{Netherlands}
+}
 
@@
 
+\section{Methodology}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{esem24-method.pdf}
+\end{figure}
+\section{Results}
+
+% TODO: for each of the feedback types; start with the descriptive and lexical analysis we performed
+% TODO: then move into the case studies, and link them to existing literature; use the fault and failure taxonomies; perhaps also link the feedbacks to ML lifecylce proposed by Amershi 2019?
+
+\subsection{Implicit feedback using \texttt{print} statements and output of code cells}
+
+% TODO: why do we check the distribution of the data, particularly during the EDA phase?
+\textbf{\emph{Data distribution check (O2, O4, O9, O14, O20, O25, O48)}}
+
+\textbf{\emph{Data relationship check (O6, 010)}} Check if feature is linearly related with the target feature (for linear regression models).
+
+\textbf{\emph{Dependency check (P71, P86)}}. P71 checks that the version of an external library while P86 checks that Cuda is available.
+
+\textbf{\emph{Execution time check (P66)}}. Print total training time of the model.
+
+% TODO: highlight the use of visualisations to check for missing values; what is the benefit of using a visualisation instead of code?
+\textbf{\emph{Missing value check (O12, O36, P74)}}. Check if missing values are present in any features of the data. After loading the dataset and in the EDA phase of analysis.
+
+\textbf{\emph{Model performance check (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)}}.
+Check various performance metrics of models on the test set.
+Typically, the classification report is used.
+Precision and recall is checked using the confusion matrix.
+Residuals of a linear model is checked.
+
+\textbf{\emph{Model architecture (P92)}}. Print the architecture of the neural network. The layers of the model, neurons in each layer and the activation functions are printed in a text diagram.
+
+\textbf{\emph{Resource check (O64, O66, P68, P82, P107)}}.
+Check is GPU or TPU resources are available.
+We suspect these checks are coming from Google Colab notebooks where computation can be off-loaded to remote hardware.
+In P107, check that the pre-trained model was successfully loaded from the file system.
+
+\textbf{\emph{Shape check (P4, P32, P117)}}. Check the number of examples in the test data; check shape of training data; check shape of training labels.
+
+\textbf{\emph{Type check (O71, P43)}}. Check that the data type of the features in the training set
+
+% NOTE: perhaps these are spot-checks?
+\textbf{\emph{Value check (O56, O60, P64, P67, P114)}}. Check that one-hot encoding worked; manually verify the output of an activation function in a neural network against expected value.
+
+\textbf{\emph{Model training (O8, O31, O42, P77)}}. Periodically print the progress made in training by printing the training loss/accuracy of model.
+
+\subsection{Explicit feedback using \texttt{assert} statements}
+
+% TODO: find out more about batch training
+\textbf{\emph{Batch size check (A21, A28, A70)}}.
+Neural Networks are trained using smaller batches that makes it more efficient use of hardware GPU cores.
+We encounter assertions that check the size of the batch prior to training or prior to calculating the score of the NN.
+
+\textbf{\emph{Data leakage check (A33)}}.
+It is a best practise to ensure that the ML model is not tested using examples that it has already encountered during training.
+This is to ensure that the model is not overfitting on the training data and is able to generalise to examples similar to the distribution of the training set, but not exactly the same.
+In A33 we see that the assert ensures that the training and validation sets do not share any examples.
+
+% TODO: what is the motivation of this assertion? Find prior work on
+% python dependency management?
+\textbf{\emph{Dependency check (A18, A67)}}.
+This is a unique type of test which occurs due to the unique workflow of notebooks.
+Typically dependencies are managed through external means such as dedicated dependency management solutions or the builtin Python method using a requirements.txt file.
+We notice assertions that check that the major version of an external dependency is equal to the specified value.
+
+% NOTE: missing-value-check and existence-check
+
+\textbf{\emph{Existence check (A86)}}.
+Existence checks are carried out in various contexts.
+Most are for checking that columns in a dataset do not contain missing values.
+Others ensure that certain columns (such as the column containing the labels of the data) exists in the dataset prior to splitting the data into training and testing sets.
+Existence checks are frequently performed after data pro-processing  steps are performed to ensure that missing values were not added to the dataset.
+
+% TODO: we need to elaborate this one with examples
+
+\textbf{\emph{Mathematical property checks (A3, A25, A56, A64)}}.
+Machine learning models are often rooted in statistical models.
+Implementing neural networks requires a graph of linear algebra and matrix multiplication.
+We find several assertions that test the mathematical properties of arrays and matrices after performing matrix operations.
+
+% TODO: there is a lot to discuss here, bring in the discussion on the
+% use of comparison operators and such into the mix here
+
+\textbf{\emph{Model performance check (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)}}.
+We see several assertions that test for the performance of the trained ML model against a validation or test test.
+The assertions often compare a performance metric (such as accuracy, precision, recall, F1, etc) against a predefined threshold.
+
+% TODO: this needs more work and investigation
+\textbf{\emph{Network architecture check (A11)}}. We only have one example here.
+
+% TODO lots to elaborate and discuss here
+\textbf{\emph{Resource check (A10, A14, A37, A60, A74)}}.
+We find assertions that check that a pre-trained model exists on the file system, or that data can be loaded from a specified file path.
+There are also checks that ensure that a visualisation created using matplotlib contains data in it.
+
+\textbf{\emph{Data shape check (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)}}. ``Swiss army knife'' of assertions!
+
+\textbf{\emph{Type check (A2, A35, A40, A81, A88)}}. Checking the data type of features and such.
+
+\textbf{\emph{Value check (A30, A41, A44--46, A52, A65, A68, A69, A73, A92)}}.
+Check that the values in a column are in the specified values (categorical) or in a specified range (normalised continuous variable) or binary (such as the label column).
+
+% NOTE: blanks and unit-tests
+\textbf{\emph{Miscellaneous checks (A6, A20, A32, A34, A36, A53, A62, A75, A83, A87, A97)}}. These feel like unit tests, performed in lieu of writing full test suites in the notebook.
+
 \bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}
-
 \end{document}

### 2024-05-06 report/emse24.tex
feat(report): init emse24

%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
    language=Python,
    basicstyle=\ttfamily,
    keywordstyle=\color{blue},
    stringstyle=\color{red},
    commentstyle=\color{green},
    morecomment=[l][\color{magenta}]{\#}
}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Exploring the Role of Feedback in Machine Learning Jupyter Notebooks}

%\titlerunning{Short form of title}        % if too long for running head

\author{Arumoy Shome\and
  Lu{\`\i}s Cruz\and
  Diomidis Spinellis\and
  Arie van Deursen
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{A. Shome \at
  Delft University of Technology\\
  \email{a.shome@tudelft.nl}
  \and
  L. Cruz \at
  Delft University of Technology\\
  \email{l.cruz@tudelft.nl}
  \and
  D. Spinellis \at
  Delft University of Technology\\
  \email{d.spinellis@tudelft.nl}
  \and
  A. V. Deursen \at
  Delft University of Technology\\
  \email{arie.vandeursen@tudelft.nl}
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
Insert your abstract here. Include keywords, PACS and mathematical
subject classification numbers as needed.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}

\section{Methodology}
% For one-column wide figures use

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{esem24-method.pdf}
  \label{fig:data-collection}
  \caption{Overview of data collection methodology used in this study}
\end{figure}

\section{Results}

% TODO: for each of the feedback types; start with the descriptive and lexical analysis we performed
% TODO: then move into the case studies, and link them to existing literature; use the fault and failure taxonomies; perhaps also link the feedbacks to ML lifecylce proposed by Amershi 2019?

\subsection{Implicit feedback using \texttt{print} statements and output of code cells}

\subsubsection{Data distribution check (O2, O4, O9, O14, O20, O25, O48)}
% a mix of visualisations and pandas dataframes that show the distribution of a column in the data
% we primarily observe these checks being used to determine the relationship of a feature with the target feature
% typically in the EDA/data exploration/understanding phase
% several examples where the distribution of a newly created feature is checked (binning a continuous feature into categorical one)
% one uses descriptive statistics to verify that data normalisation worked (again, a check after data transformation step)

We observe the use of visualisations and dataframes to check the distribution of a feature. Often, the analysis is performed in the context of a multivariate analysis, where the distribution of a feature is checked against all possible outcomes of the target feature.

Typically, this check is performed in the EDA phase where the author is trying to understand the relationship of a feature with the target feature. This done to identify features that can be used to train the ML model.

We also observe that the distribution of a feature is checked directly after creating it. For example, we observe this check after converting a continuous feature into a categorical feature using binning. Another example is the use of descriptive statistics provided by pandas to check that the prior data normalisation step worked (the author manually verified the mean and standard deviation of the feature from the output).

\subsubsection{Data relationship check (O6, O10)}

Linear regression models assume a linear relationship between the features in the training data, with the target feature. We see this being verified using lineplots to check the presence of a linear trend between a given feature and the target feature.

A special case is O10 which uses a regplot. This is a visualisation provided by the seaborn library that plots the data along with a linear regression model fit.

\subsubsection{Dependency check (P71, P86)}

We find evidence of lack of software engineering best practise. Dependencies should be manage with external programs built for this task. Python has a built-in functionality for this, where dependencies along with version can be specified in a requirements.txt file. There are also several external programs such as the use of lockfiles by pipenv and poetry which lock a specific version of a dependency such that all files are reproducible on all systems.

P71 prints the version of an external library which the author is expected to verify manually. P86 prints a confirmation message if the Cuda library is available on the system where the notebook is being executed.

\subsubsection{Execution time check (P66)}

Often several ML models are trained and for a single task. The selection is based on requirements that not only consider the performance of the model, but also the total time required to train the model. Since this is an iterative task, a model that can be more efficiently retrained, helps reduce costs in the long run, and has been proven to be more sustainable.

P66 is from a notebook where several models are trained for the same task, and their performance along with the time taken to train is printed. The cells of a notebook are also executed several times, for a long running cell (such as the ones that train the model), an indication of the total execution time helps (after the first run, the author has a baseline for the time it will take).

\subsubsection{Missing value check (O12, O36, P74)}

Check if missing values are present in any features of the data. After loading the dataset and in the EDA phase of analysis.

From our data smells paper: descriptive statistics functions typically ignore missing values; leading to biased and incorrect results and conclusions

ML models also except numerical values, a null type value raises an exception (since computations cannot be performed)

P74 and O36 is an aggregated check over the entire dataset.

% TODO: highlight the use of visualisations to check for missing values; what is the benefit of using a visualisation instead of code?
More interesting is O12 which uses a heatmap to show the distribution of the missing values across all features of the dataset.

\subsubsection{Model performance check (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)}

The performance is printed to compare against other models/variations in the experimental model development phase; during the continuous experimentation phase, the author might change parameters of the model or change the data (engineer features, etc) and then re-run the model to check if the performance improved

Mean accuracy over K-fold validation

Other performance metrics such as recall, precision and F1 also printed

Error in predictions (RMSE, MSE)

P48: prints the loss

% TODO: whats the point of this?
P54: print the intercept of linear regression model

Classification report is used to produce a summary of popular performance metrics such as accuracy, recall, precision and F1.

Precision and recall is checked using the confusion matrix.

% TODO: highlight the need for normalising vs. comparing absolute values; the data might be imbalanced so does not make sense to compare the absolute values
O3: use heatmap to visualise the confusion matrix of a multi-label classification task. The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.

O52: author wrote a custom function which gets the predictions of an image classifier (developed in the NB) on a random sample of images. The predictions along with the input image is printed. The author is manually spot-checking the accuracy of the predictions. Or perhaps trying to find examples that produce the wrong prediction?

\subsubsection{Model architecture (P92)}

Several neural network architectures are available as part of the ML library (for instance keras, pytorch and tensorflow come pre-loaded with several NNs).

Print the architecture of the neural network. The layers of the model, neurons in each layer and the activation functions are printed in a text diagram.

\subsubsection{Resource check (O64, O66, P68, P82, P107)}

Check is GPU or TPU resources are available. We suspect these checks are coming from Google Colab notebooks where computation can be off-loaded to remote hardware.

In P107, check that the pre-trained model was successfully loaded from the file system.

\subsubsection{Shape check (P4, P32, P117)}

% TODO: why is shape check important?
Checking the shape of the data is important for several reasons. If we perform some data pre-processing or transformation, we want to ensure that the shape of the data is the same as expected.

Number of features in the training and testing data must match. When using statistical ML models, the model "fits" the training data and thus expects the same shape in the testing data to make inference. When developing neural networks, architectural decisions such as the number of input neurons is made based on the shape of the data.

During the testing shape, the number of examples in the test input must be the same as the number of labels so that the performance metrics such as accuracy, can be correctly calculated.

\subsubsection{Type check (O71, P43)}

All ML models expect the input data to be entirely numerical, so that mathematical operations such as matrix multiplications may be carried out without errors.

Checking the data type in the EDA phase may be done to understand the optimal data pre-processing or transformation techniques necessary, to make the data ready for model. For instance, numerical features may need to be normalised while categorical features need to one-hot encoded.

\subsubsection{Value check/ Spot checks (O56, O60, P64, P67, P114)}
% NOTE: perhaps these are spot-checks?

O56: check the prediction of the model on a single input image

O60: check the number of features in the data is equal to the expected number after performing PCA, the author checks the head of the dataframe and manually verifies the number of features

P64: find the max in the second channel of a 3D numpy array (highlight this case study; this is done for image processing!)

P67: manually verify the output of an activation function in a neural network against expected value.

P114: check that one-hot encoding worked

\subsubsection{Model training (O8, O31, O42, P77)}

O8, P77: Periodically print the progress made in training by printing the training loss/accuracy of model.

O31: check the model fits the data without any errors; this is done in the model experimentation phase, where several models are tested and the best one based on the selection criteria is picked

\subsection{Explicit feedback using \texttt{assert} statements}

% TODO: find out more about batch training
\subsubsection{Batch size check (A21, A28, A70)}

Neural Networks are trained using smaller batches that makes it more efficient use of hardware GPU cores. During batch training the weights are not updated with every example but with the average of the entire batch, this ensures a more optimal and smoother learning curve and generally leads to better generalizability.

When trained using batches, we need to take this into account when calculating the performance of the model. We need to calculate the accuracy over multiple batches and then average the results.

We encounter assertions that check the size of the batch prior to training. This is to ensure that the batch size matches the hardware specifications? We should not have a batch size that does not fit in the memory of GPU/CPU.

\subsubsection{Data leakage check (A33)}

It is a best practise to ensure that the ML model is not tested using examples that it has already encountered during training. This is to ensure that the model is not overfitting on the training data and is able to generalise to examples similar to the distribution of the training set, but not exactly the same. In A33 we see that the assert ensures that the training and validation sets do not share any examples.

\subsubsection{Dependency check (A18, A67)}
% TODO: what is the motivation of this assertion? Find prior work on python dependency management?

This is a unique type of test which occurs due to the unique workflow of notebooks. Typically dependencies are managed through external means such as dedicated dependency management solutions or the builtin Python method using a requirements.txt file. We notice assertions that check that the major version of an external dependency is equal to the specified value.

% NOTE: missing-value-check and existence-check

\subsubsection{Existence check (A86)}

Existence checks are carried out in various contexts. Most are for checking that columns in a dataset do not contain missing values. Others ensure that certain columns (such as the column containing the labels of the data) exists in the dataset prior to splitting the data into training and testing sets. Existence checks are frequently performed after data pro-processing  steps are performed to ensure that missing values were not added to the dataset.

% TODO: we need to elaborate this one with examples

\subsubsection{Mathematical property checks (A3, A25, A56, A64)}

Machine learning models are often rooted in statistical models. Implementing neural networks requires a grasp of linear algebra and matrix multiplication. We find several assertions that test the mathematical properties of arrays and matrices after performing matrix operations.

\subsubsection{Model performance check (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)}

We see several assertions that test for the performance of the trained ML model against a validation or test test. The assertions often compare a performance metric (such as accuracy, precision, recall, F1, etc) against a predefined threshold.

% TODO: this needs more work and investigation
\subsubsection{Network architecture check (A11)}

We only have one example here.

% TODO lots to elaborate and discuss here
\subsubsection{Resource check (A10, A14, A37, A60, A74)}

We find assertions that check that a pre-trained model exists on the file system, or that data can be loaded from a specified file path. There are also checks that ensure that a visualisation created using matplotlib contains data in it.

\subsubsection{Data shape check (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)}

``Swiss army knife'' of assertions!

\subsubsection{Type check (A2, A35, A40, A81, A88)}

Checking the data type of features and such.

\subsubsection{Value check (A30, A41, A44--46, A52, A65, A68, A69, A73, A92)}

Check that the values in a column are in the specified values (categorical) or in a specified range (normalised continuous variable) or binary (such as the label column).

% NOTE: blanks and unit-tests
\subsubsection{Miscellaneous checks (A6, A20, A32, A34, A36, A53, A62, A75, A83, A87, A97)}

These feel like unit tests, performed in lieu of writing full test suites in the notebook.

\bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}
%\begin{acknowledgements}
%If you'd like to thank anyone, place your comments here
%and remove the percent signs.
%\end{acknowledgements}

% Authors must disclose all relationships or interests that 
% could have direct or potential influence or impart bias on 
% the work: 
%
% \section*{Conflict of interest}
%
% The authors declare that they have no conflict of interest.

% BibTeX users please use one of
\bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
\bibliography{bibliography}   % name your BibTeX data base

\end{document}
% end of file template.tex

### 2024-05-07 report/emse24.tex
feat(report): emse24 updates

@@
 \begin{abstract}
-Insert your abstract here. Include keywords, PACS and mathematical
-subject classification numbers as needed.
-\keywords{First keyword \and Second keyword \and More}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
+
+  % TODO update the numbers here
+  % TODO: update
+Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
+In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
+Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
+However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
+We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
+We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python.
+We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
+We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
+We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns.
+We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part.
+Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML.
+And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.
+
+\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}
 \end{abstract}
@@
 
-\section{Methodology}
-% For one-column wide figures use
+% amershi2019software: start with ML development life cycle and how everything is done in a very iterative way; things are dependant on one another; etc.
+
+% martinez-plumed2021crisp-dm: not only does ML include the crip-dm framework, but this same form of iteration is seen in the model development and experimentation phase;
+
+% this is why we see the use of computational notebooks being widely adopted by the ML community
+
+% writing ML code is more iterative and requires a constant feedback; we identify this feedback can be provided in two ways: implicitly from the output of a cell and explicitly from a failed assertion
+
+Code, data and ML model in ML-enalbed systems are highly tangled with each other.
+For instance, the decision to use a particular type of ML model, requires a deep understanding of the data which will be used to train the model.
+Data quantity and quality directly influence the complexity of the ML model.
+
+Conversely, the choice of the ML model also influences the data pre-processing and transformation steps that are required to get the data ready for training.
+For instance, using a classic statistical model requires several data transformation steps such as scaling and normalization, and converting categorical features to numerical ones using one-hot encoding.
+
+Visualisations are used throughout the ml development lifecycle to test various properties of the ml system.
+in the early stages of the ml development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
+during the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ml models.
+once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
+
+building ml systems is highly iterative and experimental.
+information gained from ml visualisations is used to make design and implementation decisions for the following steps of the ml pipeline.
+for instance, we may visualise the distribution of our training data and find that it is normally distributed.
+based on this information, we may opt for a linear regression model which assumes normality in the underlying data.
+however, such expectations regarding the data may be violated once the ml system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
+
+implicit expectations obtained from visualisations must therefore be
+translated to analytical tests that fail once our expectations
+regarding the ml system are no longer satisfied.
+
+in contrast to prior work, we approach ml testing from a new
+perspective. we conduct an empirical analysis of computational
+notebooks obtained from github, to understand the process of testing
+ml systems in practice. in particular, we focus on the combination of
+a qualitative form of testing (using visualisations) and a
+quantitative form of testing (using analytical assertions). the
+research questions along with the contributions of this paper are as
+follows.
+
+\begin{description}
+  \item[rq1.] \textbf{how frequently are analytical tests formulated
+    from visualisations created to test ml systems?}
+
+    we mine 54k jupyter notebooks from github that contain an assertion
+    written in python. we develop an automated technique to identify 1764
+    notebooks with an assertion below a visualisation. we manually analyse
+    the 1.7k notebooks and identify 269 visualisation-assertion (va) pairs
+    that are semantically related to each other. we plan to release this
+    catalogue of va pairs publicly.
+
+\item[rq2.] \textbf{what patterns are frequently observed in va pairs
+  used to test ml systems?}
+
+  we analyse the 269 va pairs from rq1, and observe three testing
+  patterns that frequently occur in the va pairs.
+
+\item[rq3.] \textbf{do the assertions capture all the information
+  obtained from the corresponding visualisation?}
+
+  we further perform an in-depth analyse of 34 va pairs (and their
+  corresponding notebooks). we observe that assertions often cannot
+  capture all the information obtained from the visualisation.
+
+\end{description}
+
+our observations indicate that formulating analytical assertions from
+visualisations is an emerging testing technique. however, the
+assertions found in this study often do not reflect the information
+obtained from their visual counter-part. this study finds the need for
+automated tools that can reduce the manual effort necessary to
+validate visualisations in ml systems. and help ml practitioners to
+formulate better analytical assertions from visualisations.
+
+the replication package for this study is made public on
+figshare\footnote{the replication package can be found at this url:
+https://figshare.com/s/9347228011999cba29f5}.
+
+\section{Preliminaries}\label{sec:prelim}
+
+This section summarises prior scientific work that has been done in ML
+Testing, computational notebooks and visual analytics. The section
+concludes with an overview of technical knowledge required for this
+paper.
+
+\subsection{ML Testing}\label{sec:ml-testing}
+
+Testing ML enabled software systems poses more challenges compared to
+their traditional counter part. While traditional software systems
+mature through change in their codebase, ML enabled systems
+additionally experience change in the training data and the machine
+learning
+model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
+With ML being adopted into safety-critical domains that can affect
+human lives, ensuring that ML systems are correct, robust to data
+perturbations and not biased by race, colour or gender is of paramount
+importance. Existing scientific literature on ML testing broadly
+focuses on two aspects. First, on functional properties such as
+correctness and robustness of the model towards unseen data. And
+second, on non-functional properties such as fairness,
+interpretability and
+privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
+
+To test the correctness of ML models, several improvements over
+existing test adequacy metrics have been proposed. Tools such as
+\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
+metrics such as neuron coverage adapted for ML enabled software
+systems~\cite{pei2017deepexplore, ma2018deepgauge,
+  gerasimou2020importance}. Formal verification methods have also been
+proposed that try to provide formal guarantee of robustness against
+adversarial examples. Such methods however are only feasible for
+statistical ML models and become computationally expensive for more
+complex models such as deep neural networks~\cite{zhu2021deepmemory,
+  baluta2021scalable}. Several works have been conducted on generating
+and detecting adversarial inputs for ML models. Data augmentation
+techniques based on fuzzing, search based software testing and
+mutation testing have been proposed to generate adversarial examples
+that can be used during model training to improve its
+robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
+  zhang2020white}. It is however not possible to include all
+variations of adversarial examples into the training data. Thus,
+methods have been proposed to detect adversarial inputs during
+runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
+  berend2020cats}.
+
+In contrast to existing scientific contributions, we conduct a
+large-scale empirical analysis of Jupyter Notebooks to understand the
+process of testing ML systems. Specifically, we focus on how
+visualisations are used to test specific properties of ML systems. We
+further investigate how frequently analytical tests are formulated
+based on the information gained from the visualisations.
+
+\subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
+
+Computational notebooks have been ubiquitously adopted by the machine
+learning community for developing ML enabled systems. Although
+originally intended to promote reproducible software, computational
+notebooks are far from being reproducible due to lack of software
+engineering best practices such as separation of concern, testing and
+versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.
+
+\emph{Psallidas et al} provide an overview of the evolving landscape
+of computational notebooks by analysing six million Python notebooks,
+two million enterprise Data Science (DS) pipelines, source code and
+metadata from over 900 releases of 12 important DS libraries. Their
+findings can be used by system builders for improving DS tools and
+also by DS practitioners to understand the current trends in
+technologies they should focus on. \emph{Pimentel et al} mined 1.4
+million notebooks from Github to conduct an empirical study on the
+coding practices in computational notebooks. Based on their analysis,
+the authors propose guidelines on improving reproducibility of
+computational notebooks. To enable future research on computational
+notebooks, \emph{Quaranta et al} mine Kaggle and present
+\textit{KGTorrent}, a public dataset consisting of approximately 2.5
+million Jupyter notebooks written in Python. The dataset also contains
+a relational database dump of metadata regarding publicly available
+notebooks on Kaggle.
+
+Studies with human subjects have been conducted to gain a deeper
+understanding of the challenges faced by ML practitioners when
+developing ML models inside notebooks. The results indicate that ML
+practitioners work in a highly iterative fashion, often experimenting
+with multiple strategies to analyse the data or produce meaningful
+visualisations~\cite{kandel2012enterprise, kery2018story,
+  liu2019understanding, chattopadhyay2020wrong}. Studies have also
+been conducted to understand how practitioners generate, evaluate and
+manage alternative hypothesis, visual designs, methods, tools,
+algorithms and data sources to arrive at the final
+implementation~\cite{liu2019understanding,kandel2012enterprise}. The
+findings from these studies can be used as guidelines for improving
+existing notebook technologies or designing new ones.
+
+To manage and prune multiple versions of code that accumulate over
+time when developing in notebooks, \emph{Head et al} propose a code
+gathering tool that allows practitioners to review and only keep the
+relevant version of code. Other tools such as \textit{WrangleDoc} and
+\textit{VizSmith} have also been proposed to aid ML practitioners when
+working within computational notebooks~\cite{yang2021subtle,
+  bavishi2021vizsmith}.
+
+Jupyter notebooks have been widely adopted by the DS and ML
+communities to develop ML
+pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Computational notebooks provide a rich source of data in the form of
+natural text, code and visualisations. Computational notebooks also
+allow ML practitioners to present the knowledge gained from the
+analysis in a narrative that can be shared with
+others~\cite{rule2018exploration}. This study leverages the
+computational narrative present within notebooks to identify
+visualisations and assertions that are semantically linked to each
+other.
+
+\subsection{Visualisations and Machine Learning}\label{sec:visualisations}
+
+As ML augment software systems in safety-critical domains, emphasis
+has been put into explainability of ML models. ML models and the
+underlying data is complex and multi-dimensional. To combat the
+``curse of dimensionality'', visual analytics has been widely adopted
+by the ML community to understand the data and the internal workings
+of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.
+
+Prior studies have been conducted to understand how visual analytics
+techniques are currently being used in ML. \emph{Yuan et al} conduct a
+systematic review of 259 papers and propose a taxonomy of visual
+analytics techniques for ML. \emph{Hohman et al} conduct a survey of
+visual analytics techniques for Deep Learning Models. The findings of
+the study indicate that visual analytics has been widely adopted in ML
+for model explanation, interpretation, debugging and improvement.
+
+Several tools have been proposed by the visual analytics research
+community to aid practitioners in understanding how their ML models
+operate. \emph{Wexler et al} propose \textit{What-If}, a visual
+analytics tool to explore alternative hypotheses, generate
+counterfactual reasoning, investigate decision boundaries of the model
+and how change in data affects the model predictions. To reduce the
+cognitive load of ML practitioners during model building phase,
+\emph{Amershi et al} propose \textit{ModelTracker}. Given a trained
+model and a sample dataset, ModelTracker presents all the information
+in traditional summary statistics along with the performance of the
+model.
+
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
+\textit{Drava} are a few other tools that have been proposed to handle
+specialised use-cases such as identifying systematic errors in ML
+models, generating counterfactual explanations for Generalised
+Additive Models, prioritising machine translation model improvements
+and relating human concepts with semantic dimensions extracted by ML
+models during disentangled representation learning\cite{ahn2023escape,
+  wang2023gam, robertson2023angler, wang2023drava}.
+
+In contrast to prior work which propose dedicated visual analytics
+tools, this study focuses on visualisations within computational
+notebooks. We focus on visualisations produced by AI practitioners
+using Python libraries (such as Matplotlib or Seaborn). Additional
+context provided by surrounding code and markdown cells are used to
+understand the motivation for creating the visualisation and the
+information gained from it.
+
+\section{Study Design}
+
+% TODO: introduce the motivation of the study along with the RQs here
+
+% role of feedback in writing ML code? (any recommendation for papers I can refer to here?)
+% implicit vs. explicit feedback (how do we establish this concept?)
+
+\subsection{Data Collection}\label{sec:data-collect}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{esem24-method.pdf}
+  \includegraphics[width=\linewidth]{data-collection.pdf}
   \label{fig:data-collection}
@@
 \end{figure}
+
+Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.
+
+We mined public repositories from Github to collect $49923$ Jupyter notebooks.
+We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them.
+The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
+
+For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github.
+To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.
+
+Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.
+
+We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}.
+Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
+
+For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells.
+We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.
+
+Similarly, for implicit feedback, we first isolate code cells that produce an output.
+The output of a cell may either be from a ``print'' statement or the last statement of the cell.
+We collect both of them using the AST of the cell.
+
+Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.
+
+\subsection{Case Studies}
+
+% TODO: might need a visual aid here
+
+We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs).
+To surface interesting candidates for the case studies, we apply NLP techniques as described below.
+
+% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
+Thats why we need to apply these tecniques to surface unique and interesting candidates!
+
+The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed.
+Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
+The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens thay appear less frequently are assigned a higher value.
+
+We apply stratified random sampling to identify the candidates for the case study analysis.
+The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
+The candidates are then divided into quartiles based on the aggregate value.
+A candidate is randomly drawn from each bin and analysed as an in-depth case study.
+The analysis is stopped when the time resource is exhausted.
+
+During each case study, we analyse the code of the candidate to understand its purpose.
+Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.
 
@@
 
-\subsection{Implicit feedback using \texttt{print} statements and output of code cells}
+\subsection{Implicit feedback from the output of code cells}
 
-\subsubsection{Data distribution check (O2, O4, O9, O14, O20, O25, O48)}
+\subsubsection{Data distribution check ($N = 7$)}
+% (O2, O4, O9, O14, O20, O25, O48)
+
+% we need to know the distribution of the data to make informed decisions regarding the data transformation steps required (for instance, do we need to scale and normalise?
+do we have outliers?); visualisations are a cheap and quick means of doing this
+
+% distribution check is also required since models such as linear regression, perform well when the data is normally distributed; if our data is not normal then we need to perform additional transformation steps to make the data ready for training
+
 % a mix of visualisations and pandas dataframes that show the distribution of a column in the data
@@
 
-\bibliographystyle{ACM-Reference-Format} \bibliography{bibliography}
 %\begin{acknowledgements}
@@
 % BibTeX users please use one of
-\bibliographystyle{spbasic}      % basic style, author-year citations
-%\bibliographystyle{spmpsci}      % mathematics and physical sciences
+% \bibliographystyle{spbasic}      % basic style, author-year citations
+\bibliographystyle{spmpsci}      % mathematics and physical sciences
 %\bibliographystyle{spphys}       % APS-like style for physics

### 2024-05-13 report/emse24.tex
feat(report): 2024-05-13 emse24 updates

@@
 %
-\documentclass{svjour3}                     % onecolumn (standard format)
+% \documentclass{svjour3}                     % onecolumn (standard format)
 %\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-% \documentclass[smallextended]{svjour3}       % onecolumn (second format)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
 %\documentclass[twocolumn]{svjour3}          % twocolumn
@@
 \usepackage{xcolor}
+\usepackage{subcaption}
+\usepackage{tabularx}
+\usepackage{booktabs}
+\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
+\usepackage{wrapfig}
 
-\lstset{
+\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
+\colorlet{punct}{red!60!black}
+\definecolor{delim}{RGB}{20,105,176}
+\colorlet{numb}{magenta!60!black}
+
+\lstdefinestyle{mystyle}{
   language=Python,
-    basicstyle=\ttfamily,
-    keywordstyle=\color{blue},
-    stringstyle=\color{red},
-    commentstyle=\color{green},
-    morecomment=[l][\color{magenta}]{\#}
+  basicstyle=\ttfamily\footnotesize,
+  keywordstyle=\color{delim},
+  stringstyle=\color{punct},
+  commentstyle=\color{numb},
+  breakatwhitespace=true,
+  breaklines=true,
+  keepspaces=true,
+  showspaces=false,
+  showstringspaces=false,
+  showtabs=false,
+  tabsize=2,
+  stepnumber=1,
+  numbersep=3pt,
+  captionpos=b,
+  numbers=left,
+  numberstyle=\tiny\color{gray},
+  backgroundcolor=\color{backcolor},
+  aboveskip=0pt,
+  belowskip=0pt
 }
+\lstset{style=mystyle}
+
+\lstset{emph={%  
+    assert%
+    },emphstyle={\color{delim}}%
+}%
 
@@
 
-  % TODO update the numbers here
-  % TODO: update
+  \todo{update the numbers here}
 Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
@@
 
-% TODO: introduce the motivation of the study along with the RQs here
+\todo{TODO: introduce the motivation of the study along with the RQs here}
 
@@
 
+\todo{TODO: justify why we didn't collect failures (strerr); that is more debugging? not the focus of this paper?}
 Similarly, for implicit feedback, we first isolate code cells that produce an output.
@@
 
-% TODO: might need a visual aid here
+\todo{TODO: might need a visual aid here}
 
@@
 
-% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
-Thats why we need to apply these tecniques to surface unique and interesting candidates!
+\todo{TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
+Thats why we need to apply these tecniques to surface unique and interesting candidates!}
 
@@
 
-% TODO: for each of the feedback types; start with the descriptive and lexical analysis we performed
-% TODO: then move into the case studies, and link them to existing literature; use the fault and failure taxonomies; perhaps also link the feedbacks to ML lifecylce proposed by Amershi 2019?
-
 \subsection{Implicit feedback from the output of code cells}
@@
 
-% we need to know the distribution of the data to make informed decisions regarding the data transformation steps required (for instance, do we need to scale and normalise?
-do we have outliers?); visualisations are a cheap and quick means of doing this
+\begin{table}
+\centering
+\caption{Examples of cell outputs used to verify distribution of data.}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.3\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}&
+\emph{\textbf{Description}}\\
+\midrule
 
-% distribution check is also required since models such as linear regression, perform well when the data is normally distributed; if our data is not normal then we need to perform additional transformation steps to make the data ready for training
+O14&
+\begin{lstlisting}
+pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])
+\end{lstlisting}&
+Check the mean value of the specified features for each label in the target feature.\\
 
-% a mix of visualisations and pandas dataframes that show the distribution of a column in the data
-% we primarily observe these checks being used to determine the relationship of a feature with the target feature
-% typically in the EDA/data exploration/understanding phase
-% several examples where the distribution of a newly created feature is checked (binning a continuous feature into categorical one)
-% one uses descriptive statistics to verify that data normalisation worked (again, a check after data transformation step)
+O48&
+\begin{lstlisting}
+x_train.describe()
+\end{lstlisting}&
+Check the descriptive statistics of the training data. The above statement is written immediately after normalising the features in the training data.\\
 
-We observe the use of visualisations and dataframes to check the distribution of a feature.
-Often, the analysis is performed in the context of a multivariate analysis, where the distribution of a feature is checked against all possible outcomes of the target feature.
+\bottomrule
+\end{tabular}
+\label{tab:distribution-check}
+\end{table}
 
-Typically, this check is performed in the EDA phase where the author is trying to understand the relationship of a feature with the target feature.
-This done to identify features that can be used to train the ML model.
+\begin{figure*}
+\centering
+\subcaptionbox{Check the distribution of the feature \texttt{Age} with-respect-to the labels in the target feature \texttt{Survived}.\label{fig:distribution-check-kdeplot}}{\includegraphics[width=0.45\linewidth]{distribution-check-kdeplot.png}}
+\subcaptionbox{Check the distribution of the feature \texttt{likes} across all values of feature \texttt{category\_id}.\label{fig:distribution-check-catplot}}{\includegraphics[width=0.45\linewidth]{distribution-check-catplot.png}}
+\caption{Examples of visualisations used to check linear relationship in data.}
+\label{fig:distribution-check}
+\end{figure*}
 
-We also observe that the distribution of a feature is checked directly after creating it.
-For example, we observe this check after converting a continuous feature into a categorical feature using binning.
-Another example is the use of descriptive statistics provided by pandas to check that the prior data normalisation step worked (the author manually verified the mean and standard deviation of the feature from the output).
+Understanding the distribution of data is crucial for making informed decisions regarding necessary data transformation steps in data analysis processes.
+For instance, identifying whether scaling, normalizing, or handling outliers is required can be efficiently and effectively determined through visualizations.
 
-\subsubsection{Data relationship check (O6, O10)}
+% This is particularly important for models like linear regression, which perform optimally when data is normally distributed.
+If the data deviates from normality, additional transformation steps may be necessary to prepare it for training.
 
-Linear regression models assume a linear relationship between the features in the training data, with the target feature.
-We see this being verified using lineplots to check the presence of a linear trend between a given feature and the target feature.
+During the exploratory data analysis (EDA) or data understanding phase, a combination of visualizations and pandas dataframes is often employed to assess the distribution of specific columns within the data.
+This step helps identify features that may have a relationship with the target variable, ultimately informing which features to include in the machine learning model training process.
+This is seen in Figure~\ref{fig:distribution-check-kdeplot} where the distribution of a feature \texttt{Age} is checked against the labels of the target feature.
+Figure~\ref{fig:distribution-check-catplot} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
 
-A special case is O10 which uses a regplot. This is a visualisation provided by the seaborn library that plots the data along with a linear regression model fit.
+For example, the distribution of a newly created feature, such as one where a continuous feature is binned into categories, is frequently examined.
 
-\subsubsection{Dependency check (P71, P86)}
+Descriptive statistics are also used post-transformation to verify that steps like data normalization have been effectively applied, ensuring that the data conforms to the expected format for further analysis.
 
-We find evidence of lack of software engineering best practise.
-Dependencies should be manage with external programs built for this task.
-Python has a built-in functionality for this, where dependencies along with version can be specified in a requirements.txt file.
-There are also several external programs such as the use of lockfiles by pipenv and poetry which lock a specific version of a dependency such that all files are reproducible on all systems.
+Furthermore, distribution checks are often performed immediately after feature creation.
+For instance, this may occur after discretizing a continuous feature using binning.
+Additionally, descriptive statistics (e.g., mean and standard deviation) obtained from libraries like pandas can be used to verify the effectiveness of data normalization steps.
 
-P71 prints the version of an external library which the author is expected to verify manually.
-P86 prints a confirmation message if the Cuda library is available on the system where the notebook is being executed.
+\subsubsection{Data relationship check ($N = 2$)}
+% (O6, O10)
+\todo{TODO: I think correlation checks should also be present in this group}
 
-\subsubsection{Execution time check (P66)}
+\begin{figure*}
+\centering
+\subcaptionbox{Using a lineplot to check linear relationship between two features in a dataset.\label{fig:linear-relation-check-lineplot}}{\includegraphics[width=0.45\linewidth]{linear-relation-check-lineplot.png}}
+\subcaptionbox{Visualisation that shows the datapoints as a scatterplot along with a linear model fit.\label{fig:linear-relation-check-regplot}}{\includegraphics[width=0.45\linewidth]{linear-relation-check-regplot.png}}
+\caption{Examples of visualisations used to check linear relationship in data.}
+\label{fig:linear-relation-check}
+\end{figure*}
 
-Often several ML models are trained and for a single task.
-The selection is based on requirements that not only consider the performance of the model, but also the total time required to train the model.
-Since this is an iterative task, a model that can be more efficiently retrained, helps reduce costs in the long run, and has been proven to be more sustainable.
+Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
+However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
+Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.
+Figure~\ref{fig:linear-relation-check} illustrates two methods for identifying these linear relationships.
+In Figure~\ref{fig:linear-relation-check-lineplot}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
+Figure~\ref{fig:linear-relation-check-regplot} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
 
-P66 is from a notebook where several models are trained for the same task, and their performance along with the time taken to train is printed.
-The cells of a notebook are also executed several times, for a long running cell (such as the ones that train the model), an indication of the total execution time helps (after the first run, the author has a baseline for the time it will take).
+\subsubsection{Resource check ($N = 7$)}
+% (P71, P86, O64, O66, P68, P82, P107)
 
-\subsubsection{Missing value check (O12, O36, P74)}
+\begin{table}
+\centering
+\caption{Examples of cell outputs used to verify if different types of resources are available on the system where the notebook is being executed.}
+\begin{tabular}{@{}m{0.1\textwidth} m{0.05\textwidth} m{0.4\textwidth} m{0.3\textwidth}@{}}
+\toprule
+\emph{\textbf{Type of resource}}&
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}&
+\emph{\textbf{Description}}\\
+\midrule
 
-Check if missing values are present in any features of the data. After loading the dataset and in the EDA phase of analysis.
+External library&
+P71&
+\begin{lstlisting}
+print('Hub version: ', hub.__version__)
+\end{lstlisting}&
+Check the version of an external Python library installed on the current system.\\
 
-From our data smells paper: descriptive statistics functions typically ignore missing values; leading to biased and incorrect results and conclusions
+Compute resources&
+P68&
+\begin{lstlisting}
+print('GPU is available')
+\end{lstlisting}&
+Check if a GPU is available on the current system.\\
 
-ML models also except numerical values, a null type value raises an exception (since computations cannot be performed)
+Data or pre-trained models&
+O66&
+\begin{lstlisting}
+prostate_cancer_df.shape
+\end{lstlisting}&
+Check that the dataset has been loaded into memory.\\
+\bottomrule
+\end{tabular}
+\label{tab:resource-check}
+\end{table}
 
-P74 and O36 is an aggregated check over the entire dataset.
+\todo{perhaps also add the keys of the case-studies here?}
+Table~\ref{tab:resource-check} presents case studies where the output is used to verify that certain resources are available on the system in which the notebook is being executed.
+We find checks for the availability of compute resources (such as a GPU or a TPU), checking that the dataset or a pre-trained model has been loaded correctly and checks to ensure that a certain version of an external library is present on the system.
 
-% TODO: highlight the use of visualisations to check for missing values; what is the benefit of using a visualisation instead of code?
-More interesting is O12 which uses a heatmap to show the distribution of the missing values across all features of the dataset.
+P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
+This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
+In this case study, managing dependencies through external tools is recommended.
+Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
+Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
 
-\subsubsection{Model performance check (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)}
+\subsubsection{Execution time check ($N = 1$)}
+% (P66)
+
+\begin{lstlisting}
+print('Total Run Time:')
+\end{lstlisting}
+
+While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
+Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run.
+The provided code snippet (Listing~\ref{lst:exec-time-check}) exemplifies this concept by showcasing a case study (P66) that tracks training times for various models using a grid search.
+This approach allows practitioners to compare model performance alongside training efficiency.
+
+\subsubsection{Missing value check ($N = 3$)}
+% (O12, O36, P74)
+
+Checking for missing values is a critical step in the preprocessing phase of a machine learning project because it significantly influences the quality and performance of the model.
+Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
+Many machine learning algorithms, including those that involve matrix multiplication operations such as linear regression and neural networks, demand complete numerical datasets to perform calculations.
+Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
+
+\begin{table}
+\centering
+\caption{Examples of cell outputs used to check for missing data.}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.6\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+P74&
+\begin{lstlisting}
+print(train_df.isnull().sum())
+\end{lstlisting}\\
+
+O12&
+\begin{lstlisting}
+sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')
+\end{lstlisting}\\
+
+O36&
+\begin{lstlisting}
+test.isna().sum().unique()
+\end{lstlisting}\\
+\end{tabular}
+\label{tab:missing-value}
+\end{table}
+
+Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
+While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
+
+\subsubsection{Model performance check ($N = 33$)}
+% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
+
+\begin{table}
+\centering
+\caption{Various cell outputs used to check the performance of an ML model after training.}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.6\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+P3&
+\begin{lstlisting}
+print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')
+\end{lstlisting}\\
+
+P6&
+\begin{lstlisting}
+print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))
+\end{lstlisting}\\
+
+P18&
+\begin{lstlisting}
+print('The Accuracy is:', accuracy_score(y_test, y_pred))
+\end{lstlisting}\\
+
+P18&
+\begin{lstlisting}
+print('Classification Report: SVM (validation data)')
+\end{lstlisting}\\
+
+P54&
+\begin{lstlisting}
+print('Intercept value:', lm.intercept_)
+\end{lstlisting}\\
+\end{tabular}
+\label{tab:model-perf}
+\end{table}
 
@@
 
-Mean accuracy over K-fold validation
+Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
+Besides \emph{accuracy}, we also see checks for the \emph{Root Mean Square Error (RMSE)} and the use of the classification report provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html}.
+In addition to the accuracy, the classification report also reports the \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
 
-Other performance metrics such as recall, precision and F1 also printed
+\emph{P54} prints the intercepts of a linear regression model.
+Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
+This insight is critical for accurate model interpretation and effective decision-making in practical applications.
 
-Error in predictions (RMSE, MSE)
+\begin{wrapfigure}{l}{0.45\linewidth}
+\includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}
+\caption{Confusion matrix}
+\label{fig:model-perf-confusion-matrix}
+\end{wrapfigure}
 
-P48: prints the loss
-
-% TODO: whats the point of this?
-P54: print the intercept of linear regression model
-
-Classification report is used to produce a summary of popular performance metrics such as accuracy, recall, precision and F1.
-
-Precision and recall is checked using the confusion matrix.
-
-% TODO: highlight the need for normalising vs. comparing absolute values; the data might be imbalanced so does not make sense to compare the absolute values
-O3: use heatmap to visualise the confusion matrix of a multi-label classification task.
+Figure~\ref{fig:model-perf-confusion-matrix} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
 The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
 
-O52: author wrote a custom function which gets the predictions of an image classifier (developed in the NB) on a random sample of images.
-The predictions along with the input image is printed.
-The author is manually spot-checking the accuracy of the predictions.
-Or perhaps trying to find examples that produce the wrong prediction?
+% TODO: this looks horrible if we are wrapping text around the figure above
+% \begin{lstlisting}[caption={Spot-check the performance of an image classification model for a random sample of input images.}, label={lst:model-perf-custom}]
+% def spot_check_recs(classifier, seed):
+%     random_images = []
+%     random_labels = []
+%     random_loc = np.random.randint(3000)
+%     distance, loc = classifier.kneighbors( embedding_model.predict( visualize[random_loc].reshape(-1,28,28)), 100)
 
-\subsubsection{Model architecture (P92)}
+%     random_images.append(visualize[random_loc])
+%     ...
+%     for rec_loc, dist in zip(loc.reshape(-1,1), distance.reshape(-1,1)):
+%         ...
 
-Several neural network architectures are available as part of the ML library (for instance keras, pytorch and tensorflow come pre-loaded with several NNs).
+%     ImageUtil.grid_display(random_images, random_labels, no_of_columns=1, figsize=(2,2), fig_title = "recs")
 
-Print the architecture of the neural network. The layers of the model, neurons in each layer and the activation functions are printed in a text diagram.
+% spot_check_recs(classifier, 910)
+% \end{lstlisting}
 
-\subsubsection{Resource check (O64, O66, P68, P82, P107)}
+% Listing~\ref{model-perf-custom} shows a function implemented by a practitioners to manually evaluate an image classification model on a random sample of input images.
+The function prints the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
 
-Check is GPU or TPU resources are available. We suspect these checks are coming from Google Colab notebooks where computation can be off-loaded to remote hardware.
+\subsubsection{Neural network architecture ($N = 1$)}
+% (P92)
 
-In P107, check that the pre-trained model was successfully loaded from the file system.
+\begin{lstlisting}[caption={Print the architecture of a neural network.}, label={lst:network-architecture}]
+print(MyNetwork)
+\end{lstlisting}
+
+Listing~\ref{lst:network-architecture} shows a print statement to show the architecture of a neural network.
+Printing the structure of a neural network serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
+It helps in verifying configurations like layer dimensions and operations before initiating computationally intensive training, ensuring the model is set up efficiently for the intended tasks.
+This practice is fundamental for maintaining accuracy, facilitating collaboration, and enhancing the reproducibility of the model.
 
@@
 
-% TODO: why is shape check important?
+\todo{TODO: why is shape check important?}
 Checking the shape of the data is important for several reasons. If we perform some data pre-processing or transformation, we want to ensure that the shape of the data is the same as expected.
@@
 
-% TODO: find out more about batch training
+% TODO: bring in the notion of pre and post condition checks (the wikipedia page on assertions is very interesting: https://en.wikipedia.org/wiki/Assertion_(software_development)); specifically the part on run-time assertions
+
+\todo{TODO: find out more about batch training}
 \subsubsection{Batch size check (A21, A28, A70)}
@@
 \subsubsection{Dependency check (A18, A67)}
-% TODO: what is the motivation of this assertion? Find prior work on python dependency management?
+\todo{TODO: what is the motivation of this assertion? Find prior work on python dependency management?}
 
@@
 
-% NOTE: missing-value-check and existence-check
+\todo{NOTE: missing-value-check and existence-check}
 
@@
 
-% TODO: we need to elaborate this one with examples
+\todo{TODO: we need to elaborate this one with examples}
 
@@
 
-% TODO: this needs more work and investigation
+\todo{TODO: this needs more work and investigation}
 \subsubsection{Network architecture check (A11)}
@@
 
+\subsection{Assertions derived from visualisations}
 \end{document}

### 2024-05-13 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsubsection{Shape check (P4, P32, P117)}
+\subsubsection{Shape check ($N = 3$}
+% (P4, P32, P117)
 
-\todo{TODO: why is shape check important?}
-Checking the shape of the data is important for several reasons. If we perform some data pre-processing or transformation, we want to ensure that the shape of the data is the same as expected.
+Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
 
-Number of features in the training and testing data must match.
-When using statistical ML models, the model "fits" the training data and thus expects the same shape in the testing data to make inference.
-When developing neural networks, architectural decisions such as the number of input neurons is made based on the shape of the data.
+Primarily, the number of features in the training set must match those in the testing set.
+This alignment is crucial because statistical machine learning models are trained on specific data dimensions and expect the same dimensional structure during inference to perform accurately.
+Similarly, in neural network architectures, the configuration of input layers depends directly on the shape of the training data, dictating the number of input neurons needed.
+Furthermore, the correspondence between the number of test examples and their respective labels is essential for accurately computing performance metrics, such as accuracy.
 
-During the testing shape, the number of examples in the test input must be the same as the number of labels so that the performance metrics such as accuracy, can be correctly calculated.
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}} & \emph{\textbf{Code}}\\
+\midrule
+P4 &
+\begin{lstlisting}
+print('no.of examples in test data : ', len(test_data))
+\end{lstlisting}\\
+P32 &
+\begin{lstlisting}
+print('Training set shape : ', x_train.shape) 
+\end{lstlisting}\\
+\end{tabular}
+\caption{Example of outputs used to check the shape of the data.}
+\label{tab:shape-check}
+\end{table}
 
-\subsubsection{Type check (O71, P43)}
+Table~\ref{tab:shape-check} shows two print statements discovered in this study which demonstrate routine checks performed by developers to ensure the integrity and applicability of their models.
+Such checks are indicative of the meticulous attention to detail necessary in the model training and validation process, underscoring the importance of shape validation in achieving reliable machine learning outcomes.
 
-All ML models expect the input data to be entirely numerical, so that mathematical operations such as matrix multiplications may be carried out without errors.
+\subsubsection{Type check ($N = 2$)}
+% (O71, P43)
 
-Checking the data type in the EDA phase may be done to understand the optimal data pre-processing or transformation techniques necessary, to make the data ready for model.
-For instance, numerical features may need to be normalised while categorical features need to one-hot encoded.
+Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis (EDA) phase of developing machine learning models.
+Since all machine learning algorithms perform mathematical operations, such as matrix multiplications, they require the input data to be entirely numerical to avoid computational errors.
+The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
+For example, while numerical features often require normalisation to bring them within a specific scale, categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
 
-\subsubsection{Value check/ Spot checks (O56, O60, P64, P67, P114)}
-% NOTE: perhaps these are spot-checks?
+This is illustrated in our analysis, where checks such as \texttt{print('data type:', images.dtype)} and \texttt{type(Y)} are employed to verify the data types of variables.
+These type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
 
-O56: check the prediction of the model on a single input image
+This proactive approach in the EDA phase facilitates smoother transitions into model training and can significantly impact the performance and effectiveness of the final machine learning models.
 
-O60: check the number of features in the data is equal to the expected number after performing PCA, the author checks the head of the dataframe and manually verifies the number of features
+\subsubsection{Value check/ Spot checks ($N = 5$)}
+% (O56, O60, P64, P67, P114)
 
-P64: find the max in the second channel of a 3D numpy array (highlight this case study; this is done for image processing!)
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
 
-P67: manually verify the output of an activation function in a neural network against expected value.
+O60 &
+\begin{lstlisting}
+X_pca.head()
+\end{lstlisting}\\
+\end{tabular}
+\caption{Example of outputs used to perform spot checks.}
+\label{tab:shape-check}
+\end{table}
 
-P114: check that one-hot encoding worked
+Table~\ref{tab:spot-check} presents outputs which were used to spot check at various stages of the ML development cycle.
+Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
+These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
+
+In \emph{O60}, the analyst verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
+This is a critical step for maintaining dimensional consistency in feature-reduced datasets.
+
+% In another case, P64, which is particularly relevant in image processing contexts, the maximum value in the second channel of a 3D numpy array is extracted and verified using np.max(cur[:, :, 1]), ensuring that the data manipulation retains its expected properties.
+
+% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
+
+% Lastly, P114 assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly as seen with print(onehot_encoded).
+Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
 

### 2024-05-13 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsubsection{Value check/ Spot checks ($N = 5$)}
+\subsubsection{Spot checks ($N = 5$)}
 % (O56, O60, P64, P67, P114)
@@
 \end{lstlisting}\\
+
+P64 &
+\begin{lstlisting}
+print(np.max(cur[:, :, 1]))
+\end{lstlisting}\\
+
+P114 &
+\begin{lstlisting}
+print(onehot_encoded)
+\end{lstlisting}\\
 \end{tabular}
 \caption{Example of outputs used to perform spot checks.}
-\label{tab:shape-check}
+\label{tab:spot-check}
 \end{table}
@@
 
+% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
+
 In \emph{O60}, the analyst verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
 This is a critical step for maintaining dimensional consistency in feature-reduced datasets.
+\emph{P64} demonstrates another case which is particularly relevant in image processing contexts.
+The maximum value in the second channel of a 3D numpy array is extracted and verified, ensuring that the data manipulation retains its expected properties.
+Lastly, \emph{P114} assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly.
 
-% In another case, P64, which is particularly relevant in image processing contexts, the maximum value in the second channel of a 3D numpy array is extracted and verified using np.max(cur[:, :, 1]), ensuring that the data manipulation retains its expected properties.
-
-% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
-
-% Lastly, P114 assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly as seen with print(onehot_encoded).
 Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
 
-\subsubsection{Model training (O8, O31, O42, P77)}
+\subsubsection{Model training ($N = 4$)}
+% (O8, O31, O42, P77)
 
-O8, P77: Periodically print the progress made in training by printing the training loss/accuracy of model.
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
 
-O31: check the model fits the data without any errors; this is done in the model experimentation phase, where several models are tested and the best one based on the selection criteria is picked
+O8 &
+\begin{lstlisting}
+autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)
+\end{lstlisting}\\
+
+O31 &
+\begin{lstlisting}
+adaBoost.fit(X_train, y_train)
+\end{lstlisting}\\
+
+O42&
+\begin{lstlisting}
+m_r.best_params_
+\end{lstlisting}\\
+\end{tabular}
+\caption{Example of outputs used to monitor model training.}
+\label{tab:model-training}
+\end{table}
+
+The ``Model Training'' phase of ML workflows involves various critical activities, each aimed at ensuring effective model performance and robustness.
+
+During training it is common practice to monitor progress periodically, as illustrated by outputs \emph{O8} in Table~\ref{tab:model-training}, where the training loss and accuracy are printed to give real-time feedback on the model's learning efficacy.
+This feedback allows for adjustments in training parameters or early stopping to prevent overfitting.
+For instance, in output O8, the autoencoder is trained using a custom callback from \texttt{keras\_utils} to provide a progress update without cluttering the output stream, which is particularly useful in lengthy training sessions.
+In contrast, output \emph{O31} represents a crucial validation step where the adaBoost model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
+Output \emph{O42} possibly checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
+
+Each of these outputs underscores different facets of the model training phase, from initialization and real-time monitoring to validation and finalization, illustrating the layered complexity of developing predictive models.
 

### 2024-05-13 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsubsection{Resource check ($N = 7$)}
+\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
 % (P71, P86, O64, O66, P68, P82, P107)
@@
 
-\todo{TODO: find out more about batch training}
-\subsubsection{Batch size check (A21, A28, A70)}
+\subsubsection{Batch size check ($N = 3$)}
+% (A21, A28, A70)
 
-Neural Networks are trained using smaller batches that makes it more efficient use of hardware GPU cores.
-During batch training the weights are not updated with every example but with the average of the entire batch, this ensures a more optimal and smoother learning curve and generally leads to better generalizability.
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
 
-When trained using batches, we need to take this into account when calculating the performance of the model. We need to calculate the accuracy over multiple batches and then average the results.
+A21 &
+\begin{lstlisting}
+assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'
+\end{lstlisting}\\
 
-We encounter assertions that check the size of the batch prior to training.
-This is to ensure that the batch size matches the hardware specifications?
-We should not have a batch size that does not fit in the memory of GPU/CPU.
+A28 &
+\begin{lstlisting}
+assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'
+\end{lstlisting}\\
 
-\subsubsection{Data leakage check (A33)}
+A70 &
+\begin{lstlisting}
+assert n_img > batch_size
+\end{lstlisting}\\
+\end{tabular}
+\caption{Assertions used to validate the batch size of input data.}
+\label{tab:batch-size}
+\end{table}
 
-It is a best practise to ensure that the ML model is not tested using examples that it has already encountered during training.
-This is to ensure that the model is not overfitting on the training data and is able to generalise to examples similar to the distribution of the training set, but not exactly the same.
-In A33 we see that the assert ensures that the training and validation sets do not share any examples.
+In the context of neural network training, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilization, particularly GPU cores.
+Batching allows for a more stable and smoother learning curve as the model's weights are updated not for each individual example, but rather based on the average gradient of a batch.
+This technique not only optimizes learning but also enhances the model's generalizability due to more robust gradient estimations.
+Given these benefits, it becomes crucial to ensure that the batch size is appropriately set to match the hardware's capacity, avoiding memory overflow issues that can halt training.
 
-\subsubsection{Dependency check (A18, A67)}
-\todo{TODO: what is the motivation of this assertion? Find prior work on python dependency management?}
+Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
+\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
+Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
+\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
 
-This is a unique type of test which occurs due to the unique workflow of notebooks.
-Typically dependencies are managed through external means such as dedicated dependency management solutions or the builtin Python method using a requirements.txt file.
-We notice assertions that check that the major version of an external dependency is equal to the specified value.
+These assertions are strategically placed to prevent common errors that could undermine training efficiency and model performance, thereby safeguarding the training process from potential disruptions due to misconfigured batch sizes.
 
-\todo{NOTE: missing-value-check and existence-check}
+\subsubsection{Data leakage check ($N = 1$)}
+% (A33)
 
-\subsubsection{Existence check (A86)}
+Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in machine learning to prevent models from overfitting.
+Overfitting occurs when a model is excessively complex, learning not only the underlying patterns but also the noise in the training data, which can degrade its performance on unseen data.
+To guarantee that the model can generalize effectively to new examples, it's crucial to verify that the training and validation sets are completely distinct with no shared examples.
 
-Existence checks are carried out in various contexts.
-Most are for checking that columns in a dataset do not contain missing values.
-Others ensure that certain columns (such as the column containing the labels of the data) exists in the dataset prior to splitting the data into training and testing sets.
-Existence checks are frequently performed after data pro-processing  steps are performed to ensure that missing values were not added to the dataset.
+This is exemplified in assertion \emph{A33}, where the code \texttt{assert len(set( tr\_df.PetID.unique()).intersection(valid\_df.PetID.unique())) == 0} checks for this exact condition.
+This assertion ensures that there are no overlapping \texttt{PetID} values between the training set (\texttt{tr\_df}) and the validation set (\texttt{valid\_df}), thereby preventing any possibility of data leakage.
 
-\todo{TODO: we need to elaborate this one with examples}
+By strictly separating these datasets, the validation phase serves its purpose of providing an unbiased evaluation of the model's ability to generalize---to data similar to, but not identical to that which it was trained on---thus maintaining the integrity and validity of the evaluation metrics derived during model testing.
+
+\subsubsection{Resource check ($N = 2$)}
+% (A18, A67)
+
+\todo{NOTE: connect this to~\ref{sec:implicit-resource-check}}
+
+In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
+Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
+
+This practice is illustrated by assertions such as those seen in \emph{A18} and \emph{A67}, where the code explicitly verifies the versions of critical libraries.
+For instance, \texttt{assert pd.\_\_version\_\_.rpartition(`.`)[0] == `1.0`} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
+
+These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
+
+\subsubsection{Existence check ($N = 8$)}
+% NOTE: this is a combination of missing-value-check and existence-check
+% (A23, A42, A43, A50, A51, A63, A79, A86)
+
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+A23 &
+\begin{lstlisting}
+assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())
+\end{lstlisting}\\
+
+A42 &
+\begin{lstlisting}
+assert not lab_s.isnull().values.any()
+\end{lstlisting}\\
+
+A43 &
+\begin{lstlisting}
+assert len(data) != 0, 'cannot divide by zero'
+\end{lstlisting}\\
+
+A50 &
+\begin{lstlisting}
+assert not np.any(np.isnan(X))
+\end{lstlisting}\\
+
+A51 &
+\begin{lstlisting}
+assert data.target.notnull().all()
+\end{lstlisting}\\
+
+A63 &
+\begin{lstlisting}
+assert X.isnull().sum().sum() == 0
+\end{lstlisting}\\
+
+A79 &
+\begin{lstlisting}
+assert not processed_data_df.isna().any().any()
+\end{lstlisting}\\
+
+A86 &
+\begin{lstlisting}
+assert p0 in poi_info.index
+\end{lstlisting}\\
+\end{tabular}
+\caption{Assertions used to perform existence checks.}
+\label{tab:existence-check}
+\end{table}
+
+Existence checks are a fundamental component of data preprocessing in machine learning, serving multiple critical purposes to ensure data integrity before analysis and modeling.
+These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
+Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values.
+
+For instance, assertions like \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
+Another common check, as seen with \emph{A43}, guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
+Similarly, checks such as \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
+Additionally, assertions like \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
+
+These checks are crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
 

### 2024-05-14 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsubsection{Mathematical property checks (A3, A25, A56, A64)}
-
-Machine learning models are often rooted in statistical models.
-Implementing neural networks requires a grasp of linear algebra and matrix multiplication.
-We find several assertions that test the mathematical properties of arrays and matrices after performing matrix operations.
-
-\subsubsection{Model performance check (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)}
-
-We see several assertions that test for the performance of the trained ML model against a validation or test test.
-The assertions often compare a performance metric (such as accuracy, precision, recall, F1, etc) against a predefined threshold.
-
-\todo{TODO: this needs more work and investigation}
-\subsubsection{Network architecture check (A11)}
-
-We only have one example here.
-
-% TODO lots to elaborate and discuss here
-\subsubsection{Resource check (A10, A14, A37, A60, A74)}
-
-We find assertions that check that a pre-trained model exists on the file system, or that data can be loaded from a specified file path.
-There are also checks that ensure that a visualisation created using matplotlib contains data in it.
-
-\subsubsection{Data shape check (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)}
-
-``Swiss army knife'' of assertions!
-
-\subsubsection{Type check (A2, A35, A40, A81, A88)}
-
-Checking the data type of features and such.
+\subsubsection{Mathematical property checks ($N = 4$)}
+% (A3, A25, A56, A64)
+
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+A3 &
+\begin{lstlisting}
+assert (xH - wH) % self.stride == 0
+\end{lstlisting}\\
+
+A25 &
+\begin{lstlisting}
+assert test_output.std() < 0.15, "Don't use batchnorm here"
+\end{lstlisting}\\
+
+A56 &
+\begin{lstlisting}
+assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))
+\end{lstlisting}\\
+
+A64 &
+\begin{lstlisting}
+assert np.allclose(T, T.T)
+\end{lstlisting}\\
+\end{tabular}
+\caption{Assertions used to validate mathematical properties of neural networks.}
+\label{tab:maths-check}
+\end{table}
+
+When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
+This rigor is captured through specific assertions aimed at validating these properties post-matrix manipulations.
+
+Assertion \emph{A3} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
+\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior.
+\emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations.
+Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.
+
+These assertions are not just precautionary but are vital for confirming that the underpinnings of machine learning algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
+
+\subsubsection{Model performance check ($N = 11$)}
+% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
+
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+A7 &
+\begin{lstlisting}
+assert len(neighbours_1) == 20, "Neighbors don't match!"
+\end{lstlisting}\\
+
+A15 &
+\begin{lstlisting}
+assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))
+\end{lstlisting}\\
+
+A19 &
+\begin{lstlisting}
+assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'
+\end{lstlisting}\\
+
+A38 &
+\begin{lstlisting}
+assert 0.75 < auc(fpr, tpr) < 0.85
+\end{lstlisting}\\
+
+A58 &
+\begin{lstlisting}
+assert np.isclose(accuracy, 0.9666666666666667)
+\end{lstlisting}\\
+\end{tabular}
+\caption{Assertions used to check performance of ML models.}
+\label{tab:model-perf-explicit}
+\end{table}
+
+This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
+
+For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors, an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
+Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
+Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity.
+Additionally, range checks like \emph{A38} confirm that the performance metrics such as AUC and Gini coefficient fall within acceptable boundaries, indicating good predictive performance without overfitting.
+Additionally, the use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
+This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
+By allowing a small tolerance in the comparison, np.isclose ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
+
+Each assertion acts as a crucial checkpoint that verifies the model's ability to generalize and perform effectively, ensuring that the outcomes are both robust and trustworthy.
+
+\subsubsection{Network architecture check ($N = 1$)}
+% (A11)
+
+When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
+This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
+
+We encountered an \texttt{assert} statement that addresses this architectural compatibility check: \lstinline{assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()}.
+This assertion verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
+The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
+By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
+
+This architectural check is crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
+
+\subsubsection{Resource check ($N = 5$)}
+% (A10, A14, A37, A60, A74)
+
+\begin{table}
+\centering
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}\\
+\midrule
+
+A10 &
+\begin{lstlisting}
+assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."
+\end{lstlisting}\\
+
+A14 &
+\begin{lstlisting}
+assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'
+\end{lstlisting}\\
+
+A37 &
+\begin{lstlisting}
+assert svm.fit_status_ == 0, 'Forgot to train the SVM!'
+\end{lstlisting}\\
+
+A60 &
+\begin{lstlisting}
+assert f2.gca().has_data()
+\end{lstlisting}\\
+
+A74 &
+\begin{lstlisting}
+assert os.path.exists(image_dir)
+\end{lstlisting}\\
+\end{tabular}
+\caption{Assertions used to validate availability of resources.}
+\label{tab:resource-check-explicit}
+\end{table}
+
+Ensuring the availability and integrity of required resources is a critical aspect of any machine learning workflow.
+Table~\ref{tab:resource-check-explicit} highlights some of the assertions we encountered that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
+These checks help prevent runtime errors and ensure the proper execution of code.
+
+One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
+These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing files.
+
+Additionally, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).
+A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
+The assertion \emph{A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
+This check prevents inconsistent model states from executing cells out of order or multiple times.
+
+Furthermore, assertions are used to verify the presence of data in visualizations generated using libraries like Matplotlib (\emph{A60}).
+These checks are crucial for ensuring that the visualizations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
+
+Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
+
+\subsubsection{Data shape check ($N = 26$)}
+% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
+
+Data shape checks can be considered the ``Swiss army knife''' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
+We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features, labels, predictions, images, sequences, and embeddings.
+These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.
+
+For instance, assertions like \emph{A17}:~\lstinline{assert X.shape[1] == 13, 'Did you drop/lose some columns in X?
+Did you properly load and split the data?'} and \emph{A90}:~\lstinline{assert len(X_train) == 2000} verify that the input features have the correct number of columns and examples, respectively.
+Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
+
+Similarly, assertions like \emph{A5}:~\lstinline{assert y_valid.shape == (1132,)} and \emph{A29}~\lstinline{assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
+
+In the context of computer vision tasks, assertions like \emph{A31}:~\lstinline{assert img.shape == (112, 92)} and \emph{A84}~\lstinline{assert red.get_shape().as_list()[1:] == [224, 224, 1]} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
+
+Furthermore, assertions like \emph{A76}:~\lstinline{assert len(encoding['token_type_ids']) == max_seq_length} and \emph{A93}:~\lstinline{assert temp_embed.shape[0] == stride} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
+
+Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
+By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
+
+\subsubsection{Type check ($N = 5$)}
+% (A2, A35, A40, A81, A88)
+
+Type checking assertions play a crucial role in maintaining the integrity and consistency of data types throughout the machine learning workflow.
+In the examined notebooks, we encountered several assert statements that verify the types of input features, models, data structures, and other objects.
+
+One common type check is performed on the input features, as demonstrated by the assertion \emph{A2}:~\lstinline{assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'}.
+This assertion ensures that the input features are represented as PyTorch float tensors, which is a requirement for many deep learning models and operations.
+
+Another type check is applied to verify the type of the machine learning models themselves.
+For instance, the \emph{A40}:~\lstinline{assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
+This type of assertion is essential when working with different types of models, as it helps prevent errors arising from using incompatible methods or attributes.
+
+In addition to input features and models, type checks are also performed on data structures and intermediate objects.
+The assertion \emph{A35}:~\lstinline{assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
+
+Type checking is not limited to objects and data structures--it can also be applied to individual columns or elements within a dataset.
+The assertion \emph{A81}:~\lstinline{assert is_all_ints(filled_df[r]) is True} checks if all the elements in a specific column (\lstinline{filled_df[r]}) of a DataFrame are integers.
+
+Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}:~\lstinline{assert isinstance(betas, np.ndarray)}.
+This assertion ensures that the \texttt{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
+
+Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behavior.
+By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.
 

### 2024-05-14 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsubsection{Network architecture check ($N = 1$)}
-% (A11)
-
+\subsubsection{Network architecture check ($N = 3$)}
+% (A11, A62, A75)
+
+% TODO: revisit this section; we have new assertions
 When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
@@
 
-\subsubsection{Value check (A30, A41, A44--46, A52, A65, A68, A69, A73, A92)}
-
-Check that the values in a column are in the specified values (categorical) or in a specified range (normalised continuous variable) or binary (such as the label column).
-
-% NOTE: blanks and unit-tests
-\subsubsection{Miscellaneous checks (A6, A20, A32, A34, A36, A53, A62, A75, A83, A87, A97)}
-
-These feel like unit tests, performed in lieu of writing full test suites in the notebook.
-
-%\begin{acknowledgements}
-%If you'd like to thank anyone, place your comments here
-%and remove the percent signs.
-%\end{acknowledgements}
-
-% Authors must disclose all relationships or interests that 
-% could have direct or potential influence or impart bias on 
-% the work: 
-%
-% \section*{Conflict of interest}
-%
-% The authors declare that they have no conflict of interest.
+\subsubsection{Data validation check ($N = 11$)}
+% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A92)
+
+% TODO: revisit, we have new assertions here; update the count
+we encountered a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
+These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
+
+One type of validation check involves verifying the presence of specific values or ranges within arrays or columns.
+For example, the assertion \emph{A41}:~\lstinline{assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))} checks if the unique values in the \texttt{smoke} column of the DataFrame \texttt{X} are limited to 0 and 1.
+Similarly, \emph{A65}~\lstinline{assert np.all(y <= nb_classes)} ensures that all values in the array \texttt{y} are less than or equal to a certain number of classes.
+
+Another common validation check is verifying the equality or closeness of values or arrays.
+The assertion \emph{A44}:~\lstinline{assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1} checks if the difference between the minimum and maximum values in a specific range of the \texttt{entity\_embedding} array is less than or equal to 1.
+Similarly, \emph{A46}:~\lstinline{assert np.isclose(stdev_norm, 1.0, atol=1e-16)} verifies that the \texttt{stdev\_norm} value is close to 1.0 within a specified absolute tolerance.
+
+Some assertions focus on validating the uniqueness or cardinality of values in arrays or DataFrames.
+For instance, \emph{A52}:~\lstinline{assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()} checks if the number of unique user IDs in the \texttt{grouped\_users} DataFrame is equal to the number of unique user IDs in the \texttt{user\_engagement} DataFrame.
+
+Other assertions validate the presence of specific values or conditions within DataFrames or other data structures.
+The assertion \emph{A73}:~\lstinline{assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \texttt{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
+
+By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
 

### 2024-05-16 report/emse24.tex
feat(report): emse24 updates

@@
 
-% amershi2019software: start with ML development life cycle and how everything is done in a very iterative way; things are dependant on one another; etc.
-
-% martinez-plumed2021crisp-dm: not only does ML include the crip-dm framework, but this same form of iteration is seen in the model development and experimentation phase;
-
-% this is why we see the use of computational notebooks being widely adopted by the ML community
-
-% writing ML code is more iterative and requires a constant feedback; we identify this feedback can be provided in two ways: implicitly from the output of a cell and explicitly from a failed assertion
-
-Code, data and ML model in ML-enalbed systems are highly tangled with each other.
-For instance, the decision to use a particular type of ML model, requires a deep understanding of the data which will be used to train the model.
-Data quantity and quality directly influence the complexity of the ML model.
-
-Conversely, the choice of the ML model also influences the data pre-processing and transformation steps that are required to get the data ready for training.
-For instance, using a classic statistical model requires several data transformation steps such as scaling and normalization, and converting categorical features to numerical ones using one-hot encoding.
-
-Visualisations are used throughout the ml development lifecycle to test various properties of the ml system.
+Machine Learning Development Lifecycle (MLDL) encompasses multifaceted series of entangled stages.
+Significant effort is invested in gathering and assembling the required data from diverse sources.
+The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages.
+MLDL is dominated by the model development phase which is highly experimental and iterative.
+This stage involves Exploratory Data Analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results.
+Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance.
+Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance.
+The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.
+
+% The feedback-driven style of development in MLDL resembles principles of Agile software development and the Cross-Industry Standard Process for Data Mining (CRISP-DM).
+Agile methodologies emphasize continuous iteration of development and testing throughout the software lifecycle.
+In Agile, software is developed in incremental, rapid cycles resulting in small, incremental releases with each release building on previous functionality.
+Each iteration is reviewed and critiqued by the project team, which includes representatives of the project's various stakeholders.
+
+The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations.
+Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring.
+Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
+
+Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks.
+Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development.
+This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development.
+This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms.
+Users can test hypotheses, debug issues, and make incremental changes with immediate feedback.
+The ability to visualize data and results inline and document the process in a narrative form further enhances their utility.
+
+% It aligns perfectly with the feedback-driven nature of machine learning development, as it supports rapid prototyping and iterative refinement of models.
+
+% TODO: back the TDD angle not only from scientific literature but also general methodologies used and adopted by the software engineering community (Xtreme programming, Kent Beck)
+
+The predominant form of obtaining feedback in Jupyter notebooks is through the output of code cells.
+These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations.
+Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code.
+This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.
+
+% TODO cite scully and breck and the data validation papers here...
+% TODO: why are we doing this? how is the prior work lacking? why is this important?
+% we use feedback in notebooks to make important decisions: what data pre-processing techniques to use, create new features to boost performance of a model, pick particular hyper-parameters.
+but these decisions only reflect the conditions when the model was created; things inevitably change: data, people, system; if we don't document these decisions then it can lead to technical debt; longer time to debug issues; which can cost companies resource (time & cost) but also imagine if the model is being used in safety critical systems: can lead to loss of human lives as well
+
+% in software engineering, tests are used to document expectations of how the code should function.
+Failing tests indicate that our expectations no longer satisfied.
+This is why we also look for assertions written in notebooks and analyse their use.
+
+% and then to tie the two things together: are assertions derived from the output of a cell?
+we still need to find a way to justify why we only look at assertions derived from visualisations specifically
+
+% By analyzing the outputs and assertions in Jupyter notebooks, we aim to provide insights into the feedback mechanisms used in machine learning development, ultimately proposing strategies to improve the efficiency and robustness of this iterative process.
+
+% Visualisations are used throughout the ml development lifecycle to test various properties of the ml system.
 in the early stages of the ml development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
@@
 
-building ml systems is highly iterative and experimental.
+% building ml systems is highly iterative and experimental.
 information gained from ml visualisations is used to make design and implementation decisions for the following steps of the ml pipeline.
@@
 
-implicit expectations obtained from visualisations must therefore be
-translated to analytical tests that fail once our expectations
-regarding the ml system are no longer satisfied.
-
-in contrast to prior work, we approach ml testing from a new
-perspective. we conduct an empirical analysis of computational
-notebooks obtained from github, to understand the process of testing
-ml systems in practice. in particular, we focus on the combination of
-a qualitative form of testing (using visualisations) and a
-quantitative form of testing (using analytical assertions). the
-research questions along with the contributions of this paper are as
-follows.
-
-\begin{description}
-  \item[rq1.] \textbf{how frequently are analytical tests formulated
-    from visualisations created to test ml systems?}
-
-    we mine 54k jupyter notebooks from github that contain an assertion
-    written in python. we develop an automated technique to identify 1764
-    notebooks with an assertion below a visualisation. we manually analyse
-    the 1.7k notebooks and identify 269 visualisation-assertion (va) pairs
-    that are semantically related to each other. we plan to release this
-    catalogue of va pairs publicly.
-
-\item[rq2.] \textbf{what patterns are frequently observed in va pairs
-  used to test ml systems?}
-
-  we analyse the 269 va pairs from rq1, and observe three testing
-  patterns that frequently occur in the va pairs.
-
-\item[rq3.] \textbf{do the assertions capture all the information
-  obtained from the corresponding visualisation?}
-
-  we further perform an in-depth analyse of 34 va pairs (and their
-  corresponding notebooks). we observe that assertions often cannot
-  capture all the information obtained from the visualisation.
-
-\end{description}
-
-our observations indicate that formulating analytical assertions from
-visualisations is an emerging testing technique. however, the
-assertions found in this study often do not reflect the information
-obtained from their visual counter-part. this study finds the need for
-automated tools that can reduce the manual effort necessary to
-validate visualisations in ml systems. and help ml practitioners to
-formulate better analytical assertions from visualisations.
-
-the replication package for this study is made public on
-figshare\footnote{the replication package can be found at this url:
-https://figshare.com/s/9347228011999cba29f5}.
+% implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ml system are no longer satisfied.
+
+% in contrast to prior work, we approach ml testing from a new perspective.
+we conduct an empirical analysis of computational notebooks obtained from github, to understand the process of testing ml systems in practice.
+in particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions).
+the research questions along with the contributions of this paper are as follows.
+
+% our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique.
+however, the assertions found in this study often do not reflect the information obtained from their visual counter-part.
+this study finds the need for automated tools that can reduce the manual effort necessary to validate visualisations in ml systems.
+and help ml practitioners to formulate better analytical assertions from visualisations.
+
+% the replication package for this study is made public on figshare\footnote{the replication package can be found at this url: https://figshare.com/s/9347228011999cba29f5}.
 
@@
 
-\todo{TODO: introduce the motivation of the study along with the RQs here}
+\subsubsection{Motivating Example}
+
+Imagine a data scientist named Sara who is developing a machine learning model to predict customer churn.
+Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells.
+Her workflow is dynamic and exploratory, typical of modern machine learning practices.
+
+Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution.
+As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features.
+This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.
+
+With data preprocessing complete, Sara proceeds to build her initial model.
+After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy and loss after the first epoch.
+The results are disappointing, prompting her to adjust the learning rate and batch size directly in subsequent cells.
+Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings.
+
+Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions.
+At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.
+
+Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively.
+Each decision in Sara's workflow—from preprocessing adjustments to hyperparameter tuning and model validation—is guided by the iterative feedback she receives from the outputs of her notebook's code cells.
+
+% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster!
+(notebook keeps running!)
+Once all experimentation and model tuning are complete, and Sara is ready to deploy the model into a production environment, she adds assertions as a final step.
+In the dynamic, interactive environment of Jupyter notebooks, where code cells can be executed in any order, there is a risk of arriving at inconsistent states that may affect the model's performance unpredictably.
+Assertions act as a safeguard against these inconsistencies by verifying that the data fed into the model adheres to expected formats and distributions and that the model's performance metrics meet predetermined standards.
+For instance, assertions are used to ensure that no data columns have null values, that the range of input data features remains within expected bounds, and that model accuracy thresholds are consistently met regardless of the execution order of the notebook cells.
+This practice not only reinforces the stability and reliability of the deployed model but also provides clear, executable documentation of the critical decisions made throughout the model's development.
+When new team members are onboarded, these assertions allow them to quickly grasp the underlying assumptions and operational parameters of the model, ensuring ongoing robustness and performance integrity.
+This is crucial in maintaining the quality and reliability of the model as it transitions from a development setting to a production environment.
+
+\subsubsection{Research Questions}
+
+\begin{description}
+  \item[RQ1.] \textbf{How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
+  \item[RQ2.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
+  \item[RQ3.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
+\end{description}
 
@@
 \subsection{Data Collection}\label{sec:data-collect}
+
+% TODO we need to explain why we only focus on the last statement of code cells (when visualisation code written anywhere produces an output); this is much harder to parse: we need to construct an exhaustive list of python visualisation libraries; people can use any naming convension to...
 
@@
 
+\subsection{Assertions derived from visualisations}
+% NOTE: following text is from the ICST paper; still need to think about how to integrate it here...
+
+\subsubsection{What patterns are frequently observed in VA pairs used to test ML systems?}
+
+We further analysed the 269 VA pairs obtained from RQ1, and identified three testing patterns which are presented in Table~\ref{tab:testing-patterns} and discussed further below.
+
+\begin{table}
+  \centering
+  \caption{Testing patterns observed in the visualisation-assertion pairs.}
+  \begin{tabularx}{0.45\textwidth}{@{}l X r r@{}}
+    \toprule
+    \textbf{ID} &
+    \textbf{Description} &
+    \multicolumn{2}{c}{\textbf{Count}}\\
+    & & Yes & No\\
+    \midrule
+    \textbf{P1} &
+    Does the assertion use a comparison operator (such as \emph{eq, lt, gt})? &
+    202 & 67\\
+    \textbf{P2} &
+    Does the assertion use magic thresholds? &
+    101 & 168\\
+    \textbf{P3} &
+    Does the assertion use an external testing library (such as \texttt{numpy.testing})? &
+    42 & 227\\
+    \bottomrule
+  \end{tabularx}
+  \label{tab:testing-patterns}
+\end{table}
+
+\begin{table}
+  \centering
+  \caption{Most frequently occurring assertions that use a comparison operator.}
+  \begin{tabular}{@{}m{0.9\linewidth}@{}}
+    \toprule
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(mean_rw_history[-10:]) > 10.
+print("That's good enough for tutorial.")\end{lstlisting}\\
+    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
+    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
+assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
+    \end{lstlisting}\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:compare-op-asserts}
+\end{table}
+
+In a total of 269 VA pairs, 202 assertions (75\%) contain a comparison operator such as \emph{equal to}, \emph{lesser than} or \emph{greater than}.
+Table~\ref{tab:compare-op-asserts} presents the most frequently observed assertions that conform to this pattern.
+
+101 of 269 assertions (37\%) compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
+This covers 46\% of the assertions with a comparison operator.
+It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
+The threshold could have been derived based on domain expertise.
+Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirements,rahimi2019toward}.
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{../catalogue/select-16.png}
+  \caption{Visualisation of the training loss of an ML model.}\label{fig:loss}
+\end{figure}
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
+In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]
+assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."
+\end{lstlisting}
+
+In contrast to the assertions presented in Table~\ref{tab:compare-op-asserts}, Listing~\ref{lst:loss} presents an assertion that ensures that the model performs well, without using a magic threshold.
+The assertion checks if the mean of the first 10 observations of the loss is higher compared to the mean of the last 10 observations.
+In other words, the assertion ensures that the loss of the model decreased after training, and that the model was able to ``learn'' from the training set.
+The assertion is derived from the visualisation presented in Figure~\ref{fig:loss}, which shows the loss of the model over the training epochs.
+
+\begin{figure}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=\linewidth]{../catalogue/select-65a.png}}
+  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=\linewidth]{../catalogue/select-65b.png}}
+  \caption{Visualisation of data before and after various data pre-processing techniques.}
+  \label{fig:data-pre-process}
+\end{figure}
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
+assert scaled_data[scaled_data.argmax()] <= 1
+assert scaled_data[scaled_data.argmix()] >= 0
+\end{lstlisting}
+
+\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
+assert normalized_data[normalized_data.argmax()] < 10
+assert normalized_data[normalized_data.argmin()] > -7
+\end{lstlisting}
+
+Visualisations in Figure~\ref{fig:data-pre-process} were obtained from a notebook that presents a tutorial on scaling and normalization---two common data pre-processing techniques in ML.
+The visualisations show a comparison of the data before and after each pre-processing technique is applied to a feature from the dataset.
+The author performs a visual check, by comparing the distribution of the data before and after applying the pre-processing technique, using a histogram.
+
+Listing~\ref{lst:scale} and~\ref{lst:normal} show the assertions that were defined below Figure~\ref{fig:scale} and~\ref{fig:normal} respectively.
+Both assertions check that the min and max of the feature is within the expected bounds after applying the respective pre-processing technique.
+In contrast to assertions presented in Table~\ref{tab:compare-op-asserts}, it is clear that the thresholds are derived from the corresponding visualisations.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]
+assert not np.allclose(x_aug, x_original)
+\end{lstlisting}
+
+22\% of the assertions did not use a comparison operator, nor a magic threshold.
+Listing~\ref{lst:no-comp-magic} highlights such an assertion which is obtained from a notebook that performs image augmentation using Weighted K-Nearest Neighbours.
+The assertion compares the matrix representation of two images, and checks that the augmented image is different from the original.
+
+Most notebooks examined in this study do not use any external testing libraries.
+227 or 84\% of the VA pairs use the low-level Python \texttt{assert} statement to define the test directly.
+This aligns with our observations that majority of the assertions compare the value of a performance metric against a user defined threshold, and do not require the use of a specialised testing library.
+
+This study found 42 VA pairs where an external testing library was used.
+Listing~\ref{lst:np-almost-equal} shows an assertion that uses the \texttt{assert\_almost\_equal} method from the \texttt{numpy.testing} module.
+The assertion checks if the loss of the model is similar to the specified value, up to a desired precision.
+Due to the stochastic nature of ML, results obtained from two separate development cycles will slightly differ.
+The assertion presented in Listing~\ref{lst:np-almost-equal} is found to be robust towards such perturbations and will only fail if there is a significant change in the performance of the model.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]
+ans4 = compute_loss(X_expanded, y, w)
+np.testing.assert_almost_equal(ans4, 0.3042764) 
+\end{lstlisting}
+
+\subsubsection{Do the assertions capture all the information obtained from the corresponding visualisation?}
+We performed an in-depth analysis of 34 unique VA pairs.
+This was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
+During the analysis, several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data were captured.
+We find several instances of VA pairs where the assertion is unable to capture all the information obtained from the visualisation.
+
+\begin{figure}
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
+  \hfill
+  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
+  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
+\end{figure}
+
+\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
+assert accuracy_score(y_test, pred) > 0.95
+\end{lstlisting}
+
+Figure~\ref{fig:svm} and Listing~\ref{lst:svm} present one such pair.
+Visualisations~\ref{fig:svm-linear} and~\ref{fig:svm-rbf} are obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
+The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel.
+
+The visualisations contain several useful information that can be used to test and debug ML models.
+For instance, using Visualisation~\ref{fig:svm}, ML practitioners can get a visual understanding of how well the model is able to classify the two classes in the dataset.
+Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
+The visualisation also allows comparison of different ML models or variants, to iteratively find a model and hyper-parameters that best fits the underlying training data.
+In this case for instance, the author of the notebook first trained the SVM with a linear kernel.
+And upon realising that the data is not linearly separable, chose the RBF kernel.
+
+Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
+The assertion checks if the model accuracy is above a threshold specified by the author.
+The assertion however, is only able to capture the information pertaining to the performance of the model.
+While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]
+r2_gru = r2_score(y_test, y_pred)
+assert r2_gru > 0.6
+\end{lstlisting}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
+  \caption{Scatter plot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
+\end{figure}
+
+Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
+The visualisation shows a scatter plot of the real and predicted values of $NO_2$ from the test set.
+A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data.
+The spread of $y_{pred}$ at any given $y_{true}$ is wide.
+For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
+This indicates that the model was not able to fit the training data that well, and can be fine-tuned.
+
+Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
+The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computes the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
+The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
+Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
+
+Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess if the model is under or overfitting the training data.
+If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
+In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
+The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
+The assertion in Listing~\ref{lst:r2} however, does not capture such information.
+It can only check if the performance of the model is better than a specific threshold.
+
+\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
+for feature in range(data_transformed.shape[1]):
+  assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
+\end{lstlisting}
+
+\begin{figure}
+  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
+  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
+\end{figure}
+
+In contrast to Listing~\ref{lst:svm} and Listing~\ref{lst:r2}, Figure~\ref{fig:kstest} and Listing~\ref{lst:kstest} present a VA pair where the assertion captures all the information obtained from the visualisation.
+
+The notebook author normalises the training data and performs a visual check using the visualisation presented in Figure~\ref{fig:kstest}.
+The visualisation shows a bell-shaped curve thus indicating that the data is normal and that the transformation was successful.
+The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
+The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
+
+Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
+Several ML models assume normality in the underlying training data.
+However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
+The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
+And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
+
+\section{Implications}\label{sec:discuss}
+% NOTE: from ICST paper, need to integrate...
+
+% NOTE: breck2017ml proposes feature and data testing; we can say that the assertions in the notebook bring the system closer to migrating them into dedicated data validation systems
+
+% NOTE: think about the actors: educators, researchers, developers...
+The results from RQ1 provide empirical evidence to support the use of
+visualisations to test various properties of ML systems.
+Visualisations are an efficient and effective medium to present
+complex information, that is easier to interpret by humans with the
+aid of visual cues. Visualisations used to test ML system properties
+however, demand manual verification effort. Information gained from
+visualisations and subsequent decisions made based on this
+information, must be manually verified in every iteration of the ML
+development cycle.
+
+% NOTE link to accumulation of technical debt due to highly
+% experimental nature, and constant change in the system
+
+Assertions or analytical tests extracted from ML visualisations can
+reduce this manual verification effort. Assertions encapsulate the
+knowledge gained from the visualisation with regards to the property
+of the ML system which is currently under test. Formal assertions
+document the observations made by the AI practitioner regarding the
+model or the data, at a specific point in time. These observations
+should not be violated by the future model development cycles unless
+the requirements of the ML system are changed. Assertions scale across
+organisational or personnel changes as future AI practitioners can
+refer to the assertions to understand how the visualisations were
+previously interpreted.
+
+The results for RQ1 indicate that the combination of ML visualisations
+with analytical assertions is an emerging testing practice within the
+machine learning community. We found only $54,070$ notebooks that
+contain the keyword ``assert'', while there are more than a million
+Jupyter notebooks available on GitHub. This suggests that writing
+assertions in computational notebooks is still relatively uncommon. We
+believe that this is due to a lack of knowledge among practitioners
+regarding testing approaches and the absence of adequate tooling to
+support the assertion of model properties. We argue that guidelines on
+ML coding practices require improvement. And advocate the need to
+steer education and research of Software Engineering practices towards
+the specific challenges faced when contributing to ML software
+systems.
+
+In the 1764 notebooks analysed in this study, several visualisations
+lack assertions. However, implicit expectations still exist within
+each visualisation which cannot be readily validated by other project
+collaborators. ML testing demands proficiency in Machine Learning,
+Software Testing, and the underlying Mathematical concepts. This
+further highlights the need for refining the education for these
+professionals and crafting tools to compensate for gaps in their
+knowledge.
+
+Results from RQ2 reveal two challenges in the current landscape of ML
+testing. The assertions in the VA pairs identified in this study lack
+dedicated testing libraries. This indicates that the current culture
+of testing ML code is still in its infancy and ML practitioners are
+not yet exploring software testing libraries. Second, the existing
+testing libraries are not tuned to the specifics of ML code and
+practitioners have to fall back to basic built-in approaches.
+
+The widespread use of ``magic'' thresholds---often rooted in domain
+expertise or experience obtained over time---poses risks to
+maintainability of ML projects. We advocate that such thresholds are
+documented for future reference. And find a clear need for automated
+tools that highlight undocumented choices or suggest standardized
+documentation in Markdown or code annotations.
+
+In the results from RQ3, we see several examples of assertions that
+fail to capture all the information gained from its visual
+counter-part. This can lead to a mismatch of expectations in future
+versions of the ML software. For instance, another project
+collaborator might run the code with a new version of the dataset or a
+different learning technique and assume that if the assertions run
+successfully, the expected patterns from the visualisation are also
+satisfied. The examples presented in Section~\ref{sec:result-rq3} show
+that deriving assertions from visualisations is not trivial. Basic
+visualisation patterns require complex mathematical techniques which
+may be beyond the expertise of an ML developer.
+
+For example, a linear correlation between the true and predicted
+labels is evident in Figure~\ref{fig:r2}. Yet, its formal assertion in
+Listing~\ref{lst:r2} necessitates defining an appropriate threshold
+for the coefficient of determination metric. Our results show that
+current software testing practices fail to capture the idiosyncrasies
+of ML code. There is a pressing need for automated tools, that bridge
+the disparity between the qualitative evaluation of ML systems using
+visualisations and the extraction of analytical assertions from these
+visuals.
+
 % BibTeX users please use one of
@@
 
-\subsection{Assertions derived from visualisations}
 \end{document}

### 2024-05-16 report/emse24.tex
Updates from Overleaf

--- 2024-05-14 report/emse24.tex
@@
 \todo{TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
-Thats why we need to apply these tecniques to surface unique and interesting candidates!}
+Thats why we need to apply these techniques to surface unique and interesting candidates!}
 
@@
 Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
-The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens thay appear less frequently are assigned a higher value.
+The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
 
@@
 
-P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
+The print statement P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
 This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
@@
 
-\subsubsection{Shape check ($N = 3$}
+\subsubsection{Shape check ($N = 3$)}
 % (P4, P32, P117)

### 2024-05-17 report/emse24.tex
feat(report): emse24 updates

@@
 % TODO: why are we doing this? how is the prior work lacking? why is this important?
-% we use feedback in notebooks to make important decisions: what data pre-processing techniques to use, create new features to boost performance of a model, pick particular hyper-parameters.
-but these decisions only reflect the conditions when the model was created; things inevitably change: data, people, system; if we don't document these decisions then it can lead to technical debt; longer time to debug issues; which can cost companies resource (time & cost) but also imagine if the model is being used in safety critical systems: can lead to loss of human lives as well
-
-% in software engineering, tests are used to document expectations of how the code should function.
-Failing tests indicate that our expectations no longer satisfied.
-This is why we also look for assertions written in notebooks and analyse their use.
+
+In Jupyter notebooks, feedback is crucial for making important decisions throughout the machine learning development lifecycle.
+Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters.
+However, these decisions are often based on the conditions present when the model was initially created.
+As data, personnel, and system environments inevitably change, undocumented decisions can lead to significant technical debt.
+This not only prolongs debugging and troubleshooting times but also incurs additional costs and resource expenditures.
+
+In software engineering, tests are employed to document the expected behavior of code.
+Failing tests indicate when these expectations are no longer met.
+Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development.
+By embedding assertions, developers ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.
+
+% TODO introduce our results and key findings...
 
@@
 
-Imagine a data scientist named Sara who is developing a machine learning model to predict customer churn.
+Sara is a data scientist who is developing a machine learning model to predict customer churn.
 Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells.
@@
 
-% role of feedback in writing ML code? (any recommendation for papers I can refer to here?)
-% implicit vs. explicit feedback (how do we establish this concept?)
-
 \subsection{Data Collection}\label{sec:data-collect}
@@
 
-\subsection{Implicit feedback from the output of code cells}
+\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
 
@@
 
+% TODO: we refer to these as outputs but they are actually the python statements that produce the outputs...do we need better terminology here?
+To me this is especially confusing because in the figure below we are actually showing the output, not the python statement (which makes sense, its more interesting to show the actual visualisation in these cases rather than the code).
 \begin{table}
@@
 \caption{Examples of cell outputs used to verify distribution of data.}
-\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.3\textwidth}@{}}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
 \toprule
@@
 \emph{\textbf{Code}}&
-\emph{\textbf{Description}}\\
+\emph{\textbf{Visual Output}}\\
 \midrule
 
+O2&
+\lstinline[]$_ = sns.catplot(x='category_id', y='likes', data=train, height=5, aspect=1.5)$&
+\includegraphics[width=\linewidth]{distribution-check-catplot.png}\\
+
+O9&
+\lstinline[]$sns.kdeplot(data=data.loc[ data['Survived'] == 0].Age, label='Died', shade=True)$&
+\includegraphics[width=\linewidth]{distribution-check-kdeplot.png}\\
+
 O14&
-\begin{lstlisting}
-pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])
-\end{lstlisting}&
-Check the mean value of the specified features for each label in the target feature.\\
+\lstinline[]$pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])$&
+\\
+
+O25&
+\lstinline[]$sns.countplot(house_pred['OverallQual'])$&
+\\
 
 O48&
-\begin{lstlisting}
-x_train.describe()
-\end{lstlisting}&
-Check the descriptive statistics of the training data. The above statement is written immediately after normalising the features in the training data.\\
+\lstinline[]$x_train.describe()$&
+\\
 
@@
 
-\begin{figure*}
-\centering
-\subcaptionbox{Check the distribution of the feature \texttt{Age} with-respect-to the labels in the target feature \texttt{Survived}.\label{fig:distribution-check-kdeplot}}{\includegraphics[width=0.45\linewidth]{distribution-check-kdeplot.png}}
-\subcaptionbox{Check the distribution of the feature \texttt{likes} across all values of feature \texttt{category\_id}.\label{fig:distribution-check-catplot}}{\includegraphics[width=0.45\linewidth]{distribution-check-catplot.png}}
-\caption{Examples of visualisations used to check linear relationship in data.}
-\label{fig:distribution-check}
-\end{figure*}
-
-Understanding the distribution of data is crucial for making informed decisions regarding necessary data transformation steps in data analysis processes.
-For instance, identifying whether scaling, normalizing, or handling outliers is required can be efficiently and effectively determined through visualizations.
-
-% This is particularly important for models like linear regression, which perform optimally when data is normally distributed.
-If the data deviates from normality, additional transformation steps may be necessary to prepare it for training.
-
-During the exploratory data analysis (EDA) or data understanding phase, a combination of visualizations and pandas dataframes is often employed to assess the distribution of specific columns within the data.
-This step helps identify features that may have a relationship with the target variable, ultimately informing which features to include in the machine learning model training process.
-This is seen in Figure~\ref{fig:distribution-check-kdeplot} where the distribution of a feature \texttt{Age} is checked against the labels of the target feature.
-Figure~\ref{fig:distribution-check-catplot} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
-
-For example, the distribution of a newly created feature, such as one where a continuous feature is binned into categories, is frequently examined.
-
-Descriptive statistics are also used post-transformation to verify that steps like data normalization have been effectively applied, ensuring that the data conforms to the expected format for further analysis.
-
-Furthermore, distribution checks are often performed immediately after feature creation.
-For instance, this may occur after discretizing a continuous feature using binning.
-Additionally, descriptive statistics (e.g., mean and standard deviation) obtained from libraries like pandas can be used to verify the effectiveness of data normalization steps.
+Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data.
+In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature.
+Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
+Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}.
+Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data.
+The statement is written immediately after normalising the features in the training data.
+Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output.
+Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.
+
+% TODO: this is probably implications/discussions material?
+Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
+For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
+During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
+Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.
 
@@
 % (O6, O10)
-\todo{TODO: I think correlation checks should also be present in this group}
-
-\begin{figure*}
+
+\begin{table}
   \centering
-\subcaptionbox{Using a lineplot to check linear relationship between two features in a dataset.\label{fig:linear-relation-check-lineplot}}{\includegraphics[width=0.45\linewidth]{linear-relation-check-lineplot.png}}
-\subcaptionbox{Visualisation that shows the datapoints as a scatterplot along with a linear model fit.\label{fig:linear-relation-check-regplot}}{\includegraphics[width=0.45\linewidth]{linear-relation-check-regplot.png}}
-\caption{Examples of visualisations used to check linear relationship in data.}
-\label{fig:linear-relation-check}
-\end{figure*}
+  \caption{Cell outputs used to verify linear relationship between features in the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}&
+    \emph{\textbf{Visual Output}}\\
+    \midrule
+
+    O6&
+    \lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$&
+    \includegraphics[width=\linewidth]{linear-relation-check-lineplot.png}\\
+
+    O10&
+    \lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$&
+    \includegraphics[width=\linewidth]{linear-relation-check-regplot.png}\\
+    \bottomrule
+  \end{tabular}
+\label{tab:linear-relation-check}
+\end{table}
+
+Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data.
+In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
+Output \emph{O10} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
 
@@
 Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.
-Figure~\ref{fig:linear-relation-check} illustrates two methods for identifying these linear relationships.
-In Figure~\ref{fig:linear-relation-check-lineplot}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
-Figure~\ref{fig:linear-relation-check-regplot} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
 
@@
 \caption{Examples of cell outputs used to verify if different types of resources are available on the system where the notebook is being executed.}
-\begin{tabular}{@{}m{0.1\textwidth} m{0.05\textwidth} m{0.4\textwidth} m{0.3\textwidth}@{}}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
 \toprule
-\emph{\textbf{Type of resource}}&
 \emph{\textbf{Key}}&
-\emph{\textbf{Code}}&
-\emph{\textbf{Description}}\\
+\emph{\textbf{Code}}\\
 \midrule
 
-External library&
+P68&
+\lstinline[]$print('GPU is available')$\\
+
 P71&
-\begin{lstlisting}
-print('Hub version: ', hub.__version__)
-\end{lstlisting}&
-Check the version of an external Python library installed on the current system.\\
-
-Compute resources&
-P68&
-\begin{lstlisting}
-print('GPU is available')
-\end{lstlisting}&
-Check if a GPU is available on the current system.\\
-
-Data or pre-trained models&
+\lstinline[]$print('Hub version: ', hub.__version__)$\\
+
+P82&
+\lstinline[]$print('Running on TPU ', tpu.master())$\\
+
+P86&
+\lstinline[]$print('Cuda is available')$\\
+
+P107&
+\lstinline[]$print('Model loaded')$\\
+
+O64&
+\lstinline[]$full_table.head(-5)$\\
+
 O66&
-\begin{lstlisting}
-prostate_cancer_df.shape
-\end{lstlisting}&
-Check that the dataset has been loaded into memory.\\
+\lstinline[]$prostate_cancer_df.shape$\\
 \bottomrule
@@
 
-\todo{perhaps also add the keys of the case-studies here?}
-Table~\ref{tab:resource-check} presents case studies where the output is used to verify that certain resources are available on the system in which the notebook is being executed.
-We find checks for the availability of compute resources (such as a GPU or a TPU), checking that the dataset or a pre-trained model has been loaded correctly and checks to ensure that a certain version of an external library is present on the system.
+% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
+Table~\ref{tab:resource-check} presents case studies where the output is used to verify availability of resources on the system in which the notebook is being executed.
+Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU.
+This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory.
+For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory.
+Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data.
+These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.
+Additionally, output \emph{P71} is used to ensure that a certain version of an external library is present on the system.
 

### 2024-05-22 report/emse24.tex
feat(report): emse24 updates

@@
 
-\begin{lstlisting}
-print('Total Run Time:')
-\end{lstlisting}
+\begin{table}
+  \centering
+  \caption{Print statement used to check the total execution time of training and cross-validating an ML model.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    P66&
+    \lstinline[]$print('Total Run Time:')$\\
+  \end{tabular}
+  \label{tab:exec-time}
+\end{table}
+
+Table~\ref{tab:exec-time} presents print statement \emph{P66} which prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
+\emph{P66} is also used in other code cells in the notebook that train other ML models.
 
 While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
-Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run.
-The provided code snippet (Listing~\ref{lst:exec-time-check}) exemplifies this concept by showcasing a case study (P66) that tracks training times for various models using a grid search.
-This approach allows practitioners to compare model performance alongside training efficiency.
+Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{CITEME}.
 
@@
 % (O12, O36, P74)
+
+\begin{table}
+\centering
+\caption{Examples of cell outputs used to check for missing data.}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}&
+\emph{\textbf{Visual Output}}\\
+\midrule
+
+P74&
+\lstinline[]$print(train_df.isnull().sum())$&
+\\
+
+O12&
+\lstinline[]$sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')$&
+\includegraphics[width=\linewidth]{missing-value-check.png}\\
+
+O36&
+\lstinline[]$test.isna().sum().unique()$&
+\\
+\bottomrule
+\end{tabular}
+\label{tab:missing-value}
+\end{table}
+
+Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
+While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
 
@@
 
-\begin{table}
-\centering
-\caption{Examples of cell outputs used to check for missing data.}
-\begin{tabular}{@{}m{0.05\textwidth} m{0.6\textwidth}@{}}
+\subsubsection{Model performance check ($N = 33$)}
+% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
+
+\begin{table}
+\centering
+\caption{Various cell outputs used to check the performance of an ML model after training.}
+\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+\toprule
+\emph{\textbf{Key}}&
+\emph{\textbf{Code}}&
+\emph{\textbf{Visual Output}}\\
+\midrule
+
+P3&
+\lstinline[]$print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')$&
+\\
+
+P6&
+\lstinline[]$print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))$&
+\\
+
+P18&
+\lstinline[]$print('The Accuracy is:', accuracy_score(y_test, y_pred))$&
+\\
+
+P50&
+\lstinline[]$print('Classification Report: SVM (validation data)')$&
+\\
+
+P54&
+\lstinline[]$print('Intercept value:', lm.intercept_)$&
+\\
+
+O3&
+\lstinline[]$skplt.metrics. plot_confusion_matrix(Y_val, Vote.predict(X_val), normalize=True, figsize=(10, 10))$&
+\includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}\\
+
+O52&
+\lstinline[]$spot_check_recs(classifier, 910)$&
+\\
+\bottomrule
+\end{tabular}
+\label{tab:model-perf}
+\end{table}
+
+Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
+Besides the \emph{accuracy} of the model (\emph{P3} and \emph{P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
+
+% TODO: this feels hallucinated; revisit
+Print statement \emph{P54} prints the intercepts of a linear regression model.
+Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
+This insight is critical for accurate model interpretation and effective decision-making in practical applications.
+
+Output \emph{O3} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
+The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
+
+Output \emph{O52} shows a function implemented by the practitioners to manually evaluate an image classification model on a random sample of input images.
+The function \texttt{spot\_check\_recs} displays the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
+
+During the model development phase, the performance of an ML model is often printed to facilitate comparisons against other models or variations.
+Throughout this continuous experimentation phase, authors may adjust model parameters or modify the data by engineering new features.
+The model is then re-trained to check if these changes lead to performance improvements.
+
+\subsubsection{Neural network architecture ($N = 1$)}
+% (P92)
+
+\begin{table}
+  \centering
+  \caption{Print the architecture of a neural network.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
     \toprule
@@
     \midrule
-
-P74&
-\begin{lstlisting}
-print(train_df.isnull().sum())
-\end{lstlisting}\\
-
-O12&
-\begin{lstlisting}
-sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')
-\end{lstlisting}\\
-
-O36&
-\begin{lstlisting}
-test.isna().sum().unique()
-\end{lstlisting}\\
-\end{tabular}
-\label{tab:missing-value}
-\end{table}
-
-Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
-While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
-
-\subsubsection{Model performance check ($N = 33$)}
-% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
-
-\begin{table}
-\centering
-\caption{Various cell outputs used to check the performance of an ML model after training.}
-\begin{tabular}{@{}m{0.05\textwidth} m{0.6\textwidth}@{}}
-\toprule
-\emph{\textbf{Key}}&
-\emph{\textbf{Code}}\\
-\midrule
-
-P3&
-\begin{lstlisting}
-print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')
-\end{lstlisting}\\
-
-P6&
-\begin{lstlisting}
-print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))
-\end{lstlisting}\\
-
-P18&
-\begin{lstlisting}
-print('The Accuracy is:', accuracy_score(y_test, y_pred))
-\end{lstlisting}\\
-
-P18&
-\begin{lstlisting}
-print('Classification Report: SVM (validation data)')
-\end{lstlisting}\\
-
-P54&
-\begin{lstlisting}
-print('Intercept value:', lm.intercept_)
-\end{lstlisting}\\
-\end{tabular}
-\label{tab:model-perf}
-\end{table}
-
-The performance is printed to compare against other models/variations in the experimental model development phase; during the continuous experimentation phase, the author might change parameters of the model or change the data (engineer features, etc) and then re-run the model to check if the performance improved
-
-Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
-Besides \emph{accuracy}, we also see checks for the \emph{Root Mean Square Error (RMSE)} and the use of the classification report provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html}.
-In addition to the accuracy, the classification report also reports the \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
-
-\emph{P54} prints the intercepts of a linear regression model.
-Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
-This insight is critical for accurate model interpretation and effective decision-making in practical applications.
-
-\begin{wrapfigure}{l}{0.45\linewidth}
-\includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}
-\caption{Confusion matrix}
-\label{fig:model-perf-confusion-matrix}
-\end{wrapfigure}
-
-Figure~\ref{fig:model-perf-confusion-matrix} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
-The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
-
-% TODO: this looks horrible if we are wrapping text around the figure above
-% \begin{lstlisting}[caption={Spot-check the performance of an image classification model for a random sample of input images.}, label={lst:model-perf-custom}]
-% def spot_check_recs(classifier, seed):
-%     random_images = []
-%     random_labels = []
-%     random_loc = np.random.randint(3000)
-%     distance, loc = classifier.kneighbors( embedding_model.predict( visualize[random_loc].reshape(-1,28,28)), 100)
-
-%     random_images.append(visualize[random_loc])
-%     ...
-%     for rec_loc, dist in zip(loc.reshape(-1,1), distance.reshape(-1,1)):
-%         ...
-
-%     ImageUtil.grid_display(random_images, random_labels, no_of_columns=1, figsize=(2,2), fig_title = "recs")
-
-% spot_check_recs(classifier, 910)
-% \end{lstlisting}
-
-% Listing~\ref{model-perf-custom} shows a function implemented by a practitioners to manually evaluate an image classification model on a random sample of input images.
-The function prints the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
-
-\subsubsection{Neural network architecture ($N = 1$)}
-% (P92)
-
-\begin{lstlisting}[caption={Print the architecture of a neural network.}, label={lst:network-architecture}]
-print(MyNetwork)
-\end{lstlisting}
-
-Listing~\ref{lst:network-architecture} shows a print statement to show the architecture of a neural network.
+    P92&
+    \lstinline[]$print(MyNetwork)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:network-architecture}
+\end{table}
+
+Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.
+
 Printing the structure of a neural network serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
@@
 
+\begin{table}
+  \centering
+  \caption{Example of outputs used to check the shape of the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}} & \emph{\textbf{Code}}\\
+    \midrule
+    P4 &
+    \lstinline[]$print('no.of examples in test data : ', len(test_data))$\\
+
+    P32 &
+    \lstinline[]$print('Training set shape : ', x_train.shape)$\\
+
+    P117&
+    \lstinline[]$print('Y_train.shape: ', Y_train.shape)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:shape-check}
+\end{table}
+
+Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data.
+\emph{P4} prints the number of rows or examples in the test set.
+\emph{P32} and \emph{117} prints the number of rows and columns (or features) in the training and test data respectively.
+
 Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
-
 Primarily, the number of features in the training set must match those in the testing set.
@@
 
-\begin{table}
-\centering
-\begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-\toprule
-\emph{\textbf{Key}} & \emph{\textbf{Code}}\\
-\midrule
-P4 &
-\begin{lstlisting}
-print('no.of examples in test data : ', len(test_data))
-\end{lstlisting}\\
-P32 &
-\begin{lstlisting}
-print('Training set shape : ', x_train.shape) 
-\end{lstlisting}\\
-\end{tabular}
-\caption{Example of outputs used to check the shape of the data.}
-\label{tab:shape-check}
-\end{table}
-
-Table~\ref{tab:shape-check} shows two print statements discovered in this study which demonstrate routine checks performed by developers to ensure the integrity and applicability of their models.
 Such checks are indicative of the meticulous attention to detail necessary in the model training and validation process, underscoring the importance of shape validation in achieving reliable machine learning outcomes.
@@
 
+% TODO: the narrative is not as strong here; it would help to also show the actual output as well?
+For instance, in P43 by itself doesn't make that much sense, the output however shows that all features in the dataset are floats which is a pre-requisite for ML training data
+\begin{table}
+  \centering
+  \caption{Outputs used to check the data type of features in the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}} &
+    \emph{\textbf{Code}}\\
+
+    P43&
+    \lstinline[]$print('data type:', images.dtype)$\\
+
+    O71&
+    \lstinline[]$type(Y)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:type-check}
+\end{table}
+
+% TODO: for O71, the output shows its a ndarray; without showing this to the reader we cannot make a strong narrative around these type checks
+Table~\ref{tab:type-check} illustrates outputs of cells found in our analysis, where checks are employed to verify the data types of the features in the dataset.
+Print statement \emph{P43} is written prior to training a neural network.
+The output validates that all features in the training data are numerical.
+Similarly, output \emph{O71} shows the data type of the target column.
+
+Type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
 Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis (EDA) phase of developing machine learning models.
@@
 
-This is illustrated in our analysis, where checks such as \texttt{print('data type:', images.dtype)} and \texttt{type(Y)} are employed to verify the data types of variables.
-These type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
-
-This proactive approach in the EDA phase facilitates smoother transitions into model training and can significantly impact the performance and effectiveness of the final machine learning models.
-
 \subsubsection{Spot checks ($N = 5$)}
@@
   \centering
+  \caption{Example of outputs used to perform spot checks.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     O60 &
-\begin{lstlisting}
-X_pca.head()
-\end{lstlisting}\\
+    \lstinline[]$X_pca.head()$\\
 
     P64 &
-\begin{lstlisting}
-print(np.max(cur[:, :, 1]))
-\end{lstlisting}\\
+    \lstinline[]$print(np.max(cur[:, :, 1]))$\\
 
     P114 &
-\begin{lstlisting}
-print(onehot_encoded)
-\end{lstlisting}\\
-\end{tabular}
-\caption{Example of outputs used to perform spot checks.}
+    \lstinline[]$print(onehot_encoded)$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:spot-check}
@@
 
-Table~\ref{tab:spot-check} presents outputs which were used to spot check at various stages of the ML development cycle.
-Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
-These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
-
-% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
-
-In \emph{O60}, the analyst verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
-This is a critical step for maintaining dimensional consistency in feature-reduced datasets.
+Table~\ref{tab:spot-check} presents outputs which were used to perform spot checks at various stages of the ML development cycle.
+In \emph{O60}, the practitioner verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
 \emph{P64} demonstrates another case which is particularly relevant in image processing contexts.
@@
 
+Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
+These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
 Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
+
+% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
 
@@
   \centering
+  \caption{Example of outputs used to monitor model training.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     O8 &
-\begin{lstlisting}
-autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)
-\end{lstlisting}\\
+    \lstinline[]$autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)$\\
 
     O31 &
-\begin{lstlisting}
-adaBoost.fit(X_train, y_train)
-\end{lstlisting}\\
+    \lstinline[]$adaBoost.fit(X_train, y_train)$\\
 
     O42&
-\begin{lstlisting}
-m_r.best_params_
-\end{lstlisting}\\
-\end{tabular}
-\caption{Example of outputs used to monitor model training.}
+    \lstinline[]$m_r.best_params_$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:model-training}
 \end{table}
-
-The ``Model Training'' phase of ML workflows involves various critical activities, each aimed at ensuring effective model performance and robustness.
 
@@
 For instance, in output O8, the autoencoder is trained using a custom callback from \texttt{keras\_utils} to provide a progress update without cluttering the output stream, which is particularly useful in lengthy training sessions.
-In contrast, output \emph{O31} represents a crucial validation step where the adaBoost model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
-Output \emph{O42} possibly checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
+In contrast, output \emph{O31} represents a crucial validation step where the \texttt{adaBoost} model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
+Output \emph{O42} checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
 
@@
 
-\subsection{Explicit feedback using \texttt{assert} statements}
+\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}
 
@@
   \centering
+  \caption{Assertions used to validate the batch size of input data.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     A21 &
-\begin{lstlisting}
-assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'
-\end{lstlisting}\\
+    \lstinline[]$assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'$\\
 
     A28 &
-\begin{lstlisting}
-assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'
-\end{lstlisting}\\
+    \lstinline[]$assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'$\\
 
     A70 &
-\begin{lstlisting}
-assert n_img > batch_size
-\end{lstlisting}\\
-\end{tabular}
-\caption{Assertions used to validate the batch size of input data.}
+    \lstinline[]$assert n_img > batch_size$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:batch-size}
 \end{table}
+
+Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
+\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
+Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
+\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
 
@@
 Given these benefits, it becomes crucial to ensure that the batch size is appropriately set to match the hardware's capacity, avoiding memory overflow issues that can halt training.
-
-Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
-\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
-Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
-\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
-
 These assertions are strategically placed to prevent common errors that could undermine training efficiency and model performance, thereby safeguarding the training process from potential disruptions due to misconfigured batch sizes.
@@
 % (A33)
+
+\begin{table}
+  \centering
+  \caption{Assert used to prevent data leakage.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A33&
+    \lstinline[]$assert len(set( tr_df.PetID.unique()).intersection(valid_df.PetID.unique())) == 0$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:data-leakage}
+\end{table}
+
+Assertion \emph{A33} in Table~\ref{tab:data-leakage} ensures that there are no overlapping \texttt{PetID} values between the training set \texttt{tr\_df} and the validation set \texttt{valid\_df}, thereby preventing any possibility of data leakage.
 
@@
 To guarantee that the model can generalize effectively to new examples, it's crucial to verify that the training and validation sets are completely distinct with no shared examples.
-
-This is exemplified in assertion \emph{A33}, where the code \texttt{assert len(set( tr\_df.PetID.unique()).intersection(valid\_df.PetID.unique())) == 0} checks for this exact condition.
-This assertion ensures that there are no overlapping \texttt{PetID} values between the training set (\texttt{tr\_df}) and the validation set (\texttt{valid\_df}), thereby preventing any possibility of data leakage.
-
 By strictly separating these datasets, the validation phase serves its purpose of providing an unbiased evaluation of the model's ability to generalize---to data similar to, but not identical to that which it was trained on---thus maintaining the integrity and validity of the evaluation metrics derived during model testing.
-
-\subsubsection{Resource check ($N = 2$)}
-% (A18, A67)
-
-\todo{NOTE: connect this to~\ref{sec:implicit-resource-check}}
-
-In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
-Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
-
-This practice is illustrated by assertions such as those seen in \emph{A18} and \emph{A67}, where the code explicitly verifies the versions of critical libraries.
-For instance, \texttt{assert pd.\_\_version\_\_.rpartition(`.`)[0] == `1.0`} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
-
-These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
 
@@
   \centering
+  \caption{Assertions used to perform existence checks.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     A23 &
-\begin{lstlisting}
-assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())
-\end{lstlisting}\\
+    \lstinline[]$assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())$\\
 
     A42 &
-\begin{lstlisting}
-assert not lab_s.isnull().values.any()
-\end{lstlisting}\\
+    \lstinline[]$assert not lab_s.isnull().values.any()$\\
 
     A43 &
-\begin{lstlisting}
-assert len(data) != 0, 'cannot divide by zero'
-\end{lstlisting}\\
+    \lstinline[]$assert len(data) != 0, 'cannot divide by zero'$\\
 
     A50 &
-\begin{lstlisting}
-assert not np.any(np.isnan(X))
-\end{lstlisting}\\
+    \lstinline[]$assert not np.any(np.isnan(X))$\\
 
     A51 &
-\begin{lstlisting}
-assert data.target.notnull().all()
-\end{lstlisting}\\
+    \lstinline[]$assert data.target.notnull().all()$\\
 
     A63 &
-\begin{lstlisting}
-assert X.isnull().sum().sum() == 0
-\end{lstlisting}\\
+    \lstinline[]$assert X.isnull().sum().sum() == 0$\\
 
     A79 &
-\begin{lstlisting}
-assert not processed_data_df.isna().any().any()
-\end{lstlisting}\\
+    \lstinline[]$assert not processed_data_df.isna().any().any()$\\
 
     A86 &
-\begin{lstlisting}
-assert p0 in poi_info.index
-\end{lstlisting}\\
-\end{tabular}
-\caption{Assertions used to perform existence checks.}
+    \lstinline[]$assert p0 in poi_info.index$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:existence-check}
@@
 
-Existence checks are a fundamental component of data preprocessing in machine learning, serving multiple critical purposes to ensure data integrity before analysis and modeling.
+Assertions \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
+Assertion \emph{A43} guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
+Assertions \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
+Assertions \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
+
+Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modeling.
 These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
 Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values.
-
-For instance, assertions like \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
-Another common check, as seen with \emph{A43}, guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
-Similarly, checks such as \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
-Additionally, assertions like \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
-
-These checks are crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
+These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
 
@@
   \centering
+  \caption{Assertions used to validate mathematical properties of neural networks.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     A3 &
-\begin{lstlisting}
-assert (xH - wH) % self.stride == 0
-\end{lstlisting}\\
+    \lstinline[]$assert (xH - wH) % self.stride == 0$\\
 
     A25 &
-\begin{lstlisting}
-assert test_output.std() < 0.15, "Don't use batchnorm here"
-\end{lstlisting}\\
+    \lstinline[]$assert test_output.std() < 0.15, "Don't use batchnorm here"$\\
 
     A56 &
-\begin{lstlisting}
-assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))
-\end{lstlisting}\\
+    \lstinline[]$assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))$\\
 
     A64 &
-\begin{lstlisting}
-assert np.allclose(T, T.T)
-\end{lstlisting}\\
-\end{tabular}
-\caption{Assertions used to validate mathematical properties of neural networks.}
+    \lstinline[]$assert np.allclose(T, T.T)$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:maths-check}
@@
 
-When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigor is captured through specific assertions aimed at validating these properties post-matrix manipulations.
-
-Assertion \emph{A3} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
+Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
 \emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior.
@@
 
-These assertions are not just precautionary but are vital for confirming that the underpinnings of machine learning algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
+When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
+This rigor is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
+These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
 
@@
 
-\subsubsection{Resource check ($N = 5$)}
-% (A10, A14, A37, A60, A74)
+\subsubsection{Resource check ($N = 7$)}
+% (A10, A14, A18, A37, A60, A67 A74)
 
@@
     A10 &
-\begin{lstlisting}
-assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."
-\end{lstlisting}\\
+    \lstinline[]$assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."$\\
 
     A14 &
-\begin{lstlisting}
-assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'
-\end{lstlisting}\\
+    \lstinline[]$assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'$\\
+
+    A18&
+    \lstinline[]$assert pd.__version__.rpartition('.')[0] == '1.0', f"Unexpected pandas version: expected 1.0, got {pd.__version__.rpartition('.')[0]}"$\\
 
     A37 &
-\begin{lstlisting}
-assert svm.fit_status_ == 0, 'Forgot to train the SVM!'
-\end{lstlisting}\\
+    \lstinline[]$assert svm.fit_status_ == 0, 'Forgot to train the SVM!'$\\
 
     A60 &
-\begin{lstlisting}
-assert f2.gca().has_data()
-\end{lstlisting}\\
+    \lstinline[]$assert f2.gca().has_data()$\\
+
+    A67&
+    \lstinline[]$assert pm.__version__ == '3.9.2'$\\
 
     A74 &
-\begin{lstlisting}
-assert os.path.exists(image_dir)
-\end{lstlisting}\\
+    \lstinline[]$assert os.path.exists(image_dir)$\\
+    \bottomrule
   \end{tabular}
   \caption{Assertions used to validate availability of resources.}
-\label{tab:resource-check-explicit}
-\end{table}
-
-Ensuring the availability and integrity of required resources is a critical aspect of any machine learning workflow.
-Table~\ref{tab:resource-check-explicit} highlights some of the assertions we encountered that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
+  \label{tab:assert-resource-check}
+\end{table}
+
+Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
 These checks help prevent runtime errors and ensure the proper execution of code.
@@
 One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
-These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing files.
+These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing resources.
 
@@
 
-Furthermore, assertions are used to verify the presence of data in visualizations generated using libraries like Matplotlib (\emph{A60}).
+Assertion \emph{A60} is used to verify the presence of data in visualizations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
 These checks are crucial for ensuring that the visualizations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
 
+Assertions \emph{A18} and \emph{A67} explicitly verify the versions of critical libraries.
+For instance, \emph{A18} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
+
 Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
+
+% NOTE: left over loads, move to implications if required
+% In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
+Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
+
+% These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
 
@@
 % NOTE: breck2017ml proposes feature and data testing; we can say that the assertions in the notebook bring the system closer to migrating them into dedicated data validation systems
+
+% NOTE: perhaps also argue the other side, the demographic using notebooks do not have SE training; they are perhaps directly sharing the notebook and the data directly...
+P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
+This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
+In this case study, managing dependencies through external tools is recommended.
+Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
+Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
 

### 2024-05-23 report/emse24.tex
feat(report): emse24 updates

@@
 
-\todo{TODO: justify why we didn't collect failures (strerr); that is more debugging? not the focus of this paper?}
 Similarly, for implicit feedback, we first isolate code cells that produce an output.
@@
 
-\subsubsection{Model performance check ($N = 11$)}
+\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
 % (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
@@
   \centering
+  \caption{Assertions used to check performance of ML models.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
@@
     A7 &
-    \begin{lstlisting}
-assert len(neighbours_1) == 20, "Neighbors don't match!"
-    \end{lstlisting}\\
+    \lstinline[]$assert len(neighbours_1) == 20, "Neighbors don't match!"$\\
 
     A15 &
-    \begin{lstlisting}
-assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))
-    \end{lstlisting}\\
+    \lstinline[]$assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))$\\
 
     A19 &
-    \begin{lstlisting}
-assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'
-    \end{lstlisting}\\
+    \lstinline[]$assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'$\\
 
     A38 &
-    \begin{lstlisting}
-assert 0.75 < auc(fpr, tpr) < 0.85
-    \end{lstlisting}\\
+    \lstinline[]$assert 0.75 < auc(fpr, tpr) < 0.85$\\
 
     A58 &
-    \begin{lstlisting}
-assert np.isclose(accuracy, 0.9666666666666667)
-    \end{lstlisting}\\
-  \end{tabular}
-  \caption{Assertions used to check performance of ML models.}
+    \lstinline[]$assert np.isclose(accuracy, 0.9666666666666667)$\\
+    \bottomrule
+  \end{tabular}
   \label{tab:model-perf-explicit}
@@
 
-For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors, an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
+% TODO: verify if this applies to recommender systems as well?
+For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
 Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
 Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity.
-Additionally, range checks like \emph{A38} confirm that the performance metrics such as AUC and Gini coefficient fall within acceptable boundaries, indicating good predictive performance without overfitting.
-Additionally, the use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
+Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.
+
+The use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
 This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
@@
 
-Each assertion acts as a crucial checkpoint that verifies the model's ability to generalize and perform effectively, ensuring that the outcomes are both robust and trustworthy.
-
 \subsubsection{Network architecture check ($N = 3$)}
@@
 
-% TODO: revisit this section; we have new assertions
+\begin{table}
+  \centering
+  \caption{Assertions used to check the architecture of neural networks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A11 &
+    \lstinline[]$assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()$\\
+
+    A62 &
+    \lstinline[]$assert activations[i + 1].shape[1] == self.encoder_layer_sizes[-1] * 2$\\
+
+    A75 &
+    \lstinline[]$assert reg in ['none', 'l2']$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-network-architecture}
+\end{table}
+
+Assertion \emph{A11} verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
+The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
+By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
+
+Assertion \emph{A62} checks that the shape of the activations at index \lstinline{i + 1} has the correct number of channels, which should be twice the size of the last encoder layer.
+This ensures that the intermediate activations maintain the expected structure necessary for the proper functioning of the subsequent layers in the network.
+Ensuring that the shape of activations is consistent throughout the network is crucial for maintaining the integrity of the data flow and preventing errors during the forward pass.
+
+Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'.
+This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behavior or suboptimal model performance.
+
 When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
 This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
-
-We encountered an \texttt{assert} statement that addresses this architectural compatibility check: \lstinline{assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()}.
-This assertion verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
-The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
-By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
-
-This architectural check is crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
+Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
 
@@
 
-Data shape checks can be considered the ``Swiss army knife''' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
+% TODO: revisit the assertions in this section; we may have misclassified assertions or need to create a new group
+
+\begin{table}
+  \centering
+  \caption{Assertions used to check the shape of the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A5 &
+    \lstinline[]$assert y_valid.shape == (1132,)$\\
+
+    A17 &
+    \lstinline[]$assert X.shape[1] == 13, 'Did you drop/lose some columns in X? Did you properly load and split the data?'$\\
+
+    A29 &
+    \lstinline[]$assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'$\\
+
+    A31 &
+    \lstinline[]$assert img.shape == (112, 92)$\\
+
+    A76 &
+    \lstinline[]$assert len(encoding['token_type_ids']) == max_seq_length$\\
+
+    A84 &
+    \lstinline[]$assert red.get_shape().as_list()[1:] == [224, 224, 1]$\\
+
+    A90 &
+    \lstinline[]$assert len(X_train) == 2000$\\
+
+    A93 &
+    \lstinline[]$assert temp_embed.shape[0] == stride$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-shape-check}
+\end{table}
+
+Assertions \emph{A17} and \emph{A90} verify that the input features have the correct number of columns and examples, respectively.
+Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
+
+Assertions \emph{A5} and \emph{A29} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
+
+In the context of computer vision tasks, assertions \emph{A31} and \emph{A84} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
+
+Furthermore, assertions \emph{A76} and \emph{A93} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
+
+Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
 We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features, labels, predictions, images, sequences, and embeddings.
@@
 
-For instance, assertions like \emph{A17}:~\lstinline{assert X.shape[1] == 13, 'Did you drop/lose some columns in X?
-Did you properly load and split the data?'} and \emph{A90}:~\lstinline{assert len(X_train) == 2000} verify that the input features have the correct number of columns and examples, respectively.
-Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
-
-Similarly, assertions like \emph{A5}:~\lstinline{assert y_valid.shape == (1132,)} and \emph{A29}~\lstinline{assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
-
-In the context of computer vision tasks, assertions like \emph{A31}:~\lstinline{assert img.shape == (112, 92)} and \emph{A84}~\lstinline{assert red.get_shape().as_list()[1:] == [224, 224, 1]} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
-
-Furthermore, assertions like \emph{A76}:~\lstinline{assert len(encoding['token_type_ids']) == max_seq_length} and \emph{A93}:~\lstinline{assert temp_embed.shape[0] == stride} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
-
 Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
@@
 
-Type checking assertions play a crucial role in maintaining the integrity and consistency of data types throughout the machine learning workflow.
-In the examined notebooks, we encountered several assert statements that verify the types of input features, models, data structures, and other objects.
-
-One common type check is performed on the input features, as demonstrated by the assertion \emph{A2}:~\lstinline{assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'}.
+\begin{table}
+  \centering
+  \caption{Assertions used to check the data type of features in the dataset.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A2 &
+    \lstinline[]$assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'$\\
+
+    A35 &
+    \lstinline[]$assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"$\\
+
+    A40 &
+    \lstinline[]$assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)$\\
+
+    A81 &
+    \lstinline[]$assert is_all_ints(filled_df[r]) is True$\\
+
+    A88 &
+    \lstinline[]$assert isinstance(betas, np.ndarray)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-type-check}
+\end{table}
+
+Assertion \emph{A2} in Table~\ref{tab:assert-type-check} demonstrates a very common type check performed on the input features.
 This assertion ensures that the input features are represented as PyTorch float tensors, which is a requirement for many deep learning models and operations.
@@
 Another type check is applied to verify the type of the machine learning models themselves.
-For instance, the \emph{A40}:~\lstinline{assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
+For instance, the \emph{A40} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
 This type of assertion is essential when working with different types of models, as it helps prevent errors arising from using incompatible methods or attributes.
@@
 In addition to input features and models, type checks are also performed on data structures and intermediate objects.
-The assertion \emph{A35}:~\lstinline{assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
+The assertion \emph{A35} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
 
 Type checking is not limited to objects and data structures--it can also be applied to individual columns or elements within a dataset.
-The assertion \emph{A81}:~\lstinline{assert is_all_ints(filled_df[r]) is True} checks if all the elements in a specific column (\lstinline{filled_df[r]}) of a DataFrame are integers.
-
-Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}:~\lstinline{assert isinstance(betas, np.ndarray)}.
-This assertion ensures that the \texttt{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
+The assertion \emph{A81} checks if all the elements in a specific column (\lstinline[]$filled_df[r]$) of a DataFrame are integers.
+
+Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}.
+This assertion ensures that the \lstinline{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
 
@@
 
-\subsubsection{Data validation check ($N = 11$)}
-% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A92)
+\subsubsection{Data validation check ($N = 14$)}
+% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to perform data validation checks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A41 &
+    \lstinline[]$assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))$\\
+
+    A44 &
+    \lstinline[]$assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1$\\
+
+    A46 &
+    \lstinline[]$assert np.isclose(stdev_norm, 1.0, atol=1e-16)$\\
+
+    A52 &
+    \lstinline[]$assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()$\\
+
+    A65 &
+    \lstinline[]$assert np.all(y <= nb_classes)$\\
+
+    A73 &
+    \lstinline[]$assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:data-validation}
+\end{table}
 
 % TODO: revisit, we have new assertions here; update the count
-we encountered a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
+Table~\ref{tab:data-validation} presents a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
 These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
@@
 One type of validation check involves verifying the presence of specific values or ranges within arrays or columns.
-For example, the assertion \emph{A41}:~\lstinline{assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))} checks if the unique values in the \texttt{smoke} column of the DataFrame \texttt{X} are limited to 0 and 1.
-Similarly, \emph{A65}~\lstinline{assert np.all(y <= nb_classes)} ensures that all values in the array \texttt{y} are less than or equal to a certain number of classes.
+For example, assertion \emph{A41} checks if the unique values in the \lstinline{smoke} column of the DataFrame \lstinline{X} are limited to 0 and 1.
+Similarly, \emph{A65} ensures that all values in the array \lstinline{y} are less than or equal to a certain number of classes.
 
 Another common validation check is verifying the equality or closeness of values or arrays.
-The assertion \emph{A44}:~\lstinline{assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1} checks if the difference between the minimum and maximum values in a specific range of the \texttt{entity\_embedding} array is less than or equal to 1.
-Similarly, \emph{A46}:~\lstinline{assert np.isclose(stdev_norm, 1.0, atol=1e-16)} verifies that the \texttt{stdev\_norm} value is close to 1.0 within a specified absolute tolerance.
+Assertion \emph{A44} checks if the difference between the minimum and maximum values in a specific range of the \lstinline{entity_embedding} array is less than or equal to 1.
+Assertion \emph{A46} verifies that the \lstinline{stdev_norm} value is close to 1.0 within a specified absolute tolerance.
 
 Some assertions focus on validating the uniqueness or cardinality of values in arrays or DataFrames.
-For instance, \emph{A52}:~\lstinline{assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()} checks if the number of unique user IDs in the \texttt{grouped\_users} DataFrame is equal to the number of unique user IDs in the \texttt{user\_engagement} DataFrame.
+For instance, \emph{A52} checks if the number of unique user IDs in the \lstinline{grouped_users} DataFrame is equal to the number of unique user IDs in the \lstinline{user_engagement} DataFrame.
 
 Other assertions validate the presence of specific values or conditions within DataFrames or other data structures.
-The assertion \emph{A73}:~\lstinline{assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \texttt{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
+The assertion \emph{A73} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \lstinline{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
 
@@
 % NOTE: breck2017ml proposes feature and data testing; we can say that the assertions in the notebook bring the system closer to migrating them into dedicated data validation systems
+
+% NOTE: refer back to sec:assert-model-perf; the use of np.isclose and robustness to stochastic nature
 

### 2024-05-24 report/emse24.tex
feat(report): emse24 updates

@@
 \usepackage{tabularx}
+\usepackage{longtable}
 \usepackage{booktabs}
@@
 
-\subsection{Assertions derived from visualisations}
-% NOTE: following text is from the ICST paper; still need to think about how to integrate it here...
-
-\subsubsection{What patterns are frequently observed in VA pairs used to test ML systems?}
-
-We further analysed the 269 VA pairs obtained from RQ1, and identified three testing patterns which are presented in Table~\ref{tab:testing-patterns} and discussed further below.
-
-\begin{table}
-  \centering
-  \caption{Testing patterns observed in the visualisation-assertion pairs.}
-  \begin{tabularx}{0.45\textwidth}{@{}l X r r@{}}
-    \toprule
-    \textbf{ID} &
-    \textbf{Description} &
-    \multicolumn{2}{c}{\textbf{Count}}\\
-    & & Yes & No\\
-    \midrule
-    \textbf{P1} &
-    Does the assertion use a comparison operator (such as \emph{eq, lt, gt})? &
-    202 & 67\\
-    \textbf{P2} &
-    Does the assertion use magic thresholds? &
-    101 & 168\\
-    \textbf{P3} &
-    Does the assertion use an external testing library (such as \texttt{numpy.testing})? &
-    42 & 227\\
-    \bottomrule
-  \end{tabularx}
-  \label{tab:testing-patterns}
-\end{table}
-
-\begin{table}
-  \centering
-  \caption{Most frequently occurring assertions that use a comparison operator.}
-  \begin{tabular}{@{}m{0.9\linewidth}@{}}
-    \toprule
-    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
-assert accuracy_score(y_test, pred) > 0.95\end{lstlisting}\\
-    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
-assert HISTORY['test_accuracy'][-1] > 0.92\end{lstlisting}\\
-    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
-assert np.mean(mean_rw_history[-10:]) > 10.
-print("That's good enough for tutorial.")\end{lstlisting}\\
-    \begin{lstlisting}[language=Python,belowskip=0pt,aboveskip=0pt]
-assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."\end{lstlisting}\\
-    \begin{lstlisting}[language=python,belowskip=0pt,aboveskip=0pt]
-assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."
-    \end{lstlisting}\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:compare-op-asserts}
-\end{table}
-
-In a total of 269 VA pairs, 202 assertions (75\%) contain a comparison operator such as \emph{equal to}, \emph{lesser than} or \emph{greater than}.
-Table~\ref{tab:compare-op-asserts} presents the most frequently observed assertions that conform to this pattern.
-
-101 of 269 assertions (37\%) compare the value of a performance metric---typically the accuracy or recall---against a ``magic threshold'' defined by the notebook author.
-This covers 46\% of the assertions with a comparison operator.
-It is unclear however, how the specified thresholds were obtained since the authors do not provide additional explanation in the notebook.
-The threshold could have been derived based on domain expertise.
-Another alternative could be to define them during the Requirements Engineering phase, through discussions with all involved stakeholders\cite{vogelsang2019requirements,rahimi2019toward}.
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{../catalogue/select-16.png}
-  \caption{Visualisation of the training loss of an ML model.}\label{fig:loss}
-\end{figure}
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that the mean of the first 10 observations of the loss is higher than the mean of the last 10 observations.
-In other words, the assertion checks if the loss function is converging to an optimal minima.}, label={lst:loss}]
-assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."
-\end{lstlisting}
-
-In contrast to the assertions presented in Table~\ref{tab:compare-op-asserts}, Listing~\ref{lst:loss} presents an assertion that ensures that the model performs well, without using a magic threshold.
-The assertion checks if the mean of the first 10 observations of the loss is higher compared to the mean of the last 10 observations.
-In other words, the assertion ensures that the loss of the model decreased after training, and that the model was able to ``learn'' from the training set.
-The assertion is derived from the visualisation presented in Figure~\ref{fig:loss}, which shows the loss of the model over the training epochs.
-
-\begin{figure}
-  \subcaptionbox{Histogram showing the distribution of a feature before and after scaling.\label{fig:scale}}{\includegraphics[width=\linewidth]{../catalogue/select-65a.png}}
-  \subcaptionbox{Histogram showing the distribution of a feature before and after normalisation.\label{fig:normal}}{\includegraphics[width=\linewidth]{../catalogue/select-65b.png}}
-  \caption{Visualisation of data before and after various data pre-processing techniques.}
-  \label{fig:data-pre-process}
-\end{figure}
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that the mix and max of a feature fall within specified threshold derived from the visualisation presented in Figure~\ref{fig:scale}.}, label={lst:scale}]
-assert scaled_data[scaled_data.argmax()] <= 1
-assert scaled_data[scaled_data.argmix()] >= 0
-\end{lstlisting}
-
-\begin{lstlisting}[language=Python, caption={Similar premise as Listing~\ref{lst:scale}, however this assertion is based on Figure~\ref{fig:normal}.}, label={lst:normal}]
-assert normalized_data[normalized_data.argmax()] < 10
-assert normalized_data[normalized_data.argmin()] > -7
-\end{lstlisting}
-
-Visualisations in Figure~\ref{fig:data-pre-process} were obtained from a notebook that presents a tutorial on scaling and normalization---two common data pre-processing techniques in ML.
-The visualisations show a comparison of the data before and after each pre-processing technique is applied to a feature from the dataset.
-The author performs a visual check, by comparing the distribution of the data before and after applying the pre-processing technique, using a histogram.
-
-Listing~\ref{lst:scale} and~\ref{lst:normal} show the assertions that were defined below Figure~\ref{fig:scale} and~\ref{fig:normal} respectively.
-Both assertions check that the min and max of the feature is within the expected bounds after applying the respective pre-processing technique.
-In contrast to assertions presented in Table~\ref{tab:compare-op-asserts}, it is clear that the thresholds are derived from the corresponding visualisations.
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that an augmented image is different from the original.}, label={lst:no-comp-magic}]
-assert not np.allclose(x_aug, x_original)
-\end{lstlisting}
-
-22\% of the assertions did not use a comparison operator, nor a magic threshold.
-Listing~\ref{lst:no-comp-magic} highlights such an assertion which is obtained from a notebook that performs image augmentation using Weighted K-Nearest Neighbours.
-The assertion compares the matrix representation of two images, and checks that the augmented image is different from the original.
-
-Most notebooks examined in this study do not use any external testing libraries.
-227 or 84\% of the VA pairs use the low-level Python \texttt{assert} statement to define the test directly.
-This aligns with our observations that majority of the assertions compare the value of a performance metric against a user defined threshold, and do not require the use of a specialised testing library.
-
-This study found 42 VA pairs where an external testing library was used.
-Listing~\ref{lst:np-almost-equal} shows an assertion that uses the \texttt{assert\_almost\_equal} method from the \texttt{numpy.testing} module.
-The assertion checks if the loss of the model is similar to the specified value, up to a desired precision.
-Due to the stochastic nature of ML, results obtained from two separate development cycles will slightly differ.
-The assertion presented in Listing~\ref{lst:np-almost-equal} is found to be robust towards such perturbations and will only fail if there is a significant change in the performance of the model.
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that the loss of the model is similar to the specified value.}, label={lst:np-almost-equal}]
-ans4 = compute_loss(X_expanded, y, w)
-np.testing.assert_almost_equal(ans4, 0.3042764) 
-\end{lstlisting}
-
-\subsubsection{Do the assertions capture all the information obtained from the corresponding visualisation?}
-We performed an in-depth analysis of 34 unique VA pairs.
-This was conducted by reading and understanding the entire contents of the notebooks in a top-down order.
-During the analysis, several data points such as the primary objective of the notebook, the type of ML problem, the type of ML model and the type of data were captured.
-We find several instances of VA pairs where the assertion is unable to capture all the information obtained from the visualisation.
-
-\begin{figure}
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a linear kernel.\label{fig:svm-linear}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-linear.png}}
-  \hfill
-  \subcaptionbox{Visualisation of the decision boundary of a SVM with a RBF kernel.\label{fig:svm-rbf}}{\includegraphics[width=0.49\linewidth]{../catalogue/select-04-rbf.png}}
-  \caption{Visualisation of the decision boundary of a SVM trained using different kernels.}\label{fig:svm}
-\end{figure}
-
-\begin{lstlisting}[language=Python, caption={Assertion on the accuracy of the ML model.}, label={lst:svm}]
-assert accuracy_score(y_test, pred) > 0.95
-\end{lstlisting}
-
-Figure~\ref{fig:svm} and Listing~\ref{lst:svm} present one such pair.
-Visualisations~\ref{fig:svm-linear} and~\ref{fig:svm-rbf} are obtained from a notebook that performs binary classification using Support Vector Machines (SVM).
-The visualisations show the two classes present in the training set, and the decision boundary of the model when trained first using a linear kernel and later with a RBF kernel.
-
-The visualisations contain several useful information that can be used to test and debug ML models.
-For instance, using Visualisation~\ref{fig:svm}, ML practitioners can get a visual understanding of how well the model is able to classify the two classes in the dataset.
-Additionally, observing the complexity of the decision boundary is a quick way to determine if the model is under or overfitting the data.
-The visualisation also allows comparison of different ML models or variants, to iteratively find a model and hyper-parameters that best fits the underlying training data.
-In this case for instance, the author of the notebook first trained the SVM with a linear kernel.
-And upon realising that the data is not linearly separable, chose the RBF kernel.
-
-Listing~\ref{lst:svm} shows the corresponding assertion that was defined below the visualisations.
-The assertion checks if the model accuracy is above a threshold specified by the author.
-The assertion however, is only able to capture the information pertaining to the performance of the model.
-While the rest of the information that is gained from the qualitative assessment using visualisation is lost.
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that the Coefficient of Determination ($R^2$) is higher than the specified threshold.}, label={lst:r2}]
-r2_gru = r2_score(y_test, y_pred)
-assert r2_gru > 0.6
-\end{lstlisting}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{../catalogue/select-332a.png}
-  \caption{Scatter plot of the actual and predicted labels of a dataset. The visualisation shows that there is a linear relationship between the true and predicted labels.}\label{fig:r2}
-\end{figure}
-
-Figure~\ref{fig:r2} shows a visualisation obtained from a notebook that trains a RNN to predict the levels of Nitrogen Dioxide ($NO_2$) in the next 24 hour window.
-The visualisation shows a scatter plot of the real and predicted values of $NO_2$ from the test set.
-A linear relationship between the ground truth and predictions is observed, indicating that the model was able to learn certain patterns from the training data.
-The spread of $y_{pred}$ at any given $y_{true}$ is wide.
-For example, at $y_{true} = 80$, the prediction can range anywhere between 30 and 120.
-This indicates that the model was not able to fit the training data that well, and can be fine-tuned.
-
-Listing~\ref{lst:r2} presents the corresponding analytical test which was defined below Figure~\ref{fig:r2}.
-The \texttt{r2\_score} method fits a linear regression model on the true and predicted labels and computes the Coefficient of Determination ($R^2$) based on the residuals of the linear regression model.
-The assertion checks that the $R^2$ score of the model is always higher than $0.6$.
-Since the best score possible is $1.0$, the specified threshold indicates that the author wants to ensure that $y_{pred}$ and $y_{true}$ are always linearly related to each other.
-
-Similar to Figure~\ref{fig:svm}, the visualisation in Figure~\ref{fig:r2} allows ML practitioners to visually assess if the model is under or overfitting the training data.
-If the spread of $y_{pred}$ is wide, it indicates that the model is under-fitting.
-In contrast, if the spread is very close to the diagonal line, it indicates that the model is most likely overfitting.
-The visualisation can also be used to verify if there are any outliers in the predictions, for a given input.
-The assertion in Listing~\ref{lst:r2} however, does not capture such information.
-It can only check if the performance of the model is better than a specific threshold.
-
-\begin{lstlisting}[language=Python, caption={Assertion to check that each feature in a dataset is normal using the Kolmogorov-Smirnov test for goodness of fit from the scipy library.}, label={lst:kstest}]
-for feature in range(data_transformed.shape[1]):
-  assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2
-\end{lstlisting}
-
-\begin{figure}
-  \includegraphics[width=\linewidth]{../catalogue/select-152a.png}
-  \caption{Visualisation of the distribution of a feature from the dataset.}\label{fig:kstest}
-\end{figure}
-
-In contrast to Listing~\ref{lst:svm} and Listing~\ref{lst:r2}, Figure~\ref{fig:kstest} and Listing~\ref{lst:kstest} present a VA pair where the assertion captures all the information obtained from the visualisation.
-
-The notebook author normalises the training data and performs a visual check using the visualisation presented in Figure~\ref{fig:kstest}.
-The visualisation shows a bell-shaped curve thus indicating that the data is normal and that the transformation was successful.
-The assertion in Listing~\ref{lst:kstest} formally verifies that all features in the dataset have a normal distribution.
-The assertion iterates through each column in the dataset and performs the Kolmogorov-Smirnov test to check that the distribution of the feature is similar to that of a normal distribution.
-
-Checking the distribution of the training data is a very common task performed during the data understanding or exploratory data analysis phase of the ML development lifecycle.
-Several ML models assume normality in the underlying training data.
-However, in a production environment where ML models are continually trained using new batches of data, this assumption of normality may not always hold.
-The formal assertion presented in Listing~\ref{lst:kstest} makes this assumption explicit.
-And causes the automated ML pipeline to fail if our expectations regarding the data are no longer satisfied.
+\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}
+
+\subsubsection{Model Performance}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used to validate the performance of a model.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used to validate the performance of a model (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA2 &
+  \includegraphics[width=\linewidth]{va2.png} &
+  \lstinline[]$assert final_score > 300, 'not good enough for DQN'$\\
+
+  VA3 &
+  \includegraphics[width=\linewidth]{va3.png} &
+  \lstinline[]$assert accuracy_score(y_test, pred) > 0.95$\\
+
+  VA9 &
+  \includegraphics[width=\linewidth]{va9.png} &
+  \lstinline[]$assert np.mean(mean_rw_history[-10:]) > 10.$\\
+
+  VA11 &
+  \includegraphics[width=\linewidth]{va11.png} &
+  \lstinline[]$assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."$\\
+
+  VA15 &
+  \includegraphics[width=\linewidth]{va15.png} &
+  \lstinline[]$assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."$\\
+
+  VA28 &
+  \includegraphics[width=\linewidth]{va28.png} &
+  \lstinline[]$assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."$\\
+
+  VA68 &
+  \includegraphics[width=\linewidth]{va68.png} &
+  \lstinline[]$np.testing.assert_almost_equal(ans4, 0.3042764)$\\
+
+  VA107 &
+  \includegraphics[width=\linewidth]{va107.png} &
+  \lstinline[]$assert r2_gru > 0.6$\\
+
+\end{longtable}\label{tab:va-model-perf}
+
+The visualizations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
+VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
+These metrics collectively indicate the model's learning progress and stability during training.
+The visualization in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
+The general upward trend indicates learning and improvement in gameplay over time.
+VA15 presents the training and development loss over time for a character-level language model, showing a gradual decrease in loss and suggesting effective learning.
+
+The visualization in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
+This visualization aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
+The visualizations in VA28 display the training loss and development recall, respectively.
+The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
+The visualizations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
+The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
+In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
+
+The corresponding assertions for these visualizations ensure that the models meet specific performance standards.
+For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
+In VA9, the assertion checks whether the mean of the last ten rewards exceeds 10, ensuring stable and satisfactory learning progress.
+For VA15, the assertion checks if the mean of the first ten training observations is greater than the mean of the last ten, ensuring loss convergence and validating the model's learning without overfitting.
+The assertion in VA3 ensures the accuracy score on the test set exceeds 0.95, confirming the model's reliability and generalization capability.
+For VA28, the assertion checks whether the average development recall in the last ten evaluations exceeds 0.85, ensuring high recall rate critical for tasks like medical diagnostics.
+In VA11, the assertion ensures the average development BLEU score in the last ten evaluations exceeds 35, confirming the model produces high-quality translations.
+Lastly, in VA107, the assertion checks whether the R2 score exceeds 0.6, validating the model's fit and reliability for making accurate predictions in real-world scenarios.
+These validations confirm that iterative training processes lead to meaningful improvements and ensure the models' readiness for deployment and practical applications.
+
+\subsubsection{Distribution Check}
+
+\begin{table}
+  \centering
+  \caption{VA pairs used to check data distribution.}
+  \begin{tabular}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Visual Output}}&
+    \emph{\textbf{Assertion}}\\
+    \midrule
+
+    VA13 &
+    \includegraphics[width=\linewidth]{va13.png} &
+    \lstinline[]$assert 80 < boot_mean_mean < 81$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:va-dist-check}
+\end{table}
+
+The visualization in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
+
+Bootstrapped sample means are calculated by repeatedly resampling the original dataset with replacement and computing the mean for each resampled dataset.
+This process generates a distribution of the mean, known as the bootstrap distribution, which helps in estimating the variability and central tendency of the original dataset's mean.
+
+The corresponding assertion checks whether the bootstrapped mean lies between 80 and 81, ensuring that the statistical estimation is consistent with expected values.
+This validation is crucial for confirming the reliability of the bootstrapped estimates and the accuracy of statistical conclusions.
+
+\subsubsection{Data Validation}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used for data validation.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used for data validation (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA30 &
+  \includegraphics[width=\linewidth]{va30.png} &
+  \lstinline[]$assert scaled_data[scaled_data.argmax()] <= 1$\\
+
+  V35 &
+  \includegraphics[width=\linewidth]{va35.png} &
+  \lstinline[]$assert(all(df_lower['inspection_score'] > 75))$\\
+
+  V57 &
+  \includegraphics[width=\linewidth]{va57.png} &
+  \lstinline[]$assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2$\\
+
+  VA62 &
+  \includegraphics[width=\linewidth]{va62.png} &
+  \lstinline[]$assert not np.allclose(x_aug, x_original)$\\
+
+\end{longtable}\label{tab:va-data-validatio}
+
+The VA pairs in Table~\ref{tab:va-data-validatio} collectively illustrate various data validation techniques essential for ensuring the integrity and quality of data used in ML models.
+
+The visualization in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
+The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
+The scaling process adjusts the data to fall within a specified range of 0 and 1.
+The corresponding assertion checks whether the maximum value in the scaled data is less than or equal to 1, ensuring that the scaling procedure was applied correctly.
+
+The visualization in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
+The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
+This comparison helps identify any disparities in health inspection scores based on income levels.
+The corresponding assertion checks that all inspection scores in the lower-income areas are above 75, ensuring that the data reflects a minimum standard of health inspection scores in these areas.
+
+The visualization in VA57 shows a histogram of transformed data after it has been normalised.
+The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
+
+The visualization in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
+This heatmap helps in visualizing the differences and similarities between the original and augmented data.
+The corresponding assertion checks that the augmented data \lstinline{x_aug} is not nearly identical to the original data \lstinline{x_original}, ensuring that the data augmentation process has introduced sufficient variability.
+
+These validations are crucial for confirming that data preprocessing techniques such as scaling, transformation, and augmentation have been correctly applied.
+Ensuring the integrity and quality of the data is essential for the subsequent steps in the machine learning pipeline, as it directly impacts the robustness, reliability, and generalization of the models developed.
+By validating these steps, we can maintain high standards of data quality and model performance, which are fundamental for accurate and meaningful analyses.
 

### 2024-06-03 report/emse24.tex
feat(report): emse24 updates

@@
 \usepackage{wrapfig}
+\usepackage{framed}
+\newcommand{\highlight}[1]{\begin{framed}%
+  \noindent#1
+\end{framed}}
 
@@
   captionpos=b,
-  numbers=left,
-  numberstyle=\tiny\color{gray},
-  backgroundcolor=\color{backcolor},
+  % numbers=left,
+  % numberstyle=\tiny\color{gray},
+  % backgroundcolor=\color{backcolor},
   aboveskip=0pt,
@@
 
-\subsubsection{Motivating Example}
+\subsection{Motivating Example}
 
@@
 
-\subsubsection{Research Questions}
+\subsection{Exploratory Analysis}
+
+We undertook an exploratory analysis of the collected data to formulate our research questions.
+
+\subsection{Research Questions}
 
@@
 
-\todo{TODO: might need a visual aid here}
+% TODO: might need a visual aid here
 
@@
 
-\todo{TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
-Thats why we need to apply these techniques to surface unique and interesting candidates!}
+% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
+Thats why we need to apply these techniques to surface unique and interesting candidates!
 
@@
 
-\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
-
-\subsubsection{Data distribution check ($N = 7$)}
+% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output.
+Somethimes actual output is not possible (eg.
+P68) because the statement may be inside an if statement.
+
+\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
+
+\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
 % (O2, O4, O9, O14, O20, O25, O48)
 
-% TODO: we refer to these as outputs but they are actually the python statements that produce the outputs...do we need better terminology here?
-To me this is especially confusing because in the figure below we are actually showing the output, not the python statement (which makes sense, its more interesting to show the actual visualisation in these cases rather than the code).
 \begin{table}
@@
     \lstinline[]$sns.countplot(house_pred['OverallQual'])$&
-\\
+    \includegraphics[width=\linewidth]{o25.png}\\
 
@@
 
-% TODO: this is probably implications/discussions material?
 Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
@@
 
-\subsubsection{Data relationship check ($N = 2$)}
+\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
+These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}
+
+\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
 % (O6, O10)
@@
     \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
+    \emph{\textbf{Output}}\\
     \midrule
@@
 
+\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
+Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
+These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
+
 \subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
@@
 % NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
-Table~\ref{tab:resource-check} presents case studies where the output is used to verify availability of resources on the system in which the notebook is being executed.
+Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed.
 Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU.
@@
 These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.
-Additionally, output \emph{P71} is used to ensure that a certain version of an external library is present on the system.
-
-The print statement P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
-This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
-In this case study, managing dependencies through external tools is recommended.
-Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
-Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
+
+Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
+This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
+
+\highlight{\textbf{Finding 1.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
+These checks include confirming the availability of compute resources like GPUs and TPUs (P68, P82 and P86), ensuring datasets or pre-trained models are successfully loaded (P107, O64 and O66), and verifying the presence of specific library versions (P71).
+These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
 
@@
     \lstinline[]$print('Total Run Time:')$\\
+    \bottomrule
   \end{tabular}
@@
 
-Table~\ref{tab:exec-time} presents print statement \emph{P66} which prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
+Table~\ref{tab:exec-time} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
 \emph{P66} is also used in other code cells in the notebook that train other ML models.
@@
 
+\highlight{\textbf{Finding 1.4} The study identifies one output (P66) used to check the total execution time of training and cross-validating an ML model.
+The print statement is used for evaluating the efficiency of model training processes, especially when multiple models are trained iteratively.}
+
 \subsubsection{Missing value check ($N = 3$)}
@@
     P74&
-\lstinline[]$print(train_df.isnull().sum())$&
-\\
+    \lstinline[]$print(train_df.isnull().sum())$\\
 
@@
     O36&
-\lstinline[]$test.isna().sum().unique()$&
-\\
+    \lstinline[]$test.isna().sum().unique()$\\
     \bottomrule
@@
 
+\highlight{\textbf{Finding 1.5} The study identifies three outputs (P74, O12 and O36) used to check for missing data.
+Missing values can lead to biased or incorrect conclusions, and many ML algorithms require complete numerical datasets for accurate calculations.
+Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
+
 \subsubsection{Model performance check ($N = 33$)}
@@
 
+\highlight{\textbf{Finding 1.6} The study identifies 33 outputs used to check the performance of trained ML models.
+These include outputs that print key performance metrics (P3, P6, P18, P50), a normalised heatmap used to visualise the confusion matrix of a model's predictions (O3) and a custom function written for manual evaluation of model predictions on random input samples (O52).
+These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
+
 \subsubsection{Neural network architecture ($N = 1$)}
@@
 
+% TODO: need more description and context around this print statement (where was it written, and why?)
 Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.
@@
 This practice is fundamental for maintaining accuracy, facilitating collaboration, and enhancing the reproducibility of the model.
+
+\highlight{\textbf{Finding 1.7} The study identifies one output (P92) used to print the architecture of a neural network.
+Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
 
@@
 
+\highlight{\textbf{Finding 1.8} The study identifies three outputs used to check the shape of data.
+These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
+
 \subsubsection{Type check ($N = 2$)}
@@
 
+\highlight{\textbf{Finding 1.9} The study identifies two outputs used to verify the data types of features in the dataset.
+These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
+Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
+
 \subsubsection{Spot checks ($N = 5$)}
@@
 
-% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.
+\highlight{\textbf{Finding 1.10} This study finds five outputs used to perform spot checks at various stages of the MLDL.
+Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
 
@@
 
-\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}
+\highlight{\textbf{Finding 1.11} The study identifies four outputs used to monitor model training progress.
+These outputs provide real-time feedback on training loss and accuracy (O8), validate model adherence to expected data patterns (O31), and identify optimal parameters through tuning methods (O42).
+These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
+
+\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
 
@@
 
+\highlight{\textbf{Finding 2.1} This study identifies three assertions validating batch size configurations.
+These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
+These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
+
 \subsubsection{Data leakage check ($N = 1$)}
@@
 
+\highlight{\textbf{Finding 2.2} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
+Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
+This practice maintains the integrity and validity of the model evaluation metrics.}
+
 \subsubsection{Existence check ($N = 8$)}
@@
 
+\highlight{\textbf{Finding 2.3} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
+These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
+
 \subsubsection{Mathematical property checks ($N = 4$)}
@@
 These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
+
+\highlight{\textbf{Finding 2.4} Four assertions validate key mathematical properties of data and model outputs.
+These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
+These checks confirm the robustness and reliability of model computations.}
 
@@
 
+\highlight{\textbf{Finding 2.5} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
+These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
+
 \subsubsection{Network architecture check ($N = 3$)}
@@
 Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
+
+\highlight{\textbf{Finding 2.6} Three assertions verify the consistency of neural network architecture.
+These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
+These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
 
@@
 
+\highlight{\textbf{Finding 2.7} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
+These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
+
 \subsubsection{Data shape check ($N = 26$)}
@@
 
+\highlight{\textbf{Finding 2.8} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
+These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
+
 \subsubsection{Type check ($N = 5$)}
@@
 
+\highlight{\textbf{Finding 2.9} Five assertions validate the data types of input features, models, and intermediate objects.
+These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
+
 \subsubsection{Data validation check ($N = 14$)}
@@
 
-\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}
-
-\subsubsection{Model Performance}
+\highlight{\textbf{Finding 2.10} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
+These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
+
+\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
+
+\subsubsection{Model Performance ($N = 15$)}
 
@@
 
-\subsubsection{Distribution Check}
+\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
+The visualizations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
+Assertions associated with these visualizations ensure models meet a specific performance criteria.}
+
+\subsubsection{Distribution Check ($N = 1$)}
 
@@
 
-\subsubsection{Data Validation}
+\highlight{\textbf{Finding 3.1} One visualization is used to display the distribution of bootstrapped sample means.
+The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
+
+\subsubsection{Data Validation ($N = 9$)}
 
@@
 
-\section{Implications}\label{sec:discuss}
-% NOTE: from ICST paper, need to integrate...
-
-% NOTE: breck2017ml proposes feature and data testing; we can say that the assertions in the notebook bring the system closer to migrating them into dedicated data validation systems
-
-% NOTE: refer back to sec:assert-model-perf; the use of np.isclose and robustness to stochastic nature
-
-% NOTE: perhaps also argue the other side, the demographic using notebooks do not have SE training; they are perhaps directly sharing the notebook and the data directly...
-P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
+\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original.
+These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}
+
+\section{Discussion}\label{sec:discuss}
+
+Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle.
+Making such decisions however, requires manual validation of the outputs.
+For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training.
+Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks.
+Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself.
+Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.
+
+Assertions can play a crucial role in automatically validating critical assumptions in ML workflows.
+As seen in the results from RQ2, assertions can verify that data distributions remain consistent with expectations or that model performance metrics meet predefined thresholds.
+Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged, reducing the reliance on manual verification and enhancing the robustness and maintainability of ML pipelines.
+Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development.
+This is particularly important in collaborative environments where multiple practitioners might work on the same project.
+Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
+Moreover, they enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells.
+
+% P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
 This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
@@
 
-% NOTE: think about the actors: educators, researchers, developers...
-The results from RQ1 provide empirical evidence to support the use of
-visualisations to test various properties of ML systems.
-Visualisations are an efficient and effective medium to present
-complex information, that is easier to interpret by humans with the
-aid of visual cues. Visualisations used to test ML system properties
-however, demand manual verification effort. Information gained from
-visualisations and subsequent decisions made based on this
-information, must be manually verified in every iteration of the ML
-development cycle.
-
-% NOTE link to accumulation of technical debt due to highly
-% experimental nature, and constant change in the system
-
-Assertions or analytical tests extracted from ML visualisations can
-reduce this manual verification effort. Assertions encapsulate the
-knowledge gained from the visualisation with regards to the property
-of the ML system which is currently under test. Formal assertions
-document the observations made by the AI practitioner regarding the
-model or the data, at a specific point in time. These observations
-should not be violated by the future model development cycles unless
-the requirements of the ML system are changed. Assertions scale across
-organisational or personnel changes as future AI practitioners can
-refer to the assertions to understand how the visualisations were
-previously interpreted.
-
-The results for RQ1 indicate that the combination of ML visualisations
-with analytical assertions is an emerging testing practice within the
-machine learning community. We found only $54,070$ notebooks that
-contain the keyword ``assert'', while there are more than a million
-Jupyter notebooks available on GitHub. This suggests that writing
-assertions in computational notebooks is still relatively uncommon. We
-believe that this is due to a lack of knowledge among practitioners
-regarding testing approaches and the absence of adequate tooling to
-support the assertion of model properties. We argue that guidelines on
-ML coding practices require improvement. And advocate the need to
-steer education and research of Software Engineering practices towards
-the specific challenges faced when contributing to ML software
-systems.
-
-In the 1764 notebooks analysed in this study, several visualisations
-lack assertions. However, implicit expectations still exist within
-each visualisation which cannot be readily validated by other project
-collaborators. ML testing demands proficiency in Machine Learning,
-Software Testing, and the underlying Mathematical concepts. This
-further highlights the need for refining the education for these
-professionals and crafting tools to compensate for gaps in their
-knowledge.
-
-Results from RQ2 reveal two challenges in the current landscape of ML
-testing. The assertions in the VA pairs identified in this study lack
-dedicated testing libraries. This indicates that the current culture
-of testing ML code is still in its infancy and ML practitioners are
-not yet exploring software testing libraries. Second, the existing
-testing libraries are not tuned to the specifics of ML code and
-practitioners have to fall back to basic built-in approaches.
-
-The widespread use of ``magic'' thresholds---often rooted in domain
-expertise or experience obtained over time---poses risks to
-maintainability of ML projects. We advocate that such thresholds are
-documented for future reference. And find a clear need for automated
-tools that highlight undocumented choices or suggest standardized
-documentation in Markdown or code annotations.
-
-In the results from RQ3, we see several examples of assertions that
-fail to capture all the information gained from its visual
-counter-part. This can lead to a mismatch of expectations in future
-versions of the ML software. For instance, another project
-collaborator might run the code with a new version of the dataset or a
-different learning technique and assume that if the assertions run
-successfully, the expected patterns from the visualisation are also
-satisfied. The examples presented in Section~\ref{sec:result-rq3} show
-that deriving assertions from visualisations is not trivial. Basic
-visualisation patterns require complex mathematical techniques which
-may be beyond the expertise of an ML developer.
-
-For example, a linear correlation between the true and predicted
-labels is evident in Figure~\ref{fig:r2}. Yet, its formal assertion in
-Listing~\ref{lst:r2} necessitates defining an appropriate threshold
-for the coefficient of determination metric. Our results show that
-current software testing practices fail to capture the idiosyncrasies
-of ML code. There is a pressing need for automated tools, that bridge
-the disparity between the qualitative evaluation of ML systems using
-visualisations and the extraction of analytical assertions from these
-visuals.
+% NOTE: here we need to bring in prior literature on ML testing; and how majority of the research is working on testing using a test dataset; however we also have data and models
+% NOTE: 3 pillars of of ML (sato2019continuous)
+
+From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions.
+This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community.
+The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.
+
+We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties.
+Current guidelines on ML coding practices require enhancement to emphasize the importance of testing.
+There is a pressing need to steer education and research in Software Engineering towards addressing the specific challenges faced when developing ML-enabled software systems.
+Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
+
+The assertions presented in RQ2 and RQ3 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
+First, the assertions identified in this study lack the support of dedicated testing libraries.
+This suggests that the culture of testing ML code is still in its infancy, and ML practitioners are not yet fully exploring or leveraging software testing libraries.
+Secondly, existing testing libraries are not specifically tuned to the needs of ML code, leading practitioners to rely on basic built-in approaches.
+This highlights the necessity for developing specialized testing frameworks that cater to the unique requirements of ML projects.
+
+Visualizations are an effective tool for interpreting complex ML models and data transformations.
+However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}.
+We argue that the need for translating visual insights into formal assertions is more pressing.
+The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners.
+Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
+
+In the VA pairs from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+This can lead to a mismatch of expectations in future versions of the ML software.
+For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
+The VA pairs presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
+Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
+For example, a linear correlation between the true and predicted labels is evident in VA107 presented in Table~\ref{tab:va-model-perf}.
+Yet, its formal assertion necessitates defining an appropriate threshold for the coefficient of determination metric.
+Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
+There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
+
+% TODO: integrate our NLBSE paper
+Our results show a clear need for developing ML-specific testing tools.
+Such tools should seamlessly integrate with computational notebooks and provide automated suggestions for assertions can greatly enhance the reliability and robustness of ML models.
+These tools can also aid in documenting and standardizing the use of domain-specific "magic" thresholds, ensuring that they are consistently applied and maintained.
+
+% NOTE: we find overlap between the manual tests performed using the output of code cells and the assertions.
+
+% NOTE: bring in prior work on reproducibility, code quality here
+% NOTE: notebooks have been adopted by a much larger scientific community; not eveybody has a background in software engineering; we need more tools to aid notebook users...
+
+\section{Implications}
+
+\paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks.
+Assertions play a crucial role in validating critical assumptions automatically, reducing the reliance on manual checks, and enhancing the robustness and maintainability of ML pipelines.
+By embedding assertions, developers can document the decisions made during model development, ensuring that these decisions are consistently applied and preserved throughout the ML lifecycle.
+This practice not only facilitates smoother transitions and handovers in collaborative environments but also improves the reproducibility of ML experiments by ensuring consistent validation checks.
+
+\paragraph{\textbf{For Researchers}} The use of assertions in Jupyter notebooks, though currently uncommon, is an emerging testing practice within the ML community.
+This study highlights the potential for assertions to enhance the reliability and robustness of ML models by automating the validation of critical assumptions.
+Researchers are encouraged to explore the development and evaluation of ML-specific testing methodologies that leverage assertions.
+This could involve creating new testing frameworks and tools tailored to the unique challenges of ML systems, thus contributing to the advancement of software engineering practices in the ML domain.
+
+\paragraph{\textbf{For Notebook Tool Developers}} There is a clear need for automated tools that suggest assertions based on the outputs of code cells in Jupyter notebooks.
+Tool developers can play a pivotal role by creating ML-specific testing tools that seamlessly integrate with computational notebooks.
+These tools should provide automated suggestions for assertions, helping practitioners to translate visual insights into formal, executable checks.
+By bridging the gap between qualitative evaluation and analytical assertions, such tools can significantly enhance the reliability and robustness of ML models, ensuring that they meet predefined performance standards consistently.
+
+\paragraph{\textbf{For Educators}} Educators should emphasize the importance of testing ML code written in Jupyter notebooks as part of the curriculum.
+By incorporating testing practices and the use of assertions into ML education, students can develop a deeper understanding of the importance of automated validation in ensuring model reliability and robustness.
+Teaching these best practices early on can help cultivate a culture of rigorous testing among future ML practitioners, preparing them to address the specific challenges faced in ML-enabled software systems.
+
+\paragraph{\textbf{For General Notebook Users}} Awareness of software engineering best practices, including the use of assertions, is crucial for all users of Jupyter notebooks.
+Assertions serve as an executable form of documentation, capturing the critical assumptions and decisions made during the development process.
+By adopting assertions, users can ensure that their notebooks are more robust, maintainable, and reproducible.
+This practice helps in maintaining consistency and reliability, especially in collaborative projects where multiple users might work on the same notebook, ensuring that the assumptions and validations are transparent and consistently applied.
+
+\section{Future Work}
+
+\section{Conclusion}
 

### 2024-06-07 report/emse24.tex
feat(report): emse24 updates

@@
 Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring.
-Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
-
+Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.
+
+Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
 Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks.
@@
 Users can test hypotheses, debug issues, and make incremental changes with immediate feedback.
-The ability to visualize data and results inline and document the process in a narrative form further enhances their utility.
-
-% It aligns perfectly with the feedback-driven nature of machine learning development, as it supports rapid prototyping and iterative refinement of models.
-
-% TODO: back the TDD angle not only from scientific literature but also general methodologies used and adopted by the software engineering community (Xtreme programming, Kent Beck)
-
-The predominant form of obtaining feedback in Jupyter notebooks is through the output of code cells.
-These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations.
-Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code.
-This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.
-
-% TODO cite scully and breck and the data validation papers here...
-% TODO: why are we doing this? how is the prior work lacking? why is this important?
-
-In Jupyter notebooks, feedback is crucial for making important decisions throughout the machine learning development lifecycle.
+The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.
+
+Feedback is crucial for making important decisions throughout the MLDL.
 Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters.
 However, these decisions are often based on the conditions present when the model was initially created.
-As data, personnel, and system environments inevitably change, undocumented decisions can lead to significant technical debt.
-This not only prolongs debugging and troubleshooting times but also incurs additional costs and resource expenditures.
+Visual outputs in particular, are open to interpretation~\cite{heer2010tour}.
+The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data.
+For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess.
+Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind.
+As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
+This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
 
@@
 Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development.
-By embedding assertions, developers ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.
-
-% TODO introduce our results and key findings...
-
-% and then to tie the two things together: are assertions derived from the output of a cell?
-we still need to find a way to justify why we only look at assertions derived from visualisations specifically
-
-% By analyzing the outputs and assertions in Jupyter notebooks, we aim to provide insights into the feedback mechanisms used in machine learning development, ultimately proposing strategies to improve the efficiency and robustness of this iterative process.
+Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.
+
+To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks.
+Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
+Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells.
+In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.
+
+% TODO: contributions?
+% TODO: key observations?
 
@@
 
-% building ml systems is highly iterative and experimental.
-information gained from ml visualisations is used to make design and implementation decisions for the following steps of the ml pipeline.
-for instance, we may visualise the distribution of our training data and find that it is normally distributed.
-based on this information, we may opt for a linear regression model which assumes normality in the underlying data.
-however, such expectations regarding the data may be violated once the ml system is deployed in production, where the data constantly changes as a reflect of the real world~\cite{amershi2019software,sambasivan2021everyone,breck2019data,baylor2017tfx}.
-
-% implicit expectations obtained from visualisations must therefore be translated to analytical tests that fail once our expectations regarding the ml system are no longer satisfied.
-
-% in contrast to prior work, we approach ml testing from a new perspective.
-we conduct an empirical analysis of computational notebooks obtained from github, to understand the process of testing ml systems in practice.
-in particular, we focus on the combination of a qualitative form of testing (using visualisations) and a quantitative form of testing (using analytical assertions).
-the research questions along with the contributions of this paper are as follows.
-
 % our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique.
@@
 
-\section{Preliminaries}\label{sec:prelim}
+\section{Background}\label{sec:background}
+
+\subsection{Jupyter Notebooks}
+
+% NOTE use background section from pimentel2019large-scale
+
+% describe the structure of jupyter notebooks (anatomy: cells, outputs); no need to go into too much technical detail regarding the internal JSON structure, just mention that the data is represented in a standardised structure internally
+
+\subsection{Python Abstract Syntax Tree}
+
+% explain the high-level way in which ast module in python works
+% explain the anatomy of assert statements and the more complicated way in which asts for print statements are represented in an AST
+
+\section{Study Design}
+
+\subsection{Motivating Example}
+
+Sara is a data scientist who is developing a machine learning model to predict customer churn.
+Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells.
+Her workflow is dynamic and exploratory, typical of modern machine learning practices.
+
+Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution.
+As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features.
+This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.
+
+With data preprocessing complete, Sara proceeds to build her initial model.
+After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy and loss after the first epoch.
+The results are disappointing, prompting her to adjust the learning rate and batch size directly in subsequent cells.
+Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings.
+
+Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions.
+At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.
+
+Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively.
+Each decision in Sara's workflow—from preprocessing adjustments to hyperparameter tuning and model validation—is guided by the iterative feedback she receives from the outputs of her notebook's code cells.
+
+% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster!
+(notebook keeps running!)
+Once all experimentation and model tuning are complete, and Sara is ready to deploy the model into a production environment, she adds assertions as a final step.
+In the dynamic, interactive environment of Jupyter notebooks, where code cells can be executed in any order, there is a risk of arriving at inconsistent states that may affect the model's performance unpredictably.
+Assertions act as a safeguard against these inconsistencies by verifying that the data fed into the model adheres to expected formats and distributions and that the model's performance metrics meet predetermined standards.
+For instance, assertions are used to ensure that no data columns have null values, that the range of input data features remains within expected bounds, and that model accuracy thresholds are consistently met regardless of the execution order of the notebook cells.
+This practice not only reinforces the stability and reliability of the deployed model but also provides clear, executable documentation of the critical decisions made throughout the model's development.
+When new team members are onboarded, these assertions allow them to quickly grasp the underlying assumptions and operational parameters of the model, ensuring ongoing robustness and performance integrity.
+This is crucial in maintaining the quality and reliability of the model as it transitions from a development setting to a production environment.
+
+\subsection{Exploratory Analysis}
+
+We undertook an exploratory analysis of the collected data to formulate our research questions.
+
+\subsection{Research Questions}
+
+\begin{description}
+  \item[RQ1.] \textbf{How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
+  \item[RQ2.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
+  \item[RQ3.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
+\end{description}
+
+\subsection{Data Collection}\label{sec:data-collect}
+
+% TODO we need to explain why we only focus on the last statement of code cells (when visualisation code written anywhere produces an output); this is much harder to parse: we need to construct an exhaustive list of python visualisation libraries; people can use any naming convension to...
+
+% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells.
+These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations.
+Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code.
+This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.
+
+\begin{figure}
+\centering
+\includegraphics[width=\linewidth]{data-collection.pdf}
+\label{fig:data-collection}
+\caption{Overview of data collection methodology used in this study}
+\end{figure}
+
+Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.
+
+We mined public repositories from Github to collect $49923$ Jupyter notebooks.
+We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them.
+The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
+
+For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github.
+To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.
+
+Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.
+
+We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}.
+Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
+
+For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells.
+We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.
+
+Similarly, for implicit feedback, we first isolate code cells that produce an output.
+The output of a cell may either be from a ``print'' statement or the last statement of the cell.
+We collect both of them using the AST of the cell.
+
+Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.
+
+\subsection{Case Studies}
+
+% TODO: might need a visual aid here
+
+We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs).
+To surface interesting candidates for the case studies, we apply NLP techniques as described below.
+
+% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
+Thats why we need to apply these techniques to surface unique and interesting candidates!
+
+The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed.
+Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
+The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
+
+We apply stratified random sampling to identify the candidates for the case study analysis.
+The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
+The candidates are then divided into quartiles based on the aggregate value.
+A candidate is randomly drawn from each bin and analysed as an in-depth case study.
+The analysis is stopped when the time resource is exhausted.
+
+During each case study, we analyse the code of the candidate to understand its purpose.
+Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.
+
+\section{Results}
+
+% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output.
+Somethimes actual output is not possible (eg.
+P68) because the statement may be inside an if statement.
+
+\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
+
+\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
+% (O2, O4, O9, O14, O20, O25, O48)
+
+\begin{table}
+  \centering
+  \caption{Examples of cell outputs used to verify distribution of data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}&
+    \emph{\textbf{Visual Output}}\\
+    \midrule
+
+    O2&
+    \lstinline[]$_ = sns.catplot(x='category_id', y='likes', data=train, height=5, aspect=1.5)$&
+    \includegraphics[width=\linewidth]{distribution-check-catplot.png}\\
+
+    O9&
+    \lstinline[]$sns.kdeplot(data=data.loc[ data['Survived'] == 0].Age, label='Died', shade=True)$&
+    \includegraphics[width=\linewidth]{distribution-check-kdeplot.png}\\
+
+    O14&
+    \lstinline[]$pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])$&
+    \\
+
+    O25&
+    \lstinline[]$sns.countplot(house_pred['OverallQual'])$&
+    \includegraphics[width=\linewidth]{o25.png}\\
+
+    O48&
+    \lstinline[]$x_train.describe()$&
+    \\
+
+    \bottomrule
+  \end{tabular}
+  \label{tab:distribution-check}
+\end{table}
+
+Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data.
+In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature.
+Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
+Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}.
+Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data.
+The statement is written immediately after normalising the features in the training data.
+Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output.
+Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.
+
+Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
+For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
+During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
+Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.
+
+\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
+These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}
+
+\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
+% (O6, O10)
+
+\begin{table}
+  \centering
+  \caption{Cell outputs used to verify linear relationship between features in the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}&
+    \emph{\textbf{Output}}\\
+    \midrule
+
+    O6&
+    \lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$&
+    \includegraphics[width=\linewidth]{linear-relation-check-lineplot.png}\\
+
+    O10&
+    \lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$&
+    \includegraphics[width=\linewidth]{linear-relation-check-regplot.png}\\
+    \bottomrule
+  \end{tabular}
+\label{tab:linear-relation-check}
+\end{table}
+
+Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data.
+In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
+Output \emph{O10} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
+
+Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
+However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
+Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.
+
+\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
+Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
+These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
+
+\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
+% (P71, P86, O64, O66, P68, P82, P107)
+
+\begin{table}
+  \centering
+  \caption{Examples of cell outputs used to verify if different types of resources are available on the system where the notebook is being executed.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    P68&
+    \lstinline[]$print('GPU is available')$\\
+
+    P71&
+    \lstinline[]$print('Hub version: ', hub.__version__)$\\
+
+    P82&
+    \lstinline[]$print('Running on TPU ', tpu.master())$\\
+
+    P86&
+    \lstinline[]$print('Cuda is available')$\\
+
+    P107&
+    \lstinline[]$print('Model loaded')$\\
+
+    O64&
+    \lstinline[]$full_table.head(-5)$\\
+
+    O66&
+    \lstinline[]$prostate_cancer_df.shape$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:resource-check}
+\end{table}
+
+% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
+Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed.
+Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU.
+This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory.
+For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory.
+Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data.
+These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.
+
+Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
+This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
+
+\highlight{\textbf{Finding 1.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
+These checks include confirming the availability of compute resources like GPUs and TPUs (P68, P82 and P86), ensuring datasets or pre-trained models are successfully loaded (P107, O64 and O66), and verifying the presence of specific library versions (P71).
+These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
+
+\subsubsection{Execution time check ($N = 1$)}
+% (P66)
+
+\begin{table}
+  \centering
+  \caption{Print statement used to check the total execution time of training and cross-validating an ML model.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    P66&
+    \lstinline[]$print('Total Run Time:')$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:exec-time}
+\end{table}
+
+Table~\ref{tab:exec-time} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
+\emph{P66} is also used in other code cells in the notebook that train other ML models.
+
+While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
+Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{CITEME}.
+
+\highlight{\textbf{Finding 1.4} The study identifies one output (P66) used to check the total execution time of training and cross-validating an ML model.
+The print statement is used for evaluating the efficiency of model training processes, especially when multiple models are trained iteratively.}
+
+\subsubsection{Missing value check ($N = 3$)}
+% (O12, O36, P74)
+
+\begin{table}
+  \centering
+  \caption{Examples of cell outputs used to check for missing data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}&
+    \emph{\textbf{Visual Output}}\\
+    \midrule
+
+    P74&
+    \lstinline[]$print(train_df.isnull().sum())$\\
+
+    O12&
+    \lstinline[]$sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')$&
+    \includegraphics[width=\linewidth]{missing-value-check.png}\\
+
+    O36&
+    \lstinline[]$test.isna().sum().unique()$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:missing-value}
+\end{table}
+
+Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
+While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
+
+Checking for missing values is a critical step in the preprocessing phase of a machine learning project because it significantly influences the quality and performance of the model.
+Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
+Many machine learning algorithms, including those that involve matrix multiplication operations such as linear regression and neural networks, demand complete numerical datasets to perform calculations.
+Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
+
+\highlight{\textbf{Finding 1.5} The study identifies three outputs (P74, O12 and O36) used to check for missing data.
+Missing values can lead to biased or incorrect conclusions, and many ML algorithms require complete numerical datasets for accurate calculations.
+Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
+
+\subsubsection{Model performance check ($N = 33$)}
+% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
+
+\begin{table}
+  \centering
+  \caption{Various cell outputs used to check the performance of an ML model after training.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}&
+    \emph{\textbf{Visual Output}}\\
+    \midrule
+
+    P3&
+    \lstinline[]$print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')$&
+    \\
+
+    P6&
+    \lstinline[]$print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))$&
+    \\
+
+    P18&
+    \lstinline[]$print('The Accuracy is:', accuracy_score(y_test, y_pred))$&
+    \\
+
+    P50&
+    \lstinline[]$print('Classification Report: SVM (validation data)')$&
+    \\
+
+    P54&
+    \lstinline[]$print('Intercept value:', lm.intercept_)$&
+    \\
+
+    O3&
+    \lstinline[]$skplt.metrics. plot_confusion_matrix(Y_val, Vote.predict(X_val), normalize=True, figsize=(10, 10))$&
+    \includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}\\
+
+    O52&
+    \lstinline[]$spot_check_recs(classifier, 910)$&
+    \\
+    \bottomrule
+  \end{tabular}
+  \label{tab:model-perf}
+\end{table}
+
+Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
+Besides the \emph{accuracy} of the model (\emph{P3} and \emph{P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
+
+% TODO: this feels hallucinated; revisit
+Print statement \emph{P54} prints the intercepts of a linear regression model.
+Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
+This insight is critical for accurate model interpretation and effective decision-making in practical applications.
+
+Output \emph{O3} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
+The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
+
+Output \emph{O52} shows a function implemented by the practitioners to manually evaluate an image classification model on a random sample of input images.
+The function \texttt{spot\_check\_recs} displays the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
+
+During the model development phase, the performance of an ML model is often printed to facilitate comparisons against other models or variations.
+Throughout this continuous experimentation phase, authors may adjust model parameters or modify the data by engineering new features.
+The model is then re-trained to check if these changes lead to performance improvements.
+
+\highlight{\textbf{Finding 1.6} The study identifies 33 outputs used to check the performance of trained ML models.
+These include outputs that print key performance metrics (P3, P6, P18, P50), a normalised heatmap used to visualise the confusion matrix of a model's predictions (O3) and a custom function written for manual evaluation of model predictions on random input samples (O52).
+These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
+
+\subsubsection{Neural network architecture ($N = 1$)}
+% (P92)
+
+\begin{table}
+  \centering
+  \caption{Print the architecture of a neural network.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+    P92&
+    \lstinline[]$print(MyNetwork)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:network-architecture}
+\end{table}
+
+% TODO: need more description and context around this print statement (where was it written, and why?)
+Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.
+
+Printing the structure of a neural network serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
+It helps in verifying configurations like layer dimensions and operations before initiating computationally intensive training, ensuring the model is set up efficiently for the intended tasks.
+This practice is fundamental for maintaining accuracy, facilitating collaboration, and enhancing the reproducibility of the model.
+
+\highlight{\textbf{Finding 1.7} The study identifies one output (P92) used to print the architecture of a neural network.
+Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
+
+\subsubsection{Shape check ($N = 3$)}
+% (P4, P32, P117)
+
+\begin{table}
+  \centering
+  \caption{Example of outputs used to check the shape of the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}} & \emph{\textbf{Code}}\\
+    \midrule
+    P4 &
+    \lstinline[]$print('no.of examples in test data : ', len(test_data))$\\
+
+    P32 &
+    \lstinline[]$print('Training set shape : ', x_train.shape)$\\
+
+    P117&
+    \lstinline[]$print('Y_train.shape: ', Y_train.shape)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:shape-check}
+\end{table}
+
+Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data.
+\emph{P4} prints the number of rows or examples in the test set.
+\emph{P32} and \emph{117} prints the number of rows and columns (or features) in the training and test data respectively.
+
+Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
+Primarily, the number of features in the training set must match those in the testing set.
+This alignment is crucial because statistical machine learning models are trained on specific data dimensions and expect the same dimensional structure during inference to perform accurately.
+Similarly, in neural network architectures, the configuration of input layers depends directly on the shape of the training data, dictating the number of input neurons needed.
+Furthermore, the correspondence between the number of test examples and their respective labels is essential for accurately computing performance metrics, such as accuracy.
+
+Such checks are indicative of the meticulous attention to detail necessary in the model training and validation process, underscoring the importance of shape validation in achieving reliable machine learning outcomes.
+
+\highlight{\textbf{Finding 1.8} The study identifies three outputs used to check the shape of data.
+These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
+
+\subsubsection{Type check ($N = 2$)}
+% (O71, P43)
+
+% TODO: the narrative is not as strong here; it would help to also show the actual output as well?
+For instance, in P43 by itself doesn't make that much sense, the output however shows that all features in the dataset are floats which is a pre-requisite for ML training data
+\begin{table}
+  \centering
+  \caption{Outputs used to check the data type of features in the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}} &
+    \emph{\textbf{Code}}\\
+
+    P43&
+    \lstinline[]$print('data type:', images.dtype)$\\
+
+    O71&
+    \lstinline[]$type(Y)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:type-check}
+\end{table}
+
+% TODO: for O71, the output shows its a ndarray; without showing this to the reader we cannot make a strong narrative around these type checks
+Table~\ref{tab:type-check} illustrates outputs of cells found in our analysis, where checks are employed to verify the data types of the features in the dataset.
+Print statement \emph{P43} is written prior to training a neural network.
+The output validates that all features in the training data are numerical.
+Similarly, output \emph{O71} shows the data type of the target column.
+
+Type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
+Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis (EDA) phase of developing machine learning models.
+Since all machine learning algorithms perform mathematical operations, such as matrix multiplications, they require the input data to be entirely numerical to avoid computational errors.
+The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
+For example, while numerical features often require normalisation to bring them within a specific scale, categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
+
+\highlight{\textbf{Finding 1.9} The study identifies two outputs used to verify the data types of features in the dataset.
+These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
+Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
+
+\subsubsection{Spot checks ($N = 5$)}
+% (O56, O60, P64, P67, P114)
+
+\begin{table}
+  \centering
+  \caption{Example of outputs used to perform spot checks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    O60 &
+    \lstinline[]$X_pca.head()$\\
+
+    P64 &
+    \lstinline[]$print(np.max(cur[:, :, 1]))$\\
+
+    P114 &
+    \lstinline[]$print(onehot_encoded)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:spot-check}
+\end{table}
+
+Table~\ref{tab:spot-check} presents outputs which were used to perform spot checks at various stages of the ML development cycle.
+In \emph{O60}, the practitioner verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
+\emph{P64} demonstrates another case which is particularly relevant in image processing contexts.
+The maximum value in the second channel of a 3D numpy array is extracted and verified, ensuring that the data manipulation retains its expected properties.
+Lastly, \emph{P114} assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly.
+
+Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
+These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
+Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
+
+\highlight{\textbf{Finding 1.10} This study finds five outputs used to perform spot checks at various stages of the MLDL.
+Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
+
+\subsubsection{Model training ($N = 4$)}
+% (O8, O31, O42, P77)
+
+\begin{table}
+  \centering
+  \caption{Example of outputs used to monitor model training.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    O8 &
+    \lstinline[]$autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)$\\
+
+    O31 &
+    \lstinline[]$adaBoost.fit(X_train, y_train)$\\
+
+    O42&
+    \lstinline[]$m_r.best_params_$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:model-training}
+\end{table}
+
+During training it is common practice to monitor progress periodically, as illustrated by outputs \emph{O8} in Table~\ref{tab:model-training}, where the training loss and accuracy are printed to give real-time feedback on the model's learning efficacy.
+This feedback allows for adjustments in training parameters or early stopping to prevent overfitting.
+For instance, in output O8, the autoencoder is trained using a custom callback from \texttt{keras\_utils} to provide a progress update without cluttering the output stream, which is particularly useful in lengthy training sessions.
+In contrast, output \emph{O31} represents a crucial validation step where the \texttt{adaBoost} model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
+Output \emph{O42} checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
+
+Each of these outputs underscores different facets of the model training phase, from initialization and real-time monitoring to validation and finalization, illustrating the layered complexity of developing predictive models.
+
+\highlight{\textbf{Finding 1.11} The study identifies four outputs used to monitor model training progress.
+These outputs provide real-time feedback on training loss and accuracy (O8), validate model adherence to expected data patterns (O31), and identify optimal parameters through tuning methods (O42).
+These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
+
+\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
+
+% TODO: bring in the notion of pre and post condition checks (the wikipedia page on assertions is very interesting: https://en.wikipedia.org/wiki/Assertion_(software_development)); specifically the part on run-time assertions
+
+\subsubsection{Batch size check ($N = 3$)}
+% (A21, A28, A70)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to validate the batch size of input data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A21 &
+    \lstinline[]$assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'$\\
+
+    A28 &
+    \lstinline[]$assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'$\\
+
+    A70 &
+    \lstinline[]$assert n_img > batch_size$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:batch-size}
+\end{table}
+
+Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
+\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
+Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
+\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
+
+In the context of neural network training, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilization, particularly GPU cores.
+Batching allows for a more stable and smoother learning curve as the model's weights are updated not for each individual example, but rather based on the average gradient of a batch.
+This technique not only optimizes learning but also enhances the model's generalizability due to more robust gradient estimations.
+Given these benefits, it becomes crucial to ensure that the batch size is appropriately set to match the hardware's capacity, avoiding memory overflow issues that can halt training.
+These assertions are strategically placed to prevent common errors that could undermine training efficiency and model performance, thereby safeguarding the training process from potential disruptions due to misconfigured batch sizes.
+
+\highlight{\textbf{Finding 2.1} This study identifies three assertions validating batch size configurations.
+These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
+These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
+
+\subsubsection{Data leakage check ($N = 1$)}
+% (A33)
+
+\begin{table}
+  \centering
+  \caption{Assert used to prevent data leakage.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A33&
+    \lstinline[]$assert len(set( tr_df.PetID.unique()).intersection(valid_df.PetID.unique())) == 0$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:data-leakage}
+\end{table}
+
+Assertion \emph{A33} in Table~\ref{tab:data-leakage} ensures that there are no overlapping \texttt{PetID} values between the training set \texttt{tr\_df} and the validation set \texttt{valid\_df}, thereby preventing any possibility of data leakage.
+
+Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in machine learning to prevent models from overfitting.
+Overfitting occurs when a model is excessively complex, learning not only the underlying patterns but also the noise in the training data, which can degrade its performance on unseen data.
+To guarantee that the model can generalize effectively to new examples, it's crucial to verify that the training and validation sets are completely distinct with no shared examples.
+By strictly separating these datasets, the validation phase serves its purpose of providing an unbiased evaluation of the model's ability to generalize---to data similar to, but not identical to that which it was trained on---thus maintaining the integrity and validity of the evaluation metrics derived during model testing.
+
+\highlight{\textbf{Finding 2.2} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
+Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
+This practice maintains the integrity and validity of the model evaluation metrics.}
+
+\subsubsection{Existence check ($N = 8$)}
+% NOTE: this is a combination of missing-value-check and existence-check
+% (A23, A42, A43, A50, A51, A63, A79, A86)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to perform existence checks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A23 &
+    \lstinline[]$assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())$\\
+
+    A42 &
+    \lstinline[]$assert not lab_s.isnull().values.any()$\\
+
+    A43 &
+    \lstinline[]$assert len(data) != 0, 'cannot divide by zero'$\\
+
+    A50 &
+    \lstinline[]$assert not np.any(np.isnan(X))$\\
+
+    A51 &
+    \lstinline[]$assert data.target.notnull().all()$\\
+
+    A63 &
+    \lstinline[]$assert X.isnull().sum().sum() == 0$\\
+
+    A79 &
+    \lstinline[]$assert not processed_data_df.isna().any().any()$\\
+
+    A86 &
+    \lstinline[]$assert p0 in poi_info.index$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:existence-check}
+\end{table}
+
+Assertions \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
+Assertion \emph{A43} guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
+Assertions \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
+Assertions \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
+
+Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modeling.
+These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
+Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values.
+These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
+
+\highlight{\textbf{Finding 2.3} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
+These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
+
+\subsubsection{Mathematical property checks ($N = 4$)}
+% (A3, A25, A56, A64)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to validate mathematical properties of neural networks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A3 &
+    \lstinline[]$assert (xH - wH) % self.stride == 0$\\
+
+    A25 &
+    \lstinline[]$assert test_output.std() < 0.15, "Don't use batchnorm here"$\\
+
+    A56 &
+    \lstinline[]$assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))$\\
+
+    A64 &
+    \lstinline[]$assert np.allclose(T, T.T)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:maths-check}
+\end{table}
+
+Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
+\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior.
+\emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations.
+Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.
+
+When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
+This rigor is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
+These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
+
+\highlight{\textbf{Finding 2.4} Four assertions validate key mathematical properties of data and model outputs.
+These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
+These checks confirm the robustness and reliability of model computations.}
+
+\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
+% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to check performance of ML models.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A7 &
+    \lstinline[]$assert len(neighbours_1) == 20, "Neighbors don't match!"$\\
+
+    A15 &
+    \lstinline[]$assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))$\\
+
+    A19 &
+    \lstinline[]$assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'$\\
+
+    A38 &
+    \lstinline[]$assert 0.75 < auc(fpr, tpr) < 0.85$\\
+
+    A58 &
+    \lstinline[]$assert np.isclose(accuracy, 0.9666666666666667)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:model-perf-explicit}
+\end{table}
+
+This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
+
+% TODO: verify if this applies to recommender systems as well?
+For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
+Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
+Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity.
+Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.
+
+The use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
+This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
+By allowing a small tolerance in the comparison, np.isclose ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
+
+\highlight{\textbf{Finding 2.5} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
+These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
+
+\subsubsection{Network architecture check ($N = 3$)}
+% (A11, A62, A75)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to check the architecture of neural networks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A11 &
+    \lstinline[]$assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()$\\
+
+    A62 &
+    \lstinline[]$assert activations[i + 1].shape[1] == self.encoder_layer_sizes[-1] * 2$\\
+
+    A75 &
+    \lstinline[]$assert reg in ['none', 'l2']$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-network-architecture}
+\end{table}
+
+Assertion \emph{A11} verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
+The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
+By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
+
+Assertion \emph{A62} checks that the shape of the activations at index \lstinline{i + 1} has the correct number of channels, which should be twice the size of the last encoder layer.
+This ensures that the intermediate activations maintain the expected structure necessary for the proper functioning of the subsequent layers in the network.
+Ensuring that the shape of activations is consistent throughout the network is crucial for maintaining the integrity of the data flow and preventing errors during the forward pass.
+
+Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'.
+This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behavior or suboptimal model performance.
+
+When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
+This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
+Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
+
+\highlight{\textbf{Finding 2.6} Three assertions verify the consistency of neural network architecture.
+These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
+These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
+
+\subsubsection{Resource check ($N = 7$)}
+% (A10, A14, A18, A37, A60, A67 A74)
+
+\begin{table}
+  \centering
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A10 &
+    \lstinline[]$assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."$\\
+
+    A14 &
+    \lstinline[]$assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'$\\
+
+    A18&
+    \lstinline[]$assert pd.__version__.rpartition('.')[0] == '1.0', f"Unexpected pandas version: expected 1.0, got {pd.__version__.rpartition('.')[0]}"$\\
+
+    A37 &
+    \lstinline[]$assert svm.fit_status_ == 0, 'Forgot to train the SVM!'$\\
+
+    A60 &
+    \lstinline[]$assert f2.gca().has_data()$\\
+
+    A67&
+    \lstinline[]$assert pm.__version__ == '3.9.2'$\\
+
+    A74 &
+    \lstinline[]$assert os.path.exists(image_dir)$\\
+    \bottomrule
+  \end{tabular}
+  \caption{Assertions used to validate availability of resources.}
+  \label{tab:assert-resource-check}
+\end{table}
+
+Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
+These checks help prevent runtime errors and ensure the proper execution of code.
+
+One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
+These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing resources.
+
+Additionally, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).
+A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
+The assertion \emph{A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
+This check prevents inconsistent model states from executing cells out of order or multiple times.
+
+Assertion \emph{A60} is used to verify the presence of data in visualizations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
+These checks are crucial for ensuring that the visualizations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
+
+Assertions \emph{A18} and \emph{A67} explicitly verify the versions of critical libraries.
+For instance, \emph{A18} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
+
+Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
+
+% NOTE: left over loads, move to implications if required
+% In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
+Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
+
+% These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
+
+\highlight{\textbf{Finding 2.7} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
+These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
+
+\subsubsection{Data shape check ($N = 26$)}
+% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
+
+% TODO: revisit the assertions in this section; we may have misclassified assertions or need to create a new group
+
+\begin{table}
+  \centering
+  \caption{Assertions used to check the shape of the data.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A5 &
+    \lstinline[]$assert y_valid.shape == (1132,)$\\
+
+    A17 &
+    \lstinline[]$assert X.shape[1] == 13, 'Did you drop/lose some columns in X? Did you properly load and split the data?'$\\
+
+    A29 &
+    \lstinline[]$assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'$\\
+
+    A31 &
+    \lstinline[]$assert img.shape == (112, 92)$\\
+
+    A76 &
+    \lstinline[]$assert len(encoding['token_type_ids']) == max_seq_length$\\
+
+    A84 &
+    \lstinline[]$assert red.get_shape().as_list()[1:] == [224, 224, 1]$\\
+
+    A90 &
+    \lstinline[]$assert len(X_train) == 2000$\\
+
+    A93 &
+    \lstinline[]$assert temp_embed.shape[0] == stride$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-shape-check}
+\end{table}
+
+Assertions \emph{A17} and \emph{A90} verify that the input features have the correct number of columns and examples, respectively.
+Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
+
+Assertions \emph{A5} and \emph{A29} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
+
+In the context of computer vision tasks, assertions \emph{A31} and \emph{A84} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
+
+Furthermore, assertions \emph{A76} and \emph{A93} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
+
+Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
+We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features, labels, predictions, images, sequences, and embeddings.
+These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.
+
+Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
+By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
+
+\highlight{\textbf{Finding 2.8} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
+These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
+
+\subsubsection{Type check ($N = 5$)}
+% (A2, A35, A40, A81, A88)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to check the data type of features in the dataset.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A2 &
+    \lstinline[]$assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'$\\
+
+    A35 &
+    \lstinline[]$assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"$\\
+
+    A40 &
+    \lstinline[]$assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)$\\
+
+    A81 &
+    \lstinline[]$assert is_all_ints(filled_df[r]) is True$\\
+
+    A88 &
+    \lstinline[]$assert isinstance(betas, np.ndarray)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:assert-type-check}
+\end{table}
+
+Assertion \emph{A2} in Table~\ref{tab:assert-type-check} demonstrates a very common type check performed on the input features.
+This assertion ensures that the input features are represented as PyTorch float tensors, which is a requirement for many deep learning models and operations.
+
+Another type check is applied to verify the type of the machine learning models themselves.
+For instance, the \emph{A40} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
+This type of assertion is essential when working with different types of models, as it helps prevent errors arising from using incompatible methods or attributes.
+
+In addition to input features and models, type checks are also performed on data structures and intermediate objects.
+The assertion \emph{A35} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
+
+Type checking is not limited to objects and data structures--it can also be applied to individual columns or elements within a dataset.
+The assertion \emph{A81} checks if all the elements in a specific column (\lstinline[]$filled_df[r]$) of a DataFrame are integers.
+
+Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}.
+This assertion ensures that the \lstinline{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
+
+Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behavior.
+By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.
+
+\highlight{\textbf{Finding 2.9} Five assertions validate the data types of input features, models, and intermediate objects.
+These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
+
+\subsubsection{Data validation check ($N = 14$)}
+% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
+
+\begin{table}
+  \centering
+  \caption{Assertions used to perform data validation checks.}
+  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Code}}\\
+    \midrule
+
+    A41 &
+    \lstinline[]$assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))$\\
+
+    A44 &
+    \lstinline[]$assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1$\\
+
+    A46 &
+    \lstinline[]$assert np.isclose(stdev_norm, 1.0, atol=1e-16)$\\
+
+    A52 &
+    \lstinline[]$assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()$\\
+
+    A65 &
+    \lstinline[]$assert np.all(y <= nb_classes)$\\
+
+    A73 &
+    \lstinline[]$assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:data-validation}
+\end{table}
+
+% TODO: revisit, we have new assertions here; update the count
+Table~\ref{tab:data-validation} presents a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
+These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
+
+One type of validation check involves verifying the presence of specific values or ranges within arrays or columns.
+For example, assertion \emph{A41} checks if the unique values in the \lstinline{smoke} column of the DataFrame \lstinline{X} are limited to 0 and 1.
+Similarly, \emph{A65} ensures that all values in the array \lstinline{y} are less than or equal to a certain number of classes.
+
+Another common validation check is verifying the equality or closeness of values or arrays.
+Assertion \emph{A44} checks if the difference between the minimum and maximum values in a specific range of the \lstinline{entity_embedding} array is less than or equal to 1.
+Assertion \emph{A46} verifies that the \lstinline{stdev_norm} value is close to 1.0 within a specified absolute tolerance.
+
+Some assertions focus on validating the uniqueness or cardinality of values in arrays or DataFrames.
+For instance, \emph{A52} checks if the number of unique user IDs in the \lstinline{grouped_users} DataFrame is equal to the number of unique user IDs in the \lstinline{user_engagement} DataFrame.
+
+Other assertions validate the presence of specific values or conditions within DataFrames or other data structures.
+The assertion \emph{A73} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \lstinline{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
+
+By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
+
+\highlight{\textbf{Finding 2.10} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
+These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
+
+\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
+
+\subsubsection{Model Performance ($N = 15$)}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used to validate the performance of a model.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used to validate the performance of a model (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA2 &
+  \includegraphics[width=\linewidth]{va2.png} &
+  \lstinline[]$assert final_score > 300, 'not good enough for DQN'$\\
+
+  VA3 &
+  \includegraphics[width=\linewidth]{va3.png} &
+  \lstinline[]$assert accuracy_score(y_test, pred) > 0.95$\\
+
+  VA9 &
+  \includegraphics[width=\linewidth]{va9.png} &
+  \lstinline[]$assert np.mean(mean_rw_history[-10:]) > 10.$\\
+
+  VA11 &
+  \includegraphics[width=\linewidth]{va11.png} &
+  \lstinline[]$assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."$\\
+
+  VA15 &
+  \includegraphics[width=\linewidth]{va15.png} &
+  \lstinline[]$assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."$\\
+
+  VA28 &
+  \includegraphics[width=\linewidth]{va28.png} &
+  \lstinline[]$assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."$\\
+
+  VA68 &
+  \includegraphics[width=\linewidth]{va68.png} &
+  \lstinline[]$np.testing.assert_almost_equal(ans4, 0.3042764)$\\
+
+  VA107 &
+  \includegraphics[width=\linewidth]{va107.png} &
+  \lstinline[]$assert r2_gru > 0.6$\\
+
+\end{longtable}\label{tab:va-model-perf}
+
+The visualizations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
+VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
+These metrics collectively indicate the model's learning progress and stability during training.
+The visualization in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
+The general upward trend indicates learning and improvement in gameplay over time.
+VA15 presents the training and development loss over time for a character-level language model, showing a gradual decrease in loss and suggesting effective learning.
+
+The visualization in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
+This visualization aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
+The visualizations in VA28 display the training loss and development recall, respectively.
+The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
+The visualizations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
+The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
+In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
+
+The corresponding assertions for these visualizations ensure that the models meet specific performance standards.
+For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
+In VA9, the assertion checks whether the mean of the last ten rewards exceeds 10, ensuring stable and satisfactory learning progress.
+For VA15, the assertion checks if the mean of the first ten training observations is greater than the mean of the last ten, ensuring loss convergence and validating the model's learning without overfitting.
+The assertion in VA3 ensures the accuracy score on the test set exceeds 0.95, confirming the model's reliability and generalization capability.
+For VA28, the assertion checks whether the average development recall in the last ten evaluations exceeds 0.85, ensuring high recall rate critical for tasks like medical diagnostics.
+In VA11, the assertion ensures the average development BLEU score in the last ten evaluations exceeds 35, confirming the model produces high-quality translations.
+Lastly, in VA107, the assertion checks whether the R2 score exceeds 0.6, validating the model's fit and reliability for making accurate predictions in real-world scenarios.
+These validations confirm that iterative training processes lead to meaningful improvements and ensure the models' readiness for deployment and practical applications.
+
+\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
+The visualizations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
+Assertions associated with these visualizations ensure models meet a specific performance criteria.}
+
+\subsubsection{Distribution Check ($N = 1$)}
+
+\begin{table}
+  \centering
+  \caption{VA pairs used to check data distribution.}
+  \begin{tabular}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Visual Output}}&
+    \emph{\textbf{Assertion}}\\
+    \midrule
+
+    VA13 &
+    \includegraphics[width=\linewidth]{va13.png} &
+    \lstinline[]$assert 80 < boot_mean_mean < 81$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:va-dist-check}
+\end{table}
+
+The visualization in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
+
+Bootstrapped sample means are calculated by repeatedly resampling the original dataset with replacement and computing the mean for each resampled dataset.
+This process generates a distribution of the mean, known as the bootstrap distribution, which helps in estimating the variability and central tendency of the original dataset's mean.
+
+The corresponding assertion checks whether the bootstrapped mean lies between 80 and 81, ensuring that the statistical estimation is consistent with expected values.
+This validation is crucial for confirming the reliability of the bootstrapped estimates and the accuracy of statistical conclusions.
+
+\highlight{\textbf{Finding 3.1} One visualization is used to display the distribution of bootstrapped sample means.
+The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
+
+\subsubsection{Data Validation ($N = 9$)}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used for data validation.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used for data validation (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA30 &
+  \includegraphics[width=\linewidth]{va30.png} &
+  \lstinline[]$assert scaled_data[scaled_data.argmax()] <= 1$\\
+
+  V35 &
+  \includegraphics[width=\linewidth]{va35.png} &
+  \lstinline[]$assert(all(df_lower['inspection_score'] > 75))$\\
+
+  V57 &
+  \includegraphics[width=\linewidth]{va57.png} &
+  \lstinline[]$assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2$\\
+
+  VA62 &
+  \includegraphics[width=\linewidth]{va62.png} &
+  \lstinline[]$assert not np.allclose(x_aug, x_original)$\\
+
+\end{longtable}\label{tab:va-data-validatio}
+
+The VA pairs in Table~\ref{tab:va-data-validatio} collectively illustrate various data validation techniques essential for ensuring the integrity and quality of data used in ML models.
+
+The visualization in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
+The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
+The scaling process adjusts the data to fall within a specified range of 0 and 1.
+The corresponding assertion checks whether the maximum value in the scaled data is less than or equal to 1, ensuring that the scaling procedure was applied correctly.
+
+The visualization in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
+The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
+This comparison helps identify any disparities in health inspection scores based on income levels.
+The corresponding assertion checks that all inspection scores in the lower-income areas are above 75, ensuring that the data reflects a minimum standard of health inspection scores in these areas.
+
+The visualization in VA57 shows a histogram of transformed data after it has been normalised.
+The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
+
+The visualization in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
+This heatmap helps in visualizing the differences and similarities between the original and augmented data.
+The corresponding assertion checks that the augmented data \lstinline{x_aug} is not nearly identical to the original data \lstinline{x_original}, ensuring that the data augmentation process has introduced sufficient variability.
+
+These validations are crucial for confirming that data preprocessing techniques such as scaling, transformation, and augmentation have been correctly applied.
+Ensuring the integrity and quality of the data is essential for the subsequent steps in the machine learning pipeline, as it directly impacts the robustness, reliability, and generalization of the models developed.
+By validating these steps, we can maintain high standards of data quality and model performance, which are fundamental for accurate and meaningful analyses.
+
+\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original.
+These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}
+
+\section{Discussion}\label{sec:discuss}
+
+Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle.
+Making such decisions however, requires manual validation of the outputs.
+For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training.
+Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks.
+Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself.
+Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.
+
+Assertions can play a crucial role in automatically validating critical assumptions in ML workflows.
+As seen in the results from RQ2, assertions can verify that data distributions remain consistent with expectations or that model performance metrics meet predefined thresholds.
+Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged, reducing the reliance on manual verification and enhancing the robustness and maintainability of ML pipelines.
+Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development.
+This is particularly important in collaborative environments where multiple practitioners might work on the same project.
+Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
+Moreover, they enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells.
+
+% P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
+This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
+In this case study, managing dependencies through external tools is recommended.
+Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
+Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
+
+% NOTE: here we need to bring in prior literature on ML testing; and how majority of the research is working on testing using a test dataset; however we also have data and models
+% NOTE: 3 pillars of of ML (sato2019continuous)
+
+From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions.
+This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community.
+The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.
+
+We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties.
+Current guidelines on ML coding practices require enhancement to emphasize the importance of testing.
+There is a pressing need to steer education and research in Software Engineering towards addressing the specific challenges faced when developing ML-enabled software systems.
+Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
+
+The assertions presented in RQ2 and RQ3 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
+First, the assertions identified in this study lack the support of dedicated testing libraries.
+This suggests that the culture of testing ML code is still in its infancy, and ML practitioners are not yet fully exploring or leveraging software testing libraries.
+Secondly, existing testing libraries are not specifically tuned to the needs of ML code, leading practitioners to rely on basic built-in approaches.
+This highlights the necessity for developing specialized testing frameworks that cater to the unique requirements of ML projects.
+
+Visualizations are an effective tool for interpreting complex ML models and data transformations.
+However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}.
+We argue that the need for translating visual insights into formal assertions is more pressing.
+The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners.
+Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
+
+In the VA pairs from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
+This can lead to a mismatch of expectations in future versions of the ML software.
+For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
+The VA pairs presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
+Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
+For example, a linear correlation between the true and predicted labels is evident in VA107 presented in Table~\ref{tab:va-model-perf}.
+Yet, its formal assertion necessitates defining an appropriate threshold for the coefficient of determination metric.
+Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
+There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
+
+% TODO: integrate our NLBSE paper
+Our results show a clear need for developing ML-specific testing tools.
+Such tools should seamlessly integrate with computational notebooks and provide automated suggestions for assertions can greatly enhance the reliability and robustness of ML models.
+These tools can also aid in documenting and standardizing the use of domain-specific "magic" thresholds, ensuring that they are consistently applied and maintained.
+
+% NOTE: we find overlap between the manual tests performed using the output of code cells and the assertions.
+
+% NOTE: bring in prior work on reproducibility, code quality here
+% NOTE: notebooks have been adopted by a much larger scientific community; not eveybody has a background in software engineering; we need more tools to aid notebook users...
+
+\section{Implications}
+
+\paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks.
+Assertions play a crucial role in validating critical assumptions automatically, reducing the reliance on manual checks, and enhancing the robustness and maintainability of ML pipelines.
+By embedding assertions, developers can document the decisions made during model development, ensuring that these decisions are consistently applied and preserved throughout the ML lifecycle.
+This practice not only facilitates smoother transitions and handovers in collaborative environments but also improves the reproducibility of ML experiments by ensuring consistent validation checks.
+
+\paragraph{\textbf{For Researchers}} The use of assertions in Jupyter notebooks, though currently uncommon, is an emerging testing practice within the ML community.
+This study highlights the potential for assertions to enhance the reliability and robustness of ML models by automating the validation of critical assumptions.
+Researchers are encouraged to explore the development and evaluation of ML-specific testing methodologies that leverage assertions.
+This could involve creating new testing frameworks and tools tailored to the unique challenges of ML systems, thus contributing to the advancement of software engineering practices in the ML domain.
+
+\paragraph{\textbf{For Notebook Tool Developers}} There is a clear need for automated tools that suggest assertions based on the outputs of code cells in Jupyter notebooks.
+Tool developers can play a pivotal role by creating ML-specific testing tools that seamlessly integrate with computational notebooks.
+These tools should provide automated suggestions for assertions, helping practitioners to translate visual insights into formal, executable checks.
+By bridging the gap between qualitative evaluation and analytical assertions, such tools can significantly enhance the reliability and robustness of ML models, ensuring that they meet predefined performance standards consistently.
+
+\paragraph{\textbf{For Educators}} Educators should emphasize the importance of testing ML code written in Jupyter notebooks as part of the curriculum.
+By incorporating testing practices and the use of assertions into ML education, students can develop a deeper understanding of the importance of automated validation in ensuring model reliability and robustness.
+Teaching these best practices early on can help cultivate a culture of rigorous testing among future ML practitioners, preparing them to address the specific challenges faced in ML-enabled software systems.
+
+\paragraph{\textbf{For General Notebook Users}} Awareness of software engineering best practices, including the use of assertions, is crucial for all users of Jupyter notebooks.
+Assertions serve as an executable form of documentation, capturing the critical assumptions and decisions made during the development process.
+By adopting assertions, users can ensure that their notebooks are more robust, maintainable, and reproducible.
+This practice helps in maintaining consistency and reliability, especially in collaborative projects where multiple users might work on the same notebook, ensuring that the assumptions and validations are transparent and consistently applied.
+
+\section{Threats to Validity}\label{sec:threats}
+TODO
+
+\section{Related Work}\label{sec:related}
 
@@
 paper.
+
+\subsection{Feedback in Software Engineering and Machine Learning}
+
+% the role of early adoption of feedback when developing software, agile
+% similar concept exists in ML: crisp-dm, mldl, sculley, amershi, sambasivan
+% feedback from one part of the pipeline feeds into another,
+
+\subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
+
+% start with a very quick review of the workflow: REPL style of development; break code into smaller chunks; faster feedack; iterative style of development
+% link to the existing work on notebooks from chi: design, benefits and challenges
+% then go into the SE side of software written in notebooks: code quality
+% adoption of notebooks by ML community
+
+% start with a quick introduction to the anatomy of a computational notebook: code is broken down into smaller chunks represented as cells, cells can contain code (which are executable and immediately display the result of the execution), or markdown cells used for the purpose of documenting or creating a narrative for collaborative work.
+
+% the move on to the benefits of this style of coding: the ML pipeline contains several components, which can be broken down into smaller, more manageable chunks.
+These can be individually developed incorporating the rapid feedback provided from the REPL style of coding.
+
+% then summarise the pain-points of working in notebooks: use chattopadhyay2020whats
+Computational notebooks have been ubiquitously adopted by the ML community for developing ML enabled systems.
+Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing, dependenvcy management and versioning~\cite{pimentel2019large-scale,wang2020better,chattopadhyay2020whats}.
+
+\emph{Psallidas et al}~\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
+Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
+\emph{Pimentel et al}~\cite{pimentel2019large-scale} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
+Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
+To enable future research on computational notebooks, \emph{Quaranta et al}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
+
+Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020whats}.
+Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
+The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
+
+To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
+
+Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.
+Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
+This study leverages the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
 
@@
 
-\subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
-
-Computational notebooks have been ubiquitously adopted by the machine
-learning community for developing ML enabled systems. Although
-originally intended to promote reproducible software, computational
-notebooks are far from being reproducible due to lack of software
-engineering best practices such as separation of concern, testing and
-versioning~\cite{pimentel2019large,wang2020better,chattopadhyay2020wrong}.
-
-\emph{Psallidas et al} provide an overview of the evolving landscape
-of computational notebooks by analysing six million Python notebooks,
-two million enterprise Data Science (DS) pipelines, source code and
-metadata from over 900 releases of 12 important DS libraries. Their
-findings can be used by system builders for improving DS tools and
-also by DS practitioners to understand the current trends in
-technologies they should focus on. \emph{Pimentel et al} mined 1.4
-million notebooks from Github to conduct an empirical study on the
-coding practices in computational notebooks. Based on their analysis,
-the authors propose guidelines on improving reproducibility of
-computational notebooks. To enable future research on computational
-notebooks, \emph{Quaranta et al} mine Kaggle and present
-\textit{KGTorrent}, a public dataset consisting of approximately 2.5
-million Jupyter notebooks written in Python. The dataset also contains
-a relational database dump of metadata regarding publicly available
-notebooks on Kaggle.
-
-Studies with human subjects have been conducted to gain a deeper
-understanding of the challenges faced by ML practitioners when
-developing ML models inside notebooks. The results indicate that ML
-practitioners work in a highly iterative fashion, often experimenting
-with multiple strategies to analyse the data or produce meaningful
-visualisations~\cite{kandel2012enterprise, kery2018story,
-  liu2019understanding, chattopadhyay2020wrong}. Studies have also
-been conducted to understand how practitioners generate, evaluate and
-manage alternative hypothesis, visual designs, methods, tools,
-algorithms and data sources to arrive at the final
-implementation~\cite{liu2019understanding,kandel2012enterprise}. The
-findings from these studies can be used as guidelines for improving
-existing notebook technologies or designing new ones.
-
-To manage and prune multiple versions of code that accumulate over
-time when developing in notebooks, \emph{Head et al} propose a code
-gathering tool that allows practitioners to review and only keep the
-relevant version of code. Other tools such as \textit{WrangleDoc} and
-\textit{VizSmith} have also been proposed to aid ML practitioners when
-working within computational notebooks~\cite{yang2021subtle,
-  bavishi2021vizsmith}.
-
-Jupyter notebooks have been widely adopted by the DS and ML
-communities to develop ML
-pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
-Computational notebooks provide a rich source of data in the form of
-natural text, code and visualisations. Computational notebooks also
-allow ML practitioners to present the knowledge gained from the
-analysis in a narrative that can be shared with
-others~\cite{rule2018exploration}. This study leverages the
-computational narrative present within notebooks to identify
-visualisations and assertions that are semantically linked to each
-other.
-
 \subsection{Visualisations and Machine Learning}\label{sec:visualisations}
@@
 
-\section{Study Design}
-
-\subsection{Motivating Example}
-
-Sara is a data scientist who is developing a machine learning model to predict customer churn.
-Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells.
-Her workflow is dynamic and exploratory, typical of modern machine learning practices.
-
-Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution.
-As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features.
-This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.
-
-With data preprocessing complete, Sara proceeds to build her initial model.
-After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy and loss after the first epoch.
-The results are disappointing, prompting her to adjust the learning rate and batch size directly in subsequent cells.
-Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings.
-
-Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions.
-At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.
-
-Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively.
-Each decision in Sara's workflow—from preprocessing adjustments to hyperparameter tuning and model validation—is guided by the iterative feedback she receives from the outputs of her notebook's code cells.
-
-% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster!
-(notebook keeps running!)
-Once all experimentation and model tuning are complete, and Sara is ready to deploy the model into a production environment, she adds assertions as a final step.
-In the dynamic, interactive environment of Jupyter notebooks, where code cells can be executed in any order, there is a risk of arriving at inconsistent states that may affect the model's performance unpredictably.
-Assertions act as a safeguard against these inconsistencies by verifying that the data fed into the model adheres to expected formats and distributions and that the model's performance metrics meet predetermined standards.
-For instance, assertions are used to ensure that no data columns have null values, that the range of input data features remains within expected bounds, and that model accuracy thresholds are consistently met regardless of the execution order of the notebook cells.
-This practice not only reinforces the stability and reliability of the deployed model but also provides clear, executable documentation of the critical decisions made throughout the model's development.
-When new team members are onboarded, these assertions allow them to quickly grasp the underlying assumptions and operational parameters of the model, ensuring ongoing robustness and performance integrity.
-This is crucial in maintaining the quality and reliability of the model as it transitions from a development setting to a production environment.
-
-\subsection{Exploratory Analysis}
-
-We undertook an exploratory analysis of the collected data to formulate our research questions.
-
-\subsection{Research Questions}
-
-\begin{description}
-  \item[RQ1.] \textbf{How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
-  \item[RQ2.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
-  \item[RQ3.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
-\end{description}
-
-\subsection{Data Collection}\label{sec:data-collect}
-
-% TODO we need to explain why we only focus on the last statement of code cells (when visualisation code written anywhere produces an output); this is much harder to parse: we need to construct an exhaustive list of python visualisation libraries; people can use any naming convension to...
-
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{data-collection.pdf}
-\label{fig:data-collection}
-\caption{Overview of data collection methodology used in this study}
-\end{figure}
-
-Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.
-
-We mined public repositories from Github to collect $49923$ Jupyter notebooks.
-We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them.
-The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
-
-For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github.
-To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.
-
-Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.
-
-We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}.
-Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
-
-For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells.
-We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.
-
-Similarly, for implicit feedback, we first isolate code cells that produce an output.
-The output of a cell may either be from a ``print'' statement or the last statement of the cell.
-We collect both of them using the AST of the cell.
-
-Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.
-
-\subsection{Case Studies}
-
-% TODO: might need a visual aid here
-
-We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs).
-To surface interesting candidates for the case studies, we apply NLP techniques as described below.
-
-% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
-Thats why we need to apply these techniques to surface unique and interesting candidates!
-
-The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed.
-Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
-The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
-
-We apply stratified random sampling to identify the candidates for the case study analysis.
-The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
-The candidates are then divided into quartiles based on the aggregate value.
-A candidate is randomly drawn from each bin and analysed as an in-depth case study.
-The analysis is stopped when the time resource is exhausted.
-
-During each case study, we analyse the code of the candidate to understand its purpose.
-Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.
-
-\section{Results}
-
-% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output.
-Somethimes actual output is not possible (eg.
-P68) because the statement may be inside an if statement.
-
-\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
-
-\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
-% (O2, O4, O9, O14, O20, O25, O48)
-
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to verify distribution of data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    O2&
-    \lstinline[]$_ = sns.catplot(x='category_id', y='likes', data=train, height=5, aspect=1.5)$&
-    \includegraphics[width=\linewidth]{distribution-check-catplot.png}\\
-
-    O9&
-    \lstinline[]$sns.kdeplot(data=data.loc[ data['Survived'] == 0].Age, label='Died', shade=True)$&
-    \includegraphics[width=\linewidth]{distribution-check-kdeplot.png}\\
-
-    O14&
-    \lstinline[]$pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])$&
-    \\
-
-    O25&
-    \lstinline[]$sns.countplot(house_pred['OverallQual'])$&
-    \includegraphics[width=\linewidth]{o25.png}\\
-
-    O48&
-    \lstinline[]$x_train.describe()$&
-    \\
-
-    \bottomrule
-  \end{tabular}
-  \label{tab:distribution-check}
-\end{table}
-
-Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data.
-In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature.
-Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
-Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}.
-Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data.
-The statement is written immediately after normalising the features in the training data.
-Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output.
-Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.
-
-Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
-For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
-During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
-Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.
-
-\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
-These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}
-
-\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
-% (O6, O10)
-
-\begin{table}
-  \centering
-  \caption{Cell outputs used to verify linear relationship between features in the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Output}}\\
-    \midrule
-
-    O6&
-    \lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$&
-    \includegraphics[width=\linewidth]{linear-relation-check-lineplot.png}\\
-
-    O10&
-    \lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$&
-    \includegraphics[width=\linewidth]{linear-relation-check-regplot.png}\\
-    \bottomrule
-  \end{tabular}
-\label{tab:linear-relation-check}
-\end{table}
-
-Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data.
-In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
-Output \emph{O10} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
-
-Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
-However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
-Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.
-
-\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
-Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
-These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
-
-\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
-% (P71, P86, O64, O66, P68, P82, P107)
-
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to verify if different types of resources are available on the system where the notebook is being executed.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    P68&
-    \lstinline[]$print('GPU is available')$\\
-
-    P71&
-    \lstinline[]$print('Hub version: ', hub.__version__)$\\
-
-    P82&
-    \lstinline[]$print('Running on TPU ', tpu.master())$\\
-
-    P86&
-    \lstinline[]$print('Cuda is available')$\\
-
-    P107&
-    \lstinline[]$print('Model loaded')$\\
-
-    O64&
-    \lstinline[]$full_table.head(-5)$\\
-
-    O66&
-    \lstinline[]$prostate_cancer_df.shape$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:resource-check}
-\end{table}
-
-% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
-Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed.
-Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU.
-This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory.
-For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory.
-Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data.
-These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.
-
-Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
-This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
-
-\highlight{\textbf{Finding 1.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
-These checks include confirming the availability of compute resources like GPUs and TPUs (P68, P82 and P86), ensuring datasets or pre-trained models are successfully loaded (P107, O64 and O66), and verifying the presence of specific library versions (P71).
-These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
-
-\subsubsection{Execution time check ($N = 1$)}
-% (P66)
-
-\begin{table}
-  \centering
-  \caption{Print statement used to check the total execution time of training and cross-validating an ML model.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    P66&
-    \lstinline[]$print('Total Run Time:')$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:exec-time}
-\end{table}
-
-Table~\ref{tab:exec-time} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
-\emph{P66} is also used in other code cells in the notebook that train other ML models.
-
-While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
-Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{CITEME}.
-
-\highlight{\textbf{Finding 1.4} The study identifies one output (P66) used to check the total execution time of training and cross-validating an ML model.
-The print statement is used for evaluating the efficiency of model training processes, especially when multiple models are trained iteratively.}
-
-\subsubsection{Missing value check ($N = 3$)}
-% (O12, O36, P74)
-
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to check for missing data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    P74&
-    \lstinline[]$print(train_df.isnull().sum())$\\
-
-    O12&
-    \lstinline[]$sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')$&
-    \includegraphics[width=\linewidth]{missing-value-check.png}\\
-
-    O36&
-    \lstinline[]$test.isna().sum().unique()$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:missing-value}
-\end{table}
-
-Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
-While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
-
-Checking for missing values is a critical step in the preprocessing phase of a machine learning project because it significantly influences the quality and performance of the model.
-Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
-Many machine learning algorithms, including those that involve matrix multiplication operations such as linear regression and neural networks, demand complete numerical datasets to perform calculations.
-Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
-
-\highlight{\textbf{Finding 1.5} The study identifies three outputs (P74, O12 and O36) used to check for missing data.
-Missing values can lead to biased or incorrect conclusions, and many ML algorithms require complete numerical datasets for accurate calculations.
-Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
-
-\subsubsection{Model performance check ($N = 33$)}
-% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
-
-\begin{table}
-  \centering
-  \caption{Various cell outputs used to check the performance of an ML model after training.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    P3&
-    \lstinline[]$print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')$&
-    \\
-
-    P6&
-    \lstinline[]$print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))$&
-    \\
-
-    P18&
-    \lstinline[]$print('The Accuracy is:', accuracy_score(y_test, y_pred))$&
-    \\
-
-    P50&
-    \lstinline[]$print('Classification Report: SVM (validation data)')$&
-    \\
-
-    P54&
-    \lstinline[]$print('Intercept value:', lm.intercept_)$&
-    \\
-
-    O3&
-    \lstinline[]$skplt.metrics. plot_confusion_matrix(Y_val, Vote.predict(X_val), normalize=True, figsize=(10, 10))$&
-    \includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}\\
-
-    O52&
-    \lstinline[]$spot_check_recs(classifier, 910)$&
-    \\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-perf}
-\end{table}
-
-Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
-Besides the \emph{accuracy} of the model (\emph{P3} and \emph{P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
-
-% TODO: this feels hallucinated; revisit
-Print statement \emph{P54} prints the intercepts of a linear regression model.
-Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
-This insight is critical for accurate model interpretation and effective decision-making in practical applications.
-
-Output \emph{O3} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
-The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
-
-Output \emph{O52} shows a function implemented by the practitioners to manually evaluate an image classification model on a random sample of input images.
-The function \texttt{spot\_check\_recs} displays the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
-
-During the model development phase, the performance of an ML model is often printed to facilitate comparisons against other models or variations.
-Throughout this continuous experimentation phase, authors may adjust model parameters or modify the data by engineering new features.
-The model is then re-trained to check if these changes lead to performance improvements.
-
-\highlight{\textbf{Finding 1.6} The study identifies 33 outputs used to check the performance of trained ML models.
-These include outputs that print key performance metrics (P3, P6, P18, P50), a normalised heatmap used to visualise the confusion matrix of a model's predictions (O3) and a custom function written for manual evaluation of model predictions on random input samples (O52).
-These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
-
-\subsubsection{Neural network architecture ($N = 1$)}
-% (P92)
-
-\begin{table}
-  \centering
-  \caption{Print the architecture of a neural network.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-    P92&
-    \lstinline[]$print(MyNetwork)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:network-architecture}
-\end{table}
-
-% TODO: need more description and context around this print statement (where was it written, and why?)
-Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.
-
-Printing the structure of a neural network serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
-It helps in verifying configurations like layer dimensions and operations before initiating computationally intensive training, ensuring the model is set up efficiently for the intended tasks.
-This practice is fundamental for maintaining accuracy, facilitating collaboration, and enhancing the reproducibility of the model.
-
-\highlight{\textbf{Finding 1.7} The study identifies one output (P92) used to print the architecture of a neural network.
-Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
-
-\subsubsection{Shape check ($N = 3$)}
-% (P4, P32, P117)
-
-\begin{table}
-  \centering
-  \caption{Example of outputs used to check the shape of the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}} & \emph{\textbf{Code}}\\
-    \midrule
-    P4 &
-    \lstinline[]$print('no.of examples in test data : ', len(test_data))$\\
-
-    P32 &
-    \lstinline[]$print('Training set shape : ', x_train.shape)$\\
-
-    P117&
-    \lstinline[]$print('Y_train.shape: ', Y_train.shape)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:shape-check}
-\end{table}
-
-Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data.
-\emph{P4} prints the number of rows or examples in the test set.
-\emph{P32} and \emph{117} prints the number of rows and columns (or features) in the training and test data respectively.
-
-Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
-Primarily, the number of features in the training set must match those in the testing set.
-This alignment is crucial because statistical machine learning models are trained on specific data dimensions and expect the same dimensional structure during inference to perform accurately.
-Similarly, in neural network architectures, the configuration of input layers depends directly on the shape of the training data, dictating the number of input neurons needed.
-Furthermore, the correspondence between the number of test examples and their respective labels is essential for accurately computing performance metrics, such as accuracy.
-
-Such checks are indicative of the meticulous attention to detail necessary in the model training and validation process, underscoring the importance of shape validation in achieving reliable machine learning outcomes.
-
-\highlight{\textbf{Finding 1.8} The study identifies three outputs used to check the shape of data.
-These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
-
-\subsubsection{Type check ($N = 2$)}
-% (O71, P43)
-
-% TODO: the narrative is not as strong here; it would help to also show the actual output as well?
-For instance, in P43 by itself doesn't make that much sense, the output however shows that all features in the dataset are floats which is a pre-requisite for ML training data
-\begin{table}
-  \centering
-  \caption{Outputs used to check the data type of features in the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}} &
-    \emph{\textbf{Code}}\\
-
-    P43&
-    \lstinline[]$print('data type:', images.dtype)$\\
-
-    O71&
-    \lstinline[]$type(Y)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:type-check}
-\end{table}
-
-% TODO: for O71, the output shows its a ndarray; without showing this to the reader we cannot make a strong narrative around these type checks
-Table~\ref{tab:type-check} illustrates outputs of cells found in our analysis, where checks are employed to verify the data types of the features in the dataset.
-Print statement \emph{P43} is written prior to training a neural network.
-The output validates that all features in the training data are numerical.
-Similarly, output \emph{O71} shows the data type of the target column.
-
-Type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
-Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis (EDA) phase of developing machine learning models.
-Since all machine learning algorithms perform mathematical operations, such as matrix multiplications, they require the input data to be entirely numerical to avoid computational errors.
-The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
-For example, while numerical features often require normalisation to bring them within a specific scale, categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
-
-\highlight{\textbf{Finding 1.9} The study identifies two outputs used to verify the data types of features in the dataset.
-These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
-Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
-
-\subsubsection{Spot checks ($N = 5$)}
-% (O56, O60, P64, P67, P114)
-
-\begin{table}
-  \centering
-  \caption{Example of outputs used to perform spot checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    O60 &
-    \lstinline[]$X_pca.head()$\\
-
-    P64 &
-    \lstinline[]$print(np.max(cur[:, :, 1]))$\\
-
-    P114 &
-    \lstinline[]$print(onehot_encoded)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:spot-check}
-\end{table}
-
-Table~\ref{tab:spot-check} presents outputs which were used to perform spot checks at various stages of the ML development cycle.
-In \emph{O60}, the practitioner verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
-\emph{P64} demonstrates another case which is particularly relevant in image processing contexts.
-The maximum value in the second channel of a 3D numpy array is extracted and verified, ensuring that the data manipulation retains its expected properties.
-Lastly, \emph{P114} assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly.
-
-Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
-These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
-Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
-
-\highlight{\textbf{Finding 1.10} This study finds five outputs used to perform spot checks at various stages of the MLDL.
-Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
-
-\subsubsection{Model training ($N = 4$)}
-% (O8, O31, O42, P77)
-
-\begin{table}
-  \centering
-  \caption{Example of outputs used to monitor model training.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    O8 &
-    \lstinline[]$autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)$\\
-
-    O31 &
-    \lstinline[]$adaBoost.fit(X_train, y_train)$\\
-
-    O42&
-    \lstinline[]$m_r.best_params_$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-training}
-\end{table}
-
-During training it is common practice to monitor progress periodically, as illustrated by outputs \emph{O8} in Table~\ref{tab:model-training}, where the training loss and accuracy are printed to give real-time feedback on the model's learning efficacy.
-This feedback allows for adjustments in training parameters or early stopping to prevent overfitting.
-For instance, in output O8, the autoencoder is trained using a custom callback from \texttt{keras\_utils} to provide a progress update without cluttering the output stream, which is particularly useful in lengthy training sessions.
-In contrast, output \emph{O31} represents a crucial validation step where the \texttt{adaBoost} model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
-Output \emph{O42} checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
-
-Each of these outputs underscores different facets of the model training phase, from initialization and real-time monitoring to validation and finalization, illustrating the layered complexity of developing predictive models.
-
-\highlight{\textbf{Finding 1.11} The study identifies four outputs used to monitor model training progress.
-These outputs provide real-time feedback on training loss and accuracy (O8), validate model adherence to expected data patterns (O31), and identify optimal parameters through tuning methods (O42).
-These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
-
-\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
-
-% TODO: bring in the notion of pre and post condition checks (the wikipedia page on assertions is very interesting: https://en.wikipedia.org/wiki/Assertion_(software_development)); specifically the part on run-time assertions
-
-\subsubsection{Batch size check ($N = 3$)}
-% (A21, A28, A70)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to validate the batch size of input data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A21 &
-    \lstinline[]$assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'$\\
-
-    A28 &
-    \lstinline[]$assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'$\\
-
-    A70 &
-    \lstinline[]$assert n_img > batch_size$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:batch-size}
-\end{table}
-
-Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
-\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
-Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
-\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
-
-In the context of neural network training, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilization, particularly GPU cores.
-Batching allows for a more stable and smoother learning curve as the model's weights are updated not for each individual example, but rather based on the average gradient of a batch.
-This technique not only optimizes learning but also enhances the model's generalizability due to more robust gradient estimations.
-Given these benefits, it becomes crucial to ensure that the batch size is appropriately set to match the hardware's capacity, avoiding memory overflow issues that can halt training.
-These assertions are strategically placed to prevent common errors that could undermine training efficiency and model performance, thereby safeguarding the training process from potential disruptions due to misconfigured batch sizes.
-
-\highlight{\textbf{Finding 2.1} This study identifies three assertions validating batch size configurations.
-These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
-These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
-
-\subsubsection{Data leakage check ($N = 1$)}
-% (A33)
-
-\begin{table}
-  \centering
-  \caption{Assert used to prevent data leakage.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A33&
-    \lstinline[]$assert len(set( tr_df.PetID.unique()).intersection(valid_df.PetID.unique())) == 0$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:data-leakage}
-\end{table}
-
-Assertion \emph{A33} in Table~\ref{tab:data-leakage} ensures that there are no overlapping \texttt{PetID} values between the training set \texttt{tr\_df} and the validation set \texttt{valid\_df}, thereby preventing any possibility of data leakage.
-
-Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in machine learning to prevent models from overfitting.
-Overfitting occurs when a model is excessively complex, learning not only the underlying patterns but also the noise in the training data, which can degrade its performance on unseen data.
-To guarantee that the model can generalize effectively to new examples, it's crucial to verify that the training and validation sets are completely distinct with no shared examples.
-By strictly separating these datasets, the validation phase serves its purpose of providing an unbiased evaluation of the model's ability to generalize---to data similar to, but not identical to that which it was trained on---thus maintaining the integrity and validity of the evaluation metrics derived during model testing.
-
-\highlight{\textbf{Finding 2.2} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
-Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
-This practice maintains the integrity and validity of the model evaluation metrics.}
-
-\subsubsection{Existence check ($N = 8$)}
-% NOTE: this is a combination of missing-value-check and existence-check
-% (A23, A42, A43, A50, A51, A63, A79, A86)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to perform existence checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A23 &
-    \lstinline[]$assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())$\\
-
-    A42 &
-    \lstinline[]$assert not lab_s.isnull().values.any()$\\
-
-    A43 &
-    \lstinline[]$assert len(data) != 0, 'cannot divide by zero'$\\
-
-    A50 &
-    \lstinline[]$assert not np.any(np.isnan(X))$\\
-
-    A51 &
-    \lstinline[]$assert data.target.notnull().all()$\\
-
-    A63 &
-    \lstinline[]$assert X.isnull().sum().sum() == 0$\\
-
-    A79 &
-    \lstinline[]$assert not processed_data_df.isna().any().any()$\\
-
-    A86 &
-    \lstinline[]$assert p0 in poi_info.index$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:existence-check}
-\end{table}
-
-Assertions \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
-Assertion \emph{A43} guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
-Assertions \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
-Assertions \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
-
-Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modeling.
-These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
-Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values.
-These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
-
-\highlight{\textbf{Finding 2.3} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
-These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
-
-\subsubsection{Mathematical property checks ($N = 4$)}
-% (A3, A25, A56, A64)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to validate mathematical properties of neural networks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A3 &
-    \lstinline[]$assert (xH - wH) % self.stride == 0$\\
-
-    A25 &
-    \lstinline[]$assert test_output.std() < 0.15, "Don't use batchnorm here"$\\
-
-    A56 &
-    \lstinline[]$assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))$\\
-
-    A64 &
-    \lstinline[]$assert np.allclose(T, T.T)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:maths-check}
-\end{table}
-
-Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
-\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior.
-\emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations.
-Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.
-
-When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigor is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
-These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
-
-\highlight{\textbf{Finding 2.4} Four assertions validate key mathematical properties of data and model outputs.
-These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
-These checks confirm the robustness and reliability of model computations.}
-
-\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
-% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check performance of ML models.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A7 &
-    \lstinline[]$assert len(neighbours_1) == 20, "Neighbors don't match!"$\\
-
-    A15 &
-    \lstinline[]$assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))$\\
-
-    A19 &
-    \lstinline[]$assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'$\\
-
-    A38 &
-    \lstinline[]$assert 0.75 < auc(fpr, tpr) < 0.85$\\
-
-    A58 &
-    \lstinline[]$assert np.isclose(accuracy, 0.9666666666666667)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-perf-explicit}
-\end{table}
-
-This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
-
-% TODO: verify if this applies to recommender systems as well?
-For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
-Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
-Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity.
-Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.
-
-The use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
-This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
-By allowing a small tolerance in the comparison, np.isclose ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
-
-\highlight{\textbf{Finding 2.5} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
-These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
-
-\subsubsection{Network architecture check ($N = 3$)}
-% (A11, A62, A75)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the architecture of neural networks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A11 &
-    \lstinline[]$assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()$\\
-
-    A62 &
-    \lstinline[]$assert activations[i + 1].shape[1] == self.encoder_layer_sizes[-1] * 2$\\
-
-    A75 &
-    \lstinline[]$assert reg in ['none', 'l2']$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-network-architecture}
-\end{table}
-
-Assertion \emph{A11} verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
-The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
-By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
-
-Assertion \emph{A62} checks that the shape of the activations at index \lstinline{i + 1} has the correct number of channels, which should be twice the size of the last encoder layer.
-This ensures that the intermediate activations maintain the expected structure necessary for the proper functioning of the subsequent layers in the network.
-Ensuring that the shape of activations is consistent throughout the network is crucial for maintaining the integrity of the data flow and preventing errors during the forward pass.
-
-Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'.
-This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behavior or suboptimal model performance.
-
-When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
-This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
-Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
-
-\highlight{\textbf{Finding 2.6} Three assertions verify the consistency of neural network architecture.
-These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
-These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
-
-\subsubsection{Resource check ($N = 7$)}
-% (A10, A14, A18, A37, A60, A67 A74)
-
-\begin{table}
-  \centering
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A10 &
-    \lstinline[]$assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."$\\
-
-    A14 &
-    \lstinline[]$assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'$\\
-
-    A18&
-    \lstinline[]$assert pd.__version__.rpartition('.')[0] == '1.0', f"Unexpected pandas version: expected 1.0, got {pd.__version__.rpartition('.')[0]}"$\\
-
-    A37 &
-    \lstinline[]$assert svm.fit_status_ == 0, 'Forgot to train the SVM!'$\\
-
-    A60 &
-    \lstinline[]$assert f2.gca().has_data()$\\
-
-    A67&
-    \lstinline[]$assert pm.__version__ == '3.9.2'$\\
-
-    A74 &
-    \lstinline[]$assert os.path.exists(image_dir)$\\
-    \bottomrule
-  \end{tabular}
-  \caption{Assertions used to validate availability of resources.}
-  \label{tab:assert-resource-check}
-\end{table}
-
-Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
-These checks help prevent runtime errors and ensure the proper execution of code.
-
-One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
-These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing resources.
-
-Additionally, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).
-A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
-The assertion \emph{A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
-This check prevents inconsistent model states from executing cells out of order or multiple times.
-
-Assertion \emph{A60} is used to verify the presence of data in visualizations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
-These checks are crucial for ensuring that the visualizations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
-
-Assertions \emph{A18} and \emph{A67} explicitly verify the versions of critical libraries.
-For instance, \emph{A18} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
-
-Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
-
-% NOTE: left over loads, move to implications if required
-% In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
-Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
-
-% These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
-
-\highlight{\textbf{Finding 2.7} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
-These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
-
-\subsubsection{Data shape check ($N = 26$)}
-% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
-
-% TODO: revisit the assertions in this section; we may have misclassified assertions or need to create a new group
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the shape of the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A5 &
-    \lstinline[]$assert y_valid.shape == (1132,)$\\
-
-    A17 &
-    \lstinline[]$assert X.shape[1] == 13, 'Did you drop/lose some columns in X? Did you properly load and split the data?'$\\
-
-    A29 &
-    \lstinline[]$assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'$\\
-
-    A31 &
-    \lstinline[]$assert img.shape == (112, 92)$\\
-
-    A76 &
-    \lstinline[]$assert len(encoding['token_type_ids']) == max_seq_length$\\
-
-    A84 &
-    \lstinline[]$assert red.get_shape().as_list()[1:] == [224, 224, 1]$\\
-
-    A90 &
-    \lstinline[]$assert len(X_train) == 2000$\\
-
-    A93 &
-    \lstinline[]$assert temp_embed.shape[0] == stride$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-shape-check}
-\end{table}
-
-Assertions \emph{A17} and \emph{A90} verify that the input features have the correct number of columns and examples, respectively.
-Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
-
-Assertions \emph{A5} and \emph{A29} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
-
-In the context of computer vision tasks, assertions \emph{A31} and \emph{A84} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
-
-Furthermore, assertions \emph{A76} and \emph{A93} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
-
-Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
-We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features, labels, predictions, images, sequences, and embeddings.
-These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.
-
-Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
-By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
-
-\highlight{\textbf{Finding 2.8} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
-These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
-
-\subsubsection{Type check ($N = 5$)}
-% (A2, A35, A40, A81, A88)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the data type of features in the dataset.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A2 &
-    \lstinline[]$assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'$\\
-
-    A35 &
-    \lstinline[]$assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"$\\
-
-    A40 &
-    \lstinline[]$assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)$\\
-
-    A81 &
-    \lstinline[]$assert is_all_ints(filled_df[r]) is True$\\
-
-    A88 &
-    \lstinline[]$assert isinstance(betas, np.ndarray)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-type-check}
-\end{table}
-
-Assertion \emph{A2} in Table~\ref{tab:assert-type-check} demonstrates a very common type check performed on the input features.
-This assertion ensures that the input features are represented as PyTorch float tensors, which is a requirement for many deep learning models and operations.
-
-Another type check is applied to verify the type of the machine learning models themselves.
-For instance, the \emph{A40} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
-This type of assertion is essential when working with different types of models, as it helps prevent errors arising from using incompatible methods or attributes.
-
-In addition to input features and models, type checks are also performed on data structures and intermediate objects.
-The assertion \emph{A35} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
-
-Type checking is not limited to objects and data structures--it can also be applied to individual columns or elements within a dataset.
-The assertion \emph{A81} checks if all the elements in a specific column (\lstinline[]$filled_df[r]$) of a DataFrame are integers.
-
-Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}.
-This assertion ensures that the \lstinline{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
-
-Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behavior.
-By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.
-
-\highlight{\textbf{Finding 2.9} Five assertions validate the data types of input features, models, and intermediate objects.
-These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
-
-\subsubsection{Data validation check ($N = 14$)}
-% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to perform data validation checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A41 &
-    \lstinline[]$assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))$\\
-
-    A44 &
-    \lstinline[]$assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1$\\
-
-    A46 &
-    \lstinline[]$assert np.isclose(stdev_norm, 1.0, atol=1e-16)$\\
-
-    A52 &
-    \lstinline[]$assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()$\\
-
-    A65 &
-    \lstinline[]$assert np.all(y <= nb_classes)$\\
-
-    A73 &
-    \lstinline[]$assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:data-validation}
-\end{table}
-
-% TODO: revisit, we have new assertions here; update the count
-Table~\ref{tab:data-validation} presents a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
-These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
-
-One type of validation check involves verifying the presence of specific values or ranges within arrays or columns.
-For example, assertion \emph{A41} checks if the unique values in the \lstinline{smoke} column of the DataFrame \lstinline{X} are limited to 0 and 1.
-Similarly, \emph{A65} ensures that all values in the array \lstinline{y} are less than or equal to a certain number of classes.
-
-Another common validation check is verifying the equality or closeness of values or arrays.
-Assertion \emph{A44} checks if the difference between the minimum and maximum values in a specific range of the \lstinline{entity_embedding} array is less than or equal to 1.
-Assertion \emph{A46} verifies that the \lstinline{stdev_norm} value is close to 1.0 within a specified absolute tolerance.
-
-Some assertions focus on validating the uniqueness or cardinality of values in arrays or DataFrames.
-For instance, \emph{A52} checks if the number of unique user IDs in the \lstinline{grouped_users} DataFrame is equal to the number of unique user IDs in the \lstinline{user_engagement} DataFrame.
-
-Other assertions validate the presence of specific values or conditions within DataFrames or other data structures.
-The assertion \emph{A73} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \lstinline{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
-
-By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
-
-\highlight{\textbf{Finding 2.10} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
-These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
-
-\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
-
-\subsubsection{Model Performance ($N = 15$)}
-
-\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-  \caption{VA pairs used to validate the performance of a model.} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endfirsthead
-
-  \caption[]{VA pairs used to validate the performance of a model (continued).} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endhead
-
-  \midrule
-  \endfoot
-
-  \bottomrule
-  \endlastfoot
-
-  VA2 &
-  \includegraphics[width=\linewidth]{va2.png} &
-  \lstinline[]$assert final_score > 300, 'not good enough for DQN'$\\
-
-  VA3 &
-  \includegraphics[width=\linewidth]{va3.png} &
-  \lstinline[]$assert accuracy_score(y_test, pred) > 0.95$\\
-
-  VA9 &
-  \includegraphics[width=\linewidth]{va9.png} &
-  \lstinline[]$assert np.mean(mean_rw_history[-10:]) > 10.$\\
-
-  VA11 &
-  \includegraphics[width=\linewidth]{va11.png} &
-  \lstinline[]$assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."$\\
-
-  VA15 &
-  \includegraphics[width=\linewidth]{va15.png} &
-  \lstinline[]$assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."$\\
-
-  VA28 &
-  \includegraphics[width=\linewidth]{va28.png} &
-  \lstinline[]$assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."$\\
-
-  VA68 &
-  \includegraphics[width=\linewidth]{va68.png} &
-  \lstinline[]$np.testing.assert_almost_equal(ans4, 0.3042764)$\\
-
-  VA107 &
-  \includegraphics[width=\linewidth]{va107.png} &
-  \lstinline[]$assert r2_gru > 0.6$\\
-
-\end{longtable}\label{tab:va-model-perf}
-
-The visualizations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
-VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
-These metrics collectively indicate the model's learning progress and stability during training.
-The visualization in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
-The general upward trend indicates learning and improvement in gameplay over time.
-VA15 presents the training and development loss over time for a character-level language model, showing a gradual decrease in loss and suggesting effective learning.
-
-The visualization in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
-This visualization aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
-The visualizations in VA28 display the training loss and development recall, respectively.
-The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
-The visualizations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
-The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
-In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
-
-The corresponding assertions for these visualizations ensure that the models meet specific performance standards.
-For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
-In VA9, the assertion checks whether the mean of the last ten rewards exceeds 10, ensuring stable and satisfactory learning progress.
-For VA15, the assertion checks if the mean of the first ten training observations is greater than the mean of the last ten, ensuring loss convergence and validating the model's learning without overfitting.
-The assertion in VA3 ensures the accuracy score on the test set exceeds 0.95, confirming the model's reliability and generalization capability.
-For VA28, the assertion checks whether the average development recall in the last ten evaluations exceeds 0.85, ensuring high recall rate critical for tasks like medical diagnostics.
-In VA11, the assertion ensures the average development BLEU score in the last ten evaluations exceeds 35, confirming the model produces high-quality translations.
-Lastly, in VA107, the assertion checks whether the R2 score exceeds 0.6, validating the model's fit and reliability for making accurate predictions in real-world scenarios.
-These validations confirm that iterative training processes lead to meaningful improvements and ensure the models' readiness for deployment and practical applications.
-
-\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
-The visualizations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
-Assertions associated with these visualizations ensure models meet a specific performance criteria.}
-
-\subsubsection{Distribution Check ($N = 1$)}
-
-\begin{table}
-  \centering
-  \caption{VA pairs used to check data distribution.}
-  \begin{tabular}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Visual Output}}&
-    \emph{\textbf{Assertion}}\\
-    \midrule
-
-    VA13 &
-    \includegraphics[width=\linewidth]{va13.png} &
-    \lstinline[]$assert 80 < boot_mean_mean < 81$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:va-dist-check}
-\end{table}
-
-The visualization in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
-
-Bootstrapped sample means are calculated by repeatedly resampling the original dataset with replacement and computing the mean for each resampled dataset.
-This process generates a distribution of the mean, known as the bootstrap distribution, which helps in estimating the variability and central tendency of the original dataset's mean.
-
-The corresponding assertion checks whether the bootstrapped mean lies between 80 and 81, ensuring that the statistical estimation is consistent with expected values.
-This validation is crucial for confirming the reliability of the bootstrapped estimates and the accuracy of statistical conclusions.
-
-\highlight{\textbf{Finding 3.1} One visualization is used to display the distribution of bootstrapped sample means.
-The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
-
-\subsubsection{Data Validation ($N = 9$)}
-
-\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-  \caption{VA pairs used for data validation.} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endfirsthead
-
-  \caption[]{VA pairs used for data validation (continued).} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endhead
-
-  \midrule
-  \endfoot
-
-  \bottomrule
-  \endlastfoot
-
-  VA30 &
-  \includegraphics[width=\linewidth]{va30.png} &
-  \lstinline[]$assert scaled_data[scaled_data.argmax()] <= 1$\\
-
-  V35 &
-  \includegraphics[width=\linewidth]{va35.png} &
-  \lstinline[]$assert(all(df_lower['inspection_score'] > 75))$\\
-
-  V57 &
-  \includegraphics[width=\linewidth]{va57.png} &
-  \lstinline[]$assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2$\\
-
-  VA62 &
-  \includegraphics[width=\linewidth]{va62.png} &
-  \lstinline[]$assert not np.allclose(x_aug, x_original)$\\
-
-\end{longtable}\label{tab:va-data-validatio}
-
-The VA pairs in Table~\ref{tab:va-data-validatio} collectively illustrate various data validation techniques essential for ensuring the integrity and quality of data used in ML models.
-
-The visualization in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
-The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
-The scaling process adjusts the data to fall within a specified range of 0 and 1.
-The corresponding assertion checks whether the maximum value in the scaled data is less than or equal to 1, ensuring that the scaling procedure was applied correctly.
-
-The visualization in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
-The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
-This comparison helps identify any disparities in health inspection scores based on income levels.
-The corresponding assertion checks that all inspection scores in the lower-income areas are above 75, ensuring that the data reflects a minimum standard of health inspection scores in these areas.
-
-The visualization in VA57 shows a histogram of transformed data after it has been normalised.
-The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
-
-The visualization in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
-This heatmap helps in visualizing the differences and similarities between the original and augmented data.
-The corresponding assertion checks that the augmented data \lstinline{x_aug} is not nearly identical to the original data \lstinline{x_original}, ensuring that the data augmentation process has introduced sufficient variability.
-
-These validations are crucial for confirming that data preprocessing techniques such as scaling, transformation, and augmentation have been correctly applied.
-Ensuring the integrity and quality of the data is essential for the subsequent steps in the machine learning pipeline, as it directly impacts the robustness, reliability, and generalization of the models developed.
-By validating these steps, we can maintain high standards of data quality and model performance, which are fundamental for accurate and meaningful analyses.
-
-\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original.
-These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}
-
-\section{Discussion}\label{sec:discuss}
-
-Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle.
-Making such decisions however, requires manual validation of the outputs.
-For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training.
-Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks.
-Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself.
-Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.
-
-Assertions can play a crucial role in automatically validating critical assumptions in ML workflows.
-As seen in the results from RQ2, assertions can verify that data distributions remain consistent with expectations or that model performance metrics meet predefined thresholds.
-Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged, reducing the reliance on manual verification and enhancing the robustness and maintainability of ML pipelines.
-Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development.
-This is particularly important in collaborative environments where multiple practitioners might work on the same project.
-Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
-Moreover, they enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells.
-
-% P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
-This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
-In this case study, managing dependencies through external tools is recommended.
-Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
-Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
-
-% NOTE: here we need to bring in prior literature on ML testing; and how majority of the research is working on testing using a test dataset; however we also have data and models
-% NOTE: 3 pillars of of ML (sato2019continuous)
-
-From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions.
-This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community.
-The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.
-
-We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties.
-Current guidelines on ML coding practices require enhancement to emphasize the importance of testing.
-There is a pressing need to steer education and research in Software Engineering towards addressing the specific challenges faced when developing ML-enabled software systems.
-Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
-
-The assertions presented in RQ2 and RQ3 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
-First, the assertions identified in this study lack the support of dedicated testing libraries.
-This suggests that the culture of testing ML code is still in its infancy, and ML practitioners are not yet fully exploring or leveraging software testing libraries.
-Secondly, existing testing libraries are not specifically tuned to the needs of ML code, leading practitioners to rely on basic built-in approaches.
-This highlights the necessity for developing specialized testing frameworks that cater to the unique requirements of ML projects.
-
-Visualizations are an effective tool for interpreting complex ML models and data transformations.
-However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}.
-We argue that the need for translating visual insights into formal assertions is more pressing.
-The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners.
-Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
-
-In the VA pairs from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-This can lead to a mismatch of expectations in future versions of the ML software.
-For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
-The VA pairs presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
-Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
-For example, a linear correlation between the true and predicted labels is evident in VA107 presented in Table~\ref{tab:va-model-perf}.
-Yet, its formal assertion necessitates defining an appropriate threshold for the coefficient of determination metric.
-Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
-There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
-
-% TODO: integrate our NLBSE paper
-Our results show a clear need for developing ML-specific testing tools.
-Such tools should seamlessly integrate with computational notebooks and provide automated suggestions for assertions can greatly enhance the reliability and robustness of ML models.
-These tools can also aid in documenting and standardizing the use of domain-specific "magic" thresholds, ensuring that they are consistently applied and maintained.
-
-% NOTE: we find overlap between the manual tests performed using the output of code cells and the assertions.
-
-% NOTE: bring in prior work on reproducibility, code quality here
-% NOTE: notebooks have been adopted by a much larger scientific community; not eveybody has a background in software engineering; we need more tools to aid notebook users...
-
-\section{Implications}
-
-\paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks.
-Assertions play a crucial role in validating critical assumptions automatically, reducing the reliance on manual checks, and enhancing the robustness and maintainability of ML pipelines.
-By embedding assertions, developers can document the decisions made during model development, ensuring that these decisions are consistently applied and preserved throughout the ML lifecycle.
-This practice not only facilitates smoother transitions and handovers in collaborative environments but also improves the reproducibility of ML experiments by ensuring consistent validation checks.
-
-\paragraph{\textbf{For Researchers}} The use of assertions in Jupyter notebooks, though currently uncommon, is an emerging testing practice within the ML community.
-This study highlights the potential for assertions to enhance the reliability and robustness of ML models by automating the validation of critical assumptions.
-Researchers are encouraged to explore the development and evaluation of ML-specific testing methodologies that leverage assertions.
-This could involve creating new testing frameworks and tools tailored to the unique challenges of ML systems, thus contributing to the advancement of software engineering practices in the ML domain.
-
-\paragraph{\textbf{For Notebook Tool Developers}} There is a clear need for automated tools that suggest assertions based on the outputs of code cells in Jupyter notebooks.
-Tool developers can play a pivotal role by creating ML-specific testing tools that seamlessly integrate with computational notebooks.
-These tools should provide automated suggestions for assertions, helping practitioners to translate visual insights into formal, executable checks.
-By bridging the gap between qualitative evaluation and analytical assertions, such tools can significantly enhance the reliability and robustness of ML models, ensuring that they meet predefined performance standards consistently.
-
-\paragraph{\textbf{For Educators}} Educators should emphasize the importance of testing ML code written in Jupyter notebooks as part of the curriculum.
-By incorporating testing practices and the use of assertions into ML education, students can develop a deeper understanding of the importance of automated validation in ensuring model reliability and robustness.
-Teaching these best practices early on can help cultivate a culture of rigorous testing among future ML practitioners, preparing them to address the specific challenges faced in ML-enabled software systems.
-
-\paragraph{\textbf{For General Notebook Users}} Awareness of software engineering best practices, including the use of assertions, is crucial for all users of Jupyter notebooks.
-Assertions serve as an executable form of documentation, capturing the critical assumptions and decisions made during the development process.
-By adopting assertions, users can ensure that their notebooks are more robust, maintainable, and reproducible.
-This practice helps in maintaining consistency and reliability, especially in collaborative projects where multiple users might work on the same notebook, ensuring that the assumptions and validations are transparent and consistently applied.
-
 \section{Future Work}
+
+% NOTE: taxonomy of assertions mapping to V&V tasks in ML
+% NOTE: automated tool that adds assertions once the contents of the NB is finalised?
 

### 2024-06-10 bin/create-master-outputs.py
feat: sync with ronaldo

@@ -1,0 +1,67 @@
+import pandas as pd
+import ast
+
+def get_ast(source: str) -> ast.Module:
+    try:
+        tree = ast.parse(source)
+    except:
+        tree = None
+    finally:
+        return tree
+
+
+def get_print_nodes(tree: ast.Module) -> list:
+    collector = PrintCollector()
+    collector.visit(tree)
+    return collector.nodes
+
+
+class PrintCollector(ast.NodeVisitor):
+    def __init__(self):
+        self.nodes = []
+
+    def visit_Call(self, node: ast.Call) -> None:
+        if (isinstance(node.func, ast.Name) and node.func.id == "print"):
+            self.nodes.append(node)
+
+if __name__ == "__main__":
+    print_frames = []
+    text_frames = []
+    for file in ["data/shome2023notebook/github-outputs.csv", "data/shome2023notebook/kaggle-outputs.csv"]:
+        with pd.read_csv(
+            file,
+            header=None,
+            names=["notebook", "source", "output_type", "text", "image", "has_html"],
+            usecols=lambda x: x in ["notebook", "source", "output_type", "text"],
+            engine="c",
+            iterator=True,
+        ) as reader:
+            chunk = reader.get_chunk(100000)
+            print(f"file: {file}, chunk: {chunk.shape}")
+            chunk.loc[:, "ast"] = chunk["source"].apply(get_ast)
+            chunk = chunk.loc[chunk["ast"].notna()]
+            chunk = chunk.loc[
+                chunk["ast"].map(lambda x: True if [node for node in ast.iter_child_nodes(x)] else False)
+            ]
+
+            prints = chunk.loc[chunk["output_type"] == "stream"]
+            prints.loc[:, "print"] = prints.loc[:, "ast"].apply(get_print_nodes)
+            prints = prints.explode("print")
+            prints = prints.loc[prints["print"].notna()]
+            prints.loc[:, "print"] = prints.loc[:, "print"].apply(lambda x: ast.unparse(x))
+            print_frames.append(prints)
+
+            texts = chunk.loc[chunk["output_type"] != "stream"]
+            texts.loc[:, "last_node"] = texts.loc[:, "ast"].apply(lambda x: x.body[-1])
+            texts.loc[:, "last_node"] = texts.loc[:, "last_node"].apply(lambda x: ast.unparse(x))
+            text_frames.append(texts)
+
+
+    prints = pd.concat(print_frames)
[…]

### 2024-06-10 bin/data-collection-asserts.py

@@ -11,9 +11,12 @@ import os
 
 def ipynb_to_dataframe() -> pd.DataFrame:
 
-    with open(args.notebook) as f:
-        cells = json.load(f)["cells"]
-        df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
+    try:
+        with open(args.notebook) as f:
+            cells = json.load(f)["cells"]
+            df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
+    except:
+        df = pd.DataFrame()
 
     # NOTE: early exit if empty notebook or no code cells
     (df.empty or df.loc[df["cell_type"] == "code"].empty) and exit()
@@ -63,31 +66,33 @@ if __name__ == "__main__":
     assert_cells.empty and exit()
 
     # populate assert_content & assert_context
-    assert_content = []
-    assert_context = []
+    asserts = []
     for idx, cell in assert_cells.iterrows():
         cell["notebook"] = args.notebook
-        assert_content.append(cell)
-        # populate assert_context
+
         try:
             above = all_cells.loc[idx - 1]
         except KeyError:
             pass
         else:
-            above["notebook"] = args.notebook
-            above["location"] = "above"
-            above["assert_cell_index"] = idx
-            assert_context.append(above)
+            # above["notebook"] = args.notebook
+            # above["location"] = "above"
+            # above["assert_cell_index"] = idx
+            # assert_context.append(above)
+            cell["above"] = above["source"]
 
         try:
             below = all_cells.loc[idx + 1]
         except KeyError:
             pass
         else:
-            below["notebook"] = args.notebook
-            below["location"] = "below"
-            below["assert_cell_index"] = idx
-            assert_context.append(below)
+            # below["notebook"] = args.notebook
+            # below["location"] = "below"
+            # below["assert_cell_index"] = idx
+            # assert_context.append(below)
+            cell["below"] = below["source"]
+
+        asserts.append(cell)
 
     # NOTE: make sure that the term `ipynb` only occurs at the end of filename
     # I did this using the following command and checking that it *does not* print anything:
@@ -108,10 +113,10 @@ if __name__ == "__main__":
     pd.DataFrame(data=[stats]).to_csv(name + "-stats.csv", index=False, header=False)
     print(f"OUTPUT:{name}" + "-stats.csv")
 
[…]

### 2024-06-10 bin/data-collection-outputs.py

@@ -7,9 +7,13 @@ import os
 
 
 def ipynb_to_dataframe() -> pd.DataFrame:
-    with open(args.notebook) as f:
-        cells = json.load(f)["cells"]
-        df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
+
+    try:
+        with open(args.notebook) as f:
+            cells = json.load(f)["cells"]
+            df = pd.read_json(StringIO(json.dumps(cells)), orient="records")
+    except:
+        df = pd.DataFrame()
 
     (df.empty or df.loc[df["cell_type"] == "code"].empty) and exit()
 
@@ -40,7 +44,7 @@ def get_outputs():
                 "source": cell.source,
                 "output_type": output["output_type"],
                 "text": None,
-                "image": None,
+                "has_image": None,
                 "has_html": None,
             }
 
@@ -49,7 +53,7 @@ def get_outputs():
 
             if output["output_type"] in ["display_data", "execute_result"]:
                 try:
-                    frame["image"] = output["data"]["image/png"]
+                    frame["has_image"] = len(output["data"]["image/png"]) != 0
                 except KeyError:
                     pass
 
@@ -68,7 +72,7 @@ def get_outputs():
     if rows:
         _indices = [v for row in rows for k, v in row.items() if k == "index"]
         rows = pd.DataFrame(data=rows, index=_indices).drop(columns="index")
-        rows = rows.loc[rows["image"].notna() | rows["text"].notna()]
+        rows = rows.loc[rows["text"].notna()]
         rows.loc[:, "text"] = rows.loc[:, "text"].apply(lambda x: "".join(x))
     else:
         rows = pd.DataFrame()

### 2024-06-10 bin/filter.py

@@ -1,214 +1,0 @@
-"""Check for assert statements in juptyer notebooks.
-
-Without any flags, the script will check for assert statements only in code cells that produce
-a visualisation. The search is not greedy by default. It will print the notebook to stdin if at
-least 1 assert statement exists in the notebook.
-
-With the `--proximity` flag, the script will also check in code cells directly below the
-visualisation cells.
-
-With the `--strict` flag, the script will only flag notebooks if the assert statements is in the
-code cell below the visualisation cell. This flag takes precedence over `--proximity`.
-"""
-
-import json
-import argparse
-import pandas as pd
-import numpy as np
-from io import StringIO
-
-pd.set_option("display.max_columns", 10)
-
-
-def _list_to_str(x):
-    return "".join(x)
-
-
-def check(df: pd.DataFrame) -> bool:
-    decision = (
-        not df.empty and df["source"].apply(_list_to_str).str.contains("assert").any()
-    )
-    return decision
-
-
-def get_cells(
-    code_cells: pd.DataFrame,
-    viz_cells: pd.DataFrame,
-    strict: bool = False,
-) -> pd.DataFrame:
-    cells = []
-    for idx, cell in viz_cells.iterrows():
-        if not strict:
-            # add cell with visualisation
-            cells.append(cell)
-
-        chunk = code_cells.loc[code_cells.index > idx]
-
-        try:
-            below = chunk.iloc[0]
-        except IndexError:
-            # chunk is empty or there were no code cells below the cell with visualistion
-            # don't do anything
-            pass
-        else:
-            cells.append(below)
-
-    return pd.DataFrame(cells)
-
-
-def filter(df: pd.DataFrame) -> pd.DataFrame:
-    filtered = df.copy()
-    if args.no_shape:
-        filtered = filtered.loc[
-            (filtered.source.str.contains("assert"))
-            & ~(filtered.source.str.contains("shape"))
-        ]
-
-    if args.max_num_lines:
-        filtered["source_num_lines"] = filtered.source.apply(lambda x: len(x))
-        filtered = filtered.loc[(filtered.source_num_lines <= args.max_num_lines)]
-
-    return filtered
-
-
-def print_match() -> None:
-    print(f"{args.notebook}")
-
-
-def ipynb_to_dataframe() -> pd.DataFrame:
-    frames = []
-
-    with open(args.notebook) as f:
-        cells = json.load(f)["cells"]
[…]

### 2024-06-10 bin/validate.py

@@ -1,16 +1,0 @@
-"""Check if the given Jupyter notebooks contains valid JSON."""
-
-import argparse
-import nbformat
-
-if __name__ == "__main__":
-    parser = argparse.ArgumentParser()
-    parser.add_argument("notebook", help="Path to Jupyter Notebook.")
-    args = parser.parse_args()
-
-    try:
-        nb_dict = nbformat.read(args.notebook, nbformat.NO_CONVERT)
-    except Exception as err:
-        print(f"ERROR\t{args.notebook}\t{err}")
-    else:
-        print(f"INFO\t{args.notebook}\tis valid")

### 2024-06-10 report/emse24.tex
feat(report): emse24 updates

@@
 
-\subsection{Jupyter Notebooks}
-
-% NOTE use background section from pimentel2019large-scale
+\subsection{Literate Programming, Computational Notebooks and Project Jupyter}
+
+% NOTE pimentel2019large-scale:
+% NOTE kery2018story
+% NOTE rule2018exploration
 
 % describe the structure of jupyter notebooks (anatomy: cells, outputs); no need to go into too much technical detail regarding the internal JSON structure, just mention that the data is represented in a standardised structure internally
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\linewidth]{nb.pdf}
+  \label{fig:nb}
+  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
+\end{figure}
+
+Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans.
+In his seminal work, Knuth adovocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}.
+Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualizations, and explanatory text within a single document.
+This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
+Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing.
+Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.
+
+% front-end which parses the JSON
+% there is also the kernel which is a process running in the background, the frontend communicates with the kernel; send input (code cell) and get output (result of execution)
+
+Figure~\ref{fig:nb} shows an Jupyter notebook with markdown and code cells.
+The data comprised within Jupyter Notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically.
+A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''.
+All cells of a notebook are stored as a list under the top-level \lstinline[language={}]$cells$ key in the JSON representation.
+Within this list, each cell is represented as a JSON object.
+The contents of the cell are stored as a list of strings inside the \lstinline[language={}]$source$ key.
 
@@
 % explain the anatomy of assert statements and the more complicated way in which asts for print statements are represented in an AST
+
+An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code.
+Each node in the tree represents a construct occurring in the source code.
+ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}.
+The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST.
+This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimization, and transformation.
+The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.
+
+In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class.
+The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails.
+\lstinline{print} statements are treated as a call to the built-in \lstinline{print} function.
+Function calls are represented in the AST by the \lstinline{Call} class.
+The \lstinline{Call} node includes \lstinline{func}, representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.
 
@@
   \centering
-\includegraphics[width=\linewidth]{data-collection.pdf}
+  \includegraphics[width=0.75\linewidth]{data-collection.pdf}
   \label{fig:data-collection}
@@
 \section{Threats to Validity}\label{sec:threats}
-TODO
+
+The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
+To the best of our knowledge,~\emph{Pimentel et al} and~\emph{Quaranta et al} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
+We faced several technical challenges while trying to access the notebooks provided by these replication packages and thus decided to mine our own data from Github.
+However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS.
+We wish to extend our analysis to notebooks collected from Kaggle since it is predominantly targeted towards the ML and DS community.
+
+In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate file type and the presence of the ``assert'' keyword.
+Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python testing libraries (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
+Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in python testing libraries.
 
@@
 
-\subsection{Feedback in Software Engineering and Machine Learning}
-
-% the role of early adoption of feedback when developing software, agile
-% similar concept exists in ML: crisp-dm, mldl, sculley, amershi, sambasivan
-% feedback from one part of the pipeline feeds into another,
-
 \subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
-
-% start with a very quick review of the workflow: REPL style of development; break code into smaller chunks; faster feedack; iterative style of development
-% link to the existing work on notebooks from chi: design, benefits and challenges
-% then go into the SE side of software written in notebooks: code quality
-% adoption of notebooks by ML community
-
-% start with a quick introduction to the anatomy of a computational notebook: code is broken down into smaller chunks represented as cells, cells can contain code (which are executable and immediately display the result of the execution), or markdown cells used for the purpose of documenting or creating a narrative for collaborative work.
-
-% the move on to the benefits of this style of coding: the ML pipeline contains several components, which can be broken down into smaller, more manageable chunks.
-These can be individually developed incorporating the rapid feedback provided from the REPL style of coding.
-
-% then summarise the pain-points of working in notebooks: use chattopadhyay2020whats
-Computational notebooks have been ubiquitously adopted by the ML community for developing ML enabled systems.
-Although originally intended to promote reproducible software, computational notebooks are far from being reproducible due to lack of software engineering best practices such as separation of concern, testing, dependenvcy management and versioning~\cite{pimentel2019large-scale,wang2020better,chattopadhyay2020whats}.
 
@@
 
-Testing ML enabled software systems poses more challenges compared to
-their traditional counter part. While traditional software systems
-mature through change in their codebase, ML enabled systems
-additionally experience change in the training data and the machine
-learning
-model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
-With ML being adopted into safety-critical domains that can affect
-human lives, ensuring that ML systems are correct, robust to data
-perturbations and not biased by race, colour or gender is of paramount
-importance. Existing scientific literature on ML testing broadly
-focuses on two aspects. First, on functional properties such as
-correctness and robustness of the model towards unseen data. And
-second, on non-functional properties such as fairness,
-interpretability and
-privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
-
-To test the correctness of ML models, several improvements over
-existing test adequacy metrics have been proposed. Tools such as
-\textit{DeepExplore} and \textit{DeepGauge} propose new test adequacy
-metrics such as neuron coverage adapted for ML enabled software
-systems~\cite{pei2017deepexplore, ma2018deepgauge,
-  gerasimou2020importance}. Formal verification methods have also been
-proposed that try to provide formal guarantee of robustness against
-adversarial examples. Such methods however are only feasible for
-statistical ML models and become computationally expensive for more
-complex models such as deep neural networks~\cite{zhu2021deepmemory,
-  baluta2021scalable}. Several works have been conducted on generating
-and detecting adversarial inputs for ML models. Data augmentation
-techniques based on fuzzing, search based software testing and
-mutation testing have been proposed to generate adversarial examples
-that can be used during model training to improve its
-robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot,
-  zhang2020white}. It is however not possible to include all
-variations of adversarial examples into the training data. Thus,
-methods have been proposed to detect adversarial inputs during
-runtime~\cite{xiao2021self, wang2020dissector, wang2019adversarial,
-  berend2020cats}.
-
-In contrast to existing scientific contributions, we conduct a
-large-scale empirical analysis of Jupyter Notebooks to understand the
-process of testing ML systems. Specifically, we focus on how
-visualisations are used to test specific properties of ML systems. We
-further investigate how frequently analytical tests are formulated
-based on the information gained from the visualisations.
+% Testing ML enabled software systems poses more challenges compared to their traditional counter part.
+While traditional software systems mature through change in their codebase, ML enabled systems additionally experience change in the training data and the machine learning model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
+With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
+
+Existing scientific literature on ML testing broadly focuses on two aspects.
+First, on functional properties such as correctness and robustness of the model towards unseen data.
+And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
+
+To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
+Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore, ma2018deepgauge, gerasimou2020importance-driven}.
+Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
+Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
+Several works have been conducted on generating and detecting adversarial inputs for ML models.
+Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white-box}.
+It is however not possible to include all variations of adversarial examples into the training data.
+Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking, wang2020dissector, wang2019adversarial, berend2020cats}.
+
+In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems.
+Specifically, we focus on how visualisations are used to test specific properties of ML systems.
+We further investigate how frequently analytical tests are formulated based on the information gained from the visualisations.
 
@@
 
-As ML augment software systems in safety-critical domains, emphasis
-has been put into explainability of ML models. ML models and the
-underlying data is complex and multi-dimensional. To combat the
-``curse of dimensionality'', visual analytics has been widely adopted
-by the ML community to understand the data and the internal workings
-of ML models~\cite{yuan2021survey,hohman2019visual,wexler2020if}.
-
-Prior studies have been conducted to understand how visual analytics
-techniques are currently being used in ML. \emph{Yuan et al} conduct a
-systematic review of 259 papers and propose a taxonomy of visual
-analytics techniques for ML. \emph{Hohman et al} conduct a survey of
-visual analytics techniques for Deep Learning Models. The findings of
-the study indicate that visual analytics has been widely adopted in ML
-for model explanation, interpretation, debugging and improvement.
-
-Several tools have been proposed by the visual analytics research
-community to aid practitioners in understanding how their ML models
-operate. \emph{Wexler et al} propose \textit{What-If}, a visual
-analytics tool to explore alternative hypotheses, generate
-counterfactual reasoning, investigate decision boundaries of the model
-and how change in data affects the model predictions. To reduce the
-cognitive load of ML practitioners during model building phase,
-\emph{Amershi et al} propose \textit{ModelTracker}. Given a trained
-model and a sample dataset, ModelTracker presents all the information
-in traditional summary statistics along with the performance of the
-model.
-
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and
-\textit{Drava} are a few other tools that have been proposed to handle
-specialised use-cases such as identifying systematic errors in ML
-models, generating counterfactual explanations for Generalised
-Additive Models, prioritising machine translation model improvements
-and relating human concepts with semantic dimensions extracted by ML
-models during disentangled representation learning\cite{ahn2023escape,
-  wang2023gam, robertson2023angler, wang2023drava}.
-
-In contrast to prior work which propose dedicated visual analytics
-tools, this study focuses on visualisations within computational
-notebooks. We focus on visualisations produced by AI practitioners
-using Python libraries (such as Matplotlib or Seaborn). Additional
-context provided by surrounding code and markdown cells are used to
-understand the motivation for creating the visualisation and the
-information gained from it.
-
-\section{Future Work}
+As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
+ML models and the underlying data is complex and multi-dimensional.
+To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2019what-if}.
+
+Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
+\emph{Yuan et al} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\emph{Hohman et al} conduct a survey of visual analytics techniques for Deep Learning Models.
+The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
+
+Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
+\emph{Wexler et al} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
+To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al} propose \textit{ModelTracker}.
+Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
+
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+
+In contrast to prior work which propose dedicated visual analytics tools, this study focuses on visualisations within computational notebooks.
+We focus on visualisations produced by AI practitioners using Python libraries (such as Matplotlib or Seaborn).
+Additional context provided by surrounding code and markdown cells are used to understand the motivation for creating the visualisation and the information gained from it.
+
+\section{Conclusion and Future Work}
 
@@
 
-\section{Conclusion}
+This study empirically supports the efficacy of visualisations for testing various properties of machine learning systems.
+Visualisations enhance the ability of ML practitioners to interpret complex information through visual cues.
+However, decisions based on such visualisations demand manual verification in each development iteration.
+Analytical assertions derived from ML visualisations can reduce this manual effort by encapsulating key insights, and ensuring continuity in interpretation across different development stages and project collaborators.
+
+Our results indicate that crafting analytical assertions from ML visualisations, is an emerging testing technique.
+Majority of the assertions however, are found to be primitive and unable to convey all information discerned from their visual counter-parts.
+This study finds a gap in the current software testing methods to address the unique challenges of ML.
+And emphasises the need for automated tools to bridge visual assessments and analytical assertions.
+
+There is a clear need for tools designed to aid ML practitioners when testing ML systems.
+These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems.
+Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.
+
+The manual translation of insights derived from visualisations into analytical assertions presents a substantial challenge.
+Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
+Generating assertions from visual information automatically, will promote adoption of assertions in ML notebooks.
+And bring the code developed within notebooks, closer to a production-ready codebase.
+
+% TODO remove this is no space left
+Extending ML testing to dynamic real-world scenarios should be explored further.
+This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
+Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycle.
 

### 2024-06-18 report/emse24.tex
feat(report): emse24 updates

@@
   \todo{update the numbers here}
-Testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
+Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
 In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
@@
 
-Machine Learning Development Lifecycle (MLDL) encompasses multifaceted series of entangled stages.
+The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages.
 Significant effort is invested in gathering and assembling the required data from diverse sources.
-The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages.
+The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages.
 MLDL is dominated by the model development phase which is highly experimental and iterative.
-This stage involves Exploratory Data Analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results.
+This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results.
 Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance.
@@
 
-The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations.
-Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring.
-Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.
-
-Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
-Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks.
-Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development.
+The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}.
+Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modeling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}.
+Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
+
+Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
+Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks.
+Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development.
 This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development.
@@
 Feedback is crucial for making important decisions throughout the MLDL.
-Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, create new features to boost model performance, and select specific hyperparameters.
+Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters.
 However, these decisions are often based on the conditions present when the model was initially created.
@@
 
-To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks.
-Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
-Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells.
+To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies.
+First, we examine the outputs of code cells in Jupyter notebooks.
+Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
+Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells.
 In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.
-
-% TODO: contributions?
-% TODO: key observations?
-
-% Visualisations are used throughout the ml development lifecycle to test various properties of the ml system.
-in the early stages of the ml development lifecycle, visualisations are extensively used to make sense of the data and verify its statistical properties.
-during the model development phase, visualisations are used to summarise metrics, contrast different learning algorithms and iteratively fine-tune ml models.
-once the model is deployed in production, visualisations are used to continually monitor their performance and trigger a new training cycle once their performance drops below a certain threshold~\cite{yuan2021survey,hohman2019visual,amershi2015modeltracker,wexler2020if}.
-
-% our observations indicate that formulating analytical assertions from visualisations is an emerging testing technique.
-however, the assertions found in this study often do not reflect the information obtained from their visual counter-part.
-this study finds the need for automated tools that can reduce the manual effort necessary to validate visualisations in ml systems.
-and help ml practitioners to formulate better analytical assertions from visualisations.
+To the best of our knowledge, we are the first to investigate how feedback derived from the output of code cells is used by ML practitioners when developing ML pipelines inside of Jupyter notebooks.
+We are also the first to investigate assertions written inside of notebooks and their intersection with the outputs of code cells.
+
+The research questions along with the contributions of this study are presented below.
+
+\begin{description}
+  \item[RQ1.] \textbf{What are the characteristics of feebback mechanisms in ML Jupyter notebooks?}
+
+  % TODO: the problem is that readers don't yet know that outputs are generated from last statement and print statements
+  We mine 300K Jupyter notebooks written in Python from Github and Kaggle.
+We analyse 2.3M code cells and create a dataset of 22.5K assertions, 25K print statements and 28K last statements.
+We further perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.
+
+  % TODO: The dataset is available publicly on ...
+
+  \item[RQ2.] \textbf{How is implicit feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
+
+  We conduct individual case-studies with XXX code cell outputs.
+During each case-study, we analyse the raw output along with the corresponding code that produced the output to understand its purpose.
+Additionally, we anayse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.
+
+  \item[RQ3.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
+
+  Similarly, we conduct case-studies with XXX assert statements. We analyse the assertion along with the surrounding code to understand its purpose.
+
+  \item[RQ4.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
+
+  Finally, we conduct case-studies with XXX visualisation and assertion (VA) pairs.
+\end{description}
+
+% TODO revisit this paragraph
+Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL.
+While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms.
+Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines.
+However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community.
+The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.
+
+% TODO update replication package
 
@@
 \section{Background}\label{sec:background}
+
+This section introduces relevant prior concepts for this study.
 
@@
 
-% describe the structure of jupyter notebooks (anatomy: cells, outputs); no need to go into too much technical detail regarding the internal JSON structure, just mention that the data is represented in a standardised structure internally
-
 \begin{figure}
   \centering
-  \includegraphics[width=0.75\linewidth]{nb.pdf}
+  \includegraphics[width=\linewidth]{nb.pdf}
   \label{fig:nb}
@@
 Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans.
-In his seminal work, Knuth adovocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}.
-Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualizations, and explanatory text within a single document.
+In his seminal work, Knuth advocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}.
+Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualisations, and explanatory text within a single document.
 This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
@@
 
-% front-end which parses the JSON
-% there is also the kernel which is a process running in the background, the frontend communicates with the kernel; send input (code cell) and get output (result of execution)
-
-Figure~\ref{fig:nb} shows an Jupyter notebook with markdown and code cells.
-The data comprised within Jupyter Notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically.
+Figure~\ref{fig:nb} shows an example Jupyter notebook with markdown and code cells.
+The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically.
 A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''.
@@
 
+\subsection{Outputs in Jupyter Notebooks}
+
+Only code cells in Jupyter notebooks can generate outputs.
+As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualisation.
+The outputs produced by code cells are stored as JSON objects under the \lstinline[language={}]$outputs$ key in the corresponding JSON representation of the notebook.
+Text outputs (such as those produced by the last statement of a cell or print statements) are stored under the \lstinline[language={}]$text$ key.
+Visual outputs have a sightly different JSON schema and are stored under the \lstinline[language={}]$data$ key.
+The image itself is represented as a base64 encoded string which is stored inside the \lstinline[language={}]$data$ object, under the \lstinline[language={}]$image/png$ key.
+
 \subsection{Python Abstract Syntax Tree}
@@
 The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST.
-This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimization, and transformation.
+This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimisation, and transformation.
 The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.
@@
 The \lstinline{Assert} node includes the \lstinline{test} attribute---which contains the condition being asserted---and the \lstinline{msg} attribute---which holds an optional message displayed if the assertion fails.
-\lstinline{print} statements are treated as a call to the built-in \lstinline{print} function.
+Print statements are treated as a call to the built-in \lstinline{print} function.
 Function calls are represented in the AST by the \lstinline{Call} class.
-The \lstinline{Call} node includes \lstinline{func}, representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.
-
-\section{Study Design}
-
-\subsection{Motivating Example}
-
-Sara is a data scientist who is developing a machine learning model to predict customer churn.
-Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells.
-Her workflow is dynamic and exploratory, typical of modern machine learning practices.
-
-Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution.
-As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features.
-This insight, derived directly from the output of the notebook cell, prompts her to apply normalization to ensure better model performance.
-
-With data preprocessing complete, Sara proceeds to build her initial model.
-After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy and loss after the first epoch.
-The results are disappointing, prompting her to adjust the learning rate and batch size directly in subsequent cells.
-Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings.
-
-Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions.
-At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.
-
-Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively.
-Each decision in Sara's workflow—from preprocessing adjustments to hyperparameter tuning and model validation—is guided by the iterative feedback she receives from the outputs of her notebook's code cells.
-
-% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster!
-(notebook keeps running!)
-Once all experimentation and model tuning are complete, and Sara is ready to deploy the model into a production environment, she adds assertions as a final step.
-In the dynamic, interactive environment of Jupyter notebooks, where code cells can be executed in any order, there is a risk of arriving at inconsistent states that may affect the model's performance unpredictably.
-Assertions act as a safeguard against these inconsistencies by verifying that the data fed into the model adheres to expected formats and distributions and that the model's performance metrics meet predetermined standards.
-For instance, assertions are used to ensure that no data columns have null values, that the range of input data features remains within expected bounds, and that model accuracy thresholds are consistently met regardless of the execution order of the notebook cells.
-This practice not only reinforces the stability and reliability of the deployed model but also provides clear, executable documentation of the critical decisions made throughout the model's development.
-When new team members are onboarded, these assertions allow them to quickly grasp the underlying assumptions and operational parameters of the model, ensuring ongoing robustness and performance integrity.
-This is crucial in maintaining the quality and reliability of the model as it transitions from a development setting to a production environment.
-
-\subsection{Exploratory Analysis}
-
-We undertook an exploratory analysis of the collected data to formulate our research questions.
-
-\subsection{Research Questions}
-
-\begin{description}
-  \item[RQ1.] \textbf{How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
-  \item[RQ2.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
-  \item[RQ3.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
-\end{description}
+The \lstinline{Call} node includes the \lstinline{func} attribute representing the \lstinline{print} function, and \lstinline{args} which lists the arguments passed.
+
+\section{Methodology}
 
@@
 % The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells.
-These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualizations.
+These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualisations.
 Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code.
@@
 
-We check the validity of the underlying JSON structure in each notebook using the nbformat tool ~\footnote{https://nbformat.readthedocs.io/en/latest/}.
+We check the validity of the underlying JSON structure in each notebook using the nbformat tool\footnote{https://nbformat.readthedocs.io/en/latest/}.
 Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
@@
 
-\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
+\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}
+
+Of the total 298K notebooks collected in this study, 25K (8.5\%) contained at least one assertion.
+From the 8.5\% notebooks, we collected 89.6K assertions.
+4.8K assertions (5\%) used methods from external testing libraries.
+Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library.
+We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{other-test-methods.pdf}
+  \caption{10 most common methods used in assertions written using external testing libraries.}
+  \label{fig:other-test-methods}
+\end{figure}
+
+Figure~\ref{fig:other-test-methods} presents the 10 most common methods used in the 5\% assertions written using external testing libraries.
+We manually tracked the source of the methods to the numpy, pandas, torch and unittest libraries.
+The most frequently used methods include \lstinline{assert_almost_equal} and \lstinline{assert_equal}, each with over 250 instances.
+Methods such as \lstinline{assert_almost_equal}, \lstinline{assert_array_almost_equal} and \lstinline{AssertAlmostEqual} accommodate the stochastic nature of ML, where results can vary slightly due to inherent randomness in data processing and model training.
+We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas data frames.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-assert-test.pdf}
+    \caption{10 most common tests used in \lstinline{assert} statements.}
+    \label{fig:common-assert-test}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-compare-op.pdf}
+    \caption{10 most common operators used in \lstinline{assert} statements that perform a comparison check.}
+    \label{fig:common-compare-op}
+  \end{minipage}
+\end{figure}
+
+% TODO: some form of background on AST representation of assert statements is missing
+
+\begin{table}
+  \centering
+  \caption{Examples of the most common tests performed in assert statements.}
+  \begin{tabular}{@{}m{0.3\textwidth} m{0.3\textwidth}@{}}
+    \toprule
+    \emph{\textbf{Comparison Check}}&
+    \emph{\textbf{Function Call}}\\
+    \midrule
+
+    \lstinline[]$type(train_dataset[0]) == tuple$&
+    \lstinline[]$(np.unique(y_test) == [-1, 1]).all()$\\
+
+    \lstinline[]$torch.__version__ > '1.10.0'$&
+    \lstinline[]$isinstance(transmat, pd.DataFrame)$\\
+
+    \lstinline[]$len(x_dev) == len(y_dev)$&
+    \lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:common-assert-test}
+\end{table}
+
+We focus the remainder of the analysis on the 95\% of the assertions written using the \lstinline{assert} statement.
+The most common test performed in the \lstinline{assert} statements include a comparison check and a function call as shown in Figure~\ref{fig:common-assert-test}.
+Table~\ref{tab:common-assert-test} provides a few examples of assertions for each of the tests above.
+In Figure~\ref{fig:common-compare-op} we see the most common operators used in assert statements that perform a comparison check.
+We observe that overwhelming majority of the asserts ensure that two values are equal to each other.
+The shape and size of data structures are frequently compared as observed in Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} which show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-compare-lhs.pdf}
+    \caption{Most common bi-grams that appear in the LHS of comparison statements.}
+    \label{fig:common-compare-lhs}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-compare-rhs.pdf}
+    \caption{Most common bi-grams that appear in the RHS of comparison statements.}
+    \label{fig:common-compare-rhs}
+  \end{minipage}
+\end{figure}
+
+Of the 95\% assert statements, 28.4K (31.7\%) have a failure message.
+Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages.
+Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9K unique messages.
+The bi-grams indicate that the shape and data type of data structures are frequently verified.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{common-assert-msgs.pdf}
+  \caption{Most common bi-grams in the failure messages of assert statements.}
+  \label{fig:common-assert-msgs}
+\end{figure}
+
+\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
 
@@
 Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
-For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
-During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
+For instance, visualisations can efficiently determine if scaling, normalizing, or handling outliers is needed.
+During the exploratory data analysis (EDA) phase, visualisations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
 Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.
 
-\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
+\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
 These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}
@@
 In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
-Output \emph{O10} depicts the visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
+Output \emph{O10} depicts the visualisation of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
 
@@
 However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
-Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\ref{shome2022data}.
+Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.
 
 \highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
-Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
+visualisations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
 These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
@@
 Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
-This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
+This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
 
@@
 While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
-Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{CITEME}.
+Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}.
 
@@
 \emph{P4} prints the number of rows or examples in the test set.
-\emph{P32} and \emph{117} prints the number of rows and columns (or features) in the training and test data respectively.
+\emph{P32} and \emph{P117} prints the number of rows and columns (or features) in the training and test data respectively.
 
@@
 
-\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
+\subsection{(RQ3) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
 
@@
 
-Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualizations, and model training status.
+Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualisations, and model training status.
 These checks help prevent runtime errors and ensure the proper execution of code.
@@
 
-Assertion \emph{A60} is used to verify the presence of data in visualizations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
-These checks are crucial for ensuring that the visualizations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
+Assertion \emph{A60} is used to verify the presence of data in visualisations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
+These checks are crucial for ensuring that the visualisations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
 
@@
 
-Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
+Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations.
 
@@
 
-\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
+\subsection{(RQ4) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
 
@@
 
-The visualizations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
+The visualisations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
 VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
 These metrics collectively indicate the model's learning progress and stability during training.
-The visualization in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
+The visualisation in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
 The general upward trend indicates learning and improvement in gameplay over time.
@@
 
-The visualization in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
-This visualization aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
-The visualizations in VA28 display the training loss and development recall, respectively.
+The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
+This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
+The visualisations in VA28 display the training loss and development recall, respectively.
 The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
-The visualizations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
+The visualisations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
 The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
@@
 
-The corresponding assertions for these visualizations ensure that the models meet specific performance standards.
+The corresponding assertions for these visualisations ensure that the models meet specific performance standards.
 For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
@@
 \highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
-The visualizations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
-Assertions associated with these visualizations ensure models meet a specific performance criteria.}
+The visualisations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
+Assertions associated with these visualisations ensure models meet a specific performance criteria.}
 
@@
 
-The visualization in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
+The visualisation in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
 
@@
 
-\highlight{\textbf{Finding 3.1} One visualization is used to display the distribution of bootstrapped sample means.
+\highlight{\textbf{Finding 3.1} One visualisation is used to display the distribution of bootstrapped sample means.
 The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
@@
 
-The visualization in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
+The visualisation in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
 The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
@@
 
-The visualization in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
+The visualisation in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
 The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
@@
 
-The visualization in VA57 shows a histogram of transformed data after it has been normalised.
+The visualisation in VA57 shows a histogram of transformed data after it has been normalised.
 The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
 
-The visualization in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
+The visualisation in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
 This heatmap helps in visualizing the differences and similarities between the original and augmented data.
@@
 Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself.
-Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.
+Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated and explicit validation mechanisms.
 
@@
 
-From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions.
+From the 3 million Jupyter notebooks collected in this study, only 30 thousand (9.7\%) were found to utilize assertions.
 This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community.
@@
 We argue that the need for translating visual insights into formal assertions is more pressing.
-The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners.
-Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
+The visual cue and trends observed in visualisations are open to interpretation, thus the interpretation of the same plot may differ amongst two practitioners.
+Translating visual insights into assertions ensures that the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
 
@@
 
+\subsection{Assertions and Software Bugs}
+ TODO
+
 \subsection{Computational Notebooks and Software Engineering}\label{sec:notebooks}
@@
 
-Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large,quaranta2021kgtorrent}.
+Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large-scale,quaranta2021kgtorrent}.
 Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.

### 2024-06-28 report/draft.tex
feat(report): emse24 updates

\subsection{Motivating Example}

Sara is a data scientist who is developing an ML model to predict customer churn. Sara begins her project in a Jupyter notebook, which supports iterative development and immediate feedback through the execution of code cells. Her workflow is dynamic and exploratory, typical of modern ML practices. Sara loads the dataset and performs Exploratory Data Analysis (EDA) to develop a high-level understanding of the data she is working with. Since most ML models expect the data to have a normal distribution, Sara generates histograms to validate this assumption. Based on the histograms, Sara confirms that the data has a normal distribution and therefore does not apply any normalisation techniques. With data preprocessing complete, Sara proceeds to build her initial model. After setting up a logistic regression classifier, she runs a training cell that outputs the model's accuracy. The results are disappointing, prompting her to adjust the hyper-parameters of the model in subsequent cells. Each modification is immediately followed by feedback on its impact, allowing Sara to quickly converge on the optimal settings. After all experimentation and model tuning are complete Sara deploys the model into a production environment where the model is periodically retrained using new training data.

% Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution. As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features. This insight, derived directly from the output of the notebook cell, prompts her to apply normalisation to ensure better model performance.

A few weeks later, Sara notices a drop in the performance of her model. To investigate, Sara revisits the notebook in which she developed the model. However all the cells in the notebook execute without any errors thus Sara needs to execute each cell individually and manually analyse the outputs which takes considerable amount of time and effort. Sara soon realises that the distribution of the new slice of training data is no longer normal. To fix the issue, Sara applies normalisation techniques. Additionally, she also adds an assertion that performs a K-test

% sara returns to the notebook since the performance of the model has suddenly dropped. However all cells in the notebook execute without any errors so sara has to execute each cell and analyse the output of the cells manually which takes a long time. Sara finally identifies the source of the problem, the training data is missing 3 features that contributed to the performance of the model. However, sara was unable to identify this since the training data is automatically pulled from a database and another team in sara's organisation changed the db schema without notifying other users

% to ensure that such issues a caught faster in the future, sara adds an assertion that fails if the number of features in the training data changes. This assertion also documents

A few weeks later, Bob is tasked with improving the performance of the model since Sara is on vacation. Bob obtains the notebook in which Sara originally developed the model however is unable to

% Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions. At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.

% Finally, Sara fine-tunes her model based on the outputs from a confusion matrix and ROC curve, identifying thresholds for classification that balance precision and recall effectively. Each decision in Sara's workflow—--from preprocessing adjustments to hyperparameter tuning and model validation—--is guided by the iterative feedback she receives from the outputs of her notebook's code cells.

% TODO: the jump to assertions is very drastic; motivate it using scenario that there is a bug in the system (use the data validation papers from google) and Sara needs to fix them and how assertions would have helped here to debug the issue faster! (notebook keeps running!)

### 2024-06-28 bin/data-collection.py

@@ -1,0 +1,168 @@
+import json
+import argparse
+import pandas as pd
+from io import StringIO
+import os
+import ast
+
+
+class PrintCollector(ast.NodeVisitor):
+    def __init__(self):
+        self.nodes = []
+
+    def visit_Call(self, node: ast.Call) -> None:
+        if isinstance(node.func, ast.Name) and node.func.id == "print":
+            self.nodes.append(node)
+
+
+class AssertionNodeCollector(ast.NodeVisitor):
+    def __init__(self):
+        self.nodes = []
+
+    def visit_Assert(self, node: ast.Assert) -> None:
+        self.nodes.append(node)
+
+    def visit_Call(self, node: ast.Call) -> None:
+        self.nodes.append(node)
+
+
+class FunctionNameCollector(ast.NodeVisitor):
+    def __init__(self):
+        self.names = []
+
+    def visit_Name(self, node: ast.Name) -> None:
+        self.names.append(node.id)
+
+    def visit_Attribute(self, node: ast.Attribute) -> None:
+        # capture function name if in module.func format
+        self.names.append(node.attr)
+
+
+def get_ast(source: str) -> ast.Module | None:
+    try:
+        tree = ast.parse(source)
+    except:
+        tree = None
+    finally:
+        return tree
+
+
+def get_assertion_nodes(tree: ast.Module) -> list:
+    collector = AssertionNodeCollector()
+    collector.visit(tree)
+    return collector.nodes
+
+
+def get_print_nodes(tree: ast.Module) -> list:
+    collector = PrintCollector()
+    collector.visit(tree)
+    return collector.nodes
+
+
+def is_assertion_node(node):
+    if isinstance(node, ast.Call):
+        collector = FunctionNameCollector()
+        collector.visit(node.func)
+        return "assert" in "".join(collector.names)
+
+    if isinstance(node, ast.Assert):
+        return True
+
+
+parser = argparse.ArgumentParser(
+    description="Compute various statistics about Python assert statements in Jupyter Notebooks"
+)
+parser.add_argument(
+    "notebook",
+    help="Path to Jupyter Notebook",
+)
+args = parser.parse_args()
+print(f"INPUT:{args.notebook}")
+
+try:
+    with open(args.notebook) as f:
[…]

### 2024-06-28 report/emse24.tex
feat(report): emse24 updates from Luis & Diomidis

--- 2024-06-10 report/emse24.tex
@@
 
-The MLDL is fundamentally iterative, closely resembling Agile software engineering principles which emphasize incorporating early feedback and working in small, manageable iterations.
-Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring.
-Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects~\cite{betz2018managing,martinez-plumed2021crisp-dm}.
-
-Computational notebooks---specifically Jupyter notebooks~\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
-Jupyter notebooks provide an interactive computing environment that allow users to break complex code into smaller, more manageable chunks.
-Each notebook is composed of a series of code cells, which can be executed independently in a Read-Eval-Print Loop (REPL) style of development.
+The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}.
+Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, through data preparation, modeling, and evaluation, to deployment and monitoring~\cite{martinez-plumed2021crisp-dm}.
+Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
+
+Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
+Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks.
+Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development.
 This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development.
@@
 
-To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct an empirical study of the outputs of code cells in Jupyter notebooks.
-Next, we conduct an empirical study of Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
-Finally, we investigate the intersection of outputs and assertions by conducting an empirical analysis of assertions that document decisions made from the outputs of code cells.
+To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies.
+First, we examine the outputs of code cells in Jupyter notebooks.
+Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
+Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells.
 In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.
@@
 
-Initially, Sara loads a dataset and uses basic descriptive statistics to understand its distribution.
+Initially, Sara loads a dataset and runs basic descriptive statistics to understand its distribution.
 As she executes a cell that calculates and displays the mean and standard deviation of each feature, she notices unusually high variance in one of the features.
@@
 
-Throughout this process, Sara heavily relies on the outputs of various code cells, not just for error correction but also for making strategic decisions about model architecture, such as adding layers or changing activation functions.
+Throughout this process, Sara heavily relies on the outputs of various code cells, not just for detecting and fixing erroneous assumptions and code, but also for making strategic decisions about the model's architecture, such as adding layers or changing activation functions.
 At one point, after noticing overfitting from a cell outputting validation loss, she decides to incorporate dropout layers, which she can quickly evaluate by rerunning the training process.

### 2024-06-28 report/emse24.tex
refactor(report): fix remaining merge conflicts

--- 2024-06-18 report/emse24.tex
@@
 
-In software engineering, tests are employed to document the expected behavior of code.
+In software engineering, tests are employed to document the expected behaviour of code.
 Failing tests indicate when these expectations are no longer met.
@@
 
-Of the total 298K notebooks collected in this study, 25K (8.5\%) contained at least one assertion.
+25K (8.5\%) notebooks contained at least one assertion.
 From the 8.5\% notebooks, we collected 89.6K assertions.
@@
 \end{figure}
+
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{common-print-nodes.pdf}
+    \caption{Most common AST nodes inside print statements}
+    \label{fig:common-print-nodes}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\textwidth]{common-print-constant-types.pdf}
+    \caption{Most common type of constant inside print statements}
+    \label{fig:common-print-constant-types}
+  \end{minipage}
+\end{figure}
+
+We collected 1.4M print statements from 180K notebooks (60\% of the total notebooks analysed in this study).
+Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls.
+Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python and output from function calls.
+Figure~\ref{fig:common-print-nodes} shows that the most common nodes appear inside of print statements are constants followed by function calls.
+Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string.
+Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes).
+Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively.
+Both figures collectively indicate that print statements are frequently used to print various performance metrics of ML models.
+
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-print-constants.pdf}
+    \caption{Most common bi-grams that appear in natural text inside of print statements.}
+    \label{fig:common-print-constants}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-print-not-constants.pdf}
+    \caption{Most common bi-grams that appear in the code of print statements.}
+    \label{fig:common-print-not-constants}
+  \end{minipage}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{common-last-nodes.pdf}
+  \caption{Most common nodes in the last statement of code cells.}
+  \label{fig:common-last-nodes}
+\end{figure}
+
+We collected 1M last statements from 119K notebooks.
+Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call.
+We further analyse the functions calls.
+Figure~\ref{fig:common-last-modules}
+
+% most common functions?
 
@@
 
-Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modeling.
+Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modelling.
 These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
@@
 Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
-\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behavior.
+\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behaviour.
 \emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations.
@@
 When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigor is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
+This rigour is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
 These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
@@
 % TODO: verify if this applies to recommender systems as well?
-For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbors as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
+For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbours as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
 Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
@@
 Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'.
-This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behavior or suboptimal model performance.
+This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behaviour or suboptimal model performance.
 
@@
 
-Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behavior.
+Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behaviour.
 By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.
@@
 
-The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background color gradient indicating the decision function of the SVM.
+The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background colour gradient indicating the decision function of the SVM.
 This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.

### 2024-07-01 report/emse24.tex
feat(report): emse24 updates

@@
 
-  \todo{update the numbers here}
+  % TODO: update the abstract
   Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
@@
 
+\end{abstract}
+
 \keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}
-\end{abstract}
 
 \section{Introduction}
+
+% TODO: feedback from Diomidis
+% - In the introduction clarify (and perhaps illustrate) the life cycle and the role of notebooks.
+Are these used to develop code that is then integrated into scripts, or is the notebook the execution vehicle, or is it informal throw-away code and rough notes on how to repeat a process in the future?
+This will also inform the discussion and the implications.
+Consider describing
+%  a) the current development model, with the role of assertions, implicit checks, and visualization,
+%  b) its limitations (which provide motivation for the study),
+%  c) a sketch for an improved development model, and
+%  d) the knowledge gaps associated with the improved model and how this study fills them.
+%
+% - The part starting "However, these decisions are often based on the conditions present when the model was initially created", which provides the motivation must be argued more convincingly.
+I think that the motivation is not so much what is described in that paragraph, as the fact that, in the absence of detailed, up-to-date, and accurate requirements, rationale, and design documentation as well as tests, many of these can be derived from notebook interactions.
 
@@
 
+% TODO: feedback from Diomidis
+% - In Section 2, rather than jumping to literate programming, it would be more appropriate to discuss the CLI interaction style, which also encompasses Jupyter.
+% - Section 2.2 (Python AST) can be shortened, shortened and moved to methods, or removed. Consider also describing the format of Jupyter notebooks.
+
 \subsection{Literate Programming, Computational Notebooks and Project Jupyter}
@@
 
+% TODO: feedback from Diomidis
+% - Sections 4.1 and 4.2 are very interesting but overly detailed.
+I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.
+In addition, consider moving the more general observations into a discussion section.
+% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
+% - I am afraid I cannot understand RQ3; consider revising it.
+I would think that visualizations are verified visually, rather than through assertions.
+You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
 % TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output.
@@
 
-25K (8.5\%) notebooks contained at least one assertion.
-From the 8.5\% notebooks, we collected 89.6K assertions.
-4.8K assertions (5\%) used methods from external testing libraries.
-Majority of the assertions however (85K or 95\%) were written using the \lstinline{assert} statement provided by the built-in Python standard library.
-We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1K assertions.
+25 thousand (8.5\%) notebooks contained at least one assertion.
+From the 8.5\% notebooks, we collected 89.6 thousand assertions.
+4.8 thousand assertions (5\%) used methods from external testing libraries.
+85 thousand assertions (95\%) however were written using the \lstinline{assert} statement provided by the built-in Python standard library.
+We removed duplicate assertions for the remainder of the analysis, thus resulting in 27.1 thousand assertions.
 
@@
   \caption{Examples of the most common tests performed in assert statements.}
-  \begin{tabular}{@{}m{0.3\textwidth} m{0.3\textwidth}@{}}
+  \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
     \toprule
@@
 
-Of the 95\% assert statements, 28.4K (31.7\%) have a failure message.
+Of the 95\% assert statements, 28.4 thousand (31.7\%) have a failure message.
 Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages.
-Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9K unique messages.
+Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9 thousand unique messages.
 The bi-grams indicate that the shape and data type of data structures are frequently verified.
@@
 
-We collected 1.4M print statements from 180K notebooks (60\% of the total notebooks analysed in this study).
+We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study).
 Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls.
-Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python and output from function calls.
-Figure~\ref{fig:common-print-nodes} shows that the most common nodes appear inside of print statements are constants followed by function calls.
+Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls.
+Figure~\ref{fig:common-print-nodes} shows that the most common nodes inside of print statements are constants followed by function calls.
 Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string.
@@
 
+We collected 1 million last statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study).
+Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call.
+We further analyse the functions calls in more detail.
+Figure~\ref{fig:common-last-modules} and Figure~\ref{fig:common-last-functions} show the most modules and functions used inside the function calls respectively.
+The figures highlight the prominence of data visualisation libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualisation functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks.
+The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently.
+The presence of \lstinline{train} and \lstinline{model} modules along with the \lstinline{fit} function indicates the use of ML libraries.
+
 \begin{figure}
@@
 
-We collected 1M last statements from 119K notebooks.
-Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call.
-We further analyse the functions calls.
-Figure~\ref{fig:common-last-modules}
-
-% most common functions?
+% TODO: we need better terminology here
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-last-modules.pdf}
+    \caption{Most common modules used in the function calls of last statements.}
+    \label{fig:common-last-modules}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-last-functions.pdf}
+    \caption{Most common functions used in the function calls of last statements.}
+    \label{fig:common-last-functions}
+  \end{minipage}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-last-df-functions.pdf}
+    \caption{Most common functions called on pandas data frames.}
+    \label{fig:common-last-df-functions}
+  \end{minipage}
+  \hfill
+  \begin{minipage}{0.49\textwidth}
+    \centering
+    \includegraphics[width=\linewidth]{common-last-visualisation-functions.pdf}
+    \caption{Most common visualisation functions used in last statements.}
+    \label{fig:common-last-visualisation-functions}
+  \end{minipage}
+\end{figure}
+
+We further analysed the most common functions called on pandas data frames by isolating the last statements where the module is \lstinline{data} or \lstinline{df}.
+This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations.
+The frequent use of functions such as \lstinline{value_counts}, \lstinline{head}, and \lstinline{unique}, indicate an emphasis on understanding the structure and content of the data.
+Visualisation functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualisations directly from data frames.
+Descriptive statistics are also prominent, with functions like \lstinline{describe}, \lstinline{mean}, and \lstinline{sum} being commonly used.
+
+Similarly, we also analysed the most common visualisation functions used in notebooks by isolating the last statements where the module is \lstinline{sns} or \lstinline{plt}.
+This is shown in Figure~\ref{fig:common-last-visualisation-functions} which indicates that categorical data visualisation appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions.
+The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualising correlations and statistical distributions.
+For continuous data, functions like \lstinline{distplot}, \lstinline{scatterplot}, and \lstinline{scatter} are commonly used.
+The presence of \lstinline{imshow} suggests that image-related visualisations are also significant in many notebooks.
 
@@
 
+% TODO: feedback from Diomidis
+% - "it represents an emerging testing practice within the ML community": I have commented before that assertions are different from tests.
+Assertions are a mechanism to increase the code's robustness, similar to type checks, array bounds checks, and thrown API exceptions.
+
 From the 3 million Jupyter notebooks collected in this study, only 30 thousand (9.7\%) were found to utilize assertions.
@@
 
+% TODO: feedback from Diomidis
+% - The implications section focuses implicitly on assertions, at the expense of the other elements discussed earlier.
+Also, it puts excessive emphasis on testing, which IIUC is not the paper's topic.
+
 \paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks.
@@
 \section{Related Work}\label{sec:related}
+
+% TODO: feedback from Diomidis
+% - In Section 8 consider adding material regarind interactive ML development.
 

### 2024-07-02 report/emse24.tex
chore(report): emse24 migrate feedback from Luis

@@
 
+% TODO: feedback from Luis
+% - not flowing with previous paragraph
+% - The previous paragraph could mention something like ad-hoc manual tests.
+And then in this paragraph we could say that software engineering proposes practices on automated testing that can be useful in this context
+% - in theory, the main goal of tests is to verify the requirements of the software.
+Documentation is a by product.
+(I like the idea in this paragraph but we dont want to touch a raw nerve from reviewers)
 In software engineering, tests are employed to document the expected behaviour of code.
@@
 
-% NOTE pimentel2019large-scale:
+% NOTE pimentel2019large-scale
 % NOTE kery2018story
@@
   \includegraphics[width=\linewidth]{nb.pdf}
+  \caption{Example Jupyter notebook with code and markdown cells, adapted ~\cite[Figure~1]{pimentel2019large-scale}.}
   \label{fig:nb}
-  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\cite[Figure~1]{pimentel2019large-scale}.}
 \end{figure}
@@
   \includegraphics[width=0.75\linewidth]{data-collection.pdf}
+  \caption{Overview of data collection methodology used in this study}
   \label{fig:data-collection}
-  \caption{Overview of data collection methodology used in this study}
 \end{figure}
@@
 
-We check the validity of the underlying JSON structure in each notebook using the nbformat tool\footnote{https://nbformat.readthedocs.io/en/latest/}.
+% TODO: feedback from Luis
+% - explain why we check the validity
+% - we cannot guarantee that the white literature will be online in the next months/years and make it difficult to collect an important detail for reproducibility.
+I would add to the text at least how many libraries we added, give a few examples and point to our replication package,
+We check the validity of the underlying JSON structure in each notebook using the nbformat tool~\footnote{https://nbformat.readthedocs.io/en/latest/}.
 Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from white literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from grey literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
 
@@
 
+% TODO: feedback from Luis
+% - only here, a reader start understanding that we want to collect assert, prints, and lasts. Perhaps we want to make that clear in the beginning of this subsection or in the one above it
 Similarly, for implicit feedback, we first isolate code cells that produce an output.
@@
 % TODO: might need a visual aid here
-
+% TODO: feedback from Luis
+% - motivate this approach
+% - not sure we want to say NLP here. I would call it text processing. NLP kinda implies that we are automatically interpreting the text
 We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs).
@@
 P68) because the statement may be inside an if statement.
-
 \subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}
@@
 \subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
+% TODO: feedback from Luis
+% -  up to this point I miss how exactly we plan to answer this question.
+Reading the results I get that we are answering by clustering outputs together based on their respective ML/DS task and then by analysing how that task is accomplished.
+We need to explain this either in the beginning of section 4.1 or in section 3
 
@@
 
+% TODO: feedback from Luis
+% - I miss a bit the link between the RQ and the different subsections 4.1.1
+% -  I also miss an explanation of the table.
+what is O2?
+We should have a sentence explaining that each of the examples in the table are identified by a key,  the respective code and visual output, if any.
+% - Explain also how the figure is extracted. This is important because it is the first one. For the next ones we can just say that we follow the same structure.
+% - Another alternative would be to explain this generally in the intro of section 4.1 and then follow it consistently.
+% - I also miss how the examples are extracted or what they mean here.
+% - I mean, are these all the examples we found for distribution check? If so, make it clear.
+% - add: "with key" to be closer to the table structure
 Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data.
@@
 
-\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48).
-These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}
+% TODO: Feedback from Luis
+% - I think you should mention the fact that there are both prints and visualisations for this task, the fact that the approach can be different not only depending on the type of data,  but also depending on the preferences of the developers (maybe we cannot really claim this last part but at least mention that there are multiple visualisations for the same scenario and developers simply need to opt for one.
+\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, kernel density estimate plots and count plots (\emph{O2}, \emph{O9} and \emph{O25}), and statistical summaries from pandas dataframes (\emph{O14} and \emph{O48}).
+These checks are used for making informed decisions about data transformations such as scaling, normalising and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
 
@@
 
+% TODO: feedback from Luis
+% - this explains what the code does but it does not pinpoint observations on how the output of code cells are used.
+E.g., here it is mostly visualisations - that's an observation.
+They are both scatterplots, one had to change the height of the plot (not sure that's relevant.
+We do not observe prints or asserts that help with this task.
 Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
@@
 \highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
-visualisations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features.
+visualisations such as a scatter plot with a regression line (\emph{O10}) and a line plot (\emph{O6}) are employed to assess the linearity between features.
 These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
@@
 % NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
+% TODO: feedback from luis
+% - I probably missed it but it should be clear why we have P and O in the key, given that they are all outputs.
 Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed.
@@
 Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
-This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
+This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
 
@@
 % (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
-
+% TODO: feedback from Luis
+% - if needed, we can merge Excution time check here. (both execution time and model correctness can be presented as a KPI)
 \begin{table}
@@
 
+% TODO: feedback from luis
+% - this cluster makes me wonder: should we make a point that some of these outputs may happen by chance and were not explicitly defined by users/developers?
 \begin{table}
@@
 Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise, kery2018story, liu2019understanding, chattopadhyay2020whats}.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
 Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
@@
 To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
-Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle, bavishi2021vizsmith}.
+Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle,bavishi2021vizsmith}.
 
@@
 To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
-Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore, ma2018deepgauge, gerasimou2020importance-driven}.
+Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore,ma2018deepgauge,gerasimou2020importance-driven}.
 Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
-Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory, baluta2021scalable}.
+Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory,baluta2021scalable}.
 Several works have been conducted on generating and detecting adversarial inputs for ML models.
-Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution, gao2020fuzz, wang2021robot, zhang2020white-box}.
+Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution,gao2020fuzz,wang2021robot,zhang2020white-box}.
 It is however not possible to include all variations of adversarial examples into the training data.
-Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking, wang2020dissector, wang2019adversarial, berend2020cats}.
+Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking,wang2020dissector,wang2019adversarial,berend2020cats}.
 
@@
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape, wang2023gam, robertson2023angler, wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
 

### 2024-07-15 report/draft.tex
feat(report): emse24 updates

@@
 (notebook keeps running!)
+
+\subsection{(RQ4) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
+
+\subsubsection{Model Performance ($N = 15$)}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used to validate the performance of a model.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used to validate the performance of a model (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA2 &
+  \includegraphics[width=\linewidth]{va2.png} &
+  \lstinline[]$assert final_score > 300, 'not good enough for DQN'$\\
+
+  VA3 &
+  \includegraphics[width=\linewidth]{va3.png} &
+  \lstinline[]$assert accuracy_score(y_test, pred) > 0.95$\\
+
+  VA9 &
+  \includegraphics[width=\linewidth]{va9.png} &
+  \lstinline[]$assert np.mean(mean_rw_history[-10:]) > 10.$\\
+
+  VA11 &
+  \includegraphics[width=\linewidth]{va11.png} &
+  \lstinline[]$assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."$\\
+
+  VA15 &
+  \includegraphics[width=\linewidth]{va15.png} &
+  \lstinline[]$assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."$\\
+
+  VA28 &
+  \includegraphics[width=\linewidth]{va28.png} &
+  \lstinline[]$assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."$\\
+
+  VA68 &
+  \includegraphics[width=\linewidth]{va68.png} &
+  \lstinline[]$np.testing.assert_almost_equal(ans4, 0.3042764)$\\
+
+  VA107 &
+  \includegraphics[width=\linewidth]{va107.png} &
+  \lstinline[]$assert r2_gru > 0.6$\\
+
+\end{longtable}\label{tab:va-model-perf}
+
+The visualisations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
+VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
+These metrics collectively indicate the model's learning progress and stability during training.
+The visualisation in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
+The general upward trend indicates learning and improvement in gameplay over time.
+VA15 presents the training and development loss over time for a character-level language model, showing a gradual decrease in loss and suggesting effective learning.
+
+The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background colour gradient indicating the decision function of the SVM.
+This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
+The visualisations in VA28 display the training loss and development recall, respectively.
+The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
+The visualisations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
+The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
+In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
+
+The corresponding assertions for these visualisations ensure that the models meet specific performance standards.
+For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
+In VA9, the assertion checks whether the mean of the last ten rewards exceeds 10, ensuring stable and satisfactory learning progress.
+For VA15, the assertion checks if the mean of the first ten training observations is greater than the mean of the last ten, ensuring loss convergence and validating the model's learning without overfitting.
+The assertion in VA3 ensures the accuracy score on the test set exceeds 0.95, confirming the model's reliability and generalization capability.
+For VA28, the assertion checks whether the average development recall in the last ten evaluations exceeds 0.85, ensuring high recall rate critical for tasks like medical diagnostics.
+In VA11, the assertion ensures the average development BLEU score in the last ten evaluations exceeds 35, confirming the model produces high-quality translations.
+Lastly, in VA107, the assertion checks whether the R2 score exceeds 0.6, validating the model's fit and reliability for making accurate predictions in real-world scenarios.
+These validations confirm that iterative training processes lead to meaningful improvements and ensure the models' readiness for deployment and practical applications.
+
+\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
+The visualisations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
+Assertions associated with these visualisations ensure models meet a specific performance criteria.}
+
+\subsubsection{Distribution Check ($N = 1$)}
+
+\begin{table}
+  \centering
+  \caption{VA pairs used to check data distribution.}
+  \begin{tabular}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+    \toprule
+    \emph{\textbf{Key}}&
+    \emph{\textbf{Visual Output}}&
+    \emph{\textbf{Assertion}}\\
+    \midrule
+
+    VA13 &
+    \includegraphics[width=\linewidth]{va13.png} &
+    \lstinline[]$assert 80 < boot_mean_mean < 81$\\
+    \bottomrule
+  \end{tabular}
+  \label{tab:va-dist-check}
+\end{table}
+
+The visualisation in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
+
+Bootstrapped sample means are calculated by repeatedly resampling the original dataset with replacement and computing the mean for each resampled dataset.
+This process generates a distribution of the mean, known as the bootstrap distribution, which helps in estimating the variability and central tendency of the original dataset's mean.
+
+The corresponding assertion checks whether the bootstrapped mean lies between 80 and 81, ensuring that the statistical estimation is consistent with expected values.
+This validation is crucial for confirming the reliability of the bootstrapped estimates and the accuracy of statistical conclusions.
+
+\highlight{\textbf{Finding 3.1} One visualisation is used to display the distribution of bootstrapped sample means.
+The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
+
+\subsubsection{Data Validation ($N = 9$)}
+
+\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
+  \caption{VA pairs used for data validation.} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endfirsthead
+
+  \caption[]{VA pairs used for data validation (continued).} \\
+  \toprule
+  \emph{\textbf{Key}} &
+  \emph{\textbf{Visual Output}} &
+  \emph{\textbf{Assertion}}\\
+  \midrule
+  \endhead
+
+  \midrule
+  \endfoot
+
+  \bottomrule
+  \endlastfoot
+
+  VA30 &
+  \includegraphics[width=\linewidth]{va30.png} &
+  \lstinline[]$assert scaled_data[scaled_data.argmax()] <= 1$\\
+
+  V35 &
+  \includegraphics[width=\linewidth]{va35.png} &
+  \lstinline[]$assert(all(df_lower['inspection_score'] > 75))$\\
+
+  V57 &
+  \includegraphics[width=\linewidth]{va57.png} &
+  \lstinline[]$assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2$\\
+
+  VA62 &
+  \includegraphics[width=\linewidth]{va62.png} &
+  \lstinline[]$assert not np.allclose(x_aug, x_original)$\\
+
+\end{longtable}\label{tab:va-data-validatio}
+
+The VA pairs in Table~\ref{tab:va-data-validatio} collectively illustrate various data validation techniques essential for ensuring the integrity and quality of data used in ML models.
+
+The visualisation in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
+The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
+The scaling process adjusts the data to fall within a specified range of 0 and 1.
+The corresponding assertion checks whether the maximum value in the scaled data is less than or equal to 1, ensuring that the scaling procedure was applied correctly.
+
+The visualisation in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
+The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
+This comparison helps identify any disparities in health inspection scores based on income levels.
+The corresponding assertion checks that all inspection scores in the lower-income areas are above 75, ensuring that the data reflects a minimum standard of health inspection scores in these areas.
+
+The visualisation in VA57 shows a histogram of transformed data after it has been normalised.
+The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
+
+The visualisation in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
+This heatmap helps in visualizing the differences and similarities between the original and augmented data.
+The corresponding assertion checks that the augmented data \lstinline{x_aug} is not nearly identical to the original data \lstinline{x_original}, ensuring that the data augmentation process has introduced sufficient variability.
+
+These validations are crucial for confirming that data preprocessing techniques such as scaling, transformation, and augmentation have been correctly applied.
+Ensuring the integrity and quality of the data is essential for the subsequent steps in the machine learning pipeline, as it directly impacts the robustness, reliability, and generalization of the models developed.
+By validating these steps, we can maintain high standards of data quality and model performance, which are fundamental for accurate and meaningful analyses.
+
+\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original.
+These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}

### 2024-07-15 report/emse24.tex
feat(report): emse24 updates

@@
   % numberstyle=\tiny\color{gray},
-  % backgroundcolor=\color{backcolor},
-  aboveskip=0pt,
-  belowskip=0pt
+  backgroundcolor=\color{backcolor},
+  % aboveskip=0pt,
+  % belowskip=0pt
 }
@@
 \section{Introduction}
-
-% TODO: feedback from Diomidis
-% - In the introduction clarify (and perhaps illustrate) the life cycle and the role of notebooks.
-Are these used to develop code that is then integrated into scripts, or is the notebook the execution vehicle, or is it informal throw-away code and rough notes on how to repeat a process in the future?
-This will also inform the discussion and the implications.
-Consider describing
-%  a) the current development model, with the role of assertions, implicit checks, and visualization,
-%  b) its limitations (which provide motivation for the study),
-%  c) a sketch for an improved development model, and
-%  d) the knowledge gaps associated with the improved model and how this study fills them.
-%
-% - The part starting "However, these decisions are often based on the conditions present when the model was initially created", which provides the motivation must be argued more convincingly.
-I think that the motivation is not so much what is described in that paragraph, as the fact that, in the absence of detailed, up-to-date, and accurate requirements, rationale, and design documentation as well as tests, many of these can be derived from notebook interactions.
-
 The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages.
@@
 MLDL is dominated by the model development phase which is highly experimental and iterative.
-This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results.
+This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results.
 Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance.
@@
 
-% The feedback-driven style of development in MLDL resembles principles of Agile software development and the Cross-Industry Standard Process for Data Mining (CRISP-DM).
-Agile methodologies emphasize continuous iteration of development and testing throughout the software lifecycle.
-In Agile, software is developed in incremental, rapid cycles resulting in small, incremental releases with each release building on previous functionality.
-Each iteration is reviewed and critiqued by the project team, which includes representatives of the project's various stakeholders.
-
 The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}.
-Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modeling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}.
+Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modelling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}.
 Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
@@
 
-Feedback is crucial for making important decisions throughout the MLDL.
-Practitioners rely on the output of cells to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters.
-However, these decisions are often based on the conditions present when the model was initially created.
-Visual outputs in particular, are open to interpretation~\cite{heer2010tour}.
-The viewer's background and expertise significantly influence how they understand and draw conclusions from visual data.
-For instance, interpreting a histogram requires knowledge of various distribution types, which someone without statistical training might not possess.
-Additionally, physical limitations such as color blindness can impede accurate interpretation if visualizations are not designed with accessibility in mind.
+% NOTE: consider starting with prints and last statements first (most obvious), then move to assertions; its fine if RQs follow different order
+Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favor of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}.
+During this process, practitioners can use various feedback mechanisms like assertions, print statements, output of code cells and visualisations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters~\cite{head2019managing,chattopadhyay2020whats}.
+Once the code is finalised, it is typically transferred into a script that is run in a production environment.
+Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration.
+To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}.
+In both scenarios, practitioners need to revisit the notebooks to obtain structured information such as up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
 As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
@@
 
-% TODO: feedback from Luis
-% - not flowing with previous paragraph
-% - The previous paragraph could mention something like ad-hoc manual tests.
-And then in this paragraph we could say that software engineering proposes practices on automated testing that can be useful in this context
-% - in theory, the main goal of tests is to verify the requirements of the software.
-Documentation is a by product.
-(I like the idea in this paragraph but we dont want to touch a raw nerve from reviewers)
-In software engineering, tests are employed to document the expected behaviour of code.
-Failing tests indicate when these expectations are no longer met.
-Similarly, assertions in Jupyter notebooks can serve as documentation for the decisions made during model development.
-Embedding assertions in the notebook can ensure that any deviations from expected data conditions or model performance are promptly detected, thus maintaining the integrity of the ML pipeline.
-
-To understand the role of feedback when developing ML pipelines in Jupyter notebooks, we conduct a series of  empirical studies.
-First, we examine the outputs of code cells in Jupyter notebooks.
-Next, we study Python assertions written in Jupyter notebooks to understand how they are used when developing ML pipelines in notebooks.
-Finally, we investigate the intersection of outputs and assertions by analysing  assertions that document decisions made from the outputs of code cells.
-In particular, we focus on assertions derived from visualisations since they pose a higher risk of technical debt.
-To the best of our knowledge, we are the first to investigate how feedback derived from the output of code cells is used by ML practitioners when developing ML pipelines inside of Jupyter notebooks.
-We are also the first to investigate assertions written inside of notebooks and their intersection with the outputs of code cells.
-
+% TODO: feedback from diomidis
+% - "We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks." Rephrase to have the knowledge gap correspond to what the RQs answer.
+Currently one may think that we aim to improve the approach, not its understanding.
+We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks.
+To address this gap, this study conducts a series of empirical studies to understand the role of various feedback mechanisms used to develop ML pipelines within Jupyter notebooks.
+We focus on implicit form of feedback obtained from print statements and output generated by the last statement in code cells, as well as explicit form of feedback obtained from assertions.
+To the best of our knowledge, we are the first to investigate how such feedback mechanisms are used by ML practitioners when developing ML pipelines inside Jupyter notebooks.
 The research questions along with the contributions of this study are presented below.
@@
 \begin{description}
-  \item[RQ1.] \textbf{What are the characteristics of feebback mechanisms in ML Jupyter notebooks?}
-
-  % TODO: the problem is that readers don't yet know that outputs are generated from last statement and print statements
-  We mine 300K Jupyter notebooks written in Python from Github and Kaggle.
-We analyse 2.3M code cells and create a dataset of 22.5K assertions, 25K print statements and 28K last statements.
-We further perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.
-
-  % TODO: The dataset is available publicly on ...
-
-  \item[RQ2.] \textbf{How is implicit feedback from the output of code cells used when writing ML code in Jupyter notebooks?}
-
-  We conduct individual case-studies with XXX code cell outputs.
-During each case-study, we analyse the raw output along with the corresponding code that produced the output to understand its purpose.
-Additionally, we anayse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.
-
-  \item[RQ3.] \textbf{How are assertions used to validate ML code written in Jupyter notebooks?}
-
-  Similarly, we conduct case-studies with XXX assert statements. We analyse the assertion along with the surrounding code to understand its purpose.
-
-  \item[RQ4.] \textbf{How are assertions used to validate visual outputs in Jupyter notebooks?}
-
-  Finally, we conduct case-studies with XXX visualisation and assertion (VA) pairs.
+  \item[RQ1.] \textbf{What feedback mechanisms are employed in the development of ML Jupyter notebooks?}
+
+    We mine 297.8 thousand Jupyter notebooks written in Python from Github and Kaggle.
+We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.
+We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.
+
+  % TODO: add citation for the dataset (make it public so that DOI is available)
+
+  \item[RQ2.] \textbf{How is explicit feedback from assert statements used to validate ML code written in Jupyter notebooks?}
+
+    To gain a deeper understanding of assert statements written in Jupter notebooks, we conduct individual case-studies with 82 assert statements.
+We analyse the assertion along with the surrounding code to understand its purpose.
+Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.
+
+  \item[RQ3.] \textbf{How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}
+
+    Similarly, we conduct individual case-studies with 44 print statements and 27 last statements.
+During each case-study, we analyse the code statement along with the raw output to understand its purpose.
 \end{description}
@@
 % TODO revisit this paragraph
+% TODO what should be done to change the things we observe? to increase the use of explicit feedback? do we know if its important or not? maybe we need to explore that?
 Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL.
@@
 
-% TODO: feedback from Diomidis
-% - In Section 2, rather than jumping to literate programming, it would be more appropriate to discuss the CLI interaction style, which also encompasses Jupyter.
-% - Section 2.2 (Python AST) can be shortened, shortened and moved to methods, or removed. Consider also describing the format of Jupyter notebooks.
-
-\subsection{Literate Programming, Computational Notebooks and Project Jupyter}
-
-% NOTE pimentel2019large-scale
-% NOTE kery2018story
-% NOTE rule2018exploration
+\subsection{REPLs, Interactive Computing and Project Jupyter}
+
+A Read-Eval-Print Loop (REPL) is an interactive programming environment that takes single user inputs (Read), executes them (Eval), returns the result to the user (Print) and then waits for the next command (Loop).
+REPLs offer an immediate feedback loop for quickly testing and debugging code snippets.
+The Python standard distribution includes a Python REPL with a command line interface to execute Python code in real-time.
+IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history.
+Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI).
+Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualisations, and narrative text.
+This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
+Jupyter Notebooks in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.
+
+\subsection{Anatomy of Jupyter Notebooks}
 
@@
 \end{figure}
-
-Literate programming is a paradigm introduced by Donald Knuth that emphasizes writing code that is understandable by humans.
-In his seminal work, Knuth advocates combining source code with documentation in a single, cohesive narrative~\cite{knuth1984literate}.
-Computational notebooks are inspired by this paradigm, allowing for the integration of executable code, scientific equations, visualisations, and explanatory text within a single document.
-This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
-Project Jupyter is a prominent example of this approach, offering a suite of open-source tools for interactive computing.
-Jupyter Notebooks, in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.
 
@@
 
-\subsection{Outputs in Jupyter Notebooks}
-
 Only code cells in Jupyter notebooks can generate outputs.
@@
 
-% explain the high-level way in which ast module in python works
-% explain the anatomy of assert statements and the more complicated way in which asts for print statements are represented in an AST
-
-An Abstract Syntax Tree (AST) is a tree representation of the abstract syntactic structure of source code.
-Each node in the tree represents a construct occurring in the source code.
-ASTs are widely used in compilers and code analysis tools to understand and manipulate code programmatically~\cite{wikipedia2024abstract}.
-The Python \emph{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an AST.
-This module allows for the examination and modification of code structure, making it valuable for tasks such as code analysis, optimisation, and transformation.
-The ast module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation~\cite{pythonast}.
-
+The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)~\cite{pythonast}.
+The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation.
 In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class.
@@
 
-% TODO we need to explain why we only focus on the last statement of code cells (when visualisation code written anywhere produces an output); this is much harder to parse: we need to construct an exhaustive list of python visualisation libraries; people can use any naming convension to...
-
-% The predominant form of obtaining feedback, while developing ML pipelines in Jupyter notebooks, is through the output of code cells.
-These outputs can come from three main sources in the source code--print statements, the last statement in code cells, and statements that create visualisations.
-Building on the principles from software engineering, we also recognize assert statements as a valuable form of feedback for writing ML code.
-This idea draws from prior literature in software engineering that connects test-driven development (TDD) as a form of developing software using feedback.
-
 \begin{figure}
@@
 
-Figure~\ref{fig:data-collection} summarises the data collection procedure used in this study. We collected Jupyter notebooks written in Python, from Kaggle and Github.
-
-We mined public repositories from Github to collect $49923$ Jupyter notebooks.
-We use the Github advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code} to isolate Jupyter Notebook that contain the keyword ``assert'' in them.
-The final search query is as follows: \texttt{"assert" language:"Jupyter Notebook"}.\footnote{Collected on June 22, 2023}
-
-For Kaggle, we used a pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} since Kaggle does not support advanced code-based search like Github.
-To the best of our knowledge, KGTorrent this is the largest dataset of Python Jupyter notebooks obtained from Kaggle consisting of $248763$ notebooks.
-
-Therefore, we start with 283GB of data comprising of $298686$ Python Jupyter notebooks.
-
-% TODO: feedback from Luis
-% - explain why we check the validity
-% - we cannot guarantee that the white literature will be online in the next months/years and make it difficult to collect an important detail for reproducibility.
-I would add to the text at least how many libraries we added, give a few examples and point to our replication package,
-We check the validity of the underlying JSON structure in each notebook using the nbformat tool~\footnote{https://nbformat.readthedocs.io/en/latest/}.
-Since the focus of this study is on the analysis of Python code, we only include notebooks that contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import popular ML libraries~\footnote{These libraries are derived from grey literature---https://www.coursera.org/articles/python-machine-learning-library, https://www.geeksforgeeks.org/best-python-libraries-for-machine-learning, https://www.datacamp.com/blog/top-python-libraries-for-data-science, https://www.kdnuggets.com/2020/11/top-python-libraries-data-science-data-visualization-machine-learning.html, https://lp.jetbrains.com/python-developers-survey-2022}.
-
-For explicit feedback using assertions, we programmatically analyse the JSON structure of the notebooks to isolate code cells.
-We collect Python assert statements from the code cells by constructing and subsequently parsing the Abstract Syntax Tree (AST) of the source code present in each cell.
-
-% TODO: feedback from Luis
-% - only here, a reader start understanding that we want to collect assert, prints, and lasts. Perhaps we want to make that clear in the beginning of this subsection or in the one above it
-Similarly, for implicit feedback, we first isolate code cells that produce an output.
-The output of a cell may either be from a ``print'' statement or the last statement of the cell.
-We collect both of them using the AST of the cell.
-
-Finally, we remove duplicate data points (asserts, prints and last statements) resulting in a final data set of $22548$ assertions, $25319$ print statements and $28522$ last statements.
+% TODO: I suspect this will raise eyebrows because the notebooks collected from Github were for the ICST paper, where we were also looking for the keyword "assert" (which does not apply to this paper)
+
+Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle.
+We used Github's advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on Github~\footnote{Collected on June 22, 2023}.
+We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub.
+To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks.
+Consequently, we started with 283GB of data comprising 297.8 thousand Python Jupyter notebooks.
+
+Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries which we identified from grey literature~\cite{datacampml,gfgml,jetbrainsml,kdnuggetml,courseraml}---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}.
+We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells.
+Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a final data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.
 
@@
 
-% TODO: might need a visual aid here
-% TODO: feedback from Luis
-% - motivate this approach
-% - not sure we want to say NLP here. I would call it text processing. NLP kinda implies that we are automatically interpreting the text
-We allocated a fixed time resource of 100 hours to conduct the case study analysis of all candidates (assertions and cell outputs).
-To surface interesting candidates for the case studies, we apply NLP techniques as described below.
-
-% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs.
-Thats why we need to apply these techniques to surface unique and interesting candidates!
+We allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last statements.
+To surface interesting candidates for the case studies, we apply text processing techniques as described below.
 
@@
 The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
-
 We apply stratified random sampling to identify the candidates for the case study analysis.
@@
 The analysis is stopped when the time resource is exhausted.
-
 During each case study, we analyse the code of the candidate to understand its purpose.
@@
 % TODO: feedback from Diomidis
-% - Sections 4.1 and 4.2 are very interesting but overly detailed.
-I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.
-In addition, consider moving the more general observations into a discussion section.
-% - Consider starting with RQ2 (making it RQ1), because assertions are a concrete feature of Python code.  Then move to RQ1, which deals with implicit assertions.
-% - I am afraid I cannot understand RQ3; consider revising it.
-I would think that visualizations are verified visually, rather than through assertions.
-You may have encountered some parts, but I would assume that the assertions mostly deal with things that can be more accurately and effectively be expressed as a predicate rather than viewed in a chart.
-% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output.
-Somethimes actual output is not possible (eg.
-P68) because the statement may be inside an if statement.
-\subsection{(RQ1) What are the characteristics of feedback mechanisms in ML Jupyter notebooks?}
+% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.
+% TODO: In addition, consider moving the more general observations into a discussion section.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{taxonomy.pdf}
+  \caption{Overview of various feedback mechanism groups identified in this study.}
+  \label{fig:taxonomy}
+\end{figure}
+
+% TODO: add citation to online appendix
+The results of the study are presented in this section.
+For the case-studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other.
+Please note that this is done only to facilitate structuring the paper in a meaningful manner and for easy of reporting the results.
+Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study.
+Additional details are provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}.
+Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples.
+These can be viewed on our online appendix.
+
+\subsection{(RQ1) What feedback mechanisms are employed in the development of ML Jupyter notebooks?}~\label{sec:result-analysis}
 
@@
 \end{figure}
-
-% TODO: some form of background on AST representation of assert statements is missing
 
@@
     \lstinline[]$torch.__version__ > '1.10.0'$&
-    \lstinline[]$isinstance(transmat, pd.DataFrame)$\\
+    \lstinline[]$isinstance(transmat, pd.dataframe)$\\
 
@@
 We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study).
-Unlike assert statements, print statements do not have a fixed lexical semantic since the Python ast module interprets them as regular function calls.
+Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls.
 Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls.
-Figure~\ref{fig:common-print-nodes} shows that the most common nodes inside of print statements are constants followed by function calls.
+Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls.
 Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string.
@@
 
-\subsection{(RQ2) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}
+\subsection{(RQ2) How is explicit feedback from assert statements used to validate ML code written in Jupyter notebooks?}~\label{sec:result-explicit}
+
+\subsubsection{Data shape check ($N = 26$)}
+% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
+
+Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the MLDL.
+We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features (\emph{A17, A90}), labels (\emph{A5, A29}), predictions, images (\emph{A76, A93}), sequences, and embeddings (\emph{A76, A93}).
+These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.
+
+Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
+By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
+
+\highlight{\textbf{Finding 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
+These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
+
+\subsubsection{Data validation check ($N = 14$)}
+% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
+
+We found assert statements that perform various validation checks on data structures, such as NumPy arrays and Pandas dataframes.
+These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
+
+Common data validation checks include verifying the presence of specific values or ranges within arrays or columns(\emph{A41, A65}) and verifying the equality or closeness of arrays to a target value within a specified absolute tolerance (\emph{A44, A46}).
+Some assertions also focus on validating the uniqueness or cardinality of values (\emph{A52}) while others validate the presence of specific values or conditions within data structures (\emph{A73}).
+
+By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
+
+\highlight{\textbf{Finding 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
+These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
+
+\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
+% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
+
+\begin{lstlisting}[caption={Assertion \emph{A58} used to check that the accuracy of a model is close to the specified value.
+The use of \lstinline{np.isclose} allows for small deviations in the accuracy thus accounting for the stocastic nature of ML models.}, label={lst:A58}]
+assert np.isclose(accuracy, 0.9666666666666667)
+\end{lstlisting}
+
+This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
+Range checks confirm that the performance metrics fall within acceptable boundaries, indicating good predictive performance without overfitting (\emph{A38}).
+Direct comparisons of model outputs with expected results, validate the model's predictive accuracy and reliability under operational conditions (\emph{A15}).
+Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}).
+Finally assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specifc scenarios like clustering (\emph{A7}).
+
+The use of \lstinline{np.isclose} in Listing~\ref{lst:A58} is particularly noteworthy.
+This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
+By allowing a small tolerance in the comparison, \lstinline{np.isclose} ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
+
+\highlight{\textbf{Finding 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
+These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
+
+\subsubsection{Existence check ($N = 8$)}
+% NOTE: this is a combination of missing-value-check and existence-check
+% (A23, A42, A43, A50, A51, A63, A79, A86)
+
+Existence checks are primarily used during the data preprocessing stage in the MLDL to ensure data integrity before analysis and modelling.
+These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns (\emph{A23, A42, A79, A86}).
+Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce \lstinline{NaN} values (\emph{A50, A51, A63}) or result in an empty dataframe (\emph{A43}).
+These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust ML applications.
+
+\highlight{\textbf{Finding 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
+These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
+
+\subsubsection{Resource check ($N = 7$)}~\label{sec:result-rq2-resource-check}
+% (A10, A14, A18, A37, A60, A67 A74)
+
+\begin{lstlisting}[caption={Assertion \emph{A37} used to ensure that an ML model has not reached an inconsistent state due to out-of-order or re-execution of code cells.}, label={lst:A37}]
+assert svm.fit_status_ == 0, 'Forgot to train the SVM!'
+\end{lstlisting}
+
+Resource check assertions are used to ensure the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations.
+One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
+Subsequently, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).
+
+A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
+Aassertion \emph{A37} presented in Listing~\ref{lst:A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
+This check prevents inconsistent model states from executing cells out of order or multiple times.
+Other interesting assertions unique to computational notebooks include verifying the presence of data in visualisations (\emph{A60}) and explicitly verifying the versions of critical libraries (\emph{A18, A67}).
+
+\highlight{\textbf{Finding 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
+These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
+
+\subsubsection{Type check ($N = 5$)}
+% (A2, A35, A40, A81, A88)
+
+Type check assertions are used to maintain data integrity, ensure compatibility between different components of the ML pipeline, and preventing errors and unexpected behaviour.
+Type checks are commonly performed to ensure that the input features are PyTorch float tensors or NumPy arrays, a requirement for many deep learning models and scientific computing operations respectively (\emph{A2, A88}).
+Type checks are also applied to very the type of the ML models (\emph{A40}), data structures and intermediate objects (\emph{A35}) and individual columns or elements within a dataset (\emph{A81}).
+
+\highlight{\textbf{Finding 2.6} Five assertions validate the data types of input features, models, and intermediate objects.
+These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
+
+\subsubsection{Mathematical property check ($N = 4$)}
+% (A3, A25, A56, A64)
+
+When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
+This rigour is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalisation (\emph{A25}).
+These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles.
+
+\highlight{\textbf{Finding 2.7} Four assertions validate key mathematical properties of data and model outputs.
+These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
+These checks confirm the robustness and reliability of model computations.}
+
+\subsubsection{Batch size check ($N = 3$)}
+% (A21, A28, A70)
+
+In the context of neural networks, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilisation.
+Batching stabilises the learning curve by updating the model weights based on the average gradient of a batch thereby optimising learning and enhancing generalisability of the model.
+Ensuring the batch size matches the hardware capacity is crucial to avoid memory overflow issues that can halt training.
+Assertions are used to check that the batch size divides evenly into the training dataset size which ensures that each training epoch is computationally efficient (\emph{A21}).
+Additionally, assertions are used to ensure that the size of the training data is larger than the batch size (\emph{A70}) and that image dimensions are suitably divisible by the patch size which is essential for certain convolutional networks (\emph{A28}).
+
+\highlight{\textbf{Finding 2.8} This study identifies three assertions validating batch size configurations.
+These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
+These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
+
+\subsubsection{Network architecture check ($N = 3$)}
+% (A11, A62, A75)
+
+When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
+This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
+Checks include the use of assertions to verify that the output and input dimensions of consecutive convolutional layers match (\emph{A11}), checking that shape of the activations have the right number of channels (\emph{A62}) and that the appropriate regularisation parameter is set (\emph{A75}).
+Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible with each other.
+
+\highlight{\textbf{Finding 2.9} Three assertions verify the consistency of neural network architecture.
+These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
+These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
+
+\subsubsection{Data leakage check ($N = 1$)}
+% (A33)
+
+\begin{lstlisting}[caption={Assertion \emph{A33} used to ensure that the training and validation sets do not contain any overlapping values.}, label={lst:A33}]
+assert len(
+  set(tr_df.PetID.unique()).intersection(valid_df.PetID.unique())
+) == 0
+\end{lstlisting}
+
+Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in ML to prevent overfitting.
+To guarantee that the model can generalise effectively to new examples, it is crucial to verify that the training and validation sets are completely distinct with no shared examples.
+By strictly separating these datasets, the validation phase provides an unbiased evaluation of the model's ability to generalise to data similar to, but not identical to that which it was trained on.
+This is demonstrated by Assertion \emph{A33} in Listing~\ref{lst:A33} ensures that there are no overlapping \lstinline{PetID} values between the training set and the validation set, thereby preventing any possibility of data leakage.
+
+\highlight{\textbf{Finding 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
+Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalise to new data.
+This practice maintains the integrity and validity of the model evaluation metrics.}
+
+\subsection{(RQ3) How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-implicit}
 % TODO: feedback from Luis
@@
 
+\subsubsection{Model performance check ($N = 33$)}
+% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
+% TODO: feedback from Luis
+% - if needed, we can merge Excution time check here. (both execution time and model correctness can be presented as a KPI)
+
+During the model development phase, the performance of an ML model is often printed to facilitate comparisons against other models or variations.
+Throughout this continuous experimentation phase, authors may adjust model parameters or modify the data by engineering new features.
+The model is then re-trained to check if these changes lead to performance improvements.
+Besides the \emph{accuracy} of the model (\emph{P3, P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
+In addition, we find the use of heatmaps (\emph{L3}) to visualise the confusion matrix of a multi-label classification task and custom functions (\emph{L52}) to manually evaluate an image classification model on a random sample of input images.
+
+\highlight{\textbf{Finding 3.1} The study identifies 33 outputs used to check the performance of trained ML models.
+These include outputs that print key performance metrics, a normalised heatmap used to visualise the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
+These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
+
 \subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
@@
 
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to verify distribution of data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    O2&
-    \lstinline[]$_ = sns.catplot(x='category_id', y='likes', data=train, height=5, aspect=1.5)$&
-    \includegraphics[width=\linewidth]{distribution-check-catplot.png}\\
-
-    O9&
-    \lstinline[]$sns.kdeplot(data=data.loc[ data['Survived'] == 0].Age, label='Died', shade=True)$&
-    \includegraphics[width=\linewidth]{distribution-check-kdeplot.png}\\
-
-    O14&
-    \lstinline[]$pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])$&
-    \\
-
-    O25&
-    \lstinline[]$sns.countplot(house_pred['OverallQual'])$&
-    \includegraphics[width=\linewidth]{o25.png}\\
-
-    O48&
-    \lstinline[]$x_train.describe()$&
-    \\
-
-    \bottomrule
-  \end{tabular}
-  \label{tab:distribution-check}
-\end{table}
-
-% TODO: feedback from Luis
-% - I miss a bit the link between the RQ and the different subsections 4.1.1
-% -  I also miss an explanation of the table.
-what is O2?
-We should have a sentence explaining that each of the examples in the table are identified by a key,  the respective code and visual output, if any.
-% - Explain also how the figure is extracted. This is important because it is the first one. For the next ones we can just say that we follow the same structure.
-% - Another alternative would be to explain this generally in the intro of section 4.1 and then follow it consistently.
-% - I also miss how the examples are extracted or what they mean here.
-% - I mean, are these all the examples we found for distribution check? If so, make it clear.
-% - add: "with key" to be closer to the table structure
-Table~\ref{tab:distribution-check} shows examples of outputs used to check the distribution of the data.
-In output \emph{O2} the distribution of feature \texttt{Age} is checked against the two labels (\texttt{Survived} and \texttt{Died}) of the target feature.
-Output \emph{O9} presents another example where the distribution of a continuous feature \texttt{likes} is checked with-respect-to all the values of another categorical feature \texttt{category\_id}.
-Output \emph{O14} produces a pandas dataframe which is used to check the mean value of the features \texttt{Age}, \texttt{SibSp}, \texttt{Parch} and \texttt{Fare} against each label in the target feature \texttt{Survived}.
-Output \emph{O48} produces a pandas dataframe with the descriptive statistics of all features in the training data.
-The statement is written immediately after normalising the features in the training data.
-Hence, the user manually verifies if the normalisation was successful by looking at the mean and standard deviation of the features in the output.
-Similarly, \emph{O25} is written after discretizing a continuous feature \texttt{OverallQual} using binning.
-
 Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
 For instance, visualisations can efficiently determine if scaling, normalizing, or handling outliers is needed.
-During the exploratory data analysis (EDA) phase, visualisations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training.
-Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.
-
-% TODO: Feedback from Luis
-% - I think you should mention the fact that there are both prints and visualisations for this task, the fact that the approach can be different not only depending on the type of data,  but also depending on the preferences of the developers (maybe we cannot really claim this last part but at least mention that there are multiple visualisations for the same scenario and developers simply need to opt for one.
-\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, kernel density estimate plots and count plots (\emph{O2}, \emph{O9} and \emph{O25}), and statistical summaries from pandas dataframes (\emph{O14} and \emph{O48}).
+During the exploratory data analysis phase, visualisations and pandas dataframes are commonly used to assess the distribution of specific columns (\emph{L9}), helping identify features related to the target variable (\emph{L2, L14})and informing feature inclusion in model training.
+Additionally, descriptive statistics are used post-transformation to manually verify the effectiveness of data transformation steps and ensuring that the data conforms to the expected format for further analysis (\emph{L48, L25}).
+
+\highlight{\textbf{Finding 3.2} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
 These checks are used for making informed decisions about data transformations such as scaling, normalising and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
+
+\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
+% (P71, P86, O64, O66, P68, P82, P107)
+
+Similar to the assertions identified in Section~\ref{sec:result-rq2-resource-check}, we also find print and last cell statements that validate the availability of essential resources on the system where the notebooks are being executed.
+This includes checks for the availability of compute resources such as a GPU or a TPU (\emph{P68, P82, P86}), verifying that a dataset or a pre-trained model has been successfully loaded into memory (\emph{P107, L66}) and ensuring that a certain version of an external library is present on the system (\emph{P71}).
+
+\highlight{\textbf{Finding 3.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
+These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions.
+These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
+
+\subsubsection{Spot check ($N = 5$)}
+% (O56, O60, P64, P67, P114)
+
+Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
+These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
+Common checks include verifying the number of features in the data matches expectations after applying feature selection or dimensionality reduction techniques (\emph{L60}), assessing the correctness of one-hot encoding (\emph{P114}) and ensuring that data retains its expected properties after manipulation (\emph{P64}).
+
+\highlight{\textbf{Finding 3.4} This study finds five outputs used to perform spot checks at various stages of the MLDL.
+Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
+
+\subsubsection{Model training check ($N = 4$)}
+% (O8, O31, O42, P77)
+
+During training it is common practice to monitor the progress periodically and use this feedback to adjustment training parameters or stop training early to prevent overfitting.
+This can be done by checking the best parameters found through tuning methods (\emph{L42}), fitting the model to the training data (\emph{L31}) or printing the training loss and accuracy to give real-time feedback on the learning efficacy of the model (\emph{L8}).
+
+\highlight{\textbf{Finding 3.5} The study identifies four outputs used to monitor model training progress.
+These outputs provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods.
+These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
+
+\subsubsection{Missing value check ($N = 3$)}
+% (O12, O36, P74)
+
+Checking for missing values is a critical step in the preprocessing phase of an ML project because it significantly influences the quality and performance of the model.
+Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
+Many ML algorithms demand complete numerical datasets to perform calculations such as matrix multiplication.
+Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
+These checks can be performed in code (\emph{P74, L36}) and using visualisations such as a heatmap (\emph{L12}).
+
+\highlight{\textbf{Finding 3.6} The study identifies three outputs used to check for missing data.
+Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations.
+Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
+
+\subsubsection{Shape check ($N = 3$)}
+% (P4, P32, P117)
+
+Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
+Primarily, the number of features in the training set must match those in the testing set (\emph{P4, P32, P117}).
+This alignment is crucial because statistical ML models are trained on specific data dimensions and expect the same dimensional structure during inference to perform accurately.
+Similarly, in neural network architectures, the configuration of input layers depends directly on the shape of the training data, dictating the number of input neurons needed.
+Furthermore, the correspondence between the number of test examples and their respective labels is essential for accurately computing performance metrics.
+
+\highlight{\textbf{Finding 3.7} The study identifies three outputs used to check the shape of data.
+These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
 
@@
 
-    O6&
+    L6&
     \lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$&
@@
 
-    O10&
+    L10&
     \lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$&
@@
 
-Table~\ref{tab:linear-relation-check} illustrates two outputs found in this study for identifying linear relationships amongst features in the data.
-In output \emph{O6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
-Output \emph{O10} depicts the visualisation of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
-
-% TODO: feedback from Luis
-% - this explains what the code does but it does not pinpoint observations on how the output of code cells are used.
-E.g., here it is mostly visualisations - that's an observation.
-They are both scatterplots, one had to change the height of the plot (not sure that's relevant.
-We do not observe prints or asserts that help with this task.
-Linear machine learning models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
+Linear ML models achieve optimal performance when the target variable can be expressed as a linear combination of the input features.
 However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
 Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.
-
-\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data.
-visualisations such as a scatter plot with a regression line (\emph{O10}) and a line plot (\emph{O6}) are employed to assess the linearity between features.
+Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships amongst features in the data.
+In addition to the code, the raw output of the cell is also shown since both case-studies produce a visualisation.
+In \emph{L6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
+\emph{L10} is a visualisation of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
+
+\highlight{\textbf{Finding 3.8} The study identifies two outputs used to verify linear relationships between features in the data.
+Visualisations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
 These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
 
-\subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
-% (P71, P86, O64, O66, P68, P82, P107)
-
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to verify if different types of resources are available on the system where the notebook is being executed.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    P68&
-    \lstinline[]$print('GPU is available')$\\
-
-    P71&
-    \lstinline[]$print('Hub version: ', hub.__version__)$\\
-
-    P82&
-    \lstinline[]$print('Running on TPU ', tpu.master())$\\
-
-    P86&
-    \lstinline[]$print('Cuda is available')$\\
-
-    P107&
-    \lstinline[]$print('Model loaded')$\\
-
-    O64&
-    \lstinline[]$full_table.head(-5)$\\
-
-    O66&
-    \lstinline[]$prostate_cancer_df.shape$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:resource-check}
-\end{table}
-
-% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
-% TODO: feedback from luis
-% - I probably missed it but it should be clear why we have P and O in the key, given that they are all outputs.
-Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed.
-Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU.
-This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory.
-For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory.
-Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data.
-These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.
-
-Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system.
-This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.
-
-\highlight{\textbf{Finding 1.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
-These checks include confirming the availability of compute resources like GPUs and TPUs (P68, P82 and P86), ensuring datasets or pre-trained models are successfully loaded (P107, O64 and O66), and verifying the presence of specific library versions (P71).
-These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
+\subsubsection{Type check ($N = 2$)}
+% (O71, P43)
+
+Type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
+Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis phase.
+Since all ML algorithms perform mathematical operations, they require the input data to be entirely numerical to avoid computational errors (\emph{P43, L71}).
+The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
+For example, numerical features often require normalisation to bring them within a specific scale and categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
+
+\highlight{\textbf{Finding 3.9} The study identifies two outputs used to verify the data types of features in the dataset.
+These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
+Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
 
@@
 % (P66)
-
-\begin{table}
-  \centering
-  \caption{Print statement used to check the total execution time of training and cross-validating an ML model.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    P66&
-    \lstinline[]$print('Total Run Time:')$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:exec-time}
-\end{table}
-
-Table~\ref{tab:exec-time} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
-\emph{P66} is also used in other code cells in the notebook that train other ML models.
-
-While training multiple machine learning models for a single task is common, choosing the best model considers not only its performance but also its training speed.
+\begin{lstlisting}[caption={Print statement \emph{P66} used to check the total execution time of training an ML model with various hyper-parameter configurations.}, label={lst:P66}]
+print('Total Run Time:')
+\end{lstlisting}
+
+While training multiple ML models for a single task is common, choosing the best model considers not only its performance but also its training speed.
 Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}.
-
-\highlight{\textbf{Finding 1.4} The study identifies one output (P66) used to check the total execution time of training and cross-validating an ML model.
-The print statement is used for evaluating the efficiency of model training processes, especially when multiple models are trained iteratively.}
-
-\subsubsection{Missing value check ($N = 3$)}
-% (O12, O36, P74)
-
-\begin{table}
-  \centering
-  \caption{Examples of cell outputs used to check for missing data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    P74&
-    \lstinline[]$print(train_df.isnull().sum())$\\
-
-    O12&
-    \lstinline[]$sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')$&
-    \includegraphics[width=\linewidth]{missing-value-check.png}\\
-
-    O36&
-    \lstinline[]$test.isna().sum().unique()$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:missing-value}
-\end{table}
-
-Table~\ref{tab:missing-value} presents all case studies where cell outputs were used to check for missing data.
-While \emph{P74} and \emph{O36} perform this check in code, \emph{O12} presents a visual representation of missing data using a heatmap.
-
-Checking for missing values is a critical step in the preprocessing phase of a machine learning project because it significantly influences the quality and performance of the model.
-Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
-Many machine learning algorithms, including those that involve matrix multiplication operations such as linear regression and neural networks, demand complete numerical datasets to perform calculations.
-Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
-
-\highlight{\textbf{Finding 1.5} The study identifies three outputs (P74, O12 and O36) used to check for missing data.
-Missing values can lead to biased or incorrect conclusions, and many ML algorithms require complete numerical datasets for accurate calculations.
-Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
-
-\subsubsection{Model performance check ($N = 33$)}
-% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
-% TODO: feedback from Luis
-% - if needed, we can merge Excution time check here. (both execution time and model correctness can be presented as a KPI)
-\begin{table}
-  \centering
-  \caption{Various cell outputs used to check the performance of an ML model after training.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Visual Output}}\\
-    \midrule
-
-    P3&
-    \lstinline[]$print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')$&
-    \\
-
-    P6&
-    \lstinline[]$print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))$&
-    \\
-
-    P18&
-    \lstinline[]$print('The Accuracy is:', accuracy_score(y_test, y_pred))$&
-    \\
-
-    P50&
-    \lstinline[]$print('Classification Report: SVM (validation data)')$&
-    \\
-
-    P54&
-    \lstinline[]$print('Intercept value:', lm.intercept_)$&
-    \\
-
-    O3&
-    \lstinline[]$skplt.metrics. plot_confusion_matrix(Y_val, Vote.predict(X_val), normalize=True, figsize=(10, 10))$&
-    \includegraphics[width=\linewidth]{model-perf-confusion-matrix.png}\\
-
-    O52&
-    \lstinline[]$spot_check_recs(classifier, 910)$&
-    \\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-perf}
-\end{table}
-
-Table~\ref{tab:model-perf} presents a few examples of cell outputs used to check the performance of a trained ML model.
-Besides the \emph{accuracy} of the model (\emph{P3} and \emph{P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
-
-% TODO: this feels hallucinated; revisit
-Print statement \emph{P54} prints the intercepts of a linear regression model.
-Checking the intercept of a linear regression model is essential for understanding the baseline prediction when all predictors are zero, interpreting the model, identifying potential data biases or missing variables, assessing model validity, and comparing baseline predictions across different models.
-This insight is critical for accurate model interpretation and effective decision-making in practical applications.
-
-Output \emph{O3} shows a heatmap used to visualise the confusion matrix of a multi-label classification task.
-The values are normalised using the actual number of examples of each class which makes the comparisons of the values across the classes possible.
-
-Output \emph{O52} shows a function implemented by the practitioners to manually evaluate an image classification model on a random sample of input images.
-The function \texttt{spot\_check\_recs} displays the input image along with the prediction of the model which the practitioner uses to spot-check the performance of the model.
-
-During the model development phase, the performance of an ML model is often printed to facilitate comparisons against other models or variations.
-Throughout this continuous experimentation phase, authors may adjust model parameters or modify the data by engineering new features.
-The model is then re-trained to check if these changes lead to performance improvements.
-
-\highlight{\textbf{Finding 1.6} The study identifies 33 outputs used to check the performance of trained ML models.
-These include outputs that print key performance metrics (P3, P6, P18, P50), a normalised heatmap used to visualise the confusion matrix of a model's predictions (O3) and a custom function written for manual evaluation of model predictions on random input samples (O52).
-These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
-
-\subsubsection{Neural network architecture ($N = 1$)}
+Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
+
+\highlight{\textbf{Finding 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model.
+The print statement can be used to evaluate the efficiency of model training, especially when multiple models are trained iteratively.}
+
+\subsubsection{Network architecture check ($N = 1$)}
 % (P92)
 
-\begin{table}
-  \centering
-  \caption{Print the architecture of a neural network.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-    P92&
-    \lstinline[]$print(MyNetwork)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:network-architecture}
-\end{table}
-
-% TODO: need more description and context around this print statement (where was it written, and why?)
-Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.
-
-Printing the structure of a neural network serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
-It helps in verifying configurations like layer dimensions and operations before initiating computationally intensive training, ensuring the model is set up efficiently for the intended tasks.
-This practice is fundamental for maintaining accuracy, facilitating collaboration, and enhancing the reproducibility of the model.
-
-\highlight{\textbf{Finding 1.7} The study identifies one output (P92) used to print the architecture of a neural network.
+Printing the structure of a neural network (\emph{P92}) serves crucial purposes such as verifying the correct implementation of the architecture, aiding in debugging by ensuring layer compatibility, and providing a clear, human-readable format for documentation and educational insights.
+It helps in verifying configurations like layer dimensions and operations before initiating computationally intensive training and ensuring that the model is set up efficiently for the intended tasks.
+
+\highlight{\textbf{Finding 3.11} The study identifies one print statement used to show the architecture of a neural network.
 Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
-
-\subsubsection{Shape check ($N = 3$)}
-% (P4, P32, P117)
-
-\begin{table}
-  \centering
-  \caption{Example of outputs used to check the shape of the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}} & \emph{\textbf{Code}}\\
-    \midrule
-    P4 &
-    \lstinline[]$print('no.of examples in test data : ', len(test_data))$\\
-
-    P32 &
-    \lstinline[]$print('Training set shape : ', x_train.shape)$\\
-
-    P117&
-    \lstinline[]$print('Y_train.shape: ', Y_train.shape)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:shape-check}
-\end{table}
-
-Table~\ref{tab:shape-check} shows print statements discovered in this study which check the shape of the data.
-\emph{P4} prints the number of rows or examples in the test set.
-\emph{P32} and \emph{P117} prints the number of rows and columns (or features) in the training and test data respectively.
-
-Ensuring that data dimensions align with expectations is important particularly following data pre-processing or transformation steps.
-Primarily, the number of features in the training set must match those in the testing set.
-This alignment is crucial because statistical machine learning models are trained on specific data dimensions and expect the same dimensional structure during inference to perform accurately.
-Similarly, in neural network architectures, the configuration of input layers depends directly on the shape of the training data, dictating the number of input neurons needed.
-Furthermore, the correspondence between the number of test examples and their respective labels is essential for accurately computing performance metrics, such as accuracy.
-
-Such checks are indicative of the meticulous attention to detail necessary in the model training and validation process, underscoring the importance of shape validation in achieving reliable machine learning outcomes.
-
-\highlight{\textbf{Finding 1.8} The study identifies three outputs used to check the shape of data.
-These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
-
-\subsubsection{Type check ($N = 2$)}
-% (O71, P43)
-
-% TODO: the narrative is not as strong here; it would help to also show the actual output as well?
-For instance, in P43 by itself doesn't make that much sense, the output however shows that all features in the dataset are floats which is a pre-requisite for ML training data
-\begin{table}
-  \centering
-  \caption{Outputs used to check the data type of features in the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}} &
-    \emph{\textbf{Code}}\\
-
-    P43&
-    \lstinline[]$print('data type:', images.dtype)$\\
-
-    O71&
-    \lstinline[]$type(Y)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:type-check}
-\end{table}
-
-% TODO: for O71, the output shows its a ndarray; without showing this to the reader we cannot make a strong narrative around these type checks
-Table~\ref{tab:type-check} illustrates outputs of cells found in our analysis, where checks are employed to verify the data types of the features in the dataset.
-Print statement \emph{P43} is written prior to training a neural network.
-The output validates that all features in the training data are numerical.
-Similarly, output \emph{O71} shows the data type of the target column.
-
-Type checks are crucial not only for confirming the current state of the data but also for determining the specific transformations required to make the data suitable for subsequent modelling steps.
-Ensuring data types align with model requirements is a fundamental step in the exploratory data analysis (EDA) phase of developing machine learning models.
-Since all machine learning algorithms perform mathematical operations, such as matrix multiplications, they require the input data to be entirely numerical to avoid computational errors.
-The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
-For example, while numerical features often require normalisation to bring them within a specific scale, categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
-
-\highlight{\textbf{Finding 1.9} The study identifies two outputs used to verify the data types of features in the dataset.
-These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
-Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
-
-\subsubsection{Spot checks ($N = 5$)}
-% (O56, O60, P64, P67, P114)
-
-\begin{table}
-  \centering
-  \caption{Example of outputs used to perform spot checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    O60 &
-    \lstinline[]$X_pca.head()$\\
-
-    P64 &
-    \lstinline[]$print(np.max(cur[:, :, 1]))$\\
-
-    P114 &
-    \lstinline[]$print(onehot_encoded)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:spot-check}
-\end{table}
-
-Table~\ref{tab:spot-check} presents outputs which were used to perform spot checks at various stages of the ML development cycle.
-In \emph{O60}, the practitioner verifies that the number of features in the data matches expectations after apply Principal Component Analysis (PCA).
-\emph{P64} demonstrates another case which is particularly relevant in image processing contexts.
-The maximum value in the second channel of a 3D numpy array is extracted and verified, ensuring that the data manipulation retains its expected properties.
-Lastly, \emph{P114} assesses the correctness of one-hot encoding, a fundamental preprocessing step for categorical data, ensuring that the transformation has been executed correctly.
-
-Performing value or spot checks can be an essential practice for ensuring data integrity and model accuracy at various stages of development.
-These checks are crucial for verifying that operations such as data transformations, model predictions, and feature engineering are functioning as expected.
-Each of these checks not only validates individual transformations or predictions but also reinforces the overall reliability and robustness of the machine learning models being developed.
-
-\highlight{\textbf{Finding 1.10} This study finds five outputs used to perform spot checks at various stages of the MLDL.
-Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
-
-\subsubsection{Model training ($N = 4$)}
-% (O8, O31, O42, P77)
-
-% TODO: feedback from luis
-% - this cluster makes me wonder: should we make a point that some of these outputs may happen by chance and were not explicitly defined by users/developers?
-\begin{table}
-  \centering
-  \caption{Example of outputs used to monitor model training.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    O8 &
-    \lstinline[]$autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)$\\
-
-    O31 &
-    \lstinline[]$adaBoost.fit(X_train, y_train)$\\
-
-    O42&
-    \lstinline[]$m_r.best_params_$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-training}
-\end{table}
-
-During training it is common practice to monitor progress periodically, as illustrated by outputs \emph{O8} in Table~\ref{tab:model-training}, where the training loss and accuracy are printed to give real-time feedback on the model's learning efficacy.
-This feedback allows for adjustments in training parameters or early stopping to prevent overfitting.
-For instance, in output O8, the autoencoder is trained using a custom callback from \texttt{keras\_utils} to provide a progress update without cluttering the output stream, which is particularly useful in lengthy training sessions.
-In contrast, output \emph{O31} represents a crucial validation step where the \texttt{adaBoost} model is fitted to the training data to ensure it adheres to the expected data patterns without errors, a typical task in the model experimentation phase where various models are assessed for their suitability.
-Output \emph{O42} checks for the best parameters found through tuning methods, highlighting the importance of parameter optimisation in achieving optimal model performance.
-
-Each of these outputs underscores different facets of the model training phase, from initialization and real-time monitoring to validation and finalization, illustrating the layered complexity of developing predictive models.
-
-\highlight{\textbf{Finding 1.11} The study identifies four outputs used to monitor model training progress.
-These outputs provide real-time feedback on training loss and accuracy (O8), validate model adherence to expected data patterns (O31), and identify optimal parameters through tuning methods (O42).
-These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
-
-\subsection{(RQ3) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}
-
-% TODO: bring in the notion of pre and post condition checks (the wikipedia page on assertions is very interesting: https://en.wikipedia.org/wiki/Assertion_(software_development)); specifically the part on run-time assertions
-
-\subsubsection{Batch size check ($N = 3$)}
-% (A21, A28, A70)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to validate the batch size of input data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A21 &
-    \lstinline[]$assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'$\\
-
-    A28 &
-    \lstinline[]$assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'$\\
-
-    A70 &
-    \lstinline[]$assert n_img > batch_size$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:batch-size}
-\end{table}
-
-Table~\ref{tab:batch-size} shows the assertions found in this study that validate the batch size.
-\emph{A21} ensure that the batch size divides evenly into the training dataset size, thereby confirming that every training epoch processes full batches without leftover data.
-Similarly, \emph{A28} checks that image dimensions are suitably divisible by the patch size, which is essential for certain types of convolutional networks.
-\emph{A70} is another critical check which confirms that the number of images is greater than the batch size, which is necessary to commence batch processing.
-
-In the context of neural network training, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilization, particularly GPU cores.
-Batching allows for a more stable and smoother learning curve as the model's weights are updated not for each individual example, but rather based on the average gradient of a batch.
-This technique not only optimizes learning but also enhances the model's generalizability due to more robust gradient estimations.
-Given these benefits, it becomes crucial to ensure that the batch size is appropriately set to match the hardware's capacity, avoiding memory overflow issues that can halt training.
-These assertions are strategically placed to prevent common errors that could undermine training efficiency and model performance, thereby safeguarding the training process from potential disruptions due to misconfigured batch sizes.
-
-\highlight{\textbf{Finding 2.1} This study identifies three assertions validating batch size configurations.
-These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
-These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
-
-\subsubsection{Data leakage check ($N = 1$)}
-% (A33)
-
-\begin{table}
-  \centering
-  \caption{Assert used to prevent data leakage.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A33&
-    \lstinline[]$assert len(set( tr_df.PetID.unique()).intersection(valid_df.PetID.unique())) == 0$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:data-leakage}
-\end{table}
-
-Assertion \emph{A33} in Table~\ref{tab:data-leakage} ensures that there are no overlapping \texttt{PetID} values between the training set \texttt{tr\_df} and the validation set \texttt{valid\_df}, thereby preventing any possibility of data leakage.
-
-Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in machine learning to prevent models from overfitting.
-Overfitting occurs when a model is excessively complex, learning not only the underlying patterns but also the noise in the training data, which can degrade its performance on unseen data.
-To guarantee that the model can generalize effectively to new examples, it's crucial to verify that the training and validation sets are completely distinct with no shared examples.
-By strictly separating these datasets, the validation phase serves its purpose of providing an unbiased evaluation of the model's ability to generalize---to data similar to, but not identical to that which it was trained on---thus maintaining the integrity and validity of the evaluation metrics derived during model testing.
-
-\highlight{\textbf{Finding 2.2} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
-Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
-This practice maintains the integrity and validity of the model evaluation metrics.}
-
-\subsubsection{Existence check ($N = 8$)}
-% NOTE: this is a combination of missing-value-check and existence-check
-% (A23, A42, A43, A50, A51, A63, A79, A86)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to perform existence checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A23 &
-    \lstinline[]$assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())$\\
-
-    A42 &
-    \lstinline[]$assert not lab_s.isnull().values.any()$\\
-
-    A43 &
-    \lstinline[]$assert len(data) != 0, 'cannot divide by zero'$\\
-
-    A50 &
-    \lstinline[]$assert not np.any(np.isnan(X))$\\
-
-    A51 &
-    \lstinline[]$assert data.target.notnull().all()$\\
-
-    A63 &
-    \lstinline[]$assert X.isnull().sum().sum() == 0$\\
-
-    A79 &
-    \lstinline[]$assert not processed_data_df.isna().any().any()$\\
-
-    A86 &
-    \lstinline[]$assert p0 in poi_info.index$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:existence-check}
-\end{table}
-
-Assertions \emph{A23} and \emph{A42} in Table~\ref{tab:existence-check} ensure that specific data columns do not contain any missing entries, which could compromise the reliability of subsequent analyses.
-Assertion \emph{A43} guards against operations on empty datasets, which are a potential risk after filtering or merging data processes.
-Assertions \emph{A50}, \emph{A51} and \emph{A63} rigorously confirm that the datasets are entirely free from \texttt{NaN} values, thus preserving the integrity of the dataset for reliable model training.
-Assertions \emph{A79} and \emph{A86} ensure that no unexpected missing values are introduced and that all expected indices are present in the dataset respectively.
-
-Existence checks are a fundamental component of data preprocessing in ML, serving multiple critical purposes to ensure data integrity before analysis and modelling.
-These checks primarily focus on verifying the presence of necessary columns and the absence of missing values within those columns.
-Such validations are especially crucial after data preprocessing steps, where transformations might inadvertently introduce null values.
-These checks are also crucial for maintaining the accuracy and efficiency of data handling processes, ensuring that the datasets are ready for robust machine learning applications.
-
-\highlight{\textbf{Finding 2.3} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
-These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
-
-\subsubsection{Mathematical property checks ($N = 4$)}
-% (A3, A25, A56, A64)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to validate mathematical properties of neural networks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A3 &
-    \lstinline[]$assert (xH - wH) % self.stride == 0$\\
-
-    A25 &
-    \lstinline[]$assert test_output.std() < 0.15, "Don't use batchnorm here"$\\
-
-    A56 &
-    \lstinline[]$assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))$\\
-
-    A64 &
-    \lstinline[]$assert np.allclose(T, T.T)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:maths-check}
-\end{table}
-
-Assertion \emph{A3} in Table~\ref{tab:maths-check} checks that the dimension differences between height of input and the filter, when divided by the stride, result in an integer, ensuring that convolution operations are dimensionally consistent.
-\emph{A25} monitors the standard deviation of outputs to decide the appropriateness of using batch normalisation, thus guiding the model optimization process based on statistical behaviour.
-\emph{A56} verifies the uniformity of certain computations, essential for maintaining consistency in state or parameter arrays over iterations.
-Similarly \emph{A64} checks for the symmetry of a matrix, a crucial property in many algorithms, particularly those involving covariance matrices and distance calculations.
-
-When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigour is captured through the assertions in Table~\ref{tab:maths-check} tha validate these properties after matrix manipulations.
-These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles, thereby ensuring the robustness and reliability of model computations and outcomes.
-
-\highlight{\textbf{Finding 2.4} Four assertions validate key mathematical properties of data and model outputs.
-These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
-These checks confirm the robustness and reliability of model computations.}
-
-\subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
-% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check performance of ML models.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A7 &
-    \lstinline[]$assert len(neighbours_1) == 20, "Neighbors don't match!"$\\
-
-    A15 &
-    \lstinline[]$assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))$\\
-
-    A19 &
-    \lstinline[]$assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'$\\
-
-    A38 &
-    \lstinline[]$assert 0.75 < auc(fpr, tpr) < 0.85$\\
-
-    A58 &
-    \lstinline[]$assert np.isclose(accuracy, 0.9666666666666667)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:model-perf-explicit}
-\end{table}
-
-This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
-
-% TODO: verify if this applies to recommender systems as well?
-For example, \emph{A7} in Table~\ref{tab:model-perf-explicit} checks for the expected number of neighbours as an indirect measure of model performance in specific scenarios like recommendation systems or clustering.
-Meanwhile, direct comparisons of model outputs with expected results, as seen in \emph{A15} validate the model's predictive accuracy and reliability under operational conditions.
-Other assertions, such as \emph{A19}, ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity.
-Range checks like \emph{A38} confirm that the performance metrics such as \emph{AUC} and \emph{Gini coefficient} fall within acceptable boundaries, indicating good predictive performance without overfitting.
-
-The use of \texttt{np.isclose} in \emph{A58} is particularly noteworthy.
-This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
-By allowing a small tolerance in the comparison, np.isclose ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
-
-\highlight{\textbf{Finding 2.5} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
-These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
-
-\subsubsection{Network architecture check ($N = 3$)}
-% (A11, A62, A75)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the architecture of neural networks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A11 &
-    \lstinline[]$assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()$\\
-
-    A62 &
-    \lstinline[]$assert activations[i + 1].shape[1] == self.encoder_layer_sizes[-1] * 2$\\
-
-    A75 &
-    \lstinline[]$assert reg in ['none', 'l2']$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-network-architecture}
-\end{table}
-
-Assertion \emph{A11} verifies that the weight dimensions of the first convolutional layer in the \lstinline{encoder_conv_01} module match the weight dimensions of the third layer in the \lstinline{vgg16.features} module.
-The \lstinline{vgg16} model is a widely-used pre-trained convolutional neural network often employed as a feature extractor or backbone for transfer learning tasks.
-By ensuring that the weight dimensions are consistent, the code guarantees that the input data to the \lstinline{encoder_conv_01} module has the expected spatial dimensions and number of channels, enabling seamless integration with the pre-trained \lstinline{vgg16} model.
-
-Assertion \emph{A62} checks that the shape of the activations at index \lstinline{i + 1} has the correct number of channels, which should be twice the size of the last encoder layer.
-This ensures that the intermediate activations maintain the expected structure necessary for the proper functioning of the subsequent layers in the network.
-Ensuring that the shape of activations is consistent throughout the network is crucial for maintaining the integrity of the data flow and preventing errors during the forward pass.
-
-Assertion \emph{A75} ensures that the regularization parameter \lstinline{reg} is either 'none' or 'l2'.
-This check is important for validating that the specified regularization method is among the acceptable options, thereby preventing potential misconfigurations that could lead to unexpected behaviour or suboptimal model performance.
-
-When integrating pre-trained models or leveraging transfer learning techniques, it is crucial to ensure the compatibility of the custom neural network architecture with the pre-trained components.
-This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
-Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible, ultimately leading to reliable and accurate model performance.
-
-\highlight{\textbf{Finding 2.6} Three assertions verify the consistency of neural network architecture.
-These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
-These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
-
-\subsubsection{Resource check ($N = 7$)}
-% (A10, A14, A18, A37, A60, A67 A74)
-
-\begin{table}
-  \centering
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A10 &
-    \lstinline[]$assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."$\\
-
-    A14 &
-    \lstinline[]$assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'$\\
-
-    A18&
-    \lstinline[]$assert pd.__version__.rpartition('.')[0] == '1.0', f"Unexpected pandas version: expected 1.0, got {pd.__version__.rpartition('.')[0]}"$\\
-
-    A37 &
-    \lstinline[]$assert svm.fit_status_ == 0, 'Forgot to train the SVM!'$\\
-
-    A60 &
-    \lstinline[]$assert f2.gca().has_data()$\\
-
-    A67&
-    \lstinline[]$assert pm.__version__ == '3.9.2'$\\
-
-    A74 &
-    \lstinline[]$assert os.path.exists(image_dir)$\\
-    \bottomrule
-  \end{tabular}
-  \caption{Assertions used to validate availability of resources.}
-  \label{tab:assert-resource-check}
-\end{table}
-
-Table~\ref{tab:assert-resource-check} persents the assertions encountered in this study, that perform resource checks encompassing pre-trained models, data paths, visualisations, and model training status.
-These checks help prevent runtime errors and ensure the proper execution of code.
-
-One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
-These checks ensure that the required resources are present before proceeding with subsequent operations, preventing potential issues caused by missing resources.
-
-Additionally, assertions are employed to validate the successful loading of models (\emph{A14}) and the completion of model training processes (\emph{A37}).
-A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
-The assertion \emph{A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
-This check prevents inconsistent model states from executing cells out of order or multiple times.
-
-Assertion \emph{A60} is used to verify the presence of data in visualisations generated using the Matplotlib~\footnote{https://matplotlib.org/} library.
-These checks are crucial for ensuring that the visualisations accurately represent the underlying data and do not contain empty or incomplete plots, which could lead to misleading interpretations.
-
-Assertions \emph{A18} and \emph{A67} explicitly verify the versions of critical libraries.
-For instance, \emph{A18} checks that the major version of pandas is precisely what is expected for the notebook’s tasks, ensuring compatibility and preventing errors that could arise from API changes in newer or older versions.
-
-Resource check assertions play a vital role in maintaining the reliability and robustness of the machine learning workflows by ensuring the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations.
-
-% NOTE: left over loads, move to implications if required
-% In the dynamic and sometimes unstructured environment of Jupyter notebooks, managing dependencies directly within the notebook becomes crucial due to the workflow peculiarities associated with interactive sessions.
-Unlike traditional software development environments where dependencies are managed through dedicated solutions like dependency management tools or by specifying requirements in a \emph{requirements.txt} file, notebooks often require on-the-fly checks to ensure the correct library versions are loaded to avoid compatibility issues.
-
-% These assertions safeguard the notebook against potential disruptions caused by version mismatches, thereby maintaining the integrity of the codebase and ensuring reproducibility across different execution environments.
-
-\highlight{\textbf{Finding 2.7} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
-These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
-
-\subsubsection{Data shape check ($N = 26$)}
-% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
-
-% TODO: revisit the assertions in this section; we may have misclassified assertions or need to create a new group
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the shape of the data.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A5 &
-    \lstinline[]$assert y_valid.shape == (1132,)$\\
-
-    A17 &
-    \lstinline[]$assert X.shape[1] == 13, 'Did you drop/lose some columns in X? Did you properly load and split the data?'$\\
-
-    A29 &
-    \lstinline[]$assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'$\\
-
-    A31 &
-    \lstinline[]$assert img.shape == (112, 92)$\\
-
-    A76 &
-    \lstinline[]$assert len(encoding['token_type_ids']) == max_seq_length$\\
-
-    A84 &
-    \lstinline[]$assert red.get_shape().as_list()[1:] == [224, 224, 1]$\\
-
-    A90 &
-    \lstinline[]$assert len(X_train) == 2000$\\
-
-    A93 &
-    \lstinline[]$assert temp_embed.shape[0] == stride$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-shape-check}
-\end{table}
-
-Assertions \emph{A17} and \emph{A90} verify that the input features have the correct number of columns and examples, respectively.
-Such checks are essential for preventing downstream errors and ensuring the compatibility of the data with the chosen model architecture and preprocessing steps.
-
-Assertions \emph{A5} and \emph{A29} ensure that the labels or predictions have the expected length or shape, which is crucial for accurate performance evaluation and model training.
-
-In the context of computer vision tasks, assertions \emph{A31} and \emph{A84} verify that the input images have the correct dimensions, which is essential for compatibility with convolutional neural network architectures and image preprocessing techniques.
-
-Furthermore, assertions \emph{A76} and \emph{A93} check the dimensions of sequences, embeddings, and other data structures used in natural language processing tasks, ensuring their compatibility with various models and algorithms.
-
-Data shape checks can be considered the ``Swiss army knife'' of assertions, as they are ubiquitous and serve multiple purposes throughout the machine learning workflow.
-We encountered a vast array of assert statements that verify the dimensions of various data structures, including input features, labels, predictions, images, sequences, and embeddings.
-These assertions ensure that the data adheres to the expected format and dimensions required by different components of the machine learning pipeline.
-
-Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
-By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
-
-\highlight{\textbf{Finding 2.8} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
-These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
-
-\subsubsection{Type check ($N = 5$)}
-% (A2, A35, A40, A81, A88)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to check the data type of features in the dataset.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A2 &
-    \lstinline[]$assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'$\\
-
-    A35 &
-    \lstinline[]$assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"$\\
-
-    A40 &
-    \lstinline[]$assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)$\\
-
-    A81 &
-    \lstinline[]$assert is_all_ints(filled_df[r]) is True$\\
-
-    A88 &
-    \lstinline[]$assert isinstance(betas, np.ndarray)$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:assert-type-check}
-\end{table}
-
-Assertion \emph{A2} in Table~\ref{tab:assert-type-check} demonstrates a very common type check performed on the input features.
-This assertion ensures that the input features are represented as PyTorch float tensors, which is a requirement for many deep learning models and operations.
-
-Another type check is applied to verify the type of the machine learning models themselves.
-For instance, the \emph{A40} checks if the \lstinline{model_3} object is an instance of the \lstinline{RandomForestClassifier} class from the scikit-learn library.
-This type of assertion is essential when working with different types of models, as it helps prevent errors arising from using incompatible methods or attributes.
-
-In addition to input features and models, type checks are also performed on data structures and intermediate objects.
-The assertion \emph{A35} verifies that the \lstinline{column_transformer} object is an instance of the \lstinline{ColumnTransformer} class, which is commonly used for preprocessing and feature engineering tasks in scikit-learn.
-
-Type checking is not limited to objects and data structures--it can also be applied to individual columns or elements within a dataset.
-The assertion \emph{A81} checks if all the elements in a specific column (\lstinline[]$filled_df[r]$) of a DataFrame are integers.
-
-Furthermore, type checks are performed on NumPy arrays, as demonstrated by the assertion \emph{A88}.
-This assertion ensures that the \lstinline{betas} object is a NumPy array, which is a fundamental data structure used in many numerical and scientific computing operations.
-
-Type check assertions are essential for maintaining data integrity, ensuring compatibility between different components of the machine learning pipeline, and preventing errors and unexpected behaviour.
-By verifying the types of input features, models, data structures, and other objects, these assertions contribute to the overall reliability and robustness of the machine learning workflow.
-
-\highlight{\textbf{Finding 2.9} Five assertions validate the data types of input features, models, and intermediate objects.
-These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
-
-\subsubsection{Data validation check ($N = 14$)}
-% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
-
-\begin{table}
-  \centering
-  \caption{Assertions used to perform data validation checks.}
-  \begin{tabular}{@{}m{0.05\textwidth} m{0.8\textwidth}@{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}\\
-    \midrule
-
-    A41 &
-    \lstinline[]$assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))$\\
-
-    A44 &
-    \lstinline[]$assert abs(min(entity_embedding[low:high]) - max(entity_embedding[low:high])) <= 1$\\
-
-    A46 &
-    \lstinline[]$assert np.isclose(stdev_norm, 1.0, atol=1e-16)$\\
-
-    A52 &
-    \lstinline[]$assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()$\\
-
-    A65 &
-    \lstinline[]$assert np.all(y <= nb_classes)$\\
-
-    A73 &
-    \lstinline[]$assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:data-validation}
-\end{table}
-
-% TODO: revisit, we have new assertions here; update the count
-Table~\ref{tab:data-validation} presents a range of assert statements that perform various validation checks on data structures, such as NumPy arrays and pandas DataFrames.
-These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
-
-One type of validation check involves verifying the presence of specific values or ranges within arrays or columns.
-For example, assertion \emph{A41} checks if the unique values in the \lstinline{smoke} column of the DataFrame \lstinline{X} are limited to 0 and 1.
-Similarly, \emph{A65} ensures that all values in the array \lstinline{y} are less than or equal to a certain number of classes.
-
-Another common validation check is verifying the equality or closeness of values or arrays.
-Assertion \emph{A44} checks if the difference between the minimum and maximum values in a specific range of the \lstinline{entity_embedding} array is less than or equal to 1.
-Assertion \emph{A46} verifies that the \lstinline{stdev_norm} value is close to 1.0 within a specified absolute tolerance.
-
-Some assertions focus on validating the uniqueness or cardinality of values in arrays or DataFrames.
-For instance, \emph{A52} checks if the number of unique user IDs in the \lstinline{grouped_users} DataFrame is equal to the number of unique user IDs in the \lstinline{user_engagement} DataFrame.
-
-Other assertions validate the presence of specific values or conditions within DataFrames or other data structures.
-The assertion \emph{A73} checks if the count of a specific value (1) in the ``clf'' column of the DataFrame \lstinline{df} is equal to the number of rows where the ``quality'' column is greater than or equal to 7.
-
-By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
-
-\highlight{\textbf{Finding 2.10} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
-These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
-
-\subsection{(RQ4) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}
-
-\subsubsection{Model Performance ($N = 15$)}
-
-\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-  \caption{VA pairs used to validate the performance of a model.} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endfirsthead
-
-  \caption[]{VA pairs used to validate the performance of a model (continued).} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endhead
-
-  \midrule
-  \endfoot
-
-  \bottomrule
-  \endlastfoot
-
-  VA2 &
-  \includegraphics[width=\linewidth]{va2.png} &
-  \lstinline[]$assert final_score > 300, 'not good enough for DQN'$\\
-
-  VA3 &
-  \includegraphics[width=\linewidth]{va3.png} &
-  \lstinline[]$assert accuracy_score(y_test, pred) > 0.95$\\
-
-  VA9 &
-  \includegraphics[width=\linewidth]{va9.png} &
-  \lstinline[]$assert np.mean(mean_rw_history[-10:]) > 10.$\\
-
-  VA11 &
-  \includegraphics[width=\linewidth]{va11.png} &
-  \lstinline[]$assert np.mean(metrics['dev_bleu'][-10:], axis=0)[1] > 35, "We kind of need a higher bleu BLEU from you. Kind of right now."$\\
-
-  VA15 &
-  \includegraphics[width=\linewidth]{va15.png} &
-  \lstinline[]$assert np.mean(train_history[:10], axis=0)[1] > np.mean(train_history[-10:], axis=0)[1], "The model didn't converge."$\\
-
-  VA28 &
-  \includegraphics[width=\linewidth]{va28.png} &
-  \lstinline[]$assert np.mean(dev_recall_history[-10:]) > 0.85, "Please train for at least 85% recall on test set. You may need to change vectorizer model for that."$\\
-
-  VA68 &
-  \includegraphics[width=\linewidth]{va68.png} &
-  \lstinline[]$np.testing.assert_almost_equal(ans4, 0.3042764)$\\
-
-  VA107 &
-  \includegraphics[width=\linewidth]{va107.png} &
-  \lstinline[]$assert r2_gru > 0.6$\\
-
-\end{longtable}\label{tab:va-model-perf}
-
-The visualisations in VA2, VA9, and VA15 provide a comprehensive view of key metrics for evaluating the performance of various machine learning models.
-VA2 includes mean reward per episode, TD loss history, initial state value (V), and gradient norm history for a Deep Q-Network (DQN) model.
-These metrics collectively indicate the model's learning progress and stability during training.
-The visualisation in VA9 shows the mean reward per game for a DQN applied to the Breakout Atari game, tracking the model's performance over 200 games.
-The general upward trend indicates learning and improvement in gameplay over time.
-VA15 presents the training and development loss over time for a character-level language model, showing a gradual decrease in loss and suggesting effective learning.
-
-The visualisation in VA3 presents a two-dimensional scatter plot showing the decision boundary of a Support Vector Machine (SVM) model, with red and yellow points representing two different classes and a background colour gradient indicating the decision function of the SVM.
-This visualisation aids in understanding the model's ability to distinguish between classes and the distance of the samples from the decision boundary.
-The visualisations in VA28 display the training loss and development recall, respectively.
-The left graph shows a decrease in training loss over time, indicating learning progress, while the right graph shows development recall over time, highlighting improvements in the model's ability to correctly identify positive instances.
-The visualisations in VA11 depict key metrics for evaluating the performance of a machine learning model in text generation or machine translation.
-The left plot shows a steady increase in the development BLEU score over the training iterations, indicating improving quality of the model's output, while the right plot displays a steady decrease in training loss, suggesting effective learning and error minimization.
-In VA107, the scatter plot displays the predicted versus true values of Nitrogen Dioxide (NO2) levels, used to assess the model's performance in predicting continuous variables.
-
-The corresponding assertions for these visualisations ensure that the models meet specific performance standards.
-For VA2, the assertion ensures that the final score exceeds 300, validating the model's performance for deployment.
-In VA9, the assertion checks whether the mean of the last ten rewards exceeds 10, ensuring stable and satisfactory learning progress.
-For VA15, the assertion checks if the mean of the first ten training observations is greater than the mean of the last ten, ensuring loss convergence and validating the model's learning without overfitting.
-The assertion in VA3 ensures the accuracy score on the test set exceeds 0.95, confirming the model's reliability and generalization capability.
-For VA28, the assertion checks whether the average development recall in the last ten evaluations exceeds 0.85, ensuring high recall rate critical for tasks like medical diagnostics.
-In VA11, the assertion ensures the average development BLEU score in the last ten evaluations exceeds 35, confirming the model produces high-quality translations.
-Lastly, in VA107, the assertion checks whether the R2 score exceeds 0.6, validating the model's fit and reliability for making accurate predictions in real-world scenarios.
-These validations confirm that iterative training processes lead to meaningful improvements and ensure the models' readiness for deployment and practical applications.
-
-\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models.
-The visualisations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores.
-Assertions associated with these visualisations ensure models meet a specific performance criteria.}
-
-\subsubsection{Distribution Check ($N = 1$)}
-
-\begin{table}
-  \centering
-  \caption{VA pairs used to check data distribution.}
-  \begin{tabular}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-    \toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Visual Output}}&
-    \emph{\textbf{Assertion}}\\
-    \midrule
-
-    VA13 &
-    \includegraphics[width=\linewidth]{va13.png} &
-    \lstinline[]$assert 80 < boot_mean_mean < 81$\\
-    \bottomrule
-  \end{tabular}
-  \label{tab:va-dist-check}
-\end{table}
-
-The visualisation in VA13 displays the distribution of bootstrapped sample means as a histogram with a superimposed kernel density estimate (KDE) curve, illustrating the central tendency and variability of the bootstrapped means.
-
-Bootstrapped sample means are calculated by repeatedly resampling the original dataset with replacement and computing the mean for each resampled dataset.
-This process generates a distribution of the mean, known as the bootstrap distribution, which helps in estimating the variability and central tendency of the original dataset's mean.
-
-The corresponding assertion checks whether the bootstrapped mean lies between 80 and 81, ensuring that the statistical estimation is consistent with expected values.
-This validation is crucial for confirming the reliability of the bootstrapped estimates and the accuracy of statistical conclusions.
-
-\highlight{\textbf{Finding 3.1} One visualisation is used to display the distribution of bootstrapped sample means.
-The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}
-
-\subsubsection{Data Validation ($N = 9$)}
-
-\begin{longtable}{@{} m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth} @{}}
-  \caption{VA pairs used for data validation.} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endfirsthead
-
-  \caption[]{VA pairs used for data validation (continued).} \\
-  \toprule
-  \emph{\textbf{Key}} &
-  \emph{\textbf{Visual Output}} &
-  \emph{\textbf{Assertion}}\\
-  \midrule
-  \endhead
-
-  \midrule
-  \endfoot
-
-  \bottomrule
-  \endlastfoot
-
-  VA30 &
-  \includegraphics[width=\linewidth]{va30.png} &
-  \lstinline[]$assert scaled_data[scaled_data.argmax()] <= 1$\\
-
-  V35 &
-  \includegraphics[width=\linewidth]{va35.png} &
-  \lstinline[]$assert(all(df_lower['inspection_score'] > 75))$\\
-
-  V57 &
-  \includegraphics[width=\linewidth]{va57.png} &
-  \lstinline[]$assert kstest(data_transformed[:, feature], 'norm').statistic < 1e-2$\\
-
-  VA62 &
-  \includegraphics[width=\linewidth]{va62.png} &
-  \lstinline[]$assert not np.allclose(x_aug, x_original)$\\
-
-\end{longtable}\label{tab:va-data-validatio}
-
-The VA pairs in Table~\ref{tab:va-data-validatio} collectively illustrate various data validation techniques essential for ensuring the integrity and quality of data used in ML models.
-
-The visualisation in VA30 consists of two histograms comparing the original and scaled distributions of an exponential dataset.
-The left plot shows the distribution of the original data, while the right plot displays the distribution after scaling.
-The scaling process adjusts the data to fall within a specified range of 0 and 1.
-The corresponding assertion checks whether the maximum value in the scaled data is less than or equal to 1, ensuring that the scaling procedure was applied correctly.
-
-The visualisation in VA35 shows the distribution of health inspection scores in areas with the bottom 25\% and top 25\% of median household income.
-The blue bars represent areas with the top 25\% median household income, while the orange bars represent areas with the bottom 25\%.
-This comparison helps identify any disparities in health inspection scores based on income levels.
-The corresponding assertion checks that all inspection scores in the lower-income areas are above 75, ensuring that the data reflects a minimum standard of health inspection scores in these areas.
-
-The visualisation in VA57 shows a histogram of transformed data after it has been normalised.
-The corresponding assertion checks that the Kolmogorov-Smirnov (KS) test statistic for the transformed data is less than $1 \times 10^{-2}$, ensuring that the data closely follows a normal distribution.
-
-The visualisation in VA62 displays a heatmap of an augmented dataset generated using a weighted k-nearest-neighbors (k-NN) method.
-This heatmap helps in visualizing the differences and similarities between the original and augmented data.
-The corresponding assertion checks that the augmented data \lstinline{x_aug} is not nearly identical to the original data \lstinline{x_original}, ensuring that the data augmentation process has introduced sufficient variability.
-
-These validations are crucial for confirming that data preprocessing techniques such as scaling, transformation, and augmentation have been correctly applied.
-Ensuring the integrity and quality of the data is essential for the subsequent steps in the machine learning pipeline, as it directly impacts the robustness, reliability, and generalization of the models developed.
-By validating these steps, we can maintain high standards of data quality and model performance, which are fundamental for accurate and meaningful analyses.
-
-\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original.
-These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}
 

### 2024-07-29 report/emse24.tex
feat: emse24 final draft

@@
 \usepackage{booktabs}
-\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
 \usepackage{wrapfig}
@@
 
-\title{Exploring the Role of Feedback in Machine Learning Jupyter Notebooks}
+\title{Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks}
 
@@
 \author{Arumoy Shome\and
-  Lu{\`\i}s Cruz\and
+Lu\`{i}s Cruz\and
   Diomidis Spinellis\and
@@
 \begin{abstract}
-
-  % TODO: update the abstract
-  Developing and testing ML systems is a highly interactive process which demands a human-in-the-loop approach.
-In addition to writing tests for the code base, practitioners are required to analyse and interpret several visualisations using their domain expertise to validate if an ML system satisfies the required set of functional and non-functional properties.
-Visualisations are frequently used to qualitatively assess various parts of an ML pipeline.
-However, implicit knowledge gained from visualisations must be translated to explicit analytical tests that fail when there is change in any component of the ML pipeline.
-We conduct an empirical analysis of Jupyter notebooks to catalogue the state-of-the-art mappings between ML visualisations and assertions.
-We mine Github to collect 54K Jupyter Notebooks that contain assertions written in Python.
-We develop a novel methodology to identify 1764 notebooks which contain an assertion below a visualisation.
-We manually analyse the 1.7K notebooks and identify 269 visualisation-assertion pairs that are semantically related to one another.
-We further investigate the 269 visualisation-assertion pairs and identify three frequently occurring testing patterns.
-We perform an in-depth analysis of 34 visualisation-assertion pairs and find that the assertions often fail to capture all the information present in their visual counter-part.
-Empirical evidence obtained in this study indicates that current software testing methods fail to address the unique challenges of ML.
-And emphasises the need for automated tools that bridge the gap between visual assessments and analytical assertions.
-
+  The machine learning development lifecycle involves iterative and exploratory processes that heavily rely on feedback mechanisms to ensure data and model integrity.
+This study investigates the use of implicit and explicit feedback within Jupyter notebooks.
+We analyse 297.8 thousand Jupyter notebooks from GitHub and Kaggle, identifying key feedback mechanisms such as print statements, outputs from last cell statements, and assertions.
+Our findings reveal extensive use of implicit feedback for critical design decisions and highlight the limited adoption of explicit feedback mechanisms like assertions.
+By conducting detailed case studies on selected feedback instances, we uncover the reliance on manual validation and the potential for automated validation to enhance robustness and maintainability.
+This study underscores the need for improved documentation and testing practices within the ML community, advocating for the integration of feedback mechanisms into the ML development workflow to mitigate technical debt and enhance reproducibility.
 \end{abstract}
 
-\keywords{ML Testing \and SE4AI \and Visualisations \and Assertions \and Computational Notebooks}
+\keywords{Assertions, Feedback Mechanisms, Jupyter notebooks, SE4AI}
 
 \section{Introduction}
+
 The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages.
 Significant effort is invested in gathering and assembling the required data from diverse sources.
-The data then undergo cleaning and preprocessing to ensure high quality and usability, in the subsequent stages.
+The data then undergoes cleaning and preprocessing to ensure high quality and usability, in the subsequent stages.
 MLDL is dominated by the model development phase which is highly experimental and iterative.
@@
 
-% NOTE: consider starting with prints and last statements first (most obvious), then move to assertions; its fine if RQs follow different order
-Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favor of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}.
-During this process, practitioners can use various feedback mechanisms like assertions, print statements, output of code cells and visualisations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and select specific hyperparameters~\cite{head2019managing,chattopadhyay2020whats}.
-Once the code is finalised, it is typically transferred into a script that is run in a production environment.
-Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration.
-To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}.
-In both scenarios, practitioners need to revisit the notebooks to obtain structured information such as up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
-As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
-This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
-
-% TODO: feedback from diomidis
-% - "We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks." Rephrase to have the knowledge gap correspond to what the RQs answer.
-Currently one may think that we aim to improve the approach, not its understanding.
-We identify a knowledge gap in the current approach used to obtain such structured information from experimental code within Jupyter notebooks.
-To address this gap, this study conducts a series of empirical studies to understand the role of various feedback mechanisms used to develop ML pipelines within Jupyter notebooks.
-We focus on implicit form of feedback obtained from print statements and output generated by the last statement in code cells, as well as explicit form of feedback obtained from assertions.
-To the best of our knowledge, we are the first to investigate how such feedback mechanisms are used by ML practitioners when developing ML pipelines inside Jupyter notebooks.
+Notebook users can obtain feedback using implicit or explicit mechanisms.
+Implicit feedback mechanisms require manual validation by the user.
+For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical.
+This can be verified with a print statement in a code cell and manually checking the output.
+This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.
+
+In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met.
+For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \lstinline{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}.
+This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.
+
+As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\cite{kery2018story,rule2018exploration}.
+In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones.
+This transition reduces the reliance on manual verification, thereby enhancing the robustness and maintainability of ML pipelines.
+
+To the best of our knowledge, no prior studies have been conducted to understand how feedback mechanisms are used by practitioners when developing ML pipelines inside Jupyter notebooks.
+To address this gap, we conduct a series of empirical studies to understand the role of various feedback mechanisms within Jupyter notebooks.
+We focus on implicit form of feedback obtained from print statements and output generated by the last statement of code cells, as well as explicit form of feedback obtained from assertions.
 The research questions along with the contributions of this study are presented below.
@@
 
-    We mine 297.8 thousand Jupyter notebooks written in Python from Github and Kaggle.
-We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.
+    We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle.
+We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements.
 We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.
-
-  % TODO: add citation for the dataset (make it public so that DOI is available)
+The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence~\cite{replication}.
 
@@
 
-    To gain a deeper understanding of assert statements written in Jupter notebooks, we conduct individual case-studies with 82 assert statements.
+    To gain a deeper understanding of Python assertions written in Jupter notebooks, we conduct individual case-studies with 82 assert statements.
 We analyse the assertion along with the surrounding code to understand its purpose.
@@
 
-  \item[RQ3.] \textbf{How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}
-
-    Similarly, we conduct individual case-studies with 44 print statements and 27 last statements.
+  \item[RQ3.] \textbf{How is implicit feedback from print statements and last cell statements used when writing ML code in Jupyter notebooks?}
+
+    Similarly, we conduct individual case-studies with 44 print statements and 27 last cell statements.
 During each case-study, we analyse the code statement along with the raw output to understand its purpose.
@@
 
-% TODO revisit this paragraph
-% TODO what should be done to change the things we observe? to increase the use of explicit feedback? do we know if its important or not? maybe we need to explore that?
-Our findings indicate that outputs from code cells are extensively utilised for making critical design and implementation decisions during the MLDL.
+% The above all motivates the research: but then the intro should also ¡®sell¡¯ the outcomes of the research. These are
+% - Identification of over 20 feedback mechanisms, half of them implicit, half of them explicit (your taxonomy)
+% - Insights in actual use of these mechanisms
+% - Recommendations for better support of these mechanisms in notebook environments / life cycle
+
+Our findings indicate that implicit feedback mechanisms are extensively utilized for making critical design and implementation decisions during the MLDL.
 While manual validation of these outputs is common, it is also prone to inconsistencies thus highlighting the need for automated validation mechanisms.
-Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds, thereby reducing reliance on manual checks and enhancing the robustness of ML pipelines.
-However, the study also notes the limited use of assertions, with only 9.7\% of the 3 million notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community.
-The findings underscore the necessity for developing ML-specific testing tools and integrating assertions more deeply into the ML development workflow to ensure consistent and reliable validation processes.
-
-% TODO update replication package
-
-% the replication package for this study is made public on figshare\footnote{the replication package can be found at this url: https://figshare.com/s/9347228011999cba29f5}.
+Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds.
+However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analysed incorporating them, suggesting a significant gap in current testing practices within the ML community.
+The findings underscore the necessity for developing ML-specific testing tools that are deeply integrated into the ML development workflow to ensure consistent and reliable validation processes.
 
@@
 Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI).
-Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualisations, and narrative text.
+Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.
 This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
@@
 Figure~\ref{fig:nb} shows an example Jupyter notebook with markdown and code cells.
-The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}~\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically.
+The data comprised within Jupyter notebooks is stored using the JSON schema specified by \emph{nbformat}\footnote{https://nbformat.readthedocs.io/en/latest/index.html}, and thus can be parsed programmatically.
 A notebook consists of \emph{cells} which can be of three types---``markdown'', ``code'' and ``raw''.
@@
 Only code cells in Jupyter notebooks can generate outputs.
-As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualisation.
+As shown in Figure~\ref{fig:nb}, outputs can come from three main sources in the source code---print statements, the last statement in the code cell, and statements that create a visualization.
 The outputs produced by code cells are stored as JSON objects under the \lstinline[language={}]$outputs$ key in the corresponding JSON representation of the notebook.
@@
 
-% TODO: I suspect this will raise eyebrows because the notebooks collected from Github were for the ICST paper, where we were also looking for the keyword "assert" (which does not apply to this paper)
-
 Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle.
-We used Github's advanced search syntax~\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on Github~\footnote{Collected on June 22, 2023}.
+We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}.
 We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub.
 To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks.
-Consequently, we started with 283GB of data comprising 297.8 thousand Python Jupyter notebooks.
+Consequently, we started with 283 GB of data comprising 297.8 thousand Python Jupyter notebooks.
 
 Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries which we identified from grey literature~\cite{datacampml,gfgml,jetbrainsml,kdnuggetml,courseraml}---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}.
+The libraries are selected based on the import analysis conducted by \emph{Psallidas et al.}~\cite{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub.
 We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells.
-Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a final data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last statements.
+Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
 
@@
 
-We allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last statements.
+For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last cell statements.
 To surface interesting candidates for the case studies, we apply text processing techniques as described below.
 
-The assertions and cell outputs are first tokenised---special characters and alpha-numeric words shorter than two characters are removed.
+The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed.
 Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
-The term frequency (TF) for all tokens is calculated and then normalised using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
+The term frequency (TF) for all tokens is calculated and then normalized using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
 We apply stratified random sampling to identify the candidates for the case study analysis.
-The sub-groups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
+The subgroups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
 The candidates are then divided into quartiles based on the aggregate value.
@@
 The analysis is stopped when the time resource is exhausted.
+
 During each case study, we analyse the code of the candidate to understand its purpose.
 Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.
+For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors.
+Further discussions were held with all authors until a consensus was reached.
 
@@
 % TODO: add citation to online appendix
+
 The results of the study are presented in this section.
-For the case-studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other.
-Please note that this is done only to facilitate structuring the paper in a meaningful manner and for easy of reporting the results.
-Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study.
-Additional details are provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}.
+For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other.
+This is done only to structure the paper in a meaningful manner and for ease of reporting the results.
+Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}.
 Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples.
@@
 \subsection{(RQ1) What feedback mechanisms are employed in the development of ML Jupyter notebooks?}~\label{sec:result-analysis}
+
+% TODO: add the findings from this part of the analysis
 
@@
 Methods such as \lstinline{assert_almost_equal}, \lstinline{assert_array_almost_equal} and \lstinline{AssertAlmostEqual} accommodate the stochastic nature of ML, where results can vary slightly due to inherent randomness in data processing and model training.
-We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas data frames.
+We also find methods such as \lstinline{assert_array_equal} and \lstinline{assert_frame_equal} that are specifically tailored for comparing high-dimensional data structures such as numpy arrays and pandas dataframes.
+
+\highlight{\textbf{Finding 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion.
+95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries.
+Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}
 
@@
     \includegraphics[width=\linewidth]{common-assert-test.pdf}
-    \caption{10 most common tests used in \lstinline{assert} statements.}
+    \caption{10 most common AST nodes in the test attribute of \lstinline{assert} statements.}
     \label{fig:common-assert-test}
@@
   \centering
-  \caption{Examples of the most common tests performed in assert statements.}
+  \caption{Examples of the most common tests performed in \lstinline{assert} statements.}
   \begin{tabular}{@{}m{0.5\textwidth} m{0.5\textwidth}@{}}
@@
 We focus the remainder of the analysis on the 95\% of the assertions written using the \lstinline{assert} statement.
-The most common test performed in the \lstinline{assert} statements include a comparison check and a function call as shown in Figure~\ref{fig:common-assert-test}.
+Figure~\ref{fig:common-assert-test} shows the most frequent AST nodes that occur in \lstinline{assert} statements.
+The figure indicates that the most common test performed in the \lstinline{assert} statements include a comparison check and a function call.
 Table~\ref{tab:common-assert-test} provides a few examples of assertions for each of the tests above.
@@
 We observe that overwhelming majority of the asserts ensure that two values are equal to each other.
-The shape and size of data structures are frequently compared as observed in Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} which show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively.
+Figure~\ref{fig:common-compare-lhs} and Figure~\ref{fig:common-compare-rhs} show the most common bi-grams that appear in the LHS and RHS of the comparison statements respectively.
+The figures collectively show that the shape and size of data structures are frequently checked.
 
@@
 The bi-grams indicate that the shape and data type of data structures are frequently verified.
+
+\highlight{\textbf{Finding 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message.
+These assertions frequently perform a comparison check or call an external function.
+The comparison checks frequently verify that the shape and size of data structures is equal to a specified value.}
 
@@
 Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string.
-Based on these observations, we disect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes).
+Based on these observations, we dissect the print statements into two components---text written in natural language (from the \emph{Constant} nodes) and code (from all other nodes).
 Figure~\ref{fig:common-print-constants} and Figure~\ref{fig:common-print-not-constants} present the most common bi-grams that appear in the natural text and code of print statements respectively.
@@
 
+\highlight{\textbf{Finding 1.3} Print statements are most commonly used to display various performance metrics of ML models.}
+
 \begin{figure}
@@
     \includegraphics[width=\linewidth]{common-print-constants.pdf}
-    \caption{Most common bi-grams that appear in natural text inside of print statements.}
+    \caption{Most common bi-grams that appear in natural text of print statements.}
     \label{fig:common-print-constants}
@@
 
-We collected 1 million last statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study).
-Figure~\ref{fig:common-last-nodes} shows that the most frequent type of code written as the final statement in a code cell is a function call.
-We further analyse the functions calls in more detail.
-Figure~\ref{fig:common-last-modules} and Figure~\ref{fig:common-last-functions} show the most modules and functions used inside the function calls respectively.
-The figures highlight the prominence of data visualisation libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualisation functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks.
+We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study).
+Figure~\ref{fig:common-last-nodes} shows that \emph{Call} nodes frequently occur in the AST of last cell statements.
+Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} show the most common functions used inside these \emph{Call} nodes.
+The figures highlight the prominence of data visualization libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualization functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks.
 The data analysis tool pandas (represented by \lstinline{df} and \lstinline{pd}) also features prominently with many data exploration and preprocessing functions such as \lstinline{head}, \lstinline{value_counts} and \lstinline{describe} being used frequently.
-The presence of \lstinline{train} and \lstinline{model} modules along with the \lstinline{fit} function indicates the use of ML libraries.
+The presence of \lstinline{train} and \lstinline{model} Python objects along with the \lstinline{fit} function indicates the use of ML libraries.
+
+\highlight{\textbf{Finding 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}
 
@@
   \includegraphics[width=0.5\textwidth]{common-last-nodes.pdf}
-  \caption{Most common nodes in the last statement of code cells.}
+  \caption{Most common AST nodes in the last cell statements.}
   \label{fig:common-last-nodes}
@@
 
-% TODO: we need better terminology here
 \begin{figure}
@@
     \includegraphics[width=\linewidth]{common-last-modules.pdf}
-    \caption{Most common modules used in the function calls of last statements.}
+    \caption{Most common Python modules and objects used in the \emph{Call} nodes of last cell statements.}
     \label{fig:common-last-modules}
@@
     \includegraphics[width=\linewidth]{common-last-functions.pdf}
-    \caption{Most common functions used in the function calls of last statements.}
+    \caption{Most common functions used in the \emph{Call} nodes of last cell statements.}
     \label{fig:common-last-functions}
@@
     \centering
-    \includegraphics[width=\linewidth]{common-last-visualisation-functions.pdf}
-    \caption{Most common visualisation functions used in last statements.}
-    \label{fig:common-last-visualisation-functions}
-  \end{minipage}
-\end{figure}
-
-We further analysed the most common functions called on pandas data frames by isolating the last statements where the module is \lstinline{data} or \lstinline{df}.
+    \includegraphics[width=\linewidth]{common-last-visualization-functions.pdf}
+    \caption{Most common visualization functions used in last statements.}
+    \label{fig:common-last-visualization-functions}
+  \end{minipage}
+\end{figure}
+
+We further analysed the most common functions called on pandas dataframes by isolating the last cell statements that contain \lstinline{data} or \lstinline{df} objects.
 This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations.
 The frequent use of functions such as \lstinline{value_counts}, \lstinline{head}, and \lstinline{unique}, indicate an emphasis on understanding the structure and content of the data.
-Visualisation functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualisations directly from data frames.
+Visualization functions like \lstinline{plot} and \lstinline{hist} suggest that practitioners often create quick visualizations directly from dataframes.
 Descriptive statistics are also prominent, with functions like \lstinline{describe}, \lstinline{mean}, and \lstinline{sum} being commonly used.
 
-Similarly, we also analysed the most common visualisation functions used in notebooks by isolating the last statements where the module is \lstinline{sns} or \lstinline{plt}.
-This is shown in Figure~\ref{fig:common-last-visualisation-functions} which indicates that categorical data visualisation appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions.
-The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualising correlations and statistical distributions.
+Similarly, we also analysed the most common visualization functions used in notebooks by isolating the last cell statements where the module is \lstinline{sns} or \lstinline{plt}.
+This is shown in Figure~\ref{fig:common-last-visualization-functions} which indicates that categorical data visualization appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions.
+The prominence of \lstinline{heatmap} and \lstinline{boxplot} functions indicates a regular need for visualizing correlations and statistical distributions.
 For continuous data, functions like \lstinline{distplot}, \lstinline{scatterplot}, and \lstinline{scatter} are commonly used.
-The presence of \lstinline{imshow} suggests that image-related visualisations are also significant in many notebooks.
+The presence of \lstinline{imshow} suggests that image-related visualizations are also significant in many notebooks.
+
+\highlight{\textbf{Finding 1.5} Common functions called on pandas dataframes focus on exploratory data analysis, statistical operations and quick visualizations.
+In visualizations, functions for categorical data, correlation and distribution are frequently used.}
 
@@
 
-Data shape check assertions serve as a crucial safeguard against data inconsistencies, ensuring the integrity and compatibility of the data throughout the machine learning workflow.
-By verifying the expected dimensions and shapes of various data structures, these assertions help prevent runtime errors, ensure accurate model training and evaluation, and maintain the reliability of the entire machine learning pipeline.
-
 \highlight{\textbf{Finding 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
@@
 We found assert statements that perform various validation checks on data structures, such as NumPy arrays and Pandas dataframes.
-These checks help ensure the data meets specific criteria or constraints, preventing potential errors or inconsistencies in downstream operations.
-
 Common data validation checks include verifying the presence of specific values or ranges within arrays or columns(\emph{A41, A65}) and verifying the equality or closeness of arrays to a target value within a specified absolute tolerance (\emph{A44, A46}).
 Some assertions also focus on validating the uniqueness or cardinality of values (\emph{A52}) while others validate the presence of specific values or conditions within data structures (\emph{A73}).
-
-By verifying specific conditions, value ranges, uniqueness, and equality constraints, these assertions help catch potential errors or inconsistencies early in the pipeline, preventing downstream issues and ensuring reliable and accurate results.
 
@@
 
-This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall, F1 score, and others.
+This study finds several assertions used to test model performance against predefined thresholds for key metrics such as accuracy, precision, recall and F1 score.
 Range checks confirm that the performance metrics fall within acceptable boundaries, indicating good predictive performance without overfitting (\emph{A38}).
@@
 Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}).
-Finally assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specifc scenarios like clustering (\emph{A7}).
+Finally, assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specific scenarios like clustering (\emph{A7}).
 
@@
 This method accounts for the stochastic nature of many ML models, where slight variations in performance metrics can occur due to differences in initial conditions, random seed settings, or inherent randomness in algorithms.
-By allowing a small tolerance in the comparison, \lstinline{np.isclose} ensures that the model's performance is consistently close to the expected benchmark, thereby affirming its reliability despite the stochastic elements.
+By allowing a small tolerance in the comparison, \lstinline{np.isclose} ensures that the model's performance is consistently close to the expected benchmark despite the stochastic elements.
 
@@
 
-Resource check assertions are used to ensure the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualisations.
+Resource check assertions are used to ensure the availability and validity of essential resources, preventing runtime errors, and maintaining the integrity of the data, models, and visualizations.
 One set of assertions verify the existence of files on the file system, such as pre-trained models (\emph{A10}) or data files (\emph{A74}).
@@
 A unique aspect of working with notebooks is the ability to re-run cells while experimenting, which can lead to unintended consequences.
-Aassertion \emph{A37} presented in Listing~\ref{lst:A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
+Assertion \emph{A37} presented in Listing~\ref{lst:A37} addresses this scenario by checking if the SVM model has been properly trained before proceeding with further operations.
 This check prevents inconsistent model states from executing cells out of order or multiple times.
-Other interesting assertions unique to computational notebooks include verifying the presence of data in visualisations (\emph{A60}) and explicitly verifying the versions of critical libraries (\emph{A18, A67}).
+Other interesting assertions unique to computational notebooks include verifying the presence of data in visualizations (\emph{A60}) and explicitly verifying the versions of critical libraries (\emph{A18, A67}).
 
@@
 Type checks are commonly performed to ensure that the input features are PyTorch float tensors or NumPy arrays, a requirement for many deep learning models and scientific computing operations respectively (\emph{A2, A88}).
-Type checks are also applied to very the type of the ML models (\emph{A40}), data structures and intermediate objects (\emph{A35}) and individual columns or elements within a dataset (\emph{A81}).
+Type checks are also applied to verify the type of the ML models (\emph{A40}), data structures and intermediate objects (\emph{A35}) and individual columns or elements within a dataset (\emph{A81}).
 
@@
 When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigour is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalisation (\emph{A25}).
+This rigour is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalization (\emph{A25}).
 These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles.
@@
 
-In the context of neural networks, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilisation.
-Batching stabilises the learning curve by updating the model weights based on the average gradient of a batch thereby optimising learning and enhancing generalisability of the model.
+In the context of neural networks, the practice of using batch processing is a key strategy to enhance computational efficiency and hardware utilization.
+Batching stabilizes the learning curve by updating the model weights based on the average gradient of a batch thereby optimizing learning and enhancing generalizability of the model.
 Ensuring the batch size matches the hardware capacity is crucial to avoid memory overflow issues that can halt training.
@@
 This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
-Checks include the use of assertions to verify that the output and input dimensions of consecutive convolutional layers match (\emph{A11}), checking that shape of the activations have the right number of channels (\emph{A62}) and that the appropriate regularisation parameter is set (\emph{A75}).
+Checks include the use of assertions to verify that the output and input dimensions of consecutive convolutional layers match (\emph{A11}), checking that shape of the activations have the right number of channels (\emph{A62}) and that the appropriate regularization parameter is set (\emph{A75}).
 Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible with each other.
@@
 Ensuring the absence of data leakage between training and validation datasets is a fundamental best practice in ML to prevent overfitting.
-To guarantee that the model can generalise effectively to new examples, it is crucial to verify that the training and validation sets are completely distinct with no shared examples.
-By strictly separating these datasets, the validation phase provides an unbiased evaluation of the model's ability to generalise to data similar to, but not identical to that which it was trained on.
+To guarantee that the model can generalize effectively to new examples, it is crucial to verify that the training and validation sets are completely distinct with no shared examples.
+By strictly separating these datasets, the validation phase provides an unbiased evaluation of the model's ability to generalize to data similar to, but not identical to that which it was trained on.
 This is demonstrated by Assertion \emph{A33} in Listing~\ref{lst:A33} ensures that there are no overlapping \lstinline{PetID} values between the training set and the validation set, thereby preventing any possibility of data leakage.
@@
 \highlight{\textbf{Finding 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
-Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalise to new data.
+Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
 This practice maintains the integrity and validity of the model evaluation metrics.}
@@
 \subsection{(RQ3) How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-implicit}
-% TODO: feedback from Luis
-% -  up to this point I miss how exactly we plan to answer this question.
-Reading the results I get that we are answering by clustering outputs together based on their respective ML/DS task and then by analysing how that task is accomplished.
-We need to explain this either in the beginning of section 4.1 or in section 3
 
@@
 % (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
-% TODO: feedback from Luis
-% - if needed, we can merge Excution time check here. (both execution time and model correctness can be presented as a KPI)
 
@@
 The model is then re-trained to check if these changes lead to performance improvements.
-Besides the \emph{accuracy} of the model (\emph{P3, P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn~\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
-In addition, we find the use of heatmaps (\emph{L3}) to visualise the confusion matrix of a multi-label classification task and custom functions (\emph{L52}) to manually evaluate an image classification model on a random sample of input images.
+Besides the \emph{accuracy} of the model (\emph{P3, P18}), we also see checks for the \emph{Root Mean Square Error (RMSE)} (\emph{P6}) and the use of the classification report (\emph{P50}) provided by scikit-learn\footnote{https://scikit-learn.org/stable/modules/generated/sklearn.metrics.classification\_report.html} which reports the \emph{accuracy}, \emph{precision}, \emph{recall} and \emph{f1 score} for all labels present in the target feature.
+In addition, we find the use of heatmaps (\emph{L3}) to visualize the confusion matrix of a multi-label classification task and custom functions (\emph{L52}) to manually evaluate an image classification model on a random sample of input images.
 
 \highlight{\textbf{Finding 3.1} The study identifies 33 outputs used to check the performance of trained ML models.
-These include outputs that print key performance metrics, a normalised heatmap used to visualise the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
+These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
 These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
@@
 Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis.
-For instance, visualisations can efficiently determine if scaling, normalizing, or handling outliers is needed.
-During the exploratory data analysis phase, visualisations and pandas dataframes are commonly used to assess the distribution of specific columns (\emph{L9}), helping identify features related to the target variable (\emph{L2, L14})and informing feature inclusion in model training.
+For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
+During the exploratory data analysis phase, visualizations and pandas dataframes are commonly used to assess the distribution of specific columns (\emph{L9}), helping identify features related to the target variable (\emph{L2, L14})and informing feature inclusion in model training.
 Additionally, descriptive statistics are used post-transformation to manually verify the effectiveness of data transformation steps and ensuring that the data conforms to the expected format for further analysis (\emph{L48, L25}).
 
-\highlight{\textbf{Finding 3.2} The study identifies seven outputs used to check data distribution, including visualisations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
-These checks are used for making informed decisions about data transformations such as scaling, normalising and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
+\highlight{\textbf{Finding 3.2} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
+These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
 
@@
 
-During training it is common practice to monitor the progress periodically and use this feedback to adjustment training parameters or stop training early to prevent overfitting.
+During training, it is common practice to monitor the progress periodically and use this feedback to adjustment training parameters or stop training early to prevent overfitting.
 This can be done by checking the best parameters found through tuning methods (\emph{L42}), fitting the model to the training data (\emph{L31}) or printing the training loss and accuracy to give real-time feedback on the learning efficacy of the model (\emph{L8}).
@@
 Missing values interrupt these calculations, leading to errors or the inability to execute algorithms entirely.
-These checks can be performed in code (\emph{P74, L36}) and using visualisations such as a heatmap (\emph{L12}).
+These checks can be performed in code (\emph{P74, L36}) and using visualizations such as a heatmap (\emph{L12}).
 
@@
 Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.
-Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships amongst features in the data.
-In addition to the code, the raw output of the cell is also shown since both case-studies produce a visualisation.
+Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships among features in the data.
+In addition to the code, the raw output of the cell is also shown since both case-studies produce a visualization.
 In \emph{L6}, the practitioner assesses the linearity between the features \texttt{Cash} and \texttt{SIZE}.
-\emph{L10} is a visualisation of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
+\emph{L10} is a visualization of the feature \texttt{X4} alongside the target variable \texttt{Y}, accompanied by a fitted linear regression model.
 
 \highlight{\textbf{Finding 3.8} The study identifies two outputs used to verify linear relationships between features in the data.
-Visualisations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
+Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
 These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
@@
 The process of checking data types helps in identifying the necessary pre-processing or transformation steps to convert the data into a form that is compatible with these models.
-For example, numerical features often require normalisation to bring them within a specific scale and categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
+For example, numerical features often require normalization to bring them within a specific scale and categorical features must be converted into a numerical format through methods like one-hot encoding to ensure they can be effectively integrated into the model.
 
 \highlight{\textbf{Finding 3.9} The study identifies two outputs used to verify the data types of features in the dataset.
-These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training.
+These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training.
 Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
@@
 Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}.
-Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search.
+Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyperparameter configurations using a grid-search.
 
@@
 
-Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle.
-Making such decisions however, requires manual validation of the outputs.
-For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training.
-Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks.
-Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself.
-Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated and explicit validation mechanisms.
-
-Assertions can play a crucial role in automatically validating critical assumptions in ML workflows.
-As seen in the results from RQ2, assertions can verify that data distributions remain consistent with expectations or that model performance metrics meet predefined thresholds.
-Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged, reducing the reliance on manual verification and enhancing the robustness and maintainability of ML pipelines.
+Drawing from the results of this study, we discuss implications and future work in this section.
+
+\subsection{Documentation and Technical Debt}
+
+Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}.
+Once the code is finalized, it is typically transferred into a script that is run in a production environment.
+Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration.
+To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}.
+In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
+
+We believe that such structured information can be derived from the feedback mechanisms available in Jupyter notebooks.
+Our results show that practitioners use assertions, print statements, output of code cells and visualizations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and check for hardware and software dependencies.
+As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
+This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
+
+Our results from RQ3 reveal that implicit feedback from print and last cell statements are extensively used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
+Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.
+
+We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations.
+Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\cite{heer2010tour}.
+By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
+Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\cite{yang2021subtle}.
+
+\subsection{Automated Data Validation}
+
+Catching data errors are critical since ML models trained using dirty data will lead to poor and incorrect predictions.
+This is a cause for concern since these models may be deployed in production and their output may be used by other services which will then also perform poorly.
+The output of the ML models are often also used to capture new training data, thus leading to a feedback loop which over time will degrade the model performance~\cite{sculley2015hidden,breck2019data}.
+
+\emph{Breck et al.
+(2017)}~\cite{breck2017ml} view ML training as analogous to compilation and recommend that training data be tested, similar to code in traditional software.
+They propose a rubric to assess the production readiness of an ML pipeline, which includes several recommendations for data testing.
+In subsequent work, \emph{Breck et al.
+(2019)}~\cite{breck2019data} introduce \emph{TensorFlow Data Validation}, an automated tool for data validation that works on principles borrowed from the database community.
+Data expectations are specified in a schema, and each new data batch is validated against this schema to ensure it meets the expected criteria or prompts schema updates if requirements change.
+
+The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows.
+Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged.
+This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines.
 Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development.
@@
 Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
-Moreover, they enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells.
-
-% P71 highlights the lack of proper software engineering best practices within Jupyter notebooks.
-This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}.
-In this case study, managing dependencies through external tools is recommended.
-Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions.
-Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.
-
-% NOTE: here we need to bring in prior literature on ML testing; and how majority of the research is working on testing using a test dataset; however we also have data and models
-% NOTE: 3 pillars of of ML (sato2019continuous)
-
-% TODO: feedback from Diomidis
-% - "it represents an emerging testing practice within the ML community": I have commented before that assertions are different from tests.
-Assertions are a mechanism to increase the code's robustness, similar to type checks, array bounds checks, and thrown API exceptions.
-
-From the 3 million Jupyter notebooks collected in this study, only 30 thousand (9.7\%) were found to utilize assertions.
-This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community.
-The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.
-
-We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties.
-Current guidelines on ML coding practices require enhancement to emphasize the importance of testing.
-There is a pressing need to steer education and research in Software Engineering towards addressing the specific challenges faced when developing ML-enabled software systems.
-Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
-
-The assertions presented in RQ2 and RQ3 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
+Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\cite{wang2020assessing}.
+
+In the results from RQ2 and RQ3, we observe that many assertions, print statements and last cell statements are used in the upstream, data intensive stages of the MLDL.
+When integrating dedicated data validation tools into an operational ML pipeline, interactions with the feedback mechanisms in notebooks can aid in writing the data validation schema.
+For instance, this study finds several assertions that perform data validation checks to ensure that data meets specific criteria or constraints.
+Similarly, we also find assertions, prints statements and last cell statements that performs type, shape and distribution checks that serve as safeguards against data inconsistencies.
+
+\subsection{ML Testing Literacy}
+
+Despite the potential benefits, our results from RQ1 reveal that assertions are not widely used in Jupyter notebooks.
+The relatively low adoption rate compared to implicit feedback mechanisms, suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.
+
+We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties~\cite{simmons2020large-scale}.
+Guidelines and curricula on ML coding practices have historically focused on building better models.
+We argue that the focus must shift towards building reliable ML-enabled systems by incorporating decades of research from the software engineering community with ML.
+Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\cite{tudremla,mlinprod,kastner2020teaching,cain2024,icse2025} Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
+
+The assertions analysed in RQ1 and RQ2 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
 First, the assertions identified in this study lack the support of dedicated testing libraries.
@@
 
-Visualizations are an effective tool for interpreting complex ML models and data transformations.
-However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}.
-We argue that the need for translating visual insights into formal assertions is more pressing.
-The visual cue and trends observed in visualisations are open to interpretation, thus the interpretation of the same plot may differ amongst two practitioners.
-Translating visual insights into assertions ensures that the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.
-
-In the VA pairs from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part.
-This can lead to a mismatch of expectations in future versions of the ML software.
-For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied.
-The VA pairs presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial.
-Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer.
-For example, a linear correlation between the true and predicted labels is evident in VA107 presented in Table~\ref{tab:va-model-perf}.
-Yet, its formal assertion necessitates defining an appropriate threshold for the coefficient of determination metric.
-Our results show that current software testing practices fail to capture the idiosyncrasies of ML code.
-There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.
-
-% TODO: integrate our NLBSE paper
-Our results show a clear need for developing ML-specific testing tools.
-Such tools should seamlessly integrate with computational notebooks and provide automated suggestions for assertions can greatly enhance the reliability and robustness of ML models.
-These tools can also aid in documenting and standardizing the use of domain-specific "magic" thresholds, ensuring that they are consistently applied and maintained.
-
-% NOTE: we find overlap between the manual tests performed using the output of code cells and the assertions.
-
-% NOTE: bring in prior work on reproducibility, code quality here
-% NOTE: notebooks have been adopted by a much larger scientific community; not eveybody has a background in software engineering; we need more tools to aid notebook users...
-
-\section{Implications}
-
-% TODO: feedback from Diomidis
-% - The implications section focuses implicitly on assertions, at the expense of the other elements discussed earlier.
-Also, it puts excessive emphasis on testing, which IIUC is not the paper's topic.
-
-\paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks.
-Assertions play a crucial role in validating critical assumptions automatically, reducing the reliance on manual checks, and enhancing the robustness and maintainability of ML pipelines.
-By embedding assertions, developers can document the decisions made during model development, ensuring that these decisions are consistently applied and preserved throughout the ML lifecycle.
-This practice not only facilitates smoother transitions and handovers in collaborative environments but also improves the reproducibility of ML experiments by ensuring consistent validation checks.
-
-\paragraph{\textbf{For Researchers}} The use of assertions in Jupyter notebooks, though currently uncommon, is an emerging testing practice within the ML community.
-This study highlights the potential for assertions to enhance the reliability and robustness of ML models by automating the validation of critical assumptions.
-Researchers are encouraged to explore the development and evaluation of ML-specific testing methodologies that leverage assertions.
-This could involve creating new testing frameworks and tools tailored to the unique challenges of ML systems, thus contributing to the advancement of software engineering practices in the ML domain.
-
-\paragraph{\textbf{For Notebook Tool Developers}} There is a clear need for automated tools that suggest assertions based on the outputs of code cells in Jupyter notebooks.
-Tool developers can play a pivotal role by creating ML-specific testing tools that seamlessly integrate with computational notebooks.
-These tools should provide automated suggestions for assertions, helping practitioners to translate visual insights into formal, executable checks.
-By bridging the gap between qualitative evaluation and analytical assertions, such tools can significantly enhance the reliability and robustness of ML models, ensuring that they meet predefined performance standards consistently.
-
-\paragraph{\textbf{For Educators}} Educators should emphasize the importance of testing ML code written in Jupyter notebooks as part of the curriculum.
-By incorporating testing practices and the use of assertions into ML education, students can develop a deeper understanding of the importance of automated validation in ensuring model reliability and robustness.
-Teaching these best practices early on can help cultivate a culture of rigorous testing among future ML practitioners, preparing them to address the specific challenges faced in ML-enabled software systems.
-
-\paragraph{\textbf{For General Notebook Users}} Awareness of software engineering best practices, including the use of assertions, is crucial for all users of Jupyter notebooks.
-Assertions serve as an executable form of documentation, capturing the critical assumptions and decisions made during the development process.
-By adopting assertions, users can ensure that their notebooks are more robust, maintainable, and reproducible.
-This practice helps in maintaining consistency and reliability, especially in collaborative projects where multiple users might work on the same notebook, ensuring that the assumptions and validations are transparent and consistently applied.
+% Our observations also indicate that it is much easier for practitioners to use implicit checks and manually validate data rather than translating these checks into assertions.
+This highlights the need for more automated tools that can help practitioners convert implicit checks into explicit assertions, thereby improving the robustness and maintainability of ML pipelines.
+
+\subsection{Machine Learning Bugs}
+
+The findings of this study align with prior research on real-world defects in ML systems.
+\emph{Humbatova et al.}~\cite{humbatova2020taxonomy} propose a taxonomy of faults in DL systems, while \emph{Morovati et al.}~\cite{morovati2023bugs} curate \emph{Defects4ML}, a dataset consisting of ML bugs.
+Both studies identify defects arising from the use of incorrect data types and shapes, deprecated APIs, and missing GPU resources.
+Our study identifies several explicit assertions that can catch such data errors.
+Additionally, we find several implicit checks performed using print and last cell statements to verify the presence of compute resources and libraries, monitor model training, and check the architecture of neural networks.
+
+This study further reveals several avenues for research.
+We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them.
+Further research is required to explore the feedback mechanisms in more detail and organize the knowledge into a taxonomy based on specific ML validation tasks.
+Future work can focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \emph{Humbatova et al.}~\cite{humbatova2020taxonomy} and \emph{Morovati et al.}~\cite{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
+
+\subsection{Code Quality and Conformance to Standards}
+
+Results from RQ1 show that majority of the assertions do not contain a failure message.
+This can lead to problems such as unclear error diagnostics, difficulty in debugging, and increased time to identify and resolve issues.
+\emph{Vidoni et al.}~\cite{vidoni2021evaluating} investigated the testing culture in open-source R packages and identified several anti-patterns that lead to the accumulation of testing technical debt.
+The dataset contributed by this study can be used by researchers to replicate the \emph{Vidoni et al.}~\cite{vidoni2021evaluating} study using assertions written in Python.
 
@@
 
-The empirical analysis presented in this study uses Jupyter notebooks mined from Github.
-To the best of our knowledge,~\emph{Pimentel et al} and~\emph{Quaranta et al} are the only two prior studies which provide a public dataset of Jupyter Notebooks.
-We faced several technical challenges while trying to access the notebooks provided by these replication packages and thus decided to mine our own data from Github.
-However, 24\% of the notebooks had to be excluded from the study since they were not related to ML or DS.
-We wish to extend our analysis to notebooks collected from Kaggle since it is predominantly targeted towards the ML and DS community.
-
-In Section~\ref{sec:data-collect}, we intentionally keep the search criteria on Github broad by only checking for the appropriate file type and the presence of the ``assert'' keyword.
-Besides being used to define tests in the Python programming language, ``assert'' also appears in several external python testing libraries (for instance nose~\footnote{http://readthedocs.org/docs/nose/}, unittest~\footnote{https://docs.python.org/3/library/unittest.html} and numpy~\footnote{https://numpy.org/doc/stable/reference/routines.testing.html}).
-Our approach captures a rich variety of assertion statements while preventing the need to craft a custom search query based on an exhaustive list of keywords that appear in python testing libraries.
+\paragraph{\textbf{Internal Validity}} This study focuses on Jupyter notebooks that focus on machine learning.
+We do this by only considering notebooks that utilize popular ML libraries.
+Consequently, we may miss notebooks that are ML-related but do not use these specific libraries.
+Additionally, notebooks that import popular ML libraries but do not actively use them in the code could lead to misclassification.
+To mitigate this threat, we specifically look for ML libraries that facilitate model training rather than more general-purpose libraries such as pandas, numpy, and scipy, which can be used for a variety of non-ML tasks.
+
+\paragraph{\textbf{External Validity}} This study examines Jupyter notebooks written in Python, specifically focusing on machine learning code.
+The analysis is based on notebooks publicly available on GitHub and Kaggle.
+Consequently, the findings may not be generalizable to Python notebooks that do not concentrate on ML.
+Additionally, the results may not be applicable to notebooks authored in other statistical programming languages (such as R or Julia), or to proprietary notebooks hosted on different platforms.
 
@@
 
-This section summarises prior scientific work that has been done in ML
-Testing, computational notebooks and visual analytics. The section
-concludes with an overview of technical knowledge required for this
-paper.
-
 \subsection{Assertions and Software Bugs}
- TODO
+
+\emph{Kochhar et al.
+(2017)}~\cite{kochhar2017revisiting} conducted a partial replication of a prior study on assertion usage, expanding the analysis to 185 Java projects from GitHub.
+\emph{Vidoni (2021)}~\cite{vidoni2021evaluating} conducted a comprehensive study on the unit testing practices within 177 systematically selected open-source R packages hosted on GitHub.
+The research addressed the quality of testing, identified testing goals, and pinpointed potential sources of testing technical debt.
+
+\emph{Humbatova et al.
+(2020)}~\cite{humbatova2020taxonomy} developed a comprehensive taxonomy of faults in deep learning systems through the manual analysis of 1059 artefacts from GitHub and Stack Overflow, complemented by structured interviews with 20 researchers and practitioners.
+The study categorized faults into five top-level categories, encompassing model, data, training, deployment, and infrastructure-related faults.
+\emph{Morovati et al.
+(2023)}~\cite{morovati2023bugs} addressed the challenge of assessing the quality and performance of tools designed to improve the reliability of machine learning components by creating a standard fault benchmark called defect4ML.
+This benchmark consists of 100 reproducible bugs reported by ML developers on GitHub and Stack Overflow, focusing on the TensorFlow and Keras ML frameworks.
+
+In contrast to prior work, our research focuses on the feedback mechanisms used within Jupyter notebooks for ML development.
+While previous work has provided insights into assertion usage and fault taxonomies, our study specifically examines the role of implicit and explicit feedback in ensuring data and model integrity within interactive development environment provided by Jupyter notebooks.
 
@@
 
-\emph{Psallidas et al}~\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
+\emph{Psallidas et al.}~\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
 Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
-\emph{Pimentel et al}~\cite{pimentel2019large-scale} mined 1.4 million notebooks from Github to conduct an empirical study on the coding practices in computational notebooks.
+\emph{Pimentel et al.}~\cite{pimentel2019large-scale} mined 1.4 million notebooks from GitHub to conduct an empirical study on the coding practices in computational notebooks.
 Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \emph{Quaranta et al}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+To enable future research on computational notebooks, \emph{Quaranta et al.}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
 The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
@@
 Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualisations~\cite{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualizations~\cite{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
 Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
@@
 
-To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al.}~\cite{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
 Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle,bavishi2021vizsmith}.
@@
 Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large-scale,quaranta2021kgtorrent}.
-Computational notebooks provide a rich source of data in the form of natural text, code and visualisations.
+Computational notebooks provide a rich source of data in the form of natural text, code and visualizations.
 Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
-This study leverages the computational narrative present within notebooks to identify visualisations and assertions that are semantically linked to each other.
+This study leverages the computational narrative present within notebooks to identify visualizations and assertions that are semantically linked to each other.
+
+Our study builds on this foundation by focusing on the specific feedback mechanisms used within Jupyter notebooks during ML development.
+We examine both implicit and explicit feedback methods, providing a detailed analysis of their usage and effectiveness in maintaining robust ML workflows.
 
 \subsection{ML Testing}\label{sec:ml-testing}
-
-% Testing ML enabled software systems poses more challenges compared to their traditional counter part.
-While traditional software systems mature through change in their codebase, ML enabled systems additionally experience change in the training data and the machine learning model~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
-With ML being adopted into safety-critical domains that can affect human lives, ensuring that ML systems are correct, robust to data perturbations and not biased by race, colour or gender is of paramount importance.
 
@@
 First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2020machine,mehrabi2021survey,chen2022fairness}.
+And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2022machine,mehrabi2021survey,chen2022fairness}.
 
@@
 
-In contrast to existing scientific contributions, we conduct a large-scale empirical analysis of Jupyter Notebooks to understand the process of testing ML systems.
-Specifically, we focus on how visualisations are used to test specific properties of ML systems.
-We further investigate how frequently analytical tests are formulated based on the information gained from the visualisations.
-
-\subsection{Visualisations and Machine Learning}\label{sec:visualisations}
+Despite these advancements, there is a notable gap in practical, day-to-day testing practices within ML development environments.
+Our study addresses this gap by analysing the feedback mechanisms in Jupyter notebooks, highlighting how assertions and print statements are used to validate data and model performance, and proposing ways to integrate these practices more effectively into the ML development lifecycle.
+
+\subsection{Visualizations and Machine Learning}\label{sec:visualizations}
 
 As ML augment software systems in safety-critical domains, emphasis has been put into explainability of ML models.
-ML models and the underlying data is complex and multi-dimensional.
+ML models and the underlying data is complex and multidimensional.
 To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2019what-if}.
@@
 Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\emph{Yuan et al} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\emph{Hohman et al} conduct a survey of visual analytics techniques for Deep Learning Models.
+\emph{Yuan et al.}~\cite{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\emph{Hohman et al.}~\cite{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
 The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
@@
 Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\emph{Wexler et al} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
-To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al} propose \textit{ModelTracker}.
+\emph{Wexler et al.}~\cite{wexler2019what-if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
+To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al.}~\cite{amershi2019software} propose \textit{ModelTracker}.
 Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialised use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalised Additive Models, prioritising machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
-
-In contrast to prior work which propose dedicated visual analytics tools, this study focuses on visualisations within computational notebooks.
-We focus on visualisations produced by AI practitioners using Python libraries (such as Matplotlib or Seaborn).
-Additional context provided by surrounding code and markdown cells are used to understand the motivation for creating the visualisation and the information gained from it.
-
-\section{Conclusion and Future Work}
-
-% NOTE: taxonomy of assertions mapping to V&V tasks in ML
-% NOTE: automated tool that adds assertions once the contents of the NB is finalised?
-
-This study empirically supports the efficacy of visualisations for testing various properties of machine learning systems.
-Visualisations enhance the ability of ML practitioners to interpret complex information through visual cues.
-However, decisions based on such visualisations demand manual verification in each development iteration.
-Analytical assertions derived from ML visualisations can reduce this manual effort by encapsulating key insights, and ensuring continuity in interpretation across different development stages and project collaborators.
-
-Our results indicate that crafting analytical assertions from ML visualisations, is an emerging testing technique.
-Majority of the assertions however, are found to be primitive and unable to convey all information discerned from their visual counter-parts.
-This study finds a gap in the current software testing methods to address the unique challenges of ML.
-And emphasises the need for automated tools to bridge visual assessments and analytical assertions.
-
-There is a clear need for tools designed to aid ML practitioners when testing ML systems.
-These tools should provide assistance in identifying the most suitable visualisations for testing specific properties of ML systems.
-Customized solutions that streamline the process of connecting visual insights to analytical assertions would significantly benefit practitioners in this domain.
-
-The manual translation of insights derived from visualisations into analytical assertions presents a substantial challenge.
-Future work should focus on the development of automated tools that can assist AI practitioners in this translation process.
-Generating assertions from visual information automatically, will promote adoption of assertions in ML notebooks.
-And bring the code developed within notebooks, closer to a production-ready codebase.
-
-% TODO remove this is no space left
-Extending ML testing to dynamic real-world scenarios should be explored further.
-This entails developing techniques and frameworks that allow assertions to adapt to evolving data distributions, changing model requirements, and continuous system updates.
-Research in this area will contribute to the development of ML systems that remain reliable and relevant throughout their operational lifecycle.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialized use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalized Additive Models, prioritizing machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
+
+Unlike prior work that focuses on standalone visual analytics tools, our study explores how visualizations within Jupyter notebooks are used by ML practitioners.
+We analyse the motivations behind creating these visualizations and the insights they provide, offering a nuanced understanding of their role in the iterative ML development process.
+
+\section{Conclusion}
+
+This study provides a comprehensive analysis of feedback mechanisms employed within Jupyter notebooks during the machine learning development lifecycle.
+Our findings indicate that while implicit feedback mechanisms such as print statements and last cell outputs are extensively utilized, there is a significant gap in the adoption of explicit feedback mechanisms like assertions.
+This reliance on manual validation highlights the potential for automated data validation to enhance the robustness and maintainability of ML pipelines.
+Additionally, our research underscores the necessity for better documentation practices to mitigate technical debt and ensure the reproducibility of ML experiments.
+By integrating feedback mechanisms into the ML development workflow, practitioners can achieve more reliable and efficient model development processes.
+Future work should focus on developing specialized tools and frameworks to support these practices, thereby advancing the field of ML engineering.
 

### 2024-07-29 docs/index.qmd
feat: init online appendix

---
title: Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks
subtitle: Online Appendix
abstract: |
  This website hosts the online appendix for the research paper titled
  *Understanding Feedback Mechanisms in Machine Learning Jupyter
  Notebooks* which has been submitted to the Journal of Empirical
  Software Engineering.
toc: true
format:
  html:
    theme: cosmo
website:
  site-url: "https://arumoy.me"
  back-to-top-navigation: true
page-footer:
  center: "All materials on this website are open sourced under the CC-BY 4.0 license."
---

# (RQ2) How is explicit feedback from assert statements used to validate ML code written in Jupyter notebooks?
## Data Shape Check ($N = 26$)

| Key | Code                                                                                                        |
|-----|-------------------------------------------------------------------------------------------------------------|
| A5  | `assert y_valid.shape == (1132,)`                                                                           |
| A17 | `assert X.shape[1] == 13, 'Did you drop/lose some columns in X? Did you properly load and split the data?'` |
| A29 | `assert len(test_y_preds) == len(test_y), 'Unexpected number of predictions.'`                              |
| A31 | `assert img.shape == (112, 92)`                                                                             |
| A76 | `assert len(encoding['token_type_ids']) == max_seq_length`                                                  |
| A84 | `assert red.get_shape().as_list()[1:] == [224, 224, 1]`                                                     |
| A90 | `assert len(X_train) == 2000`                                                                               |
| A93 | `assert temp_embed.shape[0] == stride`                                                                      |

## Data Validation Check ($N = 14$)

| Key | Code                                                                                |
|-----|-------------------------------------------------------------------------------------|
| A41 | `assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))`                   |
| A44 | `assert np.all(np.unique(X['smoke'].values) == np.array([0, 1]))`                   |
| A46 | `assert np.isclose(stdev_norm, 1.0, atol=1e-16)`                                    |
| A52 | `assert grouped_users['user_id'].nunique() == user_engagement['user_id'].nunique()` |
| A65 | `assert np.all(y <= nb_classes)`                                                    |
| A73 | `assert df['clf'].value_counts()[1] == len(df[df['quality'] >= 7])`                 |

## Model Performance Check ($N = 11$)

| Key | Code                                                                                                                     |
|-----|--------------------------------------------------------------------------------------------------------------------------|
| A7  | `assert len(neighbours_1) == 20, "Neighbors don't match!"`                                                               |
| A15 | `assert np.allclose(verify('images/camera_1.jpg', 'bertrand', database, FRmodel), (0.54364836, True))$`                  |
| A19 | `assert np.allclose(linear_model.coef_, [[1.57104472, 0.92521608]]), 'The model parameters you learned seem incorrect!'` |
| A38 | `assert 0.75 < auc(fpr, tpr) < 0.85`                                                                                     |
| A58 | `assert np.isclose(accuracy, 0.9666666666666667)`                                                                        |

## Existence Check ($N = 8$)

| Key | Code                                                                                 |
|-----|--------------------------------------------------------------------------------------|
| A23 | `assert np.all(orders.groupby('user_id') .days_since_prior_order.tail(1).notnull())` |
| A42 | `assert not lab_s.isnull().values.any()`                                             |
| A43 | `assert len(data) != 0, 'cannot divide by zero'`                                     |
| A50 | `assert not np.any(np.isnan(X))`                                                     |
| A51 | `assert data.target.notnull().all()`                                                 |
| A63 | `assert X.isnull().sum().sum() == 0`                                                 |
| A79 | `assert not processed_data_df.isna().any().any()`                                    |
| A86 | `assert p0 in poi_info.index`                                                        |

## Resource Check ($N = 7$)

| Key | Code                                                                                                                                            |
|-----|-------------------------------------------------------------------------------------------------------------------------------------------------|
| A10 | `assert le_path.is_file(), f"Label encoder file not found at {le_path}. Make sure 'label_encoder.pkl' exists in the lightning_logs directory."` |
| A14 | `assert self.model is not None, 'Model is not loaded, load it by calling .load_model()'`                                                        |
| A18 | `assert pd.__version__.rpartition('.')[0] == '1.0', f"Unexpected pandas version: expected 1.0, got {pd.__version__.rpartition('.')[0]}"`        |
| A37 | `assert svm.fit_status_ == 0, 'Forgot to train the SVM!'`                                                                                       |
| A60 | `assert f2.gca().has_data()`                                                                                                                    |
| A67 | `assert pm.__version__ == '3.9.2'`                                                                                                              |
| A74 | `assert os.path.exists(image_dir)`                                                                                                              |

## Type Check ($N = 5$)

| Key | Code                                                                                          |
|-----|-----------------------------------------------------------------------------------------------|
| A2  | `assert isinstance(X_trn, torch.FloatTensor), 'Features should be float32!'`                  |
| A35 | `assert isinstance(column_transformer, ColumnTransformer), "Input isn't a ColumnTransformer"` |
| A40 | `assert isinstance(model_3, sklearn.ensemble.RandomForestClassifier)`                         |
| A81 | `assert is_all_ints(filled_df[r]) is True`                                                    |
| A88 | `assert isinstance(betas, np.ndarray)`                                                        |

## Mathematical Property Check ($N = 4$)

| Key | Code                                                                     |
|-----|--------------------------------------------------------------------------|
| A3  | `assert (xH - wH) % self.stride == 0`                                    |
| A25 | `assert test_output.std() < 0.15, "Don't use batchnorm here"`            |
| A56 | `assert np.allclose(e_v_states[:, -1], np.ones_like(e_v_states[:, -1]))` |
| A64 | `assert np.allclose(T, T.T)`                                             |

## Batch Size Check ($N = 3$)

| Key | Code                                                                                                                                        |
|-----|---------------------------------------------------------------------------------------------------------------------------------------------|
| A21 | `assert x.size(0) % batch_size == 0, f'the first dimension of input tensor ({x.size(0)}) should be divisible by batch_size ({batch_size})'` |
| A28 | `assert image_size % patch_size_small == 0, 'Image dimensions must be divisible by the patch size.'`                                        |
| A70 | `assert n_img > batch_size`                                                                                                                 |

## Network Architecture Check ($N = 3$)

| Key | Code                                                                                   |
|-----|----------------------------------------------------------------------------------------|
| A11 | `assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()` |
| A62 | `assert self.encoder_conv_01[0].weight.size() == self.vgg16.features[2].weight.size()` |
| A75 | `assert reg in ['none', 'l2']`                                                         |

## Data Leakage Check ($N = 1$)

| Key | Code                                                                                |
|-----|-------------------------------------------------------------------------------------|
| A33 | `assert len(set( tr_df.PetID.unique()).intersection(valid_df.PetID.unique())) == 0` |

# (RQ3) How is implicit feedback from print statements and last cell statements used when writing ML code in Jupyter notebooks?

## Model Performance Check ($N = 33$)

| Key | Code                                                                                                 |
|-----|------------------------------------------------------------------------------------------------------|
| P3  | `print('The mean accuracy with 10 fold cross validation is: %s ' % round(scores * 100, 2), '%')`     |
| P6  | `print('RMSE:', np.sqrt(metrics.mean_squared_error(y_test, pred)))`                                  |
| P18 | `print('The Accuracy is:', accuracy_score(y_test, y_pred))`                                          |
| P50 | `print('Classification Report: SVM (validation data)')$`                                             |
| P54 | `print('Intercept value:', lm.intercept_)`                                                           |
| L3  | `skplt.metrics. plot_confusion_matrix(Y_val, Vote.predict(X_val), normalize=True, figsize=(10, 10))` |
| L52 | `spot_check_recs(classifier, 910)`                                                                   |

## Data Distribution ($N = 7$)

| Key | Code                                                                                |
|-----|-------------------------------------------------------------------------------------|
| L2  | `_ = sns.catplot(x='category_id', y='likes', data=train, height=5, aspect=1.5)`     |
| L9  | `sns.kdeplot(data=data.loc[ data['Survived'] == 0].Age, label='Died', shade=True)`  |
| L14 | `pd.pivot_table(train, index='Survived', values=['Age', 'SibSp', 'Parch', 'Fare'])` |
| L25 | `sns.countplot(house_pred['OverallQual'])`                                          |
| L48 | `x_train.describe()`                                                                |

## Resource Check ($N = 7$)

| Key  | Code                                      |
|------|-------------------------------------------|
| P68  | `print('GPU is available')`               |
| P71  | `print('Hub version: ', hub.__version__)` |
| P82  | `print('Running on TPU ', tpu.master())`  |
| P86  | `print('Cuda is available')`              |
| P107 | `print('Model loaded')`                   |
| L64  | `full_table.head(-5)`                     |
| L66  | `prostate_cancer_df.shape`                |

## Spot Check ($N = 5$)

| Key  | Code                          |
|------|-------------------------------|
| L60  | `X_pca.head()`                |
| P64  | `print(np.max(cur[:, :, 1]))` |
| P114 | `print(onehot_encoded)`       |

## Model Training Check ($N = 4$)

| Key | Code                                                                                                                                            |
|-----|-------------------------------------------------------------------------------------------------------------------------------------------------|
| L8  | `autoencoder.fit(x=X_train, y=X_train, epochs=15, validation_data=[X_test, X_test], callbacks=[keras_utils.TqdmProgressCallback()], verbose=0)` |
| L31 | `adaBoost.fit(X_train, y_train)`                                                                                                                |
| L42 | `m_r.best_params_`                                                                                                                              |

## Missing Value Check ($N = 3$)

| Key | Code                                                                           |
|-----|--------------------------------------------------------------------------------|
| P74 | `print(train_df.isnull().sum())`                                               |
| L12 | `sns.heatmap(test_df.isnull(), yticklabels=False, cbar=False, cmap='viridis')` |
| L36 | `test.isna().sum().unique()`                                                   |

## Shape Check ($N = 3$)

| Key  | Code                                                      |
|------|-----------------------------------------------------------|
| P4   | `print('no.of examples in test data : ', len(test_data))` |
| P32  | `print('Training set shape : ', x_train.shape)`           |
| P117 | `print('Y_train.shape: ', Y_train.shape)`                 |

## Data Relationship Check ($N = 2$)

| Key | Code                                                                                                      |
|-----|-----------------------------------------------------------------------------------------------------------|
| L6  | `b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)` |
| L10 | `sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)`             |

## Type Check ($N = 2$)

| Key | Code                                |
|-----|-------------------------------------|
| P43 | `print('data type:', images.dtype)` |
| L71 | `type(Y)`                           |

## Execution Time Check ($N = 1$)

| Key | Code                       |
|-----|----------------------------|
| P66 | `print('Total Run Time:')` |

## Network Architecture Check ($N = 1$)

| Key | Code               |
|-----|--------------------|
| P92 | `print(MyNetwork)` |

### 2024-07-30 report/emse24.tex
feat(report): emse24 updates

@@
-%
+% vim: wrap cursorline
+
 % First comes an example EPS file -- just ignore it and
@@
 Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.
-This format supports a literate approach to programming, making it easier to share and understand complex analyses and workflows.
-Jupyter Notebooks in particular, have become a popular tool in the ML community since they facilitate open and reproducible science, education, and data analysis.
 
@@
 The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)~\cite{pythonast}.
-The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree, and compile it back into executable code, providing a powerful tool for dynamic code manipulation.
+The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree and compile it back into executable code, providing a powerful tool for dynamic code manipulation.
 In Python, \lstinline{assert} statements are represented in the AST by the \lstinline{Assert} class.
@@
 
-% TODO: feedback from Diomidis
-% - Sections 4.1 and 4.2 are very interesting but overly detailed. I suggest to move most of the content to an appendix or an online appendix, and only present the uses in a table or a taxonomy tree.
-% TODO: In addition, consider moving the more general observations into a discussion section.
-
 \begin{figure}
@@
 Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}.
-Throughout the text, we reference the candidates (\textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements) using their unique identifiers as examples.
-These can be viewed on our online appendix.
+Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples.
+These can be viewed on our online appendix~\cite{appendix}.
 
@@
     \includegraphics[width=\linewidth]{common-last-df-functions.pdf}
-    \caption{Most common functions called on pandas data frames.}
+    \caption{Most common functions called on pandas dataframes.}
     \label{fig:common-last-df-functions}
@@
 \begin{lstlisting}[caption={Assertion \emph{A58} used to check that the accuracy of a model is close to the specified value.
-The use of \lstinline{np.isclose} allows for small deviations in the accuracy thus accounting for the stocastic nature of ML models.}, label={lst:A58}]
+The use of \lstinline{np.isclose} allows for small deviations in the accuracy thus accounting for the stochastic nature of ML models.}, label={lst:A58}]
 assert np.isclose(accuracy, 0.9666666666666667)
@@
 This is particularly important when combining convolutional layers from different sources, as the spatial dimensions and channel configurations must align correctly.
-Checks include the use of assertions to verify that the output and input dimensions of consecutive convolutional layers match (\emph{A11}), checking that shape of the activations have the right number of channels (\emph{A62}) and that the appropriate regularization parameter is set (\emph{A75}).
+Checks include assertions that verify the output and input dimensions of consecutive convolutional layers match (\emph{A11}), checking that shape of the activations have the right number of channels (\emph{A62}) and that the appropriate regularization parameter is set (\emph{A75}).
 Such architectural checks are crucial for preventing potential issues during the forward pass of the neural network and ensuring that the custom and pre-trained components are compatible with each other.
@@
 By strictly separating these datasets, the validation phase provides an unbiased evaluation of the model's ability to generalize to data similar to, but not identical to that which it was trained on.
-This is demonstrated by Assertion \emph{A33} in Listing~\ref{lst:A33} ensures that there are no overlapping \lstinline{PetID} values between the training set and the validation set, thereby preventing any possibility of data leakage.
+This is demonstrated by Assertion \emph{A33} in Listing~\ref{lst:A33} which ensures that there are no overlapping \lstinline{PetID} values between the training set and the validation set, thereby preventing any possibility of data leakage.
 
@@
 
-\subsection{(RQ3) How is implicit feedback from print statements and last statements of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-implicit}
+\subsection{(RQ3) How is implicit feedback from print statements and last cell statements used when writing ML code in Jupyter notebooks?}~\label{sec:result-implicit}
 
@@
 
-\highlight{\textbf{Finding 3.1} The study identifies 33 outputs used to check the performance of trained ML models.
+\highlight{\textbf{Finding 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models.
 These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
@@
 
-\highlight{\textbf{Finding 3.2} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
+\highlight{\textbf{Finding 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
 These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
@@
 
-\highlight{\textbf{Finding 3.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed.
+\highlight{\textbf{Finding 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed.
 These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions.
@@
 
-\highlight{\textbf{Finding 3.4} This study finds five outputs used to perform spot checks at various stages of the MLDL.
+\highlight{\textbf{Finding 3.4} This study finds five implicit spot checks performed at various stages of the MLDL.
 Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
@@
 
-\highlight{\textbf{Finding 3.5} The study identifies four outputs used to monitor model training progress.
-These outputs provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods.
+\highlight{\textbf{Finding 3.5} The study identifies four implicit checks used to monitor model training progress.
+These checks provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods.
 These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
@@
 
-\highlight{\textbf{Finding 3.6} The study identifies three outputs used to check for missing data.
+\highlight{\textbf{Finding 3.6} The study identifies three implicit checks for missing data.
 Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations.
@@
 
-\highlight{\textbf{Finding 3.7} The study identifies three outputs used to check the shape of data.
+\highlight{\textbf{Finding 3.7} The study identifies three implicit checks used to validate the shape of data.
 These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
@@
   \centering
-  \caption{Cell outputs used to verify linear relationship between features in the data.}
+  \caption{Last cell statements used to verify linear relationship between features in the data.}
   \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
@@
 
-\highlight{\textbf{Finding 3.8} The study identifies two outputs used to verify linear relationships between features in the data.
+\highlight{\textbf{Finding 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data.
 Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
@@
 
-\highlight{\textbf{Finding 3.9} The study identifies two outputs used to verify the data types of features in the dataset.
+\highlight{\textbf{Finding 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset.
 These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training.
@@
 We argue that the focus must shift towards building reliable ML-enabled systems by incorporating decades of research from the software engineering community with ML.
-Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\cite{tudremla,mlinprod,kastner2020teaching,cain2024,icse2025} Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
+Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\cite{tudremla,mlinprod,kastner2020teaching,cain2024,icse2025}.
+Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
 
@@
 
-\paragraph{\textbf{Internal Validity}} This study focuses on Jupyter notebooks that focus on machine learning.
+\paragraph{\textbf{Internal Validity}} This study examines Jupyter notebooks written in Python, specifically focusing on machine learning code.
 We do this by only considering notebooks that utilize popular ML libraries.
@@
 
-\paragraph{\textbf{External Validity}} This study examines Jupyter notebooks written in Python, specifically focusing on machine learning code.
-The analysis is based on notebooks publicly available on GitHub and Kaggle.
+\paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks publicly available on GitHub and Kaggle.
 Consequently, the findings may not be generalizable to Python notebooks that do not concentrate on ML.
@@
 \section{Related Work}\label{sec:related}
-
-% TODO: feedback from Diomidis
-% - In Section 8 consider adding material regarind interactive ML development.
 
@@
 Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \emph{Quaranta et al.}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 2.5 million Jupyter notebooks written in Python.
+To enable future research on computational notebooks, \emph{Quaranta et al.}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 250 thousand Jupyter notebooks written in Python.
 The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
@@
 Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
-This study leverages the computational narrative present within notebooks to identify visualizations and assertions that are semantically linked to each other.
 
@@
 Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory,baluta2021scalable}.
-Several works have been conducted on generating and detecting adversarial inputs for ML models.
+Several studies have been conducted on generating and detecting adversarial inputs for ML models.
 Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution,gao2020fuzz,wang2021robot,zhang2020white-box}.
@@
 Additionally, our research underscores the necessity for better documentation practices to mitigate technical debt and ensure the reproducibility of ML experiments.
-By integrating feedback mechanisms into the ML development workflow, practitioners can achieve more reliable and efficient model development processes.
+Feedback mechanisms are used throughout the ML development workflow to achieve more reliable and efficent model development processes.
 Future work should focus on developing specialized tools and frameworks to support these practices, thereby advancing the field of ML engineering.

### 2024-07-30 report/emse24.tex
feat(report): emse24 update t2v

@@
 
+We enforced a time resource limit of 200 hours for conducting the case studies in RQ2 and RQ3, potentially constraining the depth and breadth of our investigation.
+The stratified random sampling used for selecting the case study candidates mitigates this threat by ensuring that a representative cross-section of the population is sampled within the time constraint.
+While this sampling method enhances the generalizability of our results, the time limitation should still be considered when interpreting and applying the study's findings.
+
 \paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks publicly available on GitHub and Kaggle.

### 2024-07-30 report/emse24.tex
feat(report): emse24 t2v updates

@@
 
+The case studies conducted in RQ2 and RQ3 are subject to bias from the first author's individual perspective and understanding which could limit the interpretation of data and potentially overlook alternative viewpoints or explanations.
+To mitigate this threat, we implemented a rigorous review process involving multiple rounds of discussion with the second and third authors.
+These discussions continued until a consensus was reached on the interpretations and findings.
+
 \paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks publicly available on GitHub and Kaggle.

### 2024-07-31 report/emse24.tex
feat(report): emse24 abstract updates

@@
 %
+\usepackage{natbib}
 \usepackage{graphicx}
@@
 \author{Arumoy Shome\and
-Lu\`{i}s Cruz\and
+Lu{\'\i}s Cruz\and
   Diomidis Spinellis\and
@@
 \begin{abstract}
-  The machine learning development lifecycle involves iterative and exploratory processes that heavily rely on feedback mechanisms to ensure data and model integrity.
-This study investigates the use of implicit and explicit feedback within Jupyter notebooks.
-We analyse 297.8 thousand Jupyter notebooks from GitHub and Kaggle, identifying key feedback mechanisms such as print statements, outputs from last cell statements, and assertions.
-Our findings reveal extensive use of implicit feedback for critical design decisions and highlight the limited adoption of explicit feedback mechanisms like assertions.
-By conducting detailed case studies on selected feedback instances, we uncover the reliance on manual validation and the potential for automated validation to enhance robustness and maintainability.
-This study underscores the need for improved documentation and testing practices within the ML community, advocating for the integration of feedback mechanisms into the ML development workflow to mitigate technical debt and enhance reproducibility.
+  The machine learning development lifecycle is characterized by iterative and exploratory processes that rely on feedback mechanisms to ensure data and model integrity.
+Despite the critical role of feedback in machine learning engineering, no prior research has been conducted to identify and understand these mechanisms.
+To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyse 2.3 million code cells.
+We identify three key feedback mechanisms---assertions, print statements and last cell statements---and further categorize them into implicit and explicit forms of feedback.
+Our findings reveal extensive use of implicit feedback for critical design decisions and the relatively limited adoption of explicit feedback mechanisms.
+By conducting detailed case studies with selected feedback instances, we uncover the potential for automated validation of critical assumptions in ML workflows using assertions.
+Finally, this study underscores the need for improved documentation, and provides practical recommendations on how existing feedback mechanisms in the ML development workflow can be effectively used to mitigate technical debt and enhance reproducibility.
 \end{abstract}
 
-\keywords{Assertions, Feedback Mechanisms, Jupyter notebooks, SE4AI}
+\keywords{SE4AI, Feedback Mechanisms, Jupyter notebooks, Assertions, Technical Debt}
 
@@
 Practitioners may also engage in comparing and contrasting different models to identify the most effective approach, followed by fine-tuning selected models to optimize performance.
-The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\cite{haakman2021ai,amershi2019software,sculley2015hidden}.
-
-The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\cite{betz2018managing}.
-Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, data preparation, modelling, evaluation and finally deployment and monitoring~\cite{martinez-plumed2021crisp-dm}.
+The final model is then deployed into production where it is continuously monitored, and periodically retrained to adapt to new data or changing conditions~\citep{haakman2021ai,amershi2019software,sculley2015hidden}.
+
+The MLDL is fundamentally iterative, closely resembling Agile software engineering principles, which emphasize incorporating early feedback and working in small, manageable iterations~\citep{betz2018managing}.
+Similarly, CRISP-DM---the dominant process model in data mining---outlines a cyclical process that evolves from understanding the business objectives and data, to data preparation, modelling, evaluation and finally deployment and monitoring~\citep{martinez-plumed2021crisp-dm}.
 Both systems advocate for a dynamic, iterative approach that enhances adaptability and effectiveness in dealing with complex, changing systems like those encountered in ML projects.
 
-Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\cite{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
+Computational notebooks---specifically Jupyter notebooks\footnote{https://jupyter.org/}---have been ubiquitously adopted by the ML community for developing ML enabled systems~\citep{pimentel2019large-scale,quaranta2021kgtorrent,psallidas2019data,perkel2018why}.
 Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks.
@@
 Users can test hypotheses, debug issues, and make incremental changes with immediate feedback.
-The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\cite{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.
+The ability to visualize data and results inline and document the process in a narrative form further enhances their utility~\citep{kery2018story,head2019managing,rule2018exploration,chattopadhyay2020whats}.
 
@@
 
-As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\cite{kery2018story,rule2018exploration}.
+As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments~\citep{kery2018story,rule2018exploration}.
 In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones.
@@
     We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle.
-We analyse 2.3 million code cells obtained from the notebooks and create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements.
+We analyse 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.
+
+    We create a public dataset of 89.6 thousand assertions, 1.4 million print statements and 1 million last cell statements.
 We perform a descriptive and lexical analysis of the dataset and report the key characteristics of feedback mechanisms used in ML Jupyter notebooks.
-The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence~\cite{replication}.
+The dataset along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence\footnote{https://doi.org/10.6084/m9.figshare.26372140.v1}.
 
@@
 \end{description}
-
-% The above all motivates the research: but then the intro should also ¡®sell¡¯ the outcomes of the research. These are
-% - Identification of over 20 feedback mechanisms, half of them implicit, half of them explicit (your taxonomy)
-% - Insights in actual use of these mechanisms
-% - Recommendations for better support of these mechanisms in notebook environments / life cycle
 
@@
 Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI).
-Resembling the literate programming paradigm~\cite{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.
+Resembling the literate programming paradigm~\citep{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.
 
@@
   \includegraphics[width=\linewidth]{nb.pdf}
-  \caption{Example Jupyter notebook with code and markdown cells, adapted ~\cite[Figure~1]{pimentel2019large-scale}.}
+  \caption{Example Jupyter notebook with code and markdown cells, adapted from~\citep[Figure~1]{pimentel2019large-scale}.}
   \label{fig:nb}
@@
 
-The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)~\cite{pythonast}.
+The Python \lstinline{ast} module enables the parsing, manipulation, and analysis of Python source code by converting it into an abstract syntax tree (AST)\footnote{https://docs.python.org/3/library/ast.html}.
 The \lstinline{ast} module can parse source code strings into AST objects, traverse and modify the tree and compile it back into executable code, providing a powerful tool for dynamic code manipulation.
@@
 We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}.
-We used the pre-existing dataset KGTorrent~\cite{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub.
+We used the pre-existing dataset KGTorrent~\citep{quaranta2021kgtorrent} for Kaggle since it does not support advanced code-based search like GitHub.
 To the best of our knowledge, KGTorrent is the largest dataset of Python Jupyter notebooks obtained from Kaggle, consisting of 247.9 thousand notebooks.
@@
 Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\cite{virtanen2020scipy}, Pytorch~\cite{paszke2017automatic}, Tensorflow~\cite{abadi2015tensorflow} and Keras~\cite{chollet2015keras}.
-The libraries are selected based on the import analysis conducted by \emph{Psallidas et al.}~\cite{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub.
+Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}.
+The libraries are selected based on the import analysis conducted by \citet{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub.
 We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells.
@@
 
-% TODO: add citation to online appendix
-
 The results of the study are presented in this section.
@@
 Throughout the text, we reference the candidate \textbf{(A)}ssertions, \textbf{(P)}rint statements and \textbf{(L)}ast cell statements using their unique identifiers as examples.
-These can be viewed on our online appendix~\cite{appendix}.
+These can be viewed on our online appendix\footnote{https://arumoy.me/shome2023notebook}.
 
 \subsection{(RQ1) What feedback mechanisms are employed in the development of ML Jupyter notebooks?}~\label{sec:result-analysis}
-
-% TODO: add the findings from this part of the analysis
 
@@
 \subsubsection{Data shape check ($N = 26$)}
-% (A5, A9, A13, A16, A17, A29, A31, A61, A71, A76--78, A82, A84, A85, A89, A90, A91, A93--96, A98--101)
 
@@
 \subsubsection{Data validation check ($N = 14$)}
-% formerly value-check (A20, A30, A41, A44--46, A52, A65, A68, A69, A73, A83, A87, A92)
 
@@
 \subsubsection{Model performance check ($N = 11$)}\label{sec:assert-model-perf}
-% (A7, A15, A19, A22, A24, A26, A38, A54, A58, A59, A72)
 
@@
 \subsubsection{Existence check ($N = 8$)}
-% NOTE: this is a combination of missing-value-check and existence-check
-% (A23, A42, A43, A50, A51, A63, A79, A86)
 
@@
 \subsubsection{Resource check ($N = 7$)}~\label{sec:result-rq2-resource-check}
-% (A10, A14, A18, A37, A60, A67 A74)
 
@@
 \subsubsection{Type check ($N = 5$)}
-% (A2, A35, A40, A81, A88)
 
@@
 \subsubsection{Mathematical property check ($N = 4$)}
-% (A3, A25, A56, A64)
 
@@
 \subsubsection{Batch size check ($N = 3$)}
-% (A21, A28, A70)
 
@@
 \subsubsection{Network architecture check ($N = 3$)}
-% (A11, A62, A75)
 
@@
 \subsubsection{Data leakage check ($N = 1$)}
-% (A33)
 
@@
 \subsubsection{Model performance check ($N = 33$)}
-% (O3, O50, O52, O57, O74, P3, P6, P8--12, P16--19, P23, P24, P28, P30, P34, P42, P47, P48, P50, P51, P54, P55, P57, P58, P61, P78, P93)
 
@@
 \subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
-% (O2, O4, O9, O14, O20, O25, O48)
 
@@
 \subsubsection{Resource check ($N = 7$)}\label{sec:implicit-resource-check}
-% (P71, P86, O64, O66, P68, P82, P107)
 
@@
 \subsubsection{Spot check ($N = 5$)}
-% (O56, O60, P64, P67, P114)
 
@@
 \subsubsection{Model training check ($N = 4$)}
-% (O8, O31, O42, P77)
 
@@
 \subsubsection{Missing value check ($N = 3$)}
-% (O12, O36, P74)
 
 Checking for missing values is a critical step in the preprocessing phase of an ML project because it significantly influences the quality and performance of the model.
-Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\cite{shome2022data}.
+Missing data can lead to biased or incorrect conclusions if not handled properly, potentially skewing the model's performance by training on incomplete or non-representative samples~\citep{shome2022data}.
 Many ML algorithms demand complete numerical datasets to perform calculations such as matrix multiplication.
@@
 \subsubsection{Shape check ($N = 3$)}
-% (P4, P32, P117)
 
@@
 \subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
-% (O6, O10)
 
@@
 However, features within a dataset that exhibit a linear relationship are considered redundant as they convey the same information to the model during training.
-Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\cite{shome2022data}.
+Consequently, the feature engineering stage often involves removing such features to create a more efficient training dataset~\citep{shome2022data}.
 Table~\ref{tab:linear-relation-check} illustrates two last cell statements found in this study for identifying linear relationships among features in the data.
@@
 \subsubsection{Type check ($N = 2$)}
-% (O71, P43)
 
@@
 \subsubsection{Execution time check ($N = 1$)}
-% (P66)
 \begin{lstlisting}[caption={Print statement \emph{P66} used to check the total execution time of training an ML model with various hyper-parameter configurations.}, label={lst:P66}]
@@
 While training multiple ML models for a single task is common, choosing the best model considers not only its performance but also its training speed.
-Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\cite{shome2022data}.
+Faster training times are more cost-effective for iterative tasks and promote sustainability in the long run~\citep{shome2022data}.
 Listing~\ref{lst:P66} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyperparameter configurations using a grid-search.
@@
 \subsubsection{Network architecture check ($N = 1$)}
-% (P92)
 
@@
 
-Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\cite{kery2018story,rule2018exploration,pimentel2019large-scale}.
+Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\citep{kery2018story,rule2018exploration,pimentel2019large-scale}.
 Once the code is finalized, it is typically transferred into a script that is run in a production environment.
 Alternatively, the notebook may contain significant exploratory work such as developing a model or performing data exploration.
-To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\cite{kery2018story,rule2018exploration}.
-In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\cite{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
-
-We believe that such structured information can be derived from the feedback mechanisms available in Jupyter notebooks.
-Our results show that practitioners use assertions, print statements, output of code cells and visualizations to determine the appropriate data preprocessing techniques, detect erroneous assumptions, create new features to boost model performance, and check for hardware and software dependencies.
+To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\citep{kery2018story,rule2018exploration}.
+In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\citep{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
+
+Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks.
+The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
 As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
-This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\cite{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
-
-Our results from RQ3 reveal that implicit feedback from print and last cell statements are extensively used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
+This not only prolongs time required for debugging and troubleshooting, but also incurs additional costs and resource expenditures~\citep{sculley2015hidden,amershi2019software,sambasivan2021everyone}.
+
+Our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
 Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.
@@
 We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations.
-Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\cite{heer2010tour}.
+Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently~\citep{heer2010tour}.
 By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
-Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\cite{yang2021subtle}.
+Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.
 
@@
 This is a cause for concern since these models may be deployed in production and their output may be used by other services which will then also perform poorly.
-The output of the ML models are often also used to capture new training data, thus leading to a feedback loop which over time will degrade the model performance~\cite{sculley2015hidden,breck2019data}.
-
-\emph{Breck et al.
-(2017)}~\cite{breck2017ml} view ML training as analogous to compilation and recommend that training data be tested, similar to code in traditional software.
+The output of the ML models are often also used to capture new training data, thus leading to a feedback loop which over time will degrade the model performance~\citep{sculley2015hidden,breck2019data}.
+
+\citet{breck2017ml} view ML training as analogous to compilation and recommend that training data be tested, similar to code in traditional software.
 They propose a rubric to assess the production readiness of an ML pipeline, which includes several recommendations for data testing.
-In subsequent work, \emph{Breck et al.
-(2019)}~\cite{breck2019data} introduce \emph{TensorFlow Data Validation}, an automated tool for data validation that works on principles borrowed from the database community.
+In subsequent work, \citet{breck2019data} introduce \emph{TensorFlow Data Validation}, an automated tool for data validation that works on principles borrowed from the database community.
 Data expectations are specified in a schema, and each new data batch is validated against this schema to ensure it meets the expected criteria or prompts schema updates if requirements change.
@@
 Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
-Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\cite{wang2020assessing}.
-
-In the results from RQ2 and RQ3, we observe that many assertions, print statements and last cell statements are used in the upstream, data intensive stages of the MLDL.
-When integrating dedicated data validation tools into an operational ML pipeline, interactions with the feedback mechanisms in notebooks can aid in writing the data validation schema.
-For instance, this study finds several assertions that perform data validation checks to ensure that data meets specific criteria or constraints.
-Similarly, we also find assertions, prints statements and last cell statements that performs type, shape and distribution checks that serve as safeguards against data inconsistencies.
+Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\citep{wang2020assessing}.
+
+Results from RQ2 and RQ3 show that assertions, print statements and last cell statements are predominantly used in the upstream, data-intensive stages of the MLDL.
+For instance, this study identifies several assertions that perform data validation checks to ensure data meets specific criteria or constraints.
+Similarly, we find assertions, print statements, and last cell statements that conduct type, shape, and distribution checks to safeguard against data inconsistencies.
+These results show a missed opportunity in data validation tasks.
+We argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
 
@@
 
-We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties~\cite{simmons2020large-scale}.
+We hypothesize that this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties.
 Guidelines and curricula on ML coding practices have historically focused on building better models.
 We argue that the focus must shift towards building reliable ML-enabled systems by incorporating decades of research from the software engineering community with ML.
-Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\cite{tudremla,mlinprod,kastner2020teaching,cain2024,icse2025}.
+Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\citep{kastner2020teaching}.
 Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
@@
 
-% Our observations also indicate that it is much easier for practitioners to use implicit checks and manually validate data rather than translating these checks into assertions.
-This highlights the need for more automated tools that can help practitioners convert implicit checks into explicit assertions, thereby improving the robustness and maintainability of ML pipelines.
-
 \subsection{Machine Learning Bugs}
@@
 The findings of this study align with prior research on real-world defects in ML systems.
-\emph{Humbatova et al.}~\cite{humbatova2020taxonomy} propose a taxonomy of faults in DL systems, while \emph{Morovati et al.}~\cite{morovati2023bugs} curate \emph{Defects4ML}, a dataset consisting of ML bugs.
+\citet{humbatova2020taxonomy} propose a taxonomy of faults in DL systems, while \citet{morovati2023bugs} curate \emph{Defects4ML}, a dataset consisting of ML bugs.
 Both studies identify defects arising from the use of incorrect data types and shapes, deprecated APIs, and missing GPU resources.
@@
 Further research is required to explore the feedback mechanisms in more detail and organize the knowledge into a taxonomy based on specific ML validation tasks.
-Future work can focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \emph{Humbatova et al.}~\cite{humbatova2020taxonomy} and \emph{Morovati et al.}~\cite{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
+Future work can focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
 
@@
 
-Results from RQ1 show that majority of the assertions do not contain a failure message.
+Results from RQ1 show that 68\% of the assertions do not contain a failure message.
 This can lead to problems such as unclear error diagnostics, difficulty in debugging, and increased time to identify and resolve issues.
-\emph{Vidoni et al.}~\cite{vidoni2021evaluating} investigated the testing culture in open-source R packages and identified several anti-patterns that lead to the accumulation of testing technical debt.
-The dataset contributed by this study can be used by researchers to replicate the \emph{Vidoni et al.}~\cite{vidoni2021evaluating} study using assertions written in Python.
+\citet{vidoni2021evaluating} investigated the testing culture in open-source R packages and identified several anti-patterns that lead to the accumulation of testing technical debt.
+The dataset contributed by this study can be used by researchers to replicate the \citet{vidoni2021evaluating} study using assertions written in Python.
 
@@
 
-\emph{Kochhar et al.
-(2017)}~\cite{kochhar2017revisiting} conducted a partial replication of a prior study on assertion usage, expanding the analysis to 185 Java projects from GitHub.
-\emph{Vidoni (2021)}~\cite{vidoni2021evaluating} conducted a comprehensive study on the unit testing practices within 177 systematically selected open-source R packages hosted on GitHub.
+\citet{kochhar2017revisiting} conducted a partial replication of a prior study on assertion usage, expanding the analysis to 185 Java projects from GitHub.
+\citet{vidoni2021evaluating} conducted a comprehensive study on the unit testing practices within 177 systematically selected open-source R packages hosted on GitHub.
 The research addressed the quality of testing, identified testing goals, and pinpointed potential sources of testing technical debt.
 
-\emph{Humbatova et al.
-(2020)}~\cite{humbatova2020taxonomy} developed a comprehensive taxonomy of faults in deep learning systems through the manual analysis of 1059 artefacts from GitHub and Stack Overflow, complemented by structured interviews with 20 researchers and practitioners.
+\citet{humbatova2020taxonomy} developed a comprehensive taxonomy of faults in deep learning systems through the manual analysis of 1059 artefacts from GitHub and Stack Overflow, complemented by structured interviews with 20 researchers and practitioners.
 The study categorized faults into five top-level categories, encompassing model, data, training, deployment, and infrastructure-related faults.
-\emph{Morovati et al.
-(2023)}~\cite{morovati2023bugs} addressed the challenge of assessing the quality and performance of tools designed to improve the reliability of machine learning components by creating a standard fault benchmark called defect4ML.
+\citet{morovati2023bugs} addressed the challenge of assessing the quality and performance of tools designed to improve the reliability of machine learning components by creating a standard fault benchmark called defect4ML.
 This benchmark consists of 100 reproducible bugs reported by ML developers on GitHub and Stack Overflow, focusing on the TensorFlow and Keras ML frameworks.
@@
 
-\emph{Psallidas et al.}~\cite{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
+\citet{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
 Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
-\emph{Pimentel et al.}~\cite{pimentel2019large-scale} mined 1.4 million notebooks from GitHub to conduct an empirical study on the coding practices in computational notebooks.
+\citet{pimentel2019large-scale} mined 1.4 million notebooks from GitHub to conduct an empirical study on the coding practices in computational notebooks.
 Based on their analysis, the authors propose guidelines on improving reproducibility of computational notebooks.
-To enable future research on computational notebooks, \emph{Quaranta et al.}~\cite{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 250 thousand Jupyter notebooks written in Python.
+To enable future research on computational notebooks, \citet{quaranta2021kgtorrent} mine Kaggle and present \textit{KGTorrent}, a public dataset consisting of approximately 250 thousand Jupyter notebooks written in Python.
 The dataset also contains a relational database dump of metadata regarding publicly available notebooks on Kaggle.
@@
 Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualizations~\cite{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
-Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\cite{liu2019understanding,kandel2012enterprise}.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualizations~\citep{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
+Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\citep{liu2019understanding,kandel2012enterprise}.
 The findings from these studies can be used as guidelines for improving existing notebook technologies or designing new ones.
 
-To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \emph{Head et al.}~\cite{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
-Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\cite{yang2021subtle,bavishi2021vizsmith}.
-
-Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\cite{wang2020assessing,pimentel2019large-scale,quaranta2021kgtorrent}.
+To manage and prune multiple versions of code that accumulate over time when developing in notebooks, \citet{head2019managing} propose a code gathering tool that allows practitioners to review and only keep the relevant version of code.
+Other tools such as \textit{WrangleDoc} and \textit{VizSmith} have also been proposed to aid ML practitioners when working within computational notebooks~\citep{yang2021subtle,bavishi2021vizsmith}.
+
+Jupyter notebooks have been widely adopted by the DS and ML communities to develop ML pipelines~\citep{wang2020assessing,pimentel2019large-scale,quaranta2021kgtorrent}.
 Computational notebooks provide a rich source of data in the form of natural text, code and visualizations.
-Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\cite{rule2018exploration}.
+Computational notebooks also allow ML practitioners to present the knowledge gained from the analysis in a narrative that can be shared with others~\citep{rule2018exploration}.
 
@@
 First, on functional properties such as correctness and robustness of the model towards unseen data.
-And second, on non-functional properties such as fairness, interpretability and privacy~\cite{zhang2022machine,mehrabi2021survey,chen2022fairness}.
+And second, on non-functional properties such as fairness, interpretability and privacy~\citep{zhang2022machine,mehrabi2021survey,chen2022fairness}.
 
 To test the correctness of ML models, several improvements over existing test adequacy metrics have been proposed.
-Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\cite{pei2017deepxplore,ma2018deepgauge,gerasimou2020importance-driven}.
+Tools such as \textit{DeepXplore} and \textit{DeepGauge} propose new test adequacy metrics such as neuron coverage adapted for ML enabled software systems~\citep{pei2017deepxplore,ma2018deepgauge,gerasimou2020importance-driven}.
 Formal verification methods have also been proposed that try to provide formal guarantee of robustness against adversarial examples.
-Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\cite{zhu2021deepmemory,baluta2021scalable}.
+Such methods however are only feasible for statistical ML models and become computationally expensive for more complex models such as deep neural networks~\citep{zhu2021deepmemory,baluta2021scalable}.
 Several studies have been conducted on generating and detecting adversarial inputs for ML models.
-Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\cite{braiek2019deepevolution,gao2020fuzz,wang2021robot,zhang2020white-box}.
+Data augmentation techniques based on fuzzing, search based software testing and mutation testing have been proposed to generate adversarial examples that can be used during model training to improve its robustness~\citep{braiek2019deepevolution,gao2020fuzz,wang2021robot,zhang2020white-box}.
 It is however not possible to include all variations of adversarial examples into the training data.
-Thus, methods have been proposed to detect adversarial inputs during runtime~\cite{xiao2021self-checking,wang2020dissector,wang2019adversarial,berend2020cats}.
+Thus, methods have been proposed to detect adversarial inputs during runtime~\citep{xiao2021self-checking,wang2020dissector,wang2019adversarial,berend2020cats}.
 
@@
 ML models and the underlying data is complex and multidimensional.
-To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\cite{yuan2021survey,hohman2019visual,wexler2019what-if}.
+To combat the ``curse of dimensionality'', visual analytics has been widely adopted by the ML community to understand the data and the internal workings of ML models~\citep{yuan2021survey,hohman2019visual,wexler2019what-if}.
 
 Prior studies have been conducted to understand how visual analytics techniques are currently being used in ML.
-\emph{Yuan et al.}~\cite{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
-\emph{Hohman et al.}~\cite{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
+\citet{yuan2021survey} conduct a systematic review of 259 papers and propose a taxonomy of visual analytics techniques for ML.
+\citet{hohman2019visual} conduct a survey of visual analytics techniques for Deep Learning Models.
 The findings of the study indicate that visual analytics has been widely adopted in ML for model explanation, interpretation, debugging and improvement.
@@
 Several tools have been proposed by the visual analytics research community to aid practitioners in understanding how their ML models operate.
-\emph{Wexler et al.}~\cite{wexler2019what-if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
-To reduce the cognitive load of ML practitioners during model building phase, \emph{Amershi et al.}~\cite{amershi2019software} propose \textit{ModelTracker}.
+\citet{wexler2019what-if} propose \textit{What-If}, a visual analytics tool to explore alternative hypotheses, generate counterfactual reasoning, investigate decision boundaries of the model and how change in data affects the model predictions.
+To reduce the cognitive load of ML practitioners during model building phase, \citet{amershi2019software} propose \textit{ModelTracker}.
 Given a trained model and a sample dataset, ModelTracker presents all the information in traditional summary statistics along with the performance of the model.
 
-\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialized use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalized Additive Models, prioritizing machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning\cite{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
+\textit{ESCAPE}, \textit{GAM Coach}, \textit{Angler} and \textit{Drava} are a few other tools that have been proposed to handle specialized use-cases such as identifying systematic errors in ML models, generating counterfactual explanations for Generalized Additive Models, prioritizing machine translation model improvements and relating human concepts with semantic dimensions extracted by ML models during disentangled representation learning~\citep{ahn2023escape,wang2023gam,robertson2023angler,wang2023drava}.
 
@@
 This study provides a comprehensive analysis of feedback mechanisms employed within Jupyter notebooks during the machine learning development lifecycle.
-Our findings indicate that while implicit feedback mechanisms such as print statements and last cell outputs are extensively utilized, there is a significant gap in the adoption of explicit feedback mechanisms like assertions.
+Our findings indicate that while implicit feedback mechanisms such as print statements and last cell statements are extensively utilized, there is a significant gap in the adoption of explicit feedback mechanisms like assertions.
 This reliance on manual validation highlights the potential for automated data validation to enhance the robustness and maintainability of ML pipelines.
-Additionally, our research underscores the necessity for better documentation practices to mitigate technical debt and ensure the reproducibility of ML experiments.
+This study also underscores the necessity for better documentation practices to mitigate technical debt and ensure the reproducibility of ML experiments.
 Feedback mechanisms are used throughout the ML development workflow to achieve more reliable and efficent model development processes.
@@
 % BibTeX users please use one of
-% \bibliographystyle{spbasic}      % basic style, author-year citations
-\bibliographystyle{spmpsci}      % mathematics and physical sciences
+\bibliographystyle{spbasic}      % basic style, author-year citations
+%% \bibliographystyle{spmpsci}      % mathematics and physical sciences
 %\bibliographystyle{spphys}       % APS-like style for physics

### 2024-09-30 report/emse24.tex
Update .gitignore and emse24.tex

@@
 
-% First comes an example EPS file -- just ignore it and
-% proceed on the \documentclass line
-% your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
-%
 \RequirePackage{fix-cm}
@@
 %
+%% \usepackage{hyperref}
 \usepackage{natbib}
@@
 
+\section{Declarations}
+
+\paragraph{\textbf{Conflict of Interest}} The authors declare that they have no
+conflict of interest.
+
+\paragraph{\textbf{Data Availability Statement}} The dataset created and used
+in this study along with the replication package is shared publicly
+under the Creative Commons Attribution (CC-BY) licence:
+https://doi.org/10.6084/m9.figshare.26372140.v1. The case study
+candidates analysed in RQ2 and RQ3 can be viewed in our online
+appendix: https://arumoy.me/shome2023notebook.
+
 % BibTeX users please use one of

### 2025-02-04 report/emse24-revised.tex
feat: init emse24-revised.tex

--- 2024-09-30 report/emse24.tex
@@
 Figure~\ref{fig:data-collection} illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle.
+GitHub is a popular platform to host and collaborate on software projects.
+It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery.
+In contrast, Kaggle is focused specifically on data science and machine learning.
+It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions.
+While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.
+
 We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}.
@@
 
+\subsection{Descriptive Statistics}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
+  \caption{Number of notebooks in each category of...}
+  \label{fig:notebook-size-distribution}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
+  \caption{Distribution of number of cells for various notebook sizes.}
+  \label{fig:cell-per-notebook-size-distribution.pdf}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
+  \caption{Distribution of assertions, prints and last cell statements for various notebook sizes.}
+  \label{fig:fms-per-notebook-size-distribution}
+\end{figure}
+
+% TODO should we add the descriptive stats as a table?
+% TODO do we need to extrapolate more from figure 5?
+Figure~\ref{fig:notebook-size-distribution} shows the distribution of the number of code cells for all notebooks analyzed in this study.
+The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles.
+This is done since the data is large, non-uniform and contains many outliers.
+Figure~\ref{fig:cell-per-notebook-size-distribution.pdf} further magnifies the distribution of the number of cells within each category.
+The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.
+
+Figure~\ref{fig:fms-per-notebook-size-distribution} shows the distribution of assertions, prints and last cell statements across the different notebook size categories.
+A logarithmic scale is used for the count since the data is large and contains outliers.
+Overall, the visualization suggests an increase in the use of all feedback mechanisms as notebook size grows.
+Print statements are the most frequently used across all notebook sizes.
+
 \subsection{(RQ1) What feedback mechanisms are employed in the development of ML Jupyter notebooks?}~\label{sec:result-analysis}
@@
 
+\subsection{Complexity of Notebooks}
+
+Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research.
+We believe that the primary challenge here will be to develop an appropriate definition for notebook complexity.
+
+To the best of our knowledge, no formal definition or metrics to quantify notebook complexity exist to date.
+In their research, \citet{grotov2022large-scale} investigated the complexity of Python code written inside Jupyter notebooks using a subset of 15 metrics derived from prior literature.
+These metrics however were primarily developed for code structured in an object-oriented programming paradigm.
+Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices \cite{pimentel2019large-scale}.
+
+It is further unclear how metrics such as the size of the training data, complexity of the learning algorithm, computational costs and CO\(_2\) emissions associated with training can be incorporated into the definition for notebook complexity.
+Addressing these challenges would facilitate the stratification of our dataset into distinct categories---namely low, medium, and high complexity---which can enhance the depth of this study by enabling a more nuanced analysis of feedback mechanisms within this context.
+
 \section{Threats to Validity}\label{sec:threats}
@@
 
-\paragraph{\textbf{Conflict of Interest}} The authors declare that they have no
-conflict of interest.
-
-\paragraph{\textbf{Data Availability Statement}} The dataset created and used
-in this study along with the replication package is shared publicly
-under the Creative Commons Attribution (CC-BY) licence:
-https://doi.org/10.6084/m9.figshare.26372140.v1. The case study
-candidates analysed in RQ2 and RQ3 can be viewed in our online
-appendix: https://arumoy.me/shome2023notebook.
+\paragraph{\textbf{Conflict of Interest}} The authors declare that they have no conflict of interest.
+
+\paragraph{\textbf{Data Availability Statement}} The dataset created and used in this study along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence: https://doi.org/10.6084/m9.figshare.26372140.v1.
+The case study candidates analysed in RQ2 and RQ3 can be viewed in our online appendix: https://arumoy.me/shome2023notebook.
 

### 2025-02-04 report/emse24-revised.tex
chore(report): emse24-revised updates

@@
 \section{Introduction}
-
 The machine learning development lifecycle (MLDL) encompasses a multifaceted series of entangled stages.
@@
 Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively.
-The term frequency (TF) for all tokens is calculated and then normalized using their inter-document frequency (IDF) such that tokens that appear less frequently are assigned a higher value.
-We apply stratified random sampling to identify the candidates for the case study analysis.
-The subgroups are created by adding TF-IDF of the tokens in each candidate to produce an aggregate value.
+The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF).
+This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
+Consider for instance, the trivial print statement \lstinline{print("hello, world!")}.
+Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.
+
+To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis.
+The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate.
 The candidates are then divided into quartiles based on the aggregate value.
-A candidate is randomly drawn from each bin and analysed as an in-depth case study.
-The analysis is stopped when the time resource is exhausted.
-
+As a result, all candidates characterized by rare tokens are grouped within a single bin.
+By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.
+
+% TODO: update this paragraph, candidates were annotated by first and second author
+% we calculated IRA and obtained high agreement...
 During each case study, we analyse the code of the candidate to understand its purpose.
@@
 
-The results of the study are presented in this section.
-For the case studies conducted for RQ2 and RQ3, we group the candidates based on their similarity to each other.
-This is done only to structure the paper in a meaningful manner and for ease of reporting the results.
 Figure~\ref{fig:taxonomy} provides an overview of the various feedback mechanism groups identified in this study with additional details provided in Section~\ref{sec:result-explicit} and Section~\ref{sec:result-implicit}.
@@
 These can be viewed on our online appendix\footnote{https://arumoy.me/shome2023notebook}.
+
+Due to the exploratory nature of this study and the limited number of case studies analyzed, the observations presented should be interpreted with caution and viewed as illustrative rather than definitive.
+The case studies serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
 
@@
 
-\highlight{\textbf{Finding 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion.
+\highlight{\textbf{Observation 1.1} 25 thousand (8.5\%) notebooks contained at least one assertion.
 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries.
@@
 
-\highlight{\textbf{Finding 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message.
+\highlight{\textbf{Observation 1.2} 31.7\% of the assertions written using the built-in \lstinline{assert} statement have a failure message.
 These assertions frequently perform a comparison check or call an external function.
@@
 
-\highlight{\textbf{Finding 1.3} Print statements are most commonly used to display various performance metrics of ML models.}
+\highlight{\textbf{Observation 1.3} Print statements are most commonly used to display various performance metrics of ML models.}
 
@@
 
-\highlight{\textbf{Finding 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}
+\highlight{\textbf{Observation 1.4} Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statements.}
 
@@
 
-\highlight{\textbf{Finding 1.5} Common functions called on pandas dataframes focus on exploratory data analysis, statistical operations and quick visualizations.
+\highlight{\textbf{Observation 1.5} Common functions called on pandas dataframes focus on exploratory data analysis, statistical operations and quick visualizations.
 In visualizations, functions for categorical data, correlation and distribution are frequently used.}
@@
 
-\highlight{\textbf{Finding 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
+\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
 These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
@@
 
-\highlight{\textbf{Finding 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
+\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
 These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
@@
 
-\highlight{\textbf{Finding 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
+\highlight{\textbf{Observation 2.3} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics.
 These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}
@@
 
-\highlight{\textbf{Finding 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
+\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
 These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
@@
 
-\highlight{\textbf{Finding 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
+\highlight{\textbf{Observation 2.5} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes.
 These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}
@@
 
-\highlight{\textbf{Finding 2.6} Five assertions validate the data types of input features, models, and intermediate objects.
+\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features, models, and intermediate objects.
 These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
@@
 
-\highlight{\textbf{Finding 2.7} Four assertions validate key mathematical properties of data and model outputs.
+\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs.
 These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
@@
 
-\highlight{\textbf{Finding 2.8} This study identifies three assertions validating batch size configurations.
+\highlight{\textbf{Observation 2.8} This study identifies three assertions validating batch size configurations.
 These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
@@
 
-\highlight{\textbf{Finding 2.9} Three assertions verify the consistency of neural network architecture.
+\highlight{\textbf{Observation 2.9} Three assertions verify the consistency of neural network architecture.
 These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
@@
 
-\highlight{\textbf{Finding 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
+\highlight{\textbf{Observation 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
 Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
@@
 
-\highlight{\textbf{Finding 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models.
+\highlight{\textbf{Observation 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models.
 These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
@@
 
-\highlight{\textbf{Finding 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
+\highlight{\textbf{Observation 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
 These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
@@
 
-\highlight{\textbf{Finding 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed.
+\highlight{\textbf{Observation 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed.
 These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions.
@@
 
-\highlight{\textbf{Finding 3.4} This study finds five implicit spot checks performed at various stages of the MLDL.
+\highlight{\textbf{Observation 3.4} This study finds five implicit spot checks performed at various stages of the MLDL.
 Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
@@
 
-\highlight{\textbf{Finding 3.5} The study identifies four implicit checks used to monitor model training progress.
+\highlight{\textbf{Observation 3.5} The study identifies four implicit checks used to monitor model training progress.
 These checks provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods.
@@
 
-\highlight{\textbf{Finding 3.6} The study identifies three implicit checks for missing data.
+\highlight{\textbf{Observation 3.6} The study identifies three implicit checks for missing data.
 Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations.
@@
 
-\highlight{\textbf{Finding 3.7} The study identifies three implicit checks used to validate the shape of data.
+\highlight{\textbf{Observation 3.7} The study identifies three implicit checks used to validate the shape of data.
 These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
@@
 
-\highlight{\textbf{Finding 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data.
+\highlight{\textbf{Observation 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data.
 Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
@@
 
-\highlight{\textbf{Finding 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset.
+\highlight{\textbf{Observation 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset.
 These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training.
@@
 
-\highlight{\textbf{Finding 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model.
+\highlight{\textbf{Observation 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model.
 The print statement can be used to evaluate the efficiency of model training, especially when multiple models are trained iteratively.}
@@
 
-\highlight{\textbf{Finding 3.11} The study identifies one print statement used to show the architecture of a neural network.
+\highlight{\textbf{Observation 3.11} The study identifies one print statement used to show the architecture of a neural network.
 Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
@@
 Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers.
-Moreover, assertions may enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells~\citep{wang2020assessing}.
+Future work can focus on providing empirical evidence to demonstrate that the integration of assertions significantly improves the reliability of ML workflows in practice.
 
@@
 
-This study further reveals several avenues for research.
+This study further reveals several avenues for future research.
 We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them.
-Further research is required to explore the feedback mechanisms in more detail and organize the knowledge into a taxonomy based on specific ML validation tasks.
-Future work can focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
+Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks.
+It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering.
+Each task necessitates distinct feedback mechanisms.
+For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing.
+By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.
+
+Additionally, future work can also focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
 
@@
 Addressing these challenges would facilitate the stratification of our dataset into distinct categories---namely low, medium, and high complexity---which can enhance the depth of this study by enabling a more nuanced analysis of feedback mechanisms within this context.
+
+\subsection{Reproducibility of Jupyter Notebooks}
+
+Code written within Jupyter notebooks can reach a convoluted state since the code cells can be executed in any order.
+Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them.
+Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments~\cite{pimentel2019large-scale,wang2020assessing}.
+
+Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells~\citep{wang2020assessing}.
+Conversely, constructs like \lstinline{print("Accuracy:", accuracy)} are only effective if the cell defining \lstinline{accuracy} has been executed prior to the print statement.
+In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear.
+Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.
+
+However, there exists some ambiguity regarding how assertions interact with out-of-order execution in Jupyter notebooks.
+On one hand, our findings indicate that assertions can effectively prevent the system from reaching a convoluted state that may arise due to such execution.
 
@@
 These discussions continued until a consensus was reached on the interpretations and findings.
+
+Furthermore, we would like to reiterate that the case studies presented in RQ2 and RQ3 serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
 

### 2025-02-07 report/emse24-revised.tex
chore(report): emse24-revised updates

@@
 
-    To gain a deeper understanding of Python assertions written in Jupter notebooks, we conduct individual case-studies with 82 assert statements.
+    To gain a deeper understanding of Python assertions written in Jupyter notebooks, we conduct individual case-studies with 82 assert statements.
 We analyse the assertion along with the surrounding code to understand its purpose.
@@
 Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}.
-The libraries are selected based on the import analysis conducted by \citet{psallidas2019data} on 6 million Jupyter notebooks collected from GitHub.
+The libraries were chosen based on the import analysis conducted by \citet{psallidas2019data}, which analyzed six million Jupyter notebooks collected from GitHub.
+It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training.
+
 We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells.
@@
 
-\subsection{Case Studies}
-
-For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the case study analysis of all candidate assertions, print statements and last cell statements.
+\subsection{Exploratory Case Studies}
+
+For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements.
 To surface interesting candidates for the case studies, we apply text processing techniques as described below.
@@
 The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
+Moreover, our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
+Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.
 As data, personnel, and system environments inevitably change, undocumented decisions can lead to accumulation of technical debt.
@@
 
-Our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
-Unlike assertions, undocumented decisions based on implicit feedback mechanisms incur higher technical debt since they don't stop the execution despite violating the underlying assumptions of the code.
-
+% TODO: this paragraph needs work...
 We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations.
@@
 By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
-Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.
+Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation.
+The dataset contributed by this study provides a valuable resource to automatically generate documentaion from implicit and explicit feedback mechanisms written inside Jupyter notebooks~\cite{yang2021subtle}.
+
+% Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.
 
@@
 Additionally, we find several implicit checks performed using print and last cell statements to verify the presence of compute resources and libraries, monitor model training, and check the architecture of neural networks.
+Future work should focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
 
@@
 
-Additionally, future work can also focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by \citet{humbatova2020taxonomy} and \citet{morovati2023bugs} which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
-
 \subsection{Code Quality and Conformance to Standards}
@@
 
+% TODO: why is it important to analyze FMs wrt complexity of notebooks?
+% TBH I do not agree with the reviewer's comment about this...
 Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research.
@@
 These metrics however were primarily developed for code structured in an object-oriented programming paradigm.
-Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices \cite{pimentel2019large-scale}.
+Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices~\cite{pimentel2019large-scale}.
 
@@
 We enforced a time resource limit of 200 hours for conducting the case studies in RQ2 and RQ3, potentially constraining the depth and breadth of our investigation.
-The stratified random sampling used for selecting the case study candidates mitigates this threat by ensuring that a representative cross-section of the population is sampled within the time constraint.
+Although alternative stopping rules such as theoretical saturation are common, they are inherently subjective and can vary based on the judgement of a researcher.
+In contrast, a time-bound method ensures transparency and reflects the practical resource constraints in real-world research.
+Additionally, the stratified random sampling used for selecting the case study candidates mitigates this threat by ensuring that a representative cross-section of the population is sampled within the time constraint.
 While this sampling method enhances the generalizability of our results, the time limitation should still be considered when interpreting and applying the study's findings.
@@
 
-\paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks publicly available on GitHub and Kaggle.
-Consequently, the findings may not be generalizable to Python notebooks that do not concentrate on ML.
-Additionally, the results may not be applicable to notebooks authored in other statistical programming languages (such as R or Julia), or to proprietary notebooks hosted on different platforms.
+\paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks that are publicly available on platforms such as GitHub and Kaggle.
+As a result, the findings may not be generalizable to Python notebooks that do not center around ML applications.
+Furthermore, the applicability of the results is limited when it comes to notebooks authored in other statistical programming languages, such as R or Julia, or to proprietary notebooks hosted on different platforms.
+
+The analysis in this study is conducted only using Jupyter notebooks.
+However, the underlying workflow of users across all notebook environments is identical---that is, users execute chunks of code using a REPL mechanism.
+As such, the findings of this study may be generalizable to other notebook technologies.
 
@@
 Despite these advancements, there is a notable gap in practical, day-to-day testing practices within ML development environments.
-Our study addresses this gap by analysing the feedback mechanisms in Jupyter notebooks, highlighting how assertions and print statements are used to validate data and model performance, and proposing ways to integrate these practices more effectively into the ML development lifecycle.
+To the best of our knowledge, \citet{openja_empirical_2024} is the only study that empirically investigates the test code utilized within 11 ML projects spanning four different application domains.
+Our study builds upon the findings of \citet{openja_empirical_2024} by further analyzing the feedback mechanisms within Jupyter notebooks.
+We highlight how assertions and print statements are employed to validate data and model performance, and propose strategies for integrating these practices more effectively into the ML development lifecycle.
 

### 2025-02-13 report/latex-diff.sh
tool to diff latex files and output diff pdf

@@ -1,0 +1,26 @@
+#!/bin/bash
+
+# Check if two arguments are provided
+if [ "$#" -ne 2 ]; then
+    echo "Usage: $0 <file1.tex> <file2.tex>"
+    exit 1
+fi
+
+FILE1=$1
+FILE2=$2
+DIFF_FILE="diff.tex"
+
+# Expand LaTeX files (in case they use \input or \include)
+latexpand "$FILE1" > old.tex
+latexpand "$FILE2" > new.tex
+
+# Generate the diff file using latexdiff
+latexdiff old.tex new.tex > "$DIFF_FILE"
+
+# Compile the diff.tex file to generate the PDF
+latexmk -pdf -f "$DIFF_FILE"
+
+# Clean up temporary files
+rm old.tex new.tex "$DIFF_FILE"
+
+echo "PDF diff generated successfully!"

### 2025-02-14 report/reviews/1st round/response.tex
first draft of response with comments from reviewer 1

\documentclass[11pt,fleqn]{article}

\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{2.5cm}{2.5cm}{2.5cm}{2.5cm}%
{\baselineskip}{2\baselineskip}{baselineskip}{2\baselineskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{epsfig}
\usepackage{latexsym}
\usepackage{url}
\usepackage{tgheros}
\renewcommand*\familydefault{\sfdefault}

\newcommand{\eline}{\vspace*{.75\baselineskip}}
\newcommand{\Referee}[1]{\eline \noindent {\bf Reviewer comment #1:} \\}
\newcommand{\Us}{\eline \noindent {\bf Response:}\\}
\newcommand{\TBD}{{\bf To Be Done}}
\newcommand{\newreviewer}[1]{\section*{Reviewer #1}\vspace*{-1.05\baselineskip}}

%\usepackage{tikz}
\usepackage[skins,breakable]{tcolorbox}
\tcbset{textmarker/.style={
%
skin=enhancedmiddle jigsaw,breakable,parbox=false,before skip=-1mm, after skip=0mm,
boxrule=0mm,leftrule=2mm,rightrule=2mm,boxsep=0mm,arc=0mm,outer arc=0mm,
left=2mm,right=2mm,top=2mm,bottom=2mm,toptitle=1mm,bottomtitle=1mm,oversize}}

\newtcolorbox{rcomment}{textmarker,colback=gray!15!white,colframe=gray!80!white}

\newenvironment{revcomment}[1][]
{\Referee{#1}\begin{rcomment}}
{\end{rcomment}}

\usepackage{todo}
\usepackage{hyperref}

%customized
\usepackage{xfrac}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent\emph{#1}
\end{framed}}

\expandafter\def\expandafter\quote\expandafter{\quote\itshape}

\graphicspath{{../../}}

\usepackage{xspace}
\newcommand{\ING}{ING\xspace}

% more lenient line breaking (avoids text protruding into margins)
\sloppy

\title{\vspace*{-2cm}{\bf Authors' Response to the Review of
 EMSE-D-24-00354:\\
 ``Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks''}}

\author{Arumoy Shome, Lu\'{i}s Cruz, Diomidis Spinellis, Arie van Deursen}
\date{}

\begin{document}

\maketitle

\newreviewer{Editor}

\begin{revcomment}

Despite the paper has significant strengths, there are numerous significant weaknesses that go beyond the process and period of a major revision. Specifically, there are serious concerns about the sampling procedure and the missing details about the inter-rater reliability analysis and recommendations to practitioners.

Since this paper interests EMSE, I suggest the authors revise and resubmit it as new. When submitting as new, if the authors ask to be assigned to the same editor, I will do my best to assign it to the same reviewers to speed up the revision process.

\end{revcomment}

\Us We thank the editor and reviewers for their valuable feedback -- it helped us improve the paper in this new submission. In the following, we address point-by-point the concerns raised by the reviewers.

\newreviewer{1}

\begin{revcomment}[1.1]
The paper presents an insightful exploration of feedback mechanisms within the machine learning (ML) development lifecycle, focusing specifically on Jupyter notebooks authored in Python. By conducting an analysis of ~297.8 thousand notebooks, the authors identify three key feedback mechanisms: assertions as a form of explicit feedback; print statements and last cell statements as forms of implicit feedback. The findings underscore a widespread reliance on implicit feedback mechanisms in Jupyter notebooks. Despite their utility for automated validation, assertions are found to be underutilized, highlighting a significant gap in current testing practices.

\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[1.2]

To date, Jupyter notebooks are pervasively used in the ML development lifecycle, especially in the data preparation and model training phases. Moreover, this year, Python overtook JavaScript as the most popular programming language on GitHub, underscoring a surge in open-source data science and ML projects on the platform [1]. Hence, investigating how data scientists leverage Python Jupyter notebooks in their daily work is a timely and relevant research endeavor, which could inform the evolution of Jupyter Notebook implementations. Also, understanding the nature of feedback mechanisms used in notebooks paves the way to improve overall testing and QA practices in data science projects. In addition, as the authors suggest, this understanding can be useful for optimizing the transition from notebook-based prototyping to production-grade ML code [2].

The dataset contributed by this study—comprising assertions, print statements, and last-cell statements from notebooks—further enhances its relevance and utility, offering a rich resource for future research.

Concerning the final discussion, I wonder whether - based on their findings - the authors have thought of any recommendations for designers of notebook platforms (e.g., Jupyter Notebook, Jupyter Lab, Google Colaboratory, etc.). Integrating such recommendations could potentially enhance the practical relevance of this study.

\end{revcomment}

\Us Revised Section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook designers and developers.

Based on that, we have updated the manuscript with the following text (cf. Section XX):

\begin{quote}
GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.  
\end{quote}

\Todo{write response based on text above.}
\Todo{refer to the section where the text was updated (XX).}

\begin{revcomment}[1.3]

To the best of my knowledge, this study is original and it is the first to explore feedback mechanisms in ML Python Jupyter notebooks. Moreover, the work is well-situated within the existing literature, offering a thoughtful background and context for its findings.

\end{revcomment}

\Us 
\Todo{say something nice.}

\begin{revcomment}[1.4]
The methodology employed in this study generally appears robust and well-structured.

To address RQ1, the authors undertake an archival study drawing on two distinct sources: GitHub and Kaggle. The unit of analysis, selection criteria for notebooks, and data collection processes are all clearly delineated. However, I suggest that the authors provide additional context about the source platforms for the benefit of researchers who may not be familiar with them - especially Kaggle. Indeed, GitHub and Kaggle have different purposes and present significantly different environments for notebook authoring and sharing.
\end{revcomment}

\Us Revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github.

Based on that, we updated the manuscriptin with text below (cf. Section XX):

\begin{quote}
GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.
\end{quote}

\Todo{expand the first line into a response.}

\begin{revcomment}[1.5]

To answer RQ2 and RQ3, the authors delve in a manual analysis of a sample of notebooks, allocating a budget of 200 hours to this task. However, I have the following concerns regarding this procedure.

-  The manual task constitutes qualitative coding/annotation, which is inherently subjective. Nonetheless, only a single annotator (the first author) has been involved unless they were "unable to determine the purpose of the candidate," in which case it would be "additionally analyzed by the second and third author." However, for tasks of this type, it is usually advised to involve multiple annotators - at least to code a first, pilot subset of examples, thereby validating the annotation strategy through inter-rater reliability metrics.  I recommend that the authors elaborate more on how they addressed potential subjective bias within this annotation process.

\end{revcomment}

\Us

Section 3.3 Case studies was revised to provide additional details regarding how the exploratory case studies were conducted.

Additionally, to validate the feedback mechanism taxonomy created by the first author, the second author independently annotated the case study candidates using the codes developed by the first author. We have an inter-rater agreement score of X (highly agreeable).

\Todo{start with an agreeable statement}
\Todo{add revised text from Section 3.3, the same way we do in response 1.4}

\begin{revcomment}[1.6]

While it's reasonable to set a predefined time budget for qualitative analysis, and the sampling strategy used seems sound, the number of analyzed notebooks appears quite limited. Consequently, several findings are drawn from an exceedingly small number of ``case studies,'' sometimes as few as one. Although the exploratory nature of this study is acknowledged, deriving generalizable conclusions from such isolated instances is challenging. Therefore, it would be advisable for the authors to underscore this limitation more prominently in the Threats to Validity section, and reconsider presenting cases based solely on a single example as definitive findings.
\end{revcomment}

\Us

Revised results section to emphasize the exploratory nature of the study. We also explicitly mentioned that due to the qualitative nature of the study, the results should be interpreted with caution.

Additionally we replaced the word ``Finding'' with ``Observation'' in all summary boxes.

In the Threats to validity section, we remind the reader that the results should not be interpreted as generalizable results.

\Todo{explain why we replace Finding by Observation -- don't expect people to get it immediately if you don't explain why. }

\begin{revcomment}[1.7]
Overall, the paper is well-written and well-organized. The explanations provided are clear. The images and charts are well-crafted, providing valuable visual support to the text and enhancing comprehension.

However, the methodology section lacks specific details about the processes and steps involved in addressing RQ1.
\end{revcomment}

\Us

We have to say that we started from observations and patterns of use of notebooks from prior experience and what you typically see in books, courses and videos. Then to validate, we used X methodology...

“Clever way” to say asserts, prints and last cell statements are common patterns in notebooks

OR

Forget about it...

“ we considerably improved the methodology section...” and include text from other responses to comments

\Todo{work on response based on text above}

\begin{revcomment}[1.8]
Moreover, the authors have summarized each finding using a summary box; while this is potentially useful to improve the paper understandability, the number of these boxes is high and, in a few cases, the summarized text is shorter than the summary itself (see Sect. 4.3.3) or takes up a comparable space (see Sect 4.3.5). The result is a certain degree of repetition, while it remains challenging to keep the big picture.
\end{revcomment}

\Us
We tried our best to reduce the size of the summary boxes, especially where the boxes are the same size of larger than the main text itself. Almost all summary boxes were revised, please see the diff document for the exact changes.

\begin{revcomment}[1.9]

On a minor note, from Fig. 2 it is not clear how many notebooks were left after applying the filtering criteria.
\end{revcomment}

\Us
Figure 2 updated to reflect how many notebooks were left out after the filters.

On second thought, avoid taking action because getting this number is proving to be tricky...

\begin{revcomment}[1.10]

The methodology adopted in this study is thoroughly described. Moreover, the authors have released the dataset and replication package of this paper on Figshare, under a CC-BY license. Overall, I believe the study should be quite easy to replicate.

An additional appendix has been shared by the authors, containing the most representative examples referenced in the article. However, this appendix is hosted online on a personal website, rather than being incorporated directly at the end of the paper or included with the replication package. Given that references to this appendix are distributed throughout the paper and considering that publications are ideally self-contained, incorporating the appendix within the paper itself would be preferable. If the appendix is too extensive to be included in full, I recommend publishing it as part of the replication package on Figshare, where its long-term accessibility and persistency can be ensured.
\end{revcomment}

\Us Online appendix added to the replication package as a PDF document.

\Todo{make answer more complete}

\begin{revcomment}[1.11]

\textbf{Minor issues}

\begin{itemize}

  \item - [page 4, lines 44-46] "Jupyter notebooks […] leverage the IPython kernel to execute Python code and extends its capabilities […]" => "extend"

  \item - [page 5, Fig. 1] "Content of the cell are in the source key" => "The content of the cell is in the source key" as well as "The raw content of outputs that produce text are under the text key." => "… is …"

  \item - [page 10, lines 31-33] "Methods such as […] were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures." => "for comparing" or "to compare"

  \item - [page 13, lines 33-34] "Figure 16 show" => "Figure 16 shows"

  \item - [page 26, lines 35-36] "Catching data errors are critical" => "Catching data errors is critical"

  \item - Section 5.3: in addition to the course by Kästner and Kang, other courses covering MLOps and Software Engineering for ML are currently fostering ML testing literacy among data science students. Examples are the course on "Software Engineering for AI-enables Systems" [2, 3] held at the University of Bari (Italy) and at the Universitat Politècnica de Catalunya (Spain), as well as the course "Machine Learning Systems Design" held at ULiege (Belgium) [4].

\end{itemize}
\end{revcomment}

\Us We thank the reviewer for flagging all these minor issues. We have addressed all of them with the new version of the manuscript and hope our improvements meet your expectations.

\Todo{Address all minor issues}

%%%%%%%%%%%%%%%%%%%
\newreviewer{2}

\newreviewer{3}

\todos

\end{document}

### 2025-02-21 report/emse24-response.tex
feat: add EMSE response document
And cleanup the file structure (I like a flat directory structure).

\documentclass[11pt,fleqn]{article}

\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{2.5cm}{2.5cm}{2.5cm}{2.5cm}%
{\baselineskip}{2\baselineskip}{baselineskip}{2\baselineskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{epsfig}
\usepackage{latexsym}
\usepackage{url}
\usepackage{tgheros}
\renewcommand*\familydefault{\sfdefault}

\newcommand{\eline}{\vspace*{.75\baselineskip}}
\newcommand{\Referee}[1]{\eline \noindent {\bf Reviewer comment #1:} \\}
\newcommand{\Us}{\eline \noindent {\bf Response:}\\}
\newcommand{\TBD}{{\bf To Be Done}}
\newcommand{\newreviewer}[1]{\section*{Reviewer #1}\vspace*{-1.05\baselineskip}}

%\usepackage{tikz}
\usepackage[skins,breakable]{tcolorbox}
\tcbset{textmarker/.style={
%
skin=enhancedmiddle jigsaw,breakable,parbox=false,before skip=-1mm, after skip=0mm,
boxrule=0mm,leftrule=2mm,rightrule=2mm,boxsep=0mm,arc=0mm,outer arc=0mm,
left=2mm,right=2mm,top=2mm,bottom=2mm,toptitle=1mm,bottomtitle=1mm,oversize}}

\newtcolorbox{rcomment}{textmarker,colback=gray!15!white,colframe=gray!80!white}

\newenvironment{revcomment}[1][]
{\Referee{#1}\begin{rcomment}}
{\end{rcomment}}

\usepackage{todo}
\usepackage{hyperref}

%customized
\usepackage{xfrac}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent\emph{#1}
\end{framed}}

\expandafter\def\expandafter\quote\expandafter{\quote\itshape}

\graphicspath{{../../}}

\usepackage{xspace}
\newcommand{\ING}{ING\xspace}

% more lenient line breaking (avoids text protruding into margins)
\sloppy

\title{\vspace*{-2cm}{\bf Authors' Response to the Review of
 EMSE-D-24-00354:\\
 ``Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks''}}

\author{Arumoy Shome, Lu\'{i}s Cruz, Diomidis Spinellis, Arie van Deursen}
\date{}

\begin{document}

% TODO: make sure to update all the quoted texts after updating the discussion section to mention the observations explicitly
\maketitle

\newreviewer{Editor}

\begin{revcomment}

  Despite the paper has significant strengths, there are numerous significant weaknesses that go beyond the process and period of a major revision. Specifically, there are serious concerns about the sampling procedure and the missing details about the inter-rater reliability analysis and recommendations to practitioners.

  Since this paper interests EMSE, I suggest the authors revise and resubmit it as new. When submitting as new, if the authors ask to be assigned to the same editor, I will do my best to assign it to the same reviewers to speed up the revision process.

\end{revcomment}

\Us We thank the editor and reviewers for their valuable feedback, it helped us improve the paper in this new submission. In this document, we address all concerns raised by the reviewers. We have added our response to each comment and added relevant text from the updated manuscript where necessary.

As part of the supplemental material for this submission, we have also included a document which contains the difference between the previous and current submission to make it easier to identify relevant parts of the manuscript that were updated in the new submission.

Please note that despite our best efforts, the exerpts from the manuscript that are included in this document may be out dated. This is because both this document and the manuscript are extensive and it was challenging to keep everthing in sync. Our sincere apologies for this inconvenience. We recommend that when in doubt, the reviewers consult the diff document for the most up to date changes that were made in the new submission.

\newreviewer{1}

\begin{revcomment}[1.1]
  The paper presents an insightful exploration of feedback mechanisms within the machine learning (ML) development lifecycle, focusing specifically on Jupyter notebooks authored in Python. By conducting an analysis of ~297.8 thousand notebooks, the authors identify three key feedback mechanisms: assertions as a form of explicit feedback; print statements and last cell statements as forms of implicit feedback. The findings underscore a widespread reliance on implicit feedback mechanisms in Jupyter notebooks. Despite their utility for automated validation, assertions are found to be underutilized, highlighting a significant gap in current testing practices.

\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[1.2]
  To date, Jupyter notebooks are pervasively used in the ML development lifecycle, especially in the data preparation and model training phases. Moreover, this year, Python overtook JavaScript as the most popular programming language on GitHub, underscoring a surge in open-source data science and ML projects on the platform [1]. Hence, investigating how data scientists leverage Python Jupyter notebooks in their daily work is a timely and relevant research endeavor, which could inform the evolution of Jupyter Notebook implementations. Also, understanding the nature of feedback mechanisms used in notebooks paves the way to improve overall testing and QA practices in data science projects. In addition, as the authors suggest, this understanding can be useful for optimizing the transition from notebook-based prototyping to production-grade ML code [2].

  The dataset contributed by this study—comprising assertions, print statements, and last-cell statements from notebooks—further enhances its relevance and utility, offering a rich resource for future research.

  Concerning the final discussion, I wonder whether - based on their findings - the authors have thought of any recommendations for designers of notebook platforms (e.g., Jupyter Notebook, Jupyter Lab, Google Colaboratory, etc.). Integrating such recommendations could potentially enhance the practical relevance of this study.
\end{revcomment}

\Us Thank you for your feedback. We have revised section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook users and developers. Here is an exerpt from the manuscript that was added to address this comment.

\begin{quote}
  We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations. Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently (Heer et al., 2010). By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle. Writing documentation from implicit feedback mechanims and keeping the documentation up-to-date requires significant manual effort. Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation. We hope that the dataset contributed by this study provides a valuable resource to develop and evaluate such automated tools.
\end{quote}

\begin{revcomment}[1.3]

  To the best of my knowledge, this study is original and it is the first to explore feedback mechanisms in ML Python Jupyter notebooks. Moreover, the work is well-situated within the existing literature, offering a thoughtful background and context for its findings.

\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[1.4]
  The methodology employed in this study generally appears robust and well-structured.

  To address RQ1, the authors undertake an archival study drawing on two distinct sources: GitHub and Kaggle. The unit of analysis, selection criteria for notebooks, and data collection processes are all clearly delineated. However, I suggest that the authors provide additional context about the source platforms for the benefit of researchers who may not be familiar with them - especially Kaggle. Indeed, GitHub and Kaggle have different purposes and present significantly different environments for notebook authoring and sharing.
\end{revcomment}

\Us Thank you for your feedback. We revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github. Following is an exerpt from the manuscript that was added to address this comment.

\begin{quote}
  Figure 2 illustrates the data collection process employed in this study to gather Jupyter notebooks written in Python from GitHub and Kaggle. GitHub is a popular platform to host and collaborate on software projects. It offers extensive tools for project management such as issue tracking, pull requests, code reviews and tooling for continuous integration and delivery. In contrast, Kaggle is focused specifically on data science and machine learning. It provides integrated computational resources, a rich repository of datasets, and a platform to host machine learning competitions. While GitHub supports formal project workflows, Kaggle encourages a more informal approach to data exploration and modeling, making it a preferred choice for data science practitioners looking for hands-on experience with developing machine learning models.
\end{quote}

\begin{revcomment}[1.5]
  To answer RQ2 and RQ3, the authors delve in a manual analysis of a sample of notebooks, allocating a budget of 200 hours to this task. However, I have the following concerns regarding this procedure.

  \begin{itemize}
  \item The manual task constitutes qualitative coding/annotation, which is inherently subjective. Nonetheless, only a single annotator (the first author) has been involved unless they were ``unable to determine the purpose of the candidate,'' in which case it would be ``additionally analyzed by the second and third author.'' However, for tasks of this type, it is usually advised to involve multiple annotators - at least to code a first, pilot subset of examples, thereby validating the annotation strategy through inter-rater reliability metrics. I recommend that the authors elaborate more on how they addressed potential subjective bias within this annotation process.
  \end{itemize}
\end{revcomment}

\Us Thank you for your valuable feedback. We made several adjustments to the manuscript to address this comment. First, we revised the title of section 3.2 Case Studies to \emph{Exploratory Case Study Analysis} to be more explicit about the scientific methodology used for RQ2 and RQ3. Furthermore, we revised this section to clarify our sampling technique, the potential bias it introduces and how we mitigate this threat. Finally, to mitigate subjective bias of the first author during the first round of annotations, the second author independently conducted a second round of annotations. We have updated the text to describe this process and reported the \emph{Cohen's Kappa} metric for inter-rater agreement.

Following is the entire text from section 3.2 Exploratory Case Study Analysis. We have include the entire section here since we made several revisions to this part of the manuscript.

\begin{quote}
  For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements. We specified Jupyter notebooks written in the Python programming language as the \emph{site} for the study. To surface interesting candidates for the study, we apply text processing and sampling techniques as described below.

  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

  To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.

  During each case study, the first author categorized each candidate using \emph{inducting coding}. The first author analysed the Python code of the candidate to understand its purpose. Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analysed to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.

  A taxonomy of feedback mechanisms observed in ML Jupyter notebooks was created using the categories generated during the exploratory case study analysis. To validate the taxonomy developed by the first author, the second author independently annotated the candidates using the categories developed by the first author. The inter-rater agreement was calculated using the \emph{Cohen's Kappa} method and a score 0.8 was obtained which is regarded as highly agreeable.
\end{quote}

\begin{revcomment}[1.6]

  While it's reasonable to set a predefined time budget for qualitative analysis, and the sampling strategy used seems sound, the number of analyzed notebooks appears quite limited. Consequently, several findings are drawn from an exceedingly small number of ``case studies'', sometimes as few as one. Although the exploratory nature of this study is acknowledged, deriving generalizable conclusions from such isolated instances is challenging. Therefore, it would be advisable for the authors to underscore this limitation more prominently in the Threats to Validity section, and reconsider presenting cases based solely on a single example as definitive findings.
\end{revcomment}

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. First, the title for section 3.2 Case Studies was revised to \emph{Exploratory Case Studies} to emphasize the exploratory nature of this study. Second, we replaced the word ``Finding'' with ``Observation'' in all summary boxes to remind the reader that the results presented in this study are not meant to be generalizable to a population since we conducted qualitative research. Please note that the diff document does not highlight the changes made to the summary boxes. We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience.

Finally, we revised the section 4 Results and section 6 Threats to Validity (External Validity) to remind the reader about the qualitative nature of this study.

Following is the text that was added to section 4 Results.

\begin{quote}
  Due to the exploratory nature of this study and the limited number of case studies analyzed, the observations presented should be interpreted with caution and viewed as illustrative rather than definitive. The case studies serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
\end{quote}

Following is the text that was added to section 6 Threats to Validity (External Validity).

\begin{quote}
  Furthermore, we would like to reiterate that the case studies presented in RQ2 and RQ3 serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
\end{quote}

% TODO: I don't fully agree with this comment. Revising the title for section 3.2 and changing the word ``Finding'' to ``Observation'' should be sufficient. Since we are using case study analysis, the results are not meant to be generalizable to a population. Furthermore, the criticism regarding only have one candidate in some categories is also invalid. By changing the manuscript, we admit that the fault is on our end which is not the case. How should we handle this?

\begin{revcomment}[1.7]
  Overall, the paper is well-written and well-organized. The explanations provided are clear. The images and charts are well-crafted, providing valuable visual support to the text and enhancing comprehension.

  However, the methodology section lacks specific details about the processes and steps involved in addressing RQ1.
\end{revcomment}

\Us Thank you for your feedback. We made have revised section 3.1 Data Collection to include more details regarding the steps involved in addressing RQ1. Specifically, we have separated the section into two sub sections. Section 3.1.1 Collection of ML Jupyter Notebooks contains text from the last version of the manuscript and describes the data collection procedure to collect the Jupyter notebooks relevant to this study. We added section 3.1.2 Collection of Feedback Mechanisms to the manuscript which describes the steps involved in extracting the feedback mechanisms from the notebooks in further detail.

Following is the text that was added to section 3.1.2 Collection of Feedback Mechanisms.

\begin{quote}
  For RQ1, we extracted the contents of all code cells from the Jupyter notebooks by parsing the underlying JSON structure. We then used the \texttt{ast} module provided by the Python programming language to extract the assert, print and last statements from all code cells.

  The raw contents of each cell is then converted into an AST. The \texttt{NodeVisitor} class from the \texttt{ast} module is used to traverse the AST and collect all \texttt{Assert} nodes. To collect assertions written using Python modules not included in the standard distribution, we also collected all function \texttt{Call} nodes that contain the string ``assert'' in their name. Similarly, all print statements were collected by traversing the AST to find all function \texttt{Call} nodes with the name ``print''. To collect the last statements, we first removed all cells that did not produce any output and consequently extracted the last top-level node from the AST of the code cell. The raw Python code was regenerated from the ASTs using the \texttt{unparse} method provided by the \texttt{ast} module.

  Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
\end{quote}

\begin{revcomment}[1.8]
  Moreover, the authors have summarized each finding using a summary box; while this is potentially useful to improve the paper understandability, the number of these boxes is high and, in a few cases, the summarized text is shorter than the summary itself (see Sect. 4.3.3) or takes up a comparable space (see Sect 4.3.5). The result is a certain degree of repetition, while it remains challenging to keep the big picture.
\end{revcomment}

\Us Thank you for your feedback. We tried our best to reduce the size of the summary boxes, especially where the boxes are the same size or larger than the main text itself. Almost all summary boxes were revised. Please note however that the diff document does not highlight the changes made to the summary boxes. We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document. Our sincere apologies for this inconvenience

\begin{revcomment}[1.9]
  On a minor note, from Fig. 2 it is not clear how many notebooks were left after applying the filtering criteria.
\end{revcomment}

\Us Thank you for your feedback. We updated Figure 2 to reflect how many notebooks were excluded and the number of notebooks left after the filters. We also updated section 3.1.1 Collection of Jupyter Notebooks to explicitly mention the number of notebooks that were excluded after applying the filters and the number of notebooks in the final dataset.

Following is an exerpt from the revised text of section 3.1.1 Collection of Jupyter Notebooks relevant to this comment.

\begin{quote}
  Since the focus of this study is on the analysis of Python code, [\ldots] 107 thousand notebooks were excluded after applying the above filters, leaving a dataset 190.8 thousand notebooks.
\end{quote}

% TODO: update the figure

\begin{revcomment}[1.10]

  The methodology adopted in this study is thoroughly described. Moreover, the authors have released the dataset and replication package of this paper on Figshare, under a CC-BY license. Overall, I believe the study should be quite easy to replicate.

  An additional appendix has been shared by the authors, containing the most representative examples referenced in the article. However, this appendix is hosted online on a personal website, rather than being incorporated directly at the end of the paper or included with the replication package. Given that references to this appendix are distributed throughout the paper and considering that publications are ideally self-contained, incorporating the appendix within the paper itself would be preferable. If the appendix is too extensive to be included in full, I recommend publishing it as part of the replication package on Figshare, where its long-term accessibility and persistency can be ensured.
\end{revcomment}

\Us Thank you for your feedback. We have added the appendix to the replication package as a PDF document.

\begin{revcomment}[1.11]

  \textbf{Minor issues}
  \begin{enumerate}
  \item page 4, lines 44-46: ``Jupyter notebooks […] leverage the IPython kernel to execute Python code and extends its capabilities […]'' => ``extend''
  \item page 5, Fig. 1: ``Content of the cell are in the source key'' => ``The content of the cell is in the source key'' as well as ``The raw content of outputs that produce text are under the text key.'' => ``… is …''
  \item page 10, lines 31-33: ``Methods such as […] were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.'' => ``for comparing'' or ``to compare''
  \item page 13, lines 33-34: ``Figure 16 show'' => ``Figure 16 shows''
  \item page 26, lines 35-36: ``Catching data errors are critical'' => ``Catching data errors is critical''
  \item Section 5.3: in addition to the course by Kästner and Kang, other courses covering MLOps and Software Engineering for ML are currently fostering ML testing literacy among data science students. Examples are the course on "Software Engineering for AI-enables Systems" [2, 3] held at the University of Bari (Italy) and at the Universitat Politècnica de Catalunya (Spain), as well as the course "Machine Learning Systems Design" held at ULiege (Belgium) [4].
  \end{enumerate}
\end{revcomment}

\Us Thank you for identifying the minor issues. We have addressed them in the new version of the manuscript.

\newreviewer{2}
\begin{revcomment}[2.1]
  The paper presents an extensive analysis of feedback mechanisms in Jupyter notebooks used for machine learning workflows. By mining a significant dataset of 297.8 thousand notebooks and examining 2.3 million code cells, the authors address a critical area of research that has not received sufficient attention. The paper is well-structured, with clear research questions and a logical progression from dataset construction to findings and recommendations. It offers important insights into the role of feedback mechanisms. However, while the foundation of the work is strong, there are areas where methodological and analytical improvements could further enhance the depth and applicability of the findings.
\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[2.2]
  \textbf{Strengths of the Paper}
  \begin{enumerate}
  \item Novelty of the topic: the paper tackles an important issue in machine learning development. By focusing on Jupyter notebooks, a widely used tool in the ML community, the study addresses a highly relevant and practical domain.
  \item Thorough dataset and methodology: the authors demonstrate rigor in dataset construction, analyzing a large sample of notebooks from GitHub and Kaggle. The inclusion criteria—valid JSON notebooks with at least one code cell and imports from key ML libraries—ensure relevance to ML workflows. This filtering step is described clearly.
  \item Categorization of feedback mechanisms: by identifying three key feedback types and further categorizing them into implicit (e.g., print statements) and explicit (e.g., assertions), the paper provides a structured lens for understanding feedback practices in notebooks. The distinction between these categories is explained effectively and forms the basis of the research questions.
  \item Well-organized results: the findings are presented in a clear, systematic manner, with subsections addressing each research question. This level of detail helps readers understand the practical use cases of each mechanism.
  \item Replicability: the authors share a public dataset and replication package. This commitment to transparency strengthens the study's credibility and encourages follow-up research.
  \end{enumerate}
\end{revcomment}

\Us Thank you for your valuable comments.

\begin{revcomment}[2.3]
  \textbf{Areas for improvement}

  While the paper is well-organized and insightful, several areas could benefit from additional attention to further strengthen its contributions.

  1. Dataset and sampling methodology

  The dataset includes all notebooks with at least one code cell, but it does not provide a breakdown of the number of cells per notebook. This is a limitation because single-cell notebooks likely lack meaningful inter-cell feedback dependencies, making them fundamentally different from multi-cell workflows where dependencies and iterative processes play a crucial role. Presenting a histogram or similar visualization of the number of cells per notebook would contextualize the dataset, enabling readers to better understand its structure.
\end{revcomment}

\Us Thank you for your feedback. We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information regarding the distribution of the number of cells per notebook. Since the data is really large and contains many outliers, we categorized the notebooks into small, medium and large based on the number of cells they contain. Additionally, we also added figures 4, 5 and 6 that visualize the distibution of the number of cells and feedback mechanisms across the notebook size categories.

Following is the text from section 4.1 Descriptive Statistics which was added to the manuscript to address this comment.

\begin{quote}
  Figure 4 shows the distribution of the number of code cells for all notebooks analyzed in this study. The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles. This is done since the data is large, non-uniform and contains many outliers. Figure 5 further magnifies the distribution of the number of cells within each category. The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.

  Figure 6 shows the distribution of assertions, prints and last cell statements across the different notebook size categories. A logarithmic scale is used for the count since the data is large and contains outliers. Overall, the visualization suggests an increase in the use of all feedback mechanisms as notebook size grows. Print statements are the most frequently used across all notebook sizes.
\end{quote}

\begin{revcomment}[2.4]
  Additionally, stratifying the dataset based on notebook complexity (e.g., single-cell, small-scale, medium-scale, and large-scale notebooks) and analyzing feedback mechanisms within these clusters would enhance the study's depth.

  For instance:
  \begin{itemize}
  \item Single-cell notebooks might represent quick experiments or incomplete workflows, potentially over-relying on print statements for debugging.
  \item Small- to medium-scale notebooks could exhibit more balanced feedback mechanisms, such as assertions validating intermediate results.
  \item Large-scale notebooks might show structured practices with complete feedback mechanisms aimed at reproducibility and production-readiness.
  \end{itemize}

  By analyzing whether the findings hold consistently across these clusters, the authors could find patterns specific to notebook complexity. This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity. Including this stratified analysis would significantly strengthen the study's contributions.
\end{revcomment}

\Us Thank you for your feedback. We have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work. While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity. We agree that the number of cells (or size) of a notebook is a valid albeit not the only metric that should be considered to define notebook complexity.

Following is the text that was added to section 5.6 Complexity of Notebooks to address this comment.

\begin{quote}
  Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research. We believe that the primary challenge here will be to develop an appropriate definition for notebook complexity.

  To the best of our knowledge, no formal definition or metrics to quantify notebook complexity exist to date. In their research, Grotov et al. (2022) investigated the complexity of Python code written inside Jupyter notebooks using a subset of 15 metrics derived from prior literature. These metrics however were primarily developed for code structured in an object-oriented programming paradigm. Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices (Pimentel et al., 2019).

  It is further unclear how metrics such as the size of the training data, complexity of the learning algorithm, computational costs and CO\(_2\) emissions associated with training can be incorporated into the definition for notebook complexity. Addressing these challenges would facilitate the stratification of our dataset into distinct categories---namely low, medium, and high complexity---which can enhance the depth of this study by enabling a more nuanced analysis of feedback mechanisms within this context.

\end{quote}

\begin{revcomment}[2.5]
  The use of TF-IDF scores to sample feedback mechanisms for case studies prioritizes textual uniqueness. This approach risks overemphasizing rare but trivial feedback, such as simple debugging statements (print("Hello world")), while potentially underrepresenting more common but functionally critical feedback. Incorporating functional complexity or analyzing interdependencies between feedback mechanisms and other cells would improve sampling relevance.
\end{revcomment}

\Us Thank you for your feedback. We have revised section 3.2 Exploratory Case Study Analysis to explicitly mention the bias introduced by using TF-IDF and how subsequently the stratified random sampling is used to mitigate this threat.

Following is the relevant text from the revised version of section 3.2 Exploratory Case Study Analysis which was added to address this comment.

Please also see our response to comment 1.5 and the diff document for other revisions that were made to section 3.2 Exploratory Case Study Analysis.

\begin{quote}
  For RQ2 and RQ3, we allocated a fixed [\ldots].

  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

  To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.
\end{quote}

\begin{revcomment}[2.6]
  Machine learning workflows vary significantly by task (e.g., classification, regression, clustering). Different tasks require different feedback mechanisms—for example, assertions about image dimensions in computer vision or data distribution checks in NLP. Analyzing the dataset by ML task type would provide targeted insights and make the recommendations more actionable for practitioners.
\end{revcomment}

\Us Thank you for your feedback. We have revised section 5.4 Machine Learning Bugs to discuss feedback mechanisms and machine learning workflows.

Following is the relevant text added to section 5.4 Machine Learning Bugs.

\begin{quote}
  The findings of this study align with [\ldots].

  This study further reveals several avenues for future research. We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them. Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks. It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering. Each task necessitates distinct feedback mechanisms. For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing. By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.
\end{quote}

\begin{revcomment}[2.7]
  2. Feedback mechanisms and their context

  Jupyter notebooks allow for out-of-order execution, which impacts the effectiveness of feedback mechanisms. A statement such as print("Accuracy:", accuracy) is only useful if the cell defining accuracy has been executed first. The paper partially explore how such dependencies affect the functionality of feedback mechanisms, leaving a gap in the analysis.

\end{revcomment}

\Us Thank you for your feedback. We have added section 5.7 Reproducibility of Jupyter Notebooks to improve the discussion on feedback mechanisms and how their functionality is affected during out-of-order execution of cells.

Following is the text that was added to section 5.7 Reproducibility of Jupyter Notebooks.

% TODO: update this response once the section is finalized
\begin{quote}
  Code written within Jupyter notebooks can reach a convoluted state since the code cells can be executed in any order. Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them. Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments (Pimentel et al., 2019; Wang et al., 2020b).

  Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells (Wang et al., 2020b). Conversely, constructs like \texttt{print("Accuracy:", accuracy)} are only effective if the cell defining accuracy has been executed prior to the print statement. In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear. Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.
\end{quote}

\begin{revcomment}[2.8]
  The paper treats all print statements and last cell outputs as feedback mechanisms. However, many of these may be incidental, such as debugging statements or placeholders that do not provide actionable feedback. Distinguishing intentional feedback from debugging artifacts would refine the analysis. What about introducing criteria for identifying feedback mechanisms that actively influence decision-making versus those used for temporary debugging?
\end{revcomment}

\Us Thank you for your feedback. As per our original definition of feedback mechanisms from section 1 Introduction, all print statements are considered as feedback mechanisms in this study. Regardless, we have revised the definition of feedback mechanisms in the introduction and we hope that this is sufficient to address this comment.

Following is the revised text from section 1 Introduction that was added to address this comment.

\begin{quote}
  Computational notebooks---specifically Jupyter notebooks---have been ubiquitously adopted by the ML community for developing ML enabled systems (Pimentel et al., 2019; Quaranta et al., 2021; Psallidas et al., 2019; Perkel, 2018). Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks. Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development. This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development. This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms. Users can test hypotheses, debug issues, and make incremental changes with immediate feedback. The ability to visualize data and results inline and document the process in a narrative form further enhances their utility (Kery et al., 2018; Head et al., 2019; Rule et al., 2018; Chattopadhyay et al., 2020).

  Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues. Feedback mechanims are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program. Importantly, not all forms of feedback quality as feedback mechanisms. For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.

  Notebook users can obtain feedback using implicit or explicit mechanisms. Implicit feedback mechanisms require manual validation by the user. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

  In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \texttt{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

  It is important to contextualize feedback mechanisms within the broader framework of software testing. Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality. While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm. For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook. While such implicit feedback types can help practitioners diagnose issues within their code, they necessitate manual validation and interpretation, which can increase the potential for oversight and errors.
\end{quote}

\begin{revcomment}[2.9]
  The authors position assertions as superior to implicit mechanisms like print statements. While this aligns with software engineering principles, no empirical evidence is provided to show that assertions lead to more reliable workflows in practice. For instance, assertions may be less suitable for exploratory phases, where rapid visual feedback is often preferred.
\end{revcomment}

\Us Thank you for your feedback. We acknowledge that the current study does not provide empirical evidence to support the claim that explicit feedback mechanisms are superior to their implicit counterparts. However, the manuscript does support this statement with sufficient theory building found in section 1 Introduction and extensively throughout section 5 Discussion. Regardless, we have revised section 5.2 Automated Data Validation to explicitly state this as future work.

Following is an exerpt from section 5.2 Automated Data Validation added to address this comment.

\begin{quote}
  Breck et al. (2017) view ML training as analogous to [\ldots]

  The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged. This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Future work can focus on providing empirical evidence to demonstrate that the integration of assertions significantly improves the reliability of ML workflows in practice.

  Results from RQ2 and RQ3 show that assertions, print statements and last cell statements are predominantly [\ldots]
\end{quote}

\begin{revcomment}[2.10]
  3. Recommendations and practical relevance

  Development phase considerations: feedback needs differ across development stages. For instance, print statements may be more valuable during early exploration, while assertions are more appropriate for production workflows. Recommendations that account for these stages would make the findings more useful.
\end{revcomment}

\Us Thank you for your feedback. We agree with this point of view that as ML pipelines developed inside Jupyter notebooks mature, the implicit feedback machanisms should be converted to explicit feedback mechanisms as per software engineering best practices. Here is an exerpt from section 1 Introduction of the manuscript that motivates this recommendation.

\begin{quote}
  As machine learning pipelines developed within Jupyter notebooks become business-critical, they are frequently transitioned into automated scripts for production environments (Kery et al., 2018; Rule et al., 2018). In these contexts, it is crucial to shift from implicit feedback mechanisms to explicit ones. This transition reduces the reliance on manual verification, thereby enhancing the robustness and maintainability of ML pipelines.
\end{quote}

We also believe that existing feedback mechanisms in the notebooks (implicit and explicit) can make the transition from notebooks into a production environment smoother. In section 5.1 Documentation and Technical Debt we highlight how artifacts---such as requirements, rationale and design decisions and tests---which are a pre-requisite for all operational software systems, can be effectively derived from the feedback mechanisms used inside notebooks. Furthermore, in section 5.2 Automated Data Validation, we propose that assertions can be used to derive the data validation schema for dedicated data validation tools, which are often incorporated into operational ML pipelines.

Following are the relevant exerpts from section 5.1 Documentation and Technical Debt and 5.2 Automated Data Validation respectively that support the above statement.

\begin{quote}
  Jupyter notebooks are primarily used to explore different ideas or approaches [\ldots] In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks (Pimentel et al., 2019; Psallidas et al., 2019; Grotov et al., 2022).

  Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks. The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies. Moreover, our results from RQ3 reveal [\ldots]
\end{quote}

\begin{quote}
  Results from RQ2 and RQ3 show that assertions, [\ldots] We argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
\end{quote}

We sincerely hope that the clarification provided here is sufficient to address this comment.

\begin{revcomment}[2.11]
  4. Presentation

  Dataset characteristics: detailed statistics on the dataset, such as cell distributions or feedback mechanism frequencies, would improve transparency. These insights would help readers assess the scope and generalizability of the findings.
\end{revcomment}

\Us Thank you for your feedback. We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information on the distribution of the number of cells and distribution of feedback mechanisms across various notebook size categories (small, medium and large notebooks).

You may also see our response to comment 2.3 and the diff document for further details.

\begin{revcomment}[2.12]
  While the authors acknowledge constraints, such as the reliance on public datasets, other limitations could be better discussed. For example, the absence of task-specific analysis or a lack of industrial datasets could affect the applicability of the findings. Addressing these explicitly would strengthen the paper's credibility.
\end{revcomment}

\Us Thank you for your feedback. We have made several revisions to the manuscript to address this comment. We have revised section 3.2 Exploratory Case Study Analysis to explicitly highlight the limitations and threats of our current approach and how we mitigate this threat. All subsections under section 5 Discussion have been updated to state limitations of the current work and potential directions for future work. Two new sections 5.6 Complexity of Notebooks and 5.7 Reproducibility of Notebooks were also added to the Discussion section. Finally, we have also revised section 6 Threats to Validity the address the limitations of the study explicitly.

You may also refer to our response to comments 1.5, 2.4 and 2.7 along with the diff document for additional details.

\begin{revcomment}[2.13]
  Summary of the suggestions

  \begin{itemize}
  \item present detailed statistics on the dataset, such as the distribution of cells per notebook and task annotations.
  \item analyze feedback patterns by ML task type to provide domain-specific insights.
  \item distinguish debugging statements from intentional feedback mechanisms.
  \item provide phase-specific recommendations for exploratory versus production workflows.
  \item better discuss broader limitations, including the dataset's reliance on public repositories and its implications for generalizability.
  \end{itemize}

  Overall, this paper is a well-organized and interesting exploration of feedback mechanisms in ML workflows. Its large-scale dataset and systematic approach provide a solid starting point for understanding feedback practices in Jupyter notebooks. The suggestions outlined in this review are intended as constructive feedback to further refine the study and explore additional dimensions that could enhance its applicability and depth. The paper addresses an important and underexplored area, and it offers important insights that will likely inspire future research in this domain.
\end{revcomment}

\Us Thank you for your valuable feedback and comments.

\newreviewer{3}
\begin{revcomment}[3.1]
  \textbf{Paper Summary}

  This paper researches how machine learning (ML) developers using Jupyter Notebooks implement feedback mechanisms during the ML development process. The authors identify two primary types of feedback: implicit and explicit. The authors indicate that implicit feedback is typically provided through print statements and the output of the last cell in the notebook, while explicit feedback is provided by assert statements. To investigate these mechanisms, the authors analyzed a dataset of 297.8 thousand Jupyter Notebooks. They explain the process of gathering the data, and removing the duplicates. From this dataset, they closely examined 82 assert statements, 44 print statements, and 27 last cell outputs to further understand their purposes. The authors state that their findings reveal that implicit feedback is significantly more commonly used than explicit feedback. Additionally, the paper includes a detailed analysis of the most frequently used methods in assert statements, as well as the most common constructs employed in print statements and last-cell outputs. Finally, the author discusses some of the implications of their findings in regards to their implications with respect to different aspects like documentation and technical debt.

  \textbf{Strengths and Weaknesses}

  \textbf{Strengths:}
  \begin{itemize}
  \item The authors provide a detailed background, explaining several of the concepts introduced in the paper.
  \item The dataset used in the study is available online.
  \end{itemize}

  \textbf{Weaknesses:}
  \begin{itemize}
  \item The authors do not define the term feedback explicitly anywhere in the paper.
  \item Many claims are more general than they should be.
  \item The findings needs to have a deeper implication analysis earlier in the results section rather than later in the discussion.
  \item The reason for some study design decisions are not explained clearly like the limit of the analysis to 200 hours or use of TF-IDF for selection in the case studies section.
  \item The author claim that this is the first study to look into feedback mechanism in ML projects, they need to focus their claim on Juypter notebook rather than wider ML projects.
  \end{itemize}
\end{revcomment}

\Us Thank you for your valuable comments and feedback.

\begin{revcomment}[3.2]
  \textbf{Comments to Authors}

  I find that the paper, as it stands, doesn't read smoothly. The big picture is unclear, and I often find myself wondering, ``So what?'' It feels like the focus was too heavily placed on numbers and detailed nuances, with less attention given to the paper's narrative. For example, the term feedback is never explicitly defined. Additionally, I find that some claims in the paper, such as the one mentioned in the abstract, come across as unsupported. In addition, given that the paper's first keyword is SE4AI, I expected a stronger software engineering perspective on the process, like how the findings align or differ from best practices, what are some alternative practices not found in the results, or how effective these are mechanism and what are their limitation. Instead, the paper focuses primarily on numerical results and data analysis, with the discussion section addressing some of those aspects but not effectively.
\end{revcomment}

\Us Thank you for your feedback. We have done our best to address your comments below. We hope that our efforts satisfy your expectations.

\begin{revcomment}[3.3]
  \textbf{Main Questions}

  The following are the major concerns I have about the paper. I think the authors need to address these for the paper to be in good shape for acceptance.

  \begin{itemize}
  \item What exactly do the authors mean by feedback in the ML development process? How does it differ from testing or validation? Although the authors seem to adopt an ``implicit'' understanding of the term, I find this lack of clarity problematic. For instance, they observe ``the limited use of assertions, with only 8.5\% of the notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community.'' So, it seems they consider it from a testing perspective.
  \end{itemize}
\end{revcomment}

\Us Thank you for your feeback. We have revised section 1 Introduction to define feedback mechanims and discuss the similarities and differences between feedback mechanisms and formal testing.

Following is the text that was added to section 1 Introduction to address this comment. We sincerely hope that the new text is sufficient to address the concern raised in this comment.

\begin{quote}
  Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues. Feedback mechanims are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program. Importantly, not all forms of feedback quality as feedback mechanisms. For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.

  Notebook users can obtain feedback using implicit or explicit mechanisms. Implicit feedback mechanisms require manual validation by the user. For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical. This can be verified with a print statement in a code cell and manually checking the output. This code statement however, will continue to execute even if the data type of a feature changes in subsequent batches of training data.

  In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met. For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \texttt{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}. This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.

  It is important to contextualize feedback mechanisms within the broader framework of software testing. Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality. While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm. For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook. While such implicit feedback types can help practitioners diagnose issues within their code, they necessitate manual validation and interpretation, which can increase the potential for oversight and errors.
\end{quote}

\begin{revcomment}[3.4]
  \begin{itemize}
  \item For the print statements and last cell outputs, how do the authors determine that their purpose is related to feedback rather than other purposes, or do they consider all print statements as feedback mechanisms?
  \end{itemize}
\end{revcomment}

\Us Thank you for your feedback. Yes, as per the revised definition of feedback mechanisms in section 1 Introduction, all print statements are considered feedback mechanisms.

You may also see our response to comment 3.3 and the diff document for further details.

\begin{revcomment}[3.5]
  \begin{itemize}
  \item The authors need to justify their study choices more explicitly. For instance:
  \item Why limit the analysis to Jupyter Notebooks? While they mention their ubiquity, their claim seems to broadly target ML projects. If their scope is strictly Jupyter Notebooks, they need to focus their claims on those rather than the wider ML community.
  \end{itemize}
\end{revcomment}

\Us Thank you for your feedback.

After careful consideration, we belive that although the analysis in this study was conducted using Jupyter notebooks, the findings are applicable to other computational notebook technologies which follow the same REPL based paradigm of software development. We have therefore revised section 6 Threats to Validity (External Validity) to clarify the scope of this study.

Following is the revised text from section 6 Threats to Validity (External Validity). We hope that this revision is sufficient to address the concern raised in this comment.

\begin{quote}
  Our analysis is based on ML notebooks that are publicly available on platforms such as GitHub and Kaggle. As a result, the findings may not be generalizable to Python notebooks that do not center around ML applications. Furthermore, the applicability of the results is limited when it comes to notebooks authored in other statistical programming languages, such as R or Julia.

  The analysis in this study is conducted only using Jupyter notebooks. However, the underlying workflow of users across all notebook environments is identical---that is, users execute chunks of code using a REPL mechanism. As such, the findings of this study may be generalizable to other notebook technologies.
\end{quote}

\begin{revcomment}[3.6]
  \begin{itemize}
  \item Why include only notebooks using certain libraries? The authors reference another study, but it's unclear what that study found and how it informed their decision. Adding more detail here would strengthen their argument. They do explain this decision's implications in the threats to validity but it needs to be  explained earlier in the data-gathering part.
  \end{itemize}
\end{revcomment}

\Us Thank you for your feedback. We have revised section 3.1.1 Collection of ML Jupyter Notebooks in the manuscript to clarify the selection criteria using certain Python packages.

Following is the text that was added to section 3.1.1 Collection of ML Jupyter Notebooks to address this comment. We hope that this addition is sufficient to address the concern raised by this comment.

\begin{quote}
  Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.  Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn (Pedregora et al., 2011), Pytorch (Paszke et al., 2017), Tensorflow (Abadi et al., 2015) and Keras (Chollet et al., 2015).  The libraries were chosen based on the import analysis conducted by Psallidas at al., 2019, which analyzed six million Jupyter notebooks collected from GitHub. It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since they can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training. 107 thousand notebooks were excluded after applying the above filters, leaving a dataset of 190.8 thousand notebooks.
\end{quote}

\begin{revcomment}[3.7]
  I find significant issues with the case studies section. First, it does not include sufficient detail to qualify as a true case study; instead, it seems to me that it is an exploratory analysis of the purpose of some instances of the feedback mechanisms. Additionally, the author needs to better explain the rationale for using TF-IDF for selection, and the decision to limit the analysis to 200 hours seems arbitrary.
\end{revcomment}

\Us Thank you for your feedback. We made several adjustments to the manuscript to address this comment. We renamed the title of section 3.2 Case Studies to \emph{Exploratory Case Study Analysis} to be verbose about the methodology. We added more details regarding \emph{site} and the \emph{selection criteria} of the study. Furthermore, details regarding the sampling technique, the bias introduced, and our mitigation strategy has also been added. Finally, to validate the taxonomy of feedback mechanisms, the second author independently annotated all candidates using the codes that were generated during the first round of annotations. We have reported the inter-rater agreement which was calculated using the \emph{Cohen's Kappa} metric in the manuscript.

Following is the revised text from section 3.2 Exploratory Case Study Analysis. We sincerely hope that the revisions are sufficient to address the concerns raised by this comment.

\begin{quote}
  For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements. We specified Jupyter notebooks written in the Python programming language as the \emph{site} for the study. Given the exploratory nature of this study, we emphasized capturing a diverse range of candidates by apply text processing and sampling techniques as described below.

  The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed. Two stop words namely ``assert'' and ``print'' are removed since they appear in all assertions and print statements respectively. The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF). This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens. Consider for instance, the trivial print statement \texttt{print("hello, world!")}. Since this statement does not commonly appear in the context of machine learning Jupyter notebooks, it will have a high TF-IDF score but adds little substantive value to the analysis.

To address the potential bias introduced by this method, we employ stratified random sampling to select candidates for the case study analysis. The subgroups are created based on the aggregated TF-IDF values of the tokens in each candidate. The candidates are then divided into quartiles based on the aggregate value. As a result, all candidates characterized by rare tokens are grouped within a single bin. By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.

During each case study, the first author categorized each candidate using \emph{inducting coding}. The first author analysed the Python code of the candidate to understand its purpose. Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analysed to bring in rich context. For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors. Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.

A taxonomy of feedback mechanisms observed in ML Jupyter notebooks was created using the categories generated during the exploratory case study analysis. To validate the taxonomy developed by the first author, the second author independently annotated the candidates using the categories developed by the first author. The inter-rater agreement was calculated using the \emph{Cohen's Kappa} method and a score 0.8 was obtained which is regarded as highly agreeable.
\end{quote}

We have also revised section 6 Threats to Validity (Internal Validity) to discuss the pros and cons of using a fixed time resource. While we agree that the use of a fixed time resource has limitations, other stopping criterias, such as those based on theoretical saturation, are also subject to bias from the annotator. In contrast, a fixed time resource reflects the practical resource constraints that researchers face in the real world.

Following is the text that was added to section 6 Threats to Validity (Internal Validity) to address this comment. We hope that the revised text is sufficient to address this comment.

\begin{quote}
  We enforced a time resource limit of 200 hours for conducting the case studies in RQ2 and RQ3, potentially constraining the depth and breadth of our investigation. Although alternative stopping rules such as theoretical saturation are common, they are inherently subjective and can vary based on the judgement of a researcher. In contrast, a time-bound method ensures transparency and reflects the practical resource constraints in real-world research. Additionally, the stratified random sampling [\ldots]
\end{quote}

\begin{revcomment}[3.8]
  Some of the findings seem fairly intuitive such as ``Finding 1.3 Print statements are commonly used to display various performance metrics of ML models'' and Finding 1.4 ``Libraries and functions for data visualization, data exploration and machine learning are most common in last cell statement''. These findings align with standard practices in ML. I think the authors need to provide a deeper analysis of their implications. The author may consider examining when does these feedback mechanism happen in the development lifecycle, how effective they are, or what does these finding implies in regards to best practices in respect to software engineering for ML. I do see that they reference the findings in general in the discussion section, touching on their implications. Still, it is better to add specific implications for each finding in the results section.
\end{revcomment}

\Us Thank you for your feedback. While we agree that some observations may seem intuitive and obvious, we still report all observations of the study for completeness. We have made significant revisions to the discussion section to improve the analysis and provide deeper implications. To summarize, we revised all subsections under section 5 Discussion to include additional details such as limitations of the current work, potential directions for future work and implications for notebook users and developers. We also added two new subsections under section 5 Discussion to further enhance the implications of the study. We hope that these revisions meet your expectations.

You may also refer to our response for comments 2.9, 2.7, 2.6, 2.4 and 1.2 and the diff document for additional details.

\begin{revcomment}[3.9]
  The authors claim that there has been no prior work investigating feedback in ML projects. However, from a testing perspective, there has been relevant prior research, which the authors may have overlooked. It may be the first study to investigate them in Jupyter Notebook, though, so it is better to be specific. See for example Moses Openja, Foutse Khomh, Armstrong Foundjem, Zhen Ming (Jack) Jiang, Mouna Abidi, and Ahmed E. Hassan. 2024. An empirical study of testing machine learning in the wild. ACM Trans. Softw. Eng. Methodol.
\end{revcomment}

\Us Thank you for your feedback. We have revised section 7.3 ML Testing under related work to acknowledge the overlap with existing research efforts in ML testing.

Following is the text that was added to section 7.3 ML Testing to address this comment.

\begin{quote}
  Despite these advancements, there is a notable gap in practical, day-to-day testing practices within ML development environments. To the best of our knowledge, Openja et al. (2024) is the only study that empirically investigates the test code utilized within 11 ML projects spanning four different application domains. Our study builds upon the findings of Openja at al. (2024) by further analyzing the feedback mechanisms within Jupyter notebooks. We highlight how assertions and print statements are employed to validate data and model performance, and propose strategies for integrating these practices more effectively into the ML development lifecycle.
\end{quote}

\begin{revcomment}[3.10]
  \textbf{Questions related to the Results section:}

  \begin{itemize}
  \item The authors mention grouping candidates based on "their similarity to each other." How is similarity defined?
  \item The case study analysis process needs more detail. The authors note that the first author conducted the analysis, with additional analysis performed by the second and third authors when the author couldn't figure out the purpose of the feedback mechanism. How much of the data was analyzed solely by the first author versus collaboratively? Did the authors analyze the inter-rater agreement?
  \item Add a clearer explanation of the findings' implications.
  \end{itemize}
\end{revcomment}

\Us Thank you for your feedback. We have made several revisions to section 3.2 Exploratory Case Studies to address this comment. In summary, we added more details regarding the exploratory case study analysis and how the first round of analysis was conducted by the first author to create the taxonomy of feedback mechanisms. To validate the codes of the taxonomy, the second author independently did a second round of annotations. We calculated the inter-rater agreement using the \emph{Cohen's Kappa} metric (we report a score of 0.8, highly agreeable).

You may also refer to our response for comment 3.7 and the diff document for additional details.

\begin{revcomment}[3.11]
  \textbf{Questions related to the Discussion section:}

  Many of the implications stated in the discussion section are not directly backed up by the findings of the study or grounded in the existing literature. For example:

  \textbf{Documentation and Technical Debt:}

  \begin{itemize}
  \item The author's argument in this section is somewhat vague. They state that their study ``provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks.'' Make it clearer what these efforts are and how exactly their study contributes to them.
  \end{itemize}

  \textbf{Automated Data Validation:}

  \begin{itemize}
  \item The authors mention that ``the results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows.'' How do the results support this? Additionally, the authors suggest that these results show a ``missed opportunity'' in data validation tasks. What opportunities are they referring to? The authors need to clarify and justify their claims more.
  \end{itemize}
  \textbf{Machine Learning Bugs:}

  \begin{itemize}
  \item The discussion on ML bugs is unclear. What is the authors' main point here?
  \end{itemize}
\end{revcomment}

\Us Thank you for your feedback. We have made several revisions to section 5 Discussions to address the comments above. We hope that the changes satisfy your expectations.

\todos
\end{document}

### 2025-02-21 report/latex-diff.sh

@@ -1,26 +1,0 @@
-#!/bin/bash
-
-# Check if two arguments are provided
-if [ "$#" -ne 2 ]; then
-    echo "Usage: $0 <file1.tex> <file2.tex>"
-    exit 1
-fi
-
-FILE1=$1
-FILE2=$2
-DIFF_FILE="diff.tex"
-
-# Expand LaTeX files (in case they use \input or \include)
-latexpand "$FILE1" > old.tex
-latexpand "$FILE2" > new.tex
-
-# Generate the diff file using latexdiff
-latexdiff old.tex new.tex > "$DIFF_FILE"
-
-# Compile the diff.tex file to generate the PDF
-latexmk -pdf -f "$DIFF_FILE"
-
-# Clean up temporary files
-rm old.tex new.tex "$DIFF_FILE"
-
-echo "PDF diff generated successfully!"

### 2025-03-10 report/emse24-response.tex
chore: update EMSE response document

@@
 
-% TODO: make sure to update all the quoted texts after updating the discussion section to mention the observations explicitly
 \maketitle
@@
 
-Please note that despite our best efforts, the exerpts from the manuscript that are included in this document may be out dated.
-This is because both this document and the manuscript are extensive and it was challenging to keep everthing in sync.
+Please note that despite our best efforts, the excerpts from the manuscript that are included in this document may be outdated.
+This is because both this document and the manuscript are extensive, and it was challenging to keep everything in sync.
 Our sincere apologies for this inconvenience.
-We recommend that when in doubt, the reviewers consult the diff document for the most up to date changes that were made in the new submission.
+We recommend that when in doubt, the reviewers consult the diff document for the most up-to-date changes that were made in the new submission.
 
@@
 We have revised section 5.1 Documentation and Technical Debt to incorporate recommendation for notebook users and developers.
-Here is an exerpt from the manuscript that was added to address this comment.
+Here is an excerpt from the manuscript that was added to address this comment.
 
@@
 By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
-Writing documentation from implicit feedback mechanims and keeping the documentation up-to-date requires significant manual effort.
+Writing documentation from implicit feedback mechanisms and keeping the documentation up-to-date requires significant manual effort.
 Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation.
@@
 We revised Section 3.1 Data Collection to include more information regarding the distinction between Kaggle and Github.
-Following is an exerpt from the manuscript that was added to address this comment.
+Following is an excerpt from the manuscript that was added to address this comment.
 
@@
 
-Following is the entire text from section 3.2 Exploratory Case Study Analysis. We have include the entire section here since we made several revisions to this part of the manuscript.
+Following is the entire text from section 3.2 Exploratory Case Study Analysis. We have included the entire section here since we made several revisions to this part of the manuscript.
 
@@
 
-  During each case study, the first author categorized each candidate using \emph{inducting coding}.
-The first author analysed the Python code of the candidate to understand its purpose.
-Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analysed to bring in rich context.
-For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors.
+  During each case study, the first author categorized each candidate using \emph{inductive coding}.
+The first author analyzed the Python code of the candidate to understand its purpose.
+Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analyzed to bring in rich context.
+For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analyzed by the second and third authors.
 Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.
@@
 Second, we replaced the word ``Finding'' with ``Observation'' in all summary boxes to remind the reader that the results presented in this study are not meant to be generalizable to a population since we conducted qualitative research.
+
 Please note that the diff document does not highlight the changes made to the summary boxes.
@@
 \end{quote}
-
-% TODO: I don't fully agree with this comment.
-Revising the title for section 3.2 and changing the word ``Finding'' to ``Observation'' should be sufficient.
-Since we are using case study analysis, the results are not meant to be generalizable to a population.
-Furthermore, the criticism regarding only have one candidate in some categories is also invalid.
-By changing the manuscript, we admit that the fault is on our end which is not the case.
-How should we handle this?
 
@@
 \Us Thank you for your feedback.
-We made have revised section 3.1 Data Collection to include more details regarding the steps involved in addressing RQ1.
-Specifically, we have separated the section into two sub sections.
+We revised section 3.1 Data Collection to include more details regarding the steps involved in addressing RQ1.
+Specifically, we have separated the section into two subsections.
 Section 3.1.1 Collection of ML Jupyter Notebooks contains text from the last version of the manuscript and describes the data collection procedure to collect the Jupyter notebooks relevant to this study.
@@
 
-  The raw contents of each cell is then converted into an AST.
-The \texttt{NodeVisitor} class from the \texttt{ast} module is used to traverse the AST and collect all \texttt{Assert} nodes.
+  The raw contents of each cell were then converted into an AST.
+The \texttt{NodeVisitor} class from the \texttt{ast} module was used to traverse the AST and collect all \texttt{Assert} nodes.
 To collect assertions written using Python modules not included in the standard distribution, we also collected all function \texttt{Call} nodes that contain the string ``assert'' in their name.
@@
 
-  Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
+  Prior to conducting the analysis for RQ2 and RQ3, we removed duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
 \end{quote}
@@
 Please note however that the diff document does not highlight the changes made to the summary boxes.
-We tried several different latex implementations to generate the summary boxes but were unable to get it to show in the diff document.
-Our sincere apologies for this inconvenience
+We tried several latex implementations to generate the summary boxes but were unable to get it to show in the diff document.
+Our sincere apologies for this inconvenience.
 
@@
 \Us Thank you for your feedback.
-We updated Figure 2 to reflect how many notebooks were excluded and the number of notebooks left after the filters.
-We also updated section 3.1.1 Collection of Jupyter Notebooks to explicitly mention the number of notebooks that were excluded after applying the filters and the number of notebooks in the final dataset.
-
-Following is an exerpt from the revised text of section 3.1.1 Collection of Jupyter Notebooks relevant to this comment.
-
-\begin{quote}
-  Since the focus of this study is on the analysis of Python code, [\ldots] 107 thousand notebooks were excluded after applying the above filters, leaving a dataset 190.8 thousand notebooks.
-\end{quote}
-
-% TODO: update the figure
+We updated section 3.1.1 Collection of Jupyter Notebooks to explicitly mention the number of notebooks that were excluded after applying the filters and the number of notebooks in the final dataset.
+
+Following is an excerpt from the revised text of section 3.1.1 Collection of Jupyter Notebooks relevant to this comment.
+
+\begin{quote}
+  Since the focus of this study is on the analysis of Python code, [\ldots] 107 thousand notebooks were excluded after applying the above filters, leaving a dataset of 190.8 thousand notebooks.
+\end{quote}
 
@@
 Since the data is really large and contains many outliers, we categorized the notebooks into small, medium and large based on the number of cells they contain.
-Additionally, we also added figures 4, 5 and 6 that visualize the distibution of the number of cells and feedback mechanisms across the notebook size categories.
+Additionally, we also added figures~\ref{fig:notebook-size-distribution},~\ref{fig:cell-per-notebook-size-distribution} and~\ref{fig:fms-per-notebook-size-distribution} to the manuscript, that visualize the distribution of the number of cells and feedback mechanisms across the notebook size categories.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
+  \caption{Number of notebooks in the small, medium and large notebook size categories.}
+  \label{fig:notebook-size-distribution}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
+  \caption{Distribution of number of cells in all notebook size categories.}
+  \label{fig:cell-per-notebook-size-distribution}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
+  \caption{Distribution of assertions, prints and last cell statements across all notebook size categories.}
+  \label{fig:fms-per-notebook-size-distribution}
+\end{figure}
 
@@
 This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity.
+  % TODO: we explored this idea but we observed that large notebooks could be (uninteresting) eg.
+notebook X while smaller...
+link this with SS (gives similar weightable to all FMs without making assumptions about complexity or other properties)
+
   Including this stratified analysis would significantly strengthen the study's contributions.
@@
 
+% TODO also add the benefits of using stratified sampling to ensure diversity in population of FMs; "while not perfect, we use SS to increase diversity of our candidates"
+% TODO add example of a really large notebook with simple FM and a small one with complex FM
 \Us Thank you for your feedback.
@@
 
-Please also see our response to comment 1.5 and the diff document for other revisions that were made to section 3.2 Exploratory Case Study Analysis.
-
-\begin{quote}
-  For RQ2 and RQ3, we allocated a fixed [\ldots].
-
+\begin{quote}
   The assertions, print statements and last cell statements are first tokenized---special characters and alphanumeric words shorter than two characters are removed.
@@
 As a result, all candidates characterized by rare tokens are grouped within a single bin.
-By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of overrepresenting those that are trivial yet rare.
+By randomly selecting candidates from each bin, we ensure that the final population reflects a diverse mix of candidates, mitigating the risk of over representing those that are trivial yet rare.
 \end{quote}
@@
 \begin{quote}
-  The findings of this study align with [\ldots].
-
   This study further reveals several avenues for future research.
@@
   Jupyter notebooks allow for out-of-order execution, which impacts the effectiveness of feedback mechanisms.
-A statement such as print("Accuracy:", accuracy) is only useful if the cell defining accuracy has been executed first.
+A statement such as print(``Accuracy:'', accuracy) is only useful if the cell defining accuracy has been executed first.
 The paper partially explore how such dependencies affect the functionality of feedback mechanisms, leaving a gap in the analysis.
@@
 
-% TODO: update this response once the section is finalized
 \begin{quote}
@@
 
-  Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells (Wang et al., 2020b).
-Conversely, constructs like \texttt{print("Accuracy:", accuracy)} are only effective if the cell defining accuracy has been executed prior to the print statement.
+  The assertions presented in Section 4.3 can be used to enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells (Wang et al., 2020b).
+Conversely, implicit feedback constructs like \texttt{print(``Accuracy:'', accuracy)} are only effective if the cell defining \texttt{accuracy} has been executed prior to the print statement.
 In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear.
@@
 
-\Us Thank you for your feedback.
-As per our original definition of feedback mechanisms from section 1 Introduction, all print statements are considered as feedback mechanisms in this study.
-Regardless, we have revised the definition of feedback mechanisms in the introduction and we hope that this is sufficient to address this comment.
+\Us Thank you for your feedback. As per our original definition of feedback mechanisms from section 1 Introduction, all print statements are considered as feedback mechanisms in this study.
+
+Regardless, we have revised the definition of feedback mechanisms in the introduction, clearly emphasized the difference between implicit and explicit feedback, and clarified how they differ from testing.
 
@@
 \begin{quote}
-  Computational notebooks---specifically Jupyter notebooks---have been ubiquitously adopted by the ML community for developing ML enabled systems (Pimentel et al., 2019; Quaranta et al., 2021; Psallidas et al., 2019; Perkel, 2018).
-Jupyter notebooks provide an interactive computing environment that allow users to break the development of complex workflows into smaller, more manageable code chunks.
-Each notebook is composed of a series of code cells, which can be executed independently in a read-eval-print loop (REPL) style of interactive software development.
-This REPL approach allows users to write a piece of code, run it, and immediately see the results thus facilitating an incremental and iterative style of software development.
-This modular structure of notebooks provides users the flexibility to experiment with different approaches and algorithms.
-Users can test hypotheses, debug issues, and make incremental changes with immediate feedback.
-The ability to visualize data and results inline and document the process in a narrative form further enhances their utility (Kery et al., 2018; Head et al., 2019; Rule et al., 2018; Chattopadhyay et al., 2020).
-
   Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues.
-Feedback mechanims are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program.
+\textbf{Feedback mechanisms are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program}.
 Importantly, not all forms of feedback quality as feedback mechanisms.
@@
   Notebook users can obtain feedback using implicit or explicit mechanisms.
-Implicit feedback mechanisms require manual validation by the user.
+\textbf{Implicit feedback mechanisms require manual validation by the user}.
 For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical.
@@
 
-  In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met.
+  In contrast, \textbf{explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met}.
 For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \texttt{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}.
@@
 Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality.
-While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm.
+\textbf{While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm}.
 For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook.
@@
 
-Following is an exerpt from section 5.2 Automated Data Validation added to address this comment.
-
-\begin{quote}
-  Breck et al. (2017) view ML training as analogous to [\ldots]
-
+Following is an excerpt from section 5.2 Automated Data Validation added to address this comment.
+
+\begin{quote}
   The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows.
@@
 Future work can focus on providing empirical evidence to demonstrate that the integration of assertions significantly improves the reliability of ML workflows in practice.
-
-  Results from RQ2 and RQ3 show that assertions, print statements and last cell statements are predominantly [\ldots]
 \end{quote}
@@
 \Us Thank you for your feedback.
-We agree with this point of view that as ML pipelines developed inside Jupyter notebooks mature, the implicit feedback machanisms should be converted to explicit feedback mechanisms as per software engineering best practices.
-Here is an exerpt from section 1 Introduction of the manuscript that motivates this recommendation.
+We agree with this point of view that as ML pipelines developed inside Jupyter notebooks mature, the implicit feedback mechanisms should be converted to explicit feedback mechanisms as per software engineering best practices.
+Here is an excerpt from section 1 Introduction of the manuscript that motivates this recommendation.
 
@@
 We also believe that existing feedback mechanisms in the notebooks (implicit and explicit) can make the transition from notebooks into a production environment smoother.
-In section 5.1 Documentation and Technical Debt we highlight how artifacts---such as requirements, rationale and design decisions and tests---which are a pre-requisite for all operational software systems, can be effectively derived from the feedback mechanisms used inside notebooks.
+In section 5.1 Documentation and Technical Debt we highlight how artifacts---such as requirements, rationale and design decisions and tests---which are a prerequisite for all operational software systems, can be effectively derived from the feedback mechanisms used inside notebooks.
 Furthermore, in section 5.2 Automated Data Validation, we propose that assertions can be used to derive the data validation schema for dedicated data validation tools, which are often incorporated into operational ML pipelines.
 
-Following are the relevant exerpts from section 5.1 Documentation and Technical Debt and 5.2 Automated Data Validation respectively that support the above statement.
-
-\begin{quote}
-  Jupyter notebooks are primarily used to explore different ideas or approaches [\ldots] In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks (Pimentel et al., 2019; Psallidas et al., 2019; Grotov et al., 2022).
+Following are the relevant excerpts from section 5.1 Documentation and Technical Debt and 5.2 Automated Data Validation respectively that support the above statement.
+
+\begin{quote}
+  Jupyter notebooks are primarily used to explore different ideas or approaches [\ldots] In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks (Pimentel et al., 2019; Psallidas et al., 2019; Grotov et al., 2022).
 
   Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks.
-The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
+The results demonstrate that practitioners utilize assertions, print statements, code cell outputs, and visualizations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
 Moreover, our results from RQ3 reveal [\ldots]
@@
 \Us Thank you for your feedback.
-We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information on the distribution of the number of cells and distribution of feedback mechanisms across various notebook size categories (small, medium and large notebooks).
-
-You may also see our response to comment 2.3 and the diff document for further details.
+We have added section 4.1 Descriptive Statistics to the manuscript which provides additional information regarding the distribution of the number of cells per notebook.
+Since the data is really large and contains many outliers, we categorized the notebooks into small, medium and large based on the number of cells they contain.
+Additionally, we also added figures~\ref{fig:notebook-size-distribution},~\ref{fig:cell-per-notebook-size-distribution} and~\ref{fig:fms-per-notebook-size-distribution} to the manuscript, that visualize the distribution of the number of cells and feedback mechanisms across the notebook size categories.
+
+Following is the text from section 4.1 Descriptive Statistics which was added to the manuscript to address this comment.
+
+\begin{quote}
+  Figure 4 shows the distribution of the number of code cells for all notebooks analyzed in this study.
+The distribution is grouped into three categories---namely small, medium and large---based on the first and third quartiles.
+This is done since the data is large, non-uniform and contains many outliers.
+Figure 5 further magnifies the distribution of the number of cells within each category.
+The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.
+
+  Figure 6 shows the distribution of assertions, prints and last cell statements across the different notebook size categories.
+A logarithmic scale is used for the count since the data is large and contains outliers.
+Overall, the visualization suggests an increase in the use of all feedback mechanisms as notebook size grows.
+Print statements are the most frequently used across all notebook sizes.
+\end{quote}
 
@@
 
-\Us Thank you for your feeback. We have revised section 1 Introduction to define feedback mechanims and discuss the similarities and differences between feedback mechanisms and formal testing.
+\Us Thank you for your feedback.
+We have revised section 1 Introduction to clearly define feedback mechanisms, emphasize the difference between implicit and explicit feedback, and clarify how they differ from formal testing.
 
@@
   Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues.
-Feedback mechanims are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program.
+Feedback mechanisms are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program.
 Importantly, not all forms of feedback quality as feedback mechanisms.
@@
 
-You may also see our response to comment 3.3 and the diff document for further details.
+You may also see our response to comment 3.3 above for further details.
 
@@
 
-After careful consideration, we belive that although the analysis in this study was conducted using Jupyter notebooks, the findings are applicable to other computational notebook technologies which follow the same REPL based paradigm of software development.
+After careful consideration, we believe that although the analysis in this study was conducted using Jupyter notebooks, the findings are applicable to other computational notebook technologies which follow the same REPL based paradigm of software development.
 We have therefore revised section 6 Threats to Validity (External Validity) to clarify the scope of this study.
 
-Following is the revised text from section 6 Threats to Validity (External Validity). We hope that this revision is sufficient to address the concern raised in this comment.
+Following is the revised text from section 6 Threats to Validity (External Validity).
 
@@
 Following is the text that was added to section 3.1.1 Collection of ML Jupyter Notebooks to address this comment.
-We hope that this addition is sufficient to address the concern raised by this comment.
-
-\begin{quote}
-  Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn (Pedregora et al., 2011), Pytorch (Paszke et al., 2017), Tensorflow (Abadi et al., 2015) and Keras (Chollet et al., 2015).
-The libraries were chosen based on the import analysis conducted by Psallidas at al., 2019, which analyzed six million Jupyter notebooks collected from GitHub.
-It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since they can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training.
-107 thousand notebooks were excluded after applying the above filters, leaving a dataset of 190.8 thousand notebooks.
+
+\begin{quote}
+  This study focuses on analyzing Python code in Jupyter notebooks with valid JSON structures containing at least one code cell.
+To concentrate specifically on machine learning projects, we only included notebooks that import at least one major ML library: Scikit Learn (Pedregora et al., 2011), PyTorch (Paszke et al., 2017), TensorFlow (Abadi et al., 2015), or Keras (Chollet et al., 2015).
+While this approach may exclude some ML projects using alternative libraries, research by Psallidas et al.
+(2019) examining six million GitHub Jupyter notebooks confirms that most ML-oriented notebooks utilize these frameworks.
+Despite their widespread usage, we deliberately excluded Pandas, NumPy, and SciPy from our selection criteria as these libraries serve numerous purposes beyond machine learning and are not specifically designed for model training.
+After applying these filters, we excluded 107 thousand notebooks, resulting in a final dataset of 190.8 thousand notebooks for analysis.
 \end{quote}
@@
 We renamed the title of section 3.2 Case Studies to \emph{Exploratory Case Study Analysis} to be verbose about the methodology.
-We added more details regarding \emph{site} and the \emph{selection criteria} of the study.
+We added more details regarding the \emph{site} and the \emph{selection criteria} of the study.
 Furthermore, details regarding the sampling technique, the bias introduced, and our mitigation strategy has also been added.
@@
 
-Following is the revised text from section 3.2 Exploratory Case Study Analysis. We sincerely hope that the revisions are sufficient to address the concerns raised by this comment.
+Following is the revised text from section 3.2 Exploratory Case Study Analysis where we addressed the concerns raised in this comment.
 
@@
 The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF).
+  % TODO: after introducing TF-IDF add why this is useful for us; it allows us to find not only common FMs but also specialized ones that address corner cases
   This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
@@
 
-During each case study, the first author categorized each candidate using \emph{inducting coding}.
-The first author analysed the Python code of the candidate to understand its purpose.
-Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analysed to bring in rich context.
-For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors.
+  During each case study, the first author categorized each candidate using \emph{inductive coding}.
+The first author analyzed the Python code of the candidate to understand its purpose.
+Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analyzed to bring in rich context.
+For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analyzed by the second and third authors.
 Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.
@@
 
+% TODO: our goal was not to create an extensive taxonomy of FMs, but to validate if there is something worth future research; this supports the use of a time budget for the analysis
+% TODO: in order to scale analysis to a generalizable population, we need automated tools; our methodology does not scale; smaller but insightful analysis; to scale we need to trade-off between deep insights and generalizability
 We have also revised section 6 Threats to Validity (Internal Validity) to discuss the pros and cons of using a fixed time resource.
-While we agree that the use of a fixed time resource has limitations, other stopping criterias, such as those based on theoretical saturation, are also subject to bias from the annotator.
+While we agree that the use of a fixed time resource has limitations, other stopping criteria, such as those based on theoretical saturation, are also subject to bias from the annotator.
 In contrast, a fixed time resource reflects the practical resource constraints that researchers face in the real world.
 
-Following is the text that was added to section 6 Threats to Validity (Internal Validity) to address this comment. We hope that the revised text is sufficient to address this comment.
+Following is the text that was added to section 6 Threats to Validity (Internal Validity) to address this comment.
 
@@
 \end{quote}
+
+We sincerely hope that the revisions are sufficient to address the concerns raised by this comment.
 
@@
   \begin{itemize}
-  \item The authors mention grouping candidates based on "their similarity to each other." How is similarity defined?
+    \item The authors mention grouping candidates based on ``their similarity to each other.'' How is similarity defined?
     \item The case study analysis process needs more detail.
@@
 \Us Thank you for your feedback.
-We have made several revisions to section 3.2 Exploratory Case Studies to address this comment.
+We have made several revisions to section 3.2 Exploratory Case Study Analysis to address this comment.
 In summary, we added more details regarding the exploratory case study analysis and how the first round of analysis was conducted by the first author to create the taxonomy of feedback mechanisms.
@@
   \end{itemize}
-
+\end{revcomment}
+
+\Us Thank you for your feedback.
+We have revised section 5.1 Documentation and Technical Debt to address this comment.
+Following is an excerpt from the revised section 5.1 Documentation and Technical Debt relevant for this comment.
+
+\begin{quote}
+  We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations.
+Visual cues and trends in visualizations are open to interpretation, meaning that different practitioners may interpret the same plot differently (Heer et al., 2010).
+By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
+Writing documentation from implicit feedback mechanisms and keeping the documentation up-to-date requires significant manual effort.
+Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation.
+We hope that the dataset contributed by this study provides a valuable resource to develop and evaluate such automated tools.
+\end{quote}
+
+\begin{revcomment}[3.12]
   \textbf{Automated Data Validation:}
@@
   \end{itemize}
+\end{revcomment}
+
+\Us Thank you for your feedback. We have revised section 5.2 Automated Data Validation by making the language clearer to address this comment. Here are the relevant excerpts from the revised section.
+
+\begin{quote}
+  The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows.
+Embedding assertions into the code helps ensure that any deviation from the expected shape of the training data, presence of missing data, and absence of critical resources is promptly flagged.
+This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines.
+
+  Assertions also serve as an executable form of documentation [\ldots]
+
+  Results from RQ2 and RQ3 show that assertions, print statements and last cell statements are predominantly used in the upstream, data-intensive stages of the MLDL.
+For instance, this study identifies several assertions that perform data validation checks to ensure data meets specific criteria or constraints.
+Similarly, we find assertions, print statements, and last cell statements that conduct type, shape, and distribution checks to safeguard against data inconsistencies.
+Based on these results, we argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
+\end{quote}
+
+\begin{revcomment}[3.13]
   \textbf{Machine Learning Bugs:}
@@
 
-\Us Thank you for your feedback. We have made several revisions to section 5 Discussions to address the comments above. We hope that the changes satisfy your expectations.
-
-\todos
+\Us Thank you for your feedback.
+We have revised Section 5.4 Machine Learning Bugs to address the concern raised by this comment.
+The primary intent of the section is to highlight how implicit and explicit feedback mechanisms identified in this paper can be connected to existing literature on ML faults and bugs.
+We additionally revised the section to also highlight future research opportunities.
+
+Following is the revised text from Section 5.4 Machine Learning Bugs
+
+\begin{quote}
+  The findings of this study align with prior research on real-world defects in ML systems.
+Humbatova et al.
+(2020) propose a taxonomy of faults in DL systems, while Morovati et al.
+(2023) curate \emph{Defects4ML}, a dataset consisting of ML bugs.
+Both studies identify defects arising from the use of incorrect data types and shapes, deprecated APIs, and missing GPU resources.
+Our study identifies several explicit assertions that can catch such data errors.
+Additionally, we find several implicit checks performed using print and last cell statements to verify the presence of compute resources and libraries, monitor model training, and check the architecture of neural networks.
+Future work should focus on mapping the feedback mechanisms in Jupyter notebooks to the ML defects identified by Humbatova et al.
+(2020) and Motovati et al.
+(2023) which can aid practitioners in debugging or preventing faults using the appropriate feedback technique.
+
+  This study further reveals several avenues for future research.
+We primarily focused on identifying feedback mechanisms used when writing ML code in Jupyter notebooks and subsequently exploring a limited subset of them.
+Further research is required to investigate these feedback mechanisms in greater depth and systematically organize them into a taxonomy grounded in specific ML validation tasks.
+It is important to recognize that machine learning workflows vary significantly across different tasks---such as classification, regression, and clustering.
+Each task necessitates distinct feedback mechanisms.
+For instance, assertions regarding image dimensions are pertinent in computer vision, whereas data distribution checks are critical in natural language processing.
+By analyzing the dataset according to the specific types of ML tasks, researchers can obtain targeted insights, thereby enhancing the practicality and applicability of recommendations for practitioners.
+
+\end{quote}
+
 \end{document}

### 2025-03-10 report/emse24-revised.tex
chore: update EMSE response document

@@
 Despite the critical role of feedback in machine learning engineering, no prior research has been conducted to identify and understand these mechanisms.
-To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyse 2.3 million code cells.
+To address this knowledge gap, we mine 297.8 thousand Jupyter notebooks and analyze 2.3 million code cells.
 We identify three key feedback mechanisms---assertions, print statements and last cell statements---and further categorize them into implicit and explicit forms of feedback.
@@
 MLDL is dominated by the model development phase which is highly experimental and iterative.
-This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analysing the results.
+This stage involves exploratory data analysis (EDA) to uncover underlying patterns, fitting the data to various models, and rigorously analyzing the results.
 Often, this stage may lead to additional rounds of feature engineering, where new attributes are created based on insights gained from initial models to enhance model performance.
@@
 
+Within this environment, feedback mechanisms are essential for improving code quality and diagnosing issues.
+\textbf{Feedback mechanisms are tools or techniques intentionally integrated into the code by the developer to gain insight into the execution of the program}.
+Importantly, not all forms of feedback quality as feedback mechanisms.
+For example, a bug in the code may produce feedback indicating an issue, but because the bug was not intentionally inserted, it does not count as a feedback mechanism.
+
 Notebook users can obtain feedback using implicit or explicit mechanisms.
-Implicit feedback mechanisms require manual validation by the user.
+\textbf{Implicit feedback mechanisms require manual validation by the user}.
 For instance, a common prerequisite for all ML models is that the features in the dataset are all numerical.
@@
 
-In contrast, explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met.
+In contrast, \textbf{explicit feedback mechanisms document the expected conditions of the code and halt execution if these conditions are not met}.
 For instance, the implicit validation method mentioned above can be replaced with an explicit assertion: \lstinline{assert all(df[col].dtype in ['int64', 'float64'] for col in df.columns), "Not all columns are of numerical type"}.
 This assert statement will stop the execution if the condition is not satisfied and provide immediate feedback with a clear message explaining the issue.
+
+It is important to contextualize feedback mechanisms within the broader framework of software testing.
+Traditional testing methodologies employ formal tests to explicitly identify discrepancies between expected and actual code functionality.
+\textbf{While all tests serve as feedback mechanisms, not all feedback mechanisms operate within this formal testing paradigm}.
+For instance, print statements and outputs from the last statement of code cells offer a more informal approach to gaining insights into the execution status and data flow within a Jupyter notebook.
+While such implicit feedback types can help practitioners diagnose issues within their code, they necessitate manual validation and interpretation, which can increase the potential for oversight and errors.
 
@@
 We focus on implicit form of feedback obtained from print statements and output generated by the last statement of code cells, as well as explicit form of feedback obtained from assertions.
+
 The research questions along with the contributions of this study are presented below.
@@
 	      We mine 297.8 thousand Jupyter notebooks written in Python from GitHub and Kaggle.
-We analyse 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.
+We analyze 2.3 million code cells obtained from the notebooks and identify three key feedback mechanisms used in ML notebooks---explicit feedback from assertions and implicit feedback from print and last cell statements.
 
@@
 	      To gain a deeper understanding of Python assertions written in Jupyter notebooks, we conduct individual case-studies with 82 assert statements.
-We analyse the assertion along with the surrounding code to understand its purpose.
-Additionally, we analyse the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.
+We analyze the assertion along with the surrounding code to understand its purpose.
+Additionally, we analyze the entire code cell, the previous and next cells along with the purpose of the notebook to bring in rich context.
 
@@
 	      Similarly, we conduct individual case-studies with 44 print statements and 27 last cell statements.
-During each case-study, we analyse the code statement along with the raw output to understand its purpose.
+During each case-study, we analyze the code statement along with the raw output to understand its purpose.
 \end{description}
@@
 Assertions are identified as a potential solution, capable of verifying data distributions and ensuring model performance metrics meet predefined thresholds.
-However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analysed incorporating them, suggesting a significant gap in current testing practices within the ML community.
+However, the study also notes the limited use of assertions, with only 8.5\% of the notebooks analyzed incorporating them, suggesting a significant gap in current testing practices within the ML community.
 The findings underscore the necessity for developing ML-specific testing tools that are deeply integrated into the ML development workflow to ensure consistent and reliable validation processes.
@@
 IPython---a more advanced alternative to the Python REPL---extends the capabilities of the standard Python shell with additional features such as powerful introspection, additional shell syntax, tab completion, and command history.
-Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extends its capabilities with a graphical user interface (GUI).
+Jupyter notebooks (and its successor Jupyter lab) leverage the IPython kernel to execute Python code and extend its capabilities with a graphical user interface (GUI).
 Resembling the literate programming paradigm~\citep{knuth1984literate}, Jupyter notebooks offer a web-based interface where users can create and share documents that contain live code, equations, visualizations, and narrative text.
@@
 
+% NOTE: above number extracted using the following shell command: `cat data-collection.log |grep '^OUTPUT'|cut -d: -f2|sed -e 's/-asserts\.csv//' -e 's/-prints\.csv//' -e 's/-lasts\.csv//'|sort|uniq|wc -l`
 \begin{figure}
@@
 
+\subsubsection{Collection of ML Jupyter Notebooks}
+
 We used GitHub's advanced search syntax\footnote{https://docs.github.com/en/search-github/searching-on-github/searching-code}, with the following search query: \lstinline[language={}]$language:"Jupyter Notebook"$ to mine 49.9 thousand Jupyter notebooks from public repositories on GitHub\footnote{Collected on June 22, 2023}.
@@
 
-Since the focus of this study is on the analysis of Python code, we only include notebooks that have a valid underlying JSON structure and contain at least one code cell.
-Finally, to focus on Python code written specifically for ML projects, we only include notebooks that import at least one of the following popular ML libraries---Scikit Learn~\citep{pedregosa2011scikit-learn}, Pytorch~\citep{paszke2017automatic}, Tensorflow~\citep{abadi2015tensorflow} and Keras~\citep{chollet2015keras}.
-The libraries were chosen based on the import analysis conducted by \citet{psallidas2019data}, which analyzed six million Jupyter notebooks collected from GitHub.
-It is important to note that we excluded libraries such as Pandas, NumPy, and SciPy, since can be utilized for a wide range of tasks beyond machine learning and do not specifically facilitate model training.
-
-We use the Python \lstinline{ast} module to extract the assert, print and last statements from all code cells.
-Prior to conducting the analysis for RQ2 and RQ3, we remove duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
-
-\subsection{Exploratory Case Studies}
+This study focuses on analyzing Python code in Jupyter notebooks with valid JSON structures containing at least one code cell.
+To concentrate specifically on machine learning projects, we only included notebooks that import at least one major ML library: Scikit Learn~\citep{pedregosa2011scikit-learn}, PyTorch~\citep{paszke2017automatic}, TensorFlow~\citep{abadi2015tensorflow}, or Keras~\citep{chollet2015keras}.
+While this approach may exclude some ML projects using alternative libraries, research by \citet{psallidas2019data} examining six million GitHub Jupyter notebooks confirms that most ML-oriented notebooks utilize these frameworks.
+Despite their widespread usage, we deliberately excluded Pandas, NumPy, and SciPy from our selection criteria as these libraries serve numerous purposes beyond machine learning and are not specifically designed for model training.
+After applying these filters, we excluded 107 thousand notebooks, resulting in a final dataset of 190.8 thousand notebooks for analysis.
+
+\subsubsection{Collection of Feedback Mechanisms}
+
+For RQ1, we extracted the contents of all code cells from the Jupyter notebooks by parsing the underlying JSON structure.
+We then used the \lstinline{ast} module provided by the Python programming language to extract the assert, print and last statements from all code cells.
+
+The raw contents of each cell were then converted into an AST.
+The \lstinline{NodeVisitor} class from the \lstinline{ast} module was used to traverse the AST and collect all \lstinline{Assert} nodes.
+To collect assertions written using Python modules not included in the standard distribution, we also collected all function \lstinline{Call} nodes that contain the string ``assert'' in their name.
+Similarly, all print statements were collected by traversing the AST to find all function \lstinline{Call} nodes with the name ``print''.
+To collect the last statements, we first removed all cells that did not produce any output and consequently extracted the last top-level node from the AST of the code cell.
+The raw Python code was regenerated from the ASTs using the \lstinline{unparse} method provided by the \lstinline{ast} module.
+
+Prior to conducting the analysis for RQ2 and RQ3, we removed duplicate data points resulting in a data set of 27.1 thousand assertions, 498.3 thousand print statements and 373.6 thousand last cell statements.
+
+\subsection{Exploratory Case Study Analysis}
 
 For RQ2 and RQ3, we allocated a fixed time resource of 200 hours to conduct the exploratory case study analysis of all candidate assertions, print statements and last cell statements.
-To surface interesting candidates for the case studies, we apply text processing techniques as described below.
+We specified Jupyter notebooks written in the Python programming language as the \emph{site} for the study.
+Given the exploratory nature of this study, we emphasized capturing a diverse range of candidates by apply text processing and sampling techniques as described below.
 
@@
 
-% TODO: update this paragraph, candidates were annotated by first and second author
-% we calculated IRA and obtained high agreement...
-During each case study, we analyse the code of the candidate to understand its purpose.
-Additionally, we analyse the entire code cell, the previous cell, next cell and the notebook's purpose to bring in rich context.
-For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analysed by the second and third authors.
-Further discussions were held with all authors until a consensus was reached.
+% NOTE: Cohen's kappa: 0.8083955317337074
+
+During each case study, the first author categorized each candidate using \emph{inductive coding}.
+The first author analyzed the Python code of the candidate to understand its purpose.
+Additionally, the entire code cell, the previous cell, next cell and the notebook's purpose was analyzed to bring in rich context.
+For case studies where the first author was unable to determine the purpose of the candidate, it was additionally analyzed by the second and third authors.
+Finally, the entire research team held regular meetings throughout this iterative process to ensure consistency in the categorization of the feedback mechanisms.
+
+A taxonomy of feedback mechanisms observed in ML Jupyter notebooks was created using the categories generated during the exploratory case study analysis.
+To validate the taxonomy developed by the first author, the second author independently annotated the candidates using the categories developed by the first author.
+The inter-rater agreement was calculated using the \emph{Cohen's Kappa} method and a score 0.8 was obtained which is regarded as highly agreeable.
 
@@
 	\includegraphics[width=0.5\textwidth]{notebook-size-distribution.pdf}
-  \caption{Number of notebooks in each category of...}
+	\caption{Number of notebooks in the small, medium and large notebook size categories.}
 	\label{fig:notebook-size-distribution}
@@
 	\includegraphics[width=0.75\textwidth]{cell-per-notebook-size-distribution.pdf}
-  \caption{Distribution of number of cells for various notebook sizes.}
-  \label{fig:cell-per-notebook-size-distribution.pdf}
+	\caption{Distribution of number of cells in all notebook size categories.}
+	\label{fig:cell-per-notebook-size-distribution}
 \end{figure}
@@
 	\includegraphics[width=0.5\textwidth]{fms-per-notebook-size-distribution.pdf}
-  \caption{Distribution of assertions, prints and last cell statements for various notebook sizes.}
+	\caption{Distribution of assertions, prints and last cell statements across all notebook size categories.}
 	\label{fig:fms-per-notebook-size-distribution}
@@
 
-% TODO should we add the descriptive stats as a table?
-% TODO do we need to extrapolate more from figure 5?
 Figure~\ref{fig:notebook-size-distribution} shows the distribution of the number of code cells for all notebooks analyzed in this study.
@@
 This is done since the data is large, non-uniform and contains many outliers.
-Figure~\ref{fig:cell-per-notebook-size-distribution.pdf} further magnifies the distribution of the number of cells within each category.
+Figure~\ref{fig:cell-per-notebook-size-distribution} further magnifies the distribution of the number of cells within each category.
 The dataset mostly contains medium size notebooks (9 to 26 code cells), followed by small (1 to 8 code cells) and large (27 to 772 code cells) size notebooks.
@@
 95\% of the 89.6 thousand assertions were written using the built-in Python \lstinline{assert} statement and 5\% using external testing libraries.
-Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accomodate for the stochastic nature of ML algorithms and comparing high-dimensional data structures.}
+Methods such as \lstinline{assert_almost_equal} and \lstinline{assert_array_equal} from the numpy and pandas libraries were frequently used to accommodate for the stochastic nature of ML algorithms and to compare high-dimensional data structures.}
 
@@
 		\toprule
-    \emph{\textbf{Comparison Check}}&
-    \emph{\textbf{Function Call}}\\
+		\emph{\textbf{Comparison Check}}              &
+		\emph{\textbf{Function Call}}                             \\
 		\midrule
 
-    \lstinline[]$type(train_dataset[0]) == tuple$&
-    \lstinline[]$(np.unique(y_test) == [-1, 1]).all()$\\
-
-    \lstinline[]$torch.__version__ > '1.10.0'$&
-    \lstinline[]$isinstance(transmat, pd.dataframe)$\\
-
-    \lstinline[]$len(x_dev) == len(y_dev)$&
-    \lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$\\
+		\lstinline[]$type(train_dataset[0]) == tuple$ &
+		\lstinline[]$(np.unique(y_test) == [-1, 1]).all()$        \\
+
+		\lstinline[]$torch.__version__ > '1.10.0'$    &
+		\lstinline[]$isinstance(transmat, pd.dataframe)$          \\
+
+		\lstinline[]$len(x_dev) == len(y_dev)$        &
+		\lstinline[]$np.allclose(np.linalg.norm(wio, axis=0), 1)$ \\
 		\bottomrule
@@
 Figure~\ref{fig:common-assert-msgs} shows the most frequent bi-grams that appear in these failure messages.
-Similar to the analysis presented above, we removed the duplicate messages prior to analysing the bi-grams resulting in 5.9 thousand unique messages.
+Similar to the analysis presented above, we removed the duplicate messages prior to analyzing the bi-grams resulting in 5.9 thousand unique messages.
 The bi-grams indicate that the shape and data type of data structures are frequently verified.
@@
 
-We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analysed in this study).
+We collected 1.4 million print statements from 180 thousand notebooks (60\% of the total notebooks analyzed in this study).
 Unlike assert statements, print statements do not have a fixed lexical semantic since the Python \lstinline{ast} module interprets them as regular function calls.
 Additionally, print statements can be written in several forms---the \lstinline{print} function may have multiple arguments, any form of string interpolation supported by Python or output from function calls.
-Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside of print statements are constants followed by function calls.
+Figure~\ref{fig:common-print-nodes} shows that the most common AST nodes inside print statements are constants followed by function calls.
 Figure~\ref{fig:common-print-constant-types} further validates that the constants are almost entirely of type string.
@@
 
-We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analysed in this study).
+We collected 1 million last cell statements from 119 thousand notebooks (40\% of the total notebooks analyzed in this study).
 Figure~\ref{fig:common-last-nodes} shows that \emph{Call} nodes frequently occur in the AST of last cell statements.
-Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} show the most common functions used inside these \emph{Call} nodes.
+Figure~\ref{fig:common-last-modules} shows the most common Python modules and objects while Figure~\ref{fig:common-last-functions} shows the most common functions used inside these \emph{Call} nodes.
 The figures highlight the prominence of data visualization libraries seaborn (\lstinline{sns}) and matplotlib (\lstinline{plt}) and visualization functions such as \lstinline{plot} and \lstinline{countplot} in the notebooks.
@@
 
-We further analysed the most common functions called on pandas dataframes by isolating the last cell statements that contain \lstinline{data} or \lstinline{df} objects.
+We further analyzed the most common functions called on pandas dataframes by isolating the last cell statements that contain \lstinline{data} or \lstinline{df} objects.
 This is shown in Figure~\ref{fig:common-last-df-functions} which reveals a focus on exploratory data analysis and statistical operations.
@@
 
-Similarly, we also analysed the most common visualization functions used in notebooks by isolating the last cell statements where the module is \lstinline{sns} or \lstinline{plt}.
+Similarly, we also analyzed the most common visualization functions used in notebooks by isolating the last cell statements where the module is \lstinline{sns} or \lstinline{plt}.
 This is shown in Figure~\ref{fig:common-last-visualization-functions} which indicates that categorical data visualization appears to be a primary focus, with \lstinline{countplot} and \lstinline{barplot} ranking high among the most frequently used functions.
@@
 
-\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings.
-These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}
+\highlight{\textbf{Observation 2.1} 26 assertions verify the dimensions of various data structures to ensure data consistency and compatibility throughout the ML pipeline, prevent runtime errors and ensure accurate model training and evaluation.}
 
@@
 
-\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values.
-These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
+\highlight{\textbf{Observation 2.2} Fourteen assertions perform validation checks on data structures to catch errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}
 
@@
 Other assertions ensure that the learned parameters align closely with theoretically or empirically derived values, further cementing the model’s statistical validity (\emph{A19}).
-Finally, assertions are also used to check for the expected number of neighbours as an indirect measure of model performance in specific scenarios like clustering (\emph{A7}).
+Finally, assertions are also used to check for the expected number of neighbors as an indirect measure of model performance in specific scenarios like clustering (\emph{A7}).
 
@@
 
-\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values.
-These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}
+\highlight{\textbf{Observation 2.4} Eight assertions ensure the presence of specific columns in the dataset and the absence of \texttt{NaN} or missing values.
+These checks prevent operations on empty datasets to maintain data integrity and reliability for model training.}
 
@@
 
-Type check assertions are used to maintain data integrity, ensure compatibility between different components of the ML pipeline, and preventing errors and unexpected behaviour.
+Type check assertions are used to maintain data integrity, ensure compatibility between different components of the ML pipeline, and preventing errors and unexpected behavior.
 Type checks are commonly performed to ensure that the input features are PyTorch float tensors or NumPy arrays, a requirement for many deep learning models and scientific computing operations respectively (\emph{A2, A88}).
@@
 
-\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features, models, and intermediate objects.
-These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}
+\highlight{\textbf{Observation 2.6} Five assertions validate the data types of input features and Python objects.
+These checks prevent errors from incompatible methods or attributes, and improve the reliability of the ML workflow.}
 
@@
 When working with neural networks and other statistical models, it is imperative to ensure that mathematical properties are preserved throughout operations involving arrays and matrices.
-This rigour is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalization (\emph{A25}).
+This rigor is captured though assertions that ensure convolution operations are dimensionally consistent (\emph{A3}), checking for the symmetry of matrices (\emph{A64}) and monitoring the standard deviation of outputs to decide the appropriateness of using batch normalization (\emph{A25}).
 These assertions are not just precautionary but are vital for confirming that the underpinnings of ML algorithms align with expected mathematical principles.
 
-\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs.
-These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry.
-These checks confirm the robustness and reliability of model computations.}
+\highlight{\textbf{Observation 2.7} Four assertions validate key mathematical properties of data and model outputs to confirm the robustness and reliability of model computations.}
 
@@
 
-\highlight{\textbf{Observation 2.8} This study identifies three assertions validating batch size configurations.
-These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size.
-These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
+\highlight{\textbf{Observation 2.8} Three assertions validate the batch size configurations.
+Such checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}
 
@@
 
-\highlight{\textbf{Observation 2.9} Three assertions verify the consistency of neural network architecture.
-These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected.
-These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}
+\highlight{\textbf{Observation 2.9} Three assertions verify that the neural network is correctly configured and integrated with pre-trained models to maintain the integrity of data flow during training.}
 
@@
 
-\highlight{\textbf{Observation 2.10} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage.
-Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data.
-This practice maintains the integrity and validity of the model evaluation metrics.}
+\highlight{\textbf{Observation 2.10} One assertion ensures there are no overlapping values between the training and validation sets to prevent data leakage and subsequently overfitting during model training.}
 
@@
 
-\highlight{\textbf{Observation 3.1} The study identifies 33 implicit checks to monitor the performance of trained ML models.
-These include outputs that print key performance metrics, a normalized heatmap used to visualize the confusion matrix of a model's predictions and a custom function written for manual evaluation of model predictions on random input samples.
-These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}
+\highlight{\textbf{Observation 3.1} 33 implicit checks are used to monitor the performance of trained ML models.
+These checks are used throughout the iterative model development phase to compare different models or variations, adjust parameters, and refine features to improve performance.}
 
@@
 For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed.
-During the exploratory data analysis phase, visualizations and pandas dataframes are commonly used to assess the distribution of specific columns (\emph{L9}), helping identify features related to the target variable (\emph{L2, L14})and informing feature inclusion in model training.
+During the exploratory data analysis phase, visualizations and pandas dataframes are commonly used to assess the distribution of specific columns (\emph{L9}), helping identify features related to the target variable (\emph{L2, L14}) and informing feature inclusion in model training.
 Additionally, descriptive statistics are used post-transformation to manually verify the effectiveness of data transformation steps and ensuring that the data conforms to the expected format for further analysis (\emph{L48, L25}).
 
-\highlight{\textbf{Observation 3.2} The study identifies seven implicit data distribution checks, including visualizations like categorical plots, kernel density estimate plots and count plots, and statistical summaries from pandas dataframes.
-These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
+\highlight{\textbf{Observation 3.2} Seven implicit data distribution checks are used to make informed decisions about data transformations, ensuring data integrity and improving model training reliability during the exploratory data analysis phase.}
 
@@
 
-\highlight{\textbf{Observation 3.3} The study identifies seven implicit checks used to verify resource availability on the systems where the Jupyter notebooks are being executed.
-These checks include confirming the availability of compute resources like GPUs and TPUs, ensuring datasets or pre-trained models are successfully loaded, and verifying the presence of specific library versions.
-These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}
+\highlight{\textbf{Observation 3.3} Seven implicit checks are used to verify critical resources are available on the system.
+These checks can help ensure that the computational environment is correctly setup and prevent execution errors.}
 
@@
 
-\highlight{\textbf{Observation 3.4} This study finds five implicit spot checks performed at various stages of the MLDL.
-Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}
+\highlight{\textbf{Observation 3.4} Five implicit spot checks are performed throughout the various stages of the MLDL to verify data transformation and feature engineering steps meet expectations.}
 
@@
 
-\highlight{\textbf{Observation 3.5} The study identifies four implicit checks used to monitor model training progress.
-These checks provide real-time feedback on training loss and accuracy, validate model adherence to expected data patterns, and identify optimal parameters through tuning methods.
+\highlight{\textbf{Observation 3.5} Four implicit checks are used to monitor the model training progress.
 These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}
@@
 
-\highlight{\textbf{Observation 3.6} The study identifies three implicit checks for missing data.
-Missing values can lead to biased or incorrect conclusions as many ML algorithms require complete numerical datasets for accurate calculations.
-Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}
+\highlight{\textbf{Observation 3.6} Three implicit checks are used to identify and address missing data during preprocessing which is essential to avoid errors and improve model reliability.}
 
@@
 
-\highlight{\textbf{Observation 3.7} The study identifies three implicit checks used to validate the shape of data.
+\highlight{\textbf{Observation 3.7} Three implicit checks are used to validate the shape of data.
 These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}
@@
 		\toprule
-    \emph{\textbf{Key}}&
-    \emph{\textbf{Code}}&
-    \emph{\textbf{Output}}\\
+		\emph{\textbf{Key}}                                                                                                   &
+		\emph{\textbf{Code}}                                                                                                  &
+		\emph{\textbf{Output}}                                                                                                  \\
 		\midrule
 
-    L6&
-    \lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$&
-    \includegraphics[width=\linewidth]{linear-relation-check-lineplot.png}\\
-
-    L10&
-    \lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$&
-    \includegraphics[width=\linewidth]{linear-relation-check-regplot.png}\\
+		L6                                                                                                                    &
+		\lstinline[]$b = sns.relplot(x='SIZE', y='Cash', hue='CLARITY', alpha=0.9, palette='muted', height=8, data=raw_data)$ &
+		\includegraphics[width=\linewidth]{linear-relation-check-lineplot.png}                                                  \\
+
+		L10                                                                                                                   &
+		\lstinline[]$sns.regplot(x='X4 number of convenience stores', y='Y house price of unit area', data=data)$             &
+		\includegraphics[width=\linewidth]{linear-relation-check-regplot.png}                                                   \\
 		\bottomrule
@@
 
-\highlight{\textbf{Observation 3.8} The study identifies two last cell statements used to verify linear relationships between features in the data.
-Visualizations such as a scatter plot with a regression line and a line plot are employed to assess the linearity between features.
-These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}
+\highlight{\textbf{Observation 3.8} Two last cell statements that produce visualizations are used to verify linear relationships between features in the data.
+Such checks help identify and remove redundant features ensure that the training dataset is optimized for linear ML models.}
 
@@
 
-\highlight{\textbf{Observation 3.9} The study identifies two implicit checks used to verify the data types of features in the dataset.
-These checks are used to determine the necessary preprocessing or transformation steps required to make the data suitable for training.
-Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}
+\highlight{\textbf{Observation 3.9} Two implicit checks are used to verify the data types of features in the dataset.
+These checks inform the necessary transformation steps required to make the data suitable for training.
+Additionally, type checks ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which are fundamental to various ML models.}
 
@@
 
-\highlight{\textbf{Observation 3.10} The study identifies one print statement used to check the total execution time of training and cross-validating an ML model.
+\highlight{\textbf{Observation 3.10} One print statement was used to check the total execution time of training and cross-validating an ML model.
 The print statement can be used to evaluate the efficiency of model training, especially when multiple models are trained iteratively.}
@@
 
-\highlight{\textbf{Observation 3.11} The study identifies one print statement used to show the architecture of a neural network.
-Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}
+\highlight{\textbf{Observation 3.11} One print statement was used to check the architecture of a neural network which can facilitate validation, learning and documentation tasks.}
 
@@
 
-Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favour of rapid experimentation~\citep{kery2018story,rule2018exploration,pimentel2019large-scale}.
+Jupyter notebooks are primarily used to explore different ideas or approaches, often neglecting software development best practices in favor of rapid experimentation~\citep{kery2018story,rule2018exploration,pimentel2019large-scale}.
 Once the code is finalized, it is typically transferred into a script that is run in a production environment.
@@
 To effectively communicate the information present in the notebooks with clients, collaborators or teammates, the practitioners add annotations and structure to the notebook using markdown cells~\citep{kery2018story,rule2018exploration}.
-In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artefacts which are usually missing in the original notebooks~\citep{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
+In both scenarios, practitioners need to revisit the notebooks to obtain up-to-date and accurate requirements, their rationale and design decisions, and tests---artifacts which are usually missing in the original notebooks~\citep{pimentel2019large-scale,psallidas2019data,grotov2022large-scale}.
 
 Our findings indicate that structured information can be effectively derived from the feedback mechanisms available in Jupyter notebooks.
-The results demonstrate that practitioners utilise assertions, print statements, code cell outputs, and visualisations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
+The results demonstrate that practitioners utilize assertions, print statements, code cell outputs, and visualizations to inform data preprocessing techniques, detect erroneous assumptions, create new features to enhance model performance, and verify hardware and software dependencies.
 Moreover, our results from RQ3 reveal that implicit feedback from print and last cell statements are used in Jupyter notebooks to make design and implementation decisions throughout the machine learning development lifecycle.
@@
 
-% TODO: this paragraph needs work...
 We recommend that ML practitioners document insights derived from implicit feedback mechanisms, especially for visualizations.
@@
 By documenting these visual insights, practitioners can ensure that the knowledge encapsulated in the visualizations, along with its interpretation and the subsequent decisions made, is preserved and consistently applied throughout the ML lifecycle.
+Writing documentation from implicit feedback mechanisms and keeping the documentation up-to-date requires significant manual effort.
 Therein also lies an opportunity for notebook designers and developers to incorporate automated techniques for generating such documentation.
-The dataset contributed by this study provides a valuable resource to automatically generate documentaion from implicit and explicit feedback mechanisms written inside Jupyter notebooks~\cite{yang2021subtle}.
-
-% Although writing documentation from implicit and explicit feedback mechanisms still requires significant manual effort and resources, the dataset contributed by this study provides a valuable resource for ongoing efforts to automatically generate documentation from code written in Jupyter notebooks~\citep{yang2021subtle}.
+We hope that the dataset contributed by this study provides a valuable resource to develop and evaluate such automated tools.
 
@@
 
-Catching data errors are critical since ML models trained using dirty data will lead to poor and incorrect predictions.
+Catching data errors is critical since ML models trained using dirty data will lead to poor and incorrect predictions.
 This is a cause for concern since these models may be deployed in production and their output may be used by other services which will then also perform poorly.
@@
 The results from RQ2 indicate that assertions can be used to automatically validate critical assumptions in ML workflows.
-Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged.
+Embedding assertions into the code helps ensure that any deviation from the expected shape of the training data, presence of missing data, and absence of critical resources is promptly flagged.
 This reduces the reliance on manual verification thus enhancing the robustness and maintainability of ML pipelines.
+
 Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development.
@@
 Similarly, we find assertions, print statements, and last cell statements that conduct type, shape, and distribution checks to safeguard against data inconsistencies.
-These results show a missed opportunity in data validation tasks.
-We argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
+Based on these results, we argue that when integrating dedicated data validation tools into an operational ML pipeline, the existing feedback mechanisms in notebooks can be effectively used to write the data validation schema.
 
@@
 We argue that the focus must shift towards building reliable ML-enabled systems by incorporating decades of research from the software engineering community with ML.
-Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\citep{kastner2020teaching}.
+Our recommendation is evident in the recent developments within the software engineering research community that steer education and research towards addressing the specific challenges faced when developing ML-enabled software systems~\citep{kastner2020teaching,lanubile2021towards,lanubile2023teaching}.
 Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.
 
-The assertions analysed in RQ1 and RQ2 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
+The assertions analyzed in RQ1 and RQ2 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks.
 First, the assertions identified in this study lack the support of dedicated testing libraries.
@@
 This can lead to problems such as unclear error diagnostics, difficulty in debugging, and increased time to identify and resolve issues.
-\citet{vidoni2021evaluating} investigated the testing culture in open-source R packages and identified several anti-patterns that lead to the accumulation of testing technical debt.
+\citet{vidoni2021evaluating} investigated the testing culture in open-source R packages and identified several antipatterns that lead to the accumulation of testing technical debt.
 The dataset contributed by this study can be used by researchers to replicate the \citet{vidoni2021evaluating} study using assertions written in Python.
@@
 
-% TODO: why is it important to analyze FMs wrt complexity of notebooks?
-% TBH I do not agree with the reviewer's comment about this...
 Analyzing the feedback mechanisms with respect to the complexity of Jupyter notebooks remains a promising direction for future research.
@@
 These metrics however were primarily developed for code structured in an object-oriented programming paradigm.
-Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices~\cite{pimentel2019large-scale}.
+Consequently, the applicability of these metrics to code within notebooks is questionable, particularly given that research indicates notebooks frequently diverge from established software engineering best practices~\citep{pimentel2019large-scale}.
 
@@
 Each time a cell is run, it updates the state of the program within the kernel, leading to the potential for discrepancies if cells defining variables or functions are executed after cells that depend on them.
-Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments~\cite{pimentel2019large-scale,wang2020assessing}.
-
-Assertions may enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells~\citep{wang2020assessing}.
-Conversely, constructs like \lstinline{print("Accuracy:", accuracy)} are only effective if the cell defining \lstinline{accuracy} has been executed prior to the print statement.
+Such out-of-order execution may result in a situation where subsequent cells operate on outdated or undefined values, ultimately complicating the workflow and undermining the reproducibility of ML experiments~\citep{pimentel2019large-scale,wang2020assessing}.
+
+The assertions presented in Section~\ref{sec:result-explicit} can be used to enhance the reproducibility of ML experiments by ensuring that consistent validation checks are applied, irrespective of the execution order of the notebook cells~\citep{wang2020assessing}.
+Conversely, implicit feedback constructs like \lstinline{print("Accuracy:", accuracy)} are only effective if the cell defining \lstinline{accuracy} has been executed prior to the print statement.
 In this paper, we have partially examined how these dependencies influence the functionality of feedback mechanisms, highlighting the complexities involved in maintaining reproducibility in scenarios where cell execution order is not strictly linear.
 Further investigation in this area could provide important insights into optimizing the use of assertions and other feedback mechanisms in Jupyter notebooks, particularly in the context of diverse ML workflows.
-
-However, there exists some ambiguity regarding how assertions interact with out-of-order execution in Jupyter notebooks.
-On one hand, our findings indicate that assertions can effectively prevent the system from reaching a convoluted state that may arise due to such execution.
 
@@
 
-Furthermore, we would like to reiterate that the case studies presented in RQ2 and RQ3 serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
-
 \paragraph{\textbf{External Validity}} Our analysis is based on ML notebooks that are publicly available on platforms such as GitHub and Kaggle.
 As a result, the findings may not be generalizable to Python notebooks that do not center around ML applications.
-Furthermore, the applicability of the results is limited when it comes to notebooks authored in other statistical programming languages, such as R or Julia, or to proprietary notebooks hosted on different platforms.
+Furthermore, the applicability of the results is limited when it comes to notebooks authored in other statistical programming languages, such as R or Julia.
 
@@
 
+Finally, we would like to reiterate that the case studies presented in RQ2 and RQ3 serve to illustrate key themes, and no definitive conclusions should be drawn from individual instances.
 \section{Related Work}\label{sec:related}
@@
 
-\citet{humbatova2020taxonomy} developed a comprehensive taxonomy of faults in deep learning systems through the manual analysis of 1059 artefacts from GitHub and Stack Overflow, complemented by structured interviews with 20 researchers and practitioners.
+\citet{humbatova2020taxonomy} developed a comprehensive taxonomy of faults in deep learning systems through the manual analysis of 1059 artifacts from GitHub and Stack Overflow, complemented by structured interviews with 20 researchers and practitioners.
 The study categorized faults into five top-level categories, encompassing model, data, training, deployment, and infrastructure-related faults.
@@
 
-\citet{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analysing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
+\citet{psallidas2019data} provide an overview of the evolving landscape of computational notebooks by analyzing six million Python notebooks, two million enterprise Data Science (DS) pipelines, source code and metadata from over 900 releases of 12 important DS libraries.
 Their findings can be used by system builders for improving DS tools and also by DS practitioners to understand the current trends in technologies they should focus on.
@@
 Studies with human subjects have been conducted to gain a deeper understanding of the challenges faced by ML practitioners when developing ML models inside notebooks.
-The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyse the data or produce meaningful visualizations~\citep{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
+The results indicate that ML practitioners work in a highly iterative fashion, often experimenting with multiple strategies to analyze the data or produce meaningful visualizations~\citep{kandel2012enterprise,kery2018story,liu2019understanding,chattopadhyay2020whats}.
 Studies have also been conducted to understand how practitioners generate, evaluate and manage alternative hypothesis, visual designs, methods, tools, algorithms and data sources to arrive at the final implementation~\citep{liu2019understanding,kandel2012enterprise}.
@@
 Unlike prior work that focuses on standalone visual analytics tools, our study explores how visualizations within Jupyter notebooks are used by ML practitioners.
-We analyse the motivations behind creating these visualizations and the insights they provide, offering a nuanced understanding of their role in the iterative ML development process.
+We analyze the motivations behind creating these visualizations and the insights they provide, offering a nuanced understanding of their role in the iterative ML development process.
 
@@
 This study also underscores the necessity for better documentation practices to mitigate technical debt and ensure the reproducibility of ML experiments.
-Feedback mechanisms are used throughout the ML development workflow to achieve more reliable and efficent model development processes.
+Feedback mechanisms are used throughout the ML development workflow to achieve more reliable and efficient model development processes.
 Future work should focus on developing specialized tools and frameworks to support these practices, thereby advancing the field of ML engineering.
@@
 
-\paragraph{\textbf{Data Availability Statement}} The dataset created and used in this study along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) licence: https://doi.org/10.6084/m9.figshare.26372140.v1.
-The case study candidates analysed in RQ2 and RQ3 can be viewed in our online appendix: https://arumoy.me/shome2023notebook.
+\paragraph{\textbf{Data Availability Statement}} The dataset created and used in this study along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) license: https://doi.org/10.6084/m9.figshare.26372140.v1.
+The case study candidates analyzed in RQ2 and RQ3 can be viewed in our online appendix: https://arumoy.me/shome2023notebook.
 

### 2025-03-11 report/emse24-response.tex
chore: update emse24 revised documents

@@
 This approach would not only validate the robustness of the results but also provide more targeted recommendations for practitioners working with notebooks of varying complexity.
-  % TODO: we explored this idea but we observed that large notebooks could be (uninteresting) eg.
-notebook X while smaller...
-link this with SS (gives similar weightable to all FMs without making assumptions about complexity or other properties)
-
 Including this stratified analysis would significantly strengthen the study's contributions.
@@
 
-% TODO also add the benefits of using stratified sampling to ensure diversity in population of FMs; "while not perfect, we use SS to increase diversity of our candidates"
-% TODO add example of a really large notebook with simple FM and a small one with complex FM
-\Us Thank you for your feedback.
-We have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work.
+\Us Thank you for your feedback.
+We explored the idea of using notebook size (number of cells) as a proxy for complexity.
+However, we found that larger notebooks don't necessarily contain more complex feedback mechanisms, and smaller notebooks can sometimes contain sophisticated feedback constructs.
+For example, the largest notebook in our dataset contains over 500 simple print statements (see large-notebook.ipynb: \url{https://surfdrive.surf.nl/files/index.php/s/FDQkCFxQNAVZIcq}), while some compact notebooks implemented complex assertion patterns with custom validation logic (see small-notebook.ipynb: \url{https://surfdrive.surf.nl/files/index.php/s/FDQkCFxQNAVZIcq}).
+
+To avoid making assumptions about the relationship between notebook size and feedback mechanism complexity, we opted for stratified random sampling based on TF-IDF values.
+This approach ensures a diverse population of feedback mechanisms in our analysis, giving appropriate representation to both common patterns and specialized implementations regardless of the notebook size.
+
+Additionally, we have added section 5.6 Complexity of Notebooks that discusses the existing challenges of defining notebook complexity and its potential value to improve the depth of this study as future work.
 While we agree that analyzing the complexity of notebooks may lead to interesting patterns, it is however difficult to define notebook complexity.
@@
 We added more details regarding the \emph{site} and the \emph{selection criteria} of the study.
-Furthermore, details regarding the sampling technique, the bias introduced, and our mitigation strategy has also been added.
+Furthermore, details regarding the sampling technique using TF-IDF, the bias introduced, and our mitigation strategy using stratified random sampling has also been added.
+
 Finally, to validate the taxonomy of feedback mechanisms, the second author independently annotated all candidates using the codes that were generated during the first round of annotations.
@@
 The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF).
-  % TODO: after introducing TF-IDF add why this is useful for us; it allows us to find not only common FMs but also specialized ones that address corner cases
-  This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
+This technique allows us to surface both frequently occurring feedback mechanisms and those that are rarer in the dataset.
+
+  This normalization using IDF assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
 Consider for instance, the trivial print statement \texttt{print("hello, world!")}.
@@
 
-% TODO: our goal was not to create an extensive taxonomy of FMs, but to validate if there is something worth future research; this supports the use of a time budget for the analysis
-% TODO: in order to scale analysis to a generalizable population, we need automated tools; our methodology does not scale; smaller but insightful analysis; to scale we need to trade-off between deep insights and generalizability
 We have also revised section 6 Threats to Validity (Internal Validity) to discuss the pros and cons of using a fixed time resource.
 While we agree that the use of a fixed time resource has limitations, other stopping criteria, such as those based on theoretical saturation, are also subject to bias from the annotator.
-In contrast, a fixed time resource reflects the practical resource constraints that researchers face in the real world.
+In contrast to other methods, a fixed time resource reflects the practical resource constraints that researchers face in the real world.
+Furthermore, our goal for the study was not to create an extensive taxonomy of feedback mechanisms but to explore if this direction of research warrants future work, and the use of a time budget for the analysis also supports this goal.
 

### 2025-03-11 report/emse24-revised.tex
chore: update emse24 revised documents

@@
 The term frequency (TF) for the remaining tokens is calculated and then normalized using their inter-document frequency (IDF).
-This normalization process assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
+This technique allows us to surface both frequently occurring feedback mechanisms and those that are rarer in the dataset.
+
+The normalization using IDF assigns higher values to tokens that are less frequently encountered, which can lead to overemphasis of candidates that contain rare tokens.
 Consider for instance, the trivial print statement \lstinline{print("hello, world!")}.

### 2025-03-13 report/emse24-revised.tex
feat: add declaration section to paper

@@
 
-\paragraph{\textbf{Conflict of Interest}} The authors declare that they have no conflict of interest.
-
-\paragraph{\textbf{Data Availability Statement}} The dataset created and used in this study along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) license: https://doi.org/10.6084/m9.figshare.26372140.v1.
+\subsection{Funding}
+Open access funding provided by Delft University of Technology.
+
+\subsection{Informed Consent}
+Not applicable
+
+\subsection{Author Contributions}
+\begin{description}
+  \item[Arumoy Shome:] Conceptualization and study design. Data collection, processing and analysis. Primary author of manuscript.
+  \item[Luis Cruz:] Conceptualization and study design. Data analysis. Review and editing.
+  \item[Diomidis Spinellis:] Study design. Review and editing.
+  \item[Arie van Deursen:] Review and editing.
+\end{description}
+
+\subsection{Data Availability Statement}
+The dataset created and used in this study along with the replication package is shared publicly under the Creative Commons Attribution (CC-BY) license: https://doi.org/10.6084/m9.figshare.26372140.v1.
 The case study candidates analyzed in RQ2 and RQ3 can be viewed in our online appendix: https://arumoy.me/shome2023notebook.
+
+\subsection{Conflict of Interest}
+The authors declare that they have no conflict of interest.
+
+\subsection{Clinical Trial Number}
+Not applicable
 
